### 2015-08-12 posterior/model.tex
[posterior] initial commit
This is the current status of the notes, they are still rewritten so
they represent work in progress, especially the likelihood section

\documentclass[a4,12pt]{article}

\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[proportional]{libertine}
\usepackage[libertine]{newtxmath}

% Import the natbib package and sets a bibliography  and citation styles
\usepackage[numbers]{natbib}
\bibliographystyle{plainnat}
% \setcitestyle{numbers}

\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{paralist}
\usepackage{xcolor}

%%% math
\newcommand{\betap}{\ensuremath{\beta^\prime}}
\newcommand{\data}{\ensuremath{\mathcal{D}}}
\newcommand{\expin}{\ensuremath{\lbrace\mu_i, \sigma_i\rbrace}}
\DeclareMathOperator{\Dirichlet}{Dirichlet}
\newcommand{\given}[2]{\left(#1\, \middle| #2 \, \right)}
\newcommand{\gaussian}{\ensuremath{\mathcal{N}}}
\newcommand{\genbetapr}{\ensuremath{\mathrm{GenBetaPrime}}}
\newcommand{\likelihood}{\ensuremath{\mathcal{L}}}
\newcommand{\lumi}{\ensuremath{\ell_i}}
\newcommand{\Lumi}{\ensuremath{L_i}}
\newcommand{\real}{\ensuremath{{\rm real}}}
\newcommand{\virt}{\ensuremath{{\rm virt}}}
\newcommand{\Nreal}{\ensuremath{N_\real}}
\newcommand{\Nvirt}{\ensuremath{N_\virt}}
\newcommand{\poisson}{\ensuremath{\mathrm{Poisson}}}
\newcommand{\rmdx}[1]{\mbox{d} #1 \,} % differential
\newcommand{\vecalpha}{\ensuremath{\vec{\alpha}}}
\newcommand{\vecL}{\ensuremath{\vec{L}}}
\newcommand{\vecmu}{\ensuremath{\vec{\mu}}}
\newcommand{\vecnu}{\ensuremath{\vec{\nu}}}
\newcommand{\vecth}{\ensuremath{{\vec{\theta}}}}
\newcommand{\vecu}{\ensuremath{{\vec{u}}}}

%%% refs
\def \refeq#1{(\ref{eq:#1})}
\def \refsec#1{sec.~\ref{#1}}
\def \refSec#1{Sec.~\ref{#1}}
\def \refapp#1{app.~\ref{#1}}
\def \reffig#1{fig.~\ref{fig:#1}}

%%% comments
\newcommand{\todo}[1]{{\textsc{\color{red}#1}}}

%%% codes
\newcommand{\tardis}{TARDIS}

\title{Formulating the posterior for \tardis}
\author{Frederik Beaujean}

\begin{document}
\maketitle

\section{Input and Output}

\subsubsection*{What we are given}
\begin{itemize}
\item data $\data \sim \expin =$ set of input values from
  telescope for each frequency bin $i=1 \dots n_b$;
\item \tardis{} depends on the parameters $\vecth$ and
  produces $n_\real$ real photon packets that leave the supernova and
  $n_\virt$ virtual photon packets. When we bin the packets in the same
  frequency bins as \expin, the sum of luminosities in bin $i$ is
  $\lumi^\real$ and $\lumi^\virt$.
\end{itemize}
I try to be mostly consistent with standard notation in statistics and
denote random variables---eventually they will be integrated out---by
capital letters and numbers that are observed by lower case.

\subsubsection*{What we want}
The posterior distribution of $\vecth$ given the data $\data$,
$P(\vecth | \data)$, is given by  Bayes' theorem as
\begin{equation}
  \label{eq:Bayes-full}
  P\given{\vecth}{\data} = \frac{P\given{\data}{\vecth} P(\vecth)}{\int \rmdx{\vecth} P\given{\data}{\vecth} P(\vecth)} \,.
\end{equation}
Technically, we should write every propability (density) as
conditional on a specific statistical model $M$,
e.g. $P\given{\vecth}{\data, M}$.  At present, we are not interested
in comparing different models but rather in extracting the parameters
of one model. Thus we omit $M$ and can ignore the denominator (the
\emph{evidence}) in Bayes' theorem. All we need is to define the
likelihood $\likelihood(\vecth) \equiv P\given{\data}{\vecth}$ and the
prior $P(\vecth)$ in
\begin{equation}
  \label{eq:Bayes-simple}
  P\given{\vecth}{\data} \propto \likelihood(\vecth) P(\vecth) \,.
\end{equation}

% Notation: $\vecth=$ parameters of interest,

% The overall goal is to extract (or maximize) the posterior of
% $\vecth$, the parameters that describe the physics behind
% supernovae. Bayes' theorem is
% \begin{align}
%   \label{eq:post}
%   P(\vecth | \mathcal{D}) &\propto P(\mathcal{D} | \vecth) P_0(\vecth)
% \end{align}
\section{Likelihood}

\subsubsection*{Goals}
\begin{itemize}
\item take into account uncertainty due to the finite number of packets in the simulation
\item model the distribution of luminosities accurately
\end{itemize}

\subsection{Incorporating the observations}

Let \Lumi{} denote the predicted or true luminosity in bin $i$ and
$\vecL$ the ordered collection of luminosities. I assume from a
telescope, we get the probability of the data $\mathcal{D}$ given
$\vecth$ only as a function of $\vecL$ where the data are
implicit---we don't have them---and no correlation between bins is
given or assumed. The maximum-entropy distribution given only the
means and standard deviations $\expin, i=1\dots n_b$ is the normal or
Gaussian distribution, so we set
\begin{equation}
  \label{eq:like-L}
  P(\mathcal{D} | \vecL) = \prod_{i=1}^{n_b} \gaussian\given{\Lumi}{\mu_i, \sigma_i^2}.
\end{equation}

In principle, $\vecL(\vecth)$ is a deterministic function of
$\vecth$. But we are unable to calculate it as is; that's why we have
to estimate it from the \tardis{} simulation resulting in a
distribution $P\given{\vecL}{\vecth}$. To use Bayes'
theorem~\refeq{Bayes-full}, we need $P\given{\data}{\vecth}$ but we
only have $P\given{\data}{\vecL}$. The link between these three
densities is the law of total probability:
\begin{align}
  \label{eq:like-expanded}
  P\given{\data}{\vecth} &= \int \rmdx{\vecL} P\given{\vecL}{\vecth} P\given{\data}{\vecL}.
\end{align}
Note that in principle it should say $P\given{\data}{\vecL, \vecth}$
in \refeq{like-expanded} but in our case $\vecth$ doesn't appear
explicitly, so we can just plug in
$P\given{\data}{\vecL}$ from~\refeq{like-L}.

\subsection{Extracting information from \tardis}

To complete the specification of the likelihood, we need to formulate
$P\given{\vecL}{\vecth}$; i.e., what we learn about the luminosities
in each bin from \tardis. A simplified account of how \tardis{} is run
follows, then I will show how to phrase this in statistical terms.

\subsubsection{Running \tardis}
\todo{This is still rudimentary}
\begin{compactenum}[(I)]
  \item Adjust the temperature of an inner shell to approximately match
    the total luminosity inferred from the observations
  \item Choose a total number of real packets, and a number of virtual
    packets to send out on each scattering of a real packet.
  \item Run \tardis.
  \item Some of the real packets reenter the core and are thus
    unobservable. But $n^\real$ real and $n^\virt$ virtual packets
    leave the core.
  \item Each packet has a luminosity. The packets are put into the
    same $n_b$ bins as the observations such that
    \begin{align}
      \label{eq:bin-packets}
      n^\real = \sum_{i=1}^{n_b} n_i^\real
    \end{align}
  \item The sum of real luminosities in bin $i$ is $\lumi^\real$.
\end{compactenum}
I usually omit the superscript $\in \{\real, \virt\}$ on $n_i$,
$\lumi$, and dependent quantities but equations hold for both unless
noted otherwise.

\subsubsection{Statistical model}

Some preliminary assumptions:
\begin{compactenum}[(a)]
  \item We neglect the uncertainty on the temperature and consider it
    fixed. We might lift this in the future.
  \item \label{assum:large-n} We neglect the uncertainty on $n^\real$ and consider it
    fixed. This should be a good approximation because we make sure
    $n^\real \gg 1$.
  \item \label{assum:large-nb} There are many bins; i.e., $n_b \gg 1$. So the probability of
    a packet to end up in any one bin is $\ll 1$.
\end{compactenum}
The proper distribution to describe the allocation of $n$ packets in
$n_b$ bins is the \emph{multinomial}. But in the limit of assumptions
(\ref{assum:large-n}) and (\ref{assum:large-nb}), we can consider the
Poisson distribution as a very good approximation. This in turn allows
us to consider the bins separately such that
\begin{align}
  \label{eq:factor-L}
  P\given{\vecL}{\vecth} = \prod_i P\given{L_i}{\vecth}
\end{align}
and the $n_b$-dimensional integral in \refeq{like-expanded}
dramatically simplifies to $n_b$ one-dimensional integrals.

The two major sources of statistical uncertainty coming from the
simulation are the unknown distribution of luminosity for the packets
and the Poisson uncertainty on the number of packets $n_i$. That is,
if we repeat the \tardis{} simulation with a different random-number
seed, we would get a different number of packets in bin $i$. The goal
in the following is to properly account for these two uncertainties
from a \emph{single} simulation without running repeatedly.

Let $N_i$ be the number of packets in we \emph{could} get in bin $i$
(it is not the number of packets we \emph{did} see, that's $n_i$),
then the law of total probability gives
\begin{equation}
  \label{eq:lum}
  P(\Lumi | \vecth) = \sum_{N_i=0}^{\infty} P(\Lumi | N_i, \vecth) P(N_i | \vecth)
\end{equation}
The first term in \refeq{lum} gives the luminosity in bin $i$ given
$N_i$ packets, so $\Lumi = \sum_{j=1}^{N_i} \Lumi^j$ where $\Lumi^j$ is
the luminosity of packet $j$ in bin $i$. Obviously $P(\Lumi | 0,
\vecth) = \delta(\Lumi)$.

We thought hard and found it a fair assumption that the distribution
of $l_{ij}$ is actually independent of the frequency thus independent
of $i$ and $j$ and we can drop the subscripts for simplicity, $l_{ij}
\to r$. The reason is that each packet initially gets the same
luminosity and so far it seems the amount of luminosity it gains or
loses is independent of the frequency. The question then is what is
the functional form $P(r | \vecth) \equiv P(\Lumi | N_i=1, \vecth)
$. The distributions $P(\Lumi | N_i, \vecth)$ with $N_i > 1$ can be
determined from the case of $N_i=1$ using convolutions for small $N_i$
and the central limit theorem for $N_i \ge 15$.

An approximation of $P(r | \vecth)$---a histogram of the luminosities
of all packets that leave the supernova---is shown in
\reffig{hist}.
\begin{figure*}[h]
  \begin{center}
    \includegraphics[width=\textwidth]{fit}
  \end{center}
  \caption{
  }
  \label{fig:hist}
\end{figure*}
The prominent features are a long tail at low luminosity, a sharp
peak, and a clear cutoff at high luminosity.  After researching quite
a bit, I found a flexible unimodal univariate distribution that can
accomodate this skewed distribution, it is the generalized \betap{}
distribution that depends on the hyperparameters $\vecnu \equiv (a, s,
\alpha, \gamma, \beta)$
\begin{equation}
  \label{eq:betaprime}
  P(r | \vecth) = \genbetapr(r | \vecnu(\vecth)) \equiv \frac{1}{B(\alpha, \gamma)} \left|\frac{\beta}{s} \right| \left( \frac{r - a}{s} \right)^{\alpha \beta - 1} \left(1+\left( \frac{r - a}{s} \right)^{\beta} \right)^{-\alpha - \gamma} \,.
\end{equation}

Since we assume it is the same in all bins we can then use the entire
set of samples that depend on $\vecth$ to infer $\vecnu | \vecth$. For
simplicity, we could use a maximum-likelihood fit and simply plug in
the result neglecting the uncertainty on $\vecnu$. This should be OK
as we usually have many packets; i.e., $N \gg 1$.

With $P(r|\vecth)$ fixed, we can now look at sums luminosities. The
standard result is that if $r_1, r_2 \sim P(r|\vecth)$ then
\begin{equation}
  \label{eq:convolution}
  z_2 = r_1 + r_2 \sim \int \rmdx{r} P(r|\vecth) P(z_2 - r|\vecth) \,.
\end{equation}
Sums of $N_i$ terms are computed in identical fashion by convolution
of $P(r|\vecth)$ with the distribution of the sum with $N_i-1$
terms. For $N_i \gtrsim 15$, the central limit theorem applies and
\begin{equation}
  \label{eq:clt}
  P(\Lumi | N_i, \vecth) \approx \gaussian \left(\Lumi | N_i * E_{P(r|\vecth)}[r], \sqrt{N_i * V_{P(r|\vecth)}[r]} \right)
\end{equation}
Note that for the convolutions, we would have to use the approximate
\genbetapr{} distribution but for the mean and std. deviation in
\refeq{clt}, we should use the sample mean and std. deviation to
bypass the extra uncertainty induced by assuming $P(r | \vecth) =
\genbetapr(r | \vecnu(\vecth))$ and fixed $\vecnu$.

For $P(N_i | \vecth)$, the second term in \eqref{eq:lum}, we assume
$N_i$ follows a Poisson distribution whose rate parameter $\lambda_i =
\lambda_i(\vecth)$ has to be inferred from the simulation. We can use
the law of total probability
\begin{equation}
  P(N_i | \vecth) = \int \rmdx{\lambda_i} P(\lambda_i | \vecth) \poisson(N_i | \lambda_i, \vecth)
\end{equation}
\todo{postpone the integral over $\lambda$ then use result for
  compound Poisson distr. and finally integrate. Hopefully this can
  still all be done analytically} and Bayes' theorem
\begin{equation}
  P(\lambda_i | \vecth) \propto \poisson(n_i | \lambda_i) P_0(\lambda_i)
\end{equation}
If we choose Jeffrey's prior $P_0(\lambda_i) \propto
1/\sqrt{\lambda_i}$ then we get the analytical
result~\cite[Eq. 5]{Aggarwal:2011aa}
\begin{equation}
  P\given{N_i}{n_i} = \frac{n_i! [2 (N_i + n_i)]!}{4^{N_i} N_i! (2 n_i)! (N_i + n_i)!}
\end{equation}
where $n_i = n_i(\vecth)$ is the number of packets in bin $i$ actually
observed in the simulation at hand.  In essence, $P(N_i | \vecth)$ is
the posterior predictive distribution to have $N_i$ packets given that
the simulation spit out $n_i(\vecth)$ packages. So $n_i$ is known,
while $N_i$ is a random variable and we take into account its
uncertainty by summing over $N_i$ in \refeq{lum}. This is relevant for bins where $n_i$ is small. If we got
2 packets, it is not so unlikely we could have gotten one package. But
of course with only half the packages, the expectation for the sum of
luminosities changes significantly.

\section{Priors}

\subsection{Polychord}

The polychord program only performs sampling and integration on the
unit hypercube with coordinates $\vecu$. Furthermore, it assumes that
the prior in these coordinates is uniform
\begin{equation}
  \label{eq:hypercube-prior}
  P(\vecu) = 1.
\end{equation}
In each iteration, we have to return the log likelihood for given
$\vecu$ passed in by polychord.  As above, suppose $\vecth$ represents
the parameters that are fed into the tardis simulation, then we need
to find the transform $\vecth(\vecu)$. The prior transforms as
\begin{equation}
  \label{eq:par-trafo}
  P(\vecth) \rmdx{\vecth} =  P(\vecu) \rmdx{\vecu},
\end{equation}
so with the flat prior
\begin{equation}
  \label{eq:par-trafo-likelihood}
  P(\vecth) = \left| \frac{\partial \theta_i}{\partial u_j} \right|^{-1} .
\end{equation}
There are two options:
\begin{enumerate}
\item \label{direct-transform} We can find $\vecth(\vecu)$ such that
  the Jacobian on the right in \refeq{par-trafo-likelihood} is the
  prior we want. The we just compute $\log L(\vecth(\vecu))$ for every
  sample $\vecu$ polychord provides.
\item If \ref{direct-transform} is not possible, we need to divide
  out the Jacobian and multiply by the actual prior we want, so for
  polychord we pretend we have a uniform prior and an \emph{effective} likelihood
  \begin{equation}
    \label{eq:eff-likelihood}
    L'(\vecu) = L(\vecth(\vecu)) P(\vecth(\vecu)) \left| \frac{\partial \theta_i}{\partial u_j} \right|
  \end{equation}
\end{enumerate}
Nested sampling is most efficient if the prior is close to the
posterior as measured by the relative entropy or Kullback-Leibler
divergence. \todo{cite Skilling} We therefore prefer
option~\ref{direct-transform}. While it restricts us to simple forms
of priors: uniform, Gaussian, or Dirichlet, this, fortunately, is
enough for our purposes. The explicit transformation is given below,
separately for each block of parameters.

% We could try to work out the
% transformation $\vecth(\vecu)$ such that the transformed prior has the
% form we want but for the relative abundances I don't know how to do
% that. Hence we will keep it simple and pretend that our prior is
% uniform and move everything into the effective likelihood $L'$. This
% amounts to perhaps slightly less efficient sampling but more
% importantly we can correctly define the target density from samples
% are produced. The Jacobian determinant can be computed as the product
% over blocks of correlated parameters.

\subsection{Relative abundances}

\subsubsection{Transformation}

Suppose we have $K$ abundances subject to the constraints
\begin{align}
  \label{eq:unit-simplex}
  & 0 \leq \theta_k \leq 1, & \sum_{k}\theta_k = 1.
\end{align}
In other words, the abundances are defined on the unit simplex.  As
shown in \todo{cite ``Cruising the simplex''}, we can directly
transform $\vecu$ with $\dim \vecu = K$ to $\vecth$ as follows. First
create $K$ Gamma variates with the inverse-transform method, in python
code $\xi_k = \texttt{scipy.stats.gamma.ppf}(u_k, \alpha_k)$ with the
shape parameter $\alpha_k$ explained below. Then normalize the samples
as
\begin{equation}
  \label{eq:dirichlet-gamma}
  \theta_k = \frac{\xi_k}{\sum_{k'} \xi_{k'}}.
\end{equation}
such that
\begin{equation}
  \label{eq:dirichlet-prior}
  P(\vecth) = \Dirichlet(\vecth | \alpha)
\end{equation}

\subsubsection{Prior information}

The canonical choice of a prior for fractions such as relative
abundances is the Dirichlet
distribution~\cite[App.~B]{bishop_pattern_2006}
\begin{equation}
  \label{eq:dirichlet-def}
  \Dirichlet \given{\vecth}{\vecalpha} = \frac{\Gamma\left(\sum_k \alpha_k\right)}{\prod_k \Gamma(\alpha_k)} \prod_{k=1}^K \theta_k^{\alpha_k-1}
\end{equation}
and depends on $K$ parameters $\vecalpha$. How to choose $\alpha_k$?
The mode of $\theta_k$ is $\frac{\alpha_k - 1}{\sum_k (\alpha_k -
  1)}$, so we should choose $\alpha_k > 1$ if we want a well defined
mode. For prior ignorance, we want $\alpha_k$ identical. The canonical
choice of an uninformative distribution is a uniform distribution on
the unit simplex given by $\alpha_k=1 \, \forall \, k$~\todo{cite
  posterior predictive hints}.

Now suppose we want to include the available prior information. The
actual numbers come from table 5.1 of Wolfgang's thesis. He lists
constraints of the type
\begin{equation}
  \begin{aligned}[t]
    \label{eq:abundance-constraints}
    \mbox{C}  &< 12.5 \%,\\
    \mbox{Si} &>1 \%,\\
    1 &<\mbox{Si/S ratio} < 10 .
  \end{aligned}
\end{equation}
The most straightforward approach is to incorporate them as hard cuts
into the prior, leading to a new normalization. We will have to refine
this strategy if the regions of high posterior probability turn out to
be very close to the cuts. Formally, suppose $V$ is the subvolume of
the unit simplex in which all constraints are satistfied. Outside of
$V$, we set $P(\vecth) \equiv 0$. In the code, a sequence of
\texttt{if} statements incorporates the constraints. If any is
violated, the prior on the log scale is $-\infty$ and no tardis
simulation needs to be run. Our prior thus becomes
\begin{equation}
  \label{eq:abundances-prior}
  P(\vecth) = C_V^{-1} \mathbf{1}_V(\vecth) \Dirichlet\left(\vecmu(\vecth) \middle| \vecalpha=(10^{-4}, \dots, 10^{-4})\right) \boldsymbol{\theta}
\end{equation}
where the normalization constant $C_V$ is  determined from
\begin{equation}
  \begin{aligned}
    \label{eq:abundance-normalization}
    1 &= \int \rmdx{\vecth} P(\vecth)\\
    \Rightarrow C_V  &= \int_V \rmdx{\vecth} \Dirichlet\left(\vecmu(\vecth) \middle| \vecalpha=(10^{-4}, \dots, 10^{-4})\right)
  \end{aligned}
\end{equation}
and $\mathbf{1}_V$ is the indicator function on $V$; i.e.,
$\mathbf{1}_V(\vecth) = 1$ iff $\vecth \in V$.  If we keep the cuts
fixed, we only need to determine $C_V$ once. We can do that with good
precision without nested sampling by using the law of large numbers as
\begin{equation}
  \label{eq:abundances-large-numbers}
  C_V \approx \frac{1}{N} \sum_{i=1}^N \mathbf{1}_V(\vecth), \; \vecth \sim \Dirichlet\left(\vecmu(\vecth) \middle| \vecalpha=(10^{-4}, \dots, 10^{-4})\right)
\end{equation}
and drawing directly from $\Dirichlet(\vecalpha)$ using
\texttt{scipy.random.dirichlet}. However, we don't really need to
normalize our densities properly unless we compare different
models. In the near future, we will only infer the parameters of a
single model so there is no need to compute $C_V$.

\subsection{Other parameters}

For all parameters other than abundances we choose uniform priors to
start with. This is not necessarily noninformative but we should be a
proxy for a diffuse prior hoping that the likelihood is sharp enough
to constrain all parameters. In the end, we should study the prior
dependence by choosing some other distribution such as wide Gaussians
or Cauchys to see if it has a significant effect.

\bibliography{references}

\end{document}

### 2015-08-13 posterior/model.tex
[posterior] more structure thru subsections

@@
 
-\subsection{Extracting information from \tardis}
-
 To complete the specification of the likelihood, we need to formulate
@@
 
-\subsubsection{Running \tardis}
+\subsection{Running \tardis}
 \todo{This is still rudimentary}
@@
 
-\subsubsection{Statistical model}
+\subsection{Statistical model}
 
@@
 \end{equation}
+
+\subsubsection{Luminosity distributions}
+
 The first term in \refeq{lum} gives the luminosity in bin $i$ given
-$N_i$ packets, so $\Lumi = \sum_{j=1}^{N_i} \Lumi^j$ where $\Lumi^j$ is
-the luminosity of packet $j$ in bin $i$. Obviously $P(\Lumi | 0,
-\vecth) = \delta(\Lumi)$.
+$N_i$ packets, so $\Lumi = \sum_{j=1}^{N_i} \Lumi^j$ where $\Lumi^j$
+is the luminosity of packet $j$ in bin $i$. Obviously $P(\Lumi | 0,
+\vecth) = \delta(\Lumi)$. The crucial term is
+$P\given{L_i}{N_i=1,\vecth}$: it is the distribution of luminosity of
+a single packet in bin $i$ and determines all other terms with
+$N_i>1$.
 
@@
 
-With $P(r|\vecth)$ fixed, we can now look at sums luminosities. The
+With $P(r|\vecth)$ fixed, we can now look at sums of luminosities. The
 standard result is that if $r_1, r_2 \sim P(r|\vecth)$ then
@@
 
-For $P(N_i | \vecth)$, the second term in \eqref{eq:lum}, we assume
-$N_i$ follows a Poisson distribution whose rate parameter $\lambda_i =
-\lambda_i(\vecth)$ has to be inferred from the simulation. We can use
-the law of total probability
+\subsubsection{Poisson noise}
+
+Let us now consider $P(N_i | \vecth)$, the second term in
+\eqref{eq:lum}. As stated before, we assume the number of packets in
+the given simulation follows a Poisson distribution whose rate
+parameter $\lambda_i = \lambda_i(\vecth)$ has to be inferred from the
+observed $n_i$. We can use the law of total probability
 \begin{equation}

### 2015-08-20 posterior/model.tex
[posterior] Major overhaul of likelihood description
* discuss general and large-n separately
* add mathematica notebook to show what I tried, what worked and what
  didn't

@@
 \newcommand{\vecu}{\ensuremath{{\vec{u}}}}
+
+\DeclareMathOperator{\erf}{erf}
+\newcommand{\mui}{\mu_i}
+% \newcommand{\Ni}{N_i}
+% \newcommand{\ni}{n_i}
+\newcommand{\sigi}{\sigma_i^2}
+\DeclareMathOperator{\sgn}{sgn}
+\newcommand{\El}{E[l]}
+\newcommand{\Vl}{V[l]}
 
@@
 
+All involved integrals are (assumed) absolutely convergent so we can
+exchange the order of summation and integration, so we can combine
+\refeq{like-expanded} and \refeq{lum} and we have to compute for each bin
+\begin{align}
+  \label{eq:sum-integral}
+  \sum_{N_i=0}^{\infty} \int \rmdx{\Lumi} \gaussian\given{\Lumi}{\mu_i, \sigma_i^2}
+  P(\Lumi | N_i, \vecth) P(N_i | \vecth).
+\end{align}
+
 \subsubsection{Luminosity distributions}
 
-The first term in \refeq{lum} gives the luminosity in bin $i$ given
-$N_i$ packets, so $\Lumi = \sum_{j=1}^{N_i} \Lumi^j$ where $\Lumi^j$
-is the luminosity of packet $j$ in bin $i$. Obviously $P(\Lumi | 0,
-\vecth) = \delta(\Lumi)$. The crucial term is
-$P\given{L_i}{N_i=1,\vecth}$: it is the distribution of luminosity of
-a single packet in bin $i$ and determines all other terms with
-$N_i>1$.
+The first term in \refeq{lum}, $P(\Lumi | N_i, \vecth)$, gives the
+probability of luminosity $\Lumi$ in bin $i$ given $N_i$ packets, so
+$\Lumi = \sum_{j=1}^{N_i} \Lumi^j$ where $\Lumi^j$ is the luminosity
+of packet $j$ in bin $i$. Obviously $P(\Lumi | 0, \vecth) =
+\delta(\Lumi)$. The crucial term is
+\begin{align}
+  \label{eq:single-lum}
+  P\given{\Lumi}{1, \vecth} \equiv P\given{L_i}{N_i=1,\vecth};
+\end{align}
+it is the distribution of luminosity of a single packet in bin $i$ and
+determines all other terms with $N_i>1$ because $\forall j, L_i^j$ has
+the same distribution $P\given{L_i}{1,\vecth}$.
 
-We thought hard and found it a fair assumption that the distribution
-of $l_{ij}$ is actually independent of the frequency thus independent
-of $i$ and $j$ and we can drop the subscripts for simplicity, $l_{ij}
-\to r$. The reason is that each packet initially gets the same
-luminosity and so far it seems the amount of luminosity it gains or
-loses is independent of the frequency. The question then is what is
-the functional form $P(r | \vecth) \equiv P(\Lumi | N_i=1, \vecth)
-$. The distributions $P(\Lumi | N_i, \vecth)$ with $N_i > 1$ can be
-determined from the case of $N_i=1$ using convolutions for small $N_i$
-and the central limit theorem for $N_i \ge 15$.
+For the real packets, it seems that $P(\Lumi | 1, \vecth)$ is
+independent of the frequency bin $i$. A histogram is shown in
+\reffig{hist}.  The prominent features are a long tail at low
+luminosity, a sharp peak, and a clear cutoff at high luminosity.
 
-An approximation of $P(r | \vecth)$---a histogram of the luminosities
-of all packets that leave the supernova---is shown in
-\reffig{hist}.
 \begin{figure*}[h]
@@
   \end{center}
-  \caption{
-  }
+  \caption{Distribution of luminosities of real packets.}
   \label{fig:hist}
 \end{figure*}
-The prominent features are a long tail at low luminosity, a sharp
-peak, and a clear cutoff at high luminosity.  After researching quite
-a bit, I found a flexible unimodal univariate distribution that can
-accomodate this skewed distribution, it is the generalized \betap{}
-distribution that depends on the hyperparameters $\vecnu \equiv (a, s,
-\alpha, \gamma, \beta)$
-\begin{equation}
-  \label{eq:betaprime}
-  P(r | \vecth) = \genbetapr(r | \vecnu(\vecth)) \equiv \frac{1}{B(\alpha, \gamma)} \left|\frac{\beta}{s} \right| \left( \frac{r - a}{s} \right)^{\alpha \beta - 1} \left(1+\left( \frac{r - a}{s} \right)^{\beta} \right)^{-\alpha - \gamma} \,.
-\end{equation}
 
-Since we assume it is the same in all bins we can then use the entire
-set of samples that depend on $\vecth$ to infer $\vecnu | \vecth$. For
-simplicity, we could use a maximum-likelihood fit and simply plug in
-the result neglecting the uncertainty on $\vecnu$. This should be OK
-as we usually have many packets; i.e., $N \gg 1$.
+For virtual packets, the luminosity distribution seems to depend
+mildly on the frequency.  We do not know $P(\Lumi | 1, \vecth)$ before
+we do the simulation and we do not know its functional form
+either. But we want to minimize the assumptions on the form. Hence
+seek a nonparametric description of the samples. Obviously, if we
+observe $n_i=0$ or $n_i=1$, we cannot infer the distribution $P(\Lumi
+| 1, \vecth)$ from such little information. If we may assume that
+$P(\Lumi | 1, \vecth)$ varies only little with $i$, we may pool
+together luminosities $l_{ij}, j=1 \dots n_i$, of adjacent bins to
+form a handful of larger bins, say $10-100$, in which we assume the
+distribution is constant and we have enough packets, at least $\sum_i
+n_i \gtrsim 500$. The I propose we use Gaussian kernel density
+estimation (KDE) to obtain a smooth estimate of $P(\Lumi | 1, \vecth)$
+that we can evaluate at arbitray $\Lumi$. In python, the easiest way
+to do this is via \texttt{scipy.stats.gaussian\_kde}. The fastest to
+my knowledge is \texttt{figtree} but the difference will be mostly
+irrelevant unless there $\sum_i n_i = 10^5$ packets in a combination
+of bins.
 
-With $P(r|\vecth)$ fixed, we can now look at sums of luminosities. The
-standard result is that if $r_1, r_2 \sim P(r|\vecth)$ then
-\begin{equation}
-  \label{eq:convolution}
-  z_2 = r_1 + r_2 \sim \int \rmdx{r} P(r|\vecth) P(z_2 - r|\vecth) \,.
-\end{equation}
-Sums of $N_i$ terms are computed in identical fashion by convolution
-of $P(r|\vecth)$ with the distribution of the sum with $N_i-1$
-terms. For $N_i \gtrsim 15$, the central limit theorem applies and
-\begin{equation}
+Given the KDE estimate, we could compute the distribution for the sum
+of two variables as the convolution
+\begin{align}
+  \label{eq:sum-of-2}
+  P(\Lumi | N_i=2, \vecth) = \int \rmdx{l} P(\Lumi -l | N_i=1, \vecth) P(l | N_i=1, \vecth).
+\end{align}
+For $N_i =k$, we would repeat the procedure convolving $P(\cdot |
+N_i=k-1, \vecth)$ and $P(\Lumi | N_i=1, \vecth)$ and each time we
+would need a new KDE estimate and we would have to repeat the integral
+for different values of $\Lumi$. This quickly gets out of hand and I
+fear the errors due to KDE assumptions add up. In my attempts with
+\texttt{mathematica}, the probability for many values of $\Lumi$ was
+negative and the whole procedure is slow. Let's forget about it.
+
+The more model-independent and faster approach is to use the
+luminosity samples $\{l_j: j=1 \dots \sum n_i\}$ for all bins $i$ that
+are combined. We can create samples of the sum of $N_i=2$ luminosities
+by adding $l_1+l_2, l_3 + l_4 \dots$ and similarly for larger
+$N_i$.
+
+\subsubsection*{Approximation for large $n_i$}
+
+For $N \gtrsim 15$, we can just use the central limit theorem
+such that
+\begin{align}
   \label{eq:clt}
-  P(\Lumi | N_i, \vecth) \approx \gaussian \left(\Lumi | N_i * E_{P(r|\vecth)}[r], \sqrt{N_i * V_{P(r|\vecth)}[r]} \right)
-\end{equation}
-Note that for the convolutions, we would have to use the approximate
-\genbetapr{} distribution but for the mean and std. deviation in
-\refeq{clt}, we should use the sample mean and std. deviation to
-bypass the extra uncertainty induced by assuming $P(r | \vecth) =
-\genbetapr(r | \vecnu(\vecth))$ and fixed $\vecnu$.
+  P(\Lumi | N_i, \vecth) \approx \gaussian \given{\Lumi}{N_i \times E[l], N_i \times V[l]}
+\end{align}
+and the expecation value and variance have to be estimated by the mean
+and variance of the samples $\{l_j\}$. In fact, this paper~\cite{bohm_statistics_2014}
+shows that the \emph{scaled Poisson} is a more accurate approximation
+given the first and second moment. But the resulting integral can't be
+solved exactly, so for easier calculation we stick with the Gaussian.
 
@@
 parameter $\lambda_i = \lambda_i(\vecth)$ has to be inferred from the
-observed $n_i$. We can use the law of total probability
+observed $n_i = n_i(\vecth)$. We can use the law of total probability
 \begin{equation}
+  \label{eq:poisson-post-pred}
   P(N_i | \vecth) = \int \rmdx{\lambda_i} P(\lambda_i | \vecth) \poisson(N_i | \lambda_i, \vecth)
 \end{equation}
-\todo{postpone the integral over $\lambda$ then use result for
-  compound Poisson distr. and finally integrate. Hopefully this can
-  still all be done analytically} and Bayes' theorem
+and Bayes' theorem
 \begin{equation}
+  \label{eq:poisson-post}
   P(\lambda_i | \vecth) \propto \poisson(n_i | \lambda_i) P_0(\lambda_i)
 \end{equation}
-If we choose Jeffrey's prior $P_0(\lambda_i) \propto
-1/\sqrt{\lambda_i}$ then we get the analytical
+If we choose the noninformative and improper Jeffreys prior
+$P_0(\lambda_i) \propto 1/\sqrt{\lambda_i}$ then we get the analytical
 result~\cite[Eq. 5]{Aggarwal:2011aa}
-\begin{equation}
-  P\given{N_i}{n_i} = \frac{n_i! [2 (N_i + n_i)]!}{4^{N_i} N_i! (2 n_i)! (N_i + n_i)!}
-\end{equation}
-where $n_i = n_i(\vecth)$ is the number of packets in bin $i$ actually
-observed in the simulation at hand.  In essence, $P(N_i | \vecth)$ is
+\begin{align}
+  \label{eq:post-pred-analytical}
+  P\given{N_i}{\vecth} &= \frac{n_i! [2 (N_i + n_i)]!}{2^{N_i+n_i+1/2}4^{N_i} N_i! (2 n_i)! (N_i + n_i)!}\\
+  &= \frac{(N_i + n_i + 1)_{N_i + n_i}}{(n_i + 1)_{n_i} \sqrt{2}\, 2^{3N_i + n_i}N_i!}
+\end{align}
+where
+\begin{align}
+  \label{eq:pochhammer}
+  (a)_k \equiv a (a + 1) \dots (a + k - 1)
+\end{align}
+is the Pochhammer symbol.
+
+In essence, $P(N_i | \vecth)$ is
 the posterior predictive distribution to have $N_i$ packets given that
@@
 luminosities changes significantly.
+
+\subsubsection*{Approximation for large $n_i$}
+
+The mode and variance of $P(N_i | \vecth)$ are
+\begin{align}
+  \label{eq:poisson-mode-variance}
+  \arg \max P(N_i | \vecth) = n_i -1, && V[N_i] = 2 n_i + 1.
+\end{align}
+Why is the variance $\sim 2 n_i$ and not $n_i$ as one would expect for Poisson?
+Variances are additive and we get one factor of $n_i$ from the Poisson in \refeq{poisson-post-pred} and another one from \refeq{poisson-post}.
+
+For $n_i=0$, the distribution diverges at zero, a consequence of the
+Jeffreys prior.  For $n_i \gtrsim 30$, the distribution can be well
+approximated by a Gaussian
+\begin{align}
+  \label{eq:poisson-gauss-approx}
+  P(N_i | \vecth) \approx \gaussian \given{N_i}{n_i-1, 2 n_i + 1}
+\end{align}
+
+\subsection{Putting it all together}
+
+Now it is time to combine all the equations above to see what actually
+needs to be implemented. The likelihood is
+\begin{align}
+  \label{eq:all-likelihood}
+  P\given{\data}{\vecth} &= \prod_{i=1}^{N_b} \int \rmdx{\Lumi} P\given{\Lumi}{\vecth}   P\given{\data}{\Lumi}\\
+  &= \prod_{i=0}^{N_b} \sum_{N_i=0}^\infty \int \rmdx{\Lumi} P \given{\Lumi}{N_i}
+  P\given{N_i}{\vecth} P\given{\Lumi}{\mui, \sigi}
+\end{align}
+
+\subsubsection*{Large $n_i$}
+
+If many packets are observed in bin $i$, then with good approximation
+everything is Gaussian and we can evaluate the integral over $\Lumi$
+analytically: the convolution of two Gaussians is another Gaussian
+with variances added.
+
+\begin{align}
+  \label{eq:all-large-n}
+P\given{\data}{\vecth}  &=  \prod_{i=1}^{N_b} \sum_{N_i=0}^\infty \int \rmdx{\Lumi} \gaussian \given{N_i}{n_i-1, 2 n_i + 1} \gaussian \given{\Lumi}{N_i  E[l], N_i V[l]}\\
+  & \phantom{\prod_{i=1}^{N_b} \sum_{N_i=0}^\infty \int \rmdx{\Lumi}} \times \gaussian \given{\Lumi}{\mu_i, \sigma_i} \nonumber\\
+  & = \prod_{i=1}^{N_b} \sum_{N_i=0}^\infty \gaussian \given{N_i}{n_i-1, 2 n_i + 1} \gaussian \given{N_i \El}{\mu_i, \sigi + N_i \Vl}
+  % &= \frac{\left| \Vl \mui + \El \sigi \right|+ (\Vl \mui + \El \sigi) \erf\left(\frac{\sqrt{N_i} \left| \Vl \mui + \El \sigi \right|}{\sqrt{2 \sigi \Vl (\Vl N_i + \sigi)}}\right)}{}\\
+%   &= \frac{1 + \sgn(\rho) \erf\left(\frac{\sqrt{N_i} \left| \rho \right|}{\sqrt{2 \sigi \Vl \xi}}\right)}{4 \pi \sqrt{(2n_i + 1) \xi}} \exp \left(-\frac{ (N_i \El -\mui)^2}{2  \xi}\right)
+% \\
+%   &\phantom{=} \times \exp \left(\frac{(2n_i+1) (N_i \El)^2 + N_i^3 \Vl - 2(2n_i+1)N_i \El \mui + (2n_i+1) \mu^2}{2 (2n_i + 1) \xi}\right) \nonumber\\
+%   &\phantom{=} \times \exp \left(\frac{(n_i-1)^2 \sigi + (n_i - 1) N_i ((n_i-1)\Vl - 2\sigi) + N_i^2 (-2(n-1)\Vl + \sigi)}{2 (2n_i + 1) \xi}\right) \nonumber
+\end{align}
+% where
+% \begin{align}
+%   \label{eq:rho}
+%   \rho &\equiv  \Vl \mui + \El \sigi\\
+%   \xi &\equiv N_i \Vl  + \sigi
+% \end{align}
+
+For each bin, we have to compute a series. We have three options:
+\begin{enumerate}
+  \item The quick and dirty approach is to approximate $\sum_{N_i}$ by
+  $\int \rmdx{N_i}$ and use \texttt{scipy.integrate.quad} to
+  numerically integrate.
+  \item If the first approach turns out to be too slow, we can just
+  sum up a finite number of terms as the summand quickly falls off for
+  large $N_i$; cf. \reffig{series}. The summand/integrand is unimodal
+  and well behaved. For the numbers I tried, the mode was near $n_i$
+  as implied by the first Gaussian. The mode can be computed
+  analytically but the expression is very long, so I don't show it
+  here. We could start summing from there and grow to the left and
+  right until terms become negligibly small. In my tests in
+  mathematica, I chose the number of terms by eye.
+  \item The third and fastest way would be to compute the mode, then
+  use the Laplace approximation to estimate the integral. This
+  requires the 2nd derivative around the mode, which is available in
+  closed form but too long to report here. Check the mathematica
+  notebook for the explicit results.
+\end{enumerate}
+For tests I did, the first two agreed within 5 or 6 digits, and the
+Laplace approximation good to the 3rd digit, so we might want to go
+with that, as it should be the fastest.
+
+For $N_i \to 0$, the summand is independent of $\El$ and $\Vl$ and
+very small if $n_i$ is large, so there is fortunately no special case
+to take care of.
+
+\begin{figure}[h]
+  \centering
+  \includegraphics{series}
+  \caption{Series summand $\gaussian \given{N_i}{n_i-1, 2 n_i + 1}
+    \gaussian \given{N_i \El}{\mu_i, \sigi + N_i \Vl}$ of
+    \refeq{all-large-n} for various values of $N_i$ and $\mui=5,
+    \sigi=5^2,n_i=15, \El= 0.9 * \mui / n_i, \Vl=0.1$.}
+  \label{fig:series}
+\end{figure}
+
+\subsubsection*{General case}
+
+Plugging in the KDE estimates of $P\given{\Lumi}{N_i}$, we cannot use
+the Laplace approximation because the mode and the derivative there
+are not known. For small $n_i$, typically few terms in the series
+contribute, so the number of numerical integrals can be kept
+small. However, this depends on the experimental uncertainty. If it is
+large, then we have to take into account more terms.
+
+It is worth noting that for $N_i=0$, $\Lumi$ would be the sum of zero
+terms and that is ill defined. So we have to start summing at
+$N_i=1$. But that does not mean that $n_i=0$ is invalid, too. In fact,
+$P(N_i=1 | n_i=0) = 17.7\%$ in our model with the Jeffreys prior. So
+no packets seen means you could have seen a packet with some sizable
+probability. Of course this is a simplification. Looking at a spectrum
+in which the last 1000 bins didn't get a packet and the luminosity
+drops before that, I associate a much smaller probability for $N_i=1$
+in the very last bin than in bin $N_b-1000$. The difference is because
+I take into account extra information: smoothness of the spectrum,
+monotonicity \dots. At present, I don't know how to encode this in the
+model \emph{and} keep it computationally tractable. It is left as an
+exercise for the reader.
+
+\subsubsection*{Recommendation}
+
+The strategy would be to look at each bin separately. If $n_i \gtrsim
+5$, use the large-$n_i$ approach from above. Else start at $N_i=1$ and
+sum the numerical integrals until it is $10^{-4}$ of the maximum
+term. For 10000 bins, this works can be parallelized easily.
+
+For the real packets, I compared the general approach to the Gaussian
+large-$n_i$ approximation, using numerical integration in both cases
+in \texttt{mathematica}. For $n_i=1\dots5$, the difference is slighly
+less than 10\%. From $n_i=10$ to 15, it drops from 5\% to 3\% level,
+and for $n_i=100$, both approaches agree to within $10^{-4}$. Laplace
+can be used from $n_i=5$ on. In all cases except $n_i=100$, the
+Gaussian approximation is systematically \emph{below} the general
+result.
+
+If we are ok with less than 10\% error, we can employ the Gaussian
+approximation from $n_i\ge5$, which means we neglect the 3rd and
+higher moments of the luminosity distribution. If the virtual packets
+are even more skewed than the real packets, we might need to start at
+$n_i \sim 10$.
 
@@
 posterior as measured by the relative entropy or Kullback-Leibler
-divergence. \todo{cite Skilling} We therefore prefer
+divergence~\cite{skilling_nested_2006}. We therefore prefer
 option~\ref{direct-transform}. While it restricts us to simple forms
@@
 In other words, the abundances are defined on the unit simplex.  As
-shown in \todo{cite ``Cruising the simplex''}, we can directly
-transform $\vecu$ with $\dim \vecu = K$ to $\vecth$ as follows. First
-create $K$ Gamma variates with the inverse-transform method, in python
-code $\xi_k = \texttt{scipy.stats.gamma.ppf}(u_k, \alpha_k)$ with the
-shape parameter $\alpha_k$ explained below. Then normalize the samples
-as
+shown in~\cite{betancourt_cruising_2012}, we can directly transform
+$\vecu$ with $\dim \vecu = K$ to $\vecth$ as follows. First create $K$
+Gamma variates with the inverse-transform method, in python code
+$\xi_k = \texttt{scipy.stats.gamma.ppf}(u_k, \alpha_k)$ with the shape
+parameter $\alpha_k$ explained below. Then normalize the samples as
 \begin{equation}
@@
 choice of an uninformative distribution is a uniform distribution on
-the unit simplex given by $\alpha_k=1 \, \forall \, k$~\todo{cite
-  posterior predictive hints}.
+the unit simplex given by $\alpha_k=1 \, \forall \,
+k$~\cite{tuyl_posterior_2009}.
 

### 2015-08-20 posterior/model.tex
[posterior] Fix spelling

@@
 \end{equation}
-Technically, we should write every propability (density) as
+Technically, we should write every probability (density) as
 conditional on a specific statistical model $M$,
@@
 we do the simulation and we do not know its functional form
-either. But we want to minimize the assumptions on the form. Hence
+either. But we want to minimize the assumptions on the form. Hence we
 seek a nonparametric description of the samples. Obviously, if we
@@
 estimation (KDE) to obtain a smooth estimate of $P(\Lumi | 1, \vecth)$
-that we can evaluate at arbitray $\Lumi$. In python, the easiest way
+that we can evaluate at arbitrary $\Lumi$. In python, the easiest way
 to do this is via \texttt{scipy.stats.gaussian\_kde}. The fastest to
@@
 \end{align}
-For $N_i =k$, we would repeat the procedure convolving $P(\cdot |
+For $N_i =k$, we would repeat the procedure convoluting $P(\cdot |
 N_i=k-1, \vecth)$ and $P(\Lumi | N_i=1, \vecth)$ and each time we
@@
 \end{align}
-and the expecation value and variance have to be estimated by the mean
+and the expectation value and variance have to be estimated by the mean
 and variance of the samples $\{l_j\}$. In fact, this paper~\cite{bohm_statistics_2014}
@@
 
-For $n_i=0$, the distribution diverges at zero, a consequence of the
-Jeffreys prior.  For $n_i \gtrsim 30$, the distribution can be well
-approximated by a Gaussian
+For $n_i=0$, the distribution peaks at zero at a value of $70\%$, a
+number sensitive to the choice of prior $P(\lambda_i)$.  For $n_i \gtrsim 30$, the
+distribution can be well approximated by a Gaussian
 \begin{align}
@@
 sum the numerical integrals until it is $10^{-4}$ of the maximum
-term. For 10000 bins, this works can be parallelized easily.
+term. For 10000 bins, this work can be parallelized easily. It's all
+elementary function, so this can be done in \texttt{C} without too
+much effort.
 
@@
 large-$n_i$ approximation, using numerical integration in both cases
-in \texttt{mathematica}. For $n_i=1\dots5$, the difference is slighly
+in \texttt{mathematica}. For $n_i=1\dots5$, the difference is slightly
 less than 10\%. From $n_i=10$ to 15, it drops from 5\% to 3\% level,
@@
 
-If we are ok with less than 10\% error, we can employ the Gaussian
+If we are OK with less than 10\% error, we can employ the Gaussian
 approximation from $n_i\ge5$, which means we neglect the 3rd and
@@
 be very close to the cuts. Formally, suppose $V$ is the subvolume of
-the unit simplex in which all constraints are satistfied. Outside of
+the unit simplex in which all constraints are satisfied. Outside of
 $V$, we set $P(\vecth) \equiv 0$. In the code, a sequence of

### 2015-09-28 posterior/model.tex
[posterior] Rewrite `incorporating the observations`
Math didn't really change but I added a lot explanation due to comments
from Hans Eggers

@@
 
-Let \Lumi{} denote the predicted or true luminosity in bin $i$ and
-$\vecL$ the ordered collection of luminosities. I assume from a
-telescope, we get the probability of the data $\mathcal{D}$ given
-$\vecth$ only as a function of $\vecL$ where the data are
-implicit---we don't have them---and no correlation between bins is
-given or assumed. The maximum-entropy distribution given only the
-means and standard deviations $\expin, i=1\dots n_b$ is the normal or
-Gaussian distribution, so we set
+Let \Lumi{} denote the true (and unknown) luminosity in bin $i$ and
+and let $\vecL$ represent the ordered collection of luminosities. At a
+telescope, a possible realization of what they can observe (charges in
+CMOS sensor?) is $X$. The physicists operating the telescope have a
+model how the luminosities $\vecL$ and other parameters of the
+telescop, call them $\vecnu$, affect the outcome $X$. Quantitatively,
+in their model they write down $P \given{X}{\vecL, \vecnu}$.  We
+further assume that the telescope people average out the effect of
+their nuisance parameters to report to us
+\begin{align}
+  \label{eq:aver-like}
+  P \given{X}{\vecL} = \int \rmdx{\vecnu} P \given{X}{\vecL, \vecnu} P(\vecnu) .
+\end{align}
+Once the telescope performs the measurement, they actually observe
+$X=\data$ and they tell us that $P \given{X=\data}{\vecL}$---viewed as
+a function of $\vecL$---can be approximated by a product of Gaussians
+characterized by the numbers $\expin, i=1 \dots n_b$:
 \begin{equation}
   \label{eq:like-L}
-  P(\mathcal{D} | \vecL) = \prod_{i=1}^{n_b} \gaussian\given{\Lumi}{\mu_i, \sigma_i^2}.
+  P(X=\mathcal{D} | \vecL) \propto \prod_{i=1}^{n_b} \gaussian\given{\Lumi}{\mu_i, \sigma_i^2}.
 \end{equation}
+Note that for a different realization of $X$, $ P \given{X}{\vecL}$
+might need to be approximated by a functional form different from a
+product Gaussian. But this is of no concern to us because we only care
+about what was actually observed; i.e., $X=\data$. The numbers
+$\expin$ of course depend on $X$ (and the effect of $\vecnu$!). There
+is a $\propto$ sign in \refeq{like-L} because the lhs sign is a
+probability density for $X$ but not for $\vecL$. We don't know what
+$\data$ actually is: the telescope people don't tell us and we don't
+need to know as long as we only care about the dependence on
+$\vecL$. For our purposes, all we need are the values $\expin$ to plug
+into \refeq{like-L} under the given assumptions. We can ignore the
+normalization unless we compare different models in \tardis{}
+that give rise to different normalization constants.
 
-In principle, $\vecL(\vecth)$ is a deterministic function of
-$\vecth$. But we are unable to calculate it as is; that's why we have
-to estimate it from the \tardis{} simulation resulting in a
-distribution $P\given{\vecL}{\vecth}$. To use Bayes'
-theorem~\refeq{Bayes-full}, we need $P\given{\data}{\vecth}$ but we
-only have $P\given{\data}{\vecL}$. The link between these three
-densities is the law of total probability:
+To use Bayes' theorem~\refeq{Bayes-full}, we need
+$P\given{\data}{\vecth}$ but we only have $P\given{\data}{\vecL}$. The
+link is the law of total probability:
 \begin{align}
   \label{eq:like-expanded}
-  P\given{\data}{\vecth} &= \int \rmdx{\vecL} P\given{\vecL}{\vecth} P\given{\data}{\vecL}.
+  P\given{X}{\vecth} &= \int \rmdx{\vecL} P\given{X}{\vecL, \vecth} P\given{\vecL}{\vecth} .
 \end{align}
-Note that in principle it should say $P\given{\data}{\vecL, \vecth}$
-in \refeq{like-expanded} but in our case $\vecth$ doesn't appear
-explicitly, so we can just plug in
-$P\given{\data}{\vecL}$ from~\refeq{like-L}.
+We \emph{assume} that our parameters of interest, $\vecth$, do \emph{not}
+affect $X$ in any other way than through $\vecL$; i.e., $\vecth$
+doesn't appear explicitly
+\begin{align}
+  \label{eq:no-expl-L}
+  P\given{X}{\vecL, \vecth} = P\given{X}{\vecL}.
+\end{align}
+and we can reuse \refeq{aver-like} on the rhs.  In principle, $\vecL$
+is a deterministic function of $\vecth$, $\vecL = f(\vecth)$. But we
+are unable to simply evaluate this function; that's why we have to
+estimate it from the \tardis{} simulation resulting in a distribution
+$P\given{\vecL}{\vecth} \ne \delta(\vecL - f(\vecth))$. Specifically,
+the outcome of \tardis{} depends on the random number seed and the
+fact that only a finite number of (effective) photons are
+simulated. In the limit of infinite number of photons,
+$P\given{\vecL}{\vecth} \to \delta(\vecL - f(\vecth))$. Note that this
+does not mean that we know the true value of the luminosity in bin $i$
+for a given supernova. Instead it means that there is no uncertainty
+due to the finite simulation size. Whatever \tardis{} spits out
+depends on the physics assumptions coded up in \tardis{} and there can
+still be systematic shifts.
 
+Putting together equations~\refeq{like-L}--\refeq{no-expl-L}, the
+likelihood is
+\begin{align}
+  \label{eq:like-summary}
+  \likelihood(\vecth) \propto \int \rmdx{\vecL} \prod_{i=1}^{n_b}
+  \gaussian\given{\Lumi}{\mu_i, \sigma_i^2} P\given{\vecL}{\vecth}
+\end{align}
 To complete the specification of the likelihood, we need to formulate

### 2015-10-20 posterior/christian-robert.tex
[posterior] Quick summary for Christian Robert

\documentclass[a4,12pt]{article}

\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[proportional]{libertine}
\usepackage[libertine]{newtxmath}

% Import the natbib package and sets a bibliography  and citation styles
\usepackage[numbers]{natbib}
\bibliographystyle{plainnat}
% \setcitestyle{numbers}

\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{paralist}
\usepackage{xcolor}

%%% math
\newcommand{\betap}{\ensuremath{\beta^\prime}}
\newcommand{\data}{\ensuremath{\mathcal{D}}}
\newcommand{\expin}{\ensuremath{\lbrace\mu_i, \sigma_i\rbrace}}
\DeclareMathOperator{\Dirichlet}{Dirichlet}
\newcommand{\given}[2]{\ensuremath{(#1 | #2)}}
\newcommand{\gaussian}{\ensuremath{\mathcal{N}}}
\newcommand{\genbetapr}{\ensuremath{\mathrm{GenBetaPrime}}}
\newcommand{\likelihood}{\ensuremath{\mathcal{L}}}
\newcommand{\lumi}{\ensuremath{\ell_i}}
\newcommand{\Lumi}{\ensuremath{L_i}}
\newcommand{\real}{\ensuremath{{\rm real}}}
\newcommand{\virt}{\ensuremath{{\rm virt}}}
\newcommand{\Nreal}{\ensuremath{N_\real}}
\newcommand{\Nvirt}{\ensuremath{N_\virt}}
\newcommand{\poisson}{\ensuremath{\mathrm{Poisson}}}
\newcommand{\rmdx}[1]{\mbox{d} #1 \,} % differential

\renewcommand{\vec}[1]{\boldsymbol{#1}}
\newcommand{\vecalpha}{\ensuremath{\vec{\alpha}}}
\newcommand{\vecL}{\ensuremath{\vec{L}}}
\newcommand{\vecmu}{\ensuremath{\vec{\mu}}}
\newcommand{\vecnu}{\ensuremath{\vec{\nu}}}
\newcommand{\vecth}{\ensuremath{{\vec{\theta}}}}
\newcommand{\vecu}{\ensuremath{{\vec{u}}}}

\DeclareMathOperator{\erf}{erf}
\newcommand{\mui}{\mu_i}
% \newcommand{\Ni}{N_i}
% \newcommand{\ni}{n_i}
\newcommand{\sigi}{\sigma_i^2}
\DeclareMathOperator{\sgn}{sgn}
\newcommand{\El}{E[l]}
\newcommand{\Vl}{V[l]}

%%% refs
\def \refeq#1{(\ref{eq:#1})}
\def \refsec#1{sec.~\ref{sec:#1}}
\def \refapp#1{app.~\ref{#1}}
\def \reffig#1{fig.~\ref{fig:#1}}

%%% comments
\newcommand{\todo}[1]{{\textsc{\color{red}#1}}}

%%% codes
\newcommand{\tardis}{TARDIS}

\title{Incorporating simulation uncertainty}
\author{Frederik Beaujean}

\begin{document}
\maketitle

\noindent
\emph{I here try to formulate the problem in a way that abstracts
  nearly all the physics. Discussing with you helped a lot and I think
  I straightened it out now. I work with the supernova experts at the
  ESO and I am responsible for the statistics part.}

\section*{Overview}
My goal is to do a fit comparing a luminosity spectrum obtained from a
telescope to the model prediction. The spectrum describes the
intensity of light of various wavelenghts emitted by a distant
supernova. Our parameters of interest $\vecth$ include the abundance
of various elements produced during the supernova explosion. These
elements produce dips and other absorption features in the
spectrum.

The main difficulty is that I cannot compute the luminosity
$L(\vecth)$ directly. I have to start an expensive simulation to
compute the spectrum for the $\sim 2000$ bins. The simulation is
multithreaded and takes about 100s for a single evaluation on the
16-core machines at our C2PAP compute cluster. I was told that some
parts are not trivially parallel so it doesn't make much sense to
scale the simulation to multiple machines. Our simulation is a much
simplified version of a full-blown code that takes weeks on hundreds
of cores. Ultimately our simulation propagates $N$ photons through the
supernova. The simulation uncertainty on $L$ would tend to zero for $N
\to \infty$. But we can only afford $N \approx 40000$.

It seems to me this is a general problem a lot of people have so I
searched for a Bayesian solution but found only some frequentist
recipes that require a lot of simulated photons in each wavelenght
bin. But I want to be able to treat the case with low counts as well.

\section*{Statistical model}

Bayes' theorem

\begin{align}
  P\given{\vecth}{\data} \propto P\given{\data}{\vecth} P(\vecth)
\end{align}
For most of $\vecth$ we have relevant prior information but I won't
discuss $P(\vecth)$ further. Regarding $P\given{\data}{\vecth}$, I
have to state that we don't have the actual data $\data$, that is the
counts etc. We only get the likelihood as a function of $L$,
$P\given{\data}{L}$.

In our model, the true luminosity is a given function $f(\vecth)$ that
we cannot evaluate. But for a given random-number seed $\xi$, our
simulation represented by $g(\vecth, \xi)$ yields an estimator of
$L$. I want to marginalize over $\xi$ to take into account the
uncertainty from the simulation.
\begin{align}
  \label{eq:xi-int}
  P\given{\data}{\vecth} = \int \rmdx{\xi}  P(\xi) P\given{\data}{\vecth, \xi}
\end{align}
Formally I change variables from $\xi$ to $L=g(\vecth, \xi)$ for fixed $\vecth$
\begin{align}
  P\given{\data}{\vecth} = \int \rmdx{L} \left(\frac{\rmdx{g(\vecth, \xi)}}{\rmdx{\xi}} \right)^{-1} P(\xi(L)) P\given{\data}{\vecth, L}
\end{align}
Now I combine the first terms as a distribution over $L$ for fixed
$\vecth$ and I use the fact that $P\given{\data}{\vecth, L}$ doesn't
explicitly depend on $\vecth$ to obtain the final expression
\begin{align}
  \label{eq:L-int}
  P\given{\data}{\vecth} = \int \rmdx{L} P(L| \vecth) P\given{\data}{L}
\end{align}
I can incorporate the distribution from the telescope
$P\given{\data}{L}$ but now I need to account for the uncertainty from
the simulation through $P(L|\vecth)$. For $N \to \infty$, $P(L| \vecth) \to \delta(L -
L(\vecth))$.

What is won?  but A naive approach would be to approximate
\refeq{xi-int} by repeating the simulation for many $\xi$ but that's
expensive and we want to avoid it. Instead, I want to put in my
knowledge about the simulation to get an approximation for $P(L|
\vecth)$ to either solve \refeq{L-int} analytically or by 1D
quadrature which is much faster because both distributions under the
integral sign are assumed much faster to evaluate than $g(\vecth,
\xi)$.

\end{document}

### 2015-10-28 posterior/MyBinarySearch.m
[posterior] Validate the likelihood on 250 tardis replicas

@@ -1,0 +1,93 @@
+(* ::Package:: *)
+
+BeginPackage["MyBinarySearch`"]
+binSearchMin::usage = 
+"binSearchMin[list, key] returns the index of the smallest element of the sorted list that is larger or equal to key.
+0 is returned if no such element exists in the list. 
+For consecutive identical values, the index of the first such value is returned."
+binSearchMax::usage = 
+"binSearchMax[list, key] returns the index of the largest element of the sorted list that is less than or equal to key.
+0 is returned if no such element exists in the list. 
+For consecutive identical values, the index of the last such value is returned."
+
+Begin["`Private`"]
+
+
+(* ::Title:: *)
+(*Follow http://googleresearch.blogspot.de/2006/06/extra-extra-read-all-about-it-nearly.html*)
+
+
+(* ::Section:: *)
+(*Return first element >= key*)
+
+
+binSearchMin[list_List, key_]:=Module[{low=1,high=Length[list],mid,midVal},
+(* first element is the right one *)
+If[First[list]>=key,Return[1]];
+(* key larger than everything in list *)
+If[Last[list]<key,Return[0]];
+
+(* else it is inside the list *)
+While[low< high-1,
+mid=Quotient[low+high,2];
+midVal = list[[mid]];
+(*Print[{mid,midVal}];*)
+If[midVal>= key,high=mid,
+If[midVal<key,low=mid];
+];
+];
+high
+]
+
+
+(* ::Section:: *)
+(*Return last element <= key*)
+
+
+binSearchMax[list_List, key_]:=Module[{low=1,high=Length[list],mid,midVal},
+(* first element too large => element cannot be found *)
+If[First[list]>key,Return[0]];
+(* key larger than everything in list *)
+If[Last[list]<= key,Return[high]];
+
+(* else it is inside the list *)
+While[low< high-1,
+mid=Quotient[low+high,2];
+midVal = list[[mid]];
+(*Print[{mid,midVal}];*)
+If[midVal<=  key,low=mid,
+If[midVal> key,high=mid];
+];
+];
+low
+]
+
+
+(* ::Section:: *)
+(*Testing*)
+
+
+On[Assert];
[…]

### 2015-10-28 posterior/model.tex
[model.tex] Minor reformatting

@@
 \def \refeq#1{(\ref{eq:#1})}
-\def \refsec#1{sec.~\ref{#1}}
-\def \refSec#1{Sec.~\ref{#1}}
+\def \refsec#1{sec.~\ref{sec:#1}}
 \def \refapp#1{app.~\ref{#1}}
@@
 
-\section{Input and Output}
+\section{Overview}
+
+\subsubsection*{Notation}
+I try to be consistent with standard notation in statistics and
+denote random variables by capital letters and numbers that are
+observed by lower case.
 
@@
 \begin{itemize}
-\item data $\data \sim \expin =$ set of input values from
-  telescope for each frequency bin $i=1 \dots n_b$;
+  \item $\expin =$ the set of input values from the telescope for each
+  frequency bin $i=1 \dots n_b$. For the interpretation, see below in
+  \refsec{likelihood}.
 \item \tardis{} depends on the parameters $\vecth$ and
@@
 \end{itemize}
-I try to be mostly consistent with standard notation in statistics and
-denote random variables---eventually they will be integrated out---by
-capital letters and numbers that are observed by lower case.
 
@@
 % \end{align}
-\section{Likelihood}
+\section{Likelihood}\label{sec:likelihood}
 
@@
   \label{eq:sum-of-2}
-  P(\Lumi | N_i=2, \vecth) = \int \rmdx{l} P(\Lumi -l | N_i=1, \vecth) P(l | N_i=1, \vecth).
+  P(\Lumi | N_i=2, \vecth) = \int \rmdx{l} P(\Lumi -l | N_i=1, \vecth) P(l | N_i=1, \
+vecth).
 \end{align}

### 2015-10-28 posterior/likelihood.py
[posterior] Begin python implementation

@@ -1,0 +1,43 @@
+from astropy.modeling import Model, Parameter
+import math
+import numpy as np
+from scipy.special import gammaln
+
+
+'''
+Implement likelihood that estimates uncertainty due to tardis simulation
+'''
+
+def log_pochhammer(a, b):
+    '''
+    Pochhammer symbol (a)_b implemented for integer ``b``.
+    '''
+    assert a > 0
+    assert b >= 0
+    return gammaln(a+b) - gammaln(a)
+    # res = a
+    # for i in range(1, b):
+    #     res *= a + i
+    # return res
+
+
+def log_poisson_posterior_predictive(N, n):
+    '''
+    Poisson posterior predictive probability distribution to observe ``N`` given ``n`` on the log scale.
+
+    A Jeffrey's prior is assumed for the rate parameter. The derivation is given in arXiv:1112.2593
+    '''
+    return log_pochhammer(N + n + 1, N + n) - log_pochhammer(n + 1, n) - (3 * N + n + 0.5) * math.log(2) - gammaln(N)
+
+
+class TARDISBayesianLogLikelihood(Model):
+    inputs = ('packet_nu', 'packet_energies')
+    outputs = ('loglikelihood')
+
+    def __init__(self, wavelength, flux):
+        super(TARDISBayesianLogLikelihood, self).__init__()
+        self.wavelength = wavelength
+        self.flux = flux
+
+    def evaluate(self, packet_nu, packet_energies):
+         return 5

### 2015-10-28 posterior/likelihood_test.py

@@ -1,0 +1,7 @@
+from likelihood import *
+from numpy import testing
+
+def test_log_poisson_posterior_predictive():
+    args = ((13, 11), (11, 13), (1, 2))
+    res  = ()
+    testing.assert_almost_equal(log_poisson_posterior_predictive(13, 11), -2.63578)

### 2015-10-28 posterior/likelihood.py
(wip)[likelihood] implement first few functions

@@ -15,10 +15,6 @@ def log_pochhammer(a, b):
     assert a > 0
     assert b >= 0
     return gammaln(a+b) - gammaln(a)
-    # res = a
-    # for i in range(1, b):
-    #     res *= a + i
-    # return res
 
 
 def log_poisson_posterior_predictive(N, n):
@@ -27,7 +23,140 @@ def log_poisson_posterior_predictive(N, n):
 
     A Jeffrey's prior is assumed for the rate parameter. The derivation is given in arXiv:1112.2593
     '''
-    return log_pochhammer(N + n + 1, N + n) - log_pochhammer(n + 1, n) - (3 * N + n + 0.5) * math.log(2) - gammaln(N)
+    return log_pochhammer(N + n + 1, N + n) - log_pochhammer(n + 1, n) \
+           - (3 * N + n + 0.5) * math.log(2) - gammaln(N + 1)
+
+
+def binary_search_min(data, key, low=None, high=None):
+    '''
+    Find the index of the smallest element in ``data`` that is >= ``key``.
+    '''
+
+    # update low and high until element found
+    if low is None:
+        low = 0
+    if high is None:
+        high = len(data) - 1
+
+    assert low < high
+
+    # first element is the right one
+    if data[0] >= key:
+        return low
+    if data[-1] < key:
+        raise KeyError('Key ' + str(key) + ' exceeds all elements')
+
+    while low < (high - 1):
+        mid = (low + high) // 2
+        midval = data[mid]
+        if midval >= key:
+            high = mid
+        if midval < key:
+            low = mid
+
+    return high
+
+
+def binary_search_max(data, key, low=None, high=None):
+    '''
+    Find the index of the largest element in ``data`` that is <= ``key``.
+    '''
+
+    # update low and high until element found
+    if low is None:
+        low = 0
+    if high is None:
+        high = len(data) - 1
+
+    assert low < high
+
+    # first element is the right one
+    if data[-1] <= key:
+        return high
+    if data[0] > key:
+        raise KeyError('Key ' + str(key) + ' less than all elements')
+
+    while low < (high - 1):
+        mid = (low + high) // 2
+        midval = data[mid]
+        if midval <= key:
+            low = mid
+        if midval > key:
+            high = mid
+
+    return low
+
+
+def bin_indices(flux, data):
+    '''
+
+    :param flux:
+    :param data:
+    :return:
+    '''
+    pass
+
+def prob_L_given_N(L, N, lsamples):
+    '''
[…]

### 2015-10-28 posterior/likelihood_test.py

@@ -1,7 +1,38 @@
 from likelihood import *
 from numpy import testing
+import pytest
+
+def test_log_pochhammer():
+    args = ((13, 11), (11, 13), (1, 2))
+    res  = (31.6195, 36.5023, 0.693147)
+    for a, r in zip(args, res):
+        testing.assert_almost_equal(log_pochhammer(*a), r, decimal=3)
 
 def test_log_poisson_posterior_predictive():
     args = ((13, 11), (11, 13), (1, 2))
-    res  = ()
-    testing.assert_almost_equal(log_poisson_posterior_predictive(13, 11), -2.63578)
+    res  = (-2.63578, -2.554, -1.50972)
+    for a, r in zip(args, res):
+        testing.assert_almost_equal(log_poisson_posterior_predictive(*a), r, decimal=3)
+
+def test_binary_search():
+    data = (1, 3, 6, 7, 9, 9, 9, 9, 10)
+    keys = (0.1, 1, 2.2, 7, 9, 9.2, 11)
+    imin = (0,   0, 1,   3, 4, 8)
+    imax =      (0, 0,   3, 7, 7,   8)
+
+    for k, i in zip(keys[:-1], imin):
+        assert binary_search_min(data, k) == i
+
+    # key larger than all elements in ``data``
+    with pytest.raises(KeyError):
+        binary_search_min(data, keys[-1])
+
+    for k, i in zip(keys[1:], imax):
+        assert binary_search_max(data, k) == i
+
+    # key larger than all elements in ``data``
+    with pytest.raises(KeyError):
+        binary_search_max(data, keys[0])
+
+
+

### 2016-01-14 posterior/likelihood.py
(wip)[likelihood] add gamma dist

@@ -644,6 +644,20 @@ def gamma_max_likelihood(samples, max=None):
     return a, theta, alpha
 
 
+def gamma_log_likelihood(a, theta, alpha, x, N=1):
+    '''
+
+    :param a: location
+    :param theta: scale (<0 OK)
+    :param alpha: shape
+    :param x: random variable
+    :param N: consider ``x`` as the sum of ``N`` Gamma variates with the given parameters.
+    :return: ``log(Gamma(sgn(theta) * (x - N * a) | 0, |theta|, N * alpha)``
+    '''
+    abs_theta = math.fabs(theta)
+    scipy.stats.gamma.logpdf((theta / abs_theta) * (x - N * a), N * alpha, loc=0, scale=abs_theta)
+
+
 class TARDISBayesianLogLikelihood(object):
     def __init__(self, telescope_nus, telescope_luminosities,
                  packet_nus, packet_luminosities, window=1500):
@@ -728,6 +742,20 @@ class TARDISBayesianLogLikelihood(object):
             while len(self.sim) - self.windows[-1] < window:
                 prev_ind -= 1
                 self.windows[-1] = self.bin_indices[prev_ind]
+
+        # fit a Gamma in each window
+        # todo parallel?
+        gamma_dists = []
+        for i, _ in enumerate(self.windows[1:]):
+            if i == len(self.windows) - 1 and len(self.windows) > 2:
+                lo, hi = self.windows[i], len(self.sim)
+            else:
+                lo, hi = self.windows[i - 1], self.windows[i]
+
+            pars = gamma_max_likelihood(self.sim.energies[lo:hi])
+            gamma_dists.append()
+
+        # todo check for ``len == 2``
 
 
     def __call__(self, *args, **kwargs):

### 2016-01-14 hans/bayes-tardis-luminosity-prediction.tex
[hans] Add Hans' draft from Dec 15th, 2015 as initial version

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Bayesian prediction of luminosity distribution based on one simulation run:
% An example from TARDIS
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[11pt]{article}
\usepackage[a4paper, left=20mm, right=20mm, top=25mm, bottom=25mm]{geometry}

%% CHANGE THIS DATE AS YOU MODIFY THE FILE
\newcommand{\vdate}{151211}
\pagestyle{myheadings}
\markboth{Version \vdate} {Version \vdate}
\parindent=0mm
\def\pp{\hfill\par \vspace{\baselineskip}}
%% \def\pp{}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{bm,amssymb,amsfonts,amsmath}  % bold math symbols, AMStex
\usepackage{graphicx}    % standard graphics
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\lleq}[1]{\label{#1} }
% TO REMOVE THE LABEL LETTERS FROM THE EQUATION DISPLAY, COMMENT OUT THIS LINE:
\renewcommand{\lleq}[1]{\label{#1} {\scriptstyle {\rm (#1)}} \hspace*{2ex} }
% \renewcommand{\labelenumi}{\arabic{section}.\arabic{subsection}.\arabic{enumi}}
% \renewcommand{\thesubsubsection}{\arabic{subsubsection}}

\newcounter{mnumi}
\newenvironment{mnumerate}{
  \begin{list}{\arabic{mnumi}.}
    {\usecounter{mnumi} \setlength{\itemsep}{0pt}
      \setlength{\leftmargin}{3ex}
    }
  }
  {\end{list}}

\newenvironment{mtemize}{
  \begin{list}{$\bullet$}
    {\setlength{\itemsep}{0pt}
     \setlength{\leftmargin}{3ex}
    }
  }
  {\end{list}}

\newcommand{\hmod}  {{\mathcal{H}}}  % Model
\newcommand{\ldef}{\;{:}{=}\;}
\newcommand{\rdef}{\;{=}{:}\;}
\newcommand{\realnumbers}{\mathbb{R}}
\newcommand{\integers}{\mathbb{Z}}
\newcommand{\cond}{\,|\,}
\newcommand{\NBD}{\mathrm{NB}}
\newcommand{\var}{\mathrm{var}}

\newcommand{\smA}{{\scriptscriptstyle A}}
\newcommand{\smB}{{\scriptscriptstyle B}}
\newcommand{\smH}{{\scriptscriptstyle H}}
\newcommand{\smK}{{\scriptscriptstyle K}}
\newcommand{\smL}{{\scriptscriptstyle L}}
\newcommand{\smM}{{\scriptscriptstyle M}}
\newcommand{\smN}{{\scriptscriptstyle N}}
\newcommand{\smR}{{\scriptscriptstyle R}}

\newcommand{\smo}{{\scriptscriptstyle 1}}
\newcommand{\smt}{{\scriptscriptstyle 2}}
\newcommand{\smr}{{\scriptscriptstyle 3}}
\newcommand{\smf}{{\scriptscriptstyle 4}}
\newcommand{\smv}{{\scriptscriptstyle 5}}
\newcommand{\smx}{{\scriptscriptstyle 6}}
\newcommand{\sms}{{\scriptscriptstyle 7}}
\newcommand{\sme}{{\scriptscriptstyle 8}}
\newcommand{\smn}{{\scriptscriptstyle 9}}

\newcommand{\bma}{{\bm{a}}}
\newcommand{\bmb}{{\bm{b}}}
\newcommand{\bmc}{{\bm{c}}}
\newcommand{\bmd}{{\bm{d}}}
\newcommand{\bme}{{\bm{e}}}
\newcommand{\bmf}{{\bm{f}}}
\newcommand{\bmg}{{\bm{g}}}
\newcommand{\bmh}{{\bm{h}}}
\newcommand{\bmi}{{\bm{i}}}
\newcommand{\bmj}{{\bm{j}}}
\newcommand{\bmk}{{\bm{k}}}
\newcommand{\bml}{{\bm{\ell}}}
\newcommand{\bmm}{{\bm{m}}}
\newcommand{\bmn}{{\bm{n}}}
\newcommand{\bmo}{{\bm{o}}}
\newcommand{\bmp}{{\bm{p}}}
\newcommand{\bmq}{{\bm{q}}}
\newcommand{\bmr}{{\bm{r}}}
\newcommand{\bms}{{\bm{s}}}
\newcommand{\bmt}{{\bm{t}}}
\newcommand{\bmu}{{\bm{u}}}
\newcommand{\bmv}{{\bm{v}}}
\newcommand{\bmw}{{\bm{w}}}
\newcommand{\bmx}{{{\bm{x}}}}
\newcommand{\bmy}{{\bm{y}}}
\newcommand{\bmz}{{\bm{z}}}
\newcommand{\bmA}{{\bm{A}}}
\newcommand{\bmB}{{\bm{B}}}
\newcommand{\bmC}{{\bm{C}}}
\newcommand{\bmD}{{\bm{D}}}
\newcommand{\bmE}{{\bm{E}}}
\newcommand{\bmF}{{\bm{F}}}
\newcommand{\bmG}{{\bm{G}}}
\newcommand{\bmH}{{\bm{H}}}
\newcommand{\bmI}{{\bm{I}}}
\newcommand{\bmJ}{{\bm{J}}}
\newcommand{\bmK}{{\bm{K}}}
\newcommand{\bmL}{{\bm{L}}}
\newcommand{\bmM}{{\bm{M}}}
\newcommand{\bmN}{{\bm{N}}}
\newcommand{\bmO}{{\bm{O}}}
\newcommand{\bmP}{{\bm{P}}}
\newcommand{\bmQ}{{\bm{Q}}}
\newcommand{\bmR}{{\bm{R}}}
\newcommand{\bmS}{{\bm{S}}}
\newcommand{\bmT}{{\bm{T}}}
\newcommand{\bmU}{{\bm{U}}}
\newcommand{\bmV}{{\bm{V}}}
\newcommand{\bmW}{{\bm{W}}}
\newcommand{\bmX}{{\bm{X}}}
\newcommand{\bmY}{{\bm{Y}}}
\newcommand{\bmZ}{{\bm{Z}}}

\newcommand{\bmalpha}{{\bm{\alpha}}}
\newcommand{\bmbeta}{{\bm{\beta}}}
\newcommand{\bmchi}{{\bm{\chi}}}
\newcommand{\bmdelta}{{\bm{\delta}}}
\newcommand{\bmepsilon}{{\bm{\varepsilon}}}
\newcommand{\bmphi}{{\bm{\phi}}}
\newcommand{\bmgamma}{{\bm{\gamma}}}
\newcommand{\bmeta}{{\bm{\eta}}}
\newcommand{\bmiota}{{\bm{\iota}}}
\newcommand{\bmkappa}{{\bm{\kappa}}}
\newcommand{\bmlambda}{{\bm{\lambda}}}
\newcommand{\bmmu}{{\bm{\mu}}}
\newcommand{\bmnu}{{\bm{\nu}}}
\newcommand{\bmomega}{{\bm{\omega}}}
\newcommand{\bmpi}{{\bm{\pi}}}
\newcommand{\bmpsi}{{\bm{\psi}}}
\newcommand{\bmrho}{{\bm{\rho}}}
\newcommand{\bmsigma}{{\bm{\sigma}}}
\newcommand{\bmtau}{{\bm{\tau}}}
\newcommand{\bmtheta}{{\bm{\theta}}}
\newcommand{\bmupsilon}{{\bm{\upsilon}}}
\newcommand{\bmxi}{{\bm{\xi}}}
\newcommand{\bmzeta}{{\bm{\zeta}}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}

\begin{center}
  \textbf{\Large Bayesian prediction of supernova luminosity  distributions}\\[8pt]
  \textbf{\Large based on simulation output: An example from TARDIS}\\[12pt]
  11 December 2015
\end{center}

\textbf{Abstract} TODO

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

TODO: Wolfgang

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Bayesian prediction}

\textbf{HANS NOTE: I have just copied stuff in here which will be
  written up properly later.}
\begin{mtemize}
\item Generic definition of terms: data, parameter, variable
\item Define likelihood $p(x\cond\theta,\hmod)$
\item Product rule and sum rule (marginalisation)
\item Bayes' Theorem

\item Building a model $\hmod$ implies that we hypothesise both a
  likelihood $p(x\cond\bmtheta,\hmod)$ and a prior
  $p(\bmtheta\cond\hmod)$.

\item $p(D\cond\theta,\hmod)$ is the \textbf{likelihood} for $\theta$,
  given the data $D$. Note that, while the likelihood appears to be
  identical with the sampling distribution $p(D\cond\theta,\hmod)$
  both in notation and even in its functional form, these two
  quantities have a different purpose: The likelihood is a function of
  $\theta$ for fixed data $D$, while the forward probability is a
  function of $D$ for fixed $\theta$.

\item $p(\theta\cond\hmod)$ is the \textbf{prior probability} for
  $\theta$, before taking into account the information provided by the
  data $D$.

\item $p(D\cond\hmod)$ is sometimes called the \textbf{evidence}: it
  contains just the data, not any particular value of $\theta$. The
  only way to connect the ``evidence'' to the other probabilities is
  to expand it in terms of the sum rule, $p(D\cond\hmod) =
  \sum_{\theta'} p(D\cond\theta',\hmod)\;p(\theta'\cond\hmod)$. In
  this sense it can be thought of as the normalisation constant which
  ensures that the posterior is properly normalised.

\item $p(\theta\cond D,\hmod)$ is the \textbf{posterior probability}
  for $\theta$, after the data $D$ has been taken into account. Given
  a set of observations $D$, the $p(\theta\cond D,\hmod)$ gives the
  probability that $\theta$ takes on a particular value.

\item \texttt{COMMENT: the bracketed letters to the left of equations
    is a temporary aid to facilitate referencing. A change in
    definition of command} \verb=\lleq= \textbf{will make them revert
    to ordinary equation labels.}

\item \textbf{Bayes' Theorem}: In terms of these quantities, Bayes'
  theorem, the central theorem of inference is given by
  \begin{align}
    \lleq{bth}
    p(\theta\cond D,\hmod)
    &= \frac{p(D\cond\theta,\hmod)\;p(\theta\cond\hmod)}
    {p(D\cond\hmod)}
    = \frac{p(D\cond\theta,\hmod)\;p(\theta\cond\hmod)}
    {\int d\theta'\; p(D\cond\theta',\hmod)\;p(\theta'\cond\hmod)}
  \end{align}
  where the integral in the denominator is replaced by a sum if
  $\theta$ takes on discrete values.

\item Note that if the prior is uniform, it is independent of $\theta$
  and will cancel; we are then left with
  \begin{equation}
    \lleq{bthu}
    p(\theta\cond D,\hmod)
    = \frac{p(D\cond\theta,\hmod)} {\int d\theta'\; p(D\cond\theta',\hmod)}
  \end{equation}

\item \textbf{Prediction using Bayes' Theorem:} By contrast, the
  Bayesian prediction for $x_{n+1}$ makes full use of the information
  contained in previous data:
  \begin{align}
    p(x_{n+1}\cond D,\hmod)
    &= \int d\theta'\; p(x_{n+1}\cond\theta',D,\hmod)\;p(\theta'\cond
    D,\hmod)
  \end{align}
  i.e.\ all possible values of $\theta$ are taken into account. Using
  Bayes' theorem, this can be written also as
  \begin{align}
    p(x_{n+1}\cond D,\hmod)
    &= \frac{\int d\theta'\; p(x_{n+1}\cond\theta',D,\hmod)\;
      p(D\cond\theta',\hmod)\;p(\theta'\cond\hmod)}
    {\int d\theta'\; p(D\cond\theta',\hmod)\;p(\theta'\cond\hmod)}
  \end{align}
  which shows that the prediction for $x_{n+1}$ is a weighted average
  of $p(x_{n+1}\cond\theta',D,\hmod)$ with the weight given by
  \begin{align}
  \frac{p(D\cond\theta',\hmod)\;p(\theta'\cond\hmod)}
  {\int d\theta'\; p(D\cond\theta',\hmod)\;p(\theta'\cond\hmod)}
  &=
  \frac{p(D,\theta'\cond\hmod)}
  {\int d\theta'\; p(D,\theta'\cond\hmod)}
  \end{align}
\end{mtemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{TARDIS}

\subsection{Intro to those TARDIS aspects relevant to this paper}

TODO: Wolfgang (or in Introduction?)

\subsection{Definitions of relevant quantities}

\begin{mtemize}
\item Spectrum of luminosities in channels (or ``bins'')
  $b=0,1,\ldots,B$
\item The wavelength of channel $b$ is given by $\lambda_b =
  \lambda_0 + b\,\delta\lambda$, where $\delta\lambda$ is the bin width.

\item We treat the results of a single TARDIS run as \textit{data}
  $D$, i.e.\ fixed results which we do not question but accept as the
  true values.  Following a single run, the TARDIS ``data'' consists,
  for each bin $b$, of the number or ``multiplicity'' of packets and
  the individual packet luminosities,
\begin{align}
  \lleq{dfc}
  D &= \{n_0,\bml_0,n_1,\bml_1,\ldots,n_\smB,\bml_\smB\}
  \ =\ \{n_b,\bml_b\}_{b=0}^\smB
\end{align}
with
\begin{align}
  \lleq{dfd}
  \bml_b &= (\ell_{b,1},\ell_{b,2},\ldots,\ell_{b,n_b})
\end{align}
Given $D$, the total luminosity in bin $b$ is known,
\begin{align}
  \lleq{dfe}
  \ell_b = \sum_{j=1}^{n_b}\ell_{b,j}
\end{align}
\end{mtemize}

\subsection{Data as realization of compound process}

While we treat the TARDIS output as ``data'' as discussed above, a
repeat of the simulation with a different Monte Carlo seed would
evidently yield different ``data'' $D'$ even when all other simulation
parameters remain the same. Clearly, $n_b$ is a specific realization
of a stochastic variable $N_b$ governed by a model pdf
$p(N_b)$. Similarly, $\bml_b$ is a specific realization of a
collection stochastic variables $\bmL_b
=\{L_{b,1},\ldots,L_{b,\smN_b}\}$, where the number $N_b$ of such
variables is itself stochastic. The relevant pdf for $\bmL_b$
therefore always presupposes knowledge of $N_b$, so we must work with
the conditional probability $p(\bmL_b\cond N_b)$. This situation is
captured mathematically as $N_b \sim p(N_b), \bmL_b \sim p(\bmL_b\cond
N_b)$ where $\sim$ means ``is by hypothesis distributed in this model
according to the pdf'', i.e.\
\begin{align*}
  (N_b,\bmL_b) &\ \sim\ p(N_b,\bmL_b\cond \bmphi_b,\lambda_b)
  = p(\bmL_b\cond N_b,\bmphi_b)\; p(N_b\cond \lambda_b),
\end{align*}
where $\bmphi_b$ is the set of parameters of the model pdf which
govern $\bmL_b$ and $\lambda_b$ is the parameter of the model pdf
governing $N_b$.  To simulate $(N_b,\bmL_b)$, we clearly first need to
generate one $N_b$, followed by the generation of $N$ packet
luminosities $\bmL_b$. This situation is termed a \textit{compound
  process}.

\subsection{Construction of the model hypothesis}

\begin{mtemize}
\item In the remainder of this paper, we assume that the packet
  multiplicities $N_b$ and total luminosities $\bmL_b$ in any given
  bin $b$ are statistically independent of those in other bins. This
  means that the joint probability over all bins factorises into
  individual probabilities for each bin,
  \begin{align}
    \lleq{bsi}
    p(N_0,\bmL_0\ldots,N_b,\bmL_b,\ldots,N_\smB,\bmL_\smB
    \cond \bmtheta,\bmphi,\bmlambda,\hmod)
    &= \prod_{b=0}^B p(N_b,\bmL_b\cond \bmtheta,\bmphi,\bmlambda,\hmod)
    % p(N_0,\ldots,N_b,\ldots,N_\smB \cond \bmtheta,\bmphi,\hmod)
    % &= \prod_{b=0}^B p(N_b\cond \bmtheta,\bmphi,\hmod)\\
    % p(L_0,\ldots,L_b,\ldots,L_\smB \cond N_0,\ldots,N_\smB,\bmtheta,\bmphi,\hmod)
    % &= \prod_{b=0}^B p(L_b\cond N_b,\bmtheta,\bmphi,\hmod)
  \end{align}

\item We omit the supernova parameters $\bmtheta$ from the notation
  below.

\item Based on the assumption of bin-bin statistical independence, we
  shall henceforth drop the bin index and do all our calculations for
  an individual bin $b$ only, writing for example $L_b \to L$,
  $\ell_{b,j} \to \ell_j$ etc. The total spectrum is then obtained by
  doing any given analysis or calculation for all bins.

\item Since we shall not be doing model comparison, we can for the
  moment drop the $\hmod$ from the probability notation.

\item The \textbf{model} which we construct consists of the above
  assumptions plus
  \begin{align*}
    \text{likelihood } &\ p(\bmL\cond N,\bmphi)\\
    \text{likelihood } &\ p(N\cond\lambda)\\
    \text{prior } &\ p(\bmphi)\\
    \text{prior } &\ p(\lambda)
  \end{align*}
  which are assumed to have the same functional form in each bin,
  while of course the parameter values will change from bin to bin.

\item We choose $N$ to follow a Poisson distribution with parameter
  $\lambda$ (motivate choice),
  \begin{align}
    \lleq{pmc}
    p(N\cond\lambda) &= \frac{e^{-\lambda}\lambda^N}{N!}
    \qquad N = 0,1,\ldots,\infty,\quad \lambda>0
  \end{align}

\item We further assume that individual packets are generated
  independently,
  \begin{align}
    \lleq{pmda}
    p(\bmL\cond N,\bmphi) &= \prod_{j=1}^N p(L_j\cond \bmphi)
  \end{align}

\item The values of $\lambda$ and $\bmphi$ are unknown. It is the task
  of the analysis below to determine not just the single best values
  for these parameters, but to take into account every and all
  possible values.

\item Priors: metaparameter $a$ ***

\end{mtemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Prediction based on individual packets}

\subsection{Predicting the packet multiplicity $N$ in a bin}

\texttt{COMMENT: The derivation below is deliberately slow; the
  intention being to give the novice an easy introduction into the
  usage of Bayes before we hit the more complicated calculations
  further below.}

We first show how to use Bayesian prediction to find the probability
distribution for $N$, the number of packets in a bin, given knowledge
of ``data'' in the form of TARDIS multiplicity $n$ for that bin.
Since $n$ is a specific
realization of $N$, it results from the same Poisson distribution
\begin{align}
  \lleq{pmcb}
   p(n\cond\lambda) &= \frac{e^{-\lambda}\lambda^n}{n!}
\end{align}
with the same value for $\lambda$ as in (\ref{pmc}).  We can hence use
Bayes' Theorem to find the posterior distribution for $\lambda$ given
data $n$ and the Poisson hypothesis
\begin{align}
  \lleq{pmd}
  p(\lambda\cond n) %
  &= \frac{p(n\cond\lambda)\,p(\lambda)} {p(n)}
  \ =\ \frac{p(n\cond\lambda)\,p(\lambda)} %
  {\int d\lambda\, p(n\cond\lambda)\,p(\lambda)}
\end{align}
to find the prediction of $N$ given $n$ taking into account all
possible values of $\lambda$,
\begin{align}
  \lleq{pme}
  p(N\cond n)
  &= \int d\lambda\,p(N,\lambda\cond n)
  \ =\ \int d\lambda\,p(N\cond\lambda, n)\,p(\lambda\cond n)
  \ =\ \frac{\int d\lambda\,p(N\cond\lambda)\,p(n\cond\lambda)\,p(\lambda)}
  {\int d\lambda\, p(n\cond\lambda)\,p(\lambda)}
\end{align}
where $p(N\cond\lambda,n) = p(N\cond\lambda)$ is independent of $n$.
Taking an improper prior (see Appendix)\footnote{For a uniform prior,
  the case $a{=}0$, we require of course a maximum $\lambda_{\rm
    max}$. As long as $\lambda_{\rm max}$ is large enough, the
  integral can effectively be taken to infinity.}
\begin{align}
  \lleq{pmf}
  p(\lambda) = c\lambda^{-a} \qquad a = 0\;\text{or}\;1
\end{align}
and inserting this and (\ref{pmcb}) into (\ref{pme}), we obtain
for the evidence and prediction
\begin{align}
  \lleq{pmg}
  p(n) &= %
  \int_0^\infty d\lambda\, p(n\cond\lambda)\,p(\lambda)
  \ =\
  c \int_0^\infty d\lambda\, \frac{e^{-\lambda}\,\lambda^{n-a}}{n!}
  \ =\ c\frac{(n-a)!}{n!}
  \\
  \lleq{pmh}
  p(N\cond n) %
  &= \frac{n!}{(n-a)!} \int_0^\infty d\lambda\,
  \frac{e^{-2\lambda} \lambda^{N+n-a}}{N!\, n!}
  \ =\ \frac{(N+n-a)!}{N!\,(n-a)!\,2^{N+n+1-a}}
\end{align}
which we recognise as a Negative Binomial distribution (see appendix)
with parameters $k=n+1-a$ and $\rho=(1/2)$
\begin{align}
  \lleq{pmhb}
  p(N\cond n) &= \NBD(N\cond k{=}n{+}1{-}a, \rho{=}\tfrac{1}{2})
  \ =\ \binom{N{+}n{-}a}{N}
  \left(\frac{1}{2}\right)^N \left(\frac{1}{2}\right)^{n+1-a}
\end{align}
We shall use the properties of the Negative Binomial in the
calculations below.  Note that the distribution of $N$ given $n$ is
independent of the packets $\bmL$ and $\bml$ and their values.

\subsection{Predicting the total luminosity $L$ in a bin}

\texttt{COMMENT: We may want to eliminate some intermediate steps in
  the derivations below in the final paper. Wolfgang: how much detail
  should be supplied in the paper?}

\subsubsection{Direct probability}

Our goal of prediction of $L$ given the data $(n,\bml)$ is captured in
the probability $p(L\cond n,\bml)$ with all relevant parameters
integrated out. We first note from Eq.~(\ref{pmh}) that $p(N\cond
n,\bml) = p(N\cond n)$ does not depend on $\bml$. Furthermore, since
$L$ is fully determined by $(N,\bmL)$,
\begin{align}
  \lleq{pmjd}
  L &= \sum_{j=1}^N L_j
\end{align}
is does not depend directly on $(n,\bml)$, i.e.\
\begin{align}
  \lleq{pmi}
  p(L\cond N,\bmL,n,\bml)
  = p(L\cond N,\bmL)
  &= \delta(L - \textstyle\sum_{j=1}^N L_j).
\end{align}
Hence the prediction is given by
\begin{align}
  \lleq{pmj}
  p(L\cond n,\bml)
  &= \sum_{N=0}^\infty p(L\cond N,n,\bml)\,p(N\cond n,\bml) \\
  \lleq{pmjb}
  &= \sum_N p(N\cond n)\,\int d\bmL\, p(L\cond N,\bmL,n,\bml)\, p(\bmL\cond N,n,\bml)
  \\
  \lleq{pmjc}
  &= \sum_N p(N\cond n)\,\int d\bmL\,\delta(L - \textstyle\sum_{j=1}^N L_j)
  \, p(\bmL\cond N,n,\bml)
\end{align}
The dependence on the TARDIS ``data'' $(n,\bml)$ enters through
$p(\bmL\cond N,n,\bml)$ in terms of the parameters $\bmphi$ of
$p(\bmL\cond N,\bmphi)$,
\begin{align}
  \lleq{pmk}
  p(\bmL\cond N,n,\bml)
  &= \int d\bmphi\, p(\bmL\cond N,\bmphi,n,\bml)\,p(\bmphi\cond N,n,\bml) \\
  &= \int d\bmphi\, p(\bmL\cond N,\bmphi)\,p(\bmphi\cond N,n,\bml).
\end{align}
because the likelihood for $\bmL$ is fully determined by $N,\bmphi$
and hence independent of $(n,\bml)$.
%
Again using Bayes' Theorem (\ref{bth})
\begin{align}
  \lleq{pmkb}
  p(\bmphi\cond N,n,\bml) &=
  \frac{p(\bml\cond N,n,\bmphi)\,p(\bmphi\cond N,n)} %
  {\int d\bmphi\,p(\bml\cond N,n,\bmphi)\,p(\bmphi\cond N,n)} %
  % \ =\ \frac{p(\bml\cond N,n,\bmphi)\,p(\bmphi\cond N,n)} %
  % {p(\bml\cond n)} %
\end{align}
and using a multiplicity-independent prior for $\bmphi$
\begin{align}
  \lleq{pml}
  p(\bmphi\cond N,n) &= p(\bmphi)
\end{align}
we can write (\ref{pmk}) as
\begin{align}
  \lleq{pmn}
  p(\bmL\cond N,n,\bml)
  &= \int d\bmphi\, p(\bmL\cond N,\bmphi)\, %
  \frac{p(\bml\cond n,\bmphi)\,p(\bmphi)} %
  {\int d\bmphi\,p(\bml\cond n,\bmphi)\,p(\bmphi)} \\
  \lleq{pmo}
  &= \frac{1}  {p(\bml\cond n)} %
  \int d\bmphi\, p(\bmL\cond N,\bmphi)\, %
  p(\bml\cond n,\bmphi)\,p(\bmphi)
\end{align}
with $p(\bml\cond n) = \int d\bmphi\, p(\bml\cond
n,\bmphi)\,p(\bmphi)$ in the denominator.
Inserting this into (\ref{pmjc}), we obtain
\begin{equation}
  \lleq{pmp}
  \boxed{
  p(L\cond n,\bml)
  = \frac{1}{p(\bml\cond n)}
  \sum_N p(N\cond n)\,\int d\bmL\,\delta(L - \textstyle\sum_j L_j)
  \displaystyle\int d\bmphi\,p(\bmL\cond N,\bmphi) \, p(\bml\cond n,\bmphi)
  \, p(\phi)
  }
\end{equation}
with
\begin{align}
  p(\bmL\cond N,\bmphi) &= \prod_{j=1}^N p(L_j\cond \bmphi) \\
  p(\bml\cond n,\bmphi) &= \prod_{j=1}^n p(\ell_j\cond \bmphi)
\end{align}
In other words: Once we have decided on a prior $p(\bmphi)$ and a
specific pdf to model the likelihood $p(L_j\cond \bmphi)$, we have a
prediction for $L$ which takes into account all possible $N$,
$\lambda$ and $\bmphi$.

\subsubsection{Moment generating function}
\label{sec:mgff}

\texttt{COMMENT Since the TARDIS ``data'' for $L_j$ is highly
  asymmetric, the model likelihood\\ $p(L_j\cond \bmphi)$ must also
  be. Hence we have to determine if a description based on moments is
  relevant! It could be that our numbers are a very bad description of
  the posterior.}
\\

It may be possible to calculate the prediction in Eq.~(\ref{pmp})
numerically, but it will be slow due to the multiple $d\bmL$
integral. A partial solution lies in at least finding simple
expressions for the moments and cumulants (mean, variance, skewness,
kurtosis etc) of $p(L\cond n,\ell)$. This can be accomplished by first
finding the moment generating function (mgf) (see Appendix) which in
the present case is also the Laplace transform.  For the mgf, the
convolution is easily accomplished:
\begin{align}
  \lleq{mgc}
  M(t;L\cond n,\bml) &= M[t\cond p(L|n,\bml)]
  \ =\ \int_0^\infty dL\,p(L\cond n,\bml)\,e^{-tL} \\
  &= \frac{1}{p(\bml\cond n)}
  \sum_N p(N\cond n)\,\int dL e^{-tL} \int d\bmL\,\delta(L - \textstyle\sum_j L_j)
  \displaystyle\int d\bmphi\,p(\bmL\cond N,\bmphi) \, p(\bml\cond n,\bmphi)
  \, p(\phi)\\
  &= \frac{1}{p(\bml\cond n)}
  \sum_N p(N\cond n)\,\int d\bmphi\,
  \prod_{j=1}^N \left[ \int dL_j\,p(L_j\cond \bmphi)\,e^{-tL_j}\right]
   p(\bml\cond n,\bmphi) \, p(\phi)
\end{align}
The term in the square bracket is the moment generating function of
$p(L_j\cond \bmphi)$,
\begin{align}
  \lleq{mgcb}
  M(t; L_j\cond \bmphi) &=
  \int dL_j\,p(L_j\cond \bmphi)\,e^{-tL_j}
\end{align}
and since this is the same for all packets $L_j$, we obtain
\begin{equation}
  \lleq{mgd}
  \boxed{
  M(t;L\cond n,\bml) %
  = \frac{1}{p(\bml\cond n)}
  \sum_N p(N\cond n)\,\int d\bmphi\,
  M(t; L_j\cond \bmphi)^N
   p(\bml\cond n,\bmphi) \, p(\phi)
   }
\end{equation}
As shown in Appendix ***, order-$q$ moments (expectation values of
$L_j^q$) of $p(L\cond n,\bml)$ can be found by taking successive
derivatives with respect to $t$,
\begin{align}
  \lleq{mge}
  E[L^q\cond n,\bml]
  &= \left(\frac{-\partial}{\partial t}\right)^q M(t;L\cond n,\bml)\biggr|_{t=0}
  \qquad q = 1,2,3\ldots
\end{align}
from which we can find the mean, variance, skewness etc of
$p(L|n,\bml)$ without knowing its form explicitly. To do so, we need
to find successive derivatives of $\left[ M(t; L_j\cond \bmphi)
\right]^N$ appearing in (\ref{mgd}). Defining the moments of
$p(L_j\cond\bmphi)$ as
\begin{align}
  \lleq{mgf}
  \mu_q(\bmphi) &= \int dL_j\,p(L_j\cond\bmphi)\,L_j^q \qquad q = 1,2,\ldots,
\end{align}
we find
\begin{align}
  \lleq{mgg}
  \left(\frac{-\partial}{\partial t}\right) %
  M(t; L_j\cond \bmphi)^N\biggr|_{t=0}
  &= N \mu_1(\bmphi)\\
  \lleq{mgh}
  \left(\frac{-\partial}{\partial t}\right)^2 %
  M(t; L_j\cond \bmphi)^N\biggr|_{t=0}
  &= N^{\underline{2}}\mu_1^2(\bmphi) + N \mu_2(\bmphi) \\
  \lleq{mgi}
  \left(\frac{-\partial}{\partial t}\right)^3 %
  M(t; L_j\cond \bmphi)^N\biggr|_{t=0}
  &= N^{\underline{3}}\mu_1^3(\bmphi) + 3N^{\underline{2}}\mu_1(\bmphi)\mu_2(\bmphi)
  + N\mu_3(\bmphi)
\end{align}
where
\begin{align}
  \lleq{mgj}
  N^{\underline q} &= N(N-1)\cdots (N-q+1) = \frac{N!}{(N-q)!}
\end{align}
is the $q$th falling power of $N$. Successively inserting these and
(\ref{mgd}) into (\ref{mge}), we firstly get for the mean of $L$,
given $(n,\bml)$
\begin{align}
  \lleq{mgk}
  E[L\cond n,\bml]
  &= \frac{1}{p(\bml\cond n)}
  \sum_N p(N\cond n)\,N \int d\bmphi\, \mu_1(\bmphi) \,
  p(\bml\cond n,\bmphi) \, p(\phi).
\end{align}
Given that $p(N\cond n)$ is a Negative Binomial pdf, the sum is trivially
found to be
\begin{align}
  \lleq{mgl}
  \sum_N p(N\cond n)\,N &= n\frac{\tfrac{1}{2}}{1-\tfrac{1}{2}} = n
\end{align}
so that
\begin{align}
  \lleq{mgm}
  E[L\cond n,\bml]
  &= \frac{n}{p(\bml\cond n)}
  \int d\bmphi\, \mu_1(\bmphi) \, p(\bml\cond n,\bmphi) \, p(\phi)
\end{align}
Since the integral is an average of $\mu_1(\bmphi)$ weighted by
$p(\bml,\bmphi\cond n)$, let us give it a name. Given a function
$f(\bmphi)$, its ``$\bmphi$-expectation value'' is defined as
\begin{align}
  \lleq{mgn}
  E_\phi[ f ] &= \int d\bmphi\, f(\bmphi) \, p(\bml\cond n,\bmphi) \, p(\phi)
\end{align}
in terms of which
\begin{align}
  % \lleq{mgpa}
  % p(\bml\cond n) &= E_\phi[\mu_0],\\
  %%%%%%%
  \lleq{mgo}
  E[L\cond n,\bml]
  &= n\, \frac{E_\phi[\mu_1]}{p(\bml\cond n)}
  \ =\ n\,\frac{E_\phi[\mu_1]}{E_\phi[\mu_0]}
\end{align}
On taking the second derivative and using
\begin{align}
  \lleq{mgp}
  \sum_N p(N\cond n)\,N^2
  &= \frac{n(n+1)\left(\tfrac{1}{2}\right)^2}{\left(1-\tfrac{1}{2}\right)^2}
  = n(n+1)
\end{align}
the second moment of $p(L\cond n,\bml)$ is
\begin{align}
  \lleq{mgq}
  E[L^2\cond n,\bml]
  &= \frac{1}{p(\bml\cond n)}
  \sum_N p(N\cond n)\, \int d\bmphi\,
  \left[ N^{\underline{2}}\mu_1^2(\bmphi) + N \mu_2(\bmphi)\right]
  p(\bml\cond n,\bmphi) \, p(\phi) \\
  &= n^2\,\frac{E_\phi[\mu_1^2]}{E_\phi[1]}
  + n \frac{E_\phi[\mu_2]}{E_\phi[1]}
\end{align}
so that the variance is
\begin{align}
  \lleq{mgr}
  \var(L\cond n,\bml)
  &= n^2\, \frac{E_\phi[\mu_1^2]\,E_\phi[1] -  E_\phi[\mu_1]^2}
  {E_\phi[1]^2}
  + n\, \frac{E_\phi[\mu_2]}{E_\phi[1]}
\end{align}
It is tempting to argue that the $n^2$ term is much larger than the
$n$-order term. This depends, however, on the sizes of the respective
expectation values: no a priori assumptions can be made here.

\subsection{Third-order: skewness}

TODO Hans

This may be necessary because $p(L_j\cond\bmphi)$ is highly
asymmetric, while the pdf of the sum $L$ may be less so.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Modeling the TARDIS packet distribution}

TODO: All (numerics and analytics)

\begin{mtemize}
\item Find the most suitable likelihood for $p(L_j\cond \bmphi)$

\item Determine how far we can integrate analytically over the
parameters

\item How stable are the results to perturbations?

\item \texttt{COMMENT Since $p(L_j\cond \bmphi)$ is highly asymmetric,
    we have to determine if a description based on moments is
    relevant! It could be that our numbers are a very bad description
    of the posterior.}

\end{mtemize}

\subsection{Results for Beta distribution likelihood}

The Beta probability density function (pdf) is
\begin{align}
  \lleq{btc}
  p(x\cond a,b) &= \frac{x^{a-1}(1-x)^{b-1}}{B(a,b)} \qquad
  0 < x < 1, \quad a,b > 0.
\end{align}
where $\bmphi = (a,b)$ are the parameters and
\begin{align}
  \lleq{btd}
  B(a,b) &= \frac{\Gamma(a)\,\Gamma(b)}{\Gamma(a+b)}
\end{align}
is the beta function.
%
The form of the TARDIS single-packet distribution suggests that this
is appropriate; the strong asymmetry in the TARDIS distribution leads
us to expect most likely parameter values of $b \ll a$ so that the mode
is close to $x=1$.
%
The variable $x$ is related to packet luminosities simply by
$x = L_j/L_{\rm max}$ where $L_{\rm max}$ is the largest luminosity
in the packets i.e.\ $0 < L_j < L_{\rm max}$.
%
(Remember we are considering a single bin $b$ only, and only a single
TARDIS run.)
%
Given that we want to find the $\phi$-moments of Section \ref{sec:mgff}
in terms of TARDIS simulation ``data'', we define in this section
\begin{align}
  x_j &= \ell_j/\ell_{\rm max}
\end{align}
while $x$ without subscript is used for the generic moment
calculations. The same quantities could, of course, be used in terms
of the generic variable $x_j = L_j/L_{\rm max}$. In terms of the
original ``data'' packets, the beta distribution reads
\begin{align}
  \lleq{bte}
  p(\bml\cond \ell_{\rm max},n,a,b)
  &= \prod_{j=1}^n p(\ell_j\cond \ell_{\rm max},a,b) \\
  \lleq{btf}
  p(\ell_j\cond \ell_{\rm max},a,b)
  &= \frac{1}{\ell_{\rm max}\,B(a,b)}
  \,\left(\frac{\ell_j}{\ell_{\rm max}}\right)^{a-1}
    \left(1-\frac{\ell_j}{\ell_{\rm max}}\right)^{b-1}
  \ =\ \frac{1}{B(a,b)}
  \frac{\ell_j^{a-1}\,(\ell_{\rm max}-\ell_j)^{b-1}}
  {\ell_{\rm max}^{a+b-1}}
\end{align}
equivalent to the form (\ref{btc}) with $x\to x_j$.
% \begin{align}
%   p(x_j\cond a,b) &= \frac{1}{B(a,b)} x_j^{a-1} (1-x_j)^{b-1}
% \end{align}
NOTE: If $\ell_{\rm max}$ changes significantly from TARDIS run to
run, then the probability calculation may have to be changed from
$p(\bml\cond \ell_{\rm max},n,a,b)$ to\\
$p(\bml\cond \ell_{\rm max},n,a,b)\,p(\ell_{\rm max}\cond
n,a,b)$. However we are supposed to deal with ONE TARDIS run, so that
$\ell_{\rm max}$ is known and fixed.
\\

The moments of the beta distibution are simple,
\begin{align}
  \lleq{btg}
  \mu_q(a,b) &= \int dx\,x^q\,p(x\cond a,b)
  = \frac{(a)_q}{(a+b)_q}
  = \frac{a(a+1)\cdots(a+q-1)} {(a+b)(a+b+1)\cdots(a+b+q-1)}
  \qquad q = 0,1,2,\ldots
\end{align}
with $(a)_b$ the Pochhammer symbol.
%
\\

\textbf{Calculations for a uniform discrete prior}

The prior is as yet unspecified. Since there is no known conjugate
prior for the beta distribution, we resort to a simpler prior in terms
of discrete values for $a$ and $b$. In principle, these can be any
rational numbers, but because as stated the value of $b$ is expected
to be around 30--60, we let $a$ and $b$ be integers only, $a,b, =
1,2,3,\ldots$. Furthermore, since Eqs.~(\ref{mgo}) and (\ref{mgr})
always involve \textbf{ratios} of $\phi$-expectation values, we can
use a uniform improper prior,
\begin{align}
  p(a,b) &= C\qquad a,b = 1,2,\ldots,\infty
\end{align}
with $C$ the improper constant which will cancel in the ratio.
% With
% this prior, the $\phi$-expectation values become
% \begin{align}
%   E_\phi[\mu_q] &= C \sum_{a=1}^\infty \sum_{b=1}^\infty
%   \mu_q(a,b)\,\prod_j \frac{x_j^{a-1}(1-x_j)^{b-1}}{B(a,b)}
% \end{align}
We now calculate $\phi$-expectation values for this case, namely
\begin{align}
  \lleq{bth}
  E_\phi[\mu_q(a,b)] &= \sum_{a,b}\mu_q(a,b)\,p(\bml\cond \ell_{\rm max}, n,a,b)
  \,p(a,b)
  \\
  % = C \sum_{a,b=1}^\infty \mu_q(a,b)\,p(\bml\cond \ell_{\rm max}, n,a,b)\\
  \lleq{bti}
  &= C\sum_{a,b} \frac{(a)_q}{(a+b)_q}
  \frac{1}{B(a,b)} \prod_{j=1}^n \left(x_j^{a-1}(1-x_j)^{b-1}\right)
\end{align}
The product can be rewritten as
\begin{align}
  \lleq{btj}
  \prod_{j=1}^n \left(x_j^{a-1}(1-x_j)^{b-1}\right)
  &= \left(\prod_j x_j\right)^{a-1}\left(\prod_j(1-x_j)\right)^{b-1}
\end{align}
and defining
\begin{align}
  \lleq{btk}
  y &= \prod_{j=1}^n x_j \qquad z = \prod_{j=1}^n(1-x_j)
\end{align}
which are data quantities independent of any model or model
parameters, we get
\begin{align}
  \lleq{btl}
  E_\phi[\mu_q(a,b)] &= C\sum_{a,b} \frac{(a)_q}{(a+b)_q}
  \frac{y^{a-1}z^{b-1}}{B(a,b)}
\end{align}
With the help of mathematica we find
\begin{align}
  \lleq{btm}
  E_\phi[\mu_0] &= C\sum_{a,b=1}^\infty \frac{y^{a-1}z^{b-1}}{B(a,b)}
  = \frac{C}{(1 - y - z)^2}
  \\
  E_\phi[\mu_1] &= C\sum_{a,b} \frac{a}{(a+b)}
  \frac{y^{a-1}z^{b-1}}{B(a,b)}
  \nonumber\\
  \lleq{btn}
  &= \frac{C}{(y + z)^3}
  \left[
  \frac{(y + z) [2 y^2 + (z-y)(z-1)]}
  {(1 - y - z)^2 }
   + (z-y) \ln(1 - y - z)
   \right]
%   \\
%   E_\phi[\mu_2(a,b)] &= C\sum_{a,b} \frac{a(a+1)}{(a+b)(a+b+1)}
%   \frac{(1-y)^{a-1}(1-z)^{b-1}}{B(a,b)}
%   \\
%   E_\phi[\mu_1^2(a,b)] &= C\sum_{a,b} \left(\frac{a}{a+b}\right)^2
%   \frac{(1-y)^{a-1}(1-z)^{b-1}}{B(a,b)}
\end{align}
Taking ratios, we obtain
\begin{align}
  \lleq{bto}
  \frac{E_\phi[\mu_1(a,b)]} {E_\phi[\mu_0(a,b)]}
  &= \frac{(y + z) [2 y^2 +(z- y) (1 - z)] + (z-y) (1 - y - z)^2 \ln(1 - y - z)}
  {(y + z)^3}
\end{align}
and so
\begin{align}
  \lleq{btp}
  E[L\cond n,\bml] &= n \left[
  \frac{(y + z) [2 y^2 +(z- y) (1 - z)] + (z-y) (1 - y - z)^2 \ln(1 - y - z)}
  {(y + z)^3}\right]
\end{align}
is completely determined by the data $(n,\bml)$.  Similarly
\begin{align}
  E_\phi[\mu_2] &= C\sum_{a,b} \frac{a(a+1)}{(a+b)(a+b+1)}
  \frac{y^{a-1}z^{b-1}}{B(a,b)}
  \nonumber\\
  \lleq{btq} &= \frac{-1}{(1-y-z)^2 (y+z)^5} %
  \nonumber\\
  &\times\Bigl\{ 2 \Bigl[2 y^2(z+1) + yz(z-8) - (z-2) z^2\Bigr]
  (y+z-1)^2 \ln(-y-z+1)
  \nonumber\\
  &+ (y+z) \Bigl[
    y^4 - 2 y^3 (5z+3) + y^2[(22-19 z) z+4] - 4 yz (z-4) (z-1) + 4z^2 (z-1)^2
           \Bigr] \Bigr\}
  % \nonumber\\
  % &= -(1/((y+z-1)^2 (y+z)^5)) %
  % \nonumber\\
  % &\times \Bigl\{
  %   -4 y^3 + 6 y^4 - y^5 + 12 y^2 z - 16 y^3 z + 9 y^4 z + 12 y z^2 -
  %   42 y^2 z^2 + 29 y^3 z^2 - 4 z^3 - 12 y z^3 + 23 y^2 z^3 + 8 z^4 -
  %   4 z^5 \nonumber\\
  %   &+ \ln[1 - y - z] \Bigl[
  %   -4 y^2 + 8 y^3 - 4 y^4 + 16 y z - 28 y^2 z + 16 y^3 z -
  %     4 y^4 z - 4 z^2 - 26 y z^2 + 36 y^2 z^2 - 10 y^3 z^2 + 10 z^3 +
  %     8 y z^3 - 6 y^2 z^3 - 8 z^4 + 2 y z^4 + 2 z^5
  %     \Bigr]
  %   \Bigr\}
\end{align}
and
\begin{align}
  E_\phi[\mu_1^2] &= C\sum_{a,b} \frac{a^2}{(a+b)^2}
  \frac{y^{a-1}z^{b-1}}{B(a,b)}
  \nonumber\\
  \lleq{btr} %
  &= \frac{1}{y (y - z)^3}
  \Bigl\{
  (y - z) (-y + z + 2 y \ln[y]) + y (-2 + y + z)
  (\mathrm{PolyLog}[2, 1 - y] - \mathrm{PolyLog}[2, 1 - z])
  \Bigr\}
\end{align}
These have to be combined with the above to find the variance of
Eq.~(\ref{mgr}). The hope is that the complicated mathematical
expressions simplify. Either way, the variance is also a closed
solution in terms of $n$, $y$ and $z$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Prediction based on packet sum}

TODO: Hans. The calculations have already been done. The question is
whether we even want to bother with them since data in the form of
packet sums is less information than the vector of individual packets
as used above.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}

TODO: Fred

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion and Conclusions}

TODO: All

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Acknowledgements}

This work was supported in part by the DFG cluster of excellence
``Origin and Structure of the Universe'' (www.universe-cluster.de) and
the National Research Foundation of South Africa, etc etc.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Appendix: Relevant distributions}

TODO: Hans

\subsection{Product rule and sum rule}

\subsection{Poisson}

\subsection{Negative Binomial}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Appendix: Generating functions}

TODO: Hans
$\pi(\alpha, \beta) \propto \left( \frac{\Gamma(\alpha+\beta)}{\Gamma(\alpha) \Gamma(\beta)} \right)^{\lambda +1} (x x_0)^{\alpha-1} \left(y_0 (1-x) \right)^{\beta-1}$

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \section{}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{thebibliography}{99}
\bibitem{kendall}Stuart and Ord, \textit{Kendall's Advanced Theory of Statistics},
  Vol. 1, 5th Edition etc..

\end{thebibliography}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\end{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2016-01-15 hans/bayes-tardis-luminosity-prediction.tex
[prediction] Comment on draft

@@
 
+\usepackage[colorinlistoftodos]{todonotes} % Use 'disable' to remove todos in final version
+\newcommand{\fred}[1]{\todo[inline]{#1}} %
+\newcommand{\checked}[1]{\todo[color=green,noline]{\checkmark}} %
+\newcommand{\checkedl}[1]{\todo[color=brown]{\checkmark}} %
+
 %% CHANGE THIS DATE AS YOU MODIFY THE FILE
-\newcommand{\vdate}{151211}
+\newcommand{\vdate}{160114}
 \pagestyle{myheadings}
@@
   \textbf{\Large based on simulation output: An example from TARDIS}\\[12pt]
-  11 December 2015
 \end{center}
@@
 \end{align}
+  \checked{}
 with
@@
 \end{align}
+  \checked{}
 Given $D$, the total luminosity in bin $b$ is known,
@@
 \end{align}
+  \checked{}
 \end{mtemize}
@@
 of a stochastic variable $N_b$ governed by a model pdf
-$p(N_b)$. Similarly, $\bml_b$ is a specific realization of a
-collection stochastic variables $\bmL_b
+$p(N_b)$.
+\fred{In our general case, $\sum_b n_b$ is an instance of a random variable because some packets leave the supernova, some don't.
+I guess in most applications the total number of events would be chosen and not random.
+We are fortunate that this difference does not matter because we consider bins independently.
+} Similarly, $\bml_b$ is a specific realization of a
+collection of stochastic variables $\bmL_b
 =\{L_{b,1},\ldots,L_{b,\smN_b}\}$, where the number $N_b$ of such
@@
 \end{align*}
-where $\bmphi_b$ is the set of parameters of the model pdf which
+where $\bmphi_b$ is the set of parameters\fred{so $\bmphi$ includes random seed or not? Initially I thought so. But later on it becomes clear that you do not include it.} of the model pdf which
 govern $\bmL_b$ and $\lambda_b$ is the parameter of the model pdf
@@
   multiplicities $N_b$ and total luminosities $\bmL_b$ in any given
-  bin $b$ are statistically independent of those in other bins. This
+  bin $b$ are statistically independent of those in other bins.
+\fred{We could motivate this in the usual way as the Poisson limit of the Binomial.
+The subtlety would then the be the random total number of packets (leaving the supernova).} This
   means that the joint probability over all bins factorises into
@@
 \item We choose $N$ to follow a Poisson distribution with parameter
-  $\lambda$ (motivate choice),
+  $\lambda$ \fred{this choice linked to limit of Multinomial, see above}
   \begin{align}
@@
 
-\item Priors: metaparameter $a$ ***
+\item Priors: metaparameter $a$ \fred{?}
 
@@
   {\int d\lambda\, p(n\cond\lambda)\,p(\lambda)}
-\end{align}
+\end{align}\checked{}
 where $p(N\cond\lambda,n) = p(N\cond\lambda)$ is independent of $n$.
@@
 \end{align}
+\fred{$a=1/2 \Leftrightarrow$ Jeffrey's prior would my choice.
+We then get a Pólya distribution instead of Negative Binomial but nothing substantial changes.
+$$(pmh) \frac{\Gamma(N+n-a+1)}{N!\,\Gamma(n-a+1)\,2^{N+n+1-a}}$$ }
 which we recognise as a Negative Binomial distribution (see appendix)
+\fred{There are several definitions of the Negative Binomial, so we should specify which one we use.
+This one coincides with Wikipedia: $N$ successes given $k$ failures with probability of success $\rho$.}
 with parameters $k=n+1-a$ and $\rho=(1/2)$
@@
 calculations below.  Note that the distribution of $N$ given $n$ is
-independent of the packets $\bmL$ and $\bml$ and their values.
+independent of the packets $\bmL$ and $\bml$ by construction.
 
@@
 \end{align}
+\checked{}
 Hence the prediction is given by
@@
 \end{align}
+\checked{}
 The dependence on the TARDIS ``data'' $(n,\bml)$ enters through
@@
 \end{align}
+\fred{$p(\bmphi\cond N,n,\bml) = p(\bmphi\cond n,\bml)$ }
 and using a multiplicity-independent prior for $\bmphi$
@@
   \displaystyle\int d\bmphi\,p(\bmL\cond N,\bmphi) \, p(\bml\cond n,\bmphi)
-  \, p(\phi)
+  \, p(\bmphi)
   }
 \end{equation}
+\checked{}
 with
@@
 \end{align}
+\fred{And $p(\cdot \cond \bmphi)$ is the same function in both equations.}
 In other words: Once we have decided on a prior $p(\bmphi)$ and a
@@
 finding the moment generating function (mgf) (see Appendix) which in
-the present case is also the Laplace transform.  For the mgf, the
-convolution is easily accomplished:
+the present case is also the Laplace transform. \fred{On Wikipedia,
+  the MGF is defined as $\int_0^\infty dL\,p(L\cond
+  n,\bml)\,e^{+tL}$. Then there are less minus signs to worry about}
+For the mgf, the convolution is easily accomplished:
 \begin{align}
@@
   \displaystyle\int d\bmphi\,p(\bmL\cond N,\bmphi) \, p(\bml\cond n,\bmphi)
-  \, p(\phi)\\
+  \, p(\bmphi)\\
   &= \frac{1}{p(\bml\cond n)}
@@
   \prod_{j=1}^N \left[ \int dL_j\,p(L_j\cond \bmphi)\,e^{-tL_j}\right]
-   p(\bml\cond n,\bmphi) \, p(\phi)
+   p(\bml\cond n,\bmphi) \, p(\bmphi)
 \end{align}
@@
   M(t; L_j\cond \bmphi)^N
-   p(\bml\cond n,\bmphi) \, p(\phi)
+   p(\bml\cond n,\bmphi) \, p(\bmphi)
    }
  \end{equation}
+ \checked{}
 As shown in Appendix ***, order-$q$ moments (expectation values of
@@
 \end{align}
+\checked{}
 where
@@
   \sum_N p(N\cond n)\,N \int d\bmphi\, \mu_1(\bmphi) \,
-  p(\bml\cond n,\bmphi) \, p(\phi).
+  p(\bml\cond n,\bmphi) \, p(\bmphi).
 \end{align}
@@
 \end{align}
+\fred{The first moment under NB is $\frac{\rho k}{1-\rho}$, resulting in $n-a+1$. Here and below you seem to set $a=1$ implicitly. I want to carry the $a$ along.}
 so that
@@
   &= \frac{n}{p(\bml\cond n)}
-  \int d\bmphi\, \mu_1(\bmphi) \, p(\bml\cond n,\bmphi) \, p(\phi)
+  \int d\bmphi\, \mu_1(\bmphi) \, p(\bml\cond n,\bmphi) \, p(\bmphi)
 \end{align}
 Since the integral is an average of $\mu_1(\bmphi)$ weighted by
-$p(\bml,\bmphi\cond n)$, let us give it a name. Given a function
+$p(\bmphi\cond n, \bml)$, let us give it a name. Given a function
 $f(\bmphi)$, its ``$\bmphi$-expectation value'' is defined as
@@
   \lleq{mgn}
-  E_\phi[ f ] &= \int d\bmphi\, f(\bmphi) \, p(\bml\cond n,\bmphi) \, p(\phi)
+  E_\phi[ f ] &= \int d\bmphi\, f(\bmphi) \, p(\bml\cond n,\bmphi) \, p(\bmphi)
 \end{align}
@@
 \end{align}
+\fred{$n \to n-a+1$}
 On taking the second derivative and using
@@
 \end{align}
+\fred{I disagree. The 2nd raw moment is $\frac{k \rho (1 + k
+    \rho)}{(1-\rho)^2}$ (from Mathworld with $\rho=q,1-\rho=p, k=r$
+  and manually derived from the negative binomial MGF from
+  Wikipedia). With $k=n-a+1$ and $\rho=1/2$, I get $(n-a+1)(n-a+3)$}
 the second moment of $p(L\cond n,\bml)$ is
@@
   \left[ N^{\underline{2}}\mu_1^2(\bmphi) + N \mu_2(\bmphi)\right]
-  p(\bml\cond n,\bmphi) \, p(\phi) \\
+  p(\bml\cond n,\bmphi) \, p(\bmphi) \\
   &= n^2\,\frac{E_\phi[\mu_1^2]}{E_\phi[1]}
@@
 \end{align}
+\fred{I have
+ $$
+  \frac{n-a+1}{E_\phi[1]} \left( (n-a+2)E_\phi[\mu_1^2] + E_\phi[\mu_2]\right)
+$$
+}
 so that the variance is
@@
 expectation values: no a priori assumptions can be made here.
+\fred{I have the more complicated
+  $$
+    \var(L\cond n,\bml) = \frac{n-a+1}{E_\phi[1]^2} \left( (n-a+2)E_\phi[\mu_1^2] E_\phi[1] -(n-a+1) E_\phi[\mu_1]^2 + E_\phi[\mu_2] E_\phi[1]\right)
+  $$
+}
 
@@
 asymmetric, while the pdf of the sum $L$ may be less so.
+
+\fred{I agree but what would we do if we knew the skewness?
+For the fit, we need to evaluate the likelihood, so we'd have to build a function my moment matching.
+Of course with mean and variance alone, we'd use the Gaussian.
+But this just reduces to the central limit theorem.}
 
@@
 \end{mtemize}
+
+\fred{Yes but at least if a rather large number of events is in the
+  bin, this will be good enough}
 
@@
 \end{align}
+\checked{}
 equivalent to the form (\ref{btc}) with $x\to x_j$.
@@
 $\ell_{\rm max}$ is known and fixed.
-\\
+\fred{Rather: we approximate $p(\ell_{\rm max} \cond n, a,b)$ by $\delta(\max \ell_{b,j})$.
+This is OK is $\sum_b n_b$ is large.
+For numerical stability, we want to ensure that $x<1$, so we take the largest $\ell_{b,j}$ and multiply by $1 +\varepsilon, \varepsilon = 10^{-7}$}
 
@@
 \\
+\fred{Formula ok but to be consistent with the notation used above, we should either use the rising factorial $a^{\overline{b}}$ here or $(a)^b$ above}
 
@@
 The prior is as yet unspecified. Since there is no known conjugate
-prior for the beta distribution, we resort to a simpler prior in terms
-of discrete values for $a$ and $b$. In principle, these can be any
+prior for the beta distribution,\fred{yes it is known. But it not
+  useful because the normalization can't be calculated analytically;
+  see
+  http://mathoverflow.net/questions/20399/conjugate-prior-of-the-dirichlet-distribution}
+we resort to a simpler prior in terms of discrete values for $a$ and
+$b$ to simplify the expressions. In principle, these can be any
 rational numbers, but because as stated the value of $b$ is expected
 to be around 30--60, we let $a$ and $b$ be integers only, $a,b, =
-1,2,3,\ldots$. Furthermore, since Eqs.~(\ref{mgo}) and (\ref{mgr})
-always involve \textbf{ratios} of $\phi$-expectation values, we can
-use a uniform improper prior,
+1,2,3,\ldots$. \fred{I don't understand this argument} Furthermore,
+since Eqs.~(\ref{mgo}) and (\ref{mgr}) always involve \textbf{ratios}
+of $\phi$-expectation values, we can use a uniform improper prior,
 \begin{align}
@@
 \end{align}
+\checked{}
 The product can be rewritten as
@@
 \end{align}
+\fred{This equation can be skipped}
 and defining
@@
 \end{align}
+\checked{}
 which are data quantities independent of any model or model
@@
 \end{align}
+\checked{}
 With the help of mathematica we find
@@
  \end{align}
+ \checked{}
 Taking ratios, we obtain
@@
 as used above.
+\fred{No, we don't want to bother.}
 

### 2016-01-16 hans/bayes-tardis-luminosity-prediction.tex
[results] Add fits of beta to samples

@@
 \begin{align}
-  \lleq{bth}
+  \lleq{bthi}
   E_\phi[\mu_q(a,b)] &= \sum_{a,b}\mu_q(a,b)\,p(\bml\cond \ell_{\rm max}, n,a,b)
@@
 
-TODO: Fred
+In Figure~\ref{fig:beta} is a fit of the Beta distribution to 3000
+samples from a TARDIS run with frequencies in the range $[1.367,
+1.435] \times 10^{38}$ Hz. A characteristic feature seen in many such
+similar fits is that the Beta distribution does not give a perfect
+fit: it undershoots around the mode and overshoots around
+$x=0.96$. The reason may be that the Beta itself is not flexible
+enough to describe the data, or the distribution is simply not
+constant in frequency, so we should choose smaller bins or less
+samples. We don't expect 3000 samples in every bin, so what is the
+minimum number of samples needed to work robustly? The first numerical
+challenge is how to normalize the luminosities, I take the maximum
+across all samples (not just one bin) and multiply by $10^{-6}$ to
+avoid $x=1$ for any sample. For too few samples, the fit returns a pdf
+that peaks at the boundary $x=1$, this is not what we want. From trial
+and error, 300 samples is a solid minimum to get a robust fit if
+parameters are estimated via maximum likelihood. The method-of-moment
+estimator has the advantage of not requiring an optimization algorithm
+but for some bins it gives a solution peaking at $x=1$ even for 800
+samples so it's not robust.
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=0.48\textwidth]{beta-300}
+  \includegraphics[width=0.48\textwidth]{beta-3000}
+  \caption{Beta distribution fit. The samples are binned in the
+    histogram. The blue solid curve is a kernel-density estimate, the
+    red dashed curve is the Beta distribution. left: 300 samples,
+    right: 3000 samples}
+  \label{fig:beta}
+\end{figure}
 

### 2016-01-26 hans/bayes-tardis-luminosity-prediction.tex
[draft] separate #packets for paramater and Poisson inference

@@
 
+\fred{We should be more general: let's reserve $n$ for the number of
+  packets observed in a bin. Suppose this is $n=1$. From that, we can
+  hardly infer the luminosity distribution. Instead, we group the
+  packets in larger bins such that there are at least $m \gtrsim 300$
+  packets, and then assume the distribution doesn't change much with
+  frequency.
+
+  Hence we will learn $\bmphi$ from $m$ samples, and
+  $p(\bmphi\cond N,n,\bml) \to p(\bmphi\cond N,m,\bml)$ and we can
+  keep $\bml$ with the same meaning. So we replace $n \to m$ below
+  except in $p(N \cond n$)}
+
 Our goal of prediction of $L$ given the data $(n,\bml)$ is captured in

### 2016-01-26 hans/bayes-tardis-luminosity-prediction.tex
[hans] dos2unix line endings cleanup
[draft] Start Gamma section

@@
 
+\DeclareMathOperator{\GammaDist}{Gamma}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
+\subsection{Results for the Gamma-distribution likelihood}
+
+The Gamma distribution is with shape parameter $\alpha$ and rate
+parameter $\beta$ is defined as
+\begin{align}
+  \lleq{gma}
+  p(x | \alpha, \beta) = \frac{\beta^\alpha}{\Gamma(\alpha)} x^{\alpha-1} e^{-\beta x} .
+\end{align}
+For our purposes, the most important property is that the Gamma distribution is infinitely divisible; i.e., the sum of $N$ Gamma variates is again a Gamma variate
+\begin{align}
+  \lleq{gmb}
+  X_i \sim Gamma(\alpha, \beta) \Rightarrow \sum_{i=1}^N X_i \sim Gamma(N \alpha, \beta)
+\end{align}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2016-01-27 hans/bayes-tardis-luminosity-prediction.tex
[hans] expand gamma subsection

@@
 \usepackage[a4paper, left=20mm, right=20mm, top=25mm, bottom=25mm]{geometry}
+\usepackage[numbers]{natbib}
 
@@
 \newcommand{\fred}[1]{\todo[inline]{#1}} %
-\newcommand{\checked}[1]{\todo[color=green,noline]{\checkmark}} %
-\newcommand{\checkedl}[1]{\todo[color=brown]{\checkmark}} %
+\newcommand{\checked}{\todo[color=green,noline]{\checkmark}} %
+\newcommand{\checkedl}{\todo[color=brown]{\checkmark}} %
 
@@
 \DeclareMathOperator{\GammaDist}{Gamma}
+\newcommand{\refeq}[1]{Eq.~(\ref{#1})}
 
@@
 
+With $\bmphi = (\alpha, \beta)$, the property \refeq{gmb} is
+equivalent to
+\begin{align}
+  \lleq{gmc}
+ p(L \cond N, \bmphi) &= \int d\bmL\,\delta(L - \textstyle\sum_j L_j)
+  \,p(\bmL\cond N,\bmphi) = Gamma(L \cond N \alpha, \beta)
+\end{align}
+and this implies that the master equation~(\ref{pmp}) considerably simplifies as
+\begin{align}
+  \lleq{gmd}
+    p(L\cond n,\bml,m)
+  = \frac{1}{p(\bml\cond n)}
+  \sum_N p(N\cond n)\, \int d\bmphi\, p(L \cond \bmphi)\, p(\bml\cond m,\bmphi)
+  \, p(\bmphi).
+\end{align}
+
+To explicitly evaluate \refeq{gmd}, we have to overcome the remaining
+two difficulties: the sum over $N$ and the integral over
+$\bmphi$. Regarding the latter, there is a conjugate prior but
+unfortunately the normalization constant can not be evaluated in
+closed form. It is thus of no use for us, and we turn to a
+noninformative prior. In \cite{Moala:2013}, an overview of possible
+choices is given. They conclude that the point estimate given by the
+mode of the posterior is nearly insensitive to the prior choice for 30
+samples. We require in addition that the posterior is narrow enough so
+we group samples in bins to have at least $m=150$ samples in each
+bin. The following approximations can be used in different regions of
+the parameter space spanned by $n$ and $m$.
+
+\begin{enumerate}
+  \item
+\end{enumerate}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{thebibliography}{99}
-\bibitem{kendall}Stuart and Ord, \textit{Kendall's Advanced Theory of Statistics},
-  Vol. 1, 5th Edition etc..
-
-\end{thebibliography}
+\bibliographystyle{plain}
+\bibliography{references}
 

### 2016-01-27 hans/bayes-tardis-luminosity-prediction.tex
[hans] Add first Gamma approximations

@@
 \usepackage[numbers]{natbib}
+%\usepackage{hyperref}
 
@@
 \newcommand{\refeq}[1]{Eq.~(\ref{#1})}
+\newcommand{\lmax}{\ell_{\rm max}}
 
@@
 The Gamma distribution is with shape parameter $\alpha$ and rate
-parameter $\beta$ is defined as
+parameter $\beta$ is defined for $x \geq 0$ as
 \begin{align}
   \lleq{gma}
-  p(x | \alpha, \beta) = \frac{\beta^\alpha}{\Gamma(\alpha)} x^{\alpha-1} e^{-\beta x} .
-\end{align}
-For our purposes, the most important property is that the Gamma distribution is infinitely divisible; i.e., the sum of $N$ Gamma variates is again a Gamma variate
+  \GammaDist(x | \alpha, \beta) = \frac{\beta^\alpha}{\Gamma(\alpha)} x^{\alpha-1} e^{-\beta x} .
+\end{align}
+$\GammaDist(x \cond \alpha, \beta)$ is defined for $x \in [0, \infty]$
+whereas the luminosity of real packets is bounded from above and has a
+peak away from the bound. In order to describe the luminosity
+distribution of real packets through $\GammaDist(\alpha, \beta)$, we
+have to transform the luminosities. To this end, we first take the
+largest luminosity from the entire frequency range and increase it
+slightly to avoid numerical difficulties at the end point
+\begin{align}
+  \lleq{gme}
+  \lmax \equiv (1+\epsilon) \times \max_j \ell_j,
+\end{align}
+where $ \epsilon=10^{-6}$.
+We assume the total number of samples across all bins is large, hence
+the effect of considering $\lmax$ a constant independent of the data
+is negligible. We then transform the samples as
+\begin{align}
+  \lleq{gmf}
+  x_j \equiv 1- \frac{\ell_j}{\lmax} > 0
+\end{align}
+and similarly for $L \to X$, then we infer $P(X \cond n, \bmx,
+m)$. Only in the end, we transform back to a probability density in
+$L$ via
+\begin{align}
+  \lleq{gmg}
+  p(L  \cond n, \bml, m) = p(X  \cond n, \bmx, m) \frac{1}{\lmax}
+\end{align}
+While not obvious at first, a major numerical simplification arises
+because of the $L \to X$ transformation. Rewriting the master
+equation~(\ref{pmp}) formally as
+\begin{align}
+  \lleq{gmh}
+  p(L  \cond n, \bml, m) = \sum_{N} p(N \cond n) p(L \cond N, \bml, m)
+\end{align}
+we see that the $ p(L \cond n, \bml, m)$ is a mixture density. If the
+distribution of a single luminosity, $p(L \cond N=1, \bml, m)$ is
+narrow in the sense that the standard deviation relative to the mean
+is at the percent level, then the mixture density has one mode for
+each $N$ and all modes are clearly separated. The transformation $L
+\to x$ effects a strong overlap between adjacent $p(X \cond N, \bmx,
+m)$ and leads to a unimodal $p(X \cond n, \bmx, m)$.
+\fred{Show a plot: from hedgehog to single peak}
+
+For our purposes, the most useful property of the Gamma
+distribution is infinite divisibility; i.e., the sum of $N$ Gamma
+variates is again a Gamma variate
 \begin{align}
   \lleq{gmb}
-  X_i \sim Gamma(\alpha, \beta) \Rightarrow \sum_{i=1}^N X_i \sim Gamma(N \alpha, \beta)
-\end{align}
-
-With $\bmphi = (\alpha, \beta)$, the property \refeq{gmb} is
-equivalent to
+  X_i \sim \GammaDist(\alpha, \beta) \Rightarrow \sum_{i=1}^N X_i \sim \GammaDist(N \alpha, \beta).
+\end{align}
+With $\bmphi = (\alpha, \beta)$, \refeq{gmb} is equivalent to
 \begin{align}
   \lleq{gmc}
- p(L \cond N, \bmphi) &= \int d\bmL\,\delta(L - \textstyle\sum_j L_j)
-  \,p(\bmL\cond N,\bmphi) = Gamma(L \cond N \alpha, \beta)
+ p(X \cond N, \bmphi) &= \int d\bmX\,\delta(X - \textstyle\sum_j X_j)
+  \,p(\bmX\cond N,\bmphi) = \GammaDist(X \cond N \alpha, \beta)
 \end{align}
@@
   \lleq{gmd}
-    p(L\cond n,\bml,m)
-  = \frac{1}{p(\bml\cond n)}
-  \sum_N p(N\cond n)\, \int d\bmphi\, p(L \cond \bmphi)\, p(\bml\cond m,\bmphi)
+    p(X\cond n,\bmx,m)
+  = \frac{1}{p(\bmx\cond m)}
+  \sum_N p(N\cond n)\, \int d\bmphi\, p(X \cond N, \bmphi)\, p(\bmx\cond m,\bmphi)
   \, p(\bmphi).
@@
 closed form. It is thus of no use for us, and we turn to a
-noninformative prior. In \cite{Moala:2013}, an overview of possible
-choices is given. They conclude that the point estimate given by the
-mode of the posterior is nearly insensitive to the prior choice for 30
-samples. We require in addition that the posterior is narrow enough so
-we group samples in bins to have at least $m=150$ samples in each
-bin. The following approximations can be used in different regions of
-the parameter space spanned by $n$ and $m$.
+noninformative prior. The authors of \cite{moala2013bayesian} present
+an overview of possible choices and conclude that the point estimate
+given by the mode of the posterior is nearly insensitive to the prior
+choice for 30 samples. We require in addition that the posterior is
+narrow enough so we group samples in bins to have at least $m=150$
+samples in each bin. The following approximations can be used in
+different regions of the parameter space spanned by $n$ and $m$.
 
 \begin{enumerate}
-  \item
+  \item For $m$ large, the posterior $p(\bmphi \cond m, \bmx) =
+  p(\bmx\cond m,\bmphi) \, p(\bmphi) / p(\bmx \cond m)$ approaches a
+  multivariate Gaussian with variance tending to zero as $1/m$ in the
+  limit $m \to \infty$.  \fred{check} We can then assume the
+  parameters $\bmphi$ known and fix them at the posterior mode
+  $\bmphi^*$; i.e., $p(\bmphi \cond m, \bmx) \to \delta(\bmphi -
+  \bmphi^*)$.
+  \item For $n$ large, the Negative Binomial $p(N \cond n)$ approaches
+  a Gaussian, defining a region around $N=n$ with the dominant
+  contribution to the sum over $N$ in \refeq{gmd}. The central limit
+  theorem states that for large $N$, $P(X \cond N, \bmphi) \approx
+  \mathcal{N}(X \cond N \, E_\bmphi[X_j], N \, V_\bmphi[X_j])$. For
+  $m$ large enough, we could then plug in the sample-mean and
+  sample-variance estimates and bypass the assumption of a Gamma
+  distribution altogether.
 \end{enumerate}

### 2016-02-03 hans/bayes-tardis-luminosity-prediction.tex
[hans] Add section on approximations
* synctex support, requires pdflatex
* bibtex references

@@
 %\usepackage{hyperref}
+\synctex=1
 
@@
 \newcommand{\refeq}[1]{Eq.~(\ref{#1})}
+\newcommand{\reffig}[1]{Fig.~\ref{fig:#1}}
 \newcommand{\lmax}{\ell_{\rm max}}
@@
 
-The Gamma distribution is with shape parameter $\alpha$ and rate
-parameter $\beta$ is defined for $x \geq 0$ as
+The Gamma distribution  with shape parameter $\alpha$ and rate
+parameter $\beta$ is defined
 \begin{align}
@@
 \end{align}
-$\GammaDist(x \cond \alpha, \beta)$ is defined for $x \in [0, \infty]$
+The support of $\GammaDist(x \cond \alpha, \beta)$ is $[0, \infty]$
 whereas the luminosity of real packets is bounded from above and has a
-peak away from the bound. In order to describe the luminosity
+peak close to but not at the boundary. In order to describe the luminosity
 distribution of real packets through $\GammaDist(\alpha, \beta)$, we
@@
 \end{align}
-While not obvious at first, a major numerical simplification arises
-because of the $L \to X$ transformation. Rewriting the master
+We observe good agreement between the best fit of a Gamma distribution
+to TARDIS samples; an example is shown shown in \reffig{gamma}.  While
+not obvious at first, a major numerical simplification arises because
+of the $L \to X$ transformation. Rewriting the master
 equation~(\ref{pmp}) formally as
@@
 \end{align}
-we see that the $ p(L \cond n, \bml, m)$ is a mixture density. If the
-distribution of a single luminosity, $p(L \cond N=1, \bml, m)$ is
-narrow in the sense that the standard deviation relative to the mean
-is at the percent level, then the mixture density has one mode for
-each $N$ and all modes are clearly separated. The transformation $L
-\to x$ effects a strong overlap between adjacent $p(X \cond N, \bmx,
-m)$ and leads to a unimodal $p(X \cond n, \bmx, m)$.
-\fred{Show a plot: from hedgehog to single peak}
+we see that the right-hand side is a mixture density. In our
+application, the distribution of a single luminosity, $p(L \cond N=1,
+\bml, m)$ is narrow in the sense that the standard deviation relative
+to the mean is at the percent level. This implies that $p(L \cond n,
+\bml, m)$ has one mode from each $N$ and all modes are clearly
+separated. The transformation $L \to x$ effects a strong overlap
+between adjacent $p(X \cond N, \bmx, m)$ leading to a unimodal $p(X
+\cond n, \bmx, m)$.  \fred{Show plot: from hedgehog to single peak}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=0.48\textwidth]{gamma}
+  \caption{Gamma distribution fit. }
+  \label{fig:gamma}
+\end{figure}
 
@@
   \lleq{gmb}
-  X_i \sim \GammaDist(\alpha, \beta) \Rightarrow \sum_{i=1}^N X_i \sim \GammaDist(N \alpha, \beta).
-\end{align}
-With $\bmphi = (\alpha, \beta)$, \refeq{gmb} is equivalent to
+  X_j \sim \GammaDist(\alpha, \beta) \Rightarrow \sum_{j=1}^N X_j \sim \GammaDist(N \alpha, \beta).
+\end{align}
+With
+\begin{align}
+  \lleq{gmi}
+  \bmphi &= (\alpha, \beta),\\
+  p(X_j \cond \bmphi) &=
+  \GammaDist(X_j \cond \alpha, \beta),\\
+  p(\bmX \cond N, \bmphi) &= \prod_{j=1}^N p(X_j \cond \phi)
+\end{align}
+ \refeq{gmb} is equivalent to
 \begin{align}
@@
 
+\subsection*{Approximations}
+
 To explicitly evaluate \refeq{gmd}, we have to overcome the remaining
@@
 $\bmphi$. Regarding the latter, there is a conjugate prior but
-unfortunately the normalization constant can not be evaluated in
-closed form. It is thus of no use for us, and we turn to a
-noninformative prior. The authors of \cite{moala2013bayesian} present
-an overview of possible choices and conclude that the point estimate
-given by the mode of the posterior is nearly insensitive to the prior
-choice for 30 samples. We require in addition that the posterior is
-narrow enough so we group samples in bins to have at least $m=150$
-samples in each bin. The following approximations can be used in
-different regions of the parameter space spanned by $n$ and $m$.
-
-\begin{enumerate}
-  \item For $m$ large, the posterior $p(\bmphi \cond m, \bmx) =
+unfortunately the normalization constant cannot be evaluated in closed
+form. It is thus of no use for us, and we turn to a noninformative
+prior. \fred{Be explicit about prior choice} The authors of
+\cite{moala2013bayesian} present an overview of possible choices and
+conclude that the point estimate given by the mode of the posterior is
+nearly insensitive to the prior choice for 30 samples. We require in
+addition that the posterior is narrow enough so we group samples in
+bins to have at least $m=150$ samples in each bin. The following
+approximations can be used in different regions of the parameter space
+spanned by $n$ and $m$.
+
+\begin{mnumerate}
+  \item \label{large-m} For $m$ large, the posterior $p(\bmphi \cond m, \bmx) =
   p(\bmx\cond m,\bmphi) \, p(\bmphi) / p(\bmx \cond m)$ approaches a
-  multivariate Gaussian with variance tending to zero as $1/m$ in the
-  limit $m \to \infty$.  \fred{check} We can then assume the
-  parameters $\bmphi$ known and fix them at the posterior mode
-  $\bmphi^*$; i.e., $p(\bmphi \cond m, \bmx) \to \delta(\bmphi -
-  \bmphi^*)$.
-  \item For $n$ large, the Negative Binomial $p(N \cond n)$ approaches
-  a Gaussian, defining a region around $N=n$ with the dominant
-  contribution to the sum over $N$ in \refeq{gmd}. The central limit
-  theorem states that for large $N$, $P(X \cond N, \bmphi) \approx
-  \mathcal{N}(X \cond N \, E_\bmphi[X_j], N \, V_\bmphi[X_j])$. For
-  $m$ large enough, we could then plug in the sample-mean and
-  sample-variance estimates and bypass the assumption of a Gamma
-  distribution altogether.
-\end{enumerate}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Prediction based on packet sum}
-
-TODO: Hans. The calculations have already been done. The question is
-whether we even want to bother with them since data in the form of
-packet sums is less information than the vector of individual packets
-as used above.
-\fred{No, we don't want to bother.}
+  multivariate Gaussian. We can then compute the various moments
+  $E_\bmphi[f(x)]$ with $f(x)=1,x,x^2$ by Laplace's method to good
+  accuracy.
+  \item \label{very-large-m} In the limit $m \to \infty$, the covariance of the posterior
+  $p(\bmphi \cond m, \bmx)$ tends to zero as $1/m$ and we can then fix
+  the parameters $\bmphi$ at the posterior mode $\bmphi^*$; i.e.,
+  $p(\bmphi \cond m, \bmx) \to \delta(\bmphi - \bmphi^*)$.
+  \item \label{large-n} For $n$ large, the cumulative of the Negative
+  Binomial $p(N \cond n)$ can be approximated by a Gaussian, defining
+  a region around $N=n$ with the dominant contribution to the sum over
+  $N$ in \refeq{gmd}. The contribution around $N=0$ is negligible, and
+  we can replace $\sum_n \to \int dN$. The density $p(N \cond n) P(X
+  \cond N,m, \bmx)$ is close to a Gaussian and the integral can be
+  accurately approximated by Laplace's method which requires the
+  Hessian and the mode. We benefit from partial error cancelation
+  because we apply Laplace's method both in the numerator and
+  denominator~\cite{tierney_accurate_1986} as each moment has to be
+  normalized by the evidence $p(\bmx \cond m)$. The former is
+  available in closed form and the latter has to be determined
+  numerically. Fortunately, Newton's method leads to a solution in
+  only a few steps~\cite{minka2002estimating}. \fred{The reference is
+    for max. likelihood of a $\GammaDist(\cdot)$ but I'm sure the
+    performance is similar for max. posterior and the first moments}
+  \item \label{central-limit} The central limit theorem states that for large $N$,
+  $P(X \cond N, \bmphi) \approx \mathcal{N}(X \cond N \,
+  E_\bmphi[X_j], N \, V_\bmphi[X_j])$. For $m$ large enough, we could
+  then plug in the sample-mean and sample-variance estimates and
+  bypass the assumption of a Gamma distribution altogether.
+  \item \label{large-n-large-m} For large $n$ and $m$, $ p(X\cond
+  n,\bmx,m)$ can be approximated by a Gaussian whose mean and variance
+  are given by \refeq{mgo} and \refeq{mgr} in terms of
+  $\bmphi$-expectation values. The latter are integrals with no known
+  closed-form solutions but we can easily and accurately approximate
+  them with Laplace's method.
+\end{mnumerate}
+To simplify, we assume that $m$ is large. In our application, we may
+have to combine several frequency bins to achieve $m \gtrsim 150$ for
+the purpose of infering $\bmphi$. This is equivalent to assuming that
+the distribution changes only slowly with the frequency, a condition
+that we verified for TARDIS. It is thus justified to use
+approximations \ref{large-m} or even \ref{very-large-m}.
+
+Numerically, the most challenging case is small $n$ because the sum
+over $N$ in \refeq{gmd} has to be computed explicitly starting at
+$N=1$ until contributions become negligible. Since $p(N \cond n)$
+falls off quickly, it is sufficient in our application to do this
+until $n \lesssim 20-30$. Beyond that, we use approximations
+\ref{large-n}--\ref{large-n-large-m}.
+
+\subsection*{Comparison of approximations}
 
@@
 
-In Figure~\ref{fig:beta} is a fit of the Beta distribution to 3000
+In \reffig{beta} is a fit of the Beta distribution to 3000
 samples from a TARDIS run with frequencies in the range $[1.367,
@@
 
-\begin{figure}[h]
+\begin{figure}[ht]
   \centering
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+% Local Variables:
+% compile-command:"rubber --pdf -W refs -W misc -S bayes-tardis-luminosity-prediction"
+% End:

### 2016-02-10 hans/bayes-tardis-luminosity-prediction.tex
[hans] Revise my changes in gamma section

@@
 \documentclass[11pt]{article}
+\usepackage[utf8]{inputenc}
+
 \usepackage[a4paper, left=20mm, right=20mm, top=25mm, bottom=25mm]{geometry}
@@
 
-\item We choose $N$ to follow a Poisson distribution with parameter
-  $\lambda$ \fred{this choice linked to limit of Multinomial, see above}
+\item We choose $N$ to follow a Poisson distribution \fred{this choice
+    linked to limit of Multinomial, see above} with parameter
+  $\lambda$ \fred{don't use $\lambda$ again, it should only represent
+    the wavelength}
   \begin{align}
@@
 \end{align}
-\fred{$p(\bmphi\cond N,n,\bml) = p(\bmphi\cond n,\bml)$ }
+\fred{$p(\bmphi\cond N,n,\bml) = p(\bmphi\cond n,\bml)$. Below $N$ is (correctly) missing in \refeq{pmp} }
 and using a multiplicity-independent prior for $\bmphi$
@@
 The Gamma distribution  with shape parameter $\alpha$ and rate
-parameter $\beta$ is defined
+parameter $\beta$ is defined as
 \begin{align}
@@
 whereas the luminosity of real packets is bounded from above and has a
-peak close to but not at the boundary. In order to describe the luminosity
-distribution of real packets through $\GammaDist(\alpha, \beta)$, we
-have to transform the luminosities. To this end, we first take the
-largest luminosity from the entire frequency range and increase it
-slightly to avoid numerical difficulties at the end point
+peak close to but not at the boundary. \fred{refer to plot of real
+  luminosity packets in some frequency bin} In order to describe the
+luminosity distribution of real packets through $\GammaDist(\alpha,
+\beta)$, we have to transform the luminosities. To this end, we first
+take the largest luminosity from the entire frequency range and
+increase it slightly to avoid numerical difficulties at the end point
 \begin{align}
@@
 \end{align}
-where $ \epsilon=10^{-6}$.
-We assume the total number of samples across all bins is large, hence
-the effect of considering $\lmax$ a constant independent of the data
-is negligible. We then transform the samples as
+where $ \epsilon=10^{-6}$.  We assume the total number of samples
+across all bins is large, hence the effect of considering $\lmax$ a
+constant independent of the data is negligible. We then transform the
+samples $\bml \to \bmx$ as
 \begin{align}
@@
 separated. The transformation $L \to x$ effects a strong overlap
-between adjacent $p(X \cond N, \bmx, m)$ leading to a unimodal $p(X
-\cond n, \bmx, m)$.  \fred{Show plot: from hedgehog to single peak}
+between adjacent $p(X \cond N, \bmx, m)$, so $p(X \cond n, \bmx, m)$
+can be well approximated by a unimodal distribution; see
+\reffig{hedgehog}.  \fred{I'm not 100 \% sure but $p(X \cond n, \bmx,
+  m)$ should still have $N \to \infty$ modes but they aren't visible}
+\missingfigure{from hedgehog to single peak \label{fig:hedgehog}}
 
@@
   \includegraphics[width=0.48\textwidth]{gamma}
+  \fred{Plot with more samples to see dip at $x=0$.}
   \caption{Gamma distribution fit. }
@@
   \lleq{gmb}
-  X_j \sim \GammaDist(\alpha, \beta) \Rightarrow \sum_{j=1}^N X_j \sim \GammaDist(N \alpha, \beta).
+  X_j \sim \GammaDist(\alpha, \beta) \Rightarrow X \equiv \sum_{j=1}^N X_j \sim \GammaDist(N \alpha, \beta).
 \end{align}
@@
 unfortunately the normalization constant cannot be evaluated in closed
-form. It is thus of no use for us, and we turn to a noninformative
-prior. \fred{Be explicit about prior choice} The authors of
-\cite{moala2013bayesian} present an overview of possible choices and
-conclude that the point estimate given by the mode of the posterior is
-nearly insensitive to the prior choice for 30 samples. We require in
-addition that the posterior is narrow enough so we group samples in
-bins to have at least $m=150$ samples in each bin. The following
-approximations can be used in different regions of the parameter space
-spanned by $n$ and $m$.
+form. This conjugate prior is thus of no use for us, and we turn to a
+noninformative prior. \fred{Be explicit about prior choice} The
+authors of \cite{moala2013bayesian} present an overview of possible
+choices and conclude that the point estimate given by the mode of the
+posterior is nearly insensitive to the prior choice for 30 samples. We
+require in addition that the posterior is narrow enough so we group
+samples in bins to have at least $m=150$ samples in each bin. This
+implicitly assumes that the luminosity distribution is constant in the
+frequency range defined by $m$.
+
+Note that the bins defined by the experiment in general do not overlap
+with the bins used to infer the luminosity distribution; i.e., $n \ne
+m$.
+
+The following approximations can be used in different regions of the
+parameter space spanned by $n$ and $m$.
 

### 2016-02-10 hans/bat/Tardis.h
[hans] Add skeleton bat analysis

@@ -1,0 +1,48 @@
+// ***************************************************************
+// This file was created using the bat-project script.
+// bat-project is part of Bayesian Analysis Toolkit (BAT).
+// BAT can be downloaded from http://mpp.mpg.de/bat
+// ***************************************************************
+
+#ifndef __BAT__TARDIS__H
+#define __BAT__TARDIS__H
+
+#include <BAT/BCModel.h>
+
+#include <string>
+#include <vector>
+
+// This is a Tardis header file.
+// Model source code is located in file Tardis/Tardis.cxx
+
+// ---------------------------------------------------------
+class Tardis : public BCModel
+{
+
+public:
+    using Vec = std::vector<double>;
+
+    // Constructor
+    Tardis(const std::string& name);
+
+    // Destructor
+    ~Tardis();
+
+    void ReadData(const std::string& fileName, const std::string& dataSet, Vec& buffer);
+
+    // Overload LogLikelihood to implement model
+    double LogLikelihood(const std::vector<double>& parameters);
+
+    // Overload LogAprioriProbability if not using built-in 1D priors
+    double LogAPrioriProbability(const std::vector<double> & parameters);
+
+private:
+    double polyn(Vec::const_iterator begin, Vec::const_iterator end, const double& lambda);
+
+    std::vector<double> energies, nus;
+
+    const unsigned order;
+};
+// ---------------------------------------------------------
+
+#endif

### 2016-02-18 hans/bayes-tardis-luminosity-prediction.tex
changed one line, testing
[hans] Adjust section on TARDIS
Prepare for linear model of luminosity-distribution parameters

@@
 %% CHANGE THIS DATE AS YOU MODIFY THE FILE
-\newcommand{\vdate}{160114}
+\newcommand{\vdate}{160218}
 \pagestyle{myheadings}
@@
 
-\DeclareMathOperator{\GammaDist}{Gamma}
 \newcommand{\refeq}[1]{Eq.~(\ref{#1})}
 \newcommand{\reffig}[1]{Fig.~\ref{fig:#1}}
+
+\DeclareMathOperator{\GammaDist}{Gamma}
+\newcommand{\npack}{n_p}
 \newcommand{\lmax}{\ell_{\rm max}}
@@
 \item Spectrum of luminosities in channels (or ``bins'')
-  $b=0,1,\ldots,B$
-\item The wavelength of channel $b$ is given by $\lambda_b =
-  \lambda_0 + b\,\delta\lambda$, where $\delta\lambda$ is the bin width.
-
+  $b=0,1,\ldots,B$ is given by the telescope
+\item In the simplest case, the frequency of channel $b$ is given by
+  $\nu_b = \nu_0 + b\,\delta\nu$, where $\delta\nu$ is
+  the bin width. In the general case, the bins are not of equal
+  size. Let's then consider $\nu_b$ as the left edge of bin $b$,
+  and $\nu_{B+1}$ is the right edge of the last bin; i.e., the
+  maximum frequency.
 \item We treat the results of a single TARDIS run as \textit{data}
-  $D$, i.e.\ fixed results which we do not question but accept as the
-  true values.  Following a single run, the TARDIS ``data'' consists,
-  for each bin $b$, of the number or ``multiplicity'' of packets and
-  the individual packet luminosities,
+  $D$, i.e.\ fixed results which we do not question but accept as
+  given.  These ``data'' are the number of packets leaving the
+  supernova, $\npack$, the luminosity $\bml$, and the frequencies
+  $\bmnu$ for each packet
   \begin{align}
     \lleq{dfc}
-  D &= \{n_0,\bml_0,n_1,\bml_1,\ldots,n_\smB,\bml_\smB\}
-  \ =\ \{n_b,\bml_b\}_{b=0}^\smB
-\end{align}
-  \checked{}
-with
+    D = \{ \npack, \bml, \bmnu \} \,,
+  \end{align}
+  where
   \begin{align}
     \lleq{dfd}
-  \bml_b &= (\ell_{b,1},\ell_{b,2},\ldots,\ell_{b,n_b})
-\end{align}
-  \checked{}
-Given $D$, the total luminosity in bin $b$ is known,
+    % \bml_b &= (\ell_{b,1},\ell_{b,2},\ldots,\ell_{b,n_b})
+    \dim \bml = \dim \bmnu = \npack \,.
+  \end{align}
+  Given $D$, the number of samples in bin $b$ is
+  \begin{align}
+    \lleq{dff}
+    n_b = \sum_{j=1}^{\npack} \mathbf{1}_{[\nu_b,\nu_{b+1}]}( \nu_j )
+  \end{align}
+  and similarly the luminosity is
   \begin{align}
     \lleq{dfe}
-  \ell_b = \sum_{j=1}^{n_b}\ell_{b,j}
-\end{align}
-  \checked{}
+    \ell_b = \sum_{j=1}^{\npack}\ell_j \mathbf{1}_{[\nu_b,\nu_{b+1}]}( \nu_j ),
+  \end{align}
+  where $\mathbf{1}_V(x)$ is the indicator function requiring $x \in V$.
 \end{mtemize}
@@
 of a stochastic variable $N_b$ governed by a model pdf
-$p(N_b)$.
-\fred{In our general case, $\sum_b n_b$ is an instance of a random variable because some packets leave the supernova, some don't.
-I guess in most applications the total number of events would be chosen and not random.
-We are fortunate that this difference does not matter because we consider bins independently.
-} Similarly, $\bml_b$ is a specific realization of a
-collection of stochastic variables $\bmL_b
-=\{L_{b,1},\ldots,L_{b,\smN_b}\}$, where the number $N_b$ of such
-variables is itself stochastic. The relevant pdf for $\bmL_b$
-therefore always presupposes knowledge of $N_b$, so we must work with
-the conditional probability $p(\bmL_b\cond N_b)$. This situation is
-captured mathematically as $N_b \sim p(N_b), \bmL_b \sim p(\bmL_b\cond
-N_b)$ where $\sim$ means ``is by hypothesis distributed in this model
-according to the pdf'', i.e.\
+$p(N_b)$. \fred{In our general case, $\npack = \sum_b n_b$ is an instance of a
+  random variable because some packets leave the supernova, some
+  don't. I guess in most applications the total number of events would
+  be chosen and not random. We are fortunate that this difference does
+  not matter because we consider bins independently. } Similarly,
+$\bml_b$ is a specific realization of a collection of stochastic
+variables $\bmL_b =\{L_{b,1},\ldots,L_{b,\smN_b}\}$, where the number
+$N_b$ of such variables is itself stochastic. The relevant pdf for
+$\bmL_b$ therefore always presupposes knowledge of $N_b$, so we must
+work with the conditional probability $p(\bmL_b\cond N_b)$. This
+situation is captured mathematically as $N_b \sim p(N_b), \bmL_b \sim
+p(\bmL_b\cond N_b)$ where $\sim$ means ``is by hypothesis distributed
+in this model according to the pdf'', i.e.\
 \begin{align*}
@@
 \end{align*}
-where $\bmphi_b$ is the set of parameters\fred{so $\bmphi$ includes random seed or not? Initially I thought so. But later on it becomes clear that you do not include it.} of the model pdf which
+where $\bmphi_b$ is the set of parameters of the model pdf which
 govern $\bmL_b$ and $\lambda_b$ is the parameter of the model pdf
@@
 \begin{mtemize}
+\item We omit the dependence on supernova parameters $\bmtheta$ from
+  the notation below and instead condition on the TARDIS output
+  $\npack, \bml, \bmnu$ directly.
+
 \item In the remainder of this paper, we assume that the packet
@@
     p(N_0,\bmL_0\ldots,N_b,\bmL_b,\ldots,N_\smB,\bmL_\smB
-    \cond \bmtheta,\bmphi,\bmlambda,\hmod)
-    &= \prod_{b=0}^B p(N_b,\bmL_b\cond \bmtheta,\bmphi,\bmlambda,\hmod)
+    \cond \npack, \bml, \bmnu,\bmphi,\bmlambda,\hmod)
+    &= \prod_{b=0}^B p(N_b,\bmL_b\cond n_b, \bml, \bmnu,\bmphi,\bmlambda,\hmod)
     % p(N_0,\ldots,N_b,\ldots,N_\smB \cond \bmtheta,\bmphi,\hmod)
@@
 
-\item We omit the supernova parameters $\bmtheta$ from the notation
-  below.
-
 \item Based on the assumption of bin-bin statistical independence, we
   shall henceforth drop the bin index and do all our calculations for
-  an individual bin $b$ only, writing for example $L_b \to L$,
-  $\ell_{b,j} \to \ell_j$ etc. The total spectrum is then obtained by
-  doing any given analysis or calculation for all bins.
+  an individual bin $b$ only, writing for example $L_b \to L, n_b \to
+  n$ etc.  The total spectrum is then obtained by doing any given
+  analysis or calculation for all bins. But note that while the
+  single-bin probability $p(N_b,\bmL_b\cond n_b, \bml,
+  \bmnu,\bmphi,\bmlambda,\hmod)$ depends only on the number of packets
+  $n_b$ in the bin, it depends on the luminosities and frequencies of
+  \emph{all} packets.
 
@@
     linked to limit of Multinomial, see above} with parameter
-  $\lambda$ \fred{don't use $\lambda$ again, it should only represent
-    the wavelength}
+  $\lambda$
   \begin{align}
@@
   \end{align}
+
+\item In general, the distribution of $L_j$ could be a of a different
+  functional form in each bin. But in practice, the distribution of
+  packets does not change abruptly with frequency. We therefore assume
+  that the functional form of $p(L_j)$ is identical in each bin but it
+  is parametrized by a continuous function of frequency $f(\nu |
+  \bmphi)$. This greatly reduces the number of parameters if $\npack$
+  is large because a handful of parameters, $\dim \bmphi$, can be
+  inferred from the complete set of packets instead
 

### 2016-02-19 hans/bayes-tardis-luminosity-prediction.tex
[hans] polish `packet multiplicity` subsection

@@
 
-\item In general, the distribution of $L_j$ could be a of a different
-  functional form in each bin. But in practice, the distribution of
-  packets does not change abruptly with frequency. We therefore assume
-  that the functional form of $p(L_j)$ is identical in each bin but it
-  is parametrized by a continuous function of frequency $f(\nu |
-  \bmphi)$. This greatly reduces the number of parameters if $\npack$
-  is large because a handful of parameters, $\dim \bmphi$, can be
-  inferred from the complete set of packets instead
+  \item In general, the distribution of $L_j$ could be a of a
+  different functional form in each bin. For simplicity, we assume it
+  is identical but may depend on parameters that differ from bin to
+  bin. In practice, the distribution of packets does not change
+  abruptly with frequency. We therefore assume that any parameter
+  needed to fully specify $p(L_j)$ is given as a continuous function
+  of frequency and parameters $\bmphi$; i.e.,
+  \begin{align}
+    \lleq{pmdb}
+    p(L_j) = p(L_j \cond \nu, \bmphi)
+  \end{align}
+  For example, assume $p(L_j)$ depends on one parameter $a$. If we fit
+  separately in each bin, we would have $B+1$ fits and $B+1$
+  parameters. In each such fit, there would be only few samples to
+  constrain $a$ in each bin. Then instead of fitting $a$ separately in
+  each bin, we assume $a = a(\nu, \bmphi)$. This greatly reduces the
+  number of parameters if $B$ is large because a handful of
+  parameters, $\bmphi$, can be inferred from the complete set of
+  packets. Necessarily, $a$ is best determined for the frequency
+  around which the density of packets is highest.
 
@@
 where $p(N\cond\lambda,n) = p(N\cond\lambda)$ is independent of $n$.
-Taking an improper prior (see Appendix)\footnote{For a uniform prior,
+Taking the prior (see Appendix)\footnote{For a uniform prior,
   the case $a{=}0$, we require of course a maximum $\lambda_{\rm
@@
   \lleq{pmf}
-  p(\lambda) = c\lambda^{-a} \qquad a = 0\;\text{or}\;1
-\end{align}
-and inserting this and (\ref{pmcb}) into (\ref{pme}), we obtain
-for the evidence and prediction
+  p(\lambda) = c\lambda^{-a},
+\end{align}
+common choices are $a=0$ (uniform prior) and $a=1/2$ (Jeffreys
+prior). Both choices lead to an improper prior but the posterior
+predictive (see below) is proper.
+Inserting \refeq{pmf} and (\ref{pmcb}) into (\ref{pme}), we obtain for
+the evidence and the posterior predictive
 \begin{align}
@@
 \end{align}
-\fred{$a=1/2 \Leftrightarrow$ Jeffrey's prior would my choice.
-We then get a Pólya distribution instead of Negative Binomial but nothing substantial changes.
-$$(pmh) \frac{\Gamma(N+n-a+1)}{N!\,\Gamma(n-a+1)\,2^{N+n+1-a}}$$ }
-which we recognise as a Negative Binomial distribution (see appendix)
-\fred{There are several definitions of the Negative Binomial, so we should specify which one we use.
-This one coincides with Wikipedia: $N$ successes given $k$ failures with probability of success $\rho$.}
-with parameters $k=n+1-a$ and $\rho=(1/2)$
+which we recognise as a Negative Binomial distribution describing the
+probability to observe $N$ successes given $k=n+1-a$ failures in a
+sequence of Bernoulli trials with success probability $\rho=1/2$
 \begin{align}
@@
 We shall use the properties of the Negative Binomial in the
-calculations below.  Note that the distribution of $N$ given $n$ is
-independent of the packets $\bmL$ and $\bml$ by construction.
+calculations below. For non-integer $a$, the Negative Binomial
+generalizes immediately to the Pólya distribution by replacing $x! \to
+\Gamma(x+1)$. Note that the distribution of $N$ given $n$ is
+independent of $\bmL$ and $\bml$ by construction.
 
@@
 n,a,b)$. However we are supposed to deal with ONE TARDIS run, so that
-$\ell_{\rm max}$ is known and fixed.
-\fred{Rather: we approximate $p(\ell_{\rm max} \cond n, a,b)$ by $\delta(\max \ell_{b,j})$.
-This is OK is $\sum_b n_b$ is large.
-For numerical stability, we want to ensure that $x<1$, so we take the largest $\ell_{b,j}$ and multiply by $1 +\varepsilon, \varepsilon = 10^{-7}$}
+$\ell_{\rm max}$ is known and fixed.  \fred{Rather: we approximate
+  $p(\ell_{\rm max} \cond n, a,b)$ by $\delta(\max \ell_{b,j})$. This
+  is OK is $\sum_b n_b$ is large. For numerical stability, we want to
+  ensure that $x<1$, so we take the largest $\ell_{b,j}$ and multiply
+  by $1 +\varepsilon, \varepsilon = 10^{-7}$}
 

### 2016-02-19 hans/bayes-tardis-luminosity-prediction.tex
[hans] Add frequency depends, update assumptions

@@
   \item We treat the results of a single TARDIS run as \textit{data}
-  $D$, i.e.\ fixed results which we do not question but accept as
-  given.  These ``data'' are the number of packets leaving the
-  supernova, $\npack$, the luminosity $\bml$, and the frequencies
-  $\bmnu$ for each packet
+  \fred{If in the end we want to compare to actual data from a
+    telescope, I'd prefer to not call the TARDIS output ``data''. Why
+    don't we just call it ``TARDIS output''?}  $D$, i.e.\ fixed
+  results which we do not question but accept as given.  These
+  ``data'' are the number of packets leaving the supernova, $\npack$,
+  the luminosity $\bml$, and the frequencies $\bmnu$ for each packet
   \begin{align}
@@
 
-Our goal of prediction of $L$ given the data $(n,\bml)$ is captured in
-the probability $p(L\cond n,\bml)$ with all relevant parameters
-integrated out. We first note from Eq.~(\ref{pmh}) that $p(N\cond
-n,\bml) = p(N\cond n)$ does not depend on $\bml$. Furthermore, since
-$L$ is fully determined by $(N,\bmL)$,
+Our goal of predicting $L$ given the TARDIS output $(\npack,\bml,
+\bmnu)$ is captured in the probability $p(L\cond \npack,\bml, \bmnu)$
+with all relevant parameters integrated out. We first note from
+\refeq{pmh} that $p(N\cond \npack,\bml, \bmnu) = p(N\cond n)$ does not
+depend on $\bml$ or $\bmnu$ directly but only through $n$, the number
+of packets in the bin; cf. \refeq{dff}. Furthermore, since $L$ is
+fully determined by $(N,\bmL)$,
 \begin{align}
@@
 \end{align}
-is does not depend directly on $(n,\bml)$, i.e.\
+is does not depend directly on $(\npack),\bml,\bmnu)$, i.e.\
 \begin{align}
   \lleq{pmi}
-  p(L\cond N,\bmL,n,\bml)
+  p(L\cond N,\bmL,n,\bml, \bmnu)
   = p(L\cond N,\bmL)
@@
 \end{align}
-\checked{}
 Hence the prediction is given by
@@
   \lleq{pmj}
-  p(L\cond n,\bml)
-  &= \sum_{N=0}^\infty p(L\cond N,n,\bml)\,p(N\cond n,\bml) \\
+  p(L\cond n,\bml, \bmnu)
+  &= \sum_{N=0}^\infty p(L\cond N,n,\bml, \bmnu)\,p(N\cond n,\bml, \bmnu) \\
   \lleq{pmjb}
-  &= \sum_N p(N\cond n)\,\int d\bmL\, p(L\cond N,\bmL,n,\bml)\, p(\bmL\cond N,n,\bml)
+  &= \sum_N p(N\cond n)\,\int d\bmL\, p(L\cond N,\bmL,n,\bml, \bmnu)\, p(\bmL\cond N,n,\bml, \bmnu)
   \\
@@
   &= \sum_N p(N\cond n)\,\int d\bmL\,\delta(L - \textstyle\sum_{j=1}^N L_j)
-  \, p(\bmL\cond N,n,\bml)
-\end{align}
-\checked{}
-The dependence on the TARDIS ``data'' $(n,\bml)$ enters through
-$p(\bmL\cond N,n,\bml)$ in terms of the parameters $\bmphi$ of
-$p(\bmL\cond N,\bmphi)$,
+  \, p(\bmL\cond N,n,\bml, \bmnu)
+\end{align}
+\fred{Add missing $\bmnu$! Everything that is conditional on $\bml$ is
+  also conditional on $\bmnu$} The dependence on the TARDIS ``data''
+$(\npack,\bml, \bmnu)$ enters through $p(\bmL\cond N,\npack,\bml,
+\bmnu)$ in terms of the parameters $\bmphi$ of $p(\bmL\cond
+N,\bmphi)$,
 \begin{align}
   \lleq{pmk}
-  p(\bmL\cond N,n,\bml)
-  &= \int d\bmphi\, p(\bmL\cond N,\bmphi,n,\bml)\,p(\bmphi\cond N,n,\bml) \\
-  &= \int d\bmphi\, p(\bmL\cond N,\bmphi)\,p(\bmphi\cond N,n,\bml).
+  p(\bmL\cond N,\npack,\bml, \bmnu)
+  &= \int d\bmphi\, p(\bmL\cond N,\bmphi,\npack,\bml, \bmnu)\,p(\bmphi\cond N,\npack,\bml, \bmnu) \\
+  &= \int d\bmphi\, p(\bmL\cond N,\bmphi)\,p(\bmphi\cond N,\npack,\bml, \bmnu).
 \end{align}
 because the likelihood for $\bmL$ is fully determined by $N,\bmphi$
-and hence independent of $(n,\bml)$.
+and hence independent of $(\npack,\bml, \bmnu)$.
 %
@@
   \lleq{pmkb}
-  p(\bmphi\cond N,n,\bml) &=
-  \frac{p(\bml\cond N,n,\bmphi)\,p(\bmphi\cond N,n)} %
-  {\int d\bmphi\,p(\bml\cond N,n,\bmphi)\,p(\bmphi\cond N,n)} %
-  % \ =\ \frac{p(\bml\cond N,n,\bmphi)\,p(\bmphi\cond N,n)} %
+  p(\bmphi\cond N,\npack,\bml, \bmnu) &=
+  \frac{p(\bml\cond N,\npack,\bmphi,\bmnu)\,p(\bmphi\cond N,\npack,\bmnu)} %
+  {\int d\bmphi\,p(\bml\cond N,\npack,\bmphi,\bmnu)\,p(\bmphi\cond N,\npack,\bmnu)}, %
+  % \ =\ \frac{p(\bml\cond N,\npack,\bmphi)\,p(\bmphi\cond N,n)} %
   % {p(\bml\cond n)} %
 \end{align}
-\fred{$p(\bmphi\cond N,n,\bml) = p(\bmphi\cond n,\bml)$. Below $N$ is (correctly) missing in \refeq{pmp} }
-and using a multiplicity-independent prior for $\bmphi$
+
+ a multiplicity-independent prior for $\bmphi$
 \begin{align}
   \lleq{pml}
-  p(\bmphi\cond N,n) &= p(\bmphi)
+  p(\bmphi\cond N,\npack,\bmnu) &= p(\bmphi),
+\end{align}
+and recognizing that $N$ does not tell us anything about $\bml$
+\begin{align}
+  \lleq{pmla}
+  p(\bml\cond N,\npack,\bmphi,\bmnu) = p(\bml\cond \npack,\bmphi,\bmnu)
 \end{align}
@@
   \lleq{pmn}
-  p(\bmL\cond N,n,\bml)
+  p(\bmL\cond N,\npack,\bml, \bmnu)
   &= \int d\bmphi\, p(\bmL\cond N,\bmphi)\, %
-  \frac{p(\bml\cond n,\bmphi)\,p(\bmphi)} %
-  {\int d\bmphi\,p(\bml\cond n,\bmphi)\,p(\bmphi)} \\
+  \frac{p(\bml\cond \npack,\bmphi)\,p(\bmphi)} %
+  {\int d\bmphi\,p(\bml\cond \npack,\bmphi)\,p(\bmphi)} \\
   \lleq{pmo}
@@
   \int d\bmphi\, p(\bmL\cond N,\bmphi)\, %
-  p(\bml\cond n,\bmphi)\,p(\bmphi)
+  p(\bml\cond \npack,\bmphi)\,p(\bmphi)
 \end{align}
 with $p(\bml\cond n) = \int d\bmphi\, p(\bml\cond
-n,\bmphi)\,p(\bmphi)$ in the denominator.
+\npack,\bmphi)\,p(\bmphi)$ in the denominator.
 Inserting this into (\ref{pmjc}), we obtain
@@
   \boxed{
-  p(L\cond n,\bml)
+  p(L\cond \npack,\bml, \bmnu)
   = \frac{1}{p(\bml\cond n)}
   \sum_N p(N\cond n)\,\int d\bmL\,\delta(L - \textstyle\sum_j L_j)
-  \displaystyle\int d\bmphi\,p(\bmL\cond N,\bmphi) \, p(\bml\cond n,\bmphi)
+  \displaystyle\int d\bmphi\,p(\bmL\cond N,\bmphi) \, p(\bml\cond \npack,\bmphi)
   \, p(\bmphi)
@@
 \end{equation}
-\checked{}
 with
@@
   p(\bmL\cond N,\bmphi) &= \prod_{j=1}^N p(L_j\cond \bmphi) \\
-  p(\bml\cond n,\bmphi) &= \prod_{j=1}^n p(\ell_j\cond \bmphi)
+  p(\bml\cond \npack,\bmphi) &= \prod_{j=1}^n p(\ell_j\cond \bmphi)
 \end{align}

### 2016-02-28 hans/bayes-tardis-luminosity-prediction.tex
Inserted summary of major conceptual changes and possible consequences re (a) fitting all bins at the same time via rescaling (b) combining low-luminosity bins using the new bin index $m$. Important also the question whether we should work with frequency or wavelength.  Some comments on detail but no systematic work.

@@
 \usepackage[colorinlistoftodos]{todonotes} % Use 'disable' to remove todos in final version
-\newcommand{\fred}[1]{\todo[inline]{#1}} %
+\newcommand{\fred}[1]{\todo[color=orange!40,inline]{#1}} %
 \newcommand{\checked}{\todo[color=green,noline]{\checkmark}} %
 \newcommand{\checkedl}{\todo[color=brown]{\checkmark}} %
+\newcommand{\hans}[1]{\todo[color=yellow!30,inline]{#1}} %
+\newcommand{\hanslong}[2][]{\todo[%bordercolor=red,
+  color=white,inline,caption={2do}, #1]{
+    \begin{minipage}{\textwidth} #2\end{minipage}}} %
 
 %% CHANGE THIS DATE AS YOU MODIFY THE FILE
-\newcommand{\vdate}{160218}
+\newcommand{\vdate}{160228}
 \pagestyle{myheadings}
@@
 \newcommand{\lmax}{\ell_{\rm max}}
+\newcommand{\Lmax}{L_{\rm max}}
 
@@
 
-\textbf{Abstract} TODO
+\textbf{Abstract} TODO\\
+
+\hans{Temporary addition of TOC to make it easier to find stuff. Eventually it should
+disappear again.}
+
+\tableofcontents
+
+\newpage %
+\hanslong{ %
+  Hans 160228: \textbf{Summary of issues, based on v160218}. \\
+
+  Before getting into all sorts of detail, I wanted to understand the
+  major changes and developments. To my mind they are:
+  \begin{enumerate}
+  \item Fred discovered that the rescaling $X = L/\Lmax$ may have
+    unintended advantages. TARDIS and general supernova luminosity
+    spectra are highly variable because it consists not only of smooth
+    blackbody spectrum, but also of strong absorption lines (emission
+    also?? see 2014 MNRAS Kerzendorf, Sim et al?). Since $\Lmax$
+    varies exactly like the spectrum itself, all binwise
+    individual-packet distributions are rescaled to the interval
+    $[0,1]$ leaving only a far weaker change of \textit{shape} of the
+    distribution rather than the \textit{scale}.
+
+  \item Just below Eq. (bsi) and also in (pmda), we had made explicit
+    that we would do all the calculations for each bin separately, so
+    there would be lots of fit parameters. Fred's 160218 paragraph
+    starting with ``In general, the distribution \ldots'' and Eq
+    (pmdb) sees significant reduction in the number of parameters if
+    the above remnant dependence of $p(\bmL\cond N,\bmphi)$ in (pmda)
+    on shape could be captured in a slow-varying function
+    $a(\nu,\bmphi)$.
+
+  \item Separately, Fred introduces the idea of combining elementary
+    $b$-bins into compound bins with index $m$ as this improves the
+    statistics.
+
+  \item The simplification for large $N$ etc permits a Gaussian
+    approximation, hence Laplace, hence the key equation (pmp) may be
+    simplified to some approximation, making use of the moment
+    generating function eg Eq (mgd) unnecessary.
+
+  \item In principle, the above sounds like a very good development.
+    There are some important caveats:
+    \begin{enumerate}
+    \item I see no detail on how to find $a(\nu,\bmphi)$. The first
+      question would be if there is some numerical indication for its
+      form. This seems to be equivalent to asking what the dependence
+      of (say) the Gamma parameter $\alpha$ on $\nu$ is, and that is
+      of course of great interest anyway.
+
+    \item My second comment is that $a(\nu,\bmphi)$ is probably just
+      your normal parametrisation which enters best-fit algorithms.
+      If we are to treat it as a best-fit parametrisation with
+      metaparameters, then that may just bring back all the
+      complications which we thought had disappeared. This would have
+      to be investigated. If on the other hand you also want to
+      treat $a(\nu,\bmphi)$ as some kind of ``constant function'',
+      then one would have to set out how stable this constant function
+      would be against various underlying changes. In other words, 
+      if calculation of $a()$ is not part of the general algorithm,
+      how is it fixed? and can it be fixed for all cases?
+
+    \item I think there may be a problem of combining the two proposed
+      simplifications $L \to X$ and $m$ compound bins. Depending on
+      the variability of the parametrisation as function of $\nu$, the
+      shape of $p(\bmL\cond N,\bmphi)$ for single bins with large
+      $\Lmax$ may differ significantly from the shape for many small
+      bins combined into one using $m$. In other words, rebinning
+      and rescaling may not be compatible.
+    \end{enumerate}
+
+  \item Once these are cleared up, we can get into the details \ldots\\
+
+  \end{enumerate}
+} %
 
@@
 \subsection{Definitions of relevant quantities}
+\label{sec:dfnr}
 
@@
   maximum frequency.
+  %%%
+  \hans{The 2014 TARDIS paper I have seen uses wavelength $\lambda$,
+    not frequency in their plots. Is Wolfgang's recent data all
+    frequency data? If it is, ie if Wolfgang et al have decided to
+    change to frequency, they should explain whether that will happen
+    in their and our papers also. If they want to stick with
+    wavelength, we should too. I think this is something Wolfgang
+    should decide.
+
+    If they say we should stay with wavelength, then I would suggest
+    replacing $\nu$ by $\lambda$ here and throughout, and of course
+    change the Poisson parameter below to something like $\nu$.}
+
   \item We treat the results of a single TARDIS run as \textit{data}
+    %%%%%
     \fred{If in the end we want to compare to actual data from a
-    telescope, I'd prefer to not call the TARDIS output ``data''. Why
-    don't we just call it ``TARDIS output''?}  $D$, i.e.\ fixed
-  results which we do not question but accept as given.  These
-  ``data'' are the number of packets leaving the supernova, $\npack$,
-  the luminosity $\bml$, and the frequencies $\bmnu$ for each packet
+      telescope, I'd prefer to not call the TARDIS output
+      ``data''. Why don't we just call it ``TARDIS output''?} %
+    \hans{I agree, ``data'' is awkward, and it is indeed useful to
+      keep data and $D$ for later. ``Output'' would be ok, except that
+      we should change the symbol $D$ to something else too. Symbol
+      $O$ is not good. For the moment I would suggest $D_S$ (subscript
+      simulation) or $D_T$ since we know that $T$ will later stand for
+      temperature. Another useful term would be ``simdata'', symbol
+      $S$.}
+    %%%%%
+    $D$, i.e.\ fixed results which we do not question but accept as
+    given.  This ``data'' consists of the total number of packets
+    leaving the supernova $\npack$ plus the set of luminosities
+    $\bml$ and frequencies $\bmnu$ of each packet,
   \begin{align}
@@
   \end{align}
-  Given $D$, the number of samples in bin $b$ is
+  In terms of the $\mathbf{1}_V(x)$ the indicator function requiring
+  $x \in V$, the number of packets in bin $b$ is
   \begin{align}
@@
   \end{align}
-  and similarly the luminosity is
+  and the total luminosity in bin $b$ is
   \begin{align}
     \lleq{dfe}
-    \ell_b = \sum_{j=1}^{\npack}\ell_j \mathbf{1}_{[\nu_b,\nu_{b+1}]}( \nu_j ),
-  \end{align}
-  where $\mathbf{1}_V(x)$ is the indicator function requiring $x \in V$.
+    \ell_b = \sum_{j=1}^{\npack}\ell_j \mathbf{1}_{[\nu_b,\nu_{b+1}]}( \nu_j )\,,
+  \end{align}
+  with $\npack = \sum_b n_b$ and the total supernova luminosity $\ell
+  = \sum_b \ell_b$.
+
 \end{mtemize}
+\hans{Discussion of rebinning with index $m$?}
 
@@
 
-While we treat the TARDIS output as ``data'' as discussed above, a
-repeat of the simulation with a different Monte Carlo seed would
+While we treat the TARDIS output as fixed ``data'' as discussed above,
+a repeat of the simulation with a different Monte Carlo seed would
 evidently yield different ``data'' $D'$ even when all other simulation
 parameters remain the same. Clearly, $n_b$ is a specific realization
-of a stochastic variable $N_b$ governed by a model pdf
-$p(N_b)$. \fred{In our general case, $\npack = \sum_b n_b$ is an instance of a
+of a stochastic variable $N_b$ governed by a model pdf $p(N_b)$. %
+\fred{In our general case, $\npack = \sum_b n_b$ is an instance of a
   random variable because some packets leave the supernova, some
@@
   be chosen and not random. We are fortunate that this difference does
-  not matter because we consider bins independently. } Similarly,
-$\bml_b$ is a specific realization of a collection of stochastic
-variables $\bmL_b =\{L_{b,1},\ldots,L_{b,\smN_b}\}$, where the number
-$N_b$ of such variables is itself stochastic. The relevant pdf for
-$\bmL_b$ therefore always presupposes knowledge of $N_b$, so we must
-work with the conditional probability $p(\bmL_b\cond N_b)$. This
-situation is captured mathematically as $N_b \sim p(N_b), \bmL_b \sim
-p(\bmL_b\cond N_b)$ where $\sim$ means ``is by hypothesis distributed
-in this model according to the pdf'', i.e.\
+  not matter because we consider bins independently. } %
+\hans{I will come back to this later} %
+Similarly, $\bml_b$ is a specific realization of a collection of
+stochastic variables $\bmL_b =\{L_{b,1},\ldots,L_{b,\smN_b}\}$, where
+the number $N_b$ of such variables is itself stochastic. The relevant
+pdf for $\bmL_b$ therefore always presupposes knowledge of $N_b$, so
+we must work with the conditional probability $p(\bmL_b\cond
+N_b)$. This situation is captured mathematically as $N_b \sim p(N_b),
+\bmL_b \sim p(\bmL_b\cond N_b)$ where $\sim$ means ``is by hypothesis
+distributed in this model according to the pdf'', i.e.\
 \begin{align*}
@@
 \begin{mtemize}
-\item We omit the dependence on supernova parameters $\bmtheta$ from
-  the notation below and instead condition on the TARDIS output
-  $\npack, \bml, \bmnu$ directly.
+\item We omit the dependence on the physical supernova parameters
+  $\bmtheta$ from the notation below and instead condition on the
+  TARDIS output $\npack, \bml, \bmnu$ directly.
 
@@
   multiplicities $N_b$ and total luminosities $\bmL_b$ in any given
-  bin $b$ are statistically independent of those in other bins.
-\fred{We could motivate this in the usual way as the Poisson limit of the Binomial.
-The subtlety would then the be the random total number of packets (leaving the supernova).} This
-  means that the joint probability over all bins factorises into
+  bin $b$ are statistically independent of those in other bins. %
+  \fred{We could motivate this in the usual way as the Poisson limit
+    of the Binomial. The subtlety would then the be the random total
+    number of packets (leaving the supernova).} %
+  \hans{The limit would be from the multinomial with $B$ bins and
+    $\npack$ packtes to product of Poissonians. One could put it in,
+    but in my opinion it's not necessary. $B$ is large, so the
+    constraint plays no role.} %
+  This means that the joint probability over all bins factorises into
   individual probabilities for each bin,
@@
     linked to limit of Multinomial, see above} with parameter
-  $\lambda$
+  $\lambda$ %
+  \hans{I would change the Poisson parameter to $\nu$, see above comment.}
   \begin{align}
@@
   packets. Necessarily, $a$ is best determined for the frequency
-  around which the density of packets is highest.
+  around which the density of packets is highest. %
+  \hans{I would replace the word ``sample'' above and throughout the
+    text by ``packet''. The word ``sample'' refers in statistics to
+    the entire data set $D$, not just to one data point.} %
 
@@
 
-\item Priors: metaparameter $a$ \fred{?}
-
+\item Priors: metaparameter $a$ \fred{?}  %
+  \hans{This is a placeholder to remind us that metaparameters need to
+    be discussed at some point. Perhaps not here, but we should
+    remember.}
 \end{mtemize}
@@
 finding the moment generating function (mgf) (see Appendix) which in
-the present case is also the Laplace transform. \fred{On Wikipedia,
-  the MGF is defined as $\int_0^\infty dL\,p(L\cond
-  n,\bml)\,e^{+tL}$. Then there are less minus signs to worry about}
+the present case is also the Laplace transform. %
+\fred{On Wikipedia, the MGF is defined as $\int_0^\infty dL\,p(L\cond
+  n,\bml)\,e^{+tL}$. Then there are less minus signs to worry about} %
+\hans{Yes, most books seem to use the plus sign. It makes use of
+  Laplace transform harder, but I am happy to use plus.} %
 For the mgf, the convolution is easily accomplished:
@@
 \end{align}
+\hans{So far, $m$ appears only in your todo comment above. I would suggest
+  a short discussion of this already in Section \ref{sec:dfnr}. We
+  could then simply rewrite all equations below in terms of $m$ rather
+  than $b$.} %
 We observe good agreement between the best fit of a Gamma distribution
@@
   \includegraphics[width=0.48\textwidth]{gamma}
-  \fred{Plot with more samples to see dip at $x=0$.}
+  \fred{Plot with more samples to see dip at $x=0$.} %
+  \hans{I assume the KDE fit will disappear from the paper later as it
+    overshoots the zero point.}
   \caption{Gamma distribution fit. }
@@
 The following approximations can be used in different regions of the
-parameter space spanned by $n$ and $m$.
-
+parameter space spanned by $n$ and $m$. %
+\hans{$n$ is TARDIS output ``data'', so not a parameter. From the
+  above paragraph, $m$ seems to be a parameter governing rebinning to
+  nonuniform bins. I would have no idea how to handle $m$ formally in
+  a Bayesian way. If $m{=}150$ is the result of optimising the fitting
+  stability, then presumably it will be kept constant. In that case,
+  it would be treated like a constant rather than a parameter in the
+  usual sense.}
 \begin{mnumerate}

### 2016-03-09 hans/bayes-tardis-luminosity-prediction.tex
(wip) remove m

@@
 \usepackage[numbers]{natbib}
-%\usepackage{hyperref}
+\usepackage{hyperref}
 \synctex=1
@@
 %% CHANGE THIS DATE AS YOU MODIFY THE FILE
-\newcommand{\vdate}{160228}
+\newcommand{\vdate}{160309}
 \pagestyle{myheadings}
@@
 \newcommand{\Lmax}{L_{\rm max}}
+\newcommand{\rmdx}[1]{\mbox{d} #1 \,} % differential
 
@@
     $b$-bins into compound bins with index $m$ as this improves the
-    statistics.
+    statistics.%
+    \fred{This was an intermediate idea, I now removed any reference
+      to it. There should be no more binning when infering $\bmphi$}
 
@@
     simplified to some approximation, making use of the moment
-    generating function eg Eq (mgd) unnecessary.
+    generating function eg Eq (mgd) unnecessary.%
+    \fred{Yes, so you put in a lot of work to derive the equations and
+      I checked them carefully but I think in the end we will not need
+      them and we should then remove them from the draft}
 
@@
       of (say) the Gamma parameter $\alpha$ on $\nu$ is, and that is
-      of course of great interest anyway.
+      of course of great interest anyway.%
+      \fred{Yes, that's equivalent. I just started with a polynomial
+        for $a$ but that's ad-hoc.}
 
@@
       complications which we thought had disappeared. This would have
-      to be investigated. If on the other hand you also want to
-      treat $a(\nu,\bmphi)$ as some kind of ``constant function'',
-      then one would have to set out how stable this constant function
-      would be against various underlying changes. In other words, 
-      if calculation of $a()$ is not part of the general algorithm,
-      how is it fixed? and can it be fixed for all cases?
-
-    \item I think there may be a problem of combining the two proposed
-      simplifications $L \to X$ and $m$ compound bins. Depending on
-      the variability of the parametrisation as function of $\nu$, the
-      shape of $p(\bmL\cond N,\bmphi)$ for single bins with large
-      $\Lmax$ may differ significantly from the shape for many small
-      bins combined into one using $m$. In other words, rebinning
-      and rescaling may not be compatible.
+      to be investigated. If on the other hand you also want to treat
+      $a(\nu,\bmphi)$ as some kind of ``constant function'', then one
+      would have to set out how stable this constant function would be
+      against various underlying changes. In other words, if
+      calculation of $a()$ is not part of the general algorithm, how
+      is it fixed? and can it be fixed for all cases?%
+      \fred{I don't see what you mean by ``constant function''. The
+        choice of $a$ and how it depends on $\bmphi$ is part of the
+        overal model.}
+
+      \item I think there may be a problem of combining the two
+      proposed simplifications $L \to X$ and $m$ compound
+      bins. Depending on the variability of the parametrisation as
+      function of $\nu$, the shape of $p(\bmL\cond N,\bmphi)$ for
+      single bins with large $\Lmax$ may differ significantly from the
+      shape for many small bins combined into one using $m$. In other
+      words, rebinning and rescaling may not be compatible.%
+      \fred{No more compound bins, we just do an unbinned fit wrt
+        $\bmphi$, so there is no issue}
     \end{enumerate}
@@
 
-  \item We treat the results of a single TARDIS run as \textit{data}
+  \item We treat the results of a single TARDIS run as \textit{simulated data}
     %%%%%
@@
     %%%%%
-    $D$, i.e.\ fixed results which we do not question but accept as
-    given.  This ``data'' consists of the total number of packets
+    $S$, i.e.\ fixed results which we do not question but accept as
+    given.  These ``data'' consist of the total number of packets
     leaving the supernova $\npack$ plus the set of luminosities
@@
     \lleq{dfc}
-    D = \{ \npack, \bml, \bmnu \} \,,
+    S = \{ \npack, \bml, \bmnu \} \,,
   \end{align}
@@
   \end{align}
-  In terms of the $\mathbf{1}_V(x)$ the indicator function requiring
-  $x \in V$, the number of packets in bin $b$ is
+  In terms of $\mathbf{1}_V(x)$---the indicator function requiring
+  $x \in V$---the number of packets in bin $b$ is
   \begin{align}
@@
 \end{mtemize}
-\hans{Discussion of rebinning with index $m$?}
 
@@
 a repeat of the simulation with a different Monte Carlo seed would
-evidently yield different ``data'' $D'$ even when all other simulation
+evidently yield different ``data'' $S'$ even when all other simulation
 parameters remain the same. Clearly, $n_b$ is a specific realization
@@
   \emph{all} packets.
+  %
+  \fred{Perhaps it doesn't simplify the reading experience to omit $b$?}
 
@@
   For example, assume $p(L_j)$ depends on one parameter $a$. If we fit
-  separately in each bin, we would have $B+1$ fits and $B+1$
-  parameters. In each such fit, there would be only few samples to
-  constrain $a$ in each bin. Then instead of fitting $a$ separately in
-  each bin, we assume $a = a(\nu, \bmphi)$. This greatly reduces the
-  number of parameters if $B$ is large because a handful of
-  parameters, $\bmphi$, can be inferred from the complete set of
-  packets. Necessarily, $a$ is best determined for the frequency
-  around which the density of packets is highest. %
-  \hans{I would replace the word ``sample'' above and throughout the
-    text by ``packet''. The word ``sample'' refers in statistics to
-    the entire data set $D$, not just to one data point.} %
+  separately in each bin, we would have $B+1$ fits and $B+1$ copies of
+  $a$. In each fit, there would be only few packets to constrain
+  $a$. To avoid that, we assume $a = a(\nu, \bmphi)$. For concreteness,
+  assume $\bmphi = (a_0, a_1)$ and
+  \begin{align}
+    \lleq{pmdc}
+    a(\nu, \bmphi) = a_0 +a_1 \nu
+  \end{align}
+  This greatly reduces the number of parameters from $B+1$ to 2 and
+  ensures that $a$ is continuous in $\nu$. $a$, or equivalently
+  $\bmphi$, is now determined from a single fit to all packets. The
+  uncertainty on $a$ is at a minimum for the frequency around which
+  the density of packets is highest. \fred{Refer to plot with
+    uncertainty band}
 
@@
 
-\item Priors: metaparameter $a$ \fred{?}  %
+  \item Priors: metaparameter $a$
+  %
   \hans{This is a placeholder to remind us that metaparameters need to
@@
   {\int d\lambda\, p(n\cond\lambda)\,p(\lambda)}
-\end{align}\checked{}
+\end{align}
 where $p(N\cond\lambda,n) = p(N\cond\lambda)$ is independent of $n$.
@@
 We shall use the properties of the Negative Binomial in the
-calculations below. For non-integer $a$, the Negative Binomial
-generalizes immediately to the Pólya distribution by replacing $x! \to
-\Gamma(x+1)$. Note that the distribution of $N$ given $n$ is
-independent of $\bmL$ and $\bml$ by construction.
+calculations below. \fred{Check if we still do later} For non-integer $a$,
+the Negative Binomial generalizes immediately to the Pólya
+distribution by replacing $x! \to \Gamma(x+1)$. Note that the
+distribution of $N$ given $n$ is independent of $\bmL$ and $\bml$ by
+construction but depends on $\bmnu$ implicitly through $n$.
 
@@
 
+\fred{One thing is bugging me: To predict the luminosity in a bin, I
+  should do this $ \int_{\nu_b}^{\nu_{b+1}} d{\nu} \, p(L, \nu) =
+  \int_{\nu_b}^{\nu_{b+1}} d{\nu} \, p(L \cond \nu) p(\nu) $. We want
+  to avoid modeling $p(\nu)$ because of the absorption lines, so we
+  assume $p(L \cond \nu)$ approx. indep. of $\nu$ in the bin, so $ \approx p(L \cond
+  \nu ) \int_{\nu_b}^{\nu_{b+1}} d{\nu} \, p( \nu) \equiv p(L \cond
+  \nu) \, p_b$. Since $p_b$ is unknown, we should marginalize over it
+  using our knowledge from Tardis output:
+  $$
+  p(L \cond \mbox{bin } b, \npack, \bmnu, \bml) = \int d{p_b}\, p(L \cond \nu) \, p_b \, p(p_b \cond \npack, \bmnu, \bml)
+  $$
+
+  So how can we turn $\int \rmdx{p_b} p_b \to \sum_N$? Where does
+  $\npack$ come into play? If we can answer this, then we have one
+  more hidden assumption clarified in the derivation below that was
+  written for a binned analysis but now we want to be unbinned wrt
+  $\bmnu, \bml$.  }
+
 \subsubsection{Direct probability}
 
-\fred{We should be more general: let's reserve $n$ for the number of
-  packets observed in a bin. Suppose this is $n=1$. From that, we can
-  hardly infer the luminosity distribution. Instead, we group the
-  packets in larger bins such that there are at least $m \gtrsim 300$
-  packets, and then assume the distribution doesn't change much with
-  frequency.
-
-  Hence we will learn $\bmphi$ from $m$ samples, and
-  $p(\bmphi\cond N,n,\bml) \to p(\bmphi\cond N,m,\bml)$ and we can
-  keep $\bml$ with the same meaning. So we replace $n \to m$ below
-  except in $p(N \cond n$)}
-
-Our goal of predicting $L$ given the TARDIS output $(\npack,\bml,
-\bmnu)$ is captured in the probability $p(L\cond \npack,\bml, \bmnu)$
-with all relevant parameters integrated out. We first note from
-\refeq{pmh} that $p(N\cond \npack,\bml, \bmnu) = p(N\cond n)$ does not
-depend on $\bml$ or $\bmnu$ directly but only through $n$, the number
-of packets in the bin; cf. \refeq{dff}. Furthermore, since $L$ is
-fully determined by $(N,\bmL)$,
+% \fred{We should be more general: let's reserve $n$ for the number of
+%   packets observed in a bin. Suppose this is $n=1$. From that, we can
+%   hardly infer the luminosity distribution. Instead, we group the
+%   packets in larger bins such that there are at least $m \gtrsim 300$
+%   packets, and then assume the distribution doesn't change much with
+%   frequency.
+
+%   Hence we will learn $\bmphi$ from $m$ samples, and
+%   $p(\bmphi\cond N,n,\bml) \to p(\bmphi\cond N,m,\bml)$ and we can
+%   keep $\bml$ with the same meaning. So we replace $n \to m$ below
+%   except in $p(N \cond n$)}
+
+Our goal of predicting $L$ in bin $b$ given the TARDIS output
+$(\npack,\bml, \bmnu)$ is captured in the probability $p(L\cond \nu,
+\npack,\bml, \bmnu)$ with all relevant parameters integrated out using
+the law of total probability.
+
+To simplify the treatment, we assume that the frequency bins are
+narrow or equivalently that $p(L \cond \nu, \npack, \bml, \bmnu)$ does
+not vary within a bin. This implies that all hypothetical samples
+$L_j$ in the bin follow the same distribution.  As the reference
+frequency, we choose the bin center;
+% and avoid integration over the bin
+i.e., for each bin $b$, we set
+\begin{align}
+  \lleq{pmq}
+  p(L\cond \nu, \npack,\bml, \bmnu) = p(L\cond (\nu_b + \nu_{b+1}) / 2, \npack,\bml, \bmnu)  \; \forall \nu \in [\nu_b, \nu_{b+1}] \, .
+\end{align}
+
+We first note from \refeq{pmh} that $N$ does not depend on $\bml$ at
+all and only indirectly depends on $\bmnu$ through $n$, the number of
+packets in the bin; cf. \refeq{dff}. Furthermore, since $L$ is fully
+determined by $(N,\bmL)$,
 \begin{align}
@@
 \end{align}
-is does not depend directly on $(\npack),\bml,\bmnu)$, i.e.\
+is does not depend on $(\npack,\bml,\bmnu)$, i.e.\
 \begin{align}
   \lleq{pmi}
-  p(L\cond N,\bmL,n,\bml, \bmnu)
-  = p(L\cond N,\bmL)
-  &= \delta(L - \textstyle\sum_{j=1}^N L_j).
-\end{align}
-Hence the prediction is given by
+  p(L\cond N,\nu,\bmL,\npack,\bml, \bmnu)
+  = p(L\cond N, \bmL)
+  &= \delta(L - \textstyle\sum_{j=1}^N L_j),
+\end{align}
+where the $L_j$ are assumed to occur at $\nu$.
+% For notational simplicity, we now
+% suppress $\nu$ and remember that all probabilities below are
+% conditional on $\nu$.
+Then the prediction is given by
 \begin{align}
   \lleq{pmj}
-  p(L\cond n,\bml, \bmnu)
-  &= \sum_{N=0}^\infty p(L\cond N,n,\bml, \bmnu)\,p(N\cond n,\bml, \bmnu) \\
+  p(L\cond \nu, \npack,\bml, \bmnu)
+  &= \sum_{N=0}^\infty p(L\cond N,\nu, \npack,\bml, \bmnu)\,p(N\cond \nu, \npack,\bml, \bmnu) \\
   \lleq{pmjb}
-  &= \sum_N p(N\cond n)\,\int d\bmL\, p(L\cond N,\bmL,n,\bml, \bmnu)\, p(\bmL\cond N,n,\bml, \bmnu)
+  &= \sum_N p(N\cond n)\,\int d\bmL\, p(L\cond N,\bmL,\nu, \npack,\bml, \bmnu)\, p(\bmL\cond N,\nu, \npack,\bml, \bmnu)
   \\
@@
   &= \sum_N p(N\cond n)\,\int d\bmL\,\delta(L - \textstyle\sum_{j=1}^N L_j)
-  \, p(\bmL\cond N,n,\bml, \bmnu)
-\end{align}
-\fred{Add missing $\bmnu$! Everything that is conditional on $\bml$ is
-  also conditional on $\bmnu$} The dependence on the TARDIS ``data''
-$(\npack,\bml, \bmnu)$ enters through $p(\bmL\cond N,\npack,\bml,
-\bmnu)$ in terms of the parameters $\bmphi$ of $p(\bmL\cond
-N,\bmphi)$,
+  \, p(\bmL\cond N,\nu, \npack,\bml, \bmnu)
+\end{align}
+The dependence on the TARDIS output $(\npack,\bml, \bmnu)$ enters
+through $p(\bmL\cond N,\npack,\bml, \bmnu)$ in terms of the parameters
+$\bmphi$ of $p(\bmL\cond N,\bmphi)$,
 \begin{align}
   \lleq{pmk}
-  p(\bmL\cond N,\npack,\bml, \bmnu)
-  &= \int d\bmphi\, p(\bmL\cond N,\bmphi,\npack,\bml, \bmnu)\,p(\bmphi\cond N,\npack,\bml, \bmnu) \\
-  &= \int d\bmphi\, p(\bmL\cond N,\bmphi)\,p(\bmphi\cond N,\npack,\bml, \bmnu).
-\end{align}
-because the likelihood for $\bmL$ is fully determined by $N,\bmphi$
-and hence independent of $(\npack,\bml, \bmnu)$.
+  p(\bmL\cond N,\nu, \npack,\bml, \bmnu)
+  &= \int d\bmphi\, p(\bmL\cond N,\bmphi,\nu, \npack,\bml, \bmnu)\,p(\bmphi\cond N,\nu, \npack,\bml, \bmnu) \\
+  &= \int d\bmphi\, p(\bmL\cond N,\bmphi, \nu)\,p(\bmphi\cond N,\nu, \npack,\bml, \bmnu).
+\end{align}
+because $\bmL$ is fully determined by $N,\bmphi, \nu$ and hence independent
+of $\npack,\bml, \bmnu$.
 %
@@
   \lleq{pmkb}
-  p(\bmphi\cond N,\npack,\bml, \bmnu) &=
-  \frac{p(\bml\cond N,\npack,\bmphi,\bmnu)\,p(\bmphi\cond N,\npack,\bmnu)} %
-  {\int d\bmphi\,p(\bml\cond N,\npack,\bmphi,\bmnu)\,p(\bmphi\cond N,\npack,\bmnu)}, %
+  p(\bmphi\cond N,\nu, \npack,\bml, \bmnu) &=
+  \frac{p(\bml\cond \bmphi, N,\nu, \npack,\bmnu)\,p(\bmphi\cond N,\nu, \npack,\bmnu)} %
+  {\int d\bmphi\,p(\bml\cond \bmphi,N,\nu, \npack,\bmnu)\,p(\bmphi\cond N,\nu, \npack,\bmnu)}, %
   % \ =\ \frac{p(\bml\cond N,\npack,\bmphi)\,p(\bmphi\cond N,n)} %
@@
 \end{align}
-
- a multiplicity-independent prior for $\bmphi$
+using a multiplicity- and frequency-independent prior for $\bmphi$
 \begin{align}
   \lleq{pml}
-  p(\bmphi\cond N,\npack,\bmnu) &= p(\bmphi),
-\end{align}
-and recognizing that $N$ does not tell us anything about $\bml$
+  p(\bmphi\cond N,\nu, \npack,\bmnu) &= p(\bmphi),
+\end{align}
+and recognizing that $\bml$ is independent of $N$ and $\nu$
 \begin{align}
   \lleq{pmla}
-  p(\bml\cond N,\npack,\bmphi,\bmnu) = p(\bml\cond \npack,\bmphi,\bmnu)
+  p(\bml\cond \bmphi,N,\nu, \npack,\bmnu) = p(\bml\cond \bmphi,\npack,\bmnu)
 \end{align}
@@
   \lleq{pmn}
-  p(\bmL\cond N,\npack,\bml, \bmnu)
+  p(\bmL\cond N,\nu, \npack,\bml, \bmnu)
   &= \int d\bmphi\, p(\bmL\cond N,\bmphi)\, %
-  \frac{p(\bml\cond \npack,\bmphi)\,p(\bmphi)} %
-  {\int d\bmphi\,p(\bml\cond \npack,\bmphi)\,p(\bmphi)} \\
+  \frac{p(\bml\cond \bmphi, \npack, \bmnu)\,p(\bmphi)} %
+  {\int d\bmphi\,p(\bml\cond \bmphi,\npack,\bmnu)\,p(\bmphi)} \\
   \lleq{pmo}
-  &= \frac{1}  {p(\bml\cond n)} %
-  \int d\bmphi\, p(\bmL\cond N,\bmphi)\, %
-  p(\bml\cond \npack,\bmphi)\,p(\bmphi)
-\end{align}
-with $p(\bml\cond n) = \int d\bmphi\, p(\bml\cond
-\npack,\bmphi)\,p(\bmphi)$ in the denominator.
-Inserting this into (\ref{pmjc}), we obtain
+  &= \frac{1}  {p(\bml\cond \npack, \bmnu)} %
+  \int d\bmphi\, p(\bmL\cond \bmphi, N)\, %
+  p(\bml\cond \bmphi, \npack,\bmnu)\,p(\bmphi)
+\end{align}
+with $p(\bml\cond \npack, \bmnu) = \int d\bmphi\, p(\bml\cond \bmphi,
+\npack,\bmnu)\,p(\bmphi)$ in the denominator.  Inserting this into
+(\ref{pmjc}), we obtain the desired result:
 \begin{equation}
@@
   \boxed{
-  p(L\cond \npack,\bml, \bmnu)
-  = \frac{1}{p(\bml\cond n)}
+  p(L\cond \nu, \npack,\bml, \bmnu)
+  = \frac{1}{p(\bml\cond \npack, \bmnu)}
   \sum_N p(N\cond n)\,\int d\bmL\,\delta(L - \textstyle\sum_j L_j)
-  \displaystyle\int d\bmphi\,p(\bmL\cond N,\bmphi) \, p(\bml\cond \npack,\bmphi)
+  \displaystyle\int d\bmphi\,p(\bmL\cond N,\bmphi) \, p(\bml\cond \bmphi,\npack,\bmnu)
   \, p(\bmphi)
@@
 \begin{align}
-  p(\bmL\cond N,\bmphi) &= \prod_{j=1}^N p(L_j\cond \bmphi) \\
-  p(\bml\cond \npack,\bmphi) &= \prod_{j=1}^n p(\ell_j\cond \bmphi)
-\end{align}
-\fred{And $p(\cdot \cond \bmphi)$ is the same function in both equations.}
+  p(\bmL\cond \bmphi, N, \nu) &= \prod_{j=1}^N p(L_j\cond \bmphi, \nu), \\
+  p(\bml\cond \bmphi,\npack,\bmnu) &= \prod_{j=1}^{\npack} p(\ell_j\cond \bmphi, \nu_j)
+\end{align}
+and $p(\cdot \cond \bmphi, \cdot)$ is the same function in both equations.
 In other words: Once we have decided on a prior $p(\bmphi)$ and a
@@
 \\
+
+\fred{Dependence on $\nu$ is suppressed in this section}
 
@@
 \beta)$, we have to transform the luminosities. To this end, we first
-take the largest luminosity from the entire frequency range and
-increase it slightly to avoid numerical difficulties at the end point
+take the largest luminosity of all TARDIS packets across the entire
+frequency range and increase it slightly to avoid numerical
+difficulties at the end point
 \begin{align}
@@
 \end{align}
-and similarly for $L \to X$, then we infer $P(X \cond n, \bmx,
-m)$. Only in the end, we transform back to a probability density in
-$L$ via
+and similarly for $L \to X$, then we infer $P(X \cond \nu,  \npack, \bmx,
+\bmnu)$. Only in the end, we transform back to a probability density
+in $L$ via
 \begin{align}
   \lleq{gmg}
-  p(L  \cond n, \bml, m) = p(X  \cond n, \bmx, m) \frac{1}{\lmax}
-\end{align}
-\hans{So far, $m$ appears only in your todo comment above. I would suggest
-  a short discussion of this already in Section \ref{sec:dfnr}. We
-  could then simply rewrite all equations below in terms of $m$ rather
-  than $b$.} %
-We observe good agreement between the best fit of a Gamma distribution
-to TARDIS samples; an example is shown shown in \reffig{gamma}.  While
-not obvious at first, a major numerical simplification arises because
-of the $L \to X$ transformation. Rewriting the master
+  p(L  \cond \nu,  \npack, \bml, \bmnu) = p(X  \cond \nu,  \npack, \bmx, \bmnu) \frac{1}{\lmax}
+\end{align}
+\hans{So far, $m$ appears only in your todo comment above. I would
+  suggest a short discussion of this already in Section
+  \ref{sec:dfnr}. We could then simply rewrite all equations below in
+  terms of $m$ rather than $b$.} %
+\fred{I removed $m$ whereever I saw it} We observe good agreement
+between the best fit of a Gamma distribution to TARDIS packet
+luminosities; an example is shown shown in \reffig{gamma}.  While not
+obvious at first, a major numerical simplification arises because of
+the $L \to X$ transformation. Rewriting the master
 equation~(\ref{pmp}) formally as
@@
   \lleq{gmh}
-  p(L  \cond n, \bml, m) = \sum_{N} p(N \cond n) p(L \cond N, \bml, m)
+  p(L  \cond \nu,  \npack, \bml, \bmnu) = \sum_{N} p(N \cond \nu,  n) p(L \cond N,
+\nu, \npack, \bml, \bmnu)
 \end{align}
@@
 application, the distribution of a single luminosity, $p(L \cond N=1,
-\bml, m)$ is narrow in the sense that the standard deviation relative
-to the mean is at the percent level. This implies that $p(L \cond n,
-\bml, m)$ has one mode from each $N$ and all modes are clearly
-separated. The transformation $L \to x$ effects a strong overlap
-between adjacent $p(X \cond N, \bmx, m)$, so $p(X \cond n, \bmx, m)$
-can be well approximated by a unimodal distribution; see
-\reffig{hedgehog}.  \fred{I'm not 100 \% sure but $p(X \cond n, \bmx,
-  m)$ should still have $N \to \infty$ modes but they aren't visible}
+\nu, \bml, \bmnu)$ is narrow in the sense that the standard deviation
+relative to the mean is at the percent level. This implies that $p(L
+\cond \nu, \npack, \bml, \bmnu)$ has one mode from each $N$ and all
+modes are clearly separated. The transformation $L \to x$ effects a
+strong overlap between the distributions $p(X \cond N, \nu, \npack, \bmx, \bmnu)$
+with similar $N$, so $p(X \cond \nu, \npack, \bmx, \bmnu)$ can be well
+approximated by a unimodal distribution; see \reffig{hedgehog}.
+\fred{I'm not 100 \% sure but $p(X \cond \nu, \npack, \bmx, \bmnu)$ should still
+  have $N \to \infty$ modes but they aren't visible}
 \missingfigure{from hedgehog to single peak \label{fig:hedgehog}}
@@
   \lleq{gmb}
+  \boxed{
     X_j \sim \GammaDist(\alpha, \beta) \Rightarrow X \equiv \sum_{j=1}^N X_j \sim \GammaDist(N \alpha, \beta).
+    }
 \end{align}
@@
   \lleq{gmi}
-  \bmphi &= (\alpha, \beta),\\
-  p(X_j \cond \bmphi) &=
+  \alpha, \beta &= \alpha(\nu \cond \bmphi), \beta(\nu \cond \bmphi),\\
+  p(X_j \cond \bmphi, \nu) &=
   \GammaDist(X_j \cond \alpha, \beta),\\
-  p(\bmX \cond N, \bmphi) &= \prod_{j=1}^N p(X_j \cond \phi)
+  p(\bmX \cond N, \bmphi, \nu) &= \prod_{j=1}^N p(X_j \cond \bmphi, \nu)
 \end{align}
@@
   \lleq{gmc}
- p(X \cond N, \bmphi) &= \int d\bmX\,\delta(X - \textstyle\sum_j X_j)
-  \,p(\bmX\cond N,\bmphi) = \GammaDist(X \cond N \alpha, \beta)
+ p(X \cond N, \bmphi, \nu) &= \int d\bmX\,\delta(X - \textstyle\sum_j X_j)
+  \,p(\bmX\cond N,\bmphi, \nu) = \GammaDist(X \cond N \alpha, \beta)
 \end{align}
@@
   \lleq{gmd}
-    p(X\cond n,\bmx,m)
-  = \frac{1}{p(\bmx\cond m)}
-  \sum_N p(N\cond n)\, \int d\bmphi\, p(X \cond N, \bmphi)\, p(\bmx\cond m,\bmphi)
+    p(X\cond \nu, \npack,\bmx,\bmnu)
+  = \frac{1}{p(\bmx\cond \bmnu)}
+  \sum_N p(N\cond n)\, \int d\bmphi\, p(X \cond N, \bmphi, \nu)\, p(\bmx\cond \bmphi, \npack, \bmnu)
   \, p(\bmphi).
@@
 choices and conclude that the point estimate given by the mode of the
-posterior is nearly insensitive to the prior choice for 30 samples. We
-require in addition that the posterior is narrow enough so we group
-samples in bins to have at least $m=150$ samples in each bin. This
-implicitly assumes that the luminosity distribution is constant in the
-frequency range defined by $m$.
-
-Note that the bins defined by the experiment in general do not overlap
-with the bins used to infer the luminosity distribution; i.e., $n \ne
-m$.
+posterior is nearly insensitive to the prior choice for 30 samples in
+the iid case. Translated to our problem, this would be 30 packets at
+the same frequency. From TARDIS, we only get packets at different
+frequencies. In the following, let $m$ denote the effective number of
+packets close to $\nu$ where ``close'' is deliberately left
+unspecified.  If $m$ is large, we can well determine $\phi$
+and thus $p(X_j \cond N, \bmphi, \nu)$.
+
+% We require in addition that the posterior is narrow enough so we
+% group samples in bins to have at least $m=150$ samples in each
+% bin. This implicitly assumes that the luminosity distribution is
+% constant in the frequency range defined by $m$.
+
+% Note that the bins defined by the experiment in general do not overlap
+% with the bins used to infer the luminosity distribution; i.e., $n \ne
+% m$.
 
@@
   usual sense.}
+\fred{Yes, $n$ is fixed from TARDIS output. $m$ is defined only for this paragraph}
+\fred{Update approximations: don't need moments any longer}
 \begin{mnumerate}
-  \item \label{large-m} For $m$ large, the posterior $p(\bmphi \cond m, \bmx) =
-  p(\bmx\cond m,\bmphi) \, p(\bmphi) / p(\bmx \cond m)$ approaches a
-  multivariate Gaussian. We can then compute the various moments
-  $E_\bmphi[f(x)]$ with $f(x)=1,x,x^2$ by Laplace's method to good
-  accuracy.
-  \item \label{very-large-m} In the limit $m \to \infty$, the covariance of the posterior
-  $p(\bmphi \cond m, \bmx)$ tends to zero as $1/m$ and we can then fix
-  the parameters $\bmphi$ at the posterior mode $\bmphi^*$; i.e.,
-  $p(\bmphi \cond m, \bmx) \to \delta(\bmphi - \bmphi^*)$.
+  \item \label{large-m} For large $m$, the posterior $p(\bmphi \cond
+  \npack, \bmx, \bmnu) = p(\bmx\cond \bmphi, \npack, \bmnu) \, p(\bmphi) / p(\bmx \cond \npack, \bmnu)$
+  approaches a multivariate Gaussian. We can then compute the various
+  moments $E_\bmphi[f(x)]$ with $f(x)=1,x,x^2$ by Laplace's method to
+  good accuracy.
+  \item \label{very-large-m} In the limit $m \to \infty$, the
+  covariance of the posterior $p(\bmphi \cond \npack, \bmx, \bmnu)$
+  tends to zero as $1/m$ and we can then fix the parameters $\bmphi$
+  at the posterior mode $\bmphi^*$; i.e., $p(\bmphi \cond \npack,
+  \bmx, \bmnu) \to \delta(\bmphi - \bmphi^*)$.
   \item \label{large-n} For $n$ large, the cumulative of the Negative
@@
   $N$ in \refeq{gmd}. The contribution around $N=0$ is negligible, and
-  we can replace $\sum_n \to \int dN$. The density $p(N \cond n) P(X
+  we can replace $\sum_n \to \int \rmdx{N}$. The density $p(N \cond n) P(X
   \cond N,m, \bmx)$ is close to a Gaussian and the integral can be
@@
 % Local Variables:
-% compile-command:"rubber --pdf -W refs -W misc -S bayes-tardis-luminosity-prediction"
+% compile-command:"rubber --pdf -W refs -S bayes-tardis-luminosity-prediction"
 % End:

### 2016-03-14 hans/bayes-tardis-luminosity-prediction.tex
[draft] macro for differential

@@
     = \frac{p(D\cond\theta,\hmod)\;p(\theta\cond\hmod)}
-    {\int d\theta'\; p(D\cond\theta',\hmod)\;p(\theta'\cond\hmod)}
+    {\int \rmdx{\theta'} p(D\cond\theta',\hmod)\;p(\theta'\cond\hmod)}
   \end{align}
@@
     p(\theta\cond D,\hmod)
-    = \frac{p(D\cond\theta,\hmod)} {\int d\theta'\; p(D\cond\theta',\hmod)}
+    = \frac{p(D\cond\theta,\hmod)} {\int \rmdx{\theta'} p(D\cond\theta',\hmod)}
   \end{equation}
@@
     p(x_{n+1}\cond D,\hmod)
-    &= \int d\theta'\; p(x_{n+1}\cond\theta',D,\hmod)\;p(\theta'\cond
+    &= \int \rmdx{\theta'} p(x_{n+1}\cond\theta',D,\hmod)\;p(\theta'\cond
     D,\hmod)
@@
     p(x_{n+1}\cond D,\hmod)
-    &= \frac{\int d\theta'\; p(x_{n+1}\cond\theta',D,\hmod)\;
+    &= \frac{\int \rmdx{\theta'} p(x_{n+1}\cond\theta',D,\hmod)\;
       p(D\cond\theta',\hmod)\;p(\theta'\cond\hmod)}
-    {\int d\theta'\; p(D\cond\theta',\hmod)\;p(\theta'\cond\hmod)}
+    {\int \rmdx{\theta'} p(D\cond\theta',\hmod)\;p(\theta'\cond\hmod)}
   \end{align}
@@
   \frac{p(D\cond\theta',\hmod)\;p(\theta'\cond\hmod)}
-  {\int d\theta'\; p(D\cond\theta',\hmod)\;p(\theta'\cond\hmod)}
+  {\int \rmdx{\theta'} p(D\cond\theta',\hmod)\;p(\theta'\cond\hmod)}
   &=
   \frac{p(D,\theta'\cond\hmod)}
-  {\int d\theta'\; p(D,\theta'\cond\hmod)}
+  {\int \rmdx{\theta'} p(D,\theta'\cond\hmod)}
   \end{align}
@@
   \ =\ \frac{p(n\cond\lambda)\,p(\lambda)} %
-  {\int d\lambda\, p(n\cond\lambda)\,p(\lambda)}
+  {\int \rmdx{\lambda} p(n\cond\lambda)\,p(\lambda)}
 \end{align}
@@
   p(N\cond n)
-  &= \int d\lambda\,p(N,\lambda\cond n)
-  \ =\ \int d\lambda\,p(N\cond\lambda, n)\,p(\lambda\cond n)
-  \ =\ \frac{\int d\lambda\,p(N\cond\lambda)\,p(n\cond\lambda)\,p(\lambda)}
-  {\int d\lambda\, p(n\cond\lambda)\,p(\lambda)}
+  &= \int \rmdx{\lambda}p(N,\lambda\cond n)
+  \ =\ \int \rmdx{\lambda} p(N\cond\lambda, n)\,p(\lambda\cond n)
+  \ =\ \frac{\int \rmdx{\lambda} p(N\cond\lambda)\,p(n\cond\lambda)\,p(\lambda)}
+  {\int \rmdx{\lambda} p(n\cond\lambda)\,p(\lambda)}
 \end{align}
@@
   $$
-  p(L \cond \mbox{bin } b, \npack, \bmnu, \bml) = \int d{p_b}\, p(L \cond \nu) \, p_b \, p(p_b \cond \npack, \bmnu, \bml)
+  p(L \cond \mbox{bin } b, \npack, \bmnu, \bml) = \int \rmdx{p_b} p(L \cond \nu) \, p_b \, p(p_b \cond \npack, \bmnu, \bml)
   $$
@@
   \lleq{pmjb}
-  &= \sum_N p(N\cond n)\,\int d\bmL\, p(L\cond N,\bmL,\nu, \npack,\bml, \bmnu)\, p(\bmL\cond N,\nu, \npack,\bml, \bmnu)
+  &= \sum_N p(N\cond n)\,\int \rmdx{\bmL} p(L\cond N,\bmL,\nu, \npack,\bml, \bmnu)\, p(\bmL\cond N,\nu, \npack,\bml, \bmnu)
   \\
   \lleq{pmjc}
-  &= \sum_N p(N\cond n)\,\int d\bmL\,\delta(L - \textstyle\sum_{j=1}^N L_j)
+  &= \sum_N p(N\cond n)\,\int \rmdx{\bmL} \delta(L - \textstyle\sum_{j=1}^N L_j)
   \, p(\bmL\cond N,\nu, \npack,\bml, \bmnu)
@@
   p(\bmL\cond N,\nu, \npack,\bml, \bmnu)
-  &= \int d\bmphi\, p(\bmL\cond N,\bmphi,\nu, \npack,\bml, \bmnu)\,p(\bmphi\cond N,\nu, \npack,\bml, \bmnu) \\
-  &= \int d\bmphi\, p(\bmL\cond N,\bmphi, \nu)\,p(\bmphi\cond N,\nu, \npack,\bml, \bmnu).
+  &= \int \rmdx{\bmphi} p(\bmL\cond N,\bmphi,\nu, \npack,\bml, \bmnu)\,p(\bmphi\cond N,\nu, \npack,\bml, \bmnu) \\
+  &= \int \rmdx{\bmphi} p(\bmL\cond N,\bmphi, \nu)\,p(\bmphi\cond N,\nu, \npack,\bml, \bmnu).
 \end{align}
@@
   \frac{p(\bml\cond \bmphi, N,\nu, \npack,\bmnu)\,p(\bmphi\cond N,\nu, \npack,\bmnu)} %
-  {\int d\bmphi\,p(\bml\cond \bmphi,N,\nu, \npack,\bmnu)\,p(\bmphi\cond N,\nu, \npack,\bmnu)}, %
+  {\int \rmdx{\bmphi} p(\bml\cond \bmphi,N,\nu, \npack,\bmnu)\,p(\bmphi\cond N,\nu, \npack,\bmnu)}, %
   % \ =\ \frac{p(\bml\cond N,\npack,\bmphi)\,p(\bmphi\cond N,n)} %
@@
   p(\bmL\cond N,\nu, \npack,\bml, \bmnu)
-  &= \int d\bmphi\, p(\bmL\cond N,\bmphi)\, %
+  &= \int \rmdx{\bmphi} p(\bmL\cond N,\bmphi)\, %
   \frac{p(\bml\cond \bmphi, \npack, \bmnu)\,p(\bmphi)} %
-  {\int d\bmphi\,p(\bml\cond \bmphi,\npack,\bmnu)\,p(\bmphi)} \\
+  {\int \rmdx{\bmphi} p(\bml\cond \bmphi,\npack,\bmnu)\,p(\bmphi)} \\
   \lleq{pmo}
   &= \frac{1}  {p(\bml\cond \npack, \bmnu)} %
-  \int d\bmphi\, p(\bmL\cond \bmphi, N)\, %
+  \int \rmdx{\bmphi} p(\bmL\cond \bmphi, N)\, %
   p(\bml\cond \bmphi, \npack,\bmnu)\,p(\bmphi)
 \end{align}
-with $p(\bml\cond \npack, \bmnu) = \int d\bmphi\, p(\bml\cond \bmphi,
+with $p(\bml\cond \npack, \bmnu) = \int \rmdx{\bmphi} p(\bml\cond \bmphi,
 \npack,\bmnu)\,p(\bmphi)$ in the denominator.  Inserting this into
@@
   = \frac{1}{p(\bml\cond \npack, \bmnu)}
-  \sum_N p(N\cond n)\,\int d\bmL\,\delta(L - \textstyle\sum_j L_j)
-  \displaystyle\int d\bmphi\,p(\bmL\cond N,\bmphi) \, p(\bml\cond \bmphi,\npack,\bmnu)
+  \sum_N p(N\cond n)\,\int \rmdx{\bmL} \delta(L - \textstyle\sum_j L_j)
+  \displaystyle\int \rmdx{\bmphi} p(\bmL\cond N,\bmphi) \, p(\bml\cond \bmphi,\npack,\bmnu)
   \, p(\bmphi)
@@
   &= \frac{1}{p(\bml\cond n)}
-  \sum_N p(N\cond n)\,\int dL e^{-tL} \int d\bmL\,\delta(L - \textstyle\sum_j L_j)
-  \displaystyle\int d\bmphi\,p(\bmL\cond N,\bmphi) \, p(\bml\cond n,\bmphi)
+  \sum_N p(N\cond n)\,\int \rmdx{L} e^{-tL} \int \rmdx{\bmL} \delta(L - \textstyle\sum_j L_j)
+  \displaystyle\int \rmdx{\bmphi} p(\bmL\cond N,\bmphi) \, p(\bml\cond n,\bmphi)
   \, p(\bmphi)\\
   &= \frac{1}{p(\bml\cond n)}
-  \sum_N p(N\cond n)\,\int d\bmphi\,
-  \prod_{j=1}^N \left[ \int dL_j\,p(L_j\cond \bmphi)\,e^{-tL_j}\right]
+  \sum_N p(N\cond n)\,\int \rmdx{\bmphi}
+  \prod_{j=1}^N \left[ \int \rmdx{L_j} p(L_j\cond \bmphi)\,e^{-tL_j}\right]
    p(\bml\cond n,\bmphi) \, p(\bmphi)
@@
   M(t; L_j\cond \bmphi) &=
-  \int dL_j\,p(L_j\cond \bmphi)\,e^{-tL_j}
+  \int \rmdx{L_j} p(L_j\cond \bmphi)\,e^{-tL_j}
 \end{align}
@@
   = \frac{1}{p(\bml\cond n)}
-  \sum_N p(N\cond n)\,\int d\bmphi\,
+  \sum_N p(N\cond n)\,\int \rmdx{\bmphi}
   M(t; L_j\cond \bmphi)^N
@@
   \lleq{mgf}
-  \mu_q(\bmphi) &= \int dL_j\,p(L_j\cond\bmphi)\,L_j^q \qquad q = 1,2,\ldots,
+  \mu_q(\bmphi) &= \int \rmdx{L_j} p(L_j\cond\bmphi)\,L_j^q \qquad q = 1,2,\ldots,
 \end{align}
@@
   &= \frac{1}{p(\bml\cond n)}
-  \sum_N p(N\cond n)\,N \int d\bmphi\, \mu_1(\bmphi) \,
+  \sum_N p(N\cond n)\,N \int \rmdx{\bmphi} \mu_1(\bmphi) \,
   p(\bml\cond n,\bmphi) \, p(\bmphi).
@@
   &= \frac{n}{p(\bml\cond n)}
-  \int d\bmphi\, \mu_1(\bmphi) \, p(\bml\cond n,\bmphi) \, p(\bmphi)
+  \int \rmdx{\bmphi} \mu_1(\bmphi) \, p(\bml\cond n,\bmphi) \, p(\bmphi)
 \end{align}
@@
   \lleq{mgn}
-  E_\phi[ f ] &= \int d\bmphi\, f(\bmphi) \, p(\bml\cond n,\bmphi) \, p(\bmphi)
+  E_\phi[ f ] &= \int \rmdx{\bmphi} f(\bmphi) \, p(\bml\cond n,\bmphi) \, p(\bmphi)
 \end{align}
@@
   &= \frac{1}{p(\bml\cond n)}
-  \sum_N p(N\cond n)\, \int d\bmphi\,
+  \sum_N p(N\cond n)\, \int \rmdx{\bmphi}
   \left[ N^{\underline{2}}\mu_1^2(\bmphi) + N \mu_2(\bmphi)\right]
@@
   \lleq{btg}
-  \mu_q(a,b) &= \int dx\,x^q\,p(x\cond a,b)
+  \mu_q(a,b) &= \int \rmdx{x} x^q\,p(x\cond a,b)
   = \frac{(a)_q}{(a+b)_q}
@@
   \lleq{gmc}
- p(X \cond N, \bmphi, \nu) &= \int d\bmX\,\delta(X - \textstyle\sum_j X_j)
+ p(X \cond N, \bmphi, \nu) &= \int \rmdx{\bmX} \delta(X - \textstyle\sum_j X_j)
   \,p(\bmX\cond N,\bmphi, \nu) = \GammaDist(X \cond N \alpha, \beta)

### 2016-03-16 hans/bayes-tardis-luminosity-prediction.tex
[draft] Polish description of Gamma model

@@
 \DeclareMathOperator{\GammaDist}{Gamma}
+\newcommand{\Kalpha}{{K_\alpha}}
+\newcommand{\Kbeta}{{K_\beta}}
 \newcommand{\npack}{n_p}
@@
   terms of $m$ rather than $b$.} %
-\fred{I removed $m$ whereever I saw it} We observe good agreement
-between the best fit of a Gamma distribution to TARDIS packet
-luminosities; an example is shown shown in \reffig{gamma}.  While not
-obvious at first, a major numerical simplification arises because of
-the $L \to X$ transformation. Rewriting the master
-equation~(\ref{pmp}) formally as
+\fred{I removed $m$ whereever I saw it except below where I define it
+  explicitly}
+%
+We observe good agreement between the best fit of a Gamma distribution
+to TARDIS packet luminosities; an example is shown shown in
+\reffig{gamma}.  While not obvious at first, a major numerical
+simplification arises because of the $L \to X$
+transformation. Rewriting the master equation~(\ref{pmp}) formally as
 \begin{align}
@@
   \lleq{gmb}
-  \boxed{
     X_j \sim \GammaDist(\alpha, \beta) \Rightarrow X \equiv \sum_{j=1}^N X_j \sim \GammaDist(N \alpha, \beta).
-    }
-\end{align}
-With
+\end{align}
+Let us assume that $\alpha$ and $\beta$ are functions of $\nu$ with unknown parameters $\bmphi$
 \begin{align}
   \lleq{gmi}
-  \alpha, \beta &= \alpha(\nu \cond \bmphi), \beta(\nu \cond \bmphi),\\
+  \alpha &= \alpha(\nu \cond \bmphi),\\
+  \beta &= \beta(\nu \cond \bmphi),\\
   p(X_j \cond \bmphi, \nu) &=
   \GammaDist(X_j \cond \alpha, \beta),\\
-  p(\bmX \cond N, \bmphi, \nu) &= \prod_{j=1}^N p(X_j \cond \bmphi, \nu)
-\end{align}
- \refeq{gmb} is equivalent to
+  p(\bmX \cond N, \bmphi, \nu) &= \prod_{j=1}^N p(X_j \cond \bmphi, \nu).
+\end{align}
+With these definitions, \refeq{gmb} is equivalent to
 \begin{align}
@@
   \lleq{gmd}
+    \boxed{
     p(X\cond \nu, \npack,\bmx,\bmnu)
   = \frac{1}{p(\bmx\cond \bmnu)}
-  \sum_N p(N\cond n)\, \int d\bmphi\, p(X \cond N, \bmphi, \nu)\, p(\bmx\cond \bmphi, \npack, \bmnu)
+  \sum_N p(N\cond n)\, \int \rmdx{\bmphi} \GammaDist(X_j \cond \alpha, \beta) % p(X \cond N, \bmphi, \nu)
+  \, p(\bmx\cond \bmphi, \npack, \bmnu)
   \, p(\bmphi).
-\end{align}
-
-\subsection*{Approximations}
+  }
+\end{align}
+
+\subsubsection*{Linear model}
+
+In the absence of a physical model to determine $\alpha(\nu \cond
+\bmphi)$ and $\beta(\nu \cond \bmphi)$, we propose to start with a simple polynomial ansatz
+\begin{align}
+  \lleq{gmp}
+  \alpha(\nu \cond \bmphi) &= \sum_{k=0}^\Kalpha \alpha_k \nu^k,
+  &\beta(\nu \cond \bmphi) &= \sum_{k=0}^\Kbeta \beta_k \nu^k
+\end{align}
+such that
+\begin{align}
+  \lleq{gmq}
+  \bmphi = (\alpha_0, \dots \alpha_\Kalpha , \beta_0, \dots \beta_\Kbeta)
+\end{align}
+The usual trade off between parsimony and the ability to accurately
+model the frequency dependence determines the orders of the
+polynomials. This is outside the scope of this paper.
+
+\subsubsection*{Priors} \label{sec:priors}
+
+To explicitly evaluate \refeq{gmd}, we need to specify the prior
+$p(\bmphi)$. To gain some insight, we first consider the special case
+$\alpha = \bmphi_1$, $\beta = \bmphi_2$ where we ignore the dependence
+on $\nu$; i.e., $\Kalpha = \Kbeta = 0$. For this standard $\GammaDist$
+model, there is a conjugate prior but unfortunately the normalization
+constant cannot be evaluated in closed form, so it is thus of no use
+for us. How about a noninformative prior? The authors of
+\cite{moala2013bayesian} present an overview of possible choices and
+conclude that the point estimate given by the mode of the posterior is
+nearly insensitive to the prior choice starting from around 30
+independent identically distributed samples. We posit that there are
+this many samples available near any $\nu$ under consideration, so the
+prior is not crucial for our application.
+
+There is some explicit prior information that we want to build into
+our analysis. The distribution $p(x \cond \alpha, \beta)$ should
+always vanish at the origin. This is equivalent to requiring a maximum
+away from the boundary leading us to the condition
+\begin{align}
+  \lleq{gmn}
+  \alpha > 1.
+\end{align}
+
+For simplicity, we choose a uniform prior on $\bmphi$ on a finite box $V$ subject to the constraint \refeq{gmn}
+\begin{align}
+  p(\bmphi) = \mathbf{1}_{\alpha(\bmphi) > 1}(\bmphi) \mathbf{1}_V(\bmphi).
+\end{align}
+The box $V$ is chosen by trial and error to be large enough to contain
+the relevant region of $p(\bmx\cond \bmphi, \npack, \bmnu)$. For
+simple functions $\alpha(\nu \cond \bmphi)$, one can find the minimum
+value of $\alpha$ over the entire range of $\nu$ analytically. If this
+is too complicated, we propose to simplify by requiring $\alpha(\nu_j
+\cond \phi) > 1$ only at each packet frequency $\nu_j$. In the
+evaluation of $p(\bmx \cond \bmphi, \npack, \bmnu)$, one has to loop
+over all packets and compute $\alpha$ and $\beta$ anyways, so the
+constraint imposes essentially no computational overhead.
+
+\subsection{Approximations}
 
@@
 two difficulties: the sum over $N$ and the integral over
-$\bmphi$. Regarding the latter, there is a conjugate prior but
-unfortunately the normalization constant cannot be evaluated in closed
-form. This conjugate prior is thus of no use for us, and we turn to a
-noninformative prior. \fred{Be explicit about prior choice} The
-authors of \cite{moala2013bayesian} present an overview of possible
-choices and conclude that the point estimate given by the mode of the
-posterior is nearly insensitive to the prior choice for 30 samples in
-the iid case. Translated to our problem, this would be 30 packets at
-the same frequency. From TARDIS, we only get packets at different
-frequencies. In the following, let $m$ denote the effective number of
-packets close to $\nu$ where ``close'' is deliberately left
-unspecified.  If $m$ is large, we can well determine $\phi$
-and thus $p(X_j \cond N, \bmphi, \nu)$.
+$\bmphi$. Translated to our problem, this would be 30 packets at the
+\emph{same} frequency. But TARDIS only outputs packets at
+\emph{different} frequencies. In the following, let $m$ denote the
+effective number of packets close to $\nu$ where ``close'' is
+deliberately left unspecified.  If $m$ is large, we can well determine
+$\phi$ and thus $p(X_j \cond N, \bmphi, \nu)$.
 
@@
 The following approximations can be used in different regions of the
-parameter space spanned by $n$ and $m$. %
+parameter space spanned by $n$, $N$, and $m$. %
 \hans{$n$ is TARDIS output ``data'', so not a parameter. From the
@@
 \fred{Update approximations: don't need moments any longer}
-\begin{mnumerate}
-  \item \label{large-m} For large $m$, the posterior $p(\bmphi \cond
-  \npack, \bmx, \bmnu) = p(\bmx\cond \bmphi, \npack, \bmnu) \, p(\bmphi) / p(\bmx \cond \npack, \bmnu)$
-  approaches a multivariate Gaussian. We can then compute the various
-  moments $E_\bmphi[f(x)]$ with $f(x)=1,x,x^2$ by Laplace's method to
-  good accuracy.
-  \item \label{very-large-m} In the limit $m \to \infty$, the
-  covariance of the posterior $p(\bmphi \cond \npack, \bmx, \bmnu)$
-  tends to zero as $1/m$ and we can then fix the parameters $\bmphi$
-  at the posterior mode $\bmphi^*$; i.e., $p(\bmphi \cond \npack,
-  \bmx, \bmnu) \to \delta(\bmphi - \bmphi^*)$.
-  \item \label{large-n} For $n$ large, the cumulative of the Negative
-  Binomial $p(N \cond n)$ can be approximated by a Gaussian, defining
-  a region around $N=n$ with the dominant contribution to the sum over
-  $N$ in \refeq{gmd}. The contribution around $N=0$ is negligible, and
-  we can replace $\sum_n \to \int \rmdx{N}$. The density $p(N \cond n) P(X
-  \cond N,m, \bmx)$ is close to a Gaussian and the integral can be
-  accurately approximated by Laplace's method which requires the
-  Hessian and the mode. We benefit from partial error cancelation
-  because we apply Laplace's method both in the numerator and
-  denominator~\cite{tierney_accurate_1986} as each moment has to be
-  normalized by the evidence $p(\bmx \cond m)$. The former is
-  available in closed form and the latter has to be determined
-  numerically. Fortunately, Newton's method leads to a solution in
-  only a few steps~\cite{minka2002estimating}. \fred{The reference is
-    for max. likelihood of a $\GammaDist(\cdot)$ but I'm sure the
-    performance is similar for max. posterior and the first moments}
-  \item \label{central-limit} The central limit theorem states that for large $N$,
-  $P(X \cond N, \bmphi) \approx \mathcal{N}(X \cond N \,
-  E_\bmphi[X_j], N \, V_\bmphi[X_j])$. For $m$ large enough, we could
-  then plug in the sample-mean and sample-variance estimates and
-  bypass the assumption of a Gamma distribution altogether.
-  \item \label{large-n-large-m} For large $n$ and $m$, $ p(X\cond
-  n,\bmx,m)$ can be approximated by a Gaussian whose mean and variance
-  are given by \refeq{mgo} and \refeq{mgr} in terms of
-  $\bmphi$-expectation values. The latter are integrals with no known
+
+\subsubsection*{Large $\bmn$} \label{large-n}
+
+For large $n$, the cumulative of the Negative Binomial $p(N \cond n)$
+can be approximated by the cumulative of a Gaussian with mean $n$ and
+variance $2 (n - a + 1)$. Intuitively, this variance arises because the
+Poisson variance appears twice, once when observing $n$ and once when
+predicting $N$ given $n$.  The region around $N=n$ yields the dominant
+contribution to the sum over $N$ in \refeq{gmd}. The contribution
+around $N=0$ is then negligible, and we can approximate $\sum_N$ by
+$\int \rmdx{N}$. The density $p(N \cond n) P(X \cond N, \npack, \bmx,
+\bmnu)$ is close to a Gaussian (in the $x$ parametrization) and the
+integral can be accurately approximated by Laplace's method which
+requires the Hessian and the mode.
+  % as each moment has to be
+  % normalized by the evidence $p(\bmx \cond m)$.
+The former is available in closed form and the latter has to be
+determined numerically. Fortunately, Newton's method leads to a
+solution in only a few steps~\cite{minka2002estimating}.
+%
+\fred{The reference is for max. likelihood of a $\GammaDist(\cdot)$
+  but I'm sure the performance is similar for max. posterior and the
+  first moments} We benefit from partial error cancelation when we
+apply Laplace's method both in the numerator and
+denominator~\cite{tierney_accurate_1986}.
+  %
+\fred{Might actually evaluate numerator and denominator at posterior
+  mode to do only one instead of two optimizations;
+  cf.~\cite{lindley_approximate_1980}}
+
+\subsubsection*{Large $\bmN$} \label{large-N}
+
+The central limit theorem states that for large $N$, $P(X \cond N,
+\bmphi, \nu) \approx \mathcal{N}(X \cond N \, E_\bmphi[X_j \cond \nu],
+N \, V_\bmphi[X_j \cond \nu])$. The expectation values have no known
 closed-form solutions but we can easily and accurately approximate
 them with Laplace's method.
-\end{mnumerate}
-To simplify, we assume that $m$ is large. In our application, we may
-have to combine several frequency bins to achieve $m \gtrsim 150$ for
-the purpose of infering $\bmphi$. This is equivalent to assuming that
-the distribution changes only slowly with the frequency, a condition
-that we verified for TARDIS. It is thus justified to use
-approximations \ref{large-m} or even \ref{very-large-m}.
+% This is true independent of our model
+assumption \refeq{gmi} that $X_j \sim \GammaDist(\alpha, \beta)$.
+%
+\fred{This only helps because the Laplace approx. when combined with
+  large $n$ has an analytical result for the mode which saves us
+  numerical optimization. But it requires extra (Laplace
+  approximated?) integrals to compute $E_\bmphi[X_j \cond \nu]$ and
+  $V_\bmphi[X_j \cond \nu]$. I can't get the Hessian analytically for
+  $\npack$ samples. But we could use \texttt{Hesse} from
+  \href{http://seal.web.cern.ch/seal/snapshot/work-packages/mathlibs/minuit/}{Minuit}
+  or similar to approximate the Hessian at the mode by finite
+  differences.}
+
+\subsubsection*{Large $\bmm$}\label{large-m}
+
+By asymptotic normality, the posterior $p(\bmphi \cond \npack, \bmx,
+\bmnu) = p(\bmx\cond \bmphi, \npack, \bmnu) \, p(\bmphi) / p(\bmx
+\cond \npack, \bmnu)$ approaches a multivariate Gaussian. In the limit
+$m \to \infty$, the covariance of the posterior $p(\bmphi \cond
+\npack, \bmx, \bmnu)$ tends to zero as $1/m$ and we can then fix the
+parameters $\bmphi$ at the posterior mode $\bmphi^*$; i.e., $p(\bmphi
+\cond \npack, \bmx, \bmnu) \to \delta(\bmphi - \bmphi^*)$ which removes the integral in \refeq{gmd}.
+
+\subsubsection*{Large $\bmn, \bmN$,  and very large $\bmm$} \label{large-n-large-m}
+
+With many samples observed in a narrow bin, both $n$ and $m$ are large
+and the dominant contribution to \refeq{gmd} arises from large
+$N$. Combining all of the above approximations, we find
+\begin{align}
+  \lleq{gmr} %
+  p(X\cond \nu, \npack,\bmx,\bmnu) &= %
+  \int \rmdx{N} \mathcal{N}(N \cond n, 2 (n -a + 1)) \, \mathcal{N}(X \cond
+  N \, E_{\bmphi^*}[X_j \cond \nu], N \, V_{\bmphi^*}[X_j \cond \nu])
+\end{align}
+where
+\begin{align}
+  \lleq{gms}%
+  E_{\bmphi^*}[X_j \cond \nu] &= \frac{\alpha(\nu \cond
+    \bmphi^*)}{\beta(\nu \cond \bmphi^*)}, & %
+  V_{\bmphi^*}[X_j \cond \nu] = \frac{\alpha(\nu \cond
+    \bmphi^*)}{\beta(\nu \cond \bmphi^*)^2}.
+\end{align}
+Again, this integral does not have a closed-form expression but it is
+unimodal and can be well approximated with Laplace's method. The mode
+of the integrand is the single real root of a cubic polynomial. We
+compute the root as well as the Hessian analytically with mathematica; the results are lengthy and not shown here.
+
+\subsubsection*{Comparison of approximations}
 
@@
 \ref{large-n}--\ref{large-n-large-m}.
-
-\subsection*{Comparison of approximations}
 

### 2016-04-14 hans/bat/Tardis.h
[bat] Rescale integral

@@ -40,6 +40,14 @@ public:
     unsigned GetOrder() const
     { return order; }
 
+    /**
+     * Multiply the likelihood by this factor to avoid overflows.
+     *
+     * @param scale factor on the log(!) scale
+     */
+    void rescale(double scale)
+    { this->scale = scale; }
+
 private:
     Vec ReadData(const std::string& fileName, const std::string& dataSet);
 
@@ -64,6 +72,7 @@ private:
     unsigned order;
     const unsigned npoints;
     const double nuMax, alphaMin, betaMin;
+    double scale;
 };
 // ---------------------------------------------------------
 

### 2016-04-21 hans/bat/Tardis.h
[bat] Predict X for small n

@@ -48,8 +48,30 @@ public:
     void rescale(double scale)
     { this->scale = scale; }
 
+    /**
+     * Set state to update the prior such that minuit gets the correct
+     * target density for prediction and compute the evidence.
+     */
+    void PreparePrediction();
+
+    /**
+     * Predict X for small n at a given nu (=bin center).
+     *
+     * Algorithm: Start computing for N=n, then go to n-1,
+     * n+1. Continue until contributions are negligible. Bear in mind
+     * N>0! Stay on log scale as much as possible?
+     *
+     * @param precision Stop searching if contribution from current point is less than `precision * currentValue`
+     */
+    double PredictSmall(unsigned n, double X, double nu, double precision = 1e-2);
+
 private:
     Vec ReadData(const std::string& fileName, const std::string& dataSet);
+
+    /*
+     * @return for N posited events and n observed events, check if the contribution to res = sum_N P(X|N) is negligible with latest/ res < precision
+*/
+    bool SearchStep(unsigned N, unsigned n, double& res, double precision);
 
     /**
      * Polynomial as a function of nu with coefficients given in range
@@ -73,6 +95,14 @@ private:
     const unsigned npoints;
     const double nuMax, alphaMin, betaMin;
     double scale;
+    double evidence;
+
+    // parameter of prior on Poisson parameter
+    double a;
+
+    ///> prediction
+    double nuPrediction, XPrediction;
+    unsigned NPrediction;
 };
 // ---------------------------------------------------------
 

### 2016-04-21 hans/bat/Tardis.h
[bat] Repeat fit for multiple X
* use clang for better error message
* compute X in a bin
* sort sample

@@ -65,6 +65,11 @@ public:
      */
     double PredictSmall(unsigned n, double X, double nu, double precision = 1e-2);
 
+    /**
+     * Compute sum of X in frequency bin
+     */
+    double SumX(double numin, double numax) const;
+
 private:
     Vec ReadData(const std::string& fileName, const std::string& dataSet);
 

### 2016-04-21 hans/bat/plot.py
[bat] Add plot.py
plot text output from bat analysis

@@ -1,0 +1,33 @@
+import numpy as np
+from matplotlib import pyplot as plt
+from scipy.integrate import simps
+
+
+a = np.loadtxt("out.txt")
+
+# modify data to include the origin
+a = np.vstack((np.zeros(2), a))
+
+X = a.T[0]
+P = a.T[1]
+
+# check normalization of pdf using Simpson's rule
+norm = simps(P, X)
+print("Normalization", norm)
+
+# improve estimates by renormalizing to one
+P /= norm
+
+mean = simps(P * X, X)
+print("Mean", mean)
+print("std. dev", np.sqrt(simps(P * (X - mean)**2, X)))
+
+plt.plot(X, P)
+plt.xlabel("$X$")
+plt.ylabel("$P(X)$")
+
+plt.show()
+
+# Local Variables:
+# compile-command: "python3 plot.py"
+# End:

### 2016-04-21 hans/bat/Tardis.h
[bat] Tune performance for small n prediction

@@ -61,9 +61,10 @@ public:
      * n+1. Continue until contributions are negligible. Bear in mind
      * N>0! Stay on log scale as much as possible?
      *
+     * @param Xmean If positive, use floor(X / Xmean) as starting guess for most likely N
      * @param precision Stop searching if contribution from current point is less than `precision * currentValue`
      */
-    double PredictSmall(unsigned n, double X, double nu, double precision = 1e-2);
+    double PredictSmall(unsigned n, double X, double nu, double Xmean = -1, double precision = 1e-2);
 
     /**
      * Compute sum of X in frequency bin
@@ -74,9 +75,10 @@ private:
     Vec ReadData(const std::string& fileName, const std::string& dataSet);
 
     /*
+     * @param init use as initial position, write back updated results from minimization
      * @return for N posited events and n observed events, check if the contribution to res = sum_N P(X|N) is negligible with latest/ res < precision
 */
-    bool SearchStep(unsigned N, unsigned n, double& res, double precision);
+    bool SearchStep(unsigned N, unsigned n, double& res, double precision, std::vector<double>& init);
 
     /**
      * Polynomial as a function of nu with coefficients given in range

### 2016-04-21 hans/bat/plot.py

@@ -20,12 +20,19 @@ P /= norm
 
 mean = simps(P * X, X)
 print("Mean", mean)
-print("std. dev", np.sqrt(simps(P * (X - mean)**2, X)))
+
+stddev = np.sqrt(simps(P * (X - mean)**2, X))
+print("std. dev", stddev)
+
+mode = X[np.argmax(P)]
+print("Mode", mode)
 
 plt.plot(X, P)
 plt.xlabel("$X$")
 plt.ylabel("$P(X)$")
+plt.title("Mode = %g\nmean x = %g +- %g" % (mode, mean, stddev))
 
+plt.tight_layout()
 plt.show()
 
 # Local Variables:

### 2016-04-21 hans/bat/Tardis.h
(wip) transfer to laptop

@@ -24,7 +24,7 @@ public:
     using Vec = std::vector<double>;
 
     // Constructor
-    Tardis(const std::string& name);
+    Tardis(const std::string& name, const std::string& fileName, unsigned run = 9);
 
     // Destructor
     ~Tardis();
@@ -69,16 +69,16 @@ public:
     /**
      * Compute sum of X in frequency bin
      */
-    double SumX(double numin, double numax) const;
+    std::tuple<unsigned, double> SumX(double numin, double numax) const;
 
 private:
-    Vec ReadData(const std::string& fileName, const std::string& dataSet);
+    Vec ReadData(const std::string& fileName, const std::string& dataSet, unsigned run);
 
     /*
      * @param init use as initial position, write back updated results from minimization
      * @return for N posited events and n observed events, check if the contribution to res = sum_N P(X|N) is negligible with latest/ res < precision
 */
-    bool SearchStep(unsigned N, unsigned n, double& res, double precision, std::vector<double>& init);
+    bool SearchStep(unsigned N, unsigned n, double& res, double precision, Vec& init);
 
     /**
      * Polynomial as a function of nu with coefficients given in range

### 2016-04-21 hans/bat/plot.py

@@ -1,39 +1,47 @@
 import numpy as np
 from matplotlib import pyplot as plt
 from scipy.integrate import simps
+import sys
 
+def plot_estimate(run):
+    a = np.loadtxt("out%s.txt" % run)
 
-a = np.loadtxt("out.txt")
+    # modify data to include the origin
+    a = np.vstack((np.zeros(2), a))
 
-# modify data to include the origin
-a = np.vstack((np.zeros(2), a))
+    X = a.T[0]
+    P = a.T[1]
 
-X = a.T[0]
-P = a.T[1]
+    # check normalization of pdf using Simpson's rule
+    norm = simps(P, X)
+    print("Normalization", norm)
 
-# check normalization of pdf using Simpson's rule
-norm = simps(P, X)
-print("Normalization", norm)
+    # improve estimates by renormalizing to one
+    P /= norm
 
-# improve estimates by renormalizing to one
-P /= norm
+    mean = simps(P * X, X)
+    print("Mean", mean)
 
-mean = simps(P * X, X)
-print("Mean", mean)
+    stddev = np.sqrt(simps(P * (X - mean)**2, X))
+    print("std. dev", stddev)
 
-stddev = np.sqrt(simps(P * (X - mean)**2, X))
-print("std. dev", stddev)
+    mode = X[np.argmax(P)]
+    print("Mode", mode)
 
-mode = X[np.argmax(P)]
-print("Mode", mode)
+    plt.plot(X, P)
+    plt.xlabel("$X$")
+    plt.ylabel("$P(X)$")
+    plt.title("Mode = %g\nmean x = %g +- %g" % (mode, mean, stddev))
 
-plt.plot(X, P)
-plt.xlabel("$X$")
-plt.ylabel("$P(X)$")
-plt.title("Mode = %g\nmean x = %g +- %g" % (mode, mean, stddev))
+    replicas = np.loadtxt("X.out")
+    plt.hist(replicas, bins=15, normed=True)
 
-plt.tight_layout()
-plt.show()
+    plt.tight_layout()
+    # plt.show()
+    plt.savefig("X%s.pdf" % run)
+
+if __name__ == '__main__':
+    plot_estimate(sys.argv[1])
 
 # Local Variables:
 # compile-command: "python3 plot.py"

### 2016-04-27 hans/bat/Tardis.h
(wip)[bat] Try large-n approximation
still yields NaNs or very unsmooth results

@@ -10,10 +10,6 @@
 #include <BAT/BCModel.h>
 
 #include <string>
-#include <valarray>
-
-// This is a Tardis header file.
-// Model source code is located in file Tardis/Tardis.cxx
 
 // ---------------------------------------------------------
 class Tardis : public BCModel
@@ -36,9 +32,6 @@ public:
     virtual double LogAPrioriProbability(const std::vector<double> & parameters);
 
     virtual void CalculateObservables(const std::vector<double>& parameters);
-
-    unsigned GetOrder() const
-    { return order; }
 
     /**
      * Multiply the likelihood by this factor to avoid overflows.
@@ -66,39 +59,80 @@ public:
      */
     double PredictSmall(unsigned n, double X, double nu, double Xmean = -1, double precision = 1e-2);
 
+    double PredictMedium(unsigned n, double X, double nu);
+
+    double PredictVeryLarge(unsigned n, double X, double nu);
+
     /**
      * Compute sum of X in frequency bin
      */
     std::tuple<unsigned, double> SumX(double numin, double numax) const;
 
+    /**
+     * Polynomial as a function of nu with coefficients given in range
+     */
+    template <typename T>
+    static double Polyn(T first, T last, const double& nu)
+    {
+        double res = 0.;
+        double power = 1.;
+        for (; first != last; ++first) {
+            res += (*first) * power;
+            power *= nu;
+        }
+        return res;
+    }
+
+    double alphaNu(const Vec& x, const double& nu)
+    {
+        return Polyn(std::begin(x), std::begin(x) + orderAlpha, nu);
+    }
+
+    double betaNu(const Vec& x, const double& nu)
+    {
+        return Polyn(std::begin(x) + orderAlpha, std::begin(x) + orderAlpha + orderBeta, nu);
+    }
+
 private:
+    enum class Target { Default, Gamma, NBGamma };
+
     Vec ReadData(const std::string& fileName, const std::string& dataSet, unsigned run);
 
-    /*
+    /**
      * @param init use as initial position, write back updated results from minimization
      * @return for N posited events and n observed events, check if the contribution to res = sum_N P(X|N) is negligible with latest/ res < precision
 */
     bool SearchStep(unsigned N, unsigned n, double& res, double precision, Vec& init);
 
-    /**
-     * Polynomial as a function of nu with coefficients given in range
-     */
[…]

### 2016-04-27 hans/bat/plot.py

@@ -7,7 +7,7 @@ def plot_estimate(run):
     a = np.loadtxt("out%s.txt" % run)
 
     # modify data to include the origin
-    a = np.vstack((np.zeros(2), a))
+#     a = np.vstack((np.zeros(2), a))
 
     X = a.T[0]
     P = a.T[1]
@@ -34,11 +34,18 @@ def plot_estimate(run):
     plt.title("Mode = %g\nmean x = %g +- %g" % (mode, mean, stddev))
 
     replicas = np.loadtxt("X.out")
-    plt.hist(replicas, bins=15, normed=True)
+    print("Mean number of packets in bin", replicas.T[0].mean())
+    plt.hist(replicas.T[1], bins=15, normed=True)
 
     plt.tight_layout()
     # plt.show()
     plt.savefig("X%s.pdf" % run)
+
+    plt.clf()
+    plt.hist(replicas.T[0], bins=15, normed=False)
+    plt.xlabel("#packets in bin")
+    plt.tight_layout()
+    plt.savefig("npackets.pdf")
 
 if __name__ == '__main__':
     plot_estimate(sys.argv[1])

### 2016-04-27 hans/bayes-tardis-luminosity-prediction.tex
[draft] add hedgehog plot
[draft] hedgehog caption
[draft] Add missing N
[bayes] run dos2unix again
[draft] Clarify approximations

@@
   have $N \to \infty$ modes but they aren't visible}
-\missingfigure{from hedgehog to single peak \label{fig:hedgehog}}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=0.48\textwidth]{hedgehog}
+  \caption{Left: Prediction $p(L \cond  \nu,  \npack, \bml, \bmnu)$, right: $p(x \cond  \nu,  \npack, \bmx, \bmnu)$. I sum over $N$ in both cases. }
+  \fred{I still need to pretty this up. I used the old binned approach
+    with 50 samples in a bin from which I computed the mean and
+    variance to plug into \refeq{gmr} but I computed the sum over $N$
+    instead of the integral.}
+\label{fig:hedgehog}
+\end{figure}
 
@@
     p(X\cond \nu, \npack,\bmx,\bmnu)
-  = \frac{1}{p(\bmx\cond \bmnu)}
-  \sum_N p(N\cond n)\, \int \rmdx{\bmphi} \GammaDist(X_j \cond \alpha, \beta) % p(X \cond N, \bmphi, \nu)
+  = \frac{1}{p(\bmx\cond \npack, \bmnu)}
+  \sum_N p(N\cond n)\, \int \rmdx{\bmphi} \GammaDist(X \cond N \alpha, \beta) % p(X \cond N, \bmphi, \nu)
   \, p(\bmx\cond \bmphi, \npack, \bmnu)
@@
 can be approximated by the cumulative of a Gaussian with mean $n$ and
-variance $2 (n - a + 1)$. Intuitively, this variance arises because the
-Poisson variance appears twice, once when observing $n$ and once when
-predicting $N$ given $n$.  The region around $N=n$ yields the dominant
-contribution to the sum over $N$ in \refeq{gmd}. The contribution
-around $N=0$ is then negligible, and we can approximate $\sum_N$ by
-$\int \rmdx{N}$. The density $p(N \cond n) P(X \cond N, \npack, \bmx,
+variance $2 (n - a + 1)$. For our purposes, the variance is $ \approx
+2 n$. The intuitive reason for this value is that the Poisson variance
+appears twice, once when observing $n$ and once when predicting $N$
+given $n$.  The region around $N=n$ yields the dominant contribution
+to the sum over $N$ in \refeq{gmd}. The contribution near $N=0$ is
+then negligible, and we can approximate $\sum_N$ by $\int
+\rmdx{N}$. The density $p(N \cond n) P(X \cond N, \npack, \bmx,
 \bmnu)$ is close to a Gaussian (in the $x$ parametrization) and the
@@
 them with Laplace's method.
-% This is true independent of our model
-assumption \refeq{gmi} that $X_j \sim \GammaDist(\alpha, \beta)$.
 %
@@
 
-\subsubsection*{Large $\bmn, \bmN$,  and very large $\bmm$} \label{large-n-large-m}
+\subsubsection*{Large $\bmn, \bmN$,  and large $\bmm$} \label{large-n-large-m}
 

### 2016-05-10 real/draft.tex
[real] Add paper draft on real packets

--- 2016-03-14 hans/bayes-tardis-luminosity-prediction.tex
@@
 %% CHANGE THIS DATE AS YOU MODIFY THE FILE
-\newcommand{\vdate}{160309}
+\newcommand{\vdate}{160510}
 \pagestyle{myheadings}
@@
 \newcommand{\rmdx}[1]{\mbox{d} #1 \,} % differential
+\newcommand{\secDeriv}[1]{\frac{\partial^2}{\partial #1^2}} % second partial derivative
+\newcommand{\secPartial}[2]{\frac{\partial^2}{\partial #1 \, \partial #2}} % second partial derivative
 
@@
 
-\hans{Temporary addition of TOC to make it easier to find stuff. Eventually it should
-disappear again.}
+\section{Introduction} \label{sec:intro}
 
-\tableofcontents
+\section{Theory} \label{sec:theory}
 
-\newpage %
-\hanslong{ %
-  Hans 160228: \textbf{Summary of issues, based on v160218}. \\
+\section{Example} \label{sec:example}
 
-  Before getting into all sorts of detail, I wanted to understand the
-  major changes and developments. To my mind they are:
-  \begin{enumerate}
-  \item Fred discovered that the rescaling $X = L/\Lmax$ may have
-    unintended advantages. TARDIS and general supernova luminosity
-    spectra are highly variable because it consists not only of smooth
-    blackbody spectrum, but also of strong absorption lines (emission
-    also?? see 2014 MNRAS Kerzendorf, Sim et al?). Since $\Lmax$
-    varies exactly like the spectrum itself, all binwise
-    individual-packet distributions are rescaled to the interval
-    $[0,1]$ leaving only a far weaker change of \textit{shape} of the
-    distribution rather than the \textit{scale}.
+\subsection{Hessians} \label{sec:hessian}
 
-  \item Just below Eq. (bsi) and also in (pmda), we had made explicit
-    that we would do all the calculations for each bin separately, so
-    there would be lots of fit parameters. Fred's 160218 paragraph
-    starting with ``In general, the distribution \ldots'' and Eq
-    (pmdb) sees significant reduction in the number of parameters if
-    the above remnant dependence of $p(\bmL\cond N,\bmphi)$ in (pmda)
-    on shape could be captured in a slow-varying function
-    $a(\nu,\bmphi)$.
+\begin{align}
+  \label{eq:hess-posterior}
+   &= - \left(
+    \begin{array}{ccc}
+      \secPartial{\alpha_i}{\alpha_j} & \secPartial{\alpha_i}{\beta_j}\\
+      \dots & \secPartial{\beta_i}{\beta_j}
+    \end{array}
+  \right) \log P(\bmx \cond \npack, \bmnu, \bmphi)
+  \\
+  &= \left(
+    \begin{array}{cc}
+      \sum_n \Psi'(\alpha_n) \nu_n^{i+j} & -\sum_n \frac{\nu_n^{i+j}}{\beta_n}\\
+      \dots & \sum_n \frac{\alpha_n}{\beta_n^2} \nu_n^{i+j}
+    \end{array}
+  \right)
+\end{align}
+with $\alpha_n \equiv \alpha(\nu_n \cond \bmphi), \beta_n \equiv \beta(\nu_n \cond \bmphi)$.
 
-    \item Separately, Fred introduces the idea of combining elementary
-    $b$-bins into compound bins with index $m$ as this improves the
-    statistics.%
-    \fred{This was an intermediate idea, I now removed any reference
-      to it. There should be no more binning when infering $\bmphi$}
-
-    \item The simplification for large $N$ etc permits a Gaussian
-    approximation, hence Laplace, hence the key equation (pmp) may be
-    simplified to some approximation, making use of the moment
-    generating function eg Eq (mgd) unnecessary.%
-    \fred{Yes, so you put in a lot of work to derive the equations and
-      I checked them carefully but I think in the end we will not need
-      them and we should then remove them from the draft}
-
-  \item In principle, the above sounds like a very good development.
-    There are some important caveats:
-    \begin{enumerate}
-      \item I see no detail on how to find $a(\nu,\bmphi)$. The first
-      question would be if there is some numerical indication for its
-      form. This seems to be equivalent to asking what the dependence
-      of (say) the Gamma parameter $\alpha$ on $\nu$ is, and that is
-      of course of great interest anyway.%
-      \fred{Yes, that's equivalent. I just started with a polynomial
-        for $a$ but that's ad-hoc.}
-
-      \item My second comment is that $a(\nu,\bmphi)$ is probably just
-      your normal parametrisation which enters best-fit algorithms.
-      If we are to treat it as a best-fit parametrisation with
-      metaparameters, then that may just bring back all the
-      complications which we thought had disappeared. This would have
-      to be investigated. If on the other hand you also want to treat
-      $a(\nu,\bmphi)$ as some kind of ``constant function'', then one
-      would have to set out how stable this constant function would be
-      against various underlying changes. In other words, if
-      calculation of $a()$ is not part of the general algorithm, how
-      is it fixed? and can it be fixed for all cases?%
-      \fred{I don't see what you mean by ``constant function''. The
-        choice of $a$ and how it depends on $\bmphi$ is part of the
-        overal model.}
-
-      \item I think there may be a problem of combining the two
-      proposed simplifications $L \to X$ and $m$ compound
-      bins. Depending on the variability of the parametrisation as
-      function of $\nu$, the shape of $p(\bmL\cond N,\bmphi)$ for
-      single bins with large $\Lmax$ may differ significantly from the
-      shape for many small bins combined into one using $m$. In other
-      words, rebinning and rescaling may not be compatible.%
-      \fred{No more compound bins, we just do an unbinned fit wrt
-        $\bmphi$, so there is no issue}
-    \end{enumerate}
-
-  \item Once these are cleared up, we can get into the details \ldots\\
-
-  \end{enumerate}
-} %
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Introduction}
-
-TODO: Wolfgang
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Bayesian prediction}
-
-\textbf{HANS NOTE: I have just copied stuff in here which will be
-  written up properly later.}
-\begin{mtemize}
-\item Generic definition of terms: data, parameter, variable
-\item Define likelihood $p(x\cond\theta,\hmod)$
-\item Product rule and sum rule (marginalisation)
-\item Bayes' Theorem
-
-\item Building a model $\hmod$ implies that we hypothesise both a
-  likelihood $p(x\cond\bmtheta,\hmod)$ and a prior
-  $p(\bmtheta\cond\hmod)$.
-
-\item $p(D\cond\theta,\hmod)$ is the \textbf{likelihood} for $\theta$,
-  given the data $D$. Note that, while the likelihood appears to be
-  identical with the sampling distribution $p(D\cond\theta,\hmod)$
-  both in notation and even in its functional form, these two
-  quantities have a different purpose: The likelihood is a function of
-  $\theta$ for fixed data $D$, while the forward probability is a
-  function of $D$ for fixed $\theta$.
-
-\item $p(\theta\cond\hmod)$ is the \textbf{prior probability} for
-  $\theta$, before taking into account the information provided by the
-  data $D$.
-
-\item $p(D\cond\hmod)$ is sometimes called the \textbf{evidence}: it
-  contains just the data, not any particular value of $\theta$. The
-  only way to connect the ``evidence'' to the other probabilities is
-  to expand it in terms of the sum rule, $p(D\cond\hmod) =
-  \sum_{\theta'} p(D\cond\theta',\hmod)\;p(\theta'\cond\hmod)$. In
-  this sense it can be thought of as the normalisation constant which
-  ensures that the posterior is properly normalised.
-
-\item $p(\theta\cond D,\hmod)$ is the \textbf{posterior probability}
-  for $\theta$, after the data $D$ has been taken into account. Given
-  a set of observations $D$, the $p(\theta\cond D,\hmod)$ gives the
-  probability that $\theta$ takes on a particular value.
-
-\item \texttt{COMMENT: the bracketed letters to the left of equations
-    is a temporary aid to facilitate referencing. A change in
-    definition of command} \verb=\lleq= \textbf{will make them revert
-    to ordinary equation labels.}
-
-\item \textbf{Bayes' Theorem}: In terms of these quantities, Bayes'
-  theorem, the central theorem of inference is given by
 \begin{align}
-    \lleq{bth}
-    p(\theta\cond D,\hmod)
-    &= \frac{p(D\cond\theta,\hmod)\;p(\theta\cond\hmod)}
-    {p(D\cond\hmod)}
-    = \frac{p(D\cond\theta,\hmod)\;p(\theta\cond\hmod)}
-    {\int \rmdx{\theta'} p(D\cond\theta',\hmod)\;p(\theta'\cond\hmod)}
+  \label{eq:hess-prediction}
+  & - \left(
+    \begin{array}{ccc}
+      \secPartial{\alpha_i}{\alpha_j} & \secPartial{\alpha_i}{\beta_j} & \secPartial{\alpha_i}{N}\\
+      \dots & \secPartial{\beta_i}{\beta_j} & \secPartial{\beta_i}{N}\\
+      \dots & \dots & \secDeriv{N}
+    \end{array}
+  \right) \log P(X \cond N \alpha, \beta)
+  \\
+  &=
+  \left(
+    \begin{array}{ccc}
+      N^2 \Psi'(N \alpha) \nu^{i+j} & -\frac{N \nu^{i+j}}{\beta} & \left[ -\log \beta +\Psi(N \alpha) + N \alpha \Psi'(N \alpha) - \log X \right] \nu^{i}\\
+      \dots &  \frac{N\alpha}{\beta^2} \nu^{i+j} & -\frac{\alpha}{\beta} \nu^{i} \\
+      \dots & \dots & \alpha^2 \Psi'(N \alpha)
+    \end{array}
+  \right)
 \end{align}
-  where the integral in the denominator is replaced by a sum if
-  $\theta$ takes on discrete values.
-
-\item Note that if the prior is uniform, it is independent of $\theta$
-  and will cancel; we are then left with
-  \begin{equation}
-    \lleq{bthu}
-    p(\theta\cond D,\hmod)
-    = \frac{p(D\cond\theta,\hmod)} {\int \rmdx{\theta'} p(D\cond\theta',\hmod)}
-  \end{equation}
-
-\item \textbf{Prediction using Bayes' Theorem:} By contrast, the
-  Bayesian prediction for $x_{n+1}$ makes full use of the information
-  contained in previous data:
+with $\alpha \equiv \alpha(\nu \cond \bmphi), \beta \equiv \beta(\nu \cond \bmphi)$.
 \begin{align}
-    p(x_{n+1}\cond D,\hmod)
-    &= \int \rmdx{\theta'} p(x_{n+1}\cond\theta',D,\hmod)\;p(\theta'\cond
-    D,\hmod)
-  \end{align}
-  i.e.\ all possible values of $\theta$ are taken into account. Using
-  Bayes' theorem, this can be written also as
-  \begin{align}
-    p(x_{n+1}\cond D,\hmod)
-    &= \frac{\int \rmdx{\theta'} p(x_{n+1}\cond\theta',D,\hmod)\;
-      p(D\cond\theta',\hmod)\;p(\theta'\cond\hmod)}
-    {\int \rmdx{\theta'} p(D\cond\theta',\hmod)\;p(\theta'\cond\hmod)}
-  \end{align}
-  which shows that the prediction for $x_{n+1}$ is a weighted average
-  of $p(x_{n+1}\cond\theta',D,\hmod)$ with the weight given by
-  \begin{align}
-  \frac{p(D\cond\theta',\hmod)\;p(\theta'\cond\hmod)}
-  {\int \rmdx{\theta'} p(D\cond\theta',\hmod)\;p(\theta'\cond\hmod)}
-  &=
-  \frac{p(D,\theta'\cond\hmod)}
-  {\int \rmdx{\theta'} p(D,\theta'\cond\hmod)}
-  \end{align}
-\end{mtemize}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{TARDIS}
-
-\subsection{Intro to those TARDIS aspects relevant to this paper}
-
-TODO: Wolfgang (or in Introduction?)
-
-\subsection{Definitions of relevant quantities}
-\label{sec:dfnr}
-
-\begin{mtemize}
-\item Spectrum of luminosities in channels (or ``bins'')
-  $b=0,1,\ldots,B$ is given by the telescope
-\item In the simplest case, the frequency of channel $b$ is given by
-  $\nu_b = \nu_0 + b\,\delta\nu$, where $\delta\nu$ is
-  the bin width. In the general case, the bins are not of equal
-  size. Let's then consider $\nu_b$ as the left edge of bin $b$,
-  and $\nu_{B+1}$ is the right edge of the last bin; i.e., the
-  maximum frequency.
-  %%%
-  \hans{The 2014 TARDIS paper I have seen uses wavelength $\lambda$,
-    not frequency in their plots. Is Wolfgang's recent data all
-    frequency data? If it is, ie if Wolfgang et al have decided to
-    change to frequency, they should explain whether that will happen
-    in their and our papers also. If they want to stick with
-    wavelength, we should too. I think this is something Wolfgang
-    should decide.
-
-    If they say we should stay with wavelength, then I would suggest
-    replacing $\nu$ by $\lambda$ here and throughout, and of course
-    change the Poisson parameter below to something like $\nu$.}
-
-  \item We treat the results of a single TARDIS run as \textit{simulated data}
-    %%%%%
-    \fred{If in the end we want to compare to actual data from a
-      telescope, I'd prefer to not call the TARDIS output
-      ``data''. Why don't we just call it ``TARDIS output''?} %
-    \hans{I agree, ``data'' is awkward, and it is indeed useful to
-      keep data and $D$ for later. ``Output'' would be ok, except that
-      we should change the symbol $D$ to something else too. Symbol
-      $O$ is not good. For the moment I would suggest $D_S$ (subscript
-      simulation) or $D_T$ since we know that $T$ will later stand for
-      temperature. Another useful term would be ``simdata'', symbol
-      $S$.}
-    %%%%%
-    $S$, i.e.\ fixed results which we do not question but accept as
-    given.  These ``data'' consist of the total number of packets
-    leaving the supernova $\npack$ plus the set of luminosities
-    $\bml$ and frequencies $\bmnu$ of each packet,
-  \begin{align}
-    \lleq{dfc}
-    S = \{ \npack, \bml, \bmnu \} \,,
-  \end{align}
-  where
-  \begin{align}
-    \lleq{dfd}
-    % \bml_b &= (\ell_{b,1},\ell_{b,2},\ldots,\ell_{b,n_b})
-    \dim \bml = \dim \bmnu = \npack \,.
-  \end{align}
-  In terms of $\mathbf{1}_V(x)$---the indicator function requiring
-  $x \in V$---the number of packets in bin $b$ is
-  \begin{align}
-    \lleq{dff}
-    n_b = \sum_{j=1}^{\npack} \mathbf{1}_{[\nu_b,\nu_{b+1}]}( \nu_j )
-  \end{align}
-  and the total luminosity in bin $b$ is
-  \begin{align}
-    \lleq{dfe}
-    \ell_b = \sum_{j=1}^{\npack}\ell_j \mathbf{1}_{[\nu_b,\nu_{b+1}]}( \nu_j )\,,
-  \end{align}
-  with $\npack = \sum_b n_b$ and the total supernova luminosity $\ell
-  = \sum_b \ell_b$.
-
-\end{mtemize}
-
-\subsection{Data as realization of compound process}
-
-While we treat the TARDIS output as fixed ``data'' as discussed above,
-a repeat of the simulation with a different Monte Carlo seed would
-evidently yield different ``data'' $S'$ even when all other simulation
-parameters remain the same. Clearly, $n_b$ is a specific realization
-of a stochastic variable $N_b$ governed by a model pdf $p(N_b)$. %
-\fred{In our general case, $\npack = \sum_b n_b$ is an instance of a
-  random variable because some packets leave the supernova, some
-  don't. I guess in most applications the total number of events would
-  be chosen and not random. We are fortunate that this difference does
-  not matter because we consider bins independently. } %
-\hans{I will come back to this later} %
-Similarly, $\bml_b$ is a specific realization of a collection of
-stochastic variables $\bmL_b =\{L_{b,1},\ldots,L_{b,\smN_b}\}$, where
-the number $N_b$ of such variables is itself stochastic. The relevant
-pdf for $\bmL_b$ therefore always presupposes knowledge of $N_b$, so
-we must work with the conditional probability $p(\bmL_b\cond
-N_b)$. This situation is captured mathematically as $N_b \sim p(N_b),
-\bmL_b \sim p(\bmL_b\cond N_b)$ where $\sim$ means ``is by hypothesis
-distributed in this model according to the pdf'', i.e.\
-\begin{align*}
-  (N_b,\bmL_b) &\ \sim\ p(N_b,\bmL_b\cond \bmphi_b,\lambda_b)
-  = p(\bmL_b\cond N_b,\bmphi_b)\; p(N_b\cond \lambda_b),
-\end{align*}
-where $\bmphi_b$ is the set of parameters of the model pdf which
-govern $\bmL_b$ and $\lambda_b$ is the parameter of the model pdf
-governing $N_b$.  To simulate $(N_b,\bmL_b)$, we clearly first need to
-generate one $N_b$, followed by the generation of $N$ packet
-luminosities $\bmL_b$. This situation is termed a \textit{compound
-  process}.
-
-\subsection{Construction of the model hypothesis}
-
-\begin{mtemize}
-\item We omit the dependence on the physical supernova parameters
-  $\bmtheta$ from the notation below and instead condition on the
-  TARDIS output $\npack, \bml, \bmnu$ directly.
-
-\item In the remainder of this paper, we assume that the packet
-  multiplicities $N_b$ and total luminosities $\bmL_b$ in any given
-  bin $b$ are statistically independent of those in other bins. %
-  \fred{We could motivate this in the usual way as the Poisson limit
-    of the Binomial. The subtlety would then the be the random total
-    number of packets (leaving the supernova).} %
-  \hans{The limit would be from the multinomial with $B$ bins and
-    $\npack$ packtes to product of Poissonians. One could put it in,
-    but in my opinion it's not necessary. $B$ is large, so the
-    constraint plays no role.} %
-  This means that the joint probability over all bins factorises into
-  individual probabilities for each bin,
-  \begin{align}
-    \lleq{bsi}
-    p(N_0,\bmL_0\ldots,N_b,\bmL_b,\ldots,N_\smB,\bmL_\smB
-    \cond \npack, \bml, \bmnu,\bmphi,\bmlambda,\hmod)
-    &= \prod_{b=0}^B p(N_b,\bmL_b\cond n_b, \bml, \bmnu,\bmphi,\bmlambda,\hmod)
-    % p(N_0,\ldots,N_b,\ldots,N_\smB \cond \bmtheta,\bmphi,\hmod)
-    % &= \prod_{b=0}^B p(N_b\cond \bmtheta,\bmphi,\hmod)\\
-    % p(L_0,\ldots,L_b,\ldots,L_\smB \cond N_0,\ldots,N_\smB,\bmtheta,\bmphi,\hmod)
-    % &= \prod_{b=0}^B p(L_b\cond N_b,\bmtheta,\bmphi,\hmod)
+  \label{eq:hess-nb}
+   - \secDeriv{N} \log P(N \cond n) = \Psi'(N+1) - \Psi'(N+n-a+1)
 \end{align}
 
-\item Based on the assumption of bin-bin statistical independence, we
-  shall henceforth drop the bin index and do all our calculations for
-  an individual bin $b$ only, writing for example $L_b \to L, n_b \to
-  n$ etc.  The total spectrum is then obtained by doing any given
-  analysis or calculation for all bins. But note that while the
-  single-bin probability $p(N_b,\bmL_b\cond n_b, \bml,
-  \bmnu,\bmphi,\bmlambda,\hmod)$ depends only on the number of packets
-  $n_b$ in the bin, it depends on the luminosities and frequencies of
-  \emph{all} packets.
-  %
-  \fred{Perhaps it doesn't simplify the reading experience to omit $b$?}
+\section{Conclusion} \label{sec:conclusion}
 
-\item Since we shall not be doing model comparison, we can for the
-  moment drop the $\hmod$ from the probability notation.
-
-\item The \textbf{model} which we construct consists of the above
-  assumptions plus
-  \begin{align*}
-    \text{likelihood } &\ p(\bmL\cond N,\bmphi)\\
-    \text{likelihood } &\ p(N\cond\lambda)\\
-    \text{prior } &\ p(\bmphi)\\
-    \text{prior } &\ p(\lambda)
-  \end{align*}
-  which are assumed to have the same functional form in each bin,
-  while of course the parameter values will change from bin to bin.
-
-\item We choose $N$ to follow a Poisson distribution \fred{this choice
-    linked to limit of Multinomial, see above} with parameter
-  $\lambda$ %
-  \hans{I would change the Poisson parameter to $\nu$, see above comment.}
-  \begin{align}
-    \lleq{pmc}
-    p(N\cond\lambda) &= \frac{e^{-\lambda}\lambda^N}{N!}
-    \qquad N = 0,1,\ldots,\infty,\quad \lambda>0
-  \end{align}
-
-\item We further assume that individual packets are generated
-  independently,
-  \begin{align}
-    \lleq{pmda}
-    p(\bmL\cond N,\bmphi) &= \prod_{j=1}^N p(L_j\cond \bmphi)
-  \end{align}
-
-  \item In general, the distribution of $L_j$ could be a of a
-  different functional form in each bin. For simplicity, we assume it
-  is identical but may depend on parameters that differ from bin to
-  bin. In practice, the distribution of packets does not change
-  abruptly with frequency. We therefore assume that any parameter
-  needed to fully specify $p(L_j)$ is given as a continuous function
-  of frequency and parameters $\bmphi$; i.e.,
-  \begin{align}
-    \lleq{pmdb}
-    p(L_j) = p(L_j \cond \nu, \bmphi)
-  \end{align}
-  For example, assume $p(L_j)$ depends on one parameter $a$. If we fit
-  separately in each bin, we would have $B+1$ fits and $B+1$ copies of
-  $a$. In each fit, there would be only few packets to constrain
-  $a$. To avoid that, we assume $a = a(\nu, \bmphi)$. For concreteness,
-  assume $\bmphi = (a_0, a_1)$ and
-  \begin{align}
-    \lleq{pmdc}
-    a(\nu, \bmphi) = a_0 +a_1 \nu
-  \end{align}
-  This greatly reduces the number of parameters from $B+1$ to 2 and
-  ensures that $a$ is continuous in $\nu$. $a$, or equivalently
-  $\bmphi$, is now determined from a single fit to all packets. The
-  uncertainty on $a$ is at a minimum for the frequency around which
-  the density of packets is highest. \fred{Refer to plot with
-    uncertainty band}
-
-\item The values of $\lambda$ and $\bmphi$ are unknown. It is the task
-  of the analysis below to determine not just the single best values
-  for these parameters, but to take into account every and all
-  possible values.
-
-  \item Priors: metaparameter $a$
-  %
-  \hans{This is a placeholder to remind us that metaparameters need to
-    be discussed at some point. Perhaps not here, but we should
-    remember.}
-\end{mtemize}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Prediction based on individual packets}
-
-\subsection{Predicting the packet multiplicity $N$ in a bin}
-
-\texttt{COMMENT: The derivation below is deliberately slow; the
-  intention being to give the novice an easy introduction into the
-  usage of Bayes before we hit the more complicated calculations
-  further below.}
-
-We first show how to use Bayesian prediction to find the probability
-distribution for $N$, the number of packets in a bin, given knowledge
-of ``data'' in the form of TARDIS multiplicity $n$ for that bin.
-Since $n$ is a specific
-realization of $N$, it results from the same Poisson distribution
-\begin{align}
-  \lleq{pmcb}
-   p(n\cond\lambda) &= \frac{e^{-\lambda}\lambda^n}{n!}
-\end{align}
-with the same value for $\lambda$ as in (\ref{pmc}).  We can hence use
-Bayes' Theorem to find the posterior distribution for $\lambda$ given
-data $n$ and the Poisson hypothesis
-\begin{align}
-  \lleq{pmd}
-  p(\lambda\cond n) %
-  &= \frac{p(n\cond\lambda)\,p(\lambda)} {p(n)}
-  \ =\ \frac{p(n\cond\lambda)\,p(\lambda)} %
-  {\int \rmdx{\lambda} p(n\cond\lambda)\,p(\lambda)}
-\end{align}
-to find the prediction of $N$ given $n$ taking into account all
-possible values of $\lambda$,
-\begin{align}
-  \lleq{pme}
-  p(N\cond n)
-  &= \int \rmdx{\lambda}p(N,\lambda\cond n)
-  \ =\ \int \rmdx{\lambda} p(N\cond\lambda, n)\,p(\lambda\cond n)
-  \ =\ \frac{\int \rmdx{\lambda} p(N\cond\lambda)\,p(n\cond\lambda)\,p(\lambda)}
-  {\int \rmdx{\lambda} p(n\cond\lambda)\,p(\lambda)}
-\end{align}
-where $p(N\cond\lambda,n) = p(N\cond\lambda)$ is independent of $n$.
-Taking the prior (see Appendix)\footnote{For a uniform prior,
-  the case $a{=}0$, we require of course a maximum $\lambda_{\rm
-    max}$. As long as $\lambda_{\rm max}$ is large enough, the
-  integral can effectively be taken to infinity.}
-\begin{align}
-  \lleq{pmf}
-  p(\lambda) = c\lambda^{-a},
-\end{align}
-common choices are $a=0$ (uniform prior) and $a=1/2$ (Jeffreys
-prior). Both choices lead to an improper prior but the posterior
-predictive (see below) is proper.
-Inserting \refeq{pmf} and (\ref{pmcb}) into (\ref{pme}), we obtain for
-the evidence and the posterior predictive
-\begin{align}
-  \lleq{pmg}
-  p(n) &= %
-  \int_0^\infty d\lambda\, p(n\cond\lambda)\,p(\lambda)
-  \ =\
-  c \int_0^\infty d\lambda\, \frac{e^{-\lambda}\,\lambda^{n-a}}{n!}
-  \ =\ c\frac{(n-a)!}{n!}
-  \\
-  \lleq{pmh}
-  p(N\cond n) %
-  &= \frac{n!}{(n-a)!} \int_0^\infty d\lambda\,
-  \frac{e^{-2\lambda} \lambda^{N+n-a}}{N!\, n!}
-  \ =\ \frac{(N+n-a)!}{N!\,(n-a)!\,2^{N+n+1-a}}
-\end{align}
-which we recognise as a Negative Binomial distribution describing the
-probability to observe $N$ successes given $k=n+1-a$ failures in a
-sequence of Bernoulli trials with success probability $\rho=1/2$
-\begin{align}
-  \lleq{pmhb}
-  p(N\cond n) &= \NBD(N\cond k{=}n{+}1{-}a, \rho{=}\tfrac{1}{2})
-  \ =\ \binom{N{+}n{-}a}{N}
-  \left(\frac{1}{2}\right)^N \left(\frac{1}{2}\right)^{n+1-a}
-\end{align}
-We shall use the properties of the Negative Binomial in the
-calculations below. \fred{Check if we still do later} For non-integer $a$,
-the Negative Binomial generalizes immediately to the Pólya
-distribution by replacing $x! \to \Gamma(x+1)$. Note that the
-distribution of $N$ given $n$ is independent of $\bmL$ and $\bml$ by
-construction but depends on $\bmnu$ implicitly through $n$.
-
-\subsection{Predicting the total luminosity $L$ in a bin}
-
-\texttt{COMMENT: We may want to eliminate some intermediate steps in
-  the derivations below in the final paper. Wolfgang: how much detail
-  should be supplied in the paper?}
-
-\fred{One thing is bugging me: To predict the luminosity in a bin, I
-  should do this $ \int_{\nu_b}^{\nu_{b+1}} d{\nu} \, p(L, \nu) =
-  \int_{\nu_b}^{\nu_{b+1}} d{\nu} \, p(L \cond \nu) p(\nu) $. We want
-  to avoid modeling $p(\nu)$ because of the absorption lines, so we
-  assume $p(L \cond \nu)$ approx. indep. of $\nu$ in the bin, so $ \approx p(L \cond
-  \nu ) \int_{\nu_b}^{\nu_{b+1}} d{\nu} \, p( \nu) \equiv p(L \cond
-  \nu) \, p_b$. Since $p_b$ is unknown, we should marginalize over it
-  using our knowledge from Tardis output:
-  $$
-  p(L \cond \mbox{bin } b, \npack, \bmnu, \bml) = \int \rmdx{p_b} p(L \cond \nu) \, p_b \, p(p_b \cond \npack, \bmnu, \bml)
-  $$
-
-  So how can we turn $\int \rmdx{p_b} p_b \to \sum_N$? Where does
-  $\npack$ come into play? If we can answer this, then we have one
-  more hidden assumption clarified in the derivation below that was
-  written for a binned analysis but now we want to be unbinned wrt
-  $\bmnu, \bml$.  }
-
-\subsubsection{Direct probability}
-
-% \fred{We should be more general: let's reserve $n$ for the number of
-%   packets observed in a bin. Suppose this is $n=1$. From that, we can
-%   hardly infer the luminosity distribution. Instead, we group the
-%   packets in larger bins such that there are at least $m \gtrsim 300$
-%   packets, and then assume the distribution doesn't change much with
-%   frequency.
-
-%   Hence we will learn $\bmphi$ from $m$ samples, and
-%   $p(\bmphi\cond N,n,\bml) \to p(\bmphi\cond N,m,\bml)$ and we can
-%   keep $\bml$ with the same meaning. So we replace $n \to m$ below
-%   except in $p(N \cond n$)}
-
-Our goal of predicting $L$ in bin $b$ given the TARDIS output
-$(\npack,\bml, \bmnu)$ is captured in the probability $p(L\cond \nu,
-\npack,\bml, \bmnu)$ with all relevant parameters integrated out using
-the law of total probability.
-
-To simplify the treatment, we assume that the frequency bins are
-narrow or equivalently that $p(L \cond \nu, \npack, \bml, \bmnu)$ does
-not vary within a bin. This implies that all hypothetical samples
-$L_j$ in the bin follow the same distribution.  As the reference
-frequency, we choose the bin center;
-% and avoid integration over the bin
-i.e., for each bin $b$, we set
-\begin{align}
-  \lleq{pmq}
-  p(L\cond \nu, \npack,\bml, \bmnu) = p(L\cond (\nu_b + \nu_{b+1}) / 2, \npack,\bml, \bmnu)  \; \forall \nu \in [\nu_b, \nu_{b+1}] \, .
-\end{align}
-
-We first note from \refeq{pmh} that $N$ does not depend on $\bml$ at
-all and only indirectly depends on $\bmnu$ through $n$, the number of
-packets in the bin; cf. \refeq{dff}. Furthermore, since $L$ is fully
-determined by $(N,\bmL)$,
-\begin{align}
-  \lleq{pmjd}
-  L &= \sum_{j=1}^N L_j
-\end{align}
-is does not depend on $(\npack,\bml,\bmnu)$, i.e.\
-\begin{align}
-  \lleq{pmi}
-  p(L\cond N,\nu,\bmL,\npack,\bml, \bmnu)
-  = p(L\cond N, \bmL)
-  &= \delta(L - \textstyle\sum_{j=1}^N L_j),
-\end{align}
-where the $L_j$ are assumed to occur at $\nu$.
-% For notational simplicity, we now
-% suppress $\nu$ and remember that all probabilities below are
-% conditional on $\nu$.
-Then the prediction is given by
-\begin{align}
-  \lleq{pmj}
-  p(L\cond \nu, \npack,\bml, \bmnu)
-  &= \sum_{N=0}^\infty p(L\cond N,\nu, \npack,\bml, \bmnu)\,p(N\cond \nu, \npack,\bml, \bmnu) \\
-  \lleq{pmjb}
-  &= \sum_N p(N\cond n)\,\int \rmdx{\bmL} p(L\cond N,\bmL,\nu, \npack,\bml, \bmnu)\, p(\bmL\cond N,\nu, \npack,\bml, \bmnu)
-  \\
-  \lleq{pmjc}
-  &= \sum_N p(N\cond n)\,\int \rmdx{\bmL} \delta(L - \textstyle\sum_{j=1}^N L_j)
-  \, p(\bmL\cond N,\nu, \npack,\bml, \bmnu)
-\end{align}
-The dependence on the TARDIS output $(\npack,\bml, \bmnu)$ enters
-through $p(\bmL\cond N,\npack,\bml, \bmnu)$ in terms of the parameters
-$\bmphi$ of $p(\bmL\cond N,\bmphi)$,
-\begin{align}
-  \lleq{pmk}
-  p(\bmL\cond N,\nu, \npack,\bml, \bmnu)
-  &= \int \rmdx{\bmphi} p(\bmL\cond N,\bmphi,\nu, \npack,\bml, \bmnu)\,p(\bmphi\cond N,\nu, \npack,\bml, \bmnu) \\
-  &= \int \rmdx{\bmphi} p(\bmL\cond N,\bmphi, \nu)\,p(\bmphi\cond N,\nu, \npack,\bml, \bmnu).
-\end{align}
-because $\bmL$ is fully determined by $N,\bmphi, \nu$ and hence independent
-of $\npack,\bml, \bmnu$.
-%
-Again using Bayes' Theorem (\ref{bth})
-\begin{align}
-  \lleq{pmkb}
-  p(\bmphi\cond N,\nu, \npack,\bml, \bmnu) &=
-  \frac{p(\bml\cond \bmphi, N,\nu, \npack,\bmnu)\,p(\bmphi\cond N,\nu, \npack,\bmnu)} %
-  {\int \rmdx{\bmphi} p(\bml\cond \bmphi,N,\nu, \npack,\bmnu)\,p(\bmphi\cond N,\nu, \npack,\bmnu)}, %
-  % \ =\ \frac{p(\bml\cond N,\npack,\bmphi)\,p(\bmphi\cond N,n)} %
-  % {p(\bml\cond n)} %
-\end{align}
-using a multiplicity- and frequency-independent prior for $\bmphi$
-\begin{align}
-  \lleq{pml}
-  p(\bmphi\cond N,\nu, \npack,\bmnu) &= p(\bmphi),
-\end{align}
-and recognizing that $\bml$ is independent of $N$ and $\nu$
-\begin{align}
-  \lleq{pmla}
-  p(\bml\cond \bmphi,N,\nu, \npack,\bmnu) = p(\bml\cond \bmphi,\npack,\bmnu)
-\end{align}
-we can write (\ref{pmk}) as
-\begin{align}
-  \lleq{pmn}
-  p(\bmL\cond N,\nu, \npack,\bml, \bmnu)
-  &= \int \rmdx{\bmphi} p(\bmL\cond N,\bmphi)\, %
-  \frac{p(\bml\cond \bmphi, \npack, \bmnu)\,p(\bmphi)} %
-  {\int \rmdx{\bmphi} p(\bml\cond \bmphi,\npack,\bmnu)\,p(\bmphi)} \\
-  \lleq{pmo}
-  &= \frac{1}  {p(\bml\cond \npack, \bmnu)} %
-  \int \rmdx{\bmphi} p(\bmL\cond \bmphi, N)\, %
-  p(\bml\cond \bmphi, \npack,\bmnu)\,p(\bmphi)
-\end{align}
-with $p(\bml\cond \npack, \bmnu) = \int \rmdx{\bmphi} p(\bml\cond \bmphi,
-\npack,\bmnu)\,p(\bmphi)$ in the denominator.  Inserting this into
-(\ref{pmjc}), we obtain the desired result:
-\begin{equation}
-  \lleq{pmp}
-  \boxed{
-  p(L\cond \nu, \npack,\bml, \bmnu)
-  = \frac{1}{p(\bml\cond \npack, \bmnu)}
-  \sum_N p(N\cond n)\,\int \rmdx{\bmL} \delta(L - \textstyle\sum_j L_j)
-  \displaystyle\int \rmdx{\bmphi} p(\bmL\cond N,\bmphi) \, p(\bml\cond \bmphi,\npack,\bmnu)
-  \, p(\bmphi)
-  }
-\end{equation}
-with
-\begin{align}
-  p(\bmL\cond \bmphi, N, \nu) &= \prod_{j=1}^N p(L_j\cond \bmphi, \nu), \\
-  p(\bml\cond \bmphi,\npack,\bmnu) &= \prod_{j=1}^{\npack} p(\ell_j\cond \bmphi, \nu_j)
-\end{align}
-and $p(\cdot \cond \bmphi, \cdot)$ is the same function in both equations.
-In other words: Once we have decided on a prior $p(\bmphi)$ and a
-specific pdf to model the likelihood $p(L_j\cond \bmphi)$, we have a
-prediction for $L$ which takes into account all possible $N$,
-$\lambda$ and $\bmphi$.
-
-\subsubsection{Moment generating function}
-\label{sec:mgff}
-
-\texttt{COMMENT Since the TARDIS ``data'' for $L_j$ is highly
-  asymmetric, the model likelihood\\ $p(L_j\cond \bmphi)$ must also
-  be. Hence we have to determine if a description based on moments is
-  relevant! It could be that our numbers are a very bad description of
-  the posterior.}
-\\
-
-\fred{Dependence on $\nu$ is suppressed in this section}
-
-It may be possible to calculate the prediction in Eq.~(\ref{pmp})
-numerically, but it will be slow due to the multiple $d\bmL$
-integral. A partial solution lies in at least finding simple
-expressions for the moments and cumulants (mean, variance, skewness,
-kurtosis etc) of $p(L\cond n,\ell)$. This can be accomplished by first
-finding the moment generating function (mgf) (see Appendix) which in
-the present case is also the Laplace transform. %
-\fred{On Wikipedia, the MGF is defined as $\int_0^\infty dL\,p(L\cond
-  n,\bml)\,e^{+tL}$. Then there are less minus signs to worry about} %
-\hans{Yes, most books seem to use the plus sign. It makes use of
-  Laplace transform harder, but I am happy to use plus.} %
-For the mgf, the convolution is easily accomplished:
-\begin{align}
-  \lleq{mgc}
-  M(t;L\cond n,\bml) &= M[t\cond p(L|n,\bml)]
-  \ =\ \int_0^\infty dL\,p(L\cond n,\bml)\,e^{-tL} \\
-  &= \frac{1}{p(\bml\cond n)}
-  \sum_N p(N\cond n)\,\int \rmdx{L} e^{-tL} \int \rmdx{\bmL} \delta(L - \textstyle\sum_j L_j)
-  \displaystyle\int \rmdx{\bmphi} p(\bmL\cond N,\bmphi) \, p(\bml\cond n,\bmphi)
-  \, p(\bmphi)\\
-  &= \frac{1}{p(\bml\cond n)}
-  \sum_N p(N\cond n)\,\int \rmdx{\bmphi}
-  \prod_{j=1}^N \left[ \int \rmdx{L_j} p(L_j\cond \bmphi)\,e^{-tL_j}\right]
-   p(\bml\cond n,\bmphi) \, p(\bmphi)
-\end{align}
-The term in the square bracket is the moment generating function of
-$p(L_j\cond \bmphi)$,
-\begin{align}
-  \lleq{mgcb}
-  M(t; L_j\cond \bmphi) &=
-  \int \rmdx{L_j} p(L_j\cond \bmphi)\,e^{-tL_j}
-\end{align}
-and since this is the same for all packets $L_j$, we obtain
-\begin{equation}
-  \lleq{mgd}
-  \boxed{
-  M(t;L\cond n,\bml) %
-  = \frac{1}{p(\bml\cond n)}
-  \sum_N p(N\cond n)\,\int \rmdx{\bmphi}
-  M(t; L_j\cond \bmphi)^N
-   p(\bml\cond n,\bmphi) \, p(\bmphi)
-   }
- \end{equation}
- \checked{}
-As shown in Appendix ***, order-$q$ moments (expectation values of
-$L_j^q$) of $p(L\cond n,\bml)$ can be found by taking successive
-derivatives with respect to $t$,
-\begin{align}
-  \lleq{mge}
-  E[L^q\cond n,\bml]
-  &= \left(\frac{-\partial}{\partial t}\right)^q M(t;L\cond n,\bml)\biggr|_{t=0}
-  \qquad q = 1,2,3\ldots
-\end{align}
-from which we can find the mean, variance, skewness etc of
-$p(L|n,\bml)$ without knowing its form explicitly. To do so, we need
-to find successive derivatives of $\left[ M(t; L_j\cond \bmphi)
-\right]^N$ appearing in (\ref{mgd}). Defining the moments of
-$p(L_j\cond\bmphi)$ as
-\begin{align}
-  \lleq{mgf}
-  \mu_q(\bmphi) &= \int \rmdx{L_j} p(L_j\cond\bmphi)\,L_j^q \qquad q = 1,2,\ldots,
-\end{align}
-we find
-\begin{align}
-  \lleq{mgg}
-  \left(\frac{-\partial}{\partial t}\right) %
-  M(t; L_j\cond \bmphi)^N\biggr|_{t=0}
-  &= N \mu_1(\bmphi)\\
-  \lleq{mgh}
-  \left(\frac{-\partial}{\partial t}\right)^2 %
-  M(t; L_j\cond \bmphi)^N\biggr|_{t=0}
-  &= N^{\underline{2}}\mu_1^2(\bmphi) + N \mu_2(\bmphi) \\
-  \lleq{mgi}
-  \left(\frac{-\partial}{\partial t}\right)^3 %
-  M(t; L_j\cond \bmphi)^N\biggr|_{t=0}
-  &= N^{\underline{3}}\mu_1^3(\bmphi) + 3N^{\underline{2}}\mu_1(\bmphi)\mu_2(\bmphi)
-  + N\mu_3(\bmphi)
-\end{align}
-\checked{}
-where
-\begin{align}
-  \lleq{mgj}
-  N^{\underline q} &= N(N-1)\cdots (N-q+1) = \frac{N!}{(N-q)!}
-\end{align}
-is the $q$th falling power of $N$. Successively inserting these and
-(\ref{mgd}) into (\ref{mge}), we firstly get for the mean of $L$,
-given $(n,\bml)$
-\begin{align}
-  \lleq{mgk}
-  E[L\cond n,\bml]
-  &= \frac{1}{p(\bml\cond n)}
-  \sum_N p(N\cond n)\,N \int \rmdx{\bmphi} \mu_1(\bmphi) \,
-  p(\bml\cond n,\bmphi) \, p(\bmphi).
-\end{align}
-Given that $p(N\cond n)$ is a Negative Binomial pdf, the sum is trivially
-found to be
-\begin{align}
-  \lleq{mgl}
-  \sum_N p(N\cond n)\,N &= n\frac{\tfrac{1}{2}}{1-\tfrac{1}{2}} = n
-\end{align}
-\fred{The first moment under NB is $\frac{\rho k}{1-\rho}$, resulting in $n-a+1$. Here and below you seem to set $a=1$ implicitly. I want to carry the $a$ along.}
-so that
-\begin{align}
-  \lleq{mgm}
-  E[L\cond n,\bml]
-  &= \frac{n}{p(\bml\cond n)}
-  \int \rmdx{\bmphi} \mu_1(\bmphi) \, p(\bml\cond n,\bmphi) \, p(\bmphi)
-\end{align}
-Since the integral is an average of $\mu_1(\bmphi)$ weighted by
-$p(\bmphi\cond n, \bml)$, let us give it a name. Given a function
-$f(\bmphi)$, its ``$\bmphi$-expectation value'' is defined as
-\begin{align}
-  \lleq{mgn}
-  E_\phi[ f ] &= \int \rmdx{\bmphi} f(\bmphi) \, p(\bml\cond n,\bmphi) \, p(\bmphi)
-\end{align}
-in terms of which
-\begin{align}
-  % \lleq{mgpa}
-  % p(\bml\cond n) &= E_\phi[\mu_0],\\
-  %%%%%%%
-  \lleq{mgo}
-  E[L\cond n,\bml]
-  &= n\, \frac{E_\phi[\mu_1]}{p(\bml\cond n)}
-  \ =\ n\,\frac{E_\phi[\mu_1]}{E_\phi[\mu_0]}
-\end{align}
-\fred{$n \to n-a+1$}
-On taking the second derivative and using
-\begin{align}
-  \lleq{mgp}
-  \sum_N p(N\cond n)\,N^2
-  &= \frac{n(n+1)\left(\tfrac{1}{2}\right)^2}{\left(1-\tfrac{1}{2}\right)^2}
-  = n(n+1)
-\end{align}
-\fred{I disagree. The 2nd raw moment is $\frac{k \rho (1 + k
-    \rho)}{(1-\rho)^2}$ (from Mathworld with $\rho=q,1-\rho=p, k=r$
-  and manually derived from the negative binomial MGF from
-  Wikipedia). With $k=n-a+1$ and $\rho=1/2$, I get $(n-a+1)(n-a+3)$}
-the second moment of $p(L\cond n,\bml)$ is
-\begin{align}
-  \lleq{mgq}
-  E[L^2\cond n,\bml]
-  &= \frac{1}{p(\bml\cond n)}
-  \sum_N p(N\cond n)\, \int \rmdx{\bmphi}
-  \left[ N^{\underline{2}}\mu_1^2(\bmphi) + N \mu_2(\bmphi)\right]
-  p(\bml\cond n,\bmphi) \, p(\bmphi) \\
-  &= n^2\,\frac{E_\phi[\mu_1^2]}{E_\phi[1]}
-  + n \frac{E_\phi[\mu_2]}{E_\phi[1]}
-\end{align}
-\fred{I have
- $$
-  \frac{n-a+1}{E_\phi[1]} \left( (n-a+2)E_\phi[\mu_1^2] + E_\phi[\mu_2]\right)
-$$
-}
-so that the variance is
-\begin{align}
-  \lleq{mgr}
-  \var(L\cond n,\bml)
-  &= n^2\, \frac{E_\phi[\mu_1^2]\,E_\phi[1] -  E_\phi[\mu_1]^2}
-  {E_\phi[1]^2}
-  + n\, \frac{E_\phi[\mu_2]}{E_\phi[1]}
-\end{align}
-It is tempting to argue that the $n^2$ term is much larger than the
-$n$-order term. This depends, however, on the sizes of the respective
-expectation values: no a priori assumptions can be made here.
-\fred{I have the more complicated
-  $$
-    \var(L\cond n,\bml) = \frac{n-a+1}{E_\phi[1]^2} \left( (n-a+2)E_\phi[\mu_1^2] E_\phi[1] -(n-a+1) E_\phi[\mu_1]^2 + E_\phi[\mu_2] E_\phi[1]\right)
-  $$
-}
-
-\subsection{Third-order: skewness}
-
-TODO Hans
-
-This may be necessary because $p(L_j\cond\bmphi)$ is highly
-asymmetric, while the pdf of the sum $L$ may be less so.
-
-\fred{I agree but what would we do if we knew the skewness?
-For the fit, we need to evaluate the likelihood, so we'd have to build a function my moment matching.
-Of course with mean and variance alone, we'd use the Gaussian.
-But this just reduces to the central limit theorem.}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Modeling the TARDIS packet distribution}
-
-TODO: All (numerics and analytics)
-
-\begin{mtemize}
-\item Find the most suitable likelihood for $p(L_j\cond \bmphi)$
-
-\item Determine how far we can integrate analytically over the
-parameters
-
-\item How stable are the results to perturbations?
-
-\item \texttt{COMMENT Since $p(L_j\cond \bmphi)$ is highly asymmetric,
-    we have to determine if a description based on moments is
-    relevant! It could be that our numbers are a very bad description
-    of the posterior.}
-
-\end{mtemize}
-
-\fred{Yes but at least if a rather large number of events is in the
-  bin, this will be good enough}
-
-\subsection{Results for Beta distribution likelihood}
-
-The Beta probability density function (pdf) is
-\begin{align}
-  \lleq{btc}
-  p(x\cond a,b) &= \frac{x^{a-1}(1-x)^{b-1}}{B(a,b)} \qquad
-  0 < x < 1, \quad a,b > 0.
-\end{align}
-where $\bmphi = (a,b)$ are the parameters and
-\begin{align}
-  \lleq{btd}
-  B(a,b) &= \frac{\Gamma(a)\,\Gamma(b)}{\Gamma(a+b)}
-\end{align}
-is the beta function.
-%
-The form of the TARDIS single-packet distribution suggests that this
-is appropriate; the strong asymmetry in the TARDIS distribution leads
-us to expect most likely parameter values of $b \ll a$ so that the mode
-is close to $x=1$.
-%
-The variable $x$ is related to packet luminosities simply by
-$x = L_j/L_{\rm max}$ where $L_{\rm max}$ is the largest luminosity
-in the packets i.e.\ $0 < L_j < L_{\rm max}$.
-%
-(Remember we are considering a single bin $b$ only, and only a single
-TARDIS run.)
-%
-Given that we want to find the $\phi$-moments of Section \ref{sec:mgff}
-in terms of TARDIS simulation ``data'', we define in this section
-\begin{align}
-  x_j &= \ell_j/\ell_{\rm max}
-\end{align}
-while $x$ without subscript is used for the generic moment
-calculations. The same quantities could, of course, be used in terms
-of the generic variable $x_j = L_j/L_{\rm max}$. In terms of the
-original ``data'' packets, the beta distribution reads
-\begin{align}
-  \lleq{bte}
-  p(\bml\cond \ell_{\rm max},n,a,b)
-  &= \prod_{j=1}^n p(\ell_j\cond \ell_{\rm max},a,b) \\
-  \lleq{btf}
-  p(\ell_j\cond \ell_{\rm max},a,b)
-  &= \frac{1}{\ell_{\rm max}\,B(a,b)}
-  \,\left(\frac{\ell_j}{\ell_{\rm max}}\right)^{a-1}
-    \left(1-\frac{\ell_j}{\ell_{\rm max}}\right)^{b-1}
-  \ =\ \frac{1}{B(a,b)}
-  \frac{\ell_j^{a-1}\,(\ell_{\rm max}-\ell_j)^{b-1}}
-  {\ell_{\rm max}^{a+b-1}}
-\end{align}
-\checked{}
-equivalent to the form (\ref{btc}) with $x\to x_j$.
-% \begin{align}
-%   p(x_j\cond a,b) &= \frac{1}{B(a,b)} x_j^{a-1} (1-x_j)^{b-1}
-% \end{align}
-NOTE: If $\ell_{\rm max}$ changes significantly from TARDIS run to
-run, then the probability calculation may have to be changed from
-$p(\bml\cond \ell_{\rm max},n,a,b)$ to\\
-$p(\bml\cond \ell_{\rm max},n,a,b)\,p(\ell_{\rm max}\cond
-n,a,b)$. However we are supposed to deal with ONE TARDIS run, so that
-$\ell_{\rm max}$ is known and fixed.  \fred{Rather: we approximate
-  $p(\ell_{\rm max} \cond n, a,b)$ by $\delta(\max \ell_{b,j})$. This
-  is OK is $\sum_b n_b$ is large. For numerical stability, we want to
-  ensure that $x<1$, so we take the largest $\ell_{b,j}$ and multiply
-  by $1 +\varepsilon, \varepsilon = 10^{-7}$}
-
-The moments of the beta distibution are simple,
-\begin{align}
-  \lleq{btg}
-  \mu_q(a,b) &= \int \rmdx{x} x^q\,p(x\cond a,b)
-  = \frac{(a)_q}{(a+b)_q}
-  = \frac{a(a+1)\cdots(a+q-1)} {(a+b)(a+b+1)\cdots(a+b+q-1)}
-  \qquad q = 0,1,2,\ldots
-\end{align}
-with $(a)_b$ the Pochhammer symbol.
-%
-\\
-\fred{Formula ok but to be consistent with the notation used above, we should either use the rising factorial $a^{\overline{b}}$ here or $(a)^b$ above}
-
-\textbf{Calculations for a uniform discrete prior}
-
-The prior is as yet unspecified. Since there is no known conjugate
-prior for the beta distribution,\fred{yes it is known. But it not
-  useful because the normalization can't be calculated analytically;
-  see
-  http://mathoverflow.net/questions/20399/conjugate-prior-of-the-dirichlet-distribution}
-we resort to a simpler prior in terms of discrete values for $a$ and
-$b$ to simplify the expressions. In principle, these can be any
-rational numbers, but because as stated the value of $b$ is expected
-to be around 30--60, we let $a$ and $b$ be integers only, $a,b, =
-1,2,3,\ldots$. \fred{I don't understand this argument} Furthermore,
-since Eqs.~(\ref{mgo}) and (\ref{mgr}) always involve \textbf{ratios}
-of $\phi$-expectation values, we can use a uniform improper prior,
-\begin{align}
-  p(a,b) &= C\qquad a,b = 1,2,\ldots,\infty
-\end{align}
-with $C$ the improper constant which will cancel in the ratio.
-% With
-% this prior, the $\phi$-expectation values become
-% \begin{align}
-%   E_\phi[\mu_q] &= C \sum_{a=1}^\infty \sum_{b=1}^\infty
-%   \mu_q(a,b)\,\prod_j \frac{x_j^{a-1}(1-x_j)^{b-1}}{B(a,b)}
-% \end{align}
-We now calculate $\phi$-expectation values for this case, namely
-\begin{align}
-  \lleq{bthi}
-  E_\phi[\mu_q(a,b)] &= \sum_{a,b}\mu_q(a,b)\,p(\bml\cond \ell_{\rm max}, n,a,b)
-  \,p(a,b)
-  \\
-  % = C \sum_{a,b=1}^\infty \mu_q(a,b)\,p(\bml\cond \ell_{\rm max}, n,a,b)\\
-  \lleq{bti}
-  &= C\sum_{a,b} \frac{(a)_q}{(a+b)_q}
-  \frac{1}{B(a,b)} \prod_{j=1}^n \left(x_j^{a-1}(1-x_j)^{b-1}\right)
-\end{align}
-\checked{}
-The product can be rewritten as
-\begin{align}
-  \lleq{btj}
-  \prod_{j=1}^n \left(x_j^{a-1}(1-x_j)^{b-1}\right)
-  &= \left(\prod_j x_j\right)^{a-1}\left(\prod_j(1-x_j)\right)^{b-1}
-\end{align}
-\fred{This equation can be skipped}
-and defining
-\begin{align}
-  \lleq{btk}
-  y &= \prod_{j=1}^n x_j \qquad z = \prod_{j=1}^n(1-x_j)
-\end{align}
-\checked{}
-which are data quantities independent of any model or model
-parameters, we get
-\begin{align}
-  \lleq{btl}
-  E_\phi[\mu_q(a,b)] &= C\sum_{a,b} \frac{(a)_q}{(a+b)_q}
-  \frac{y^{a-1}z^{b-1}}{B(a,b)}
-\end{align}
-\checked{}
-With the help of mathematica we find
-\begin{align}
-  \lleq{btm}
-  E_\phi[\mu_0] &= C\sum_{a,b=1}^\infty \frac{y^{a-1}z^{b-1}}{B(a,b)}
-  = \frac{C}{(1 - y - z)^2}
-  \\
-  E_\phi[\mu_1] &= C\sum_{a,b} \frac{a}{(a+b)}
-  \frac{y^{a-1}z^{b-1}}{B(a,b)}
-  \nonumber\\
-  \lleq{btn}
-  &= \frac{C}{(y + z)^3}
-  \left[
-  \frac{(y + z) [2 y^2 + (z-y)(z-1)]}
-  {(1 - y - z)^2 }
-   + (z-y) \ln(1 - y - z)
-   \right]
-%   \\
-%   E_\phi[\mu_2(a,b)] &= C\sum_{a,b} \frac{a(a+1)}{(a+b)(a+b+1)}
-%   \frac{(1-y)^{a-1}(1-z)^{b-1}}{B(a,b)}
-%   \\
-%   E_\phi[\mu_1^2(a,b)] &= C\sum_{a,b} \left(\frac{a}{a+b}\right)^2
-%   \frac{(1-y)^{a-1}(1-z)^{b-1}}{B(a,b)}
- \end{align}
- \checked{}
-Taking ratios, we obtain
-\begin{align}
-  \lleq{bto}
-  \frac{E_\phi[\mu_1(a,b)]} {E_\phi[\mu_0(a,b)]}
-  &= \frac{(y + z) [2 y^2 +(z- y) (1 - z)] + (z-y) (1 - y - z)^2 \ln(1 - y - z)}
-  {(y + z)^3}
-\end{align}
-and so
-\begin{align}
-  \lleq{btp}
-  E[L\cond n,\bml] &= n \left[
-  \frac{(y + z) [2 y^2 +(z- y) (1 - z)] + (z-y) (1 - y - z)^2 \ln(1 - y - z)}
-  {(y + z)^3}\right]
-\end{align}
-is completely determined by the data $(n,\bml)$.  Similarly
-\begin{align}
-  E_\phi[\mu_2] &= C\sum_{a,b} \frac{a(a+1)}{(a+b)(a+b+1)}
-  \frac{y^{a-1}z^{b-1}}{B(a,b)}
-  \nonumber\\
-  \lleq{btq} &= \frac{-1}{(1-y-z)^2 (y+z)^5} %
-  \nonumber\\
-  &\times\Bigl\{ 2 \Bigl[2 y^2(z+1) + yz(z-8) - (z-2) z^2\Bigr]
-  (y+z-1)^2 \ln(-y-z+1)
-  \nonumber\\
-  &+ (y+z) \Bigl[
-    y^4 - 2 y^3 (5z+3) + y^2[(22-19 z) z+4] - 4 yz (z-4) (z-1) + 4z^2 (z-1)^2
-           \Bigr] \Bigr\}
-  % \nonumber\\
-  % &= -(1/((y+z-1)^2 (y+z)^5)) %
-  % \nonumber\\
-  % &\times \Bigl\{
-  %   -4 y^3 + 6 y^4 - y^5 + 12 y^2 z - 16 y^3 z + 9 y^4 z + 12 y z^2 -
-  %   42 y^2 z^2 + 29 y^3 z^2 - 4 z^3 - 12 y z^3 + 23 y^2 z^3 + 8 z^4 -
-  %   4 z^5 \nonumber\\
-  %   &+ \ln[1 - y - z] \Bigl[
-  %   -4 y^2 + 8 y^3 - 4 y^4 + 16 y z - 28 y^2 z + 16 y^3 z -
-  %     4 y^4 z - 4 z^2 - 26 y z^2 + 36 y^2 z^2 - 10 y^3 z^2 + 10 z^3 +
-  %     8 y z^3 - 6 y^2 z^3 - 8 z^4 + 2 y z^4 + 2 z^5
-  %     \Bigr]
-  %   \Bigr\}
-\end{align}
-and
-\begin{align}
-  E_\phi[\mu_1^2] &= C\sum_{a,b} \frac{a^2}{(a+b)^2}
-  \frac{y^{a-1}z^{b-1}}{B(a,b)}
-  \nonumber\\
-  \lleq{btr} %
-  &= \frac{1}{y (y - z)^3}
-  \Bigl\{
-  (y - z) (-y + z + 2 y \ln[y]) + y (-2 + y + z)
-  (\mathrm{PolyLog}[2, 1 - y] - \mathrm{PolyLog}[2, 1 - z])
-  \Bigr\}
-\end{align}
-These have to be combined with the above to find the variance of
-Eq.~(\ref{mgr}). The hope is that the complicated mathematical
-expressions simplify. Either way, the variance is also a closed
-solution in terms of $n$, $y$ and $z$.
-
-\subsection{Results for the Gamma-distribution likelihood}
-
-The Gamma distribution  with shape parameter $\alpha$ and rate
-parameter $\beta$ is defined as
-\begin{align}
-  \lleq{gma}
-  \GammaDist(x | \alpha, \beta) = \frac{\beta^\alpha}{\Gamma(\alpha)} x^{\alpha-1} e^{-\beta x} .
-\end{align}
-The support of $\GammaDist(x \cond \alpha, \beta)$ is $[0, \infty]$
-whereas the luminosity of real packets is bounded from above and has a
-peak close to but not at the boundary. \fred{refer to plot of real
-  luminosity packets in some frequency bin} In order to describe the
-luminosity distribution of real packets through $\GammaDist(\alpha,
-\beta)$, we have to transform the luminosities. To this end, we first
-take the largest luminosity of all TARDIS packets across the entire
-frequency range and increase it slightly to avoid numerical
-difficulties at the end point
-\begin{align}
-  \lleq{gme}
-  \lmax \equiv (1+\epsilon) \times \max_j \ell_j,
-\end{align}
-where $ \epsilon=10^{-6}$.  We assume the total number of samples
-across all bins is large, hence the effect of considering $\lmax$ a
-constant independent of the data is negligible. We then transform the
-samples $\bml \to \bmx$ as
-\begin{align}
-  \lleq{gmf}
-  x_j \equiv 1- \frac{\ell_j}{\lmax} > 0
-\end{align}
-and similarly for $L \to X$, then we infer $P(X \cond \nu,  \npack, \bmx,
-\bmnu)$. Only in the end, we transform back to a probability density
-in $L$ via
-\begin{align}
-  \lleq{gmg}
-  p(L  \cond \nu,  \npack, \bml, \bmnu) = p(X  \cond \nu,  \npack, \bmx, \bmnu) \frac{1}{\lmax}
-\end{align}
-\hans{So far, $m$ appears only in your todo comment above. I would
-  suggest a short discussion of this already in Section
-  \ref{sec:dfnr}. We could then simply rewrite all equations below in
-  terms of $m$ rather than $b$.} %
-\fred{I removed $m$ whereever I saw it} We observe good agreement
-between the best fit of a Gamma distribution to TARDIS packet
-luminosities; an example is shown shown in \reffig{gamma}.  While not
-obvious at first, a major numerical simplification arises because of
-the $L \to X$ transformation. Rewriting the master
-equation~(\ref{pmp}) formally as
-\begin{align}
-  \lleq{gmh}
-  p(L  \cond \nu,  \npack, \bml, \bmnu) = \sum_{N} p(N \cond \nu,  n) p(L \cond N,
-\nu, \npack, \bml, \bmnu)
-\end{align}
-we see that the right-hand side is a mixture density. In our
-application, the distribution of a single luminosity, $p(L \cond N=1,
-\nu, \bml, \bmnu)$ is narrow in the sense that the standard deviation
-relative to the mean is at the percent level. This implies that $p(L
-\cond \nu, \npack, \bml, \bmnu)$ has one mode from each $N$ and all
-modes are clearly separated. The transformation $L \to x$ effects a
-strong overlap between the distributions $p(X \cond N, \nu, \npack, \bmx, \bmnu)$
-with similar $N$, so $p(X \cond \nu, \npack, \bmx, \bmnu)$ can be well
-approximated by a unimodal distribution; see \reffig{hedgehog}.
-\fred{I'm not 100 \% sure but $p(X \cond \nu, \npack, \bmx, \bmnu)$ should still
-  have $N \to \infty$ modes but they aren't visible}
-\missingfigure{from hedgehog to single peak \label{fig:hedgehog}}
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width=0.48\textwidth]{gamma}
-  \fred{Plot with more samples to see dip at $x=0$.} %
-  \hans{I assume the KDE fit will disappear from the paper later as it
-    overshoots the zero point.}
-  \caption{Gamma distribution fit. }
-  \label{fig:gamma}
-\end{figure}
-
-For our purposes, the most useful property of the Gamma
-distribution is infinite divisibility; i.e., the sum of $N$ Gamma
-variates is again a Gamma variate
-\begin{align}
-  \lleq{gmb}
-  \boxed{
-    X_j \sim \GammaDist(\alpha, \beta) \Rightarrow X \equiv \sum_{j=1}^N X_j \sim \GammaDist(N \alpha, \beta).
-    }
-\end{align}
-With
-\begin{align}
-  \lleq{gmi}
-  \alpha, \beta &= \alpha(\nu \cond \bmphi), \beta(\nu \cond \bmphi),\\
-  p(X_j \cond \bmphi, \nu) &=
-  \GammaDist(X_j \cond \alpha, \beta),\\
-  p(\bmX \cond N, \bmphi, \nu) &= \prod_{j=1}^N p(X_j \cond \bmphi, \nu)
-\end{align}
- \refeq{gmb} is equivalent to
-\begin{align}
-  \lleq{gmc}
- p(X \cond N, \bmphi, \nu) &= \int \rmdx{\bmX} \delta(X - \textstyle\sum_j X_j)
-  \,p(\bmX\cond N,\bmphi, \nu) = \GammaDist(X \cond N \alpha, \beta)
-\end{align}
-and this implies that the master equation~(\ref{pmp}) considerably simplifies as
-\begin{align}
-  \lleq{gmd}
-    p(X\cond \nu, \npack,\bmx,\bmnu)
-  = \frac{1}{p(\bmx\cond \bmnu)}
-  \sum_N p(N\cond n)\, \int d\bmphi\, p(X \cond N, \bmphi, \nu)\, p(\bmx\cond \bmphi, \npack, \bmnu)
-  \, p(\bmphi).
-\end{align}
-
-\subsection*{Approximations}
-
-To explicitly evaluate \refeq{gmd}, we have to overcome the remaining
-two difficulties: the sum over $N$ and the integral over
-$\bmphi$. Regarding the latter, there is a conjugate prior but
-unfortunately the normalization constant cannot be evaluated in closed
-form. This conjugate prior is thus of no use for us, and we turn to a
-noninformative prior. \fred{Be explicit about prior choice} The
-authors of \cite{moala2013bayesian} present an overview of possible
-choices and conclude that the point estimate given by the mode of the
-posterior is nearly insensitive to the prior choice for 30 samples in
-the iid case. Translated to our problem, this would be 30 packets at
-the same frequency. From TARDIS, we only get packets at different
-frequencies. In the following, let $m$ denote the effective number of
-packets close to $\nu$ where ``close'' is deliberately left
-unspecified.  If $m$ is large, we can well determine $\phi$
-and thus $p(X_j \cond N, \bmphi, \nu)$.
-
-% We require in addition that the posterior is narrow enough so we
-% group samples in bins to have at least $m=150$ samples in each
-% bin. This implicitly assumes that the luminosity distribution is
-% constant in the frequency range defined by $m$.
-
-% Note that the bins defined by the experiment in general do not overlap
-% with the bins used to infer the luminosity distribution; i.e., $n \ne
-% m$.
-
-The following approximations can be used in different regions of the
-parameter space spanned by $n$ and $m$. %
-\hans{$n$ is TARDIS output ``data'', so not a parameter. From the
-  above paragraph, $m$ seems to be a parameter governing rebinning to
-  nonuniform bins. I would have no idea how to handle $m$ formally in
-  a Bayesian way. If $m{=}150$ is the result of optimising the fitting
-  stability, then presumably it will be kept constant. In that case,
-  it would be treated like a constant rather than a parameter in the
-  usual sense.}
-\fred{Yes, $n$ is fixed from TARDIS output. $m$ is defined only for this paragraph}
-\fred{Update approximations: don't need moments any longer}
-\begin{mnumerate}
-  \item \label{large-m} For large $m$, the posterior $p(\bmphi \cond
-  \npack, \bmx, \bmnu) = p(\bmx\cond \bmphi, \npack, \bmnu) \, p(\bmphi) / p(\bmx \cond \npack, \bmnu)$
-  approaches a multivariate Gaussian. We can then compute the various
-  moments $E_\bmphi[f(x)]$ with $f(x)=1,x,x^2$ by Laplace's method to
-  good accuracy.
-  \item \label{very-large-m} In the limit $m \to \infty$, the
-  covariance of the posterior $p(\bmphi \cond \npack, \bmx, \bmnu)$
-  tends to zero as $1/m$ and we can then fix the parameters $\bmphi$
-  at the posterior mode $\bmphi^*$; i.e., $p(\bmphi \cond \npack,
-  \bmx, \bmnu) \to \delta(\bmphi - \bmphi^*)$.
-  \item \label{large-n} For $n$ large, the cumulative of the Negative
-  Binomial $p(N \cond n)$ can be approximated by a Gaussian, defining
-  a region around $N=n$ with the dominant contribution to the sum over
-  $N$ in \refeq{gmd}. The contribution around $N=0$ is negligible, and
-  we can replace $\sum_n \to \int \rmdx{N}$. The density $p(N \cond n) P(X
-  \cond N,m, \bmx)$ is close to a Gaussian and the integral can be
-  accurately approximated by Laplace's method which requires the
-  Hessian and the mode. We benefit from partial error cancelation
-  because we apply Laplace's method both in the numerator and
-  denominator~\cite{tierney_accurate_1986} as each moment has to be
-  normalized by the evidence $p(\bmx \cond m)$. The former is
-  available in closed form and the latter has to be determined
-  numerically. Fortunately, Newton's method leads to a solution in
-  only a few steps~\cite{minka2002estimating}. \fred{The reference is
-    for max. likelihood of a $\GammaDist(\cdot)$ but I'm sure the
-    performance is similar for max. posterior and the first moments}
-  \item \label{central-limit} The central limit theorem states that for large $N$,
-  $P(X \cond N, \bmphi) \approx \mathcal{N}(X \cond N \,
-  E_\bmphi[X_j], N \, V_\bmphi[X_j])$. For $m$ large enough, we could
-  then plug in the sample-mean and sample-variance estimates and
-  bypass the assumption of a Gamma distribution altogether.
-  \item \label{large-n-large-m} For large $n$ and $m$, $ p(X\cond
-  n,\bmx,m)$ can be approximated by a Gaussian whose mean and variance
-  are given by \refeq{mgo} and \refeq{mgr} in terms of
-  $\bmphi$-expectation values. The latter are integrals with no known
-  closed-form solutions but we can easily and accurately approximate
-  them with Laplace's method.
-\end{mnumerate}
-To simplify, we assume that $m$ is large. In our application, we may
-have to combine several frequency bins to achieve $m \gtrsim 150$ for
-the purpose of infering $\bmphi$. This is equivalent to assuming that
-the distribution changes only slowly with the frequency, a condition
-that we verified for TARDIS. It is thus justified to use
-approximations \ref{large-m} or even \ref{very-large-m}.
-
-Numerically, the most challenging case is small $n$ because the sum
-over $N$ in \refeq{gmd} has to be computed explicitly starting at
-$N=1$ until contributions become negligible. Since $p(N \cond n)$
-falls off quickly, it is sufficient in our application to do this
-until $n \lesssim 20-30$. Beyond that, we use approximations
-\ref{large-n}--\ref{large-n-large-m}.
-
-\subsection*{Comparison of approximations}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Results}
-
-In \reffig{beta} is a fit of the Beta distribution to 3000
-samples from a TARDIS run with frequencies in the range $[1.367,
-1.435] \times 10^{38}$ Hz. A characteristic feature seen in many such
-similar fits is that the Beta distribution does not give a perfect
-fit: it undershoots around the mode and overshoots around
-$x=0.96$. The reason may be that the Beta itself is not flexible
-enough to describe the data, or the distribution is simply not
-constant in frequency, so we should choose smaller bins or less
-samples. We don't expect 3000 samples in every bin, so what is the
-minimum number of samples needed to work robustly? The first numerical
-challenge is how to normalize the luminosities, I take the maximum
-across all samples (not just one bin) and multiply by $10^{-6}$ to
-avoid $x=1$ for any sample. For too few samples, the fit returns a pdf
-that peaks at the boundary $x=1$, this is not what we want. From trial
-and error, 300 samples is a solid minimum to get a robust fit if
-parameters are estimated via maximum likelihood. The method-of-moment
-estimator has the advantage of not requiring an optimization algorithm
-but for some bins it gives a solution peaking at $x=1$ even for 800
-samples so it's not robust.
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width=0.48\textwidth]{beta-300}
-  \includegraphics[width=0.48\textwidth]{beta-3000}
-  \caption{Beta distribution fit. The samples are binned in the
-    histogram. The blue solid curve is a kernel-density estimate, the
-    red dashed curve is the Beta distribution. left: 300 samples,
-    right: 3000 samples}
-  \label{fig:beta}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Discussion and Conclusions}
-
-TODO: All
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section*{Acknowledgements}
-
-This work was supported in part by the DFG cluster of excellence
-``Origin and Structure of the Universe'' (www.universe-cluster.de) and
-the National Research Foundation of South Africa, etc etc.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Appendix: Relevant distributions}
-
-TODO: Hans
-
-\subsection{Product rule and sum rule}
-
-\subsection{Poisson}
-
-\subsection{Negative Binomial}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Appendix: Generating functions}
-
-TODO: Hans
-$\pi(\alpha, \beta) \propto \left( \frac{\Gamma(\alpha+\beta)}{\Gamma(\alpha) \Gamma(\beta)} \right)^{\lambda +1} (x x_0)^{\alpha-1} \left(y_0 (1-x) \right)^{\beta-1}$
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% \section{}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\bibliographystyle{plain}
-\bibliography{references}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \end{document}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
 % Local Variables:
-% compile-command:"rubber --pdf -W refs -S bayes-tardis-luminosity-prediction"
+% compile-command:"rubber --pdf -W refs -S draft"
 % End:

### 2016-05-10 real/code/Tardis.h
[real] Add code/
First version essentially copied from bat project

@@ -1,0 +1,158 @@
+#pragma once
+
+#include <BAT/BCModel.h>
+
+#include <string>
+
+// ---------------------------------------------------------
+class Tardis : public BCModel
+{
+
+public:
+    /* using Vec = std::valarray<double>; */
+    using Vec = std::vector<double>;
+
+    // Constructor
+    Tardis(const std::string& name, const std::string& fileName, unsigned run = 9, unsigned maxElements = 0);
+
+    // Destructor
+    ~Tardis();
+
+    // Overload LogLikelihood to implement model
+    virtual double LogLikelihood(const std::vector<double>& parameters);
+
+    // Overload LogAprioriProbability if not using built-in 1D priors
+    virtual double LogAPrioriProbability(const std::vector<double> & parameters);
+
+    virtual void CalculateObservables(const std::vector<double>& parameters);
+
+    /**
+     * Multiply the likelihood by this factor to avoid overflows.
+     *
+     * @param scale factor on the log(!) scale
+     */
+    void rescale(double scale)
+    { this->scale = scale; }
+
+    /**
+     * Set state to update the prior such that minuit gets the correct
+     * target density for prediction and compute the evidence.
+     */
+    void PreparePrediction();
+
+    /**
+     * Predict X for small n at a given nu (=bin center).
+     *
+     * Algorithm: Start computing for N=n, then go to n-1,
+     * n+1. Continue until contributions are negligible. Bear in mind
+     * N>0! Stay on log scale as much as possible?
+     *
+     * @param Xmean If positive, use floor(X / Xmean) as starting guess for most likely N
+     * @param precision Stop searching if contribution from current point is less than `precision * currentValue`
+     */
+    double PredictSmall(unsigned n, double X, double nu, double Xmean = -1, double precision = 1e-2);
+
+    double PredictMedium(unsigned n, double X, double nu);
+
+    double PredictVeryLarge(unsigned n, double X, double nu);
+
+    /**
+     * Compute sum of X in frequency bin
+     */
+    std::tuple<unsigned, double> SumX(double numin, double numax) const;
+
+    /**
+     * Polynomial as a function of nu with coefficients given in range
+     */
+    template <typename T>
+    static double Polyn(T first, T last, const double& nu)
+    {
+        double res = 0.;
+        double power = 1.;
+        for (; first != last; ++first) {
+            res += (*first) * power;
+            power *= nu;
+        }
[…]

### 2016-05-10 real/code/Tardis.h
[code] First version of function + gradient

@@ -61,6 +61,11 @@ public:
      */
     std::tuple<unsigned, double> SumX(double numin, double numax) const;
 
+    double * begin(double * const x)
+    {
+        return x;
+    }
+
     /**
      * Polynomial as a function of nu with coefficients given in range
      */
@@ -76,14 +81,28 @@ public:
         return res;
     }
 
-    double alphaNu(const Vec& x, const double& nu)
+    template <typename T>
+    double alphaNu(const T& x, const double& nu)
     {
-        return Polyn(std::begin(x), std::begin(x) + orderAlpha, nu);
+        using std::begin;
+        return Polyn(begin(x), begin(x) + orderAlpha, nu);
     }
 
-    double betaNu(const Vec& x, const double& nu)
+    template <typename T>
+    double betaNu(const T& x, const double& nu)
     {
-        return Polyn(std::begin(x) + orderAlpha, std::begin(x) + orderAlpha + orderBeta, nu);
+        using std::begin;
+        return Polyn(begin(x) + orderAlpha, begin(x) + orderAlpha + orderBeta, nu);
+    }
+
+    double alphaNu_p(double* x, const double& nu)
+    {
+        return Polyn(x, x + orderAlpha, nu);
+    }
+
+    double betaNu_p(double* x, const double& nu)
+    {
+        return Polyn(x + orderAlpha, x + orderAlpha + orderBeta, nu);
     }
 
     size_t Nsamples() const
@@ -101,7 +120,6 @@ public:
         return sum / Nsamples();
     }
 
-private:
     enum class Target { Default, Gamma, NBGamma };
 
     Vec ReadData(const std::string& fileName, const std::string& dataSet, unsigned run, unsigned maxElements = 0);
@@ -136,6 +154,11 @@ private:
         nPrediction = 0;
         XPrediction = -1;
         nuPrediction = -1;
+    }
+
+    unsigned orderN() const
+    {
+        return orderAlpha + orderBeta;
     }
 
     std::vector<Point> samples;

### 2016-05-10 hans/bat/Tardis.h
[bat] Upload last changes to bat version

@@ -98,7 +98,16 @@ public:
         return samples.size();
     }
 
-private:
+    double mean() const
+    {
+        const double sum = std::accumulate(samples.begin(), samples.end(), 0.0,
+                [](double value, const Point& s1)
+                {
+                    return value + s1.en;
+                });
+        return sum / Nsamples();
+    }
+
     enum class Target { Default, Gamma, NBGamma };
 
     Vec ReadData(const std::string& fileName, const std::string& dataSet, unsigned run, unsigned maxElements = 0);

### 2016-05-10 hans/bat/plot.py

@@ -47,8 +47,11 @@ def plot_estimate(runfile):
 
     plt.title("mode = %g\nmean x = %g +- %g\nsample mean %g +- %g" % (mode, mean, stddev, replicas.T[1].mean(), replicas.T[1].std()))
 
+    plt.xlim(X[1], X[-1])
+    plt.ylim(0.0)
     plt.tight_layout()
     # plt.show()
+
     plt.savefig(s + ".pdf")
 
     plt.clf()

### 2016-05-10 real/code/Tardis.h
[real] First working version
not thread safe: don't modify gradient gsl_vector in loop

@@ -19,20 +19,12 @@ public:
     ~Tardis();
 
     // Overload LogLikelihood to implement model
-    virtual double LogLikelihood(const std::vector<double>& parameters);
+    virtual double LogLikelihood(const std::vector<double>& parameters)
+    { return 0.0; }
 
     // Overload LogAprioriProbability if not using built-in 1D priors
-    virtual double LogAPrioriProbability(const std::vector<double> & parameters);
-
-    virtual void CalculateObservables(const std::vector<double>& parameters);
-
-    /**
-     * Multiply the likelihood by this factor to avoid overflows.
-     *
-     * @param scale factor on the log(!) scale
-     */
-    void rescale(double scale)
-    { this->scale = scale; }
+    virtual double LogAPrioriProbability(const std::vector<double> & parameters)
+    { return 0.0; }
 
     /**
      * Set state to update the prior such that minuit gets the correct
@@ -56,15 +48,12 @@ public:
 
     double PredictVeryLarge(unsigned n, double X, double nu);
 
+    void minimize_gsl();
+
     /**
      * Compute sum of X in frequency bin
      */
     std::tuple<unsigned, double> SumX(double numin, double numax) const;
-
-    double * begin(double * const x)
-    {
-        return x;
-    }
 
     /**
      * Polynomial as a function of nu with coefficients given in range
@@ -167,7 +156,6 @@ public:
     // unsigned orderEnmax;
     const unsigned npoints;
     const double nuMax, alphaMin, betaMin;
-    double scale;
     double evidence;
 
     // parameter of prior on Poisson parameter

### 2016-05-13 real/code/Tardis.h
(wip)[hans,real]  transfer to laptop

@@ -1,33 +1,25 @@
 #pragma once
 
-#include <BAT/BCModel.h>
+#include <Minuit2/MnUserParameters.h>
+#include <Minuit2/FunctionMinimum.h>
 
 #include <gsl/gsl_vector.h>
 #include <gsl/gsl_multimin.h>
 
 #include <string>
+#include <vector>
 
-// ---------------------------------------------------------
-class Tardis : public BCModel
+class Tardis
 {
-
 public:
     /* using Vec = std::valarray<double>; */
     using Vec = std::vector<double>;
 
     // Constructor
-    Tardis(const std::string& name, const std::string& fileName, unsigned run = 9, unsigned maxElements = 0);
+    Tardis(const std::string& fileName, unsigned run = 9, unsigned maxElements = 0);
 
     // Destructor
     ~Tardis();
-
-    // Overload LogLikelihood to implement model
-    virtual double LogLikelihood(const std::vector<double>&)
-    { return 0.0; }
-
-    // Overload LogAprioriProbability if not using built-in 1D priors
-    virtual double LogAPrioriProbability(const std::vector<double>&)
-    { return 0.0; }
 
     /**
      * Set state to update the prior such that minuit gets the correct
@@ -60,6 +52,7 @@ public:
     public:
         static OptimOptions DefaultSimplex();
         static OptimOptions DefaultLBFGS();
+        static OptimOptions DefaultMinuit();
         double eps, step_size, tol;
         unsigned iter_min, iter_max;
     private:
@@ -70,14 +63,14 @@ public:
     gsl_multimin_fminimizer* minimizeSimplex(gsl_multimin_function, gsl_vector* initial, OptimOptions o);
     gsl_multimin_fdfminimizer* minimizeLBFGS(gsl_multimin_function_fdf, gsl_vector* initial, OptimOptions o);
 
-    gsl_multimin_fminimizer* minimizeSimplex(gsl_vector* initial, OptimOptions o);
-    gsl_multimin_fdfminimizer* minimizeLBFGS(gsl_vector* initial, OptimOptions o);
+    gsl_multimin_fminimizer* minimizeSimplex(gsl_vector* initial, OptimOptions);
+    gsl_multimin_fdfminimizer* minimizeLBFGS(gsl_vector* initial, OptimOptions);
 
     gsl_multimin_fminimizer* minimizeSimplex(gsl_vector* initial = nullptr);
[…]

### 2016-05-13 real/draft.tex
[real] add gradients
[real] fix gradients
[code] First version of function + gradient
[draft] copy and paste

@@
 \newcommand{\rmdx}[1]{\mbox{d} #1 \,} % differential
+\newcommand{\firstDeriv}[1]{\frac{\partial}{\partial #1}}
 \newcommand{\secDeriv}[1]{\frac{\partial^2}{\partial #1^2}} % second partial derivative
 \newcommand{\secPartial}[2]{\frac{\partial^2}{\partial #1 \, \partial #2}} % second partial derivative
+\newcommand{\tardis}{TARDIS}
 
@@
   \textbf{\Large Bayesian prediction of supernova luminosity  distributions}\\[8pt]
-  \textbf{\Large based on simulation output: An example from TARDIS}\\[12pt]
+  \textbf{\Large based on simulation output: An example from \tardis}\\[12pt]
 \end{center}
@@
 
-\section{Theory} \label{sec:theory}
+\section{Notation} \label{sec:notation}
 
-\section{Example} \label{sec:example}
+We consider a computer simulation that, given input parameters
+$\bmtheta$ and a random-number seed, produces a set of simulated data
+$ S = \{ (\nu_j, \ell_j): j=1\dots\npack \}$ consisting of $\npack$
+packets, each with a frequency and a luminosity. We distinguish
+between a random variable $L$ and its concrete realization $\ell$.  To
+learn about $\bmtheta$, we need to compare to data from a telescope
+available as luminosities in frequency bins, $D = \{ ( \nu_b, \ell_b):
+b=1 \dots B\}$. The natural estimate from a single simulation is to
+add up the luminosities of all packets in a bin.  With a different
+seed but identical $\bmtheta$, both the number of packets in the bin
+and their luminosities could be different. Our goal is to estimate the
+distribution based on a \emph{single} simulation without recourse to
+expensive repetitions or asymptotic approximations; i.e., we seek
+$P(L_b \cond \nu_b, \npack, \bmnu, \bml )$. In an actual physics
+analysis, one would proceed to plug in the observed value $\ell_b$ and
+perform an analysis to learn about $\bmtheta$. But here we focus on
+the specific problem of $P(L_b \cond \nu_b, \npack, \bmnu, \bml )$ and
+hence omit $\bmtheta$ throughout.
 
-\subsection{Hessians} \label{sec:hessian}
+Various features, for example absorption lines, usually make it
+difficult to explicitly model the spectrum---the luminosity as a
+function of frequency. In a binned analysis, we can avoid these
+difficulties associated to modeling $n_b$. Our approach, however,
+requires an explicit expression of how the distribution of the
+luminosity of a single packet in our simulation changes with $\nu$. It
+is this crucial piece of information that allows us to extend previous
+methods \fred{Refs: Zech, maybe others, too} in dealing with the
+extreme cases when the number of packets in a bin, $n_b$, is one or
+even zero.
+
+\section{Model} \label{sec:model}
+
+We consider the total number of packets $\npack$ fixed, for example
+chosen by a user. In contrast, the number of packets in a bin, $n_b$
+is a realization of a random variable $N_b$ which we assume to follow
+a Poisson distribution. With this assumption, we neglect bin-by-bin
+correlations and treat each bin independently which allows us to focus
+on a single bin and to drop the subscript $b$ in the remainder. Hence
+\begin{align}
+  \lleq{caa}
+  p(N\cond\lambda) &= \frac{e^{-\lambda}\lambda^N}{N!}
+  \qquad N = 0,1,\ldots,\infty ,
+\end{align}
+where the dependency of the Poisson mean $\lambda>0$ on $\nu$ is of no
+concern to us. Let $\bmL$ denote the collection of luminosities of
+packets in a certain bin. The total luminosity in that bin is then
+given by
+\begin{align}
+  \lleq{cab}
+  L = \sum_{j=1}^{N} L_j\,.
+\end{align}
+We assume that each $L_j$ has a distribution depending on $\nu$ and
+other parameters $\bmphi$ and that each packet is statistically
+independent. Since $N$ is random, $L$ is said to follow a compound
+Poisson process. The joint distribution then is
+\begin{align*}
+  \label{cac}
+  (N,\bmL) &\ \sim\ p(N,\bmL\cond \bmphi,\lambda, \nu)
+  = p(\bmL\cond N,\bmphi, \nu)\; P(N\cond \lambda) = \prod_{j=1}^N p(L_j\cond \bmphi, \nu) \; P(N\cond \lambda),
+\end{align*}
+where we used that the Poisson term does not depend on $\nu$
+explicitly. In general, the distribution of $L_j$ could be a of a
+different functional form in each bin. For simplicity, we assume it is
+identical but may depend on parameters that differ from bin to bin. In
+practice, the distribution of packets does not change abruptly with
+frequency. We therefore assume that any parameter needed to fully
+specify $p(L_j)$ is given explicitly as a continuous function of $\nu$
+and the unknown parameters $\bmphi$. This allows us to use a small
+number of parameters to describe $P(L)$ across the entire frequency
+range and in turn lets us determine $\bmphi$ from fitting all
+simulated packets simultaneously whereas $\lambda$ is given only by
+the packets in a bin.
+
+Our goal of predicting $L$ in bin $b$ given the simulation output
+$(\npack,\bml, \bmnu)$ is captured in the probability $p(L\cond \nu,
+\npack,\bml, \bmnu)$ with all relevant parameters integrated out using
+the law of total probability.  To simplify the treatment, we assume
+that the frequency bins are narrow or equivalently that $p(L \cond
+\nu, \npack, \bml, \bmnu)$ does not vary within a bin. This implies
+that all every sample $L_j$ in the bin follows the same distribution.
+As the reference frequency, we choose the bin center.
+
+We first note from \refeq{caa} that $N$ does not depend on $\bml$ at
+all and only indirectly depends on $\bmnu$ through $n$. Since $L$ is
+fully determined by $(N,\bmL)$, it does not depend on
+$(\npack,\bml,\bmnu)$, i.e.
+\begin{align}
+  \lleq{pmi}
+  p(L\cond N,\bmL,\nu,\npack,\bml, \bmnu)
+  = p(L\cond N, \bmL)
+  &= \delta(L - \textstyle\sum_{j=1}^N L_j),
+\end{align}
+\fred{Uniform order: $\nu$ first, then $\ell$, compare to $S$}
+where the $L_j$ are assumed to occur at $\nu$.  Using the law of total
+probability repeatedly, we can write the desired distribution
+\fred{Discuss $N=0$ case somewhere}
+\begin{align}
+  \lleq{pmj}
+  p(L\cond \nu, \npack,\bml, \bmnu)
+  &= \sum_N p(N\cond \nu, \npack,\bml, \bmnu) \, p(L\cond N,\nu, \npack,\bml, \bmnu) \\
+  \lleq{pmjb}
+  &= \sum_N p(N\cond n)\,\int \rmdx{\bmL} p(L\cond N,\bmL,\nu, \npack,\bml, \bmnu)\, p(\bmL\cond N,\nu, \npack,\bml, \bmnu)
+  \\
+  \lleq{pmjc}
+  &= \sum_N p(N\cond n)\,\int \rmdx{\bmL} \delta(L - \textstyle\sum_{j=1}^N L_j)
+  \, p(\bmL\cond N,\nu, \npack,\bml, \bmnu)
+\end{align}
+The dependence on the simulation output $(\npack,\bml, \bmnu)$ enters
+through $p(\bmL\cond N,\npack,\bml, \bmnu)$ in terms of the parameters
+$\bmphi$ of $p(\bmL\cond N,\bmphi)$,
+\begin{align}
+  p(\bmL\cond N,\nu, \npack,\bml, \bmnu)
+  &= \int \rmdx{\bmphi} p(\bmL\cond \bmphi,N, \nu, \npack,\bml, \bmnu)\,p(\bmphi\cond N,\nu, \npack,\bml, \bmnu) \\
+  \lleq{pmk}
+  &= \int \rmdx{\bmphi} p(\bmL\cond \bmphi, N, \nu)\,p(\bmphi\cond N,\nu, \npack,\bml, \bmnu).
+\end{align}
+because $\bmL$ is fully determined by $N,\bmphi, \nu$ and hence
+independent of $\npack,\bml, \bmnu$.  With Bayes' theorem we learn about
+$\bmphi$ from the simulation output
+\begin{align}
+  \lleq{pmkb}
+  p(\bmphi\cond N,\nu, \npack,\bml, \bmnu) &=
+  \frac{p(\bml\cond \bmphi, N,\nu, \npack,\bmnu)\,p(\bmphi\cond N,\nu, \npack,\bmnu)} %
+  {\int \rmdx{\bmphi} p(\bml\cond \bmphi,N,\nu, \npack,\bmnu)\,p(\bmphi\cond N,\nu, \npack,\bmnu)}.
+\end{align}
+By using a multiplicity- and frequency-independent prior for $\bmphi$,
+\begin{align}
+  \lleq{pml}
+  p(\bmphi\cond N,\nu, \npack,\bmnu) &= p(\bmphi),
+\end{align}
+and by recognizing that $\bml$ is independent of $N$ and $\nu$
+\begin{align}
+  \lleq{pmla}
+  p(\bml\cond \bmphi,N,\nu, \npack,\bmnu) = p(\bml\cond \bmphi,\npack,\bmnu)
+\end{align}
+we can write \refeq{pmk} as
+\begin{align}
+  \lleq{pmn}
+  p(\bmL\cond N,\nu, \npack,\bml, \bmnu)
+  &= \int \rmdx{\bmphi} p(\bmL\cond N,\bmphi)\, %
+  \frac{p(\bml\cond \bmphi, \npack, \bmnu)\,p(\bmphi)} %
+  {\int \rmdx{\bmphi} p(\bml\cond \bmphi,\npack,\bmnu)\,p(\bmphi)} \\
+  \lleq{pmo}
+  &= \frac{1}  {p(\bml\cond \npack, \bmnu)} %
+  \int \rmdx{\bmphi} p(\bmL\cond \bmphi, N)\, %
+  p(\bml\cond \bmphi, \npack,\bmnu)\,p(\bmphi)
+\end{align}
+with $p(\bml\cond \npack, \bmnu) = \int \rmdx{\bmphi} p(\bml\cond
+\bmphi, \npack,\bmnu)\,p(\bmphi)$ is the evidence in the denominator.
+Inserting this into (\ref{pmjc}), we obtain the desired master equation:
+\begin{equation}
+  \lleq{pmp}
+  \boxed{
+  p(L\cond \nu, \npack,\bml, \bmnu)
+  = \frac{1}{p(\bml\cond \npack, \bmnu)}
+  \sum_N p(N\cond n)\,\int \rmdx{\bmL} \delta(L - \textstyle\sum_j L_j)
+  \displaystyle\int \rmdx{\bmphi} p(\bmL\cond \bmphi, N, \nu) \, p(\bml\cond \bmphi,\npack,\bmnu)
+  \, p(\bmphi)
+  }
+\end{equation}
+with
+\begin{align}
+  \lleq{pmq}
+  p(\bmL\cond \bmphi, N, \nu) &= \prod_{j=1}^N p(L_j\cond \bmphi, \nu), \\
+  \lleq{pmr}
+  p(\bml\cond \bmphi,\npack,\bmnu) &= \prod_{j=1}^{\npack} p(\ell_j\cond \bmphi, \nu_j)
+\end{align}
+and $p(\cdot \cond \bmphi, \cdot)$ is the same function in both
+equations.  In the next section, we compute the general result for
+$P(N \cond n)$. Equation \refeq{pmp} implies that once we specify the
+prior $p(\bmphi)$ and the likelihood $p(L_j\cond \bmphi)$, we have a
+prediction for $L$ which takes into account all possible $N$,
+$\lambda$ and $\bmphi$. Both choices depend on the simulation at hand
+but a general consideration to render our approach tractable is that
+the distribution of a single packet luminosity $P(L_j \cond \bmphi,
+\nu)$ should be such that the integral over $\bmL$ has a closed-form
+solution. In other words, the distribution of $L = \sum_N L_j$ should
+be known.
+%  An important special case is the class of \emph{stable}
+% distributions.
+
+\subsection{Predicting the number of packets in a bin}
+\fred{Show reference Poisson analysis here of before?}
+
+We use Bayes' theorem to find the posterior distribution for $\lambda$
+given data $n$ and the Poisson hypothesis
+\begin{align}
+  \lleq{pmd}
+  p(\lambda\cond n) %
+  &= \frac{p(n\cond\lambda)\,p(\lambda)} {p(n)}
+  \ =\ \frac{p(n\cond\lambda)\,p(\lambda)} %
+  {\int \rmdx{\lambda} p(n\cond\lambda)\,p(\lambda)}
+\end{align}
+to find the posterior predictive of $N$ given $n$ taking into account
+all possible values of $\lambda$,
+\begin{align}
+  \lleq{pme}
+  p(N\cond n)
+  &= \int \rmdx{\lambda}p(N,\lambda\cond n)
+  \ =\ \int \rmdx{\lambda} p(N\cond\lambda, n)\,p(\lambda\cond n)
+  \ =\ \frac{\int \rmdx{\lambda} p(N\cond\lambda)\,p(n\cond\lambda)\,p(\lambda)}
+  {\int \rmdx{\lambda} p(n\cond\lambda)\,p(\lambda)}
+\end{align}
+where $p(N\cond\lambda,n) = p(N\cond\lambda)$ is independent of $n$.
+A common choice is the prior
+\begin{align}
+  \lleq{pmf}
+  p(\lambda) = c\lambda^{-a},
+\end{align}
+where $a=0$ gives the uniform prior\footnote{For a uniform prior, the
+  case $a{=}0$, we require of course a maximum $\lambda_{\rm max}$. As
+  long as $\lambda_{\rm max}$ is large enough, the integral can
+  effectively be taken to infinity.} and $a=1/2$ gives the Jeffreys
+and reference prior. Both choices lead to an improper prior but the
+posterior predictive is proper provided that $n>0$ or $N>0$.  The
+standard results for the evidence and the posterior predictive are
+\begin{align}
+  \lleq{pmg}
+  p(n) &= %
+  \int_0^\infty d\lambda\, p(n\cond\lambda)\,p(\lambda)
+  \ =\
+  c \int_0^\infty d\lambda\, \frac{e^{-\lambda}\,\lambda^{n-a}}{n!}
+  \ =\ c\frac{(n-a)!}{n!}
+  \\
+  \lleq{pmh}
+  p(N\cond n) %
+  &= \frac{n!}{(n-a)!} \int_0^\infty d\lambda\,
+  \frac{e^{-2\lambda} \lambda^{N+n-a}}{N!\, n!}
+  \ =\ \frac{(N+n-a)!}{N!\,(n-a)!\,2^{N+n+1-a}}
+\end{align}
+which we recognise as a Negative Binomial distribution describing the
+probability to observe $N$ successes given $k=n+1-a$ failures in a
+sequence of Bernoulli trials with success probability $\rho=1/2$
+\begin{align}
+  \lleq{pmhb}
+  p(N\cond n) &= \NBD(N\cond k{=}n{+}1{-}a, \rho{=}\tfrac{1}{2})
+  \ =\ \binom{N{+}n{-}a}{N}
+  \left(\frac{1}{2}\right)^N \left(\frac{1}{2}\right)^{n+1-a}
+\end{align}
+For non-integer $a$, the Negative Binomial generalizes immediately to
+the Pólya distribution by replacing $x! \to \Gamma(x+1)$. Note that
+the distribution of $N$ given $n$ is independent of $\bmL$ and $\bml$
+by construction but depends on $\bmnu$ implicitly through $n$.
+
+\section{Example application: \tardis} \label{sec:example}
+
+The \tardis{} packages simulates supernovae with \fred{Wolfgang's
+  favorite description of his package}. Here we focus on the real
+packets as opposed to the virtual packets output by tardis.
+
+Based on staring at the distribution of luminosities,
+
+We use the shorthand $\alpha_n \equiv \alpha(\nu_n \cond
+\bmphi), \beta_n \equiv \beta(\nu_n \cond \bmphi), \alpha \equiv
+\alpha(\nu \cond \bmphi), \beta \equiv \beta(\nu \cond \bmphi)$.
+
+\subsection{Gradients} \label{sec:gradients}
+
+\begin{align}
+  \label{eq:grad-posterior}
+  \left( \firstDeriv{\alpha_i}, \firstDeriv{\beta_i} \right) \log P(\bmx \cond \alpha, \beta) &=
+  \left(
+    \sum_n \left[ \log \beta_n - \Psi(\alpha_n) + \log x_n \right] \nu_n^i \;,
+    \sum_n \left[ \frac{\alpha_n}{\beta_n} - x_n \right]\nu_n^i
+  \right)
+\end{align}
+
+\begin{align}
+  \label{eq:grad-prediction}
+  \left( \firstDeriv{\alpha_i}, \firstDeriv{\beta_i}, \firstDeriv{N} \right) \log P(X \cond N \alpha, \beta)  &=
+  \left(
+    \left[ \log \beta - \Psi(N \alpha) + \log X \right] N \nu^i,
+    \left[ \frac{N\alpha}{\beta} - X \right]\nu^i,
+    \left[ \log \beta -\Psi(N \alpha) + \log X \right] \alpha
+  \right)
+\end{align}
+
+\begin{align}
+  \label{eq:grad-nb}
+  \firstDeriv{N} \log P(N \cond n) = \Psi(N+n-a+1) - \Psi(N+1) - \log 2
+\end{align}
+
+\subsection{Hessians} \label{sec:hessians}
 
@@
   \label{eq:hess-posterior}
-   &= - \left(
+    - \left(
     \begin{array}{ccc}
@@
     \end{array}
-  \right) \log P(\bmx \cond \npack, \bmnu, \bmphi)
-  \\
+  \right) \log P(\bmx \cond \alpha, \beta)
     &= \left(

### 2016-05-15 real/draft.tex
[draft] Vastly expand on train Hamburg -> Munich

@@
 \newcommand{\reffig}[1]{Fig.~\ref{fig:#1}}
+\newcommand{\refsec}[1]{Sec.~\ref{sec:#1}}
 
@@
 \section{Introduction} \label{sec:intro}
+
+\begin{itemize}
+  \item radiative transfer
+  \item get uncertainty from single run
+  \item cover case when few packets in a bin
+\end{itemize}
 
@@
 Poisson process. The joint distribution then is
-\begin{align*}
-  \label{cac}
+\begin{align}
+  \lleq{cac}
   (N,\bmL) &\ \sim\ p(N,\bmL\cond \bmphi,\lambda, \nu)
   = p(\bmL\cond N,\bmphi, \nu)\; P(N\cond \lambda) = \prod_{j=1}^N p(L_j\cond \bmphi, \nu) \; P(N\cond \lambda),
-\end{align*}
+\end{align}
 where we used that the Poisson term does not depend on $\nu$
@@
 equations.  In the next section, we compute the general result for
-$P(N \cond n)$. Equation \refeq{pmp} implies that once we specify the
-prior $p(\bmphi)$ and the likelihood $p(L_j\cond \bmphi)$, we have a
-prediction for $L$ which takes into account all possible $N$,
+$P(N \cond n)$. Equation \refeq{pmp} then implies that once we specify
+the prior $p(\bmphi)$ and the likelihood $p(L_j\cond \bmphi, \nu)$, we
+have a prediction for $L$ which takes into account all possible $N$,
 $\lambda$ and $\bmphi$. Both choices depend on the simulation at hand
@@
 \nu)$ should be such that the integral over $\bmL$ has a closed-form
-solution. In other words, the distribution of $L = \sum_N L_j$ should
-be known.
+solution. In other words, for a given distribution of $L_j$, the
+distribution of $L = \sum_N L_j$ should be known.
 %  An important special case is the class of \emph{stable}
@@
   c \int_0^\infty d\lambda\, \frac{e^{-\lambda}\,\lambda^{n-a}}{n!}
-  \ =\ c\frac{(n-a)!}{n!}
+  \ =\ c\frac{(n-a)!}{n!} \, ,
   \\
@@
   \frac{e^{-2\lambda} \lambda^{N+n-a}}{N!\, n!}
-  \ =\ \frac{(N+n-a)!}{N!\,(n-a)!\,2^{N+n+1-a}}
+  \ =\ \frac{(N+n-a)!}{N!\,(n-a)!\,2^{N+n+1-a}} \, .
 \end{align}
-which we recognise as a Negative Binomial distribution describing the
-probability to observe $N$ successes given $k=n+1-a$ failures in a
-sequence of Bernoulli trials with success probability $\rho=1/2$
-\begin{align}
-  \lleq{pmhb}
-  p(N\cond n) &= \NBD(N\cond k{=}n{+}1{-}a, \rho{=}\tfrac{1}{2})
-  \ =\ \binom{N{+}n{-}a}{N}
-  \left(\frac{1}{2}\right)^N \left(\frac{1}{2}\right)^{n+1-a}
-\end{align}
-For non-integer $a$, the Negative Binomial generalizes immediately to
-the Pólya distribution by replacing $x! \to \Gamma(x+1)$. Note that
-the distribution of $N$ given $n$ is independent of $\bmL$ and $\bml$
-by construction but depends on $\bmnu$ implicitly through $n$.
 
@@
 
-Based on staring at the distribution of luminosities,
+% Based on staring at the distribution of luminosities,
+To apply \refeq{pmp}, we first need to choose $P(L_j \cond \bmphi,
+\nu)$. A histogram of the luminosities from a \tardis{} run indicates
+a strongly skewed distribution with a single mode that is close to but
+not at the sharp end point $\lmax$. There is a long tail for small
+$\ell$. These features can be accomodated by a $\GammaDist$
+distribution with shape parameter $\alpha$ and rate
+parameter $\beta$
+\begin{align}
+  \lleq{gma}
+  \GammaDist(x | \alpha, \beta) = \frac{\beta^\alpha}{\Gamma(\alpha)} x^{\alpha-1} e^{-\beta x}\, , \; x > 0 .
+\end{align}
+Since $\ell$ seems bounded from above, we have to transform the data.
+We first take the largest luminosity of all \tardis{} packets across the
+entire frequency range and increase it slightly to avoid numerical
+difficulties at the end point
+\begin{align}
+  \lleq{gme}
+  \lmax \equiv (1+\epsilon) \times \max_j \ell_j,
+\end{align}
+where $ \epsilon=10^{-6}$.  We assume the total number of samples
+across all bins is large, hence the effect of considering $\lmax$ a
+constant independent of the data is negligible. We checked that
+leaving $\lmax$ as a free parameter in a fit gave equivalent
+results. We define
+\begin{align}
+  \lleq{gmf}
+  x_j \equiv 1- \frac{\ell_j}{\lmax} > 0
+\end{align}
+and similarly for $L \to X$, then we infer $P(X \cond \nu, \npack,
+\bmx, \bmnu)$. If necessary, we transform back to a probability
+density in $L$ via
+\begin{align}
+  \lleq{gmg}
+  p(L  \cond \nu,  \npack, \bml, \bmnu) = p(X  \cond \nu,  \npack, \bmx, \bmnu) \frac{1}{\lmax} \,.
+\end{align}
+To simplify the numerics, we also rescale the frequencies by a fixed
+constant from $\mathcal{O}(10^{15})$ to $\mathcal{O}(1)$.  A major
+numerical simplification arises because of the $L \to X$
+transformation. Rewriting the master equation~(\ref{pmp}) formally as
+\begin{align}
+  \lleq{gmh}
+  p(L  \cond \nu,  \npack, \bml, \bmnu) = \sum_{N} p(N \cond \nu,  n) p(L \cond N,
+\nu, \npack, \bml, \bmnu)
+\end{align}
+we see that the right-hand side is a mixture density. In our
+application, the distribution of a single packet luminosity is narrow
+in the sense that the standard deviation relative to the mean is at
+the percent level. This implies that $p(L \cond \nu, \npack, \bml,
+\bmnu)$ has one mode from each $N$ and all modes are clearly
+separated. The transformation $L \to x$ effects a strong overlap
+between the distributions $p(X \cond N, \nu, \npack, \bmx, \bmnu)$
+with similar $N$, so $p(X \cond \nu, \npack, \bmx, \bmnu)$ can be well
+approximated by a unimodal distribution; see \reffig{hedgehog}.
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=0.48\textwidth]{hedgehog}
+  \caption{Left: Prediction $p(L \cond  \nu,  \npack, \bml, \bmnu)$, right: $p(x \cond  \nu,  \npack, \bmx, \bmnu)$. I sum over $N$ in both cases. }
+  \fred{I still need to pretty this up. I used the old binned approach
+    with 50 samples in a bin from which I computed the mean and
+    variance to plug into \refeq{gmr} but I computed the sum over $N$
+    instead of the integral.}
+\label{fig:hedgehog}
+\end{figure}
+
+We prefer the Gamma distribution because it is \emph{stable}; i.e.,
+the sum of $N$ Gamma variates is again a Gamma variate
+\begin{align}
+  \lleq{gmb}
+    X_j \sim \GammaDist(\alpha, \beta) \Rightarrow X \equiv \sum_{j=1}^N X_j \sim \GammaDist(N \alpha, \beta).
+\end{align}
+Assuming $\alpha = \alpha(\nu \cond \bmphi)$ and $\beta = \beta(\nu
+\cond \bmphi)$ are functions of $\nu$ that depend on parameters
+$\bmphi$, we set
+\begin{align}
+  \lleq{gmi}
+  p(\bmX \cond \bmphi, N, \nu) &= \prod_{j=1}^N \GammaDist(X_j \cond \alpha, \beta ).
+\end{align}
+and equivalently $P(\bmx \cond \bmphi, \npack, \bmnu)$.  Combining
+\refeq{gmb} and \refeq{gmi}, we can solve the integral
+\begin{align}
+  \lleq{gmc}
+  \int \rmdx{\bmX} \delta(X - \textstyle\sum_j X_j)
+  \,p(\bmX\cond N,\bmphi, \nu) = \GammaDist(X \cond N \alpha, \beta)
+\end{align}
+implying that the master equation~(\ref{pmp}) considerably simplifies
+as
+\begin{align}
+  \lleq{gmd}
+    \boxed{
+    p(X\cond \nu, \npack,\bmx,\bmnu)
+  = \frac{1}{p(\bmx\cond \npack, \bmnu)}
+  \sum_N p(N\cond n)\, \int \rmdx{\bmphi} \GammaDist(X \cond N \alpha, \beta) % p(X \cond N, \bmphi, \nu)
+  \, p(\bmx\cond \bmphi, \npack, \bmnu)
+  \, p(\bmphi).
+  }
+\end{align}
+
+\subsection{Linear model}
+
+In the absence of a physical model to determine $\alpha(\nu \cond
+\bmphi)$ and $\beta(\nu \cond \bmphi)$, we propose to start with a simple polynomial ansatz   $\alpha(\nu \cond \bmphi) = \sum_{k=0}^\Kalpha \alpha_k \nu^k, \beta(\nu \cond \bmphi) = \sum_{k=0}^\Kbeta \beta_k \nu^k$ such that
+\begin{align}
+  \lleq{gmp}
+  \bmphi = (\alpha_0, \dots \alpha_\Kalpha , \beta_0, \dots \beta_\Kbeta)
+\end{align}
+The usual trade off between parsimony and the ability to accurately
+model the frequency dependence determines the orders of the
+polynomials, $\Kalpha$ and $\Kbeta$. This is outside the scope of this
+paper. Here we choose $\Kalpha = 3$ and $\Kbeta = 1$.
+
+\subsection{Prior} \label{sec:priors}
+
+The only missing piece to explicitly evaluate \refeq{gmd} is the prior
+$p(\bmphi)$. To gain some insight, consider the standard $\GammaDist$
+model with $\Kalpha = \Kbeta = 0$.
+% There is a conjugate prior but
+% unfortunately the normalization constant cannot be evaluated in closed
+% form, so it is thus of no use for us.
+How about a noninformative prior? The authors of
+\cite{moala2013bayesian} present an overview of possible choices and
+conclude that the point estimate given by the mode of the posterior is
+nearly insensitive to the prior choice starting from around 30
+independent identically distributed samples. We posit that there are
+this many samples available near any $\nu$ under consideration, so the
+form of the prior is not crucial for our application. But there is
+some explicit prior information that we want to build into our
+analysis. The distribution $p(x \cond \alpha, \beta)$ should always
+vanish at the origin. This is equivalent to requiring a maximum away
+from the boundary leading to $\alpha > 1$. We choose a uniform prior
+on $\bmphi$ on a finite box $V$ through indicator functions
+\begin{align}
+  p(\bmphi) = \mathbf{1}_{\alpha(\bmphi) > 1}(\bmphi) \mathbf{1}_V(\bmphi).
+\end{align}
+
+\subsection{Numerical implementation} \label{sec:numerical}
+
+The box $V$ is chosen by trial and error to be large enough to contain
+the relevant region of $p(\bmx\cond \bmphi, \npack, \bmnu)$. For
+simple functions $\alpha(\nu \cond \bmphi)$, one can find the minimum
+value of $\alpha$ over the entire range of $\nu$ analytically. If this
+is too complicated, we propose to require $\alpha(\nu_j \cond \bmphi)
+> 1$ only at each packet frequency $\nu_j$. In the evaluation of
+$p(\bmx \cond \bmphi, \npack, \bmnu)$, one has to loop over all
+packets and compute $\alpha$ and $\beta$ anyways, so the constraint
+imposes essentially no computational overhead.  To explicitly evaluate
+\refeq{gmd}, we have to overcome the remaining two difficulties: the
+integral over $\bmphi$ and the sum over $N$.
+
+As for the former, the posterior $p(\bmphi \cond \npack, \bmx, \bmnu)$
+is approximately Gaussian in $\phi$ for $\npack$ large, so we can
+apply Laplace's method to perform the $\bmphi$ integrals for both the
+evidence $p(\bmx \cond \npack, \bmnu)$ and the posterior mean $p(X
+\cond N, \npack, \bmx, \bmnu)$. This requires the Hessian at the mode
+and the gradient helps in locating the mode; the analytic expressions
+are summarized in \refsec{appendix}.
+
+Regarding the sum over $N$, we start at $N = \lfloor n-a+1 \rfloor =
+\arg \max_N p(N|n)$ \fred{Actually we should say $p(N \cond n, a)$}
+and continue to larger and smaller values of $N$ until the
+contribution to $P(X \cond \nu, \npack, \bmx, \bmnu)$ becomes
+negligible. In case the relevant contribution is from large $N$ and
+$n$, we interpret $N$ as a continuous variable, replace factorials by
+$\Gamma$ function in \refeq{pmh}, and apply Laplace's method to
+integrate over $N$ and $\bmphi$.
+
+\fred{Our implementation is available online on \href{https://github.com/tardis-sn/XXX}{github}}
+
+\subsection{Graphical output}
+
+\section{Conclusion} \label{sec:conclusion}
+
+\section{Appendix} \label{sec:appendix}
 
@@
   \label{eq:grad-posterior}
-  \left( \firstDeriv{\alpha_i}, \firstDeriv{\beta_i} \right) \log P(\bmx \cond \alpha, \beta) &=
+  \left( \firstDeriv{\alpha_i}, \firstDeriv{\beta_i} \right) \log p(\bmx \cond \alpha, \beta) &=
   \left(
@@
   \label{eq:grad-prediction}
-  \left( \firstDeriv{\alpha_i}, \firstDeriv{\beta_i}, \firstDeriv{N} \right) \log P(X \cond N \alpha, \beta)  &=
+  \left( \firstDeriv{\alpha_i}, \firstDeriv{\beta_i}, \firstDeriv{N} \right) \log p(X \cond N \alpha, \beta)  &=
   \left(
@@
   \label{eq:grad-nb}
-  \firstDeriv{N} \log P(N \cond n) = \Psi(N+n-a+1) - \Psi(N+1) - \log 2
+  \firstDeriv{N} \log p(N \cond n) = \Psi(N+n-a+1) - \Psi(N+1) - \log 2
 \end{align}
@@
     \end{array}
-  \right) \log P(\bmx \cond \alpha, \beta)
+  \right) \log p(\bmx \cond \alpha, \beta)
     &= \left(
@@
     \end{array}
-  \right) \log P(X \cond N \alpha, \beta)
+  \right) \log p(X \cond N \alpha, \beta)
   \\
@@
   \label{eq:hess-nb}
-   - \secDeriv{N} \log P(N \cond n) = \Psi'(N+1) - \Psi'(N+n-a+1)
+   - \secDeriv{N} \log p(N \cond n) = \Psi'(N+1) - \Psi'(N+n-a+1)
 \end{align}
 
-\section{Conclusion} \label{sec:conclusion}
+\bibliographystyle{plain}
+\bibliography{references}
 

### 2016-07-22 real/draft.tex
[real/draft] Consistent notation

@@
 \newcommand{\fred}[1]{\todo[color=orange!40,inline]{#1}} %
+\newcommand{\fredmargin}[1]{\todo[color=orange!40]{#1}} %
 \newcommand{\checked}{\todo[color=green,noline]{\checkmark}} %
@@
 $\bmtheta$ and a random-number seed, produces a set of simulated data
-$ S = \{ (\nu_j, \ell_j): j=1\dots\npack \}$ consisting of $\npack$
-packets, each with a frequency and a luminosity. We distinguish
-between a random variable $L$ and its concrete realization $\ell$.  To
-learn about $\bmtheta$, we need to compare to data from a telescope
-available as luminosities in frequency bins, $D = \{ ( \nu_b, \ell_b):
-b=1 \dots B\}$. The natural estimate from a single simulation is to
-add up the luminosities of all packets in a bin.  With a different
-seed but identical $\bmtheta$, both the number of packets in the bin
-and their luminosities could be different. Our goal is to estimate the
-distribution based on a \emph{single} simulation without recourse to
-expensive repetitions or asymptotic approximations; i.e., we seek
+$ S = \{ (\nu_i, \ell_i): i=1\dots\npack \} = (\bmnu, \bml)$
+consisting of $\npack$ packets, each with a frequency and a
+luminosity. We distinguish between a random variable $L$ and its
+concrete realization $\ell$.  To learn about $\bmtheta$, we need to
+compare to data from a telescope available as luminosities in
+frequency bins, $D = \{ ( \nu_b, \ell_b): b=1 \dots B\}$. The natural
+estimate from a single simulation is to add up the luminosities of all
+packets in a bin to estimate $L_b$.  With a different seed but
+identical $\bmtheta$, both the number of packets in the bin and their
+luminosities are usually different. Our goal is to capture what is
+known about $L$ based on a \emph{single} simulation without recourse
+to expensive repetitions or asymptotic approximations; i.e., we seek
 $P(L_b \cond \nu_b, \npack, \bmnu, \bml )$. In an actual physics
@@
 Various features, for example absorption lines, usually make it
-difficult to explicitly model the spectrum---the luminosity as a
-function of frequency. In a binned analysis, we can avoid these
+difficult to explicitly model the spectrum---the (total) luminosity as
+a function of frequency. In a binned analysis, we can avoid these
 difficulties associated to modeling $n_b$. Our approach, however,
@@
 We assume that each $L_j$ has a distribution depending on $\nu$ and
-other parameters $\bmphi$ and that each packet is statistically
-independent. Since $N$ is random, $L$ is said to follow a compound
-Poisson process. The joint distribution then is
+other parameters $\bmphi$, and that each packet is logically
+independent. Since $N$ follows a Poisson distribution, $L$ is said to
+follow a compound Poisson process. The joint distribution then is
 \begin{align}
@@
   (N,\bmL) &\ \sim\ p(N,\bmL\cond \bmphi,\lambda, \nu)
-  = p(\bmL\cond N,\bmphi, \nu)\; P(N\cond \lambda) = \prod_{j=1}^N p(L_j\cond \bmphi, \nu) \; P(N\cond \lambda),
+  = p(\bmL\cond N,\bmphi, \nu)\; p(N\cond \lambda) = \prod_{j=1}^N p(L_j\cond \bmphi, \nu) \; p(N\cond \lambda),
 \end{align}
@@
 identical but may depend on parameters that differ from bin to bin. In
-practice, the distribution of packets does not change abruptly with
-frequency. We therefore assume that any parameter needed to fully
+practice, the distribution of packets usually does not change abruptly
+with frequency. We therefore assume that any parameter needed to fully
 specify $p(L_j)$ is given explicitly as a continuous function of $\nu$
 and the unknown parameters $\bmphi$. This allows us to use a small
-number of parameters to describe $P(L)$ across the entire frequency
-range and in turn lets us determine $\bmphi$ from fitting all
-simulated packets simultaneously whereas $\lambda$ is given only by
-the packets in a bin.
+number of parameters to describe $p(L)$ across the entire frequency
+range instead of having a few parameters per bin which could lead to
+tens of thousands of parameters for the applications we have in
+mind. So we determine $\bmphi$ from fitting all simulated packets
+simultaneously whereas $\lambda$ is a per-bin quantity.
 
-Our goal of predicting $L$ in bin $b$ given the simulation output
-$(\npack,\bml, \bmnu)$ is captured in the probability $p(L\cond \nu,
-\npack,\bml, \bmnu)$ with all relevant parameters integrated out using
-the law of total probability.  To simplify the treatment, we assume
-that the frequency bins are narrow or equivalently that $p(L \cond
-\nu, \npack, \bml, \bmnu)$ does not vary within a bin. This implies
-that all every sample $L_j$ in the bin follows the same distribution.
-As the reference frequency, we choose the bin center.
+Our goal of predicting $L$ in some bin given the simulation output
+$(\npack,\bmnu, \bml)$ is captured in the probability density
+$p(L\cond \nu, \npack,\bmnu, \bml)$ with all relevant parameters
+integrated out. To simplify the treatment, we assume that the
+frequency bins are narrow or equivalently that $p(L \cond \nu, \npack,
+\bmnu, \bml)$ does not vary within a bin. This implies that every
+sample $L_j$ in the bin follows the same distribution.  As the
+reference frequency, we choose the bin center.
 
-We first note from \refeq{caa} that $N$ does not depend on $\bml$ at
-all and only indirectly depends on $\bmnu$ through $n$. Since $L$ is
-fully determined by $(N,\bmL)$, it does not depend on
+% From \refeq{caa}, $N$ does not depend on $\bml$ at all and only
+% indirectly depends on $\bmnu$ through $n$.
+Since $L$ is fully determined by $(N,\bmL)$, it does not depend on
 $(\npack,\bml,\bmnu)$, i.e.
@@
   \lleq{pmi}
-  p(L\cond N,\bmL,\nu,\npack,\bml, \bmnu)
+  p(L\cond N,\bmL,\nu,\npack,\bmnu, \bml)
   = p(L\cond N, \bmL)
@@
 \end{align}
-\fred{Uniform order: $\nu$ first, then $\ell$, compare to $S$}
-where the $L_j$ are assumed to occur at $\nu$.  Using the law of total
-probability repeatedly, we can write the desired distribution
+\fred{Uniform order: $\nu$ first, then $\ell$, compare to $S$} where
+each $L_j$ is assumed to occur at $\nu$.  Using the law of total
+probability repeatedly, we can write the desired distribution as
 \fred{Discuss $N=0$ case somewhere}
@@
   \lleq{pmj}
-  p(L\cond \nu, \npack,\bml, \bmnu)
-  &= \sum_N p(N\cond \nu, \npack,\bml, \bmnu) \, p(L\cond N,\nu, \npack,\bml, \bmnu) \\
+  p(L\cond \nu, \npack,\bmnu, \bml)
+  &= \sum_N p(N\cond \nu, \npack,\bmnu, \bml) \, p(L\cond N,\nu, \npack,\bmnu, \bml) \\
   \lleq{pmjb}
-  &= \sum_N p(N\cond n)\,\int \rmdx{\bmL} p(L\cond N,\bmL,\nu, \npack,\bml, \bmnu)\, p(\bmL\cond N,\nu, \npack,\bml, \bmnu)
+  &= \sum_N p(N\cond n)\,\int \rmdx{\bmL} p(L\cond N,\bmL,\nu, \npack,\bmnu, \bml)\, p(\bmL\cond N,\nu, \npack,\bmnu, \bml)
   \\
@@
   &= \sum_N p(N\cond n)\,\int \rmdx{\bmL} \delta(L - \textstyle\sum_{j=1}^N L_j)
-  \, p(\bmL\cond N,\nu, \npack,\bml, \bmnu)
+  \, p(\bmL\cond N,\nu, \npack,\bmnu, \bml).
 \end{align}
-The dependence on the simulation output $(\npack,\bml, \bmnu)$ enters
-through $p(\bmL\cond N,\npack,\bml, \bmnu)$ in terms of the parameters
+The dependence on the simulation output $(\npack,\bmnu, \bml)$ enters
+through $p(\bmL\cond N,\npack,\bmnu, \bml)$ in terms of the parameters
 $\bmphi$ of $p(\bmL\cond N,\bmphi)$,
 \begin{align}
-  p(\bmL\cond N,\nu, \npack,\bml, \bmnu)
-  &= \int \rmdx{\bmphi} p(\bmL\cond \bmphi,N, \nu, \npack,\bml, \bmnu)\,p(\bmphi\cond N,\nu, \npack,\bml, \bmnu) \\
+  p(\bmL\cond N,\nu, \npack,\bmnu, \bml)
+  &= \int \rmdx{\bmphi} p(\bmL\cond \bmphi,N, \nu, \npack,\bmnu, \bml)\,p(\bmphi\cond N,\nu, \npack,\bmnu, \bml) \\
   \lleq{pmk}
-  &= \int \rmdx{\bmphi} p(\bmL\cond \bmphi, N, \nu)\,p(\bmphi\cond N,\nu, \npack,\bml, \bmnu).
+  &= \int \rmdx{\bmphi} p(\bmL\cond \bmphi, N, \nu)\,p(\bmphi\cond N,\nu, \npack,\bmnu, \bml).
 \end{align}
 because $\bmL$ is fully determined by $N,\bmphi, \nu$ and hence
-independent of $\npack,\bml, \bmnu$.  With Bayes' theorem we learn about
+independent of $\npack,\bmnu, \bml$.  With Bayes' theorem we learn about
 $\bmphi$ from the simulation output
@@
   \lleq{pmkb}
-  p(\bmphi\cond N,\nu, \npack,\bml, \bmnu) &=
+  p(\bmphi\cond N,\nu, \npack,\bmnu, \bml) &=
   \frac{p(\bml\cond \bmphi, N,\nu, \npack,\bmnu)\,p(\bmphi\cond N,\nu, \npack,\bmnu)} %
@@
   \lleq{pmn}
-  p(\bmL\cond N,\nu, \npack,\bml, \bmnu)
+  p(\bmL\cond N,\nu, \npack,\bmnu, \bml)
   &= \int \rmdx{\bmphi} p(\bmL\cond N,\bmphi)\, %
@@
 \end{align}
-with $p(\bml\cond \npack, \bmnu) = \int \rmdx{\bmphi} p(\bml\cond
-\bmphi, \npack,\bmnu)\,p(\bmphi)$ is the evidence in the denominator.
-Inserting this into (\ref{pmjc}), we obtain the desired master equation:
+where $p(\bml\cond \npack, \bmnu) = \int \rmdx{\bmphi} p(\bml\cond
+\bmphi, \npack,\bmnu)\,p(\bmphi)$ is the evidence.  Inserting \refeq{pmo}
+into \refeq{pmjc}, we obtain the desired master equation:
 \begin{equation}
@@
   \boxed{
-  p(L\cond \nu, \npack,\bml, \bmnu)
+  p(L\cond \nu, \npack,\bmnu, \bml)
   = \frac{1}{p(\bml\cond \npack, \bmnu)}
@@
   \lleq{pmr}
-  p(\bml\cond \bmphi,\npack,\bmnu) &= \prod_{j=1}^{\npack} p(\ell_j\cond \bmphi, \nu_j)
+  p(\bml\cond \bmphi,\npack,\bmnu) &= \prod_{i=1}^{\npack} p(\ell_i\cond \bmphi, \nu_i)
 \end{align}
@@
 have a prediction for $L$ which takes into account all possible $N$,
-$\lambda$ and $\bmphi$. Both choices depend on the simulation at hand
-but a general consideration to render our approach tractable is that
-the distribution of a single packet luminosity $P(L_j \cond \bmphi,
-\nu)$ should be such that the integral over $\bmL$ has a closed-form
-solution. In other words, for a given distribution of $L_j$, the
-distribution of $L = \sum_N L_j$ should be known.
+$\lambda$ and $\bmphi$. Both likelihood and prior depend on the
+simulation under consideration but a general requirement to render our
+approach tractable is that the distribution of a single packet
+luminosity $P(L_j \cond \bmphi, \nu)$ should be such that the integral
+over $\bmL$ has a closed-form solution. In other words, for a given
+distribution of $L_j$, the distribution of $L = \sum_N L_j$ should be
+known.
 %  An important special case is the class of \emph{stable}
@@
 where $p(N\cond\lambda,n) = p(N\cond\lambda)$ is independent of $n$.
-A common choice is the prior
+Common uninformative priors are
 \begin{align}
@@
   long as $\lambda_{\rm max}$ is large enough, the integral can
-  effectively be taken to infinity.} and $a=1/2$ gives the Jeffreys
-and reference prior. Both choices lead to an improper prior but the
-posterior predictive is proper provided that $n>0$ or $N>0$.  The
-standard results for the evidence and the posterior predictive are
+  effectively be taken to infinity.}, $a=1/2$ gives the Jeffreys and
+reference prior, and $a=1$ is the transformation-group result
+advocated by Jaynes\fredmargin{book 12.4.2}. All choices lead to an
+improper prior but the posterior predictive is proper provided that $n
+\ge 1$; i.e., at least one packet is observed in the bin. The standard
+results for the evidence and the posterior predictive are
 \begin{align}
@@
 \end{align}
-Since $\ell$ seems bounded from above, we have to transform the data.
-We first take the largest luminosity of all \tardis{} packets across the
-entire frequency range and increase it slightly to avoid numerical
-difficulties at the end point
+if we transform the data as
+\begin{align}
+  \lleq{gmf}
+  x_i \equiv 1- \frac{\ell_i}{\lmax} > 0
+\end{align}
+and similarly for $L \to X$ because $\ell$ is bounded from above and
+$x$ is bounded from below.  We first take the largest luminosity of
+all \tardis{} packets across the entire frequency range and increase
+it slightly to avoid numerical difficulties at the end point
 \begin{align}
   \lleq{gme}
-  \lmax \equiv (1+\epsilon) \times \max_j \ell_j,
+  \lmax \equiv (1+\epsilon) \times \max_i \ell_i,
 \end{align}
@@
 constant independent of the data is negligible. We checked that
-leaving $\lmax$ as a free parameter in a fit gave equivalent
-results. We define
-\begin{align}
-  \lleq{gmf}
-  x_j \equiv 1- \frac{\ell_j}{\lmax} > 0
-\end{align}
-and similarly for $L \to X$, then we infer $P(X \cond \nu, \npack,
-\bmx, \bmnu)$. If necessary, we transform back to a probability
-density in $L$ via
+leaving $\lmax$ as a free parameter in a fit gave equivalent results.
+
+We infer $P(X \cond \nu, \npack, \bmnu, \bmx)$ and only in the end
+transform back to a probability density in $L$ via
 \begin{align}
   \lleq{gmg}
-  p(L  \cond \nu,  \npack, \bml, \bmnu) = p(X  \cond \nu,  \npack, \bmx, \bmnu) \frac{1}{\lmax} \,.
+  p(L  \cond \nu,  \npack, \bmnu, \bml) = p(X  \cond \nu,  \npack, \bmnu, \bmx) \frac{1}{\lmax} \,.
 \end{align}
-To simplify the numerics, we also rescale the frequencies by a fixed
+To simplify the numerics, we also rescale the frequencies by a
 constant from $\mathcal{O}(10^{15})$ to $\mathcal{O}(1)$.  A major
 numerical simplification arises because of the $L \to X$
-transformation. Rewriting the master equation~(\ref{pmp}) formally as
+transformation. Rewriting the master~\refeq{pmp} formally as
 \begin{align}
   \lleq{gmh}
-  p(L  \cond \nu,  \npack, \bml, \bmnu) = \sum_{N} p(N \cond \nu,  n) p(L \cond N,
-\nu, \npack, \bml, \bmnu)
+  p(L  \cond \nu,  \npack, \bmnu, \bml) = \sum_{N} p(N \cond \nu,  n) p(L \cond N,
+\nu, \npack, \bmnu, \bml),
 \end{align}
 we see that the right-hand side is a mixture density. In our
-application, the distribution of a single packet luminosity is narrow
+application, the distribution of a single-packet luminosity is narrow
 in the sense that the standard deviation relative to the mean is at
-the percent level. This implies that $p(L \cond \nu, \npack, \bml,
-\bmnu)$ has one mode from each $N$ and all modes are clearly
+the percent level. This implies that $p(L \cond \nu, \npack,
+\bmnu, \bml)$ has one mode from each $N$ and all modes are clearly
 separated. The transformation $L \to x$ effects a strong overlap
-between the distributions $p(X \cond N, \nu, \npack, \bmx, \bmnu)$
-with similar $N$, so $p(X \cond \nu, \npack, \bmx, \bmnu)$ can be well
+between the distributions $p(X \cond N, \nu, \npack, \bmnu, \bmx)$
+with similar $N$, so $p(X \cond \nu, \npack, \bmnu, \bmx)$ can be well
 approximated by a unimodal distribution; see \reffig{hedgehog}.
@@
   \centering
-  \includegraphics[width=0.48\textwidth]{hedgehog}
-  \caption{Left: Prediction $p(L \cond  \nu,  \npack, \bml, \bmnu)$, right: $p(x \cond  \nu,  \npack, \bmx, \bmnu)$. I sum over $N$ in both cases. }
+  % \includegraphics[width=0.48\textwidth]{hedgehog}
+  \caption{Left: Prediction $p(L \cond  \nu,  \npack, \bmnu, \bml)$, right: $p(x \cond  \nu,  \npack, \bmnu, \bmx)$. I sum over $N$ in both cases. }
   \fred{I still need to pretty this up. I used the old binned approach
@@
 
-We prefer the Gamma distribution because it is \emph{stable}; i.e.,
+We chose the Gamma distribution because it is \emph{stable}; i.e.,
 the sum of $N$ Gamma variates is again a Gamma variate
@@
   \lleq{gmi}
-  p(\bmX \cond \bmphi, N, \nu) &= \prod_{j=1}^N \GammaDist(X_j \cond \alpha, \beta ).
+  p(\bmX \cond \bmphi, N, \nu) &= \prod_{j=1}^N \GammaDist(X_j \cond \alpha, \beta ),\\
+  p(\bmx \cond \bmphi, \npack, \bmnu) &= \prod_{i=1}^{\npack} \GammaDist(x_i \cond \alpha_i, \beta_i).
 \end{align}
-and equivalently $P(\bmx \cond \bmphi, \npack, \bmnu)$.  Combining
-\refeq{gmb} and \refeq{gmi}, we can solve the integral
+where $\alpha_i \equiv \alpha(\nu_i \cond \bmphi), \beta_i \equiv
+\beta(\nu_i \cond \bmphi)$.  Combining \refeq{gmb} and \refeq{gmi}, we
+can solve the integral
 \begin{align}
@@
 value of $\alpha$ over the entire range of $\nu$ analytically. If this
-is too complicated, we propose to require $\alpha(\nu_j \cond \bmphi)
-> 1$ only at each packet frequency $\nu_j$. In the evaluation of
+is too complicated, we propose to require $\alpha(\nu_i \cond \bmphi)
+> 1$ only at each packet frequency $\nu_i$. In the evaluation of
 $p(\bmx \cond \bmphi, \npack, \bmnu)$, one has to loop over all
@@
 
-As for the former, the posterior $p(\bmphi \cond \npack, \bmx, \bmnu)$
+As for the former, the posterior $p(\bmphi \cond \npack, \bmnu, \bmx)$
 is approximately Gaussian in $\phi$ for $\npack$ large, so we can
@@
 evidence $p(\bmx \cond \npack, \bmnu)$ and the posterior mean $p(X
-\cond N, \npack, \bmx, \bmnu)$. This requires the Hessian at the mode
+\cond N, \npack, \bmnu, \bmx)$. This requires the Hessian at the mode
 and the gradient helps in locating the mode; the analytic expressions
@@
 and continue to larger and smaller values of $N$ until the
-contribution to $P(X \cond \nu, \npack, \bmx, \bmnu)$ becomes
+contribution to $P(X \cond \nu, \npack, \bmnu, \bmx)$ becomes
 negligible. In case the relevant contribution is from large $N$ and
@@
 \end{align}
-with $\alpha_n \equiv \alpha(\nu_n \cond \bmphi), \beta_n \equiv \beta(\nu_n \cond \bmphi)$.
 
@@
 \end{align}
-with $\alpha \equiv \alpha(\nu \cond \bmphi), \beta \equiv \beta(\nu \cond \bmphi)$.
 \begin{align}
@@
 \bibliographystyle{plain}
-\bibliography{references}
+% \bibliography{references}
 

### 2016-07-22 real/code/tardis.jl
(wip) Rewrite problem in julia

@@ -1,0 +1,130 @@
+using DataFrames
+using HDF5
+using Optim
+using Polynomials
+
+# read in data
+# function readdata(run, maxelements=false)
+#     # todo use maxelements
+#     file = h5open("../../posterior/real_tardis_250.h5", "r")
+#     en = file["energies"][:, run]
+#     nus = file["nus"][:, run]
+#     close(file)
+#     return nus, en
+# end
+
+"""
+read energies and frequencies from a particular run
+"""
+
+function readdata(runid=10, filename="../../posterior/real_tardis_250.h5")
+    h5open(filename, "r") do file
+        # read only a single run
+        file["nus"][:,runid], file["energies"][:,runid]
+    end
+end
+
+"""
+    filter_positive(x, y)
+
+For two arrays of same length, filter out elements with negative x and
+return a data frame with columns (nu, energies)
+
+"""
+
+function filter_positive(nus, energies)
+    assert(length(energies) == length(nus))
+
+    mask = (energies .> 0.0)
+    DataFrame(nus=nus[mask], energies=energies[mask])
+end
+
+"""
+
+Sort by frequency, then rescale and flip the energies, then rescale
+the frequencies.
+
+"""
+
+function transform_data!(frame::DataFrame, nuscale::Float64=1e15)
+    sort!(frame, cols=:nus)
+
+    # update energies
+
+    # reference to the frame if we index directly
+    # things like en /= 2 operate on a copy of the data
+    en = frame[:energies]
+    maxen = (1.0 + 1e-6) * maximum(en)
+    for i in eachindex(en)
+        en[i] = 1.0 - en[i] / maxen
+    end
+
+    frame[:nus] /= nuscale
+
+    println("min nu = $(frame[:nus][1]), max en = $(maximum(en))")
+
+    # return nothing, just mutate input
+    nothing
+end
+
+function loggamma(x::Float64, α::Float64, β::Float64)
+  -α * log(β) - lgamma(α) + (α-1)*log(x) - x / β
+end
+
+"""
+
+Create all target functions, return triples of log(f), grad(f),
+Hessian(f).
+
+"""
+
+function targetfactory(frame::DataFrame)
+    αOrder::Int32 = 1
+    βOrder::Int32 = 1
+    αPoly = Poly(Vector{typeof(θ[1])}(αOrder))
+    βPoly = Poly(Vector{typeof(θ[1])}(βOrder))
+    ν::Vector{Float64} = frame[:nus]
+    x::Vector{Float64} = frame[:energies]
+
[…]

### 2016-08-05 real/code/minpoly2.jl
[minpoly2] Make state available outside

@@ -1,20 +1,33 @@
+module MinPoly
+
 using NLopt
 using Polynomials
 
-type PolyOpt
+type OptData
     p::Poly
     dp::Poly
     count::Unsigned
 
-    function PolyOpt(θ::Vector)
+    function OptData(θ::Vector)
         p = Poly(θ)
         dp = polyder(p)
         new(p, dp, 0)
     end
 end
 
+function update_poly!(popt::OptData, θ::Vector)
+    # first the actual polynomial
+    # then its derivative
+    for i in eachindex(popt.p)
+        popt.p[i] = θ[i]
+        if i > 1
+            popt.dp[i]
+        end
+    end
+end
+
 function factory(θ)
-    popt = PolyOpt(θ)
+    popt = OptData(θ)
     myfunc = function(x::Vector, grad::Vector)
         if length(grad) > 0
             grad[1] = polyval(popt.dp, x[1])
@@ -25,17 +38,26 @@ function factory(θ)
     popt, myfunc
 end
 
-θ = [1, -2, 0, 1.34]
-opt = Opt(:LD_VAR2, 1)
-xtol_rel!(opt, 1e-16)
-lower_bounds!(opt, [0.0])
-upper_bounds!(opt, [1.0])
-xinit = [0.5]
-popt, f = factory(θ)
-println(f([0.1], [0.3]))
-min_objective!(opt, f)
-@time minf, minx, ret = optimize(opt, xinit)
-println("got $minf at $minx after $(popt.count) iterations (return $ret)")
+function create_optimizer(θ::Vector, xmin::Real, xmax::Real)
+    opt = Opt(:LD_VAR2, 1)
+    xtol_rel!(opt, 1e-16)
+    lower_bounds!(opt, [xmin])
+    upper_bounds!(opt, [xmax])
+    popt, f = factory(θ)
+    min_objective!(opt, f)
+    return opt, popt
+end
+
+function test()
+    θ = [1, 0.0, -2, 0]
+    opt, popt = create_optimizer(θ, 0.0, 1.0)
+    @time minf, minx, ret = optimize(opt, [0.5])
+    println("got $minf at $minx after $(popt.count) iterations (return $ret)")
+end
+
+end # module MinPoly
+
+MinPoly.test()
 
 # Local Variables:
 # compile-command:"julia minpoly2.jl"

### 2016-08-05 real/code/optim.jl
[tardis, optim] Fix posterior mean, nlopt easier to use

@@ -1,4 +1,5 @@
-include("tardis.jl")
+module optim
+
 using tardis
 using DataFrames
 using Optim
@@ -13,21 +14,13 @@ function problem(;run=11, npackets=typemax(Int64), αOrder=2, βOrder=1)
     return frame, targetfactory(frame, αOrder, βOrder)
 end
 
-function tryoptim()
-    frame, f, ∇f!, Hf! = problem()
-
-    # initial value
-    θ = [1.3, 0, 54]
-
-    res = Optim.optimize(
-                   f, θ, NelderMead(),
-                   ## f, ∇f!, θ, BFGS(),
-                   ## f, ∇f!, Hf!, θ, Newton(),
-                   OptimizationOptions(iterations=1000))
-
-    println(res)
-    θ = Optim.minimizer(res)
-    println("f($(θ)) = $(f(θ))")
+function nlopt_factory(f, ∇f!)
+    function(x::Vector, grad::Vector)
+        if length(grad) > 0
+            ∇f!(x, grad)
+        end
+        f(x)
+    end
 end
 
 function trynlopt()
@@ -42,13 +35,7 @@ function trynlopt()
     ## Plots.pdf("optim.pdf")
     ## return
 
-    function myfunc(x::Vector, grad::Vector)
-        if length(grad) > 0
-            ∇f!(x, grad)
-        end
-
-        return f(x)
-    end
+    myfunc = nlopt_factory(f, ∇f!)
 
     """
     Prior constraint for α > 1 and β > 0 in the form g(θ) ≤ 0
@@ -73,7 +60,7 @@ function trynlopt()
                 grad[offset + m] = -ν[imin]^(m-1)
             end
         end
-        return 1.0 - zmin
+        return minvalue - zmin
     end
 
     # GN_DIRECT_L, GN_ORIG_DIRECT_L
@@ -90,8 +77,8 @@ function trynlopt()
     upperbounds[αOrder + 1] = 200
     ## lower_bounds!(opt, [1.0, 0.0])
     ## upper_bounds!(opt, [4.0, 100.0])
-    ## xtol_rel!(opt, 1e-10)
-    ftol_rel!(opt, 1e-7)
+    xtol_rel!(opt, 1e-5)
+    # ftol_rel!(opt, 1e-7)
     ## maxeval!(opt, 2500)
     max_objective!(opt, myfunc)
 
@@ -112,9 +99,127 @@ function trynlopt()
     Hf!(minx, Hres)
[…]

### 2016-08-05 real/code/tardis.jl

@@ -5,7 +5,7 @@ using HDF5
 using Optim
 using Polynomials
 
-export allocations, filter_positive, laplace, loggamma, νpower, readdata, symmetrize!, targetfactory, transform_data!, update_polynomials!
+export allocations, filter_positive, laplace, loggamma, lognegativebinomial, νpower, PosteriorMean, readdata, symmetrize!, targetfactory, transform_data!, update_polynomials!
 
 """
 read energies and frequencies from a particular run. Limit to read in at most npackets
@@ -65,6 +65,11 @@ function loggamma(x::Float64, α::Float64, β::Float64)
     α * log(β) - lgamma(α) + (α-1)*log(x) - β * x
 end
 
+function lognegativebinomial(N::Real, n::Unsigned, a::Real)
+    tmp = N + n -a + 1
+    lgamma(tmp) - lgamma(N+1) - lgamma(n-a+1) - tmp * log(2)
+end
+
 """
 Reset the coefficients of the polynomials to θ
 """
@@ -103,12 +108,18 @@ Raise ν to the power of m + n, where m and n are 1-based indices.
 
 @enum STATE kLikelihood kMean
 
+type PosteriorMean
+    ν::Real
+    X::Real
+    N::Real
+end
+
 """
 Create all target functions, return triples of log(f), grad(f),
 Hessian(f). Hessian is -∂²log(f), note the minus sign.
 """
 function targetfactory(frame::DataFrame, αOrder::Int64, βOrder::Int64;
-                       X=Nullable{Float64}, N=Nullable{Int64}, νbin=Nullable{Float64})
+                       pm=nothing, evidence=nothing)
     # references
     ν::Vector{Float64} = frame[:nus]
     x::Vector{Float64} = frame[:energies]
@@ -117,17 +128,24 @@ function targetfactory(frame::DataFrame, αOrder::Int64, βOrder::Int64;
 
     # possible input states
     local state::STATE
-    if X == Nullable{Float64} && N == Nullable{Int64} && νbin == Nullable{Float64}
+    if pm === nothing
         state = kLikelihood
-    elseif X > 0 && N >= 0 && νbin >= 0
+    elseif pm.X > 0 && pm.N >= 0 && pm.ν >= 0
         state = kMean
     else
-        error("Invalid input values: X = $X, N = $N, ν = $νbin")
+        error("Invalid input values: X = $(pm.X), N = $(pm.N), ν = $(pm.ν)")
[…]

### 2016-08-05 real/code/MaxPlugin.jl
[MaxPlugin] rename, unsigned -> integer

@@ -1,5 +1,3 @@
-include("tardis.jl")
-
 """
 Find the mode of NegativeBinomial(N|n-a+1, 1/2)*Gamma(X|Nα, β) using Brent's method.
 
@@ -18,10 +16,10 @@ type OptData
     α::Real
     β::Real
     a::Real
-    n::Unsigned
-    count::Unsigned
+    n::Integer
+    count::Integer
 
-    function OptData(X::Real, α::Real, β::Real, a::Real, n::Unsigned)
+    function OptData(X::Real, α::Real, β::Real, a::Real, n::Integer)
         if X < 0 Error("Invalid X < 0: $X") end
         if α < 1 Error("Invalid α < 1: $α") end
         if β < 0 Error("Invalid β < 0: $β") end
@@ -33,17 +31,17 @@ end
 """
 Create the target with data embedded via a closure
 """
-function factory(X::Real, α::Real, β::Real, a::Real, n::Unsigned)
+function factory(X::Real, α::Real, β::Real, a::Real, n::Integer)
     popt = OptData(X, α, β, a, n)
     f(N::Real) = (popt.count += 1; -tardis.lognegativebinomial(N, popt.n, popt.a) - loggamma(popt.X, N * popt.α, popt.β))
     return popt, f
 end
 
-function solve(X::Real, α::Real, β::Real, a::Real, n::Unsigned;
-               ε=1e-2, min=0.0, max=Nullable{Float64})
+function solve(X::Real, α::Real, β::Real, a::Real, n::Integer;
+               ε=1e-2, min=0.0, max=0.0, trace=false)
     popt, f = factory(X, α, β, a, n)
-    if max == Nullable{Float64} max = 10n end
-    optimize(f, min, max, rel_tol=ε)
+    if (max <= 0.0) max = 10n end
+    optimize(f, min, max, rel_tol=ε, show_trace=trace, extended_trace=trace)
 end
 
 function test()
@@ -57,8 +55,7 @@ end
 
 end # module MaxPlugin
 
-
-MaxPlugin.test()
+# MaxPlugin.test()
 
 # Local Variables:
 # compile-command:"julia maxplugin.jl"

### 2016-08-05 real/code/optim.jl
[tardis,optim] First plot of predictions for various X

@@ -2,10 +2,12 @@ module optim
 
 using tardis
 using DataFrames
-using Optim
+using MaxPlugin
 using NLopt
-using Plots;
+using Optim
+using Plots; gr() #pyplot()
 using Polynomials
+using PyCall
 
 function problem(;run=11, npackets=typemax(Int64), αOrder=2, βOrder=1)
     raw_data = readdata(run, npackets=npackets)
@@ -21,96 +23,96 @@ function nlopt_factory(f, ∇f!)
         end
         f(x)
     end
-end
-
-function trynlopt()
-    αOrder = 1
-    βOrder = 1
-    dim = αOrder + βOrder
-    frame, (f, ∇f!, Hf!) = problem(αOrder=αOrder, βOrder=βOrder, run=99)
-
-    ## histogram(frame[:energies], normalized=true)
-    ## α = 1.331; β=58.3
-    ## plot!(x -> exp(loggamma(x, α, β)), linspace(0, 0.2, 500))
-    ## Plots.pdf("optim.pdf")
-    ## return
-
-    myfunc = nlopt_factory(f, ∇f!)
-
-    """
-    Prior constraint for α > 1 and β > 0 in the form g(θ) ≤ 0
-    """
-    function myconstraint(x::Vector, grad::Vector, frame::DataFrame, p::Poly,
-                          minvalue::Float64, offset::Int64)
-        ν::Vector{Float64} = frame[:nus]
-        z::Float64 = 0
-        zmin::Float64 = oftype(1.0, Inf)
-        imin::Int64 = 0
-
-        for i in eachindex(ν)
-            @inbounds z = polyval(p, ν[i])
-            if z < zmin
-                zmin = z
-                imin = i
-            end
-
-        end
-        if length(grad) > 0
-            for m in 1:length(p)
-                grad[offset + m] = -ν[imin]^(m-1)
-            end
-        end
-        return minvalue - zmin
     end
 
-    # GN_DIRECT_L, GN_ORIG_DIRECT_L
-    # LN_BOBYQA, LN_COBYLA
-    # LD_LBFGS, LD_MMA
-    opt = Opt(:LN_COBYLA, dim)
-    lowerbounds = zeros(dim)
-    upperbounds = zeros(dim)
-    lowerbounds[1] = 1.0 # α₀
-    lowerbounds[2:end] = -50.0
-    lowerbounds[αOrder + 1] = 0.0 # β₀
[…]

### 2016-08-05 real/code/tardis.jl

@@ -28,8 +28,7 @@ return a data frame with columns (nu, energies)
 """
 
 function filter_positive(nus, energies)
-    assert(length(energies) == length(nus))
-
+    @assert length(energies) == length(nus)
     mask = (energies .> 0.0)
     DataFrame(nus=nus[mask], energies=energies[mask])
 end
@@ -65,7 +64,7 @@ function loggamma(x::Float64, α::Float64, β::Float64)
     α * log(β) - lgamma(α) + (α-1)*log(x) - β * x
 end
 
-function lognegativebinomial(N::Real, n::Unsigned, a::Real)
+function lognegativebinomial(N::Real, n::Integer, a::Real)
     tmp = N + n -a + 1
     lgamma(tmp) - lgamma(N+1) - lgamma(n-a+1) - tmp * log(2)
 end
@@ -97,7 +96,7 @@ function symmetrize!(H::Array)
     end
 end
 
-function allocations(dim::Int64)
+function allocations(dim::Integer)
     return zeros(dim), zeros((dim, dim))
 end
 
@@ -317,62 +316,5 @@ the mode, -Hf(θ). Both f and Hf are on the log scale.
 """
         laplace(logf::Real, log_det_H::Real, dim::Integer) = logf + dim/2 * log(2pi) - 1/2*log_det_H
         laplace(logf::Real, H::Matrix) = laplace(logf, log(det(H)), size(H)[1])
-
-
-"""
-opt: NLopt
-f: function(x::Vector, grad::Vector, N::Integer)
-"""
-        function search_step!(opt, f, N::Integer, res::Real, θ::Vector, ε::Real)
-            # N>0 required to search
-            if N == 0
-                return res, false
-            end
-
-            # optimize and Laplace integrate
-            # max_objective!(opt, (x, grad) -> f(x, grad, N))
-            # maxf, mode, ret = NLopt.optimize(opt, θ)
-            # integral = laplace(maxf, hessian(mode))
-            maxf, mode, integral = opt_laplace(N, θ)
-
-            # copy back mode
-            θ[:] = mode
-
-            latest = nb(N) + integral
-            res += latest
-
-            return res, (latest / res) > ε
-        end
-
-        function predict_small(frame::DataFrame, αOrder::Int64, βOrder::Int64,
-                               evidence::Real, θ::Vector, X::Real, n::Integer, a::Real; ε=1e-5)
-            # initialization
-            res = 0.0
-            N = Ninit
[…]

### 2016-11-28 real/code/Replica.jl
[replica] separate module

@@ -1,4 +1,5 @@
-push!(LOAD_PATH, pwd())
+module Replica
+
 using tardis
 using Base.Test
 import ConjugatePriors
@@ -10,16 +11,20 @@ using Plots
 import ForwardDiff
 chunk = ForwardDiff.Chunk{3}()
 
+"""
+
+"""
 function plotprediction(sums, predict, stddev, i=1)
-    pyplot()
     Plots.histogram(sums, normed=true, lab="$(length(sums)) replicas")
     vline!([sums[i]], line=:red, lab="lum. in replica $i")
     dist=Distributions.Normal(predict[i], stddev[i])
     plot!(x->Distributions.pdf(dist, x), linspace(predict[i] - 4*stddev[i], predict[i] + 4*stddev[i]), lab="prediction from replica $i", line=:black)
     Plots.pdf("replica-prediction.pdf")
+    # replot to show it interactively
+    plot!()
 end
 
-function read_real(nsim)
+function read_real(nsim; npackets=typemax(Int64))
     n = Array(Int64, (nsim,))
     x = Array(Float64, (nsim,))
     means = zeros(x)
@@ -37,7 +42,7 @@ function read_real(nsim)
     end
 
     function read_run(run)
-        raw_data = readdata(run)
+        raw_data = readdata(run; npackets=npackets)
         frame = filter_positive(raw_data...)
         scale_data!(frame)
         nmin=searchsortedfirst(frame[:nus], νmin)
@@ -202,8 +207,11 @@ function uncertainty(n::Int64, xmean::Float64, xsumsq::Float64, a::Float64=1.0)
     mean, sqrt(exp(logsecond) - mean^2)
 end
 
-function analyze(nsim, readf)
-    n, sums, means, sumsqdiff = readf(nsim)
+"""
+Analyze all replicas and compare mean and standard deviation from the replicas and the model
+"""
+function analyze(nsim, readf; npackets=typemax(Int64))
+    n, sums, means, sumsqdiff = readf(nsim; npackets=npackets)
 
     predict, stddev = begin
         a = zeros(means)
@@ -211,14 +219,14 @@ function analyze(nsim, readf)
 
         for i in 1:nsim
             a[i], b[i] = uncertainty(n[i], means[i], sumsqdiff[i])
-            println("$(i): $(a[i]), $(b[i])")
+           # println("$(i): $(a[i]), $(b[i])")
         end
         a, b
     end
[…]

### 2016-11-29 real/draft.tex
[draft] Reactivate references
[draft] Rename L->Q, start work on asymptotic approx

@@
 
+\DeclareMathOperator{\Expect}{\mathbb{E}}
+\newcommand{\expect}[1]{\Expect\left[#1\right]}
 \DeclareMathOperator{\GammaDist}{Gamma}
+\DeclareMathOperator{\GaussianDist}{\mathcal{N}}
 \newcommand{\Kalpha}{{K_\alpha}}
@@
 \newcommand{\Lmax}{L_{\rm max}}
+\newcommand{\Lumtot}{Q}
+\newcommand{\lumtot}{q}
+\newcommand{\Lum}{L}
+\newcommand{\lum}{\ell}
 \newcommand{\rmdx}[1]{\mbox{d} #1 \,} % differential
@@
 estimate from a single simulation is to add up the luminosities of all
-packets in a bin to estimate $L_b$.  With a different seed but
+packets in a bin to estimate $\Lumtot_b$.  With a different seed but
 identical $\bmtheta$, both the number of packets in the bin and their
 luminosities are usually different. Our goal is to capture what is
-known about $L$ based on a \emph{single} simulation without recourse
+known about $\Lumtot$ based on a \emph{single} simulation without recourse
 to expensive repetitions or asymptotic approximations; i.e., we seek
-$P(L_b \cond \nu_b, \npack, \bmnu, \bml )$. In an actual physics
+$P(\Lumtot_b \cond \nu_b, \npack, \bmnu, \bml )$. In an actual physics
 analysis, one would proceed to plug in the observed value $\ell_b$ and
 perform an analysis to learn about $\bmtheta$. But here we focus on
-the specific problem of $P(L_b \cond \nu_b, \npack, \bmnu, \bml )$ and
+the specific problem of $P(\Lumtot_b \cond \nu_b, \npack, \bmnu, \bml )$ and
 hence omit $\bmtheta$ throughout.
@@
 \end{align}
-where the dependency of the Poisson mean $\lambda>0$ on $\nu$ is of no
-concern to us. Let $\bmL$ denote the collection of luminosities of
-packets in a certain bin. The total luminosity in that bin is then
-given by
+where the dependency of the Poisson mean, $\lambda>0$, on $\nu$ is
+implicit. Let $\bmL$ denote the collection of luminosities of packets
+in a certain bin. The total luminosity in that bin is then given by
 \begin{align}
   \lleq{cab}
-  L = \sum_{j=1}^{N} L_j\,.
+  \Lumtot = \sum_{j=1}^{N} \Lum_j\,.
 \end{align}
-We assume that each $L_j$ has a distribution depending on $\nu$ and
+We assume that each luminosity $\Lum_j$, and we usually omit the subscript $j$ to refer to a single luminosity, has a distribution depending on $\nu$ and
 other parameters $\bmphi$, and that each packet is logically
-independent. Since $N$ follows a Poisson distribution, $L$ is said to
-follow a compound Poisson process. The joint distribution then is
+independent. Since $N$ follows a Poisson distribution, $\Lumtot$
+follows a \emph{compound Poisson process}. The joint distribution then is
 \begin{align}
@@
   (N,\bmL) &\ \sim\ p(N,\bmL\cond \bmphi,\lambda, \nu)
-  = p(\bmL\cond N,\bmphi, \nu)\; p(N\cond \lambda) = \prod_{j=1}^N p(L_j\cond \bmphi, \nu) \; p(N\cond \lambda),
+  = p(\bmL\cond N,\bmphi, \nu)\; p(N\cond \lambda) = \prod_{j=1}^N p(\Lum_j\cond \bmphi, \nu) \; p(N\cond \lambda),
 \end{align}
 where we used that the Poisson term does not depend on $\nu$
-explicitly. In general, the distribution of $L_j$ could be a of a
+explicitly. In general, the distribution of $\Lum_j$ could be a of a
 different functional form in each bin. For simplicity, we assume it is
@@
 with frequency. We therefore assume that any parameter needed to fully
-specify $p(L_j)$ is given explicitly as a continuous function of $\nu$
+specify $p(\Lum_j)$ is given explicitly as a continuous function of $\nu$
 and the unknown parameters $\bmphi$. This allows us to use a small
-number of parameters to describe $p(L)$ across the entire frequency
+number of parameters to describe $p(\Lumtot)$ across the entire frequency
 range instead of having a few parameters per bin which could lead to
@@
 simultaneously whereas $\lambda$ is a per-bin quantity.
+\fred{Reformulate when asymptotic approx. included: it has thousands of parameters}
 
-Our goal of predicting $L$ in some bin given the simulation output
+Our goal of predicting $\Lumtot$ in some bin given the simulation output
 $(\npack,\bmnu, \bml)$ is captured in the probability density
-$p(L\cond \nu, \npack,\bmnu, \bml)$ with all relevant parameters
+$p(\Lumtot\cond \nu, \npack,\bmnu, \bml)$ with all relevant parameters
 integrated out. To simplify the treatment, we assume that the
-frequency bins are narrow or equivalently that $p(L \cond \nu, \npack,
+frequency bins are narrow or equivalently that $p(\Lumtot \cond \nu, \npack,
 \bmnu, \bml)$ does not vary within a bin. This implies that every
-sample $L_j$ in the bin follows the same distribution.  As the
+sample $\Lum_j$ in the bin follows the same distribution.  As the
 reference frequency, we choose the bin center.
@@
 % indirectly depends on $\bmnu$ through $n$.
-Since $L$ is fully determined by $(N,\bmL)$, it does not depend on
+Since $\Lumtot$ is fully determined by $(N,\bmL)$, it does not depend on
 $(\npack,\bml,\bmnu)$, i.e.
@@
   \lleq{pmi}
-  p(L\cond N,\bmL,\nu,\npack,\bmnu, \bml)
-  = p(L\cond N, \bmL)
-  &= \delta(L - \textstyle\sum_{j=1}^N L_j),
+  p(\Lumtot\cond N,\bmL,\nu,\npack,\bmnu, \bml)
+  = p(\Lumtot\cond N, \bmL)
+  &= \delta(\Lumtot - \textstyle\sum_{j=1}^N \Lum_j),
 \end{align}
 \fred{Uniform order: $\nu$ first, then $\ell$, compare to $S$} where
-each $L_j$ is assumed to occur at $\nu$.  Using the law of total
+each $\Lum_j$ is assumed to occur at $\nu$.  Using the law of total
 probability repeatedly, we can write the desired distribution as
@@
   \lleq{pmj}
-  p(L\cond \nu, \npack,\bmnu, \bml)
-  &= \sum_N p(N\cond \nu, \npack,\bmnu, \bml) \, p(L\cond N,\nu, \npack,\bmnu, \bml) \\
+  p(\Lumtot\cond \nu, \npack,\bmnu, \bml)
+  &= \sum_N p(N\cond \nu, \npack,\bmnu, \bml) \, p(\Lumtot\cond N,\nu, \npack,\bmnu, \bml) \\
   \lleq{pmjb}
-  &= \sum_N p(N\cond n)\,\int \rmdx{\bmL} p(L\cond N,\bmL,\nu, \npack,\bmnu, \bml)\, p(\bmL\cond N,\nu, \npack,\bmnu, \bml)
+  &= \sum_N p(N\cond n)\,\int \rmdx{\bmL} p(\Lumtot\cond N,\bmL,\nu, \npack,\bmnu, \bml)\, p(\bmL\cond N,\nu, \npack,\bmnu, \bml)
   \\
   \lleq{pmjc}
-  &= \sum_N p(N\cond n)\,\int \rmdx{\bmL} \delta(L - \textstyle\sum_{j=1}^N L_j)
+  &= \sum_N p(N\cond n)\,\int \rmdx{\bmL} \delta(\Lumtot - \textstyle\sum_{j=1}^N \Lum_j)
   \, p(\bmL\cond N,\nu, \npack,\bmnu, \bml).
@@
   \boxed{
-  p(L\cond \nu, \npack,\bmnu, \bml)
+  p(\Lumtot\cond \nu, \npack,\bmnu, \bml)
   = \frac{1}{p(\bml\cond \npack, \bmnu)}
-  \sum_N p(N\cond n)\,\int \rmdx{\bmL} \delta(L - \textstyle\sum_j L_j)
+  \sum_N p(N\cond n)\,\int \rmdx{\bmL} \delta(\Lumtot - \textstyle\sum_j \Lum_j)
   \displaystyle\int \rmdx{\bmphi} p(\bmL\cond \bmphi, N, \nu) \, p(\bml\cond \bmphi,\npack,\bmnu)
@@
   \lleq{pmq}
-  p(\bmL\cond \bmphi, N, \nu) &= \prod_{j=1}^N p(L_j\cond \bmphi, \nu), \\
+  p(\bmL\cond \bmphi, N, \nu) &= \prod_{j=1}^N p(\Lum_j\cond \bmphi, \nu), \\
   \lleq{pmr}
@@
 $P(N \cond n)$. Equation \refeq{pmp} then implies that once we specify
-the prior $p(\bmphi)$ and the likelihood $p(L_j\cond \bmphi, \nu)$, we
-have a prediction for $L$ which takes into account all possible $N$,
+the prior $p(\bmphi)$ and the likelihood $p(\Lum_j\cond \bmphi, \nu)$, we
+have a prediction for $\Lumtot$ which takes into account all possible $N$,
 $\lambda$ and $\bmphi$. Both likelihood and prior depend on the
@@
 approach tractable is that the distribution of a single packet
-luminosity $P(L_j \cond \bmphi, \nu)$ should be such that the integral
+luminosity $P(\Lum_j \cond \bmphi, \nu)$ should be such that the integral
 over $\bmL$ has a closed-form solution. In other words, for a given
-distribution of $L_j$, the distribution of $L = \sum_N L_j$ should be
+distribution of $\Lum_j$, the distribution of $\Lumtot = \sum_N \Lum_j$ should be
 known.
@@
 % Based on staring at the distribution of luminosities,
-To apply \refeq{pmp}, we first need to choose $P(L_j \cond \bmphi,
+To apply \refeq{pmp}, we first need to choose $P(\Lum_j \cond \bmphi,
 \nu)$. A histogram of the luminosities from a \tardis{} run indicates
@@
 \end{align}
-and similarly for $L \to X$ because $\ell$ is bounded from above and
+and similarly for $\Lumtot \to X$ because $\ell$ is bounded from above and
 $x$ is bounded from below.  We first take the largest luminosity of
@@
 We infer $P(X \cond \nu, \npack, \bmnu, \bmx)$ and only in the end
-transform back to a probability density in $L$ via
+transform back to a probability density in $\Lumtot$ via
 \begin{align}
   \lleq{gmg}
-  p(L  \cond \nu,  \npack, \bmnu, \bml) = p(X  \cond \nu,  \npack, \bmnu, \bmx) \frac{1}{\lmax} \,.
+  p(\Lumtot  \cond \nu,  \npack, \bmnu, \bml) = p(X  \cond \nu,  \npack, \bmnu, \bmx) \frac{1}{\lmax} \,.
 \end{align}
@@
 constant from $\mathcal{O}(10^{15})$ to $\mathcal{O}(1)$.  A major
-numerical simplification arises because of the $L \to X$
+numerical simplification arises because of the $\Lumtot \to X$
 transformation. Rewriting the master~\refeq{pmp} formally as
@@
   \lleq{gmh}
-  p(L  \cond \nu,  \npack, \bmnu, \bml) = \sum_{N} p(N \cond \nu,  n) p(L \cond N,
+  p(\Lumtot  \cond \nu,  \npack, \bmnu, \bml) = \sum_{N} p(N \cond \nu,  n) p(\Lumtot \cond N,
 \nu, \npack, \bmnu, \bml),
@@
 in the sense that the standard deviation relative to the mean is at
-the percent level. This implies that $p(L \cond \nu, \npack,
+the percent level. This implies that $p(\Lumtot \cond \nu, \npack,
 \bmnu, \bml)$ has one mode from each $N$ and all modes are clearly
-separated. The transformation $L \to x$ effects a strong overlap
+separated. The transformation $\Lumtot \to x$ effects a strong overlap
 between the distributions $p(X \cond N, \nu, \npack, \bmnu, \bmx)$
@@
   % \includegraphics[width=0.48\textwidth]{hedgehog}
-  \caption{Left: Prediction $p(L \cond  \nu,  \npack, \bmnu, \bml)$, right: $p(x \cond  \nu,  \npack, \bmnu, \bmx)$. I sum over $N$ in both cases. }
+  \caption{Left: Prediction $p(\Lumtot \cond  \nu,  \npack, \bmnu, \bml)$, right: $p(x \cond  \nu,  \npack, \bmnu, \bmx)$. I sum over $N$ in both cases. }
   \fred{I still need to pretty this up. I used the old binned approach
@@
 
+\section{Asymptotic approximation} \label{sec:asymptotic}
+
+In the case that there are many packets in every bin, some
+simplifications arise as the distribution of individual packets
+$p(\Lum_j \cond \bmphi, \nu)$ becomes irrelevant in the asymptotic regime
+$n \to \infty$, which implies $\lambda \to \infty$. In that case, the
+compound Poisson distribution can be approximated by a Gaussian
+distribution provided that the second moment of $\Lum_j, \expect{\Lum_j^2}$
+is finite. Theorem 4.3.1 of \cite{bening2002generalized} states
+that
+% \begin{align}
+%   \label{eq:asympt-conv}
+%   P\left(\frac{L- \lambda \expect{\Lum_j}}{\sqrt{\lambda \expect{\Lum_j^2}}}  < z \right)  \Rightarrow \Phi(z) \mbox{ as } \lambda \to \infty,
+% \end{align}
+the asymptotic distribution of $\Lumtot$ when $\lambda \to \infty$, given
+the first and second moment of $\Lum_j$, is
+\begin{align}
+  \label{eq:asympt-normal}
+  P(\Lumtot \cond \bmphi, \nu) = \GaussianDist \left( \Lumtot \cond \lambda \expect{\Lum_j}, \lambda \expect{\Lum_j^2} \right).
+\end{align}
+That allows us to model $\Lumtot$ in each bin \emph{without} making concrete
+assumptions about the functional form of $P(\Lum_j \cond \bmphi, \nu)$; i.e., we may now view $\bmphi$ as the collection of first and second moment in each bin
+\begin{align}
+  \label{eq:phi-bin-moments}
+  \bmphi = \left(\expect{\Lum_j(\nu_b)}, \expect{\Lum_j^2(\nu_b)} \right)_{b=1}^{n_b}.
+\end{align}
+The price for this flexibility is that $\bmphi$ may now contain many
+parameters, and each one is determined only from the samples in a
+single bin.
+
 \section{Conclusion} \label{sec:conclusion}
@@
 \bibliographystyle{plain}
-% \bibliography{references}
+\bibliography{references}
 

### 2016-11-29 real/draft.tex
[draft] move bold symbols into separate file
[draft] Include Hans's changes, rewrite basic formulae

@@
 
-\newcommand{\hmod}  {{\mathcal{H}}}  % Model
+\input{symbols}
+
+\newcommand{\hypo}  {{\mathcal{H}}}  % Model
 \newcommand{\ldef}{\;{:}{=}\;}
@@
 
-\newcommand{\smA}{{\scriptscriptstyle A}}
-\newcommand{\smB}{{\scriptscriptstyle B}}
-\newcommand{\smH}{{\scriptscriptstyle H}}
-\newcommand{\smK}{{\scriptscriptstyle K}}
-\newcommand{\smL}{{\scriptscriptstyle L}}
-\newcommand{\smM}{{\scriptscriptstyle M}}
-\newcommand{\smN}{{\scriptscriptstyle N}}
-\newcommand{\smR}{{\scriptscriptstyle R}}
-
-\newcommand{\smo}{{\scriptscriptstyle 1}}
-\newcommand{\smt}{{\scriptscriptstyle 2}}
-\newcommand{\smr}{{\scriptscriptstyle 3}}
-\newcommand{\smf}{{\scriptscriptstyle 4}}
-\newcommand{\smv}{{\scriptscriptstyle 5}}
-\newcommand{\smx}{{\scriptscriptstyle 6}}
-\newcommand{\sms}{{\scriptscriptstyle 7}}
-\newcommand{\sme}{{\scriptscriptstyle 8}}
-\newcommand{\smn}{{\scriptscriptstyle 9}}
-
-\newcommand{\bma}{{\bm{a}}}
-\newcommand{\bmb}{{\bm{b}}}
-\newcommand{\bmc}{{\bm{c}}}
-\newcommand{\bmd}{{\bm{d}}}
-\newcommand{\bme}{{\bm{e}}}
-\newcommand{\bmf}{{\bm{f}}}
-\newcommand{\bmg}{{\bm{g}}}
-\newcommand{\bmh}{{\bm{h}}}
-\newcommand{\bmi}{{\bm{i}}}
-\newcommand{\bmj}{{\bm{j}}}
-\newcommand{\bmk}{{\bm{k}}}
-\newcommand{\bml}{{\bm{\ell}}}
-\newcommand{\bmm}{{\bm{m}}}
-\newcommand{\bmn}{{\bm{n}}}
-\newcommand{\bmo}{{\bm{o}}}
-\newcommand{\bmp}{{\bm{p}}}
-\newcommand{\bmq}{{\bm{q}}}
-\newcommand{\bmr}{{\bm{r}}}
-\newcommand{\bms}{{\bm{s}}}
-\newcommand{\bmt}{{\bm{t}}}
-\newcommand{\bmu}{{\bm{u}}}
-\newcommand{\bmv}{{\bm{v}}}
-\newcommand{\bmw}{{\bm{w}}}
-\newcommand{\bmx}{{{\bm{x}}}}
-\newcommand{\bmy}{{\bm{y}}}
-\newcommand{\bmz}{{\bm{z}}}
-\newcommand{\bmA}{{\bm{A}}}
-\newcommand{\bmB}{{\bm{B}}}
-\newcommand{\bmC}{{\bm{C}}}
-\newcommand{\bmD}{{\bm{D}}}
-\newcommand{\bmE}{{\bm{E}}}
-\newcommand{\bmF}{{\bm{F}}}
-\newcommand{\bmG}{{\bm{G}}}
-\newcommand{\bmH}{{\bm{H}}}
-\newcommand{\bmI}{{\bm{I}}}
-\newcommand{\bmJ}{{\bm{J}}}
-\newcommand{\bmK}{{\bm{K}}}
-\newcommand{\bmL}{{\bm{L}}}
-\newcommand{\bmM}{{\bm{M}}}
-\newcommand{\bmN}{{\bm{N}}}
-\newcommand{\bmO}{{\bm{O}}}
-\newcommand{\bmP}{{\bm{P}}}
-\newcommand{\bmQ}{{\bm{Q}}}
-\newcommand{\bmR}{{\bm{R}}}
-\newcommand{\bmS}{{\bm{S}}}
-\newcommand{\bmT}{{\bm{T}}}
-\newcommand{\bmU}{{\bm{U}}}
-\newcommand{\bmV}{{\bm{V}}}
-\newcommand{\bmW}{{\bm{W}}}
-\newcommand{\bmX}{{\bm{X}}}
-\newcommand{\bmY}{{\bm{Y}}}
-\newcommand{\bmZ}{{\bm{Z}}}
-
-\newcommand{\bmalpha}{{\bm{\alpha}}}
-\newcommand{\bmbeta}{{\bm{\beta}}}
-\newcommand{\bmchi}{{\bm{\chi}}}
-\newcommand{\bmdelta}{{\bm{\delta}}}
-\newcommand{\bmepsilon}{{\bm{\varepsilon}}}
-\newcommand{\bmphi}{{\bm{\phi}}}
-\newcommand{\bmgamma}{{\bm{\gamma}}}
-\newcommand{\bmeta}{{\bm{\eta}}}
-\newcommand{\bmiota}{{\bm{\iota}}}
-\newcommand{\bmkappa}{{\bm{\kappa}}}
-\newcommand{\bmlambda}{{\bm{\lambda}}}
-\newcommand{\bmmu}{{\bm{\mu}}}
-\newcommand{\bmnu}{{\bm{\nu}}}
-\newcommand{\bmomega}{{\bm{\omega}}}
-\newcommand{\bmpi}{{\bm{\pi}}}
-\newcommand{\bmpsi}{{\bm{\psi}}}
-\newcommand{\bmrho}{{\bm{\rho}}}
-\newcommand{\bmsigma}{{\bm{\sigma}}}
-\newcommand{\bmtau}{{\bm{\tau}}}
-\newcommand{\bmtheta}{{\bm{\theta}}}
-\newcommand{\bmupsilon}{{\bm{\upsilon}}}
-\newcommand{\bmxi}{{\bm{\xi}}}
-\newcommand{\bmzeta}{{\bm{\zeta}}}
-
 \newcommand{\refeq}[1]{Eq.~(\ref{#1})}
@@
 \newcommand{\expect}[1]{\Expect\left[#1\right]}
+\newcommand{\expectest}[1]{\widehat{\Expect\left[#1\right]}}
 \DeclareMathOperator{\GammaDist}{Gamma}
 \DeclareMathOperator{\GaussianDist}{\mathcal{N}}
+\DeclareMathOperator{\InvGammaDist}{InvGamma}
 \newcommand{\Kalpha}{{K_\alpha}}
 \newcommand{\Kbeta}{{K_\beta}}
-\newcommand{\npack}{n_p}
+\newcommand{\npack}{{n_p}}
 \newcommand{\lmax}{\ell_{\rm max}}
-\newcommand{\Lmax}{L_{\rm max}}
+\newcommand{\Lmax}{{L_{\rm max}}}
 \newcommand{\Lumtot}{Q}
@@
 \newcommand{\tardis}{TARDIS}
+\DeclareMathOperator{\Variance}{\mathbb{V}}
+\newcommand{\variance}[1]{\Variance\left[#1\right]}
 
@@
 
-\section{Introduction} \label{sec:intro}
+\section{Introduction (Wolfgang)}
 
-\begin{itemize}
-  \item radiative transfer
-  \item get uncertainty from single run
-  \item cover case when few packets in a bin
-\end{itemize}
+\subsection{Astrophysical context}
+
+\textbf{Define here?}
+\begin{enumerate}
+\item Basic notation for packets, luminosity as used below
+\end{enumerate}
+
+\textbf{Issues that need to be resolved}
+\begin{enumerate}
+\item Write everything in terms of frequency $\nu$ or wavelength $\lambda$
+\end{enumerate}
+
+\subsection*{Main points of this paper}
+
+\begin{enumerate}
+\item The first goal of this paper is to show that the Bayesian
+  approach to this problem is well-founded and robust.
+
+\item Luminosity data generated by a single simulation provides a
+  basis for predicting its uncertainty, i.e.\ how much this luminosity
+  can be expected to vary from simulation to simulation.
+
+\item The uncertainty in the luminosity calculated in the Bayesian
+  approach is larger than anticipated because it takes into account
+  not only the variability of the luminosity per packet, but also the
+  distribution of the number of packets.
+
+\item Traditional methods using conventional counting ratio
+  (``relative frequency'') can also make predictions for such
+  uncertainties, but only if the number of packets is large. By
+  contrast, the Bayesian method presented here remains valid also for
+  small packet multiplicities.
+
+\end{enumerate}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Bayesian prediction}
+
+We first sketch the generic notation and relations for Bayesian
+probability theory.
+%
+A particular experiment or simulation generates \textit{data} $D$
+which will often come in the form of a list of $n$ measurements,
+$D = \bmy = (y_1,y_2,\ldots,y_n)$. In order to understand or explain
+this data, the observer constructs one or more \textit{models} or
+\textit{hypotheses} $\hypo$. Central to this construction is the
+specification of a \textit{likelihood} $p(\bmy\cond \bmtheta,\hypo)$
+where $\bmtheta = (\theta_1,\theta_2,\ldots,\theta_K)$ is the set of
+unknown parameters specified by the model. If and when the individual
+measurements $y_i$ are independent, the joint likelihood is the
+product
+\begin{align}
+  p(\bmy\cond \bmtheta) &= \prod_{i=1}^n p(y_i\cond \bmtheta).
+\end{align}
+At this level of analysis, the goal
+is to find which values for these parameters provide a satisfactory
+description of the data. In the Bayesian context, these parameter
+values are best described in terms of the \textit{posterior}, which
+is related to the likelihood by
+\begin{align}
+  \lleq{bth}
+  p(\bmtheta\cond \bmy)
+  &= \frac{p(\bmy\cond\bmtheta)\;p(\bmtheta)}
+    {p(\bmy)}
+    %= \frac{p(\bmy\cond\bmtheta)\;p(\bmtheta)}
+    %{\int d\bmtheta'\, p(\bmy\cond\bmtheta')\;p(\bmtheta')}
+\end{align}
+where $p(\bmtheta)$, the so-called \textit{prior}, is the probability
+assigned to the values of the parameters before any data was taken
+into account and the \textit{evidence}
+\begin{align}
+  p(\bmy) &= \int d\bmtheta'\, p(\bmy\cond\bmtheta')\;p(\bmtheta'),
+\end{align}
+is fully determined by the likelihood and prior.
+
+The scope of this paper is limited to the relationship between
+simulated data generated by radiative transfer simulation codes.
+Likewise, all parameters introduced and calculated relate strictly to
+model descriptions of the simulation.
+%
+While Bayesian concepts and methods can be applied to actual
+experimental data and related physics parameters, we shall not do so
+here.
+
+\hanslong{
+Note that if the prior is uniform, it is independent of $\theta$ and
+will cancel,
+\begin{align}
+  \lleq{bthu}
+  p(\theta\cond \bmy)
+  &= \frac{p(\bmy\cond\theta)} {\int d\theta'\, p(\bmy\cond\theta')}.
+\end{align}
+}
+
+\hanslong{Bayesian prediction?}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{TARDIS}
+\fred{I don't think TARDIS is a good title, too specific}
+
+We consider a computer simulation which, based on tunable input
+parameters $\bmtheta$, generates $\npack$ photon packets, each with a
+particular frequency $\nu_i$ and luminosity $\ell_i$. Together, these
+frequencies and luminosities form the simulation data
+$\bmy = \{ (\nu_i,\ell_i): i = 1,\ldots,\npack\}$.  We divide up the
+frequency spectrum into bins
+$\Delta_b = [\nu_b {-} \tfrac{1}{2} \Delta\nu \,,\, \nu_b {+}
+\tfrac{1}{2} \Delta\nu], b = 1,2,\ldots, B$ of width $\Delta\nu$
+centered on frequencies $\nu_b$. Let $n_b$ be the number of packets
+having a frequency falling into a particular bin $b$, and let $\bml_b$
+be the set of luminosities of those $n_b$ packets. The total
+luminosity in bin $b$ is the sum of luminosities of those packets
+whose frequency falls into $b$,
+\begin{align}
+  \ell_b = \sum_{i=1}^\npack \bm{1}_{\Delta_b}(\nu_i)\,\ell_i
+\end{align}
+where the indicator function
+$\bm{1}_{[a,b]}(x) = 1$ when $x$ is in the interval
+$[a,b]$ and 0 otherwise.
 
 \section{Notation} \label{sec:notation}
+
+\fred{Merge this section with previous section}
 
@@
 a Poisson distribution. With this assumption, we neglect bin-by-bin
-correlations and treat each bin independently which allows us to focus
-on a single bin and to drop the subscript $b$ in the remainder. Hence
+correlations and treat each bin independently. From now on, we focus
+on a single bin and drop the subscript $b$. Hence
 \begin{align}
@@
 \end{align}
-where the dependency of the Poisson mean, $\lambda>0$, on $\nu$ is
-implicit. Let $\bmL$ denote the collection of luminosities of packets
-in a certain bin. The total luminosity in that bin is then given by
+where the dependency of the Poisson mean, $\lambda>0$, on the bin, or
+respectively the frequency $\nu$, is implicit. Let $\bmL = (\Lum_1,\dotsc,\Lum_N)$ denote the
+collection of luminosities of packets in the bin, then the total
+luminosity in that bin is
+% and its realization based on the data $\bmy$ are
 \begin{align}
   \lleq{cab}
-  \Lumtot = \sum_{j=1}^{N} \Lum_j\,.
+  \Lumtot &= \sum_{j=1}^{N} \Lum_j%, \qquad \lumtot = \sum_{j=1}^{n} \lum_j
+  \,.
 \end{align}
-We assume that each luminosity $\Lum_j$, and we usually omit the subscript $j$ to refer to a single luminosity, has a distribution depending on $\nu$ and
-other parameters $\bmphi$, and that each packet is logically
-independent. Since $N$ follows a Poisson distribution, $\Lumtot$
-follows a \emph{compound Poisson process}. The joint distribution then is
+The distribution of the luminosity for each packet,
+$p(\Lum \cond \bmphi)$, depends on parameters $\bmphi$ and potentially
+on the frequency but we suppress the latter dependency.  The packets
+are logically independent. Since $N$ follows a Poisson distribution,
+$\Lumtot$ follows a \emph{compound Poisson process}.
+%
+The key quantity of interest is the distribution of $\Lumtot$ given
+the data $\bmy$, $p(\Lumtot \cond \bmy)$ for the compound Poisson
+process. In the following, we rewrite $p(\Lumtot \cond \bmy)$ to
+obtain an expression that we can readily evaluate numerically. Using
+the law of total probability,
 \begin{align}
-  \lleq{cac}
-  (N,\bmL) &\ \sim\ p(N,\bmL\cond \bmphi,\lambda, \nu)
-  = p(\bmL\cond N,\bmphi, \nu)\; p(N\cond \lambda) = \prod_{j=1}^N p(\Lum_j\cond \bmphi, \nu) \; p(N\cond \lambda),
+  \lleq{qaa}
+  p(\Lumtot \cond \bmy) &= \int \rmdx{\bmphi} p(\bmphi \cond \bmy) p(\Lumtot \cond \bmphi, \bmy) \,,
 \end{align}
-where we used that the Poisson term does not depend on $\nu$
-explicitly. In general, the distribution of $\Lum_j$ could be a of a
+where the first term in the integral is the posterior $p(\bmphi \cond \bmy)$ after observing $\bmy$. We can decompose the second term as
+\begin{align}
+  \lleq{qab}
+  p(\Lumtot \cond \bmphi, \bmy) &= \sum_{N=0}^\infty  p(N \cond \bmphi, \bmy) p(\Lumtot \cond N, \bmphi, \bmy)\,,
+\end{align}
+where $p(N \cond \bmphi, \bmy) = p(N \cond \lambda)$ is given by
+\refeq{caa} and the second term in the sum is
+\begin{align}
+  \lleq{qac}
+  p(\Lumtot \cond N, \bmphi, \bmy) &= \int \rmdx{\bmL} p(\Lumtot \cond \bmL, N, \bmphi, \bmy) p(\bmL \cond N, \bmphi, \bmy)\\
+  \lleq{qaf}
+  &= \int \rmdx{\bmL} \delta \left( \Lumtot - \sum_{j=1}^N \Lum_j  \right) \prod_{j=1}^N p(\Lum_j \cond \bmphi)\,,
+\end{align}
+hence $Q$ does not depend on $\bmy$ explicitly. In the next
+subsections, we provide more details on our assumptions and provide
+explicit expressions for \refeq{qaa}--\refeq{qaf}.
+
+% As the reference frequency, we choose the bin center.
+
+\subsection{Modeling the packet luminosity} \label{sec:pack-lum}
+
+In general, the distribution $p(\Lum_j \cond \bmphi)$ could be a of a
 different functional form in each bin. For simplicity, we assume it is
-identical but may depend on parameters that differ from bin to bin. In
-practice, the distribution of packets usually does not change abruptly
-with frequency. We therefore assume that any parameter needed to fully
-specify $p(\Lum_j)$ is given explicitly as a continuous function of $\nu$
-and the unknown parameters $\bmphi$. This allows us to use a small
-number of parameters to describe $p(\Lumtot)$ across the entire frequency
-range instead of having a few parameters per bin which could lead to
-tens of thousands of parameters for the applications we have in
-mind. So we determine $\bmphi$ from fitting all simulated packets
-simultaneously whereas $\lambda$ is a per-bin quantity.
-\fred{Reformulate when asymptotic approx. included: it has thousands of parameters}
+identical but may depend on parameters $\bmphi$ that need to be
+inferred from $\bmy$.  To simplify the treatment, we assume that the
+frequency bins are narrow or equivalently that $p(\Lumtot \cond \bmy)$
+does not vary within a bin. This implies that every sample $\Lum_j$ in
+the bin follows the same distribution.
 
-Our goal of predicting $\Lumtot$ in some bin given the simulation output
-$(\npack,\bmnu, \bml)$ is captured in the probability density
-$p(\Lumtot\cond \nu, \npack,\bmnu, \bml)$ with all relevant parameters
-integrated out. To simplify the treatment, we assume that the
-frequency bins are narrow or equivalently that $p(\Lumtot \cond \nu, \npack,
-\bmnu, \bml)$ does not vary within a bin. This implies that every
-sample $\Lum_j$ in the bin follows the same distribution.  As the
-reference frequency, we choose the bin center.
+What is the role of $\bmphi$? It contains all parameters needed to
+model the entire spectrum that have to be learned from $\bmy$. To
+begin, we include the Poisson means for each bin
+\begin{align}
+  \lleq{qad} \bmphi = \left( \bmlambda, \bmxi \right) = \left(
+    {\lambda_1,\dotsc,\lambda_{n_b}}, \bmxi \right)\,.
+\end{align}
+\fred{How }
+The number and the luminosity of the packets are assumed independent,
+so the posterior factorizes as
+\begin{align}
+  \lleq{qadd}
+  p(\bmphi \cond \bmy) = p(\bmlambda \cond \bmy) p(\bmxi \cond \bmy) \,.
+\end{align}
 
-% From \refeq{caa}, $N$ does not depend on $\bml$ at all and only
-% indirectly depends on $\bmnu$ through $n$.
-Since $\Lumtot$ is fully determined by $(N,\bmL)$, it does not depend on
-$(\npack,\bml,\bmnu)$, i.e.
+Regarding the additional parameters $\bmxi$, we have essentially two
+extreme options; both illustrated below in \refsec{example} and
+\refsec{asymptotic}. The first option is to assume an explicit
+\emph{parametric} form, for example some function $f$, such that
 \begin{align}
-  \lleq{pmi}
-  p(\Lumtot\cond N,\bmL,\nu,\npack,\bmnu, \bml)
-  = p(\Lumtot\cond N, \bmL)
-  &= \delta(\Lumtot - \textstyle\sum_{j=1}^N \Lum_j),
+  \lleq{qae}
+  p(\Lum \cond \bmphi) = f(\Lum \cond \bmxi, \nu) \,,
 \end{align}
-\fred{Uniform order: $\nu$ first, then $\ell$, compare to $S$} where
-each $\Lum_j$ is assumed to occur at $\nu$.  Using the law of total
-probability repeatedly, we can write the desired distribution as
-\fred{Discuss $N=0$ case somewhere}
-\begin{align}
-  \lleq{pmj}
-  p(\Lumtot\cond \nu, \npack,\bmnu, \bml)
-  &= \sum_N p(N\cond \nu, \npack,\bmnu, \bml) \, p(\Lumtot\cond N,\nu, \npack,\bmnu, \bml) \\
-  \lleq{pmjb}
-  &= \sum_N p(N\cond n)\,\int \rmdx{\bmL} p(\Lumtot\cond N,\bmL,\nu, \npack,\bmnu, \bml)\, p(\bmL\cond N,\nu, \npack,\bmnu, \bml)
-  \\
-  \lleq{pmjc}
-  &= \sum_N p(N\cond n)\,\int \rmdx{\bmL} \delta(\Lumtot - \textstyle\sum_{j=1}^N \Lum_j)
-  \, p(\bmL\cond N,\nu, \npack,\bmnu, \bml).
-\end{align}
-The dependence on the simulation output $(\npack,\bmnu, \bml)$ enters
-through $p(\bmL\cond N,\npack,\bmnu, \bml)$ in terms of the parameters
-$\bmphi$ of $p(\bmL\cond N,\bmphi)$,
-\begin{align}
-  p(\bmL\cond N,\nu, \npack,\bmnu, \bml)
-  &= \int \rmdx{\bmphi} p(\bmL\cond \bmphi,N, \nu, \npack,\bmnu, \bml)\,p(\bmphi\cond N,\nu, \npack,\bmnu, \bml) \\
-  \lleq{pmk}
-  &= \int \rmdx{\bmphi} p(\bmL\cond \bmphi, N, \nu)\,p(\bmphi\cond N,\nu, \npack,\bmnu, \bml).
-\end{align}
-because $\bmL$ is fully determined by $N,\bmphi, \nu$ and hence
-independent of $\npack,\bmnu, \bml$.  With Bayes' theorem we learn about
-$\bmphi$ from the simulation output
-\begin{align}
-  \lleq{pmkb}
-  p(\bmphi\cond N,\nu, \npack,\bmnu, \bml) &=
-  \frac{p(\bml\cond \bmphi, N,\nu, \npack,\bmnu)\,p(\bmphi\cond N,\nu, \npack,\bmnu)} %
-  {\int \rmdx{\bmphi} p(\bml\cond \bmphi,N,\nu, \npack,\bmnu)\,p(\bmphi\cond N,\nu, \npack,\bmnu)}.
-\end{align}
-By using a multiplicity- and frequency-independent prior for $\bmphi$,
-\begin{align}
-  \lleq{pml}
-  p(\bmphi\cond N,\nu, \npack,\bmnu) &= p(\bmphi),
-\end{align}
-and by recognizing that $\bml$ is independent of $N$ and $\nu$
-\begin{align}
-  \lleq{pmla}
-  p(\bml\cond \bmphi,N,\nu, \npack,\bmnu) = p(\bml\cond \bmphi,\npack,\bmnu)
-\end{align}
-we can write \refeq{pmk} as
-\begin{align}
-  \lleq{pmn}
-  p(\bmL\cond N,\nu, \npack,\bmnu, \bml)
-  &= \int \rmdx{\bmphi} p(\bmL\cond N,\bmphi)\, %
-  \frac{p(\bml\cond \bmphi, \npack, \bmnu)\,p(\bmphi)} %
-  {\int \rmdx{\bmphi} p(\bml\cond \bmphi,\npack,\bmnu)\,p(\bmphi)} \\
-  \lleq{pmo}
-  &= \frac{1}  {p(\bml\cond \npack, \bmnu)} %
-  \int \rmdx{\bmphi} p(\bmL\cond \bmphi, N)\, %
-  p(\bml\cond \bmphi, \npack,\bmnu)\,p(\bmphi)
-\end{align}
-where $p(\bml\cond \npack, \bmnu) = \int \rmdx{\bmphi} p(\bml\cond
-\bmphi, \npack,\bmnu)\,p(\bmphi)$ is the evidence.  Inserting \refeq{pmo}
-into \refeq{pmjc}, we obtain the desired master equation:
-\begin{equation}
-  \lleq{pmp}
-  \boxed{
-  p(\Lumtot\cond \nu, \npack,\bmnu, \bml)
-  = \frac{1}{p(\bml\cond \npack, \bmnu)}
-  \sum_N p(N\cond n)\,\int \rmdx{\bmL} \delta(\Lumtot - \textstyle\sum_j \Lum_j)
-  \displaystyle\int \rmdx{\bmphi} p(\bmL\cond \bmphi, N, \nu) \, p(\bml\cond \bmphi,\npack,\bmnu)
-  \, p(\bmphi)
-  }
-\end{equation}
-with
-\begin{align}
-  \lleq{pmq}
-  p(\bmL\cond \bmphi, N, \nu) &= \prod_{j=1}^N p(\Lum_j\cond \bmphi, \nu), \\
-  \lleq{pmr}
-  p(\bml\cond \bmphi,\npack,\bmnu) &= \prod_{i=1}^{\npack} p(\ell_i\cond \bmphi, \nu_i)
-\end{align}
-and $p(\cdot \cond \bmphi, \cdot)$ is the same function in both
-equations.  In the next section, we compute the general result for
-$P(N \cond n)$. Equation \refeq{pmp} then implies that once we specify
-the prior $p(\bmphi)$ and the likelihood $p(\Lum_j\cond \bmphi, \nu)$, we
-have a prediction for $\Lumtot$ which takes into account all possible $N$,
-$\lambda$ and $\bmphi$. Both likelihood and prior depend on the
-simulation under consideration but a general requirement to render our
-approach tractable is that the distribution of a single packet
-luminosity $P(\Lum_j \cond \bmphi, \nu)$ should be such that the integral
-over $\bmL$ has a closed-form solution. In other words, for a given
-distribution of $\Lum_j$, the distribution of $\Lumtot = \sum_N \Lum_j$ should be
-known.
-%  An important special case is the class of \emph{stable}
-% distributions.
+where we now made the dependency on $\nu$ explicit again.  In this
+way, we assume the same distribution for all frequencies and we only
+need to learn the probable values of $\bmxi$ from the data $\bmy$,
+$p(\bmxi \cond \bmy)$. The advantage is that $f(\Lum \cond \bmxi, \nu)$ may
+be well constrained by $\bmy$ if $\npack$ is large and it is easy to
+include further assumptions; e.g., $f$ should be continuous in $\nu$.
 
-\subsection{Predicting the number of packets in a bin}
-\fred{Show reference Poisson analysis here of before?}
+The second option is preferable if it is impossible to elicit an
+appropriate explicit expression for $p(\Lum \cond \bmphi)$ \emph{and}
+each bin has many packets; i.e., $n \gg 1$. Then we can include one or
+more parameters for each bin, consider each bin independently, and
+perform a \emph{nonparametric} analysis. Compared to the parametric
+approach, we now typically have many more parameters, and each one is
+determined only by the packets in one instead of all bins.
 
-We use Bayes' theorem to find the posterior distribution for $\lambda$
-given data $n$ and the Poisson hypothesis
+\subsection{Poisson analysis}
+
+In order to constrain the Poisson mean $\lambda$ in a bin, we have one
+measurement, the number of packets $n$ in that bin. Using Bayes' theorem, we find the posterior
 \begin{align}
@@
   \ =\ \frac{p(n\cond\lambda)\,p(\lambda)} %
-  {\int \rmdx{\lambda} p(n\cond\lambda)\,p(\lambda)}
+  {\int \rmdx{\lambda} p(n\cond\lambda)\,p(\lambda)} \,.
 \end{align}
-to find the posterior predictive of $N$ given $n$ taking into account
-all possible values of $\lambda$,
-\begin{align}
-  \lleq{pme}
-  p(N\cond n)
-  &= \int \rmdx{\lambda}p(N,\lambda\cond n)
-  \ =\ \int \rmdx{\lambda} p(N\cond\lambda, n)\,p(\lambda\cond n)
-  \ =\ \frac{\int \rmdx{\lambda} p(N\cond\lambda)\,p(n\cond\lambda)\,p(\lambda)}
-  {\int \rmdx{\lambda} p(n\cond\lambda)\,p(\lambda)}
-\end{align}
-where $p(N\cond\lambda,n) = p(N\cond\lambda)$ is independent of $n$.
+% to find the posterior predictive of $N$ given $n$ taking into account
+% all possible values of $\lambda$,
+% \begin{align}
+%   \lleq{pme}
+%   p(N\cond n)
+%   &= \int \rmdx{\lambda}p(N,\lambda\cond n)
+%   \ =\ \int \rmdx{\lambda} p(N\cond\lambda, n)\,p(\lambda\cond n)
+%   \ =\ \frac{\int \rmdx{\lambda} p(N\cond\lambda)\,p(n\cond\lambda)\,p(\lambda)}
+%   {\int \rmdx{\lambda} p(n\cond\lambda)\,p(\lambda)}
+% \end{align}
+% where $p(N\cond\lambda,n) = p(N\cond\lambda)$ is independent of $n$.
 Common uninformative priors are
@@
   \lleq{pmf}
-  p(\lambda) = c\lambda^{-a},
+  p(\lambda) = c \lambda^{-a},
 \end{align}
@@
 reference prior, and $a=1$ is the transformation-group result
-advocated by Jaynes\fredmargin{book 12.4.2}. All choices lead to an
-improper prior but the posterior predictive is proper provided that $n
-\ge 1$; i.e., at least one packet is observed in the bin. The standard
-results for the evidence and the posterior predictive are
+advocated by Jaynes~\cite[Ch. 12]{jaynes2003probability}. All choices
+lead to an improper prior but the posterior is proper provided that
+$n \ge 1$; i.e., at least one packet is observed in the bin. The
+standard result for the posterior is
 \begin{align}
-  \lleq{pmg}
-  p(n) &= %
-  \int_0^\infty d\lambda\, p(n\cond\lambda)\,p(\lambda)
-  \ =\
-  c \int_0^\infty d\lambda\, \frac{e^{-\lambda}\,\lambda^{n-a}}{n!}
-  \ =\ c\frac{(n-a)!}{n!} \, ,
-  \\
-  \lleq{pmh}
-  p(N\cond n) %
-  &= \frac{n!}{(n-a)!} \int_0^\infty d\lambda\,
-  \frac{e^{-2\lambda} \lambda^{N+n-a}}{N!\, n!}
-  \ =\ \frac{(N+n-a)!}{N!\,(n-a)!\,2^{N+n+1-a}} \, .
+  \lleq{pms}
+  p(\lambda\cond n) &= \GammaDist(\lambda \cond \alpha = n-a+1, \beta = 1)\,,
 \end{align}
+where the  $\GammaDist$ distribution with shape parameter $\alpha$ and rate
+parameter $\beta$ is defined as
+\begin{align}
+  \lleq{gma}
+  \GammaDist(\lambda | \alpha, \beta) = \frac{\beta^\alpha}{\Gamma(\alpha)} x^{\alpha-1} e^{-\beta x}\, , \; x > 0 .
+\end{align}
+% \begin{align}
+%   \lleq{pmg}
+%   p(n) &= %
+%   \int_0^\infty d\lambda\, p(n\cond\lambda)\,p(\lambda)
+%   \ =\
+%   c \int_0^\infty d\lambda\, \frac{e^{-\lambda}\,\lambda^{n-a}}{n!}
+%   \ =\ c\frac{(n-a)!}{n!} \, ,
+%   \\
+%   \lleq{pmh}
+%   p(N\cond n) %
+%   &= \frac{n!}{(n-a)!} \int_0^\infty d\lambda\,
+%   \frac{e^{-2\lambda} \lambda^{N+n-a}}{N!\, n!}
+%   \ =\ \frac{(N+n-a)!}{N!\,(n-a)!\,2^{N+n+1-a}} \, .
+% \end{align}
 
@@
 In the case that there are many packets in every bin, some
-simplifications arise as the distribution of individual packets
-$p(\Lum_j \cond \bmphi, \nu)$ becomes irrelevant in the asymptotic regime
-$n \to \infty$, which implies $\lambda \to \infty$. In that case, the
-compound Poisson distribution can be approximated by a Gaussian
-distribution provided that the second moment of $\Lum_j, \expect{\Lum_j^2}$
-is finite. Theorem 4.3.1 of \cite{bening2002generalized} states
-that
-% \begin{align}
-%   \label{eq:asympt-conv}
-%   P\left(\frac{L- \lambda \expect{\Lum_j}}{\sqrt{\lambda \expect{\Lum_j^2}}}  < z \right)  \Rightarrow \Phi(z) \mbox{ as } \lambda \to \infty,
-% \end{align}
-the asymptotic distribution of $\Lumtot$ when $\lambda \to \infty$, given
-the first and second moment of $\Lum_j$, is
+simplifications arise as the exact distribution of individual packets
+$p(\Lum \cond \bmphi, \nu)$ becomes irrelevant in the asymptotic
+regime $n \to \infty$, which implies $\lambda \to \infty$. In that
+case, the compound Poisson distribution can be approximated by a
+Gaussian distribution provided that the second moment of
+$\Lum, \expect{\Lum^2}$ is finite. For a compound Poisson process with Poisson mean $\lambda$,
+the general result is that, even for $\lambda$ small,
 \begin{align}
-  \label{eq:asympt-normal}
-  P(\Lumtot \cond \bmphi, \nu) = \GaussianDist \left( \Lumtot \cond \lambda \expect{\Lum_j}, \lambda \expect{\Lum_j^2} \right).
+  \lleq{raa}
+  \expect{Q} = \lambda \expect{L}, \qquad \variance{Q} = \lambda \expect{L^2} \,.
 \end{align}
+
+From Theorem 4.3.1 of \cite{bening2002generalized}, we can conclude
+that the asymptotic distribution of $\Lumtot$ when
+$\lambda \to \infty$, given the first and second moment of $\Lum$, is
+\begin{align}
+  \lleq{rab}
+  P(\Lumtot \cond \bmphi) = \GaussianDist \left( \Lumtot \cond \expect{Q}, \variance{Q} \right) = \GaussianDist \left( \Lumtot \cond \lambda \expect{\Lum}, \lambda \expect{\Lum^2} \right).
+\end{align}
+This result is a drastic simplification as it eliminates both the sum over $N$ in \refeq{qab} and the integral over $\bmL$ in \refeq{qaf}.
 That allows us to model $\Lumtot$ in each bin \emph{without} making concrete
-assumptions about the functional form of $P(\Lum_j \cond \bmphi, \nu)$; i.e., we may now view $\bmphi$ as the collection of first and second moment in each bin
+assumptions about the functional form of $P(\Lum \cond \bmphi)$; i.e., we may now view $\bmphi$ as the collection of first and second moment in each bin in addition to $\lambda$.
+Continuing the notation of \refeq{qadd},
 \begin{align}
-  \label{eq:phi-bin-moments}
-  \bmphi = \left(\expect{\Lum_j(\nu_b)}, \expect{\Lum_j^2(\nu_b)} \right)_{b=1}^{n_b}.
+  \lleq{rac}
+  \bmphi =  \left( \bmlambda, \bmxi \right) = \left(\lambda_b, \expect{\Lum_b}, \expect{\Lum^2_b} \right)_{b=1}^B.
 \end{align}
@@
 parameters, and each one is determined only from the samples in a
-single bin.
+single bin. To obtain the posterior $p(\bmxi \cond \bmy)$, we may focus on
+a single bin again
+\begin{align}
+  \lleq{rad}
+  p(\bmxi \cond \bmy) = \prod_{b=1}^B p\left(\expect{\Lum_b}, \expect{\Lum^2_b} \cond \bmy\right)
+\end{align}
+and omit the subscript $b$. To learn the moments of $\Lum$, we follow
+Jaynes~\cite[Ch. 7.10]{jaynes2003probability} and assign a Gaussian
+sampling distribution
+\begin{align}
+  \lleq{rae}
+  p(\Lum \cond \expect{\Lum}, \expect{\Lum^2} )
+  &= \GaussianDist(\Lum \cond \mu = \expect{\Lum}, \sigma^2
+  = \expect{\Lum^2} - \expect{\Lum}^2)
+\end{align}
+although the actual frequencies will be quite different from a
+Gaussian. Our motivation is that this Gaussian represents our state of
+knowledge \emph{before} observing $\bmy$ well. To make the
+calculations easier, we assume a conjugate prior for unknown $\mu$ and
+$\sigma^2$. The standard result for the posterior is
+\begin{align}
+  \lleq{raf}
+  p(\mu, \sigma^2) &= \GaussianDist(\mu \cond \mu_0^\prime, \frac{\sigma^2}{n_0^\prime}) \InvGammaDist\left( \sigma^2 \cond \frac{\nu_0^\prime}{2}, \frac{\nu_0^\prime \sigma_0^{\prime 2}}{2} \right)
+\end{align}
+where the hyperparameters are updated from prior to posterior as
+\begin{align}
+  \lleq{rag}
+  \expectest{\lum} &= \frac{1}{n} \sum_{i=1}^n \lum_i,\\
+  \lleq{rah}  n_0^\prime &= n_0 + n,\\
+  \lleq{rai}    \mu_0^\prime &= \frac{n_0 \mu_0 + n \expectest{\lum}}{n^\prime},\\
+  \lleq{raj}    \nu_0^\prime &= \nu_0 + n,\\
+  \lleq{rak}    \nu_0^\prime \sigma_0^{\prime 2} &= \nu_0 \sigma_0^{ 2} + \sum_{i=1}^n \left(\lum_j - \expectest{\lum}\right)^2 + \frac{n_0 n}{n^\prime} \left(\mu_0 - \expectest{\lum}\right)\,.
+  \end{align}
+  By default, we take a noninformative prior and set
+  $n_0 = \mu_0 = \nu_0 = 0$, so the value $\sigma_0^2>0$ is
+  irrelevant. The posterior
+  $ p(\expect{\Lum}, \expect{\Lum^2} \cond \bmy)$ is given by
+  \refeq{raf} with $\mu, \sigma^2$ replaced by
+  $\expect{\Lum}, \expect{\Lum^2}$ as indicated in \refeq{rae} because
+  the Jacobian determinant is one.
+
+  Now all ingredients for the prediction of $\Lumtot$ in one bin are
+  ready, and \refeq{qaa} becomes
+\begin{align}
+  \lleq{ral}
+  p(\Lumtot \cond \bmy) &= \int \rmdx{\lambda} \rmdx{\expect{\Lum}} \rmdx{\expect{\Lum^2}} \GaussianDist \left( \Lumtot \cond \lambda \expect{\Lum}, \lambda \expect{\Lum^2} \right) p(\expect{\Lum}, \expect{\Lum^2} \cond \bmy) \,.
+\end{align}
+\fred{$\int \rmdx{\expect{\Lum}}$ looks ugly, I'd rather have other
+  symbols, perhaps just use $\mu$ and $\sigma^2$. The transformation
+  is one-to-one anyways.}
+\fred{Could integrate directly, or use mgf to get moments of $  p(\Lumtot \cond \bmy)$. Then do mode finding, describe initial values and Laplace approximation. }
 
@@
 \end{document}
+\subsection{Old}
+%old text
+The joint distribution then is
+\begin{align}
+  \lleq{cac}
+  % Q, N \sim P(Q, N \cond \bmphi, )
+  (N,\bmL) &\ \sim\ p(N,\bmL\cond \bmphi,\lambda, \nu)
+  = p(\bmL\cond N,\bmphi, \nu)\; p(N\cond \lambda) = \prod_{j=1}^N p(\Lum_j\cond \bmphi, \nu) \; p(N\cond \lambda),
+\end{align}
+\fred{Do we need this equation?}
+where we used that the Poisson term does not depend on $\nu$
+explicitly. In general, the distribution of $\Lum_j$ could be a of a
+different functional form in each bin. For simplicity, we assume it is
+identical but may depend on parameters that differ from bin to bin. In
+practice, the distribution of packets usually does not change abruptly
+with frequency. We therefore assume that any parameter needed to fully
+specify $p(\Lum_j)$ is given explicitly as a continuous function of $\nu$
+and the unknown parameters $\bmphi$. This allows us to use a small
+number of parameters to describe $p(\Lumtot)$ across the entire frequency
+range instead of having a few parameters per bin which could lead to
+tens of thousands of parameters for the applications we have in
+mind. So we determine $\bmphi$ from fitting all simulated packets
+simultaneously whereas $\lambda$ is a per-bin quantity.
+\fred{Reformulate when asymptotic approx. included: it has thousands of parameters}
+
+Our goal of predicting $\Lumtot$ in some bin given the simulation output
+$(\npack,\bmnu, \bml)$ is captured in the probability density
+$p(\Lumtot\cond \nu, \npack,\bmnu, \bml)$ with all relevant parameters
+integrated out. To simplify the treatment, we assume that the
+frequency bins are narrow or equivalently that $p(\Lumtot \cond \nu, \npack,
+\bmnu, \bml)$ does not vary within a bin. This implies that every
+sample $\Lum_j$ in the bin follows the same distribution.  As the
+reference frequency, we choose the bin center.
+
+% From \refeq{caa}, $N$ does not depend on $\bml$ at all and only
+% indirectly depends on $\bmnu$ through $n$.
+Since $\Lumtot$ is fully determined by $(N,\bmL)$, it does not depend on
+$(\npack,\bml,\bmnu)$, i.e.
+\begin{align}
+  \lleq{pmi}
+  p(\Lumtot\cond N,\bmL,\nu,\npack,\bmnu, \bml)
+  = p(\Lumtot\cond N, \bmL)
+  &= \delta(\Lumtot - \textstyle\sum_{j=1}^N \Lum_j),
+\end{align}
+\fred{Uniform order: $\nu$ first, then $\ell$, compare to $S$} where
+each $\Lum_j$ is assumed to occur at $\nu$.  Using the law of total
+probability repeatedly, we can write the desired distribution as
+\fred{Discuss $N=0$ case somewhere}
+\begin{align}
+  \lleq{pmj}
+  p(\Lumtot\cond \nu, \npack,\bmnu, \bml)
+  &= \sum_N p(N\cond \nu, \npack,\bmnu, \bml) \, p(\Lumtot\cond N,\nu, \npack,\bmnu, \bml) \\
+  \lleq{pmjb}
+  &= \sum_N p(N\cond n)\,\int \rmdx{\bmL} p(\Lumtot\cond N,\bmL,\nu, \npack,\bmnu, \bml)\, p(\bmL\cond N,\nu, \npack,\bmnu, \bml)
+  \\
+  \lleq{pmjc}
+  &= \sum_N p(N\cond n)\,\int \rmdx{\bmL} \delta(\Lumtot - \textstyle\sum_{j=1}^N \Lum_j)
+  \, p(\bmL\cond N,\nu, \npack,\bmnu, \bml).
+\end{align}
+The dependence on the simulation output $(\npack,\bmnu, \bml)$ enters
+through $p(\bmL\cond N,\npack,\bmnu, \bml)$ in terms of the parameters
+$\bmphi$ of $p(\bmL\cond N,\bmphi)$,
+\begin{align}
+  p(\bmL\cond N,\nu, \npack,\bmnu, \bml)
+  &= \int \rmdx{\bmphi} p(\bmL\cond \bmphi,N, \nu, \npack,\bmnu, \bml)\,p(\bmphi\cond N,\nu, \npack,\bmnu, \bml) \\
+  \lleq{pmk}
+  &= \int \rmdx{\bmphi} p(\bmL\cond \bmphi, N, \nu)\,p(\bmphi\cond N,\nu, \npack,\bmnu, \bml).
+\end{align}
+because $\bmL$ is fully determined by $N,\bmphi, \nu$ and hence
+independent of $\npack,\bmnu, \bml$.  With Bayes' theorem we learn about
+$\bmphi$ from the simulation output
+\begin{align}
+  \lleq{pmkb}
+  p(\bmphi\cond N,\nu, \npack,\bmnu, \bml) &=
+  \frac{p(\bml\cond \bmphi, N,\nu, \npack,\bmnu)\,p(\bmphi\cond N,\nu, \npack,\bmnu)} %
+  {\int \rmdx{\bmphi} p(\bml\cond \bmphi,N,\nu, \npack,\bmnu)\,p(\bmphi\cond N,\nu, \npack,\bmnu)}.
+\end{align}
+By using a multiplicity- and frequency-independent prior for $\bmphi$,
+\begin{align}
+  \lleq{pml}
+  p(\bmphi\cond N,\nu, \npack,\bmnu) &= p(\bmphi),
+\end{align}
+and by recognizing that $\bml$ is independent of $N$ and $\nu$
+\begin{align}
+  \lleq{pmla}
+  p(\bml\cond \bmphi,N,\nu, \npack,\bmnu) = p(\bml\cond \bmphi,\npack,\bmnu)
+\end{align}
+we can write \refeq{pmk} as
+\begin{align}
+  \lleq{pmn}
+  p(\bmL\cond N,\nu, \npack,\bmnu, \bml)
+  &= \int \rmdx{\bmphi} p(\bmL\cond N,\bmphi)\, %
+  \frac{p(\bml\cond \bmphi, \npack, \bmnu)\,p(\bmphi)} %
+  {\int \rmdx{\bmphi} p(\bml\cond \bmphi,\npack,\bmnu)\,p(\bmphi)} \\
+  \lleq{pmo}
+  &= \frac{1}  {p(\bml\cond \npack, \bmnu)} %
+  \int \rmdx{\bmphi} p(\bmL\cond \bmphi, N)\, %
+  p(\bml\cond \bmphi, \npack,\bmnu)\,p(\bmphi)
+\end{align}
+where $p(\bml\cond \npack, \bmnu) = \int \rmdx{\bmphi} p(\bml\cond
+\bmphi, \npack,\bmnu)\,p(\bmphi)$ is the evidence.  Inserting \refeq{pmo}
+into \refeq{pmjc}, we obtain the desired master equation:
+\begin{equation}
+  \lleq{pmp}
+  \boxed{
+  p(\Lumtot\cond \nu, \npack,\bmnu, \bml)
+  = \frac{1}{p(\bml\cond \npack, \bmnu)}
+  \sum_N p(N\cond n)\,\int \rmdx{\bmL} \delta(\Lumtot - \textstyle\sum_j \Lum_j)
+  \displaystyle\int \rmdx{\bmphi} p(\bmL\cond \bmphi, N, \nu) \, p(\bml\cond \bmphi,\npack,\bmnu)
+  \, p(\bmphi)
+  }
+\end{equation}
+with
+\begin{align}
+  \lleq{pmq}
+  p(\bmL\cond \bmphi, N, \nu) &= \prod_{j=1}^N p(\Lum_j\cond \bmphi, \nu), \\
+  \lleq{pmr}
+  p(\bml\cond \bmphi,\npack,\bmnu) &= \prod_{i=1}^{\npack} p(\ell_i\cond \bmphi, \nu_i)
+\end{align}
+and $p(\cdot \cond \bmphi, \cdot)$ is the same function in both
+equations.  In the next section, we compute the general result for
+$P(N \cond n)$. Equation \refeq{pmp} then implies that once we specify
+the prior $p(\bmphi)$ and the likelihood $p(\Lum_j\cond \bmphi, \nu)$, we
+have a prediction for $\Lumtot$ which takes into account all possible $N$,
+$\lambda$ and $\bmphi$. Both likelihood and prior depend on the
+simulation under consideration but a general requirement to render our
+approach tractable is that the distribution of a single packet
+luminosity $P(\Lum_j \cond \bmphi, \nu)$ should be such that the integral
+over $\bmL$ has a closed-form solution. In other words, for a given
+distribution of $\Lum_j$, the distribution of $\Lumtot = \sum_N \Lum_j$ should be
+known.
+%  An important special case is the class of \emph{stable}
+% distributions.
+
+We use Bayes' theorem to find the posterior distribution for
 

### 2016-11-30 real/draft.tex
[draft] rewrite asympt. with \mu, \sigma^2 instead of moments

@@
 Gaussian distribution provided that the second moment of
-$\Lum, \expect{\Lum^2}$ is finite. For a compound Poisson process with Poisson mean $\lambda$,
-the general result is that, even for $\lambda$ small,
+$\Lum, \expect{\Lum^2}$ is finite. To see this in more detail, we define
+\begin{align}
+  \lleq{ram}
+  \mu \equiv \expect{\Lum}, \qquad \sigma^2 \equiv \variance{\Lum}\,.
+\end{align}
+For a compound Poisson process with arbitrary Poisson mean $\lambda > 0$,
 \begin{align}
   \lleq{raa}
-  \expect{Q} = \lambda \expect{L}, \qquad \variance{Q} = \lambda \expect{L^2} \,.
+  \expect{Q} =\lambda \expect{L} = \lambda \mu, \qquad \variance{Q} = \lambda \expect{L^2} = \lambda (\mu^2 + \sigma^2) \,.
 \end{align}
@@
   \lleq{rab}
-  P(\Lumtot \cond \bmphi) = \GaussianDist \left( \Lumtot \cond \expect{Q}, \variance{Q} \right) = \GaussianDist \left( \Lumtot \cond \lambda \expect{\Lum}, \lambda \expect{\Lum^2} \right).
+  P(\Lumtot \cond \bmphi) = \GaussianDist \left( \Lumtot \cond \expect{Q}, \variance{Q} \right) = \GaussianDist \left( \Lumtot \cond \lambda \mu, \lambda (\mu^2 + \sigma^2) \right).
 \end{align}
@@
 That allows us to model $\Lumtot$ in each bin \emph{without} making concrete
-assumptions about the functional form of $P(\Lum \cond \bmphi)$; i.e., we may now view $\bmphi$ as the collection of first and second moment in each bin in addition to $\lambda$.
+assumptions about the functional form of $P(\Lum \cond \bmphi)$; i.e., we may now view $\bmphi$ as the collection of first and second moments in each bin in addition to $\lambda$.
 Continuing the notation of \refeq{qadd},
@@
   \lleq{rac}
-  \bmphi =  \left( \bmlambda, \bmxi \right) = \left(\lambda_b, \expect{\Lum_b}, \expect{\Lum^2_b} \right)_{b=1}^B.
+  \bmphi =  \left(\lambda_b, \mu_b, \sigma^2_b \right)_{b=1}^B.
 \end{align}
@@
 parameters, and each one is determined only from the samples in a
-single bin. To obtain the posterior $p(\bmxi \cond \bmy)$, we may focus on
-a single bin again
+single bin. The joint posterior is
 \begin{align}
   \lleq{rad}
-  p(\bmxi \cond \bmy) = \prod_{b=1}^B p\left(\expect{\Lum_b}, \expect{\Lum^2_b} \cond \bmy\right)
+  p(\bmphi \cond \bmy ) = \prod_{b=1}^B p\left(\lambda_b, \mu_b, \sigma^2_b \cond \bmy\right),
 \end{align}
-and omit the subscript $b$. To learn the moments of $\Lum$, we follow
+Focussing on a single bin, we omit the subscript $b$. As before, we
+further assume that
+\begin{align}
+  \lleq{rao}
+p(\lambda, \mu, \sigma^2 \cond \bmy) = p(\lambda \cond \bmy) p(\mu,
+\sigma \cond \bmy).
+\end{align}
+The posterior for $\lambda$ is given by \refeq{pms}.  To learn mean
+and variance of $\Lum$ from $\bmy$, we follow
 Jaynes~\cite[Ch. 7.10]{jaynes2003probability} and assign a Gaussian
@@
   \lleq{rae}
-  p(\Lum \cond \expect{\Lum}, \expect{\Lum^2} )
-  &= \GaussianDist(\Lum \cond \mu = \expect{\Lum}, \sigma^2
-  = \expect{\Lum^2} - \expect{\Lum}^2)
+  p(\Lum \cond \mu, \sigma^2 ) &= \GaussianDist(\Lum \cond \mu, \sigma^2)
 \end{align}
-although the actual frequencies will be quite different from a
+although the actual frequencies may be quite different from a
 Gaussian. Our motivation is that this Gaussian represents our state of
@@
   \lleq{raf}
-  p(\mu, \sigma^2) &= \GaussianDist(\mu \cond \mu_0^\prime, \frac{\sigma^2}{n_0^\prime}) \InvGammaDist\left( \sigma^2 \cond \frac{\nu_0^\prime}{2}, \frac{\nu_0^\prime \sigma_0^{\prime 2}}{2} \right)
+  p(\mu, \sigma^2 \cond \bmy) &= \GaussianDist(\mu \cond \mu_0^\prime, \frac{\sigma^2}{n_0^\prime}) \InvGammaDist\left( \sigma^2 \cond \frac{\nu_0^\prime}{2}, \frac{\nu_0^\prime \sigma_0^{\prime 2}}{2} \right)
 \end{align}
-where the hyperparameters are updated from prior to posterior as
+where the hyperparameters are updated from prior to posterior based on $\bmy$ as
 \begin{align}
@@
   $n_0 = \mu_0 = \nu_0 = 0$, so the value $\sigma_0^2>0$ is
-  irrelevant. The posterior
-  $ p(\expect{\Lum}, \expect{\Lum^2} \cond \bmy)$ is given by
-  \refeq{raf} with $\mu, \sigma^2$ replaced by
-  $\expect{\Lum}, \expect{\Lum^2}$ as indicated in \refeq{rae} because
-  the Jacobian determinant is one.
+  irrelevant.
+  % The posterior
+  % $ p(\expect{\Lum}, \expect{\Lum^2} \cond \bmy)$ is given by
+  % \refeq{raf} with $\mu, \sigma^2$ replaced by
+  % $\expect{\Lum}, \expect{\Lum^2}$ as indicated in \refeq{rae} because
+  % the Jacobian determinant is one.
 
   Now all ingredients for the prediction of $\Lumtot$ in one bin are
-  ready, and \refeq{qaa} becomes
+  ready, and \refeq{qaa} simplifies to
 \begin{align}
   \lleq{ral}
-  p(\Lumtot \cond \bmy) &= \int \rmdx{\lambda} \rmdx{\expect{\Lum}} \rmdx{\expect{\Lum^2}} \GaussianDist \left( \Lumtot \cond \lambda \expect{\Lum}, \lambda \expect{\Lum^2} \right) p(\expect{\Lum}, \expect{\Lum^2} \cond \bmy) \,.
+  p(\Lumtot \cond \bmy) &= \int \rmdx{\lambda} \rmdx{\mu} \rmdx{\sigma^2} \GaussianDist \left( \Lumtot \cond \lambda \mu, \lambda \sigma^2 \right) p(\lambda \cond \bmy) p( \mu, \sigma^2 \cond \bmy) \,,
 \end{align}
-\fred{$\int \rmdx{\expect{\Lum}}$ looks ugly, I'd rather have other
-  symbols, perhaps just use $\mu$ and $\sigma^2$. The transformation
-  is one-to-one anyways.}
+where $ p(\lambda \cond \bmy)$ is given by \refeq{pms} and $ p( \mu, \sigma^2 \cond \bmy)$ by \refeq{raf}--\refeq{rak}.
 \fred{Could integrate directly, or use mgf to get moments of $  p(\Lumtot \cond \bmy)$. Then do mode finding, describe initial values and Laplace approximation. }

### 2016-11-30 real/draft.tex
[draft] Finish asympt. section

@@
 \usepackage{hyperref}
-\synctex=1
+\usepackage{physics}
 
@@
 \newcommand{\NBD}{\mathrm{NB}}
-\newcommand{\var}{\mathrm{var}}
 
@@
 \newcommand{\lum}{\ell}
-\newcommand{\rmdx}[1]{\mbox{d} #1 \,} % differential
+% \newcommand{\rmdx}[1]{\mbox{d} #1 \,} % differential
+\newcommand{\rmdx}[1]{\dd{#1}} % differential
 \newcommand{\firstDeriv}[1]{\frac{\partial}{\partial #1}}
@@
 
-  Now all ingredients for the prediction of $\Lumtot$ in one bin are
+  Finally all ingredients for the prediction of $\Lumtot$ in one bin are
   ready, and \refeq{qaa} simplifies to
@@
 \end{align}
-where $ p(\lambda \cond \bmy)$ is given by \refeq{pms} and $ p( \mu, \sigma^2 \cond \bmy)$ by \refeq{raf}--\refeq{rak}.
-\fred{Could integrate directly, or use mgf to get moments of $  p(\Lumtot \cond \bmy)$. Then do mode finding, describe initial values and Laplace approximation. }
+where $ p(\lambda \cond \bmy)$ is given by \refeq{pms} and
+$ p( \mu, \sigma^2 \cond \bmy)$ by
+\refeq{raf}--\refeq{rak}.
+
+\subsection{Numerical implementation}\label{sec:asympt-numeric}
+
+Unfortunately we could not integrate \refeq{ral} analytically but
+since the integrand is unimodal and strongly peaked the Laplace or
+saddle-point approximation~\cite[Ch. 27]{mackay2003information} can be
+applied in two ways. First, if only the value
+$p(\Lumtot=\lumtot \cond \bmy)$ for some particular $q$ is sought,
+then one can approximate it directly. Second, we can compute the
+moments of $\Lumtot$ through the \emph{moment-generating
+  function}\fredmargin{ref}
+\begin{align}
+  \lleq{rap}
+  M(t \cond \bmy) = \int_0^{\infty} \rmdx{\Lumtot} p(\Lumtot \cond \bmy) e^{-t \Lumtot} \,
+\end{align}
+from which the moments are accessible by differentiation
+\begin{align}
+  \lleq{raq}
+  \expect{\Lumtot^k} = (-1)^k \eval{\dv[k]{M(t \cond \bmy)}{t}}_{t=0}
+\end{align}
+
+Assuming all involved integrals are absolutely convergent, we can
+exchange the order of integration and find
+\begin{align}
+  \lleq{rar}
+  M(t \cond \bmy) &= \int \rmdx{\lambda} \rmdx{\mu} \rmdx{\sigma^2} p(\lambda \cond \bmy) p( \mu, \sigma^2 \cond \bmy)  f(t \cond \lambda, \mu, \sigma^2),
+  \intertext{where }
+  f\left(t \cond  \lambda, \mu, \sigma^2\right) &\equiv \int \rmdx{\Lumtot} \GaussianDist \left( \Lumtot \cond \lambda \mu, \lambda \sigma^2 \right) e^{-t \Lumtot}= \frac{1}{2} e^{t \lambda (t \sigma^2/2 - \mu)} \left[ 1 + \erf\qty( \frac{\sqrt{\lambda}(\mu - t \sigma^2)}{\sqrt{2 \sigma^2}}) \right].
+\end{align}
+This implies that the mean and variance of $\Lumtot$ require an integral over the three parameters
+\begin{align}
+  \lleq{ras}
+  \expect{\Lumtot} &= \int \rmdx{\lambda} \rmdx{\mu} \rmdx{\sigma^2} p(\lambda \cond \bmy) p( \mu, \sigma^2 \cond \bmy) \qty(-\eval{\dv{f(t \cond  \lambda, \mu, \sigma^2)}{t}}_{t=0})\\
+  \variance{\Lumtot} &= \int \rmdx{\lambda} \rmdx{\mu} \rmdx{\sigma^2} p(\lambda \cond \bmy) p( \mu, \sigma^2 \cond \bmy) \qty(\eval{\dv[2]{f(t \cond  \lambda, \mu, \sigma^2)}{t}}_{t=0}) - \expect{\Lumtot}^2
+\end{align}
+We implemented the above formulae in the julia language~\cite{julia14}
+and used Newton's method from the \texttt{Optim.jl} package in
+combination with automatic
+differentiation~\cite{RevelsLubinPapamarkou2016} to optimize the
+respective integrand and compute the Hessian at the mode. As initial
+values, we slightly offset the sample estimates to avoid issues if the
+optimizer cannot improve on the initial guess and found convergence
+within $\order{10}$ steps:
+\begin{align}
+  \lleq{rat}
+  \qty(\lambda, \mu, \sigma^2) = 1.0005 (n, \expectest{\lum}, \widehat{\variance{\lum}}).
+\end{align}
+\fred{Did we define expectation value and variance? Should be in notation or intro section}
+\fred{At least for us, I should do a comparison plot to evaluate $p(\Lumtot \cond \bmy)$ with both methods }
 

### 2016-11-30 real/code/Replica.jl
[Replica] add doc strings, fix test case

@@ -87,8 +87,14 @@ function read_virtual(nsim; νmin=1.00677e15, νmax=1.023018e15, cutoff=1e-20)
     n, x, means, sumsqdiff
 end
 
+"""
+Enumeration to specify which target function to create
+"""
 @enum TARGET Posterior FirstMoment SecondMoment
 
+"""
+Create a function for Laplace approximation
+"""
 function targetfactory(target::TARGET, n::Int64, xmean::Float64, xsumsq::Float64, a::Float64=1)
     0 <= a <= 1 || error("need a in [0,1]")
 
@@ -126,8 +132,9 @@ function targetfactory(target::TARGET, n::Int64, xmean::Float64, xsumsq::Float64
         log(1/2 * tmp)
     end
 
+    # optimizer minimizes => add minus sign at the last possible moment
     if target === Posterior
-        return -logposterior
+        return x -> -logposterior(x)
     elseif target === FirstMoment
         return x -> -(logposterior(x) + logfirst(x))
     elseif target === SecondMoment
@@ -139,8 +146,9 @@ end
 function test()
     n, sums, means, sumsqdiff = read_real(10)
     i = 10
-    logposterior=logposterior_factory(n[i], means[i], sumsqdiff[i])
-    target(x) = -logposterior(x)
+    a = 1.0
+    target = targetfactory(Posterior, n[i], means[i], sumsqdiff[i], a)
+    # target(x) = -logposterior(x)
     x = [501.0,1.406,0.001]
     res = optimize(target, x, Newton(), OptimizationOptions(autodiff = true))
     # compare to mathematica
@@ -148,22 +156,27 @@ function test()
 
     H = zeros((3,3))
     # measured, chunk size of 3 is twice as fast as 1 or 2
-    ForwardDiff.hessian!(H,logposterior, φ, chunk)
+    ForwardDiff.hessian!(H,target, Optim.minimizer(res), chunk)
+
+    println("Posterior")
+    println(res)
 
     # need a really good starting point, else run into
     # ERROR: DomainError:
     # in logposterior at none:12
 
     # minimize the negative to maximize
-    target(x) = -logposterior(x) - logfirst(x)
+    target = targetfactory(FirstMoment, n[i], means[i], sumsqdiff[i], a)
[…]

### 2016-11-30 real/draft.tex
[draft] Introductory paragraphs edited, still in progress

@@
 %% CHANGE THIS DATE AS YOU MODIFY THE FILE
-\newcommand{\vdate}{160510}
+\newcommand{\vdate}{161130}
 \pagestyle{myheadings}
 \markboth{Version \vdate} {Version \vdate}
-\parindent=0mm
-\def\pp{\hfill\par \vspace{\baselineskip}}
-%% \def\pp{}
+
+\parindent=3ex
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \renewcommand{\lleq}[1]{\label{#1} {\scriptstyle {\rm (#1)}} \hspace*{2ex} }
-% \renewcommand{\labelenumi}{\arabic{section}.\arabic{subsection}.\arabic{enumi}}
-% \renewcommand{\thesubsubsection}{\arabic{subsubsection}}
-
-\newcounter{mnumi}
-\newenvironment{mnumerate}{
-  \begin{list}{\arabic{mnumi}.}
-    {\usecounter{mnumi} \setlength{\itemsep}{0pt}
-      \setlength{\leftmargin}{3ex}
-    }
-  }
-  {\end{list}}
-
-\newenvironment{mtemize}{
-  \begin{list}{$\bullet$}
-    {\setlength{\itemsep}{0pt}
-     \setlength{\leftmargin}{3ex}
-    }
-  }
-  {\end{list}}
 
@@
 \subsection{Bayesian prediction}
+\label{sec:bayp}
 
@@
 %
-A particular experiment or simulation generates \textit{data} $D$
+A particular experiment or simulation generates
+\textit{data}
 which will often come in the form of a list of $n$ measurements,
-$D = \bmy = (y_1,y_2,\ldots,y_n)$. In order to understand or explain
-this data, the observer constructs one or more \textit{models} or
+$\bmy = (y_1,y_2,\ldots,y_n)$. In order to understand or explain this
+data, the observer constructs one or more \textit{models} or
 \textit{hypotheses} $\hypo$. Central to this construction is the
@@
 where $\bmtheta = (\theta_1,\theta_2,\ldots,\theta_K)$ is the set of
-unknown parameters specified by the model. If and when the individual
-measurements $y_i$ are independent, the joint likelihood is the
-product
+unknown parameters specified by the model. The likelihood is the
+probability assigned by the model $\hypo$ to the data $\bmy$ when the
+parameters $\bmtheta$ are fixed and known.
+% 
+If and when the individual measurements $y_i$ are independent, the
+joint likelihood is the product
 \begin{align}
+  \lleq{bta}
   p(\bmy\cond \bmtheta) &= \prod_{i=1}^n p(y_i\cond \bmtheta).
 \end{align}
-At this level of analysis, the goal
-is to find which values for these parameters provide a satisfactory
-description of the data. In the Bayesian context, these parameter
-values are best described in terms of the \textit{posterior}, which
-is related to the likelihood by
+At this level of analysis, the goal is to find which values for these
+parameters provide a satisfactory description of the data. In the
+Bayesian context, these parameter values are best described in terms
+of the probability of the parameters given the data,
+$p(\bmtheta\cond \bmy)$. This so-called \textit{posterior} is related
+to the likelihood by Bayes' Theorem,
 \begin{align}
@@
 \begin{align}
+  \lleq{bti}
   p(\bmy) &= \int d\bmtheta'\, p(\bmy\cond\bmtheta')\;p(\bmtheta'),
@@
 is fully determined by the likelihood and prior. 
-
-The scope of this paper is limited to the relationship between
+% 
+Prediction of any ``future'' data $Y$ is made not by considering the
+likelihood $p(Y\cond\bmtheta_{\rm max})$ using single best-fit
+parameter values $\bmtheta_{\rm max}$, but by integrating over all
+possible values of the $\bmtheta$ in terms of the posterior
+$p(\bmtheta\cond n,\bml)$,
+\begin{align}
+  \lleq{btj}
+  p(Y\cond \bmy) &= \int d\bmtheta\,p(Y\cond\bmtheta)\,p(\bmtheta\cond \bmy).
+\end{align}
+The above relations are true for any type of data and can be applied
+to actual experimental data and the pertinent physical parameters. The
+scope of this paper, however, is limited to the relationship between
 simulated data generated by radiative transfer simulation codes.
-Likewise, all parameters introduced and calculated relate strictly to
-model descriptions of the simulation.
-%
-While Bayesian concepts and methods can be applied to actual
-experimental data and related physics parameters, we shall not do so
-here.
-
-\hanslong{
-Note that if the prior is uniform, it is independent of $\theta$ and
-will cancel,
-\begin{align}
-  \lleq{bthu}
-  p(\theta\cond \bmy)
-  &= \frac{p(\bmy\cond\theta)} {\int d\theta'\, p(\bmy\cond\theta')}.
-\end{align}
-}
-
-\hanslong{Bayesian prediction?}
+Likewise, all parameters considered below relate strictly to model
+descriptions of the simulation.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{TARDIS}
-\fred{I don't think TARDIS is a good title, too specific}
+\section{From single-simulation data to many-simulation prediction}
 
-We consider a computer simulation which, based on tunable input
-parameters $\bmtheta$, generates $\npack$ photon packets, each with a
-particular frequency $\nu_i$ and luminosity $\ell_i$. Together, these
-frequencies and luminosities form the simulation data
-$\bmy = \{ (\nu_i,\ell_i): i = 1,\ldots,\npack\}$.  We divide up the
-frequency spectrum into bins
+% \subsection{Output data from radiative transfer simulations}
+\subsection{Simulation data}
+
+We consider a single computer simulation which generates a luminosity
+spectrum from $\npack$ photon packets, each with a particular
+frequency $\nu_i$ and luminosity $\lum_i$. Together, these frequencies
+and luminosities form the raw simulation data
+$\bmy = \{ (\nu_i,\lum_i): i = 1,\ldots,\npack\}$.  We divide up the
+frequency spectrum into bins of width $\Delta\nu$ centered on
+frequencies $\nu_b$ resulting in bin intervals
 $\Delta_b = [\nu_b {-} \tfrac{1}{2} \Delta\nu \,,\, \nu_b {+}
-\tfrac{1}{2} \Delta\nu], b = 1,2,\ldots, B$ of width $\Delta\nu$
-centered on frequencies $\nu_b$. Let $n_b$ be the number of packets
-having a frequency falling into a particular bin $b$, and let $\bml_b$
-be the set of luminosities of those $n_b$ packets. The total
-luminosity in bin $b$ is the sum of luminosities of those packets
-whose frequency falls into $b$,
+\tfrac{1}{2} \Delta\nu], b = 1,2,\ldots, B$. Let $n_b$ be the number
+of packets having a frequency falling into a particular bin $b$, and
+let the vector $\bml_b = \{(\ell_{b,j}): j = 1,\ldots,n_b\}$ be the
+set of luminosities of those packets. The data set is correspondingly
+cast into the more compact histogram form
 \begin{align}
-  \ell_b = \sum_{i=1}^\npack \bm{1}_{\Delta_b}(\nu_i)\,\ell_i
+  \lleq{sdb}
+  \bmy = \{(n_b,\bml_b): b = 1,\ldots,B\}
 \end{align}
-where the indicator function
-$\bm{1}_{[a,b]}(x) = 1$ when $x$ is in the interval
-$[a,b]$ and 0 otherwise.
+with loss of information limited to replacing the exact packet
+frequencies $\nu_i$ by the appropriate $\nu_b$. The total luminosity
+in bin $b$ is the sum of luminosities of those packets whose frequency
+falls into $\Delta_b$,
+\begin{align}
+  \lleq{sdc}
+  \lumtot_b 
+  % &= \sum_{i=1}^\npack \bm{1}_{\Delta_b}(\nu_i)\,\ell_i
+  &= \sum_{j=1}^{n_b} \lum_{b,j}.
+\end{align}
+% where the indicator function $\bm{1}_{[a,b]}(x) = 1$ when $x$ is in
+% the interval $[a,b]$ and 0 otherwise.
 
-\section{Notation} \label{sec:notation}
+\subsection{Prediction of luminosity spectrum}
 
-\fred{Merge this section with previous section}
+Within a given model $\hypo$, the data $\bmy$ is only one of many
+possible realisations of the corresponding random variables; for
+example, starting the simulation with a different seed would result in
+different output data. We denote the general random variables by the
+corresponding upper-case symbols,
+\begin{align}
+  \bmL_b &= \{(\Lum_{b,j}): j = 1,\ldots,N_b\}\\
+  \bmY &= \{(N_b, \bmL_b): b = 1,\ldots,B\},\\
+  \Lumtot_b &= \sum_{j=1}^{N_b} \Lum_{b,j},
+\end{align}
+with corresponding likelihood $p(\bmY\cond\bmtheta)$ and posterior
+$p(\bmtheta\cond\bmY)$ for any $\bmY$. The lower-case $(n_b,\bml_b)$
+are strictly reserved for the simulated data at hand.
 
-We consider a computer simulation that, given input parameters
-$\bmtheta$ and a random-number seed, produces a set of simulated data
-$ S = \{ (\nu_i, \ell_i): i=1\dots\npack \} = (\bmnu, \bml)$
-consisting of $\npack$ packets, each with a frequency and a
-luminosity. We distinguish between a random variable $L$ and its
-concrete realization $\ell$.  To learn about $\bmtheta$, we need to
-compare to data from a telescope available as luminosities in
-frequency bins, $D = \{ ( \nu_b, \ell_b): b=1 \dots B\}$. The natural
-estimate from a single simulation is to add up the luminosities of all
-packets in a bin to estimate $\Lumtot_b$.  With a different seed but
-identical $\bmtheta$, both the number of packets in the bin and their
-luminosities are usually different. Our goal is to capture what is
-known about $\Lumtot$ based on a \emph{single} simulation without recourse
-to expensive repetitions or asymptotic approximations; i.e., we seek
-$P(\Lumtot_b \cond \nu_b, \npack, \bmnu, \bml )$. In an actual physics
-analysis, one would proceed to plug in the observed value $\ell_b$ and
-perform an analysis to learn about $\bmtheta$. But here we focus on
-the specific problem of $P(\Lumtot_b \cond \nu_b, \npack, \bmnu, \bml )$ and
-hence omit $\bmtheta$ throughout.
+The goal of this paper is to illustrate how the single-simulation data
+$\bmy$ can be used to predict the distribution of total bin
+luminosities $Q_b, b = 1,\ldots,B$ if many simulations had been
+carried out.
 
-Various features, for example absorption lines, usually make it
-difficult to explicitly model the spectrum---the (total) luminosity as
-a function of frequency. In a binned analysis, we can avoid these
+Various features such as absorption lines make the luminosity spectrum
+highly variable. The model we develop below takes into account this
+variability in a two-step process. First, the number of packets $N_b$
+is modelled by means of a separate Poisson distribution
+$p(N_b\cond\lambda_b)$ for each bin, where the parameter $\lambda_b$
+will capture most of the variability of the spectrum.  Second, the
+single-packet luminosity distribution is modelled by means of a packet
+likelihood for $\Lum_{b,j}$ with packet parameters, a packet prior and,
+of course, packet parameter posterior. The packet parameters, which
+depend on the frequency bin, will be calculated in two ways: first by
+drawing them from a ``parametrisation of the parameters'' across all
+bins as shown in \refsec{example} and secondly by assuming that
+\refsec{asymptotic}
+
+Our approach
+is to predict each bin's luminosity $Q_b$ separately using the same
+model but allowing its parameters to vary independently in each bin.
+Connections between them
+
+function of frequency. In a binned analysis, we can avoid these
 difficulties associated to modeling $n_b$. Our approach, however,
@@
 is this crucial piece of information that allows us to extend previous
-methods \fred{Refs: Zech, maybe others, too} in dealing with the
-extreme cases when the number of packets in a bin, $n_b$, is one or
-even zero.
+methods in dealing with the extreme cases when the number of packets
+in a bin, $n_b$, is one or even zero.
+
+To keep it simple, we shall assume that all bins are independent,
+meaning that
+\begin{align}
+  p(Q_1,Q_2,\ldots,Q_B\cond\bmy) = \prod_{b=1}^B p(Q_b\cond n_b,\bml_b).
+\end{align}
+Below, we calculate $p(Q_b\cond n_b,\bml_b)$ for fixed $b$ and hence
+drop the subscript, i.e.\ $Q_b\to Q, n_b \to n, N_b\to N$ etc.
+
+%%%%%%%%%%%%%%%%%%%
+\fred{Refs: Zech, maybe others, too}
+
+In terms of the generic notation of Section \ref{sec:bayp}, Bayesian
+prediction of the distribution of the bin's luminosity
+$Q = \sum_{j=1}^n L_j$ given the bin's data $(n,\bml)$ is obtained by
+considering not only the likelihood $p(Q\cond\bmtheta_{\rm max})$
+using best-fit parameter values $\bmtheta_{\rm max}$, but by
+integrating over all possible values of the $\bmtheta$ in terms of the
+posterior $p(\bmtheta\cond n,\bml)$,
+\begin{align}
+  p(Q\cond n,\bml) &= \int d\bmtheta\,p(Q\cond\bmtheta)\,p(\bmtheta\cond n,\bml).
+\end{align}
+The model's parameters $\bmtheta$ and calculation of this prediction
+will be set out in detail below.
 

### 2016-11-30 real/code/optim.jl
[code] Add gamma
Update code to renormalize Gamma distribution

@@ -16,181 +16,186 @@ function problem(;run=11, npackets=typemax(Int64), αOrder=2, βOrder=1)
     return frame, targetfactory(frame, αOrder, βOrder)
 end
 
-function nlopt_factory(f, ∇f!)
-    function(x::Vector, grad::Vector)
+nlopt_factory(f, ∇f!) = function(x::Vector, grad::Vector)
+    if length(grad) > 0
+        ∇f!(x, grad)
+    end
+    f(x)
+end
+
+function trynlopt()
+    αOrder = 1
+    βOrder = 1
+    dim = αOrder + βOrder
+    frame, (f, ∇f!, Hf!) = problem(αOrder=αOrder, βOrder=βOrder, run=99)
+
+    ## histogram(frame[:energies], normalized=true)
+    ## α = 1.331; β=58.3
+    ## plot!(x -> exp(loggamma(x, α, β)), linspace(0, 0.2, 500))
+    ## Plots.pdf("optim.pdf")
+    ## return
+
+    myfunc = nlopt_factory(f, ∇f!)
+
+    """
+    Prior constraint for α > 1 and β > 0 in the form g(θ) ≤ 0
+    """
+    function myconstraint(x::Vector, grad::Vector, frame::DataFrame, p::Poly,
+                          minvalue::Float64, offset::Int64)
+        ν::Vector{Float64} = frame[:nus]
+        z::Float64 = 0
+        zmin::Float64 = oftype(1.0, Inf)
+        imin::Int64 = 0
+
+        for i in eachindex(ν)
+            @inbounds z = polyval(p, ν[i])
+            if z < zmin
+                zmin = z
+                imin = i
+            end
+
+        end
         if length(grad) > 0
-            ∇f!(x, grad)
-        end
-        f(x)
-    end
-    end
-
-    function trynlopt()
-        αOrder = 1
-        βOrder = 1
-        dim = αOrder + βOrder
-        frame, (f, ∇f!, Hf!) = problem(αOrder=αOrder, βOrder=βOrder, run=99)
-
-        ## histogram(frame[:energies], normalized=true)
-        ## α = 1.331; β=58.3
-        ## plot!(x -> exp(loggamma(x, α, β)), linspace(0, 0.2, 500))
-        ## Plots.pdf("optim.pdf")
-        ## return
-
-        myfunc = nlopt_factory(f, ∇f!)
-
-        """
-        Prior constraint for α > 1 and β > 0 in the form g(θ) ≤ 0
-        """
[…]

### 2016-11-30 real/code/tardis.jl

@@ -1,9 +1,7 @@
 module tardis
 
-using DataFrames
-using HDF5
-using Optim
-using Polynomials
+using DataFrames, HDF5, Optim, Polynomials
+import GSL, StatsFuns
 
 export allocations, filter_positive, laplace, loggamma, lognegativebinomial, NegBinom, νpower, PosteriorMean, readdata, symmetrize!, targetfactory, transform_data!, update_polynomials!
 
@@ -61,7 +59,25 @@ function transform_data!(frame::DataFrame, nuscale::Float64=1e15)
 end
 
 function loggamma(x::Float64, α::Float64, β::Float64)
-    α * log(β) - lgamma(α) + (α-1)*log(x) - β * x
+    if isnan(α) || isnan(β) === NaN
+        return -Inf
+    end
+    if α >= 171
+        # x ∈ [0, ∞]
+        return α * log(β) - lgamma(α) + (α-1)*log(x) - β * x
+    end
+    # renormalize to a Gamma distribution for x ∈ [0, 1]
+    gammainc = 0.0
+    try
+        gammainc = GSL.sf_gamma_inc(α, β)
+    catch err
+        println("GSL problem at α = $α, β = $β")
+        throw( err)
+    end
+    if gamma(α)-gammainc < 0.0
+        return -Inf
+    end
+    res = α * log(β) - log(gamma(α)-gammainc) + (α-1)*log(x) - β * x
 end
 
 function lognegativebinomial(N::Real, n::Integer, a::Real)

### 2016-11-30 real/code/tst.jl

@@ -27,11 +27,11 @@ function main()
 
     return
 
-    # nb = NegBinom(5, 0.5)
-    # Pall, ∇Pall!, HPall! = targetfactory(frame, αOrder, βOrder, evidence=evidence, pm=pm, nb=nb)
-    # @time maxPall, mode, ret = run_nlopt(frame, Pall, ∇Pall!, HPall!, αOrder, βOrder,
-    #                                       init=vcat(mode, 10), xtol_rel=1e-8, nb=nb)
-    # println("max of normalized all posterior = $maxPall at $mode")
+    nb = NegBinom(5, 0.5)
+    Pall, ∇Pall!, HPall! = targetfactory(frame, αOrder, βOrder, evidence=evidence, pm=pm, nb=nb)
+    @time maxPall, mode, ret = run_nlopt(frame, Pall, ∇Pall!, HPall!, αOrder, βOrder,
+                                          init=vcat(mode, 10), xtol_rel=1e-8, nb=nb)
+    println("max of normalized all posterior = $maxPall at $mode")
     # return
 
     points = collect(linspace(0.0, 0.6, 25))

### 2016-12-01 real/code/Replica.jl
[code] Add gamma null hypothesis, many fixes
* remove loggamma fixed to x in [0,1]
* joint plot to compare Gamma and asymptotic approx.

@@ -21,16 +21,15 @@ function plotprediction(sums, predict, stddev, i=1)
     plot!(x->Distributions.pdf(dist, x), linspace(predict[i] - 4*stddev[i], predict[i] + 4*stddev[i]), lab="prediction from replica $i", line=:black)
     Plots.pdf("replica-prediction.pdf")
     # replot to show it interactively
+    info("Replica $i: $(sums[i]) vs $(predict[i]) ± $(stddev[i])")
     plot!()
 end
 
-function read_real(nsim; npackets=typemax(Int64))
+function read_real(nsim; npackets=typemax(Int64), νmin=1.00677, νmax=1.023018)
     n = Array(Int64, (nsim,))
     x = Array(Float64, (nsim,))
     means = zeros(x)
     sumsqdiff = zeros(x)
-
-    νmin, νmax = 1.00677, 1.023018
 
     function scale_data!(frame::DataFrame, nuscale::Float64=1e15, xscale::Float64=1e38)
         sort!(frame, cols=:nus)
@@ -44,14 +43,19 @@ function read_real(nsim; npackets=typemax(Int64))
     function read_run(run)
         raw_data = readdata(run; npackets=npackets)
         frame = filter_positive(raw_data...)
-        scale_data!(frame)
+        # scale_data!(frame)
+        transform_data!(frame)
         nmin=searchsortedfirst(frame[:nus], νmin)
         nmax=searchsortedlast(frame[:nus], νmax)
         n[run] = nmax - nmin + 1
         x[run] = sum(frame[:energies][nmin:nmax])
-        means[run] = mean(frame[:energies][nmin:nmax])
+        means[run] = mean(frame[:energies][nmin:nmax]) # 1/n[run] * x[run]
         # remove Bessel's correction
         sumsqdiff[run] = (n[run]-1) * var(frame[:energies][nmin:nmax])
+        # if run == 1
+        #     info("first run: n=$(n[1]), sum=$(x[1]), mean=$(means[1]), sum^2=$(sumsqdiff[1])")
+        #     println(frame[:energies][nmin:nmax])
+        # end
     end
 
     for i in 1:nsim
@@ -61,24 +65,26 @@ function read_real(nsim; npackets=typemax(Int64))
     n, x, means, sumsqdiff
 end
 
-function read_virtual(nsim; νmin=1.00677e15, νmax=1.023018e15, cutoff=1e-20)
[…]

### 2016-12-01 real/code/optim.jl

@@ -23,95 +23,6 @@ nlopt_factory(f, ∇f!) = function(x::Vector, grad::Vector)
     f(x)
 end
 
-function trynlopt()
-    αOrder = 1
-    βOrder = 1
-    dim = αOrder + βOrder
-    frame, (f, ∇f!, Hf!) = problem(αOrder=αOrder, βOrder=βOrder, run=99)
-
-    ## histogram(frame[:energies], normalized=true)
-    ## α = 1.331; β=58.3
-    ## plot!(x -> exp(loggamma(x, α, β)), linspace(0, 0.2, 500))
-    ## Plots.pdf("optim.pdf")
-    ## return
-
-    myfunc = nlopt_factory(f, ∇f!)
-
-    """
-    Prior constraint for α > 1 and β > 0 in the form g(θ) ≤ 0
-    """
-    function myconstraint(x::Vector, grad::Vector, frame::DataFrame, p::Poly,
-                          minvalue::Float64, offset::Int64)
-        ν::Vector{Float64} = frame[:nus]
-        z::Float64 = 0
-        zmin::Float64 = oftype(1.0, Inf)
-        imin::Int64 = 0
-
-        for i in eachindex(ν)
-            @inbounds z = polyval(p, ν[i])
-            if z < zmin
-                zmin = z
-                imin = i
-            end
-
-        end
-        if length(grad) > 0
-            for m in 1:length(p)
-                grad[offset + m] = -ν[imin]^(m-1)
-            end
-        end
-        return minvalue - zmin
-    end
-
-    # GN_DIRECT_L, GN_ORIG_DIRECT_L
-    # LN_BOBYQA, LN_COBYLA
-    # LD_LBFGS, LD_MMA
-    opt = Opt(:LN_COBYLA, dim)
-    lowerbounds = zeros(dim)
-    upperbounds = zeros(dim)
-    lowerbounds[1] = 1.0 # α₀
-    lowerbounds[2:end] = -50.0
-    lowerbounds[αOrder + 1] = 0.0 # β₀
-    upperbounds = -1 * lowerbounds
-    upperbounds[1] = 4
-    upperbounds[αOrder + 1] = 200
-    ## lower_bounds!(opt, [1.0, 0.0])
-    ## upper_bounds!(opt, [4.0, 100.0])
-    xtol_rel!(opt, 1e-5)
-    # ftol_rel!(opt, 1e-7)
-    ## maxeval!(opt, 2500)
-    max_objective!(opt, myfunc)
-
[…]

### 2016-12-01 real/code/tardis.jl

@@ -50,7 +50,7 @@ function transform_data!(frame::DataFrame, nuscale::Float64=1e15)
         en[i] = 1.0 - en[i] / maxen
     end
 
-    frame[:nus] /= nuscale
+    scale!(frame[:nus], 1/nuscale)
 
     # println("min nu = $(frame[:nus][1]), max  max en = $(maximum(en))")
 
@@ -62,22 +62,31 @@ function loggamma(x::Float64, α::Float64, β::Float64)
     if isnan(α) || isnan(β) === NaN
         return -Inf
     end
-    if α >= 171
-        # x ∈ [0, ∞]
-        return α * log(β) - lgamma(α) + (α-1)*log(x) - β * x
-    end
-    # renormalize to a Gamma distribution for x ∈ [0, 1]
-    gammainc = 0.0
-    try
-        gammainc = GSL.sf_gamma_inc(α, β)
-    catch err
-        println("GSL problem at α = $α, β = $β")
-        throw( err)
-    end
-    if gamma(α)-gammainc < 0.0
-        return -Inf
-    end
-    res = α * log(β) - log(gamma(α)-gammainc) + (α-1)*log(x) - β * x
+    return α * log(β) - lgamma(α) + (α-1)*log(x) - β * x
+
+    # # ugly modifications to allow limiting range of GammaDist x ∈ [0,1]
+    # if α >= 171
+    #     # x ∈ [0, ∞]
+    #     return α * log(β) - lgamma(α) + (α-1)*log(x) - β * x
+    # end
+    # # renormalize to a Gamma distribution for x ∈ [0, 1]
+    # gammainc = 0.0
+    # try
+    #     gammainc = GSL.sf_gamma_inc(α, β)
+    # catch err
+    #     println("GSL problem at α = $α, β = $β")
+    #     throw( err)
+    # end
+    # # avoid Domain error with negative and divergence with == 0
+    # # because 8.54e202 - 8.64e202=0 with double precision
+    # if gamma(α)-gammainc <= 0.0
+    #     return -Inf
+    # end
+    # res = α * log(β) - log(gamma(α)-gammainc) + (α-1)*log(x) - β * x
+    # if res == Inf
+    #     error("loggamma diverges to ∞")
+    # end
+    # return res
 end
 
 function lognegativebinomial(N::Real, n::Integer, a::Real)
@@ -201,6 +210,7 @@ function targetfactory(frame::DataFrame, αOrder::Int64, βOrder::Int64;
                 end
 
[…]

### 2016-12-02 real/draft.tex
[draft] Add two notes
[draft] add todo notes

@@
 
+\section{What's left to do}
+
+\subsection{why asymptotic approx. has larger variance}
+\begin{enumerate}
+  \item For $\alpha=1.5, \beta=60, n=14000$ still get 50 \% larger uncertainty from asymptotic compared to summation over $N$.
+Is it due very asymmetric one-packet distribution?
+Try larger $\alpha$, for example $\alpha = 30$.
+  \item  Another option: $\int \rmdx{\alpha} \rmdx{\beta}$, use true values in simulation. Where does uncertainty come from? From $\lambda$ or $\mu, \sigma^2$?
+  \item Are finite number of terms in $\sum_N$ a problem? Check
+\end{enumerate}
+
+\subsection{Plots}
+
+\begin{enumerate}
+  \item generate a toy spectrum, take frequencies from actual tardis,
+  generate luminosities from a $\GammaDist(\alpha, \beta)$, no $\nu$
+  dependence of $\alpha, \beta$ for clarity , plot spectrum with
+  uncertainty bands, show it early in the paper, then describe where
+  it comes from
+  \item compare asymptotic and $\sum_N$ $p(\Lumtot \cond \bmy)$ on a
+  plot, add what happens if only $\sqrt{n}$ uncertainty is used: that
+  would neglect uncertainty on the single-packet distribution, show
+  case when parameters determined from a single bin with few samples,
+  should make large difference. Fix $\lambda$, what are the uncertainties now?
+  \item Statistics of replicas: again compare \refeq{spe} to \refeq{ral}. Teach a lesson: from a single data set, prediction will be centered there
+  \item Question: does Wolfgang want to see a plot in which ``$\lambda$ is divided out''? That is , just plot how mean
+\end{enumerate}
+
+\subsection{tardis}
+
+move tardis section into separate document in this git repo.
+Better to keep it out of the paper.
+Cite tardis when showing the toy spectrum.
+Mention that in reality $\alpha$ and $\beta$ are $\nu$ dependent
+
 \section{Introduction (Wolfgang)}
@@
 \fred{At least for us, I should do a comparison plot to evaluate $p(\Lumtot \cond \bmy)$ with both methods }
+\fred{We need a plot, the one to compare predictions and 250 real replicas or 100 virtual replicas?}
+\fred{Advantage of this approach: no need to store all packets anymore, sum and sum of squares can be accumulated online $\Rightarrow$ reduced computation and storage requirements}
 

### 2016-12-02 real/code/Asymptotic.jl
[code] move stuff into asymptotic.jl

@@ -1,0 +1,124 @@
+module Asymptotic
+
+using tardis, Optim
+import ConjugatePriors
+import ForwardDiff
+chunk = ForwardDiff.Chunk{3}()
+
+
+"""
+Enumeration to specify which target function to create
+"""
+@enum TARGET Posterior FirstMoment SecondMoment ThirdMoment FourthMoment
+
+"""
+Create a function for Laplace approximation
+"""
+function targetfactory(target::TARGET, n::Int64, xmean::Float64, xsumsq::Float64, a::Float64=1)
+    0 <= a <= 1 || error("need a in [0,1]")
+    (xmean > 0) || error("nonpositive xmean $xmean")
+    (xsumsq > 0) || error("nonpositive xsumsq $xsumsq")
+
+    function logposterior(x::Vector)
+        λ, μ, σSq = x
+
+        # gamma(λ), set β=1
+        α = n-a+1
+        resgamma = - lgamma(α) + (α-1)*log(λ) - λ
+        ## println(resgamma)
+
+        # normal-inversegamma
+        d = ConjugatePriors.NormalInverseGamma(xmean, n, n/2, xsumsq/2)
+
+        resnormal = -1/2 * (log(2*π*σSq / d.v0) + (μ-d.mu)^2 * d.v0 / σSq)
+        resinvgamma = -lgamma(d.shape) + d.shape * log(d.scale) - (d.shape + 1) * log(σSq) - d.scale / σSq
+
+        ## println(resinvgamma," ", resnormal)
+        return resgamma + resnormal + resinvgamma
+    end
+
+    """log(-d/dt f(target=0|x))"""
+    function logfirst(x::Vector)
+        λ, μ, σSq = x
+        exponent = λ*μ^2 / (2*σSq)
+        tmp = exp(-exponent) * sqrt(λ * σSq / (2 * π)) + 1/2 * λ * μ * (1 + erf(sqrt(exponent)))
+        log(tmp)
+    end
+
+    """log(d²/dt² f(target=0|x))"""
+    function logsecond(x::Vector)
+        λ, μ, σSq = x
+        exponent = λ*μ^2 / (2*σSq)
+        tmp = exp(-exponent) * sqrt(2 / π * λ^3 * σSq) * μ + λ*(λ*μ^2 + σSq)*(1 + erf(sqrt(exponent)))
+        log(1/2 * tmp)
+    end
+
+    """log(-d^3/dt^3 f(target=0|x))"""
+    function logthird(x::Vector)
+        λ, μ, σSq = x
[…]

### 2016-12-02 real/code/Replica.jl

@@ -1,15 +1,13 @@
 module Replica
 
 using tardis
+using Asymptotic
 using Base.Test
 import ConjugatePriors
 import Distributions
 using DataFrames
 using HDF5
-using Optim
 using Plots
-import ForwardDiff
-chunk = ForwardDiff.Chunk{3}()
 
 """
 
@@ -91,121 +89,6 @@ function read_virtual(nsim; νmin::Float64=1.00677, νmax::Float64=1.023018, νs
         end
     end
     n, x, means, sumsqdiff
-end
-
-"""
-Enumeration to specify which target function to create
-"""
-@enum TARGET Posterior FirstMoment SecondMoment ThirdMoment FourthMoment
-
-"""
-Create a function for Laplace approximation
-"""
-function targetfactory(target::TARGET, n::Int64, xmean::Float64, xsumsq::Float64, a::Float64=1)
-    0 <= a <= 1 || error("need a in [0,1]")
-    (xmean > 0) || error("nonpositive xmean $xmean")
-    (xsumsq > 0) || error("nonpositive xsumsq $xsumsq")
-
-    function logposterior(x::Vector)
-        λ, μ, σSq = x
-
-        # gamma(λ), set β=1
-        α = n-a+1
-        resgamma = - lgamma(α) + (α-1)*log(λ) - λ
-        ## println(resgamma)
-
-        # normal-inversegamma
-        d = ConjugatePriors.NormalInverseGamma(xmean, n, n/2, xsumsq/2)
-
-        resnormal = -1/2 * (log(2*π*σSq / d.v0) + (μ-d.mu)^2 * d.v0 / σSq)
-        resinvgamma = -lgamma(d.shape) + d.shape * log(d.scale) - (d.shape + 1) * log(σSq) - d.scale / σSq
-
-        ## println(resinvgamma," ", resnormal)
-        return resgamma + resnormal + resinvgamma
-    end
-
-    """log(-d/dt f(target=0|x))"""
-    function logfirst(x::Vector)
-        λ, μ, σSq = x
-        exponent = λ*μ^2 / (2*σSq)
-        tmp = exp(-exponent) * sqrt(λ * σSq / (2 * π)) + 1/2 * λ * μ * (1 + erf(sqrt(exponent)))
-        log(tmp)
-    end
-
-    """log(d²/dt² f(target=0|x))"""
-    function logsecond(x::Vector)
-        λ, μ, σSq = x
[…]

### 2016-12-02 real/code/tardis.jl

@@ -1,9 +1,10 @@
 module tardis
 
 using DataFrames, HDF5, Optim, Polynomials
-import GSL, StatsFuns
+import PyCall
+# import GSL, StatsFuns
 
-export allocations, filter_positive, laplace, loggamma, lognegativebinomial, NegBinom, νpower, PosteriorMean, readdata, symmetrize!, targetfactory, transform_data!, update_polynomials!
+export allocations, filter_positive, laplace, loggamma, lognegativebinomial, NegBinom, νpower, PosteriorMean, readdata, simpson, symmetrize!, targetfactory, transform_data!, update_polynomials!
 
 """
 read energies and frequencies from a particular run. Limit to read in at most npackets
@@ -59,7 +60,7 @@ function transform_data!(frame::DataFrame, nuscale::Float64=1e15)
 end
 
 function loggamma(x::Float64, α::Float64, β::Float64)
-    if isnan(α) || isnan(β) === NaN
+    if isnan(α) || isnan(β)
         return -Inf
     end
     return α * log(β) - lgamma(α) + (α-1)*log(x) - β * x
@@ -426,5 +427,20 @@ the mode, -Hf(θ). Both f and Hf are on the log scale.
 """
         laplace(logf::Real, log_det_H::Real, dim::Integer) = logf + dim/2 * log(2pi) - 1/2*log_det_H
         laplace(logf::Real, H::Matrix) = laplace(logf, log(det(H)), size(H)[1])
+
+
+"""
+Use Simpson's rule to evalue the integral of a function with
+`values` evaluated at `points`, and compute the mean and standard deviation.
+
+"""
+
+            function simpson(points, values)
+                PyCall.@pyimport scipy.integrate as si
+                normalization = si.simps(values, points)
+                μ = si.simps(values .* points, points)
+                σ2 = si.simps(values .* (points - μ).^2, points)
+                normalization, μ, sqrt(σ2)
+            end
 
     end # module

### 2016-12-02 real/draft.tex
draft.tex extensively rewritten
This rewrite accommodate the double limits for the same expansion of
p(Q|data). This draft is as yet incomplete and must still be finished in
Section 5.

@@
 %% CHANGE THIS DATE AS YOU MODIFY THE FILE
-\newcommand{\vdate}{161130}
+\newcommand{\vdate}{161202}
 \pagestyle{myheadings}
@@
 
-\item Traditional methods using conventional counting ratio
-  (``relative frequency'') can also make predictions for such
+\item Traditional methods using conventional counting ratios
+  (``relative frequencies'') can also make predictions for such
   uncertainties, but only if the number of packets is large. By
@@
 Prediction of any ``future'' data $Y$ is made not by considering the
-likelihood $p(Y\cond\bmtheta_{\rm max})$ using single best-fit
-parameter values $\bmtheta_{\rm max}$, but by integrating over all
-possible values of the $\bmtheta$ in terms of the posterior
-$p(\bmtheta\cond n,\bml)$,
+likelihood $p(Y\cond\bmtheta_{\rm max})$ using single
+maximum-likelihood parameter values $\bmtheta_{\rm max}$, but by
+integrating over all possible values of the likelihood, weighted by
+the posterior,
 \begin{align}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{From single-simulation data to many-simulation prediction}
+\section{Single-simulation data and multi-simulation prediction}
 
 % \subsection{Output data from radiative transfer simulations}
-\subsection{Simulation data}
+% \subsection{Simulation data}
 
 We consider a single computer simulation which generates a luminosity
-spectrum from $\npack$ photon packets, each with a particular
+spectrum from $\npack$ photon packets, each having a particular
 frequency $\nu_i$ and luminosity $\lum_i$. Together, these frequencies
@@
 with loss of information limited to replacing the exact packet
-frequencies $\nu_i$ by the appropriate $\nu_b$. The total luminosity
-in bin $b$ is the sum of luminosities of those packets whose frequency
-falls into $\Delta_b$,
+frequencies $\nu_i$ by the approximate bin frequency $\nu_b$. The
+total luminosity in bin $b$ is
 \begin{align}
@@
   \lumtot_b 
-  % &= \sum_{i=1}^\npack \bm{1}_{\Delta_b}(\nu_i)\,\ell_i
   &= \sum_{j=1}^{n_b} \lum_{b,j}.
 \end{align}
-% where the indicator function $\bm{1}_{[a,b]}(x) = 1$ when $x$ is in
-% the interval $[a,b]$ and 0 otherwise.
-
-\subsection{Prediction of luminosity spectrum}
-
 Within a given model $\hypo$, the data $\bmy$ is only one of many
@@
 \begin{align}
+  \lleq{sdd}
   \bmL_b &= \{(\Lum_{b,j}): j = 1,\ldots,N_b\}\\
+  \lleq{sde}
   \bmY &= \{(N_b, \bmL_b): b = 1,\ldots,B\},\\
+  \lleq{sdf}
   \Lumtot_b &= \sum_{j=1}^{N_b} \Lum_{b,j},
 \end{align}
-with corresponding likelihood $p(\bmY\cond\bmtheta)$ and posterior
-$p(\bmtheta\cond\bmY)$ for any $\bmY$. The lower-case $(n_b,\bml_b)$
-are strictly reserved for the simulated data at hand.
+Lower-case symbols such as $\bmy, n_b,\bml_b$ etc.\ are strictly
+reserved for the simulated data at hand.
 
@@
 luminosities $Q_b, b = 1,\ldots,B$ if many simulations had been
-carried out.
-
-Various features such as absorption lines make the luminosity spectrum
-highly variable. The model we develop below takes into account this
-variability in a two-step process. First, the number of packets $N_b$
-is modelled by means of a separate Poisson distribution
-$p(N_b\cond\lambda_b)$ for each bin, where the parameter $\lambda_b$
-will capture most of the variability of the spectrum.  Second, the
-single-packet luminosity distribution is modelled by means of a packet
-likelihood for $\Lum_{b,j}$ with packet parameters, a packet prior and,
-of course, packet parameter posterior. The packet parameters, which
-depend on the frequency bin, will be calculated in two ways: first by
-drawing them from a ``parametrisation of the parameters'' across all
-bins as shown in \refsec{example} and secondly by assuming that
-\refsec{asymptotic}
-
-Our approach
-is to predict each bin's luminosity $Q_b$ separately using the same
-model but allowing its parameters to vary independently in each bin.
-Connections between them
-
-function of frequency. In a binned analysis, we can avoid these
-difficulties associated to modeling $n_b$. Our approach, however,
-requires an explicit expression of how the distribution of the
-luminosity of a single packet in our simulation changes with $\nu$. It
-is this crucial piece of information that allows us to extend previous
-methods in dealing with the extreme cases when the number of packets
-in a bin, $n_b$, is one or even zero.
-
-To keep it simple, we shall assume that all bins are independent,
-meaning that
-\begin{align}
-  p(Q_1,Q_2,\ldots,Q_B\cond\bmy) = \prod_{b=1}^B p(Q_b\cond n_b,\bml_b).
-\end{align}
-Below, we calculate $p(Q_b\cond n_b,\bml_b)$ for fixed $b$ and hence
-drop the subscript, i.e.\ $Q_b\to Q, n_b \to n, N_b\to N$ etc.
-
+carried out. At this introductory level, we assume that the different
+luminosities are independent, so that the prediction of the entire
+spectrum (including all its uncertainty) is the product of predictions
+for each frequency bin,
+\begin{align}
+  \lleq{sdg}
+  p(Q_1,Q_2,\ldots,Q_B\cond\bmy) = \prod_{b=1}^B p(Q_b\cond \bmy).
+\end{align}
+
+\section{The generic model} \label{sec:model}
+
+Various physical processes such as absorption result in luminosity
+spectra with strong features. The model we develop below takes into
+account the variability introduced by these features in a two-step
+process termed a \textit{compound process} in the statistics
+literature. 
+% 
+In the first step, the number of packets $N$ is modelled by means of a
+separate Poisson distribution $p(N\cond\lambda)$ for each bin, where
+the magnitude of $\lambda$ will capture most of the variability of the
+spectrum. Given $N$, the single-packet luminosity distribution is
+modelled in the second step by means of a packet likelihood for
+$\Lum_{b,j}$ with packet parameters, a packet prior and, of course,
+packet parameter posterior. The single-packet distribution contains
+the crucial information which allows us to extend previous methods to
+cases where the number of packets in a bin is small.
 %%%%%%%%%%%%%%%%%%%
@@
 
-In terms of the generic notation of Section \ref{sec:bayp}, Bayesian
-prediction of the distribution of the bin's luminosity
-$Q = \sum_{j=1}^n L_j$ given the bin's data $(n,\bml)$ is obtained by
-considering not only the likelihood $p(Q\cond\bmtheta_{\rm max})$
-using best-fit parameter values $\bmtheta_{\rm max}$, but by
-integrating over all possible values of the $\bmtheta$ in terms of the
-posterior $p(\bmtheta\cond n,\bml)$,
-\begin{align}
-  p(Q\cond n,\bml) &= \int d\bmtheta\,p(Q\cond\bmtheta)\,p(\bmtheta\cond n,\bml).
-\end{align}
-The model's parameters $\bmtheta$ and calculation of this prediction
-will be set out in detail below.
-
-\section{Model} \label{sec:model}
-
-We consider the total number of packets $\npack$ fixed, for example
-chosen by a user. In contrast, the number of packets in a bin, $n_b$
-is a realization of a random variable $N_b$ which we assume to follow
-a Poisson distribution. With this assumption, we neglect bin-by-bin
-correlations and treat each bin independently. From now on, we focus
-on a single bin and drop the subscript $b$. Hence
+We derive in this section the generic equations of the compound
+process, and then proceed to show in Sections \ref{sec:example} and
+\ref{sec:asymptotic} how these equations can be used in two different
+ways.
+% 
+We consider for the moment quantities in one bin $b$ only, and hence
+drop the $b$-subscripts, e.g.\
+$Q_b \to Q, n_b \to n, N_b\to N, \bml_b\to\bml, L_{b,j} \to L_j$ etc.
+
+We seek a Bayesian prediction for the total luminosity in one bin,
+$p(Q\cond n,\bml)$, one of the factors in Eq.~(\ref{sdg}).  To do so,
+we must take into account all possible values of variables and
+parameters entering our model.  This involves multiple use of the
+\textit{marginalisation rule}, which states that, to obtain the
+probability for a variable $U$ when we are not interested in another
+variable $V$, the marginal probability for $U$ is %
+$p(U) = \int dV\,p(U,V) = \int dV\,p(U\cond V)\,p(V)$ for continuous
+$V$ and $p(U) = \sum_V p(U,V) = \sum_V p(U\cond V)\,p(V)$ for discrete
+$V$ respectively.
+% 
+The uninteresting variables $V$ in the present case are the number of
+packets $N$, the set of luminosities $\bmL$ of the $N$ packets
+whose likelihood is modelled by a distribution determined by
+likelihood parameters $\bmphi$,
+\begin{align}
+  \lleq{slk}
+  p(\bmL\cond N,\bmphi)
+  &= \prod_{j=1}^{N} p(L_j\cond \bmphi)
+\end{align}
+and the parameter $\lambda$ governing the likelihood for the number of
+packets, $p(N\cond\lambda)$. Multiple use of the marginalisation rule
+on $p(Q\cond n,\bml)$ results in
+\begin{align}
+  \lleq{sll}
+  p(Q\cond n,\bml)
+  &= \sum_{N=0}^\infty \;\int_0^\infty d\lambda\;
+    p(N\cond\lambda)\;
+    p(\lambda\cond n)
+    \int d\bmL\;
+    p(Q\cond N,\bmL)\;
+    \int d\bmphi\;
+    p(\bmL\cond N,\bmphi)\,
+    p(\bmphi\cond n,\bml),
+\end{align}
+where we have omitted from the probabilities all quantities to the
+right of the conditional line which are irrelevant to the variable at
+hand. For example, Eq.~(\ref{sdf}) shows that the total bin luminosity
+$Q$ is fully determined by $N$ and $\bmL$ and so we can shorten
+$p(Q\cond N,\bmL,\lambda,\bmphi,n,\bml)$ to $p(Q\cond N,\bmL)$. %
+Likewise the probability for $\bmL$ is fully determined by $N$ and the
+parameters $\bmphi$ and does not depend on the data $n,\bml$.  The
+bounds of $\int d\bmL$ and $\int d\bmphi$ are determined by the
+specific model and data for which $p(Q\cond n,\bml)$ is to be
+calculated.
+% 
+We note that a part of Eq.~(\ref{sll}) can be seen as a convolution
+since $p(Q\cond N,\bmL)$ can be written as a Dirac delta function,
+\begin{align}
+  \lleq{ddf}
+  \int d\bmL\; p(Q\cond N,\bmL)\; p(\bmL\cond N,\bmphi)
+  &= \int d\bmL\;\delta(Q - \textstyle\sum_{j=1}^N L_j)\; \prod_j p(L_j\cond \bmphi).
+\end{align}
+Eq.~(\ref{sll}) is exact, and in principle all that remains is to make
+specific choices for the various probabilities which are judged best
+to represent the data at hand, and thence to carry out the sum and
+multiple integrals. In practice, of course, such calculations may be
+time-consuming, and there will often be more than one choice for the
+functional form of each probability. In Section \ref{sec:example}, we
+shall therefore set out a specific example
+
+\section{Example A: modelling single-packet luminosity distributions} 
+\label{sec:example}
+
+In this section, we show by example how information on the
+single-packet luminosity distributions can be utilised to find exact
+expressions for Eq.~(\ref{sll}).
+
+\subsection{Multiplicity prediction}
+
+First, the parameter $\lambda$ is eliminated as follows. The number of
+packets $N$ is Poisson-distributed
 \begin{align}
   \lleq{caa}
-  p(N\cond\lambda) &= \frac{e^{-\lambda}\lambda^N}{N!}
-  \qquad N = 0,1,\ldots,\infty ,
-\end{align}
-where the dependency of the Poisson mean, $\lambda>0$, on the bin, or
-respectively the frequency $\nu$, is implicit. Let $\bmL = (\Lum_1,\dotsc,\Lum_N)$ denote the
-collection of luminosities of packets in the bin, then the total
-luminosity in that bin is
-% and its realization based on the data $\bmy$ are
-\begin{align}
-  \lleq{cab}
-  \Lumtot &= \sum_{j=1}^{N} \Lum_j%, \qquad \lumtot = \sum_{j=1}^{n} \lum_j
-  \,.
-\end{align}
-The distribution of the luminosity for each packet,
-$p(\Lum \cond \bmphi)$, depends on parameters $\bmphi$ and potentially
-on the frequency but we suppress the latter dependency.  The packets
-are logically independent. Since $N$ follows a Poisson distribution,
-$\Lumtot$ follows a \emph{compound Poisson process}.
-%
-The key quantity of interest is the distribution of $\Lumtot$ given
-the data $\bmy$, $p(\Lumtot \cond \bmy)$ for the compound Poisson
-process. In the following, we rewrite $p(\Lumtot \cond \bmy)$ to
-obtain an expression that we can readily evaluate numerically. Using
-the law of total probability,
-\begin{align}
-  \lleq{qaa}
-  p(\Lumtot \cond \bmy) &= \int \rmdx{\bmphi} p(\bmphi \cond \bmy) p(\Lumtot \cond \bmphi, \bmy) \,,
-\end{align}
-where the first term in the integral is the posterior $p(\bmphi \cond \bmy)$ after observing $\bmy$. We can decompose the second term as
-\begin{align}
-  \lleq{qab}
-  p(\Lumtot \cond \bmphi, \bmy) &= \sum_{N=0}^\infty  p(N \cond \bmphi, \bmy) p(\Lumtot \cond N, \bmphi, \bmy)\,,
-\end{align}
-where $p(N \cond \bmphi, \bmy) = p(N \cond \lambda)$ is given by
-\refeq{caa} and the second term in the sum is
-\begin{align}
-  \lleq{qac}
-  p(\Lumtot \cond N, \bmphi, \bmy) &= \int \rmdx{\bmL} p(\Lumtot \cond \bmL, N, \bmphi, \bmy) p(\bmL \cond N, \bmphi, \bmy)\\
-  \lleq{qaf}
-  &= \int \rmdx{\bmL} \delta \left( \Lumtot - \sum_{j=1}^N \Lum_j  \right) \prod_{j=1}^N p(\Lum_j \cond \bmphi)\,,
-\end{align}
-hence $Q$ does not depend on $\bmy$ explicitly. In the next
-subsections, we provide more details on our assumptions and provide
-explicit expressions for \refeq{qaa}--\refeq{qaf}.
-
-% As the reference frequency, we choose the bin center.
-
-\subsection{Modeling the packet luminosity} \label{sec:pack-lum}
-
-In general, the distribution $p(\Lum_j \cond \bmphi)$ could be a of a
-different functional form in each bin. For simplicity, we assume it is
-identical but may depend on parameters $\bmphi$ that need to be
-inferred from $\bmy$.  To simplify the treatment, we assume that the
-frequency bins are narrow or equivalently that $p(\Lumtot \cond \bmy)$
-does not vary within a bin. This implies that every sample $\Lum_j$ in
-the bin follows the same distribution.
-
-What is the role of $\bmphi$? It contains all parameters needed to
-model the entire spectrum that have to be learned from $\bmy$. To
-begin, we include the Poisson means for each bin
-\begin{align}
-  \lleq{qad} \bmphi = \left( \bmlambda, \bmxi \right) = \left(
-    {\lambda_1,\dotsc,\lambda_{n_b}}, \bmxi \right)\,.
-\end{align}
-\fred{How }
-The number and the luminosity of the packets are assumed independent,
-so the posterior factorizes as
-\begin{align}
-  \lleq{qadd}
-  p(\bmphi \cond \bmy) = p(\bmlambda \cond \bmy) p(\bmxi \cond \bmy) \,.
-\end{align}
-
-Regarding the additional parameters $\bmxi$, we have essentially two
-extreme options; both illustrated below in \refsec{example} and
-\refsec{asymptotic}. The first option is to assume an explicit
-\emph{parametric} form, for example some function $f$, such that
-\begin{align}
-  \lleq{qae}
-  p(\Lum \cond \bmphi) = f(\Lum \cond \bmxi, \nu) \,,
-\end{align}
-where we now made the dependency on $\nu$ explicit again.  In this
-way, we assume the same distribution for all frequencies and we only
-need to learn the probable values of $\bmxi$ from the data $\bmy$,
-$p(\bmxi \cond \bmy)$. The advantage is that $f(\Lum \cond \bmxi, \nu)$ may
-be well constrained by $\bmy$ if $\npack$ is large and it is easy to
-include further assumptions; e.g., $f$ should be continuous in $\nu$.
-
-The second option is preferable if it is impossible to elicit an
-appropriate explicit expression for $p(\Lum \cond \bmphi)$ \emph{and}
-each bin has many packets; i.e., $n \gg 1$. Then we can include one or
-more parameters for each bin, consider each bin independently, and
-perform a \emph{nonparametric} analysis. Compared to the parametric
-approach, we now typically have many more parameters, and each one is
-determined only by the packets in one instead of all bins.
-
-\subsection{Poisson analysis}
-
-In order to constrain the Poisson mean $\lambda$ in a bin, we have one
-measurement, the number of packets $n$ in that bin. Using Bayes' theorem, we find the posterior
+  p(N\cond\lambda) &= \frac{e^{-\lambda}\lambda^{N}}{N!}
+  \quad N = 0,1,\ldots,\infty.
+\end{align}
+Data $n$ is naturally presumed to have arisen from the same Poisson
+distribution with the same $\lambda$, so $p(n\cond\lambda) =
+e^{-\lambda}\lambda^{n}/n!$. %
+% and since $n$ is just one realisation of $N$, it likewise is
+% poissonian $p(n\cond\lambda) = e^{-\lambda}\lambda^{n}/n!$ with the
+% same $\lambda$.
+The posterior for $\lambda$ is found from the data $n$ via Bayes'
+Theorem,
 \begin{align}
@@
 \end{align}
-% to find the posterior predictive of $N$ given $n$ taking into account
-% all possible values of $\lambda$,
-% \begin{align}
-%   \lleq{pme}
-%   p(N\cond n)
-%   &= \int \rmdx{\lambda}p(N,\lambda\cond n)
-%   \ =\ \int \rmdx{\lambda} p(N\cond\lambda, n)\,p(\lambda\cond n)
-%   \ =\ \frac{\int \rmdx{\lambda} p(N\cond\lambda)\,p(n\cond\lambda)\,p(\lambda)}
-%   {\int \rmdx{\lambda} p(n\cond\lambda)\,p(\lambda)}
-% \end{align}
-% where $p(N\cond\lambda,n) = p(N\cond\lambda)$ is independent of $n$.
-Common uninformative priors are
+We use an improper prior 
 \begin{align}
@@
 \end{align}
-where $a=0$ gives the uniform prior\footnote{For a uniform prior, the
-  case $a{=}0$, we require of course a maximum $\lambda_{\rm max}$. As
-  long as $\lambda_{\rm max}$ is large enough, the integral can
-  effectively be taken to infinity.}, $a=1/2$ gives the Jeffreys and
-reference prior, and $a=1$ is the transformation-group result
-advocated by Jaynes~\cite[Ch. 12]{jaynes2003probability}. All choices
-lead to an improper prior but the posterior is proper provided that
-$n \ge 1$; i.e., at least one packet is observed in the bin. The
-standard result for the posterior is
+meaning that $\int_0^\infty d\lambda\,p(\lambda)$ is known only up to
+a constant $c$. A hyperparameter choice $a=0$ yields the uniform
+prior, $a=1/2$ gives the Jeffreys and reference prior, and $a=1$ is
+the transformation-group result advocated by
+Jaynes~\cite[Ch. 12]{jaynes2003probability}. An improper prior is not
+a problem because $c$ cancels out when calculating the posterior,
+provided that $n \ge 1$; i.e., at least one packet is observed in the
+bin.\footnote{For a uniform prior, the case $a{=}0$, we require of
+  course a parameter $\lambda_{\rm max}$. As long as
+  $\lambda_{\rm max}$ is large enough, it too would cancel.}, The
+posterior is
 \begin{align}
   \lleq{pms}
-  p(\lambda\cond n) &= \GammaDist(\lambda \cond \alpha = n-a+1, \beta = 1)\,,
-\end{align}
-where the  $\GammaDist$ distribution with shape parameter $\alpha$ and rate
-parameter $\beta$ is defined as
+  p(\lambda\cond n) 
+  &= \GammaDist(\lambda \cond \alpha = n{-}a{+}1 \,,\, \beta = 1)\,,
+\end{align}
+where the $\GammaDist$ distribution with shape parameter $\alpha > 0$ and
+rate parameter $\beta > 0$ is defined as
 \begin{align}
   \lleq{gma}
-  \GammaDist(\lambda | \alpha, \beta) = \frac{\beta^\alpha}{\Gamma(\alpha)} x^{\alpha-1} e^{-\beta x}\, , \; x > 0 .
-\end{align}
+  \GammaDist(x\cond \alpha, \beta) 
+  &= \frac{\beta^\alpha}{\Gamma(\alpha)} x^{\alpha-1} e^{-\beta x}\,,
+  \quad 0 \leq x < \infty.
+\end{align}
+The evidence is 
+$p(n) =  \int_0^\infty d\lambda\, p(n\cond\lambda)\,p(\lambda) = c\,(n-a)!/n!$
 % \begin{align}
 %   \lleq{pmg}
-%   p(n) &= %
-%   \int_0^\infty d\lambda\, p(n\cond\lambda)\,p(\lambda)
-%   \ =\
-%   c \int_0^\infty d\lambda\, \frac{e^{-\lambda}\,\lambda^{n-a}}{n!}
-%   \ =\ c\frac{(n-a)!}{n!} \, ,
-%   \\
-%   \lleq{pmh}
-%   p(N\cond n) %
-%   &= \frac{n!}{(n-a)!} \int_0^\infty d\lambda\,
-%   \frac{e^{-2\lambda} \lambda^{N+n-a}}{N!\, n!}
-%   \ =\ \frac{(N+n-a)!}{N!\,(n-a)!\,2^{N+n+1-a}} \, .
+%   p(n) 
+%   &=  \int_0^\infty d\lambda\, p(n\cond\lambda)\,p(\lambda)
+%     \ =\ c\frac{(n-a)!}{n!} \,, \\
 % \end{align}
-
-\section{Example application: \tardis} \label{sec:example}
-
- The \tardis{} packages simulates supernovae with \fred{Wolfgang's
-  favorite description of his package}. Here we focus on the real
-packets as opposed to the virtual packets output by tardis.
-
-% Based on staring at the distribution of luminosities,
-To apply \refeq{pmp}, we first need to choose $P(\Lum_j \cond \bmphi,
-\nu)$. A histogram of the luminosities from a \tardis{} run indicates
-a strongly skewed distribution with a single mode that is close to but
-not at the sharp end point $\lmax$. There is a long tail for small
-$\ell$. These features can be accomodated by a $\GammaDist$
-distribution with shape parameter $\alpha$ and rate
-parameter $\beta$
-\begin{align}
-  \lleq{gma}
-  \GammaDist(x | \alpha, \beta) = \frac{\beta^\alpha}{\Gamma(\alpha)} x^{\alpha-1} e^{-\beta x}\, , \; x > 0 .
-\end{align}
-if we transform the data as
+and inserting Eqs.~(\ref{caa}) and (\ref{pmd}), the prediction for $N$
+given $n$ is
+\begin{align}
+  \lleq{pmh}
+  p(N\cond n) %
+  &= \int_0^\infty d\lambda\, p(N\cond\lambda)\,p(\lambda\cond n)
+  \ =\ \frac{(N+n-a)!}{N!\,(n-a)!\,2^{N+n+1-a}} \, .
+\end{align}
+so that Eq.~(\ref{sll}) simplifies to
+\begin{align}
+  \lleq{gmh}
+  p(Q\cond n,\bml)
+  &= \sum_{N=0}^\infty \;
+    p(N\cond n)\;
+    \int d\bmL\;
+    p(Q\cond N,\bmL)\;
+    \int d\bmphi\;
+    p(\bmL\cond N,\bmphi)\,
+    p(\bmphi\cond n,\bml),
+\end{align}
+
+\subsection{Single-packet description}
+
+Next, we consider detailed modelling of $p(\bmL\cond N,\bmphi)$ which
+was written in Eq.~(\ref{slk}) as a product single-packet
+distributions $p(L_j\cond \bmphi)$, meaning that single packets are
+assumed to be mutually independent and to follow the same distribution
+independent of $N$.
+
+Which specific probability function to choose for $p(L\cond\bmtheta)$
+depends, of course, on the specific simulation data. In this section,
+we choose it to be the same Gamma distribution for all $L_j$,
+with parameters $\bmphi = (\alpha,\beta)$,
+\begin{align}
+  \lleq{spc}
+  p(L_j\cond \bmphi) &\sim \GammaDist(L_j\cond \alpha,\beta)
+\end{align}
+which has the nice property that the convolution (\ref{ddf}) results
+in yet another Gamma distribution,
+\begin{align}
+  \lleq{spd}
+  \int d\bmL\; p(Q\cond N,\bmL)\; p(\bmL\cond N,\bmphi)
+  &= \GammaDist(Q\cond N\alpha, \beta).
+\end{align}
+and so (\ref{gmh}) becomes in this example
+\begin{align}
+  \lleq{spe}
+  p(Q\cond n,\bml)
+  &= \sum_{N=0}^\infty \;
+    p(N\cond n)\;
+    \int d\alpha\,d\beta\;
+    \GammaDist(Q\cond N\alpha, \beta)\;
+    p(\alpha,\beta\cond n,\bml).
+\end{align}
+The posterior for the parameters is calculated by yet another 
+application of Bayes' Theorem,
+\begin{align}
+  \lleq{spf}
+  p(\alpha,\beta\cond n,\bml)
+  &= \frac{p(n,\bml\cond \alpha,\beta)\,p(\alpha,\beta)}{p(n,\bml)}
+\end{align}
+where the data $(n,\bml)$ must naturally follow the same
+Gamma-distribution likelihood as $(N,\bmL)$, i.e.\
+\begin{align}
+  \lleq{spg}
+  p(n,\bml\cond \alpha,\beta) 
+  & = \prod_{j=1}^n \GammaDist(\ell_j\cond\alpha,\beta)
+    \ =\ \frac{\beta^{n\alpha} p^{\alpha-1}\,e^{-\beta q}}{[\Gamma(\alpha)]^n}
+\end{align}
+with compact representation of the data in terms of
+$q = \sum_{j=1}^n \ell_j$ and $r = \prod_{j=1}^n \ell_j$.
+
+The only missing piece to explicitly evaluate \refeq{spf} is the prior
+$p(\alpha,\beta)$. 
+% To gain some insight, consider the standard $\GammaDist$ model with
+% $\Kalpha = \Kbeta = 0$.
+% There is a conjugate prior but
+% unfortunately the normalization constant cannot be evaluated in closed
+% form, so it is thus of no use for us.
+% How about a noninformative prior? 
+The authors of \cite{moala2013bayesian} present an overview of
+possible choices and conclude that the point estimate given by the
+mode of the posterior is nearly insensitive to the prior choice
+starting from around $n = 30$. This requirement is nearly always
+satisfied and so the functional form of $p(\alpha,\beta)$ is not
+important. There is, however, some explicit prior information which we
+want to build into our analysis. The distribution
+$p(\ell \cond \alpha, \beta)$ should always vanish at the origin,
+requiring that $\alpha$ must be larger than 1.  For $\beta$, we
+likewise choose a uniform prior
+\begin{align}
+  \lleq{sph}
+  p(\alpha,\beta) 
+  &= \frac{1}{(\alpha_{\rm max}{-}1)\beta_{\rm max}}
+  \qquad 1 < \alpha < \alpha_{\rm max},\; 0 < \beta < \beta_{\rm max}.
+\end{align}
+% since $\alpha_{\rm max}$ and $\beta_{\rm max}$ cancel in (\ref{spf}).
+The posterior is
+\begin{align}
+  \lleq{spi}
+  p(\alpha,\beta\cond n,\bml)
+  &= \frac{1}{C}\; \frac{\beta^{n\alpha} r^{\alpha-1}\,e^{-\beta q}}{[\Gamma(\alpha)]^n}\\
+  \lleq{spj}
+  C &= \int_1^{\alpha_{\rm max}} d\alpha \int_0^{\beta_{\rm max}} d\beta\;
+    \frac{\beta^{n\alpha} r^{\alpha-1}\,e^{-\beta q}}{[\Gamma(\alpha)]^n}
+\end{align}
+and hence
+\begin{align}
+  \lleq{spj}
+  p(Q\cond n,\bml)
+  &= \frac{1}{C} 
+    \sum_{N=0}^\infty \;
+    p(N\cond n)\;
+    \int d\alpha\,d\beta\;
+    \GammaDist(Q\cond N\alpha, \beta)\;
+    \frac{\beta^{n\alpha} p^{\alpha-1}\,e^{-\beta q}}{[\Gamma(\alpha)]^n}
+\end{align}
+
+\subsection{Numerical implementation} \label{sec:numerical}
+
+\fred{Editing will be done by Fred}
+
+In our numerical implementation, the constants
+$(\alpha_{\rm max}, \beta_{\rm max})$ are chosen by trial and error to
+be large enough to contain the relevant region of
+$p(\ell\cond \alpha,\beta)$.
+% 
+To perform the integrals over $(\alpha,\beta)$ to find $C$ and the
+Gamma-weighted numerator in (\ref{spe}), we used the fact that the
+posterior $p(\alpha,\beta \cond n,\bml)$ is approximately Gaussian in
+both $\alpha$ and $\beta$ for large $n$, so we can apply Laplace's
+method for both. This requires the Hessian at the mode and the
+gradient helps in locating the mode. The relevant analytic expressions
+are summarized in \refsec{appendix}.
+
+Regarding the sum over $N$, we start at
+$N = \lfloor n-a+1 \rfloor = \arg \max_N p(N|n)$ %
+\fred{Actually we should say $p(N \cond n, a)$} %
+and continue to larger and smaller values of $N$ until the
+contribution to $P(Q \cond n,\bml)$ becomes negligible. In case the
+relevant contribution is from large $N$ and $n$, we interpret $N$ as a
+continuous variable, replace factorials by $\Gamma$ function in
+\refeq{spj}, and apply Laplace's method to integrate over $N$ also.
+
+\fred{Our implementation is available online on \href{https://github.com/tardis-sn/XXX}{github}}
+
+\subsection{Results} \label{sec:gammaresults}
+
+\fred{Graphics output, comments, conclusions by Fred}
+
+\subsection{Specifics of \tardis} \label{sec:tardis}
+
+\hans{H: This subsection needs much more work. I have just copied some
+  relevant text in here which may be useful. Much of it should be
+  removed.}
+
+The \tardis{} packages simulates supernovae with %
+\fred{Wolfgang's favorite description of his package}. %
+Here we focus on the real packets as opposed to the virtual packets
+output by tardis.
+
+A histogram of the luminosities from a \tardis{} run indicates a
+distribution strongly skewed with a mode close to the maximum
+luminosity $\lmax$, while there is a long tail for small $\ell$. These
+features can be accomodated by a $\GammaDist$ distribution if we
+transform the luminosities to
 \begin{align}
   \lleq{gmf}
-  x_i \equiv 1- \frac{\ell_i}{\lmax} > 0
-\end{align}
-and similarly for $\Lumtot \to X$ because $\ell$ is bounded from above and
-$x$ is bounded from below.  We first take the largest luminosity of
-all \tardis{} packets across the entire frequency range and increase
-it slightly to avoid numerical difficulties at the end point
+  x_j \equiv 1- \frac{\ell_j}{\lmax} > 0
+\end{align}
+and similarly $X_j = 1 - (L_j/\lmax)$ and $\Lumtot \to X$. We assume
+the total number of samples across all bins is large, hence the effect
+of considering $\lmax$ a constant independent of the data is
+negligible. We checked that leaving $\lmax$ as a free parameter in a
+fit gave equivalent results.
+% 
+Numerically, we take the largest luminosity of all \tardis{} packets
+across the entire frequency range and increase it slightly to avoid
+numerical difficulties at the end point
 \begin{align}
@@
 \end{align}
-where $ \epsilon=10^{-6}$.  We assume the total number of samples
-across all bins is large, hence the effect of considering $\lmax$ a
-constant independent of the data is negligible. We checked that
-leaving $\lmax$ as a free parameter in a fit gave equivalent results.
-
-We infer $P(X \cond \nu, \npack, \bmnu, \bmx)$ and only in the end
-transform back to a probability density in $\Lumtot$ via
+where $ \epsilon=10^{-6}$.
+
+We infer $P(X \cond n,\bml)$ and only in the end transform back to a
+probability density in $\Lumtot$ via
 \begin{align}
   \lleq{gmg}
-  p(\Lumtot  \cond \nu,  \npack, \bmnu, \bml) = p(X  \cond \nu,  \npack, \bmnu, \bmx) \frac{1}{\lmax} \,.
+  p(\Lumtot  \cond n, \bml) 
+  &= p(X  \cond n, \bmx) \frac{1}{\lmax} \,.
 \end{align}
 To simplify the numerics, we also rescale the frequencies by a
-constant from $\mathcal{O}(10^{15})$ to $\mathcal{O}(1)$.  A major
-numerical simplification arises because of the $\Lumtot \to X$
-transformation. Rewriting the master~\refeq{pmp} formally as
-\begin{align}
-  \lleq{gmh}
-  p(\Lumtot  \cond \nu,  \npack, \bmnu, \bml) = \sum_{N} p(N \cond \nu,  n) p(\Lumtot \cond N,
-\nu, \npack, \bmnu, \bml),
+constant from $\mathcal{O}(10^{15})$ to $\mathcal{O}(1)$.  %
+
+A major numerical simplification arises because of the $\Lumtot \to X$
+transformation. Rewriting the \refeq{sll} formally as
+\begin{align}
+  \lleq{gmhb}
+  p(Q  \cond n,\bml) 
+  & = \sum_{N} p(N \cond n)\; p(Q \cond N, n, \bml),
 \end{align}
@@
 in the sense that the standard deviation relative to the mean is at
-the percent level. This implies that $p(\Lumtot \cond \nu, \npack,
-\bmnu, \bml)$ has one mode from each $N$ and all modes are clearly
-separated. The transformation $\Lumtot \to x$ effects a strong overlap
-between the distributions $p(X \cond N, \nu, \npack, \bmnu, \bmx)$
-with similar $N$, so $p(X \cond \nu, \npack, \bmnu, \bmx)$ can be well
-approximated by a unimodal distribution; see \reffig{hedgehog}.
+the percent level. This implies that $p(Q \cond N, n, \bml)$ has one
+mode from each $N$ and all modes are clearly separated. The
+transformation $\Lumtot \to X$ effects a strong overlap between the
+distributions $p(X \cond N, n, \bmx)$ with similar $N$, so
+$p(X \cond n, \bmx)$ can be well approximated by a unimodal
+distribution; see \reffig{hedgehog}.
 
@@
 \begin{align}
-  \lleq{gmi}
+  \lleq{gmib}
   p(\bmX \cond \bmphi, N, \nu) &= \prod_{j=1}^N \GammaDist(X_j \cond \alpha, \beta ),\\
@@
 where $\alpha_i \equiv \alpha(\nu_i \cond \bmphi), \beta_i \equiv
-\beta(\nu_i \cond \bmphi)$.  Combining \refeq{gmb} and \refeq{gmi}, we
-can solve the integral
-\begin{align}
-  \lleq{gmc}
-  \int \rmdx{\bmX} \delta(X - \textstyle\sum_j X_j)
-  \,p(\bmX\cond N,\bmphi, \nu) = \GammaDist(X \cond N \alpha, \beta)
-\end{align}
-implying that the master equation~(\ref{pmp}) considerably simplifies
-as
-\begin{align}
-  \lleq{gmd}
-    \boxed{
-    p(X\cond \nu, \npack,\bmx,\bmnu)
-  = \frac{1}{p(\bmx\cond \npack, \bmnu)}
-  \sum_N p(N\cond n)\, \int \rmdx{\bmphi} \GammaDist(X \cond N \alpha, \beta) % p(X \cond N, \bmphi, \nu)
-  \, p(\bmx\cond \bmphi, \npack, \bmnu)
-  \, p(\bmphi).
-  }
-\end{align}
+\beta(\nu_i \cond \bmphi)$.
 
@@
 
-In the absence of a physical model to determine $\alpha(\nu \cond
-\bmphi)$ and $\beta(\nu \cond \bmphi)$, we propose to start with a simple polynomial ansatz   $\alpha(\nu \cond \bmphi) = \sum_{k=0}^\Kalpha \alpha_k \nu^k, \beta(\nu \cond \bmphi) = \sum_{k=0}^\Kbeta \beta_k \nu^k$ such that
+\fred{Parts may still be relevant, especially if we put a plot of an
+  entire spectrum in the paper.}
+
+For simple functions $\alpha(\nu \cond \bmphi)$, one can find the
+minimum value of $\alpha$ over the entire range of $\nu$
+analytically. If this is too complicated, we propose to require
+$\alpha(\nu_i \cond \bmphi) > 1$ only at each packet frequency
+$\nu_i$. In the evaluation of $p(\bmx \cond \bmphi, \npack, \bmnu)$,
+one has to loop over all packets and compute $\alpha$ and $\beta$
+anyways, so the constraint imposes essentially no computational
+overhead.
+
+In the absence of a physical model to determine
+$\alpha(\nu \cond \bmphi)$ and $\beta(\nu \cond \bmphi)$, we propose
+to start with a simple polynomial ansatz
+$\alpha(\nu \cond \bmphi) = \sum_{k=0}^\Kalpha \alpha_k \nu^k,
+\beta(\nu \cond \bmphi) = \sum_{k=0}^\Kbeta \beta_k \nu^k$ such that
 \begin{align}
@@
 
-\subsection{Prior} \label{sec:priors}
-
-The only missing piece to explicitly evaluate \refeq{gmd} is the prior
-$p(\bmphi)$. To gain some insight, consider the standard $\GammaDist$
-model with $\Kalpha = \Kbeta = 0$.
-% There is a conjugate prior but
-% unfortunately the normalization constant cannot be evaluated in closed
-% form, so it is thus of no use for us.
-How about a noninformative prior? The authors of
-\cite{moala2013bayesian} present an overview of possible choices and
-conclude that the point estimate given by the mode of the posterior is
-nearly insensitive to the prior choice starting from around 30
-independent identically distributed samples. We posit that there are
-this many samples available near any $\nu$ under consideration, so the
-form of the prior is not crucial for our application. But there is
-some explicit prior information that we want to build into our
-analysis. The distribution $p(x \cond \alpha, \beta)$ should always
-vanish at the origin. This is equivalent to requiring a maximum away
-from the boundary leading to $\alpha > 1$. We choose a uniform prior
-on $\bmphi$ on a finite box $V$ through indicator functions
-\begin{align}
-  p(\bmphi) = \mathbf{1}_{\alpha(\bmphi) > 1}(\bmphi) \mathbf{1}_V(\bmphi).
-\end{align}
-
-\subsection{Numerical implementation} \label{sec:numerical}
-
-The box $V$ is chosen by trial and error to be large enough to contain
-the relevant region of $p(\bmx\cond \bmphi, \npack, \bmnu)$. For
-simple functions $\alpha(\nu \cond \bmphi)$, one can find the minimum
-value of $\alpha$ over the entire range of $\nu$ analytically. If this
-is too complicated, we propose to require $\alpha(\nu_i \cond \bmphi)
-> 1$ only at each packet frequency $\nu_i$. In the evaluation of
-$p(\bmx \cond \bmphi, \npack, \bmnu)$, one has to loop over all
-packets and compute $\alpha$ and $\beta$ anyways, so the constraint
-imposes essentially no computational overhead.  To explicitly evaluate
-\refeq{gmd}, we have to overcome the remaining two difficulties: the
-integral over $\bmphi$ and the sum over $N$.
-
-As for the former, the posterior $p(\bmphi \cond \npack, \bmnu, \bmx)$
-is approximately Gaussian in $\phi$ for $\npack$ large, so we can
-apply Laplace's method to perform the $\bmphi$ integrals for both the
-evidence $p(\bmx \cond \npack, \bmnu)$ and the posterior mean $p(X
-\cond N, \npack, \bmnu, \bmx)$. This requires the Hessian at the mode
-and the gradient helps in locating the mode; the analytic expressions
-are summarized in \refsec{appendix}.
-
-Regarding the sum over $N$, we start at $N = \lfloor n-a+1 \rfloor =
-\arg \max_N p(N|n)$ \fred{Actually we should say $p(N \cond n, a)$}
-and continue to larger and smaller values of $N$ until the
-contribution to $P(X \cond \nu, \npack, \bmnu, \bmx)$ becomes
-negligible. In case the relevant contribution is from large $N$ and
-$n$, we interpret $N$ as a continuous variable, replace factorials by
-$\Gamma$ function in \refeq{pmh}, and apply Laplace's method to
-integrate over $N$ and $\bmphi$.
-
-\fred{Our implementation is available online on \href{https://github.com/tardis-sn/XXX}{github}}
-
-\subsection{Graphical output}
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Asymptotic approximation} \label{sec:asymptotic}
 
-In the case that there are many packets in every bin, some
-simplifications arise as the exact distribution of individual packets
-$p(\Lum \cond \bmphi, \nu)$ becomes irrelevant in the asymptotic
-regime $n \to \infty$, which implies $\lambda \to \infty$. In that
-case, the compound Poisson distribution can be approximated by a
-Gaussian distribution provided that the second moment of
-$\Lum, \expect{\Lum^2}$ is finite. To see this in more detail, we define
+We return to the general formulation of \refeq{sll}, whose factors we 
+rearrange to write it as 
+\begin{align}
+  \lleq{asc}
+  p(Q\cond n,\bml) 
+  &= \int d\lambda\; d\bmphi\; 
+    p(\lambda\cond n)\;
+    p(\bmphi\cond n,\bml)\;
+    p(Q\cond \lambda,\bmphi) \\
+  \lleq{asd}
+  p(Q\cond \lambda,\bmphi)
+  &= \sum_N 
+    p(N\cond \lambda)\;
+    \int d\bmL\; 
+    p(Q\cond N,\bmL)\;
+    p(\bmL\cond n,\bmphi)
+\end{align}
+with $p(\bmL\cond n,\bmphi) = \prod_{j=1}^n p(L\cond\bmphi)$ as
+before.  If, as assumed, $p(N\cond\lambda)$ is Poissonian, there is a
+simple relation between their respective moments\footnote{The $k$-th
+  moment of a variable is defined as
+  $\mu_k = \expect{x^k} = \int dx\,p(x)\,x^k, k = 1,2,\ldots$, while
+  the variance is equal to $\sigma^2 = \mu_2 - \mu_1^2$.} %
+for $p(Q\cond \lambda,\bmphi)$ and $p(L\cond \bmphi)$ as follows.
+%%%%%%%%%%%%%%%
+\fred{citation} %
+%%%%%%%%%%%%%%%
+In terms of the moments and variance of $L$,
 \begin{align}
   \lleq{ram}
-  \mu \equiv \expect{\Lum}, \qquad \sigma^2 \equiv \variance{\Lum}\,.
-\end{align}
-For a compound Poisson process with arbitrary Poisson mean $\lambda > 0$,
+  \mu_1 &\equiv \expect{\Lum}, \qquad 
+  \mu_2 \equiv \expect{\Lum^2}, \\
+  \sigma^2 &\equiv \variance{\Lum} 
+  = \expect{\Lum^2} - \expect{\Lum}^2\,,
+\end{align}
+the first moment and variance of $Q$ are 
 \begin{align}
   \lleq{raa}
-  \expect{Q} =\lambda \expect{L} = \lambda \mu, \qquad \variance{Q} = \lambda \expect{L^2} = \lambda (\mu^2 + \sigma^2) \,.
-\end{align}
-
+  \expect{Q} = \lambda\, \mu_1, \qquad 
+  \variance{Q} = \lambda \mu_2 = \lambda (\mu_1^2 + \sigma^2) \,.
+\end{align}
+for \textit{any} $p(L\cond\bmphi)$. 
+%
+\hans{Stopped editing HERE 161202 18h30}
+Furthermore, higher-order quantities such as the skewness and kurtosis
+are suppressed by powers of $\lambda^{-1/2}$.
+% 
 From Theorem 4.3.1 of \cite{bening2002generalized}, we can conclude
@@
   \lleq{rab}
-  P(\Lumtot \cond \bmphi) = \GaussianDist \left( \Lumtot \cond \expect{Q}, \variance{Q} \right) = \GaussianDist \left( \Lumtot \cond \lambda \mu, \lambda (\mu^2 + \sigma^2) \right).
-\end{align}
-This result is a drastic simplification as it eliminates both the sum over $N$ in \refeq{qab} and the integral over $\bmL$ in \refeq{qaf}.
-That allows us to model $\Lumtot$ in each bin \emph{without} making concrete
-assumptions about the functional form of $P(\Lum \cond \bmphi)$; i.e., we may now view $\bmphi$ as the collection of first and second moments in each bin in addition to $\lambda$.
-Continuing the notation of \refeq{qadd},
+  P(\Lumtot \cond \bmphi) 
+  &= \GaussianDist \left( \Lumtot \cond \expect{Q}, \variance{Q} \right) %
+    = \GaussianDist \left( \Lumtot \cond \lambda \mu, \lambda (\mu^2 + \sigma^2) \right).
+\end{align}
+This result is a drastic simplification as it eliminates both the sum
+over $N$ in \refeq{qab} and the integral over $\bmL$ in \refeq{qaf}.
+That allows us to model $\Lumtot$ in each bin \emph{without} making
+concrete assumptions about the functional form of
+$P(\Lum \cond \bmphi)$; i.e., we may now view $\bmphi$ as the
+collection of first and second moments in each bin in addition to
+$\lambda$. Continuing the notation of \refeq{qadd},
 \begin{align}

### 2016-12-03 real/draft.tex
[real] Fix bug in asymptotic variance

@@
 %% CHANGE THIS DATE AS YOU MODIFY THE FILE
-\newcommand{\vdate}{161202}
+\newcommand{\vdate}{161206}
 \pagestyle{myheadings}
@@
   \textbf{\Large Bayesian prediction of supernova luminosity  distributions}\\[8pt]
-  \textbf{\Large based on simulation output: An example from \tardis}\\[12pt]
+  \textbf{\Large based on simulation output}\\[12pt]
 \end{center}
@@
   \item Are finite number of terms in $\sum_N$ a problem? Check
+
 \end{enumerate}
@@
 
-\subsection{tardis}
-
-move tardis section into separate document in this git repo.
-Better to keep it out of the paper.
-Cite tardis when showing the toy spectrum.
-Mention that in reality $\alpha$ and $\beta$ are $\nu$ dependent
+\subsection{Other issues that need to be resolved}
+
+\begin{enumerate}
+
+\item Clean up integrals using \verb=\rmdx= rather than $d$.
+
+\item Write everything in terms of frequency $\nu$ or wavelength
+  $\lambda$
+
+\item Now that we have removed direct analysis of \tardis\ from the
+  paper, there is no need to define an overall parameterization of the
+  Gamma parameters $\alpha(\nu_b) = \sum_k a_k \nu_b^k$ and
+  $\beta(\nu_b) = \sum_k b_k \nu_b^k$. However, the possibility of
+  linking the analyses of all the bins by this and similar methods
+  should be highlighted. Perhaps in the Conclusions?
+
+\item \textbf{\tardis} move tardis section into separate document in
+  this git repo. Better to keep it out of the paper. Cite tardis when
+  showing the toy spectrum. Mention that in reality $\alpha$ and
+  $\beta$ are $\nu$ dependent
+
+\end{enumerate}
 
@@
 
-\textbf{Define here?}
-\begin{enumerate}
-\item Basic notation for packets, luminosity as used below
-\end{enumerate}
-
-\textbf{Issues that need to be resolved}
-\begin{enumerate}
-\item Write everything in terms of frequency $\nu$ or wavelength $\lambda$
-\end{enumerate}
+Wolfgang's text
 
@@
     %= \frac{p(\bmy\cond\bmtheta)\;p(\bmtheta)}
-    %{\int d\bmtheta'\, p(\bmy\cond\bmtheta')\;p(\bmtheta')}
+    %{\int \rmdx{\bmtheta}'\, p(\bmy\cond\bmtheta')\;p(\bmtheta')}
 \end{align}
@@
   \lleq{bti}
-  p(\bmy) &= \int d\bmtheta'\, p(\bmy\cond\bmtheta')\;p(\bmtheta'),
+  p(\bmy) &= \int \rmdx{\bmtheta}'\, p(\bmy\cond\bmtheta')\;p(\bmtheta'),
 \end{align}
@@
   \lleq{btj}
-  p(Y\cond \bmy) &= \int d\bmtheta\,p(Y\cond\bmtheta)\,p(\bmtheta\cond \bmy).
+  p(Y\cond \bmy) &= \int \rmdx{\bmtheta}\,p(Y\cond\bmtheta)\,p(\bmtheta\cond \bmy).
 \end{align}
@@
   p(Q\cond n,\bml)
-  &= \sum_{N=0}^\infty \;\int_0^\infty d\lambda\;
+  &= \sum_{N=0}^\infty \;\int_0^\infty \rmdx{\lambda}\;
     p(N\cond\lambda)\;
     p(\lambda\cond n)
-    \int d\bmL\;
+    \int \rmdx{\bmL}\;
     p(Q\cond N,\bmL)\;
-    \int d\bmphi\;
+    \int \rmdx{\bmphi}\;
     p(\bmL\cond N,\bmphi)\,
@@
 parameters $\bmphi$ and does not depend on the data $n,\bml$.  The
-bounds of $\int d\bmL$ and $\int d\bmphi$ are determined by the
+bounds of $\int \rmdx{\bmL}$ and $\int \rmdx{\bmphi}$ are determined by the
 specific model and data for which $p(Q\cond n,\bml)$ is to be
@@
 \begin{align}
-  \lleq{ddf}
-  \int d\bmL\; p(Q\cond N,\bmL)\; p(\bmL\cond N,\bmphi)
-  &= \int d\bmL\;\delta(Q - \textstyle\sum_{j=1}^N L_j)\; \prod_j p(L_j\cond \bmphi).
+  \lleq{slm}
+  \int \rmdx{\bmL}\; p(Q\cond N,\bmL)\; p(\bmL\cond N,\bmphi)
+  &= \int \rmdx{\bmL}\;\delta(Q - \textstyle\sum_{j=1}^N L_j)\; \prod_j p(L_j\cond \bmphi).
 \end{align}
@@
 \begin{align}
-  \lleq{caa}
+  \lleq{pmb}
   p(N\cond\lambda) &= \frac{e^{-\lambda}\lambda^{N}}{N!}
@@
 \begin{align}
-  \lleq{pmd}
+  \lleq{pmc}
   p(\lambda\cond n) %
@@
 \end{align}
-meaning that $\int_0^\infty d\lambda\,p(\lambda)$ is known only up to
+meaning that $\int_0^\infty \rmdx{\lambda}\,p(\lambda)$ is known only up to
 a constant $c$. A hyperparameter choice $a=0$ yields the uniform
@@
 The evidence is
-$p(n) =  \int_0^\infty d\lambda\, p(n\cond\lambda)\,p(\lambda) = c\,(n-a)!/n!$
+$p(n) =  \int_0^\infty \rmdx{\lambda}\, p(n\cond\lambda)\,p(\lambda) = c\,(n-a)!/n!$
 % \begin{align}
@@
 %   p(n)
-%   &=  \int_0^\infty d\lambda\, p(n\cond\lambda)\,p(\lambda)
+%   &=  \int_0^\infty \rmdx{\lambda}\, p(n\cond\lambda)\,p(\lambda)
 %     \ =\ c\frac{(n-a)!}{n!} \,, \\
 % \end{align}
-and inserting Eqs.~(\ref{caa}) and (\ref{pmd}), the prediction for $N$
+and inserting Eqs.~(\ref{pmb}) and (\ref{pmc}), the prediction for $N$
 given $n$ is
@@
   p(N\cond n) %
-  &= \int_0^\infty d\lambda\, p(N\cond\lambda)\,p(\lambda\cond n)
-  \ =\ \frac{(N+n-a)!}{N!\,(n-a)!\,2^{N+n+1-a}} \, .
+  &= \int_0^\infty \rmdx{\lambda}\, p(N\cond\lambda)\,p(\lambda\cond n)
+  \ =\ \binom{N{+}n{-}a}{N} \left(\frac{1}{2}\right)^{N+n-a+1}
+    %\ =\ \frac{(N+n-a)!}{N!\,(n-a)!\,2^{N+n+1-a}} \, .
 \end{align}
@@
     p(N\cond n)\;
-    \int d\bmL\;
+    \int \rmdx{\bmL}\;
     p(Q\cond N,\bmL)\;
-    \int d\bmphi\;
+    \int \rmdx{\bmphi}\;
     p(\bmL\cond N,\bmphi)\,
@@
 \end{align}
-which has the nice property that the convolution (\ref{ddf}) results
+which has the nice property that the convolution (\ref{slm}) results
 in yet another Gamma distribution,
@@
   \lleq{spd}
-  \int d\bmL\; p(Q\cond N,\bmL)\; p(\bmL\cond N,\bmphi)
+  \int \rmdx{\bmL}\; p(Q\cond N,\bmL)\; p(\bmL\cond N,\bmphi)
   &= \GammaDist(Q\cond N\alpha, \beta).
@@
     p(N\cond n)\;
-    \int d\alpha\,d\beta\;
+    \int \rmdx{\alpha}\,\rmdx{\beta}\;
     \GammaDist(Q\cond N\alpha, \beta)\;
@@
 \begin{align}
-  \lleq{spj}
+  \lleq{spk}
   p(Q\cond n,\bml)
@@
   relevant text in here which may be useful. Much of it should be
-  removed.}
+  removed. Some connection to TARDIS should preferably be kept in the
+  paper, in my view, eg the definition of $X = 1 - L/\ell_{\rm max}$.}
 
@@
 
-We return to the general formulation of \refeq{sll}, whose factors we 
-rearrange to write it as 
+\subsection{Derivation of $p(Q\,|\,n,\boldmath{\ell})$ for large $n$}
+
+We return to the general formulation of \refeq{sll} and re-arrange its
+factors into
 \begin{align}
@@
   p(Q\cond n,\bml)
-  &= \int d\lambda\; d\bmphi\; 
+  &= \int \rmdx{\lambda}\; \rmdx{\bmphi}\;
     p(\lambda\cond n)\;
@@
     p(N\cond \lambda)\;
-    \int d\bmL\; 
+    \int \rmdx{\bmL}\;
     p(Q\cond N,\bmL)\;
@@
 \end{align}
-with $p(\bmL\cond n,\bmphi) = \prod_{j=1}^n p(L\cond\bmphi)$ as
-before.  If, as assumed, $p(N\cond\lambda)$ is Poissonian, there is a
-simple relation between their respective moments\footnote{The $k$-th
-  moment of a variable is defined as
-  $\mu_k = \expect{x^k} = \int dx\,p(x)\,x^k, k = 1,2,\ldots$, while
-  the variance is equal to $\sigma^2 = \mu_2 - \mu_1^2$.} %
-for $p(Q\cond \lambda,\bmphi)$ and $p(L\cond \bmphi)$ as follows.
-%%%%%%%%%%%%%%%
-\fred{citation} %
-%%%%%%%%%%%%%%%
-In terms of the moments and variance of $L$,
+with $p(\bmL\cond n,\bmphi) = \prod_{j=1}^n p(L_j\cond\bmphi)$ as
+before. Again all quantities are related to a single bin in frequency.
+Writing the moments and variance of $L$ as\footnote{The $k$-th moment
+  of variable $x$ is defined as the expectation value $\mu_k =
+  \expect{x^k} = \int \rmdx{x}\,x^k\,p(x),\; k = 0,1,2,\ldots$.} %
 \begin{align}
   \lleq{ram}
-  \mu_1 &\equiv \expect{\Lum}, \qquad 
+  \mu =  \mu_1 &\equiv \expect{\Lum}, \qquad
   \mu_2 \equiv \expect{\Lum^2}, \\
   \sigma^2 &\equiv \variance{\Lum}
-  = \expect{\Lum^2} - \expect{\Lum}^2\,,
-\end{align}
-the first moment and variance of $Q$ are 
+  = \expect{\Lum^2} - \expect{\Lum}^2,
+\end{align}
+it is easy to show that, for the Poissonian compound process
+(\ref{asd}), the first moment and variance of $Q$ are related to those
+of $L$ by
+%%%%%%%%%%%%%%%%
+\fredmargin{citation} %
+%%%%%%%%%%%%%%%
 \begin{align}
   \lleq{raa}
-  \expect{Q} = \lambda\, \mu_1, \qquad 
-  \variance{Q} = \lambda \mu_2 = \lambda (\mu_1^2 + \sigma^2) \,.
-\end{align}
-for \textit{any} $p(L\cond\bmphi)$. 
-%
-\hans{Stopped editing HERE 161202 18h30}
-Furthermore, higher-order quantities such as the skewness and kurtosis
+  \expect{Q} = \lambda \mu, \qquad
+  \variance{Q} = \lambda \mu_2 = \lambda (\mu^2 + \sigma^2) \,.
+\end{align}
+for any probability $p(L\cond\bmphi)$.
+%
+Higher-order quantities such as the skewness and kurtosis
 are suppressed by powers of $\lambda^{-1/2}$.
-% 
-From Theorem 4.3.1 of \cite{bening2002generalized}, we can conclude
+%%%%%%%%%%%
+\hans{The variance is the second so-called \textit{cumulant}. The
+  relation for all orders is that the $k$-th order cumulant
+  $\kappa_q(Q)$ is equal to $\lambda$ times the $k$-th order moment of
+  $L$, $\mu_k$. The skewness is therefore
+  $\kappa_3(Q)/\kappa_2(Q)^{3/2} = \lambda \mu_3/(\lambda \mu_2)^{3/2}
+  = \lambda^{-1/2} \mu_3/\mu_2^{3/2}$. The magnitude of the correction
+  to (rab) due to $\kappa_3$ and higher orders is unclear.}
+%%%%%%%%%%%
+From Theorem 4.3.1 of Ref.~\cite{bening2002generalized}, we conclude
 that the asymptotic distribution of $\Lumtot$ when
 $\lambda \to \infty$, given the first and second moment of $\Lum$, is
+a gaussian or normal distribution,
 \begin{align}
   \lleq{rab}
-  P(\Lumtot \cond \bmphi) 
+  P(\Lumtot \cond \lambda,\bmphi)
   &= \GaussianDist \left( \Lumtot \cond \expect{Q}, \variance{Q} \right) %
-    = \GaussianDist \left( \Lumtot \cond \lambda \mu, \lambda (\mu^2 + \sigma^2) \right).
-\end{align}
-This result is a drastic simplification as it eliminates both the sum
-over $N$ in \refeq{qab} and the integral over $\bmL$ in \refeq{qaf}.
-That allows us to model $\Lumtot$ in each bin \emph{without} making
-concrete assumptions about the functional form of
-$P(\Lum \cond \bmphi)$; i.e., we may now view $\bmphi$ as the
-collection of first and second moments in each bin in addition to
-$\lambda$. Continuing the notation of \refeq{qadd},
+    = \GaussianDist \left( \Lumtot \cond \lambda \mu_1, \lambda (\mu^2 {+} \sigma^2) \right).
+\end{align}
+This drastic simplification eliminates both the sum over $N$ and the
+integral over $\bmL$ in \refeq{asd}.
+
+It remains to find the integrals over $\lambda$ and $\bmphi$ in
+\refeq{asc}. As before, the posterior for $\lambda$ is the Gamma
+distribution of \refeq{pms}. %
+%
+To determine $p(\bmphi\cond n,\bml)$, we note that the result
+(\ref{rab}) also simplifies the calculation of the parameter posterior
+$p(\bmphi\cond n,\bml)$ since, independently of the actual functional
+model for $p(L\cond\bmphi)$ the only relevant information is contained
+in the first moment $\mu$ and variance $\sigma^2$.  As set out in
+Jaynes~\cite[Ch. 7.10]{jaynes2003probability}, knowledge of $\mu$ and
+$\sigma^2$ implies that we can assign a Gaussian distribution to
+$p(L\cond\bmphi)$
+\begin{align}
+  \lleq{rae}
+  p(\Lum \cond \bmphi) =
+  p(\Lum \cond \mu, \sigma^2 ) &= \GaussianDist(\Lum \cond \mu, \sigma^2)
+\end{align}
+even though the actual data histograms may not be Gaussian. The reason
+is that a Gaussian represents our state of knowledge \emph{before}
+data $\bmy$ is observed. With this assignment, the definition of
+$\bmphi$ reduces to $(\mu,\sigma^2)$ and the posterior becomes, using
+Bayes' Theorem,
 \begin{align}
   \lleq{rac}
-  \bmphi =  \left(\lambda_b, \mu_b, \sigma^2_b \right)_{b=1}^B.
-\end{align}
-The price for this flexibility is that $\bmphi$ may now contain many
-parameters, and each one is determined only from the samples in a
-single bin. The joint posterior is
+  p(\mu,\sigma^2\cond n,\bml)
+  &= \frac{
+    \GaussianDist(\bml\cond\mu,\sigma^2)\;
+    p(\mu\cond n)\;
+    p(\sigma^2\cond n)}
+  {p(\bml\cond n)}
+\end{align}
+\hans{I have inserted a uniform prior for $\mu$ here; see comments
+  below. The original Gaussian prior has been retained below but I
+  would suggest that the compliations introduced by it are
+  unnecessary.}  %
+In keeping with prior information that luminositites are always
+positive, we assign an improper uniform prior
+$p(\mu\cond\mu_{\rm max}) = 1/\mu_{\rm max}$ independent of $n$; as
+always, the posterior is independent of $\mu_{\rm max}$. For
+$\sigma^2$, we use an $n$-independent Inverse Gamma distribution prior
+with hyperparameters $(a_0,b_0)$ which reads
 \begin{align}
   \lleq{rad}
-  p(\bmphi \cond \bmy ) = \prod_{b=1}^B p\left(\lambda_b, \mu_b, \sigma^2_b \cond \bmy\right),
-\end{align}
-Focussing on a single bin, we omit the subscript $b$. As before, we
-further assume that
-\begin{align}
-  \lleq{rao}
-p(\lambda, \mu, \sigma^2 \cond \bmy) = p(\lambda \cond \bmy) p(\mu,
-\sigma \cond \bmy).
-\end{align}
-The posterior for $\lambda$ is given by \refeq{pms}.  To learn mean
-and variance of $\Lum$ from $\bmy$, we follow
-Jaynes~\cite[Ch. 7.10]{jaynes2003probability} and assign a Gaussian
-sampling distribution
-\begin{align}
-  \lleq{rae}
-  p(\Lum \cond \mu, \sigma^2 ) &= \GaussianDist(\Lum \cond \mu, \sigma^2)
-\end{align}
-although the actual frequencies may be quite different from a
-Gaussian. Our motivation is that this Gaussian represents our state of
-knowledge \emph{before} observing $\bmy$ well. To make the
-calculations easier, we assume a conjugate prior for unknown $\mu$ and
-$\sigma^2$. The standard result for the posterior is
+  p(\sigma^2\cond a_0,b_0)
+  = \InvGammaDist(\sigma^2\cond a_0,b_0)
+  &= \frac{b_0^{a_0}}{\Gamma(a_0)} (\sigma^2)^{-a_0-1} \exp(-b_0/\sigma^2)
+\end{align}
+and is ``conjugate'' to the normal distribution likelihood of
+\refeq{rae} in that the posterior for $(\mu,\sigma^2)$ is also a
+Gaussian in $\mu$ and an Inverse Gamma in $\sigma^2$, but with
+parameters updated by the data,
+\begin{align}
+  \lleq{rbd}
+  p(\mu,\sigma^2\cond n,\bml,a_0,b_0)
+  &= \GaussianDist(\mu\cond \mu_n,\sigma_n^2)
+  \cdot
+  \InvGammaDist(\sigma^2\cond a_n,b_n) \\
+  \mu_n &= \langle\ell\rangle\\
+  \sigma_n^2 &= \sigma^2/n \\
+  a_n &= a_0 + \tfrac{1}{2}n \\
+  b_n &= b_0 + \tfrac{1}{2} n\left( \langle\ell^2\rangle - \langle\ell\rangle^2\right)
+\end{align}
+with data sample moments $\langle\lum^k\rangle$ defined as $(1/n)
+\sum_{j=1}^n \lum_j^k$.
+%
+Inserting Eqs.~(\ref{pms}) and (\ref{rbd}) into \refeq{asc}, we obtain
+\begin{align}
+  \lleq{ral}
+  p(\Lumtot \cond n,\bml)
+  &= \int \rmdx{\lambda}\; \rmdx{\mu}\; \rmdx{\sigma^2}\;
+  \GaussianDist \left( \Lumtot \cond \lambda \mu, \lambda (\sigma^2{+}\mu^2) \right)
+  \times\\
+  &\quad\times
+  \GammaDist(\lambda \cond n{-}a{+}1 , 1) \cdot
+  \GaussianDist(\mu\cond \mu_n,\sigma_n^2) \cdot
+  \InvGammaDist(\sigma^2\cond a_n,b_n) \;
+  \nonumber
+\end{align}
+While the integral cannot be carried out analytically, the integrand
+is unimodal and strongly peaked, so that the Laplace or saddle-point
+approximation~\cite[Ch. 27]{mackay2003information} can be applied.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\hans{Some comments on the numerical values of hyperparameters $a$ in Poisson, $a,b$
+  in inverse gamma needed.} %
+\hans{I would omit the text and calculations below. Apart from the
+  extra parameters, the main reason is that the Gaussian prior in
+  $\mu$ is not consistent with the fact that all luminosities are
+  positive, while the gaussian permits negative $\mu$. This may not be
+  important numerically, but the uniform prior in $\mu$ does
+  circumvent this problem and is moreover less informative than the
+  Gaussian prior. } %
+\hans{Note also that letting $\mu_0{=}0$ and $ n_0\to 0$ below implies
+  a prior gaussian of infinite width centered on zero, i.e.\ half of
+  the values of $\mu$ will be negative. Could this be the reason why
+  the numerics pick up a factor two?}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+To simplify the calculations, we assume a conjugate prior for unknown
+$\mu$ and $\sigma^2$. The standard result for the posterior is a
+Gaussian for $\mu$ and an Inverse Gamma distribution for $\sigma^2$,
 \begin{align}
   \lleq{raf}
-  p(\mu, \sigma^2 \cond \bmy) &= \GaussianDist(\mu \cond \mu_0^\prime, \frac{\sigma^2}{n_0^\prime}) \InvGammaDist\left( \sigma^2 \cond \frac{\nu_0^\prime}{2}, \frac{\nu_0^\prime \sigma_0^{\prime 2}}{2} \right)
-\end{align}
-where the hyperparameters are updated from prior to posterior based on $\bmy$ as
-\begin{align}
-  \lleq{rag}
-  \expectest{\lum} &= \frac{1}{n} \sum_{i=1}^n \lum_i,\\
-  \lleq{rah}  n_0^\prime &= n_0 + n,\\
-  \lleq{rai}    \mu_0^\prime &= \frac{n_0 \mu_0 + n \expectest{\lum}}{n^\prime},\\
-  \lleq{raj}    \nu_0^\prime &= \nu_0 + n,\\
-  \lleq{rak}    \nu_0^\prime \sigma_0^{\prime 2} &= \nu_0 \sigma_0^{ 2} + \sum_{i=1}^n \left(\lum_j - \expectest{\lum}\right)^2 + \frac{n_0 n}{n^\prime} \left(\mu_0 - \expectest{\lum}\right)\,.
-  \end{align}
-  By default, we take a noninformative prior and set
-  $n_0 = \mu_0 = \nu_0 = 0$, so the value $\sigma_0^2>0$ is
-  irrelevant.
+  p(\mu, \sigma^2 \cond \bmy)
+  &= \GaussianDist\left(\mu \cond \mu_0^\prime, \sigma^2/n_0^\prime\right) %
+    \InvGammaDist\left( \sigma^2 \,\Bigl|\, \tfrac{1}{2}\nu_0^\prime,
+    \tfrac{1}{2}\nu_0^\prime \sigma_0^{\prime 2} \right)
+\end{align}
+where the hyperparameters $(n_0',\mu_0',\nu_0')$ are updated from
+prior to posterior based on $\bmy$ with the sample mean
+$\langle\lum\rangle = (1/n) \sum_{j=1}^n \lum_j$,
+\begin{align}
+  %\lleq{rag}
+  \lleq{rah}  n_0^\prime
+  &= n_0 + n,\\
+  \lleq{rai}    \mu_0^\prime
+  &= \frac{n_0 \mu_0 + n \langle\lum\rangle}{n_0^\prime},\\
+  \lleq{raj}    \nu_0^\prime
+  &= \nu_0 + n,\\
+  \lleq{rak}
+  \nu_0^\prime \sigma_0^{\prime 2}
+  &= \nu_0 \sigma_0^{ 2}
+    + \sum_{j=1}^n \left(\lum_j - \langle\lum\rangle\right)^2
+    + \frac{n_0 n}{n_0^\prime} \left(\mu_0 - \langle\lum\rangle\right)^2\,.
+\end{align}
+By default, we take a noninformative prior and set $n_0 = \mu_0 =
+\nu_0 = 0$, so the value $\sigma_0^2>0$ is irrelevant. %
   % The posterior
@@
   % the Jacobian determinant is one.
-
-  Finally all ingredients for the prediction of $\Lumtot$ in one bin are
-  ready, and \refeq{qaa} simplifies to
+The above equations also show that the data $\bmy$ appears only in the
+form of $n$ and the first two sample moments
+$m_1 = \langle\ell\rangle$ and $m_2 = \langle\ell^2\rangle$.
+Inserting all these ingredients into Eq.~(\ref{asc}), we obtain
 \begin{align}
   \lleq{ral}
-  p(\Lumtot \cond \bmy) &= \int \rmdx{\lambda} \rmdx{\mu} \rmdx{\sigma^2} \GaussianDist \left( \Lumtot \cond \lambda \mu, \lambda \sigma^2 \right) p(\lambda \cond \bmy) p( \mu, \sigma^2 \cond \bmy) \,,
-\end{align}
-where $ p(\lambda \cond \bmy)$ is given by \refeq{pms} and
-$ p( \mu, \sigma^2 \cond \bmy)$ by
-\refeq{raf}--\refeq{rak}.
+  p(\Lumtot \cond \bmy)
+  &= \int \rmdx{\lambda}\; \rmdx{\mu}\; \rmdx{\sigma^2}\;
+    p(\lambda \cond n)\; p( \mu, \sigma^2 \cond \bmy)\;
+    \GaussianDist \left( \Lumtot \cond \lambda \mu, \lambda (\sigma^2{+}\mu^2) \right)
+    \,,
+\end{align}
+where $ p(\lambda \cond n)$ is given by \refeq{pms} and
+$ p( \mu, \sigma^2 \cond \bmy)$ by Eqs.~(\ref{raf})--(\ref{rak}).
 
@@
 
-Unfortunately we could not integrate \refeq{ral} analytically but
-since the integrand is unimodal and strongly peaked the Laplace or
+\hans{H: If threefold integrals must be carried out for the
+  expectation values, why not do so for Eq (ral) directly rather than
+  just the expectation values?} %
+
+While the integral in \refeq{ral} cannot be integrated analytically,
+the integrand is unimodal and strongly peaked, so that the Laplace or
 saddle-point approximation~\cite[Ch. 27]{mackay2003information} can be
-applied in two ways. First, if only the value
-$p(\Lumtot=\lumtot \cond \bmy)$ for some particular $q$ is sought,
-then one can approximate it directly. Second, we can compute the
-moments of $\Lumtot$ through the \emph{moment-generating
-  function}\fredmargin{ref}
+applied. If only the value $p(\Lumtot{=}\lumtot \cond n,m_1,m_2)$ at the
+total data luminosity $q$ is sought, then one can approximate it
+directly. In addition, we can compute the moments of $\Lumtot$ through
+the \emph{moment generating function} \cite{stuart1994kendall}
 \begin{align}
@@
 \end{align}
-
 Assuming all involved integrals are absolutely convergent, we can
@@
   \lleq{ras}
-  \expect{\Lumtot} &= \int \rmdx{\lambda} \rmdx{\mu} \rmdx{\sigma^2} p(\lambda \cond \bmy) p( \mu, \sigma^2 \cond \bmy) \qty(-\eval{\dv{f(t \cond  \lambda, \mu, \sigma^2)}{t}}_{t=0})\\
-  \variance{\Lumtot} &= \int \rmdx{\lambda} \rmdx{\mu} \rmdx{\sigma^2} p(\lambda \cond \bmy) p( \mu, \sigma^2 \cond \bmy) \qty(\eval{\dv[2]{f(t \cond  \lambda, \mu, \sigma^2)}{t}}_{t=0}) - \expect{\Lumtot}^2
+  \expect{\Lumtot} &= \int \rmdx{\lambda} \rmdx{\mu} \rmdx{\sigma^2} p(\lambda \cond n) p( \mu, \sigma^2 \cond \bmy) \qty(-\eval{\dv{f(t \cond  \lambda, \mu, \sigma^2)}{t}}_{t=0})\\
+  \variance{\Lumtot} &= \int \rmdx{\lambda} \rmdx{\mu} \rmdx{\sigma^2} p(\lambda \cond n) p( \mu, \sigma^2 \cond \bmy) \qty(\eval{\dv[2]{f(t \cond  \lambda, \mu, \sigma^2)}{t}}_{t=0}) - \expect{\Lumtot}^2
 \end{align}
@@
   \lleq{rat}
-  \qty(\lambda, \mu, \sigma^2) = 1.0005 (n, \expectest{\lum}, \widehat{\variance{\lum}}).
-\end{align}
-\fred{Did we define expectation value and variance? Should be in notation or intro section}
-\fred{At least for us, I should do a comparison plot to evaluate $p(\Lumtot \cond \bmy)$ with both methods }
-\fred{We need a plot, the one to compare predictions and 250 real replicas or 100 virtual replicas?}
+  \qty(\lambda, \mu, \sigma^2) = 1.0005 (n, \langle\lum\rangle, \langle\lum^2\rangle{-}\langle\lum\rangle^2).
+\end{align}
+\fred{Did we define expectation value and variance? Should be in notation or intro section. Hans: see footnote 2.} %
+\fred{At least for us, I should do a comparison plot to evaluate $p(\Lumtot \cond \bmy)$ with both methods } %
+\fred{We need a plot, the one to compare predictions and 250 real replicas or 100 virtual replicas?} %
 \fred{Advantage of this approach: no need to store all packets anymore, sum and sum of squares can be accumulated online $\Rightarrow$ reduced computation and storage requirements}
@@
 
+TODO
+
 \section{Appendix} \label{sec:appendix}
+
+\fred{Will be removed later}
 
@@
 \end{document}
-\subsection{Old}
-%old text
-The joint distribution then is
-\begin{align}
-  \lleq{cac}
-  % Q, N \sim P(Q, N \cond \bmphi, )
-  (N,\bmL) &\ \sim\ p(N,\bmL\cond \bmphi,\lambda, \nu)
-  = p(\bmL\cond N,\bmphi, \nu)\; p(N\cond \lambda) = \prod_{j=1}^N p(\Lum_j\cond \bmphi, \nu) \; p(N\cond \lambda),
-\end{align}
-\fred{Do we need this equation?}
-where we used that the Poisson term does not depend on $\nu$
-explicitly. In general, the distribution of $\Lum_j$ could be a of a
-different functional form in each bin. For simplicity, we assume it is
-identical but may depend on parameters that differ from bin to bin. In
-practice, the distribution of packets usually does not change abruptly
-with frequency. We therefore assume that any parameter needed to fully
-specify $p(\Lum_j)$ is given explicitly as a continuous function of $\nu$
-and the unknown parameters $\bmphi$. This allows us to use a small
-number of parameters to describe $p(\Lumtot)$ across the entire frequency
-range instead of having a few parameters per bin which could lead to
-tens of thousands of parameters for the applications we have in
-mind. So we determine $\bmphi$ from fitting all simulated packets
-simultaneously whereas $\lambda$ is a per-bin quantity.
-\fred{Reformulate when asymptotic approx. included: it has thousands of parameters}
-
-Our goal of predicting $\Lumtot$ in some bin given the simulation output
-$(\npack,\bmnu, \bml)$ is captured in the probability density
-$p(\Lumtot\cond \nu, \npack,\bmnu, \bml)$ with all relevant parameters
-integrated out. To simplify the treatment, we assume that the
-frequency bins are narrow or equivalently that $p(\Lumtot \cond \nu, \npack,
-\bmnu, \bml)$ does not vary within a bin. This implies that every
-sample $\Lum_j$ in the bin follows the same distribution.  As the
-reference frequency, we choose the bin center.
-
-% From \refeq{caa}, $N$ does not depend on $\bml$ at all and only
-% indirectly depends on $\bmnu$ through $n$.
-Since $\Lumtot$ is fully determined by $(N,\bmL)$, it does not depend on
-$(\npack,\bml,\bmnu)$, i.e.
-\begin{align}
-  \lleq{pmi}
-  p(\Lumtot\cond N,\bmL,\nu,\npack,\bmnu, \bml)
-  = p(\Lumtot\cond N, \bmL)
-  &= \delta(\Lumtot - \textstyle\sum_{j=1}^N \Lum_j),
-\end{align}
-\fred{Uniform order: $\nu$ first, then $\ell$, compare to $S$} where
-each $\Lum_j$ is assumed to occur at $\nu$.  Using the law of total
-probability repeatedly, we can write the desired distribution as
-\fred{Discuss $N=0$ case somewhere}
-\begin{align}
-  \lleq{pmj}
-  p(\Lumtot\cond \nu, \npack,\bmnu, \bml)
-  &= \sum_N p(N\cond \nu, \npack,\bmnu, \bml) \, p(\Lumtot\cond N,\nu, \npack,\bmnu, \bml) \\
-  \lleq{pmjb}
-  &= \sum_N p(N\cond n)\,\int \rmdx{\bmL} p(\Lumtot\cond N,\bmL,\nu, \npack,\bmnu, \bml)\, p(\bmL\cond N,\nu, \npack,\bmnu, \bml)
-  \\
-  \lleq{pmjc}
-  &= \sum_N p(N\cond n)\,\int \rmdx{\bmL} \delta(\Lumtot - \textstyle\sum_{j=1}^N \Lum_j)
-  \, p(\bmL\cond N,\nu, \npack,\bmnu, \bml).
-\end{align}
-The dependence on the simulation output $(\npack,\bmnu, \bml)$ enters
-through $p(\bmL\cond N,\npack,\bmnu, \bml)$ in terms of the parameters
-$\bmphi$ of $p(\bmL\cond N,\bmphi)$,
-\begin{align}
-  p(\bmL\cond N,\nu, \npack,\bmnu, \bml)
-  &= \int \rmdx{\bmphi} p(\bmL\cond \bmphi,N, \nu, \npack,\bmnu, \bml)\,p(\bmphi\cond N,\nu, \npack,\bmnu, \bml) \\
-  \lleq{pmk}
-  &= \int \rmdx{\bmphi} p(\bmL\cond \bmphi, N, \nu)\,p(\bmphi\cond N,\nu, \npack,\bmnu, \bml).
-\end{align}
-because $\bmL$ is fully determined by $N,\bmphi, \nu$ and hence
-independent of $\npack,\bmnu, \bml$.  With Bayes' theorem we learn about
-$\bmphi$ from the simulation output
-\begin{align}
-  \lleq{pmkb}
-  p(\bmphi\cond N,\nu, \npack,\bmnu, \bml) &=
-  \frac{p(\bml\cond \bmphi, N,\nu, \npack,\bmnu)\,p(\bmphi\cond N,\nu, \npack,\bmnu)} %
-  {\int \rmdx{\bmphi} p(\bml\cond \bmphi,N,\nu, \npack,\bmnu)\,p(\bmphi\cond N,\nu, \npack,\bmnu)}.
-\end{align}
-By using a multiplicity- and frequency-independent prior for $\bmphi$,
-\begin{align}
-  \lleq{pml}
-  p(\bmphi\cond N,\nu, \npack,\bmnu) &= p(\bmphi),
-\end{align}
-and by recognizing that $\bml$ is independent of $N$ and $\nu$
-\begin{align}
-  \lleq{pmla}
-  p(\bml\cond \bmphi,N,\nu, \npack,\bmnu) = p(\bml\cond \bmphi,\npack,\bmnu)
-\end{align}
-we can write \refeq{pmk} as
-\begin{align}
-  \lleq{pmn}
-  p(\bmL\cond N,\nu, \npack,\bmnu, \bml)
-  &= \int \rmdx{\bmphi} p(\bmL\cond N,\bmphi)\, %
-  \frac{p(\bml\cond \bmphi, \npack, \bmnu)\,p(\bmphi)} %
-  {\int \rmdx{\bmphi} p(\bml\cond \bmphi,\npack,\bmnu)\,p(\bmphi)} \\
-  \lleq{pmo}
-  &= \frac{1}  {p(\bml\cond \npack, \bmnu)} %
-  \int \rmdx{\bmphi} p(\bmL\cond \bmphi, N)\, %
-  p(\bml\cond \bmphi, \npack,\bmnu)\,p(\bmphi)
-\end{align}
-where $p(\bml\cond \npack, \bmnu) = \int \rmdx{\bmphi} p(\bml\cond
-\bmphi, \npack,\bmnu)\,p(\bmphi)$ is the evidence.  Inserting \refeq{pmo}
-into \refeq{pmjc}, we obtain the desired master equation:
-\begin{equation}
-  \lleq{pmp}
-  \boxed{
-  p(\Lumtot\cond \nu, \npack,\bmnu, \bml)
-  = \frac{1}{p(\bml\cond \npack, \bmnu)}
-  \sum_N p(N\cond n)\,\int \rmdx{\bmL} \delta(\Lumtot - \textstyle\sum_j \Lum_j)
-  \displaystyle\int \rmdx{\bmphi} p(\bmL\cond \bmphi, N, \nu) \, p(\bml\cond \bmphi,\npack,\bmnu)
-  \, p(\bmphi)
-  }
-\end{equation}
-with
-\begin{align}
-  \lleq{pmq}
-  p(\bmL\cond \bmphi, N, \nu) &= \prod_{j=1}^N p(\Lum_j\cond \bmphi, \nu), \\
-  \lleq{pmr}
-  p(\bml\cond \bmphi,\npack,\bmnu) &= \prod_{i=1}^{\npack} p(\ell_i\cond \bmphi, \nu_i)
-\end{align}
-and $p(\cdot \cond \bmphi, \cdot)$ is the same function in both
-equations.  In the next section, we compute the general result for
-$P(N \cond n)$. Equation \refeq{pmp} then implies that once we specify
-the prior $p(\bmphi)$ and the likelihood $p(\Lum_j\cond \bmphi, \nu)$, we
-have a prediction for $\Lumtot$ which takes into account all possible $N$,
-$\lambda$ and $\bmphi$. Both likelihood and prior depend on the
-simulation under consideration but a general requirement to render our
-approach tractable is that the distribution of a single packet
-luminosity $P(\Lum_j \cond \bmphi, \nu)$ should be such that the integral
-over $\bmL$ has a closed-form solution. In other words, for a given
-distribution of $\Lum_j$, the distribution of $\Lumtot = \sum_N \Lum_j$ should be
-known.
-%  An important special case is the class of \emph{stable}
-% distributions.
-
-We use Bayes' theorem to find the posterior distribution for
-
-% Local Variables:
-% compile-command:"rubber --pdf -W refs -S draft"
-% End:

### 2016-12-03 real/code/Asymptotic.jl

@@ -19,6 +19,7 @@ function targetfactory(target::TARGET, n::Int64, xmean::Float64, xsumsq::Float64
     (xmean > 0) || error("nonpositive xmean $xmean")
     (xsumsq > 0) || error("nonpositive xsumsq $xsumsq")
 
+    """ with X (or Q) dependence, only contains λ, μ, σSq posterior"""
     function logposterior(x::Vector)
         λ, μ, σSq = x
 
@@ -40,6 +41,8 @@ function targetfactory(target::TARGET, n::Int64, xmean::Float64, xsumsq::Float64
     """log(-d/dt f(target=0|x))"""
     function logfirst(x::Vector)
         λ, μ, σSq = x
+        # bug fix
+        σSq += μ^2
         exponent = λ*μ^2 / (2*σSq)
         tmp = exp(-exponent) * sqrt(λ * σSq / (2 * π)) + 1/2 * λ * μ * (1 + erf(sqrt(exponent)))
         log(tmp)
@@ -48,6 +51,8 @@ function targetfactory(target::TARGET, n::Int64, xmean::Float64, xsumsq::Float64
     """log(d²/dt² f(target=0|x))"""
     function logsecond(x::Vector)
         λ, μ, σSq = x
+        # bug fix
+        σSq += μ^2
         exponent = λ*μ^2 / (2*σSq)
         tmp = exp(-exponent) * sqrt(2 / π * λ^3 * σSq) * μ + λ*(λ*μ^2 + σSq)*(1 + erf(sqrt(exponent)))
         log(1/2 * tmp)
@@ -56,6 +61,8 @@ function targetfactory(target::TARGET, n::Int64, xmean::Float64, xsumsq::Float64
     """log(-d^3/dt^3 f(target=0|x))"""
     function logthird(x::Vector)
         λ, μ, σSq = x
+        # bug fix
+        σSq += μ^2
         exponent = λ*μ^2 / (2*σSq)
         tmp = exp(-exponent) / sqrt(2π) * λ^(3/2) * sqrt(σSq) * (λ*μ^2 + 2σSq)
         tmp += 1/2 * λ * μ * (λ*μ^2 + 3σSq)*(1 + erf(sqrt(exponent)))
@@ -65,6 +72,8 @@ function targetfactory(target::TARGET, n::Int64, xmean::Float64, xsumsq::Float64
[…]

### 2016-12-03 real/draft.tex
[draft] [references] update asymptotic approximation

--- 2016-12-02 real/draft.tex
@@
 %% CHANGE THIS DATE AS YOU MODIFY THE FILE
-\newcommand{\vdate}{161202}
+\newcommand{\vdate}{161203}
 \pagestyle{myheadings}
@@
 \begin{align}
-  \lleq{pmd}
+  \lleq{cab}
   p(\lambda\cond n) %
@@
 % \end{align}
-and inserting Eqs.~(\ref{caa}) and (\ref{pmd}), the prediction for $N$
+and inserting Eqs.~(\ref{caa}) and (\ref{cab}), the prediction for $N$
 given $n$ is
@@
 \end{align}
-with $p(\bmL\cond n,\bmphi) = \prod_{j=1}^n p(L\cond\bmphi)$ as
-before.  If, as assumed, $p(N\cond\lambda)$ is Poissonian, there is a
-simple relation between their respective moments\footnote{The $k$-th
-  moment of a variable is defined as
-  $\mu_k = \expect{x^k} = \int dx\,p(x)\,x^k, k = 1,2,\ldots$, while
-  the variance is equal to $\sigma^2 = \mu_2 - \mu_1^2$.} %
-for $p(Q\cond \lambda,\bmphi)$ and $p(L\cond \bmphi)$ as follows.
-%%%%%%%%%%%%%%%
-\fred{citation} %
-%%%%%%%%%%%%%%%
-In terms of the moments and variance of $L$,
+with $p(\bmL\cond n,\bmphi) = \prod_{j=1}^n p(L_j\cond\bmphi)$ as
+before. Writing the moments and variance of $L$ as\footnote{The $k$-th
+  moment of variable $x$ is defined as the expectation value
+  $\mu_k = \expect{x^k} = \int dx\,x^k\,p(x),\; k = 0,1,2,\ldots$.} %
 \begin{align}
   \lleq{ram}
-  \mu_1 &\equiv \expect{\Lum}, \qquad 
+  \mu =  \mu_1 &\equiv \expect{\Lum}, \qquad 
   \mu_2 \equiv \expect{\Lum^2}, \\
   \sigma^2 &\equiv \variance{\Lum} 
-  = \expect{\Lum^2} - \expect{\Lum}^2\,,
-\end{align}
-the first moment and variance of $Q$ are 
+  = \expect{\Lum^2} - \expect{\Lum}^2,
+\end{align}
+it is easy to show that, for the Poissonian compound process
+(\ref{asd}), the first moment and variance of $Q$ are related to those
+of $L$ by
+%%%%%%%%%%%%%%%%
+\fredmargin{citation} %
+%%%%%%%%%%%%%%%
 \begin{align}
   \lleq{raa}
-  \expect{Q} = \lambda\, \mu_1, \qquad 
-  \variance{Q} = \lambda \mu_2 = \lambda (\mu_1^2 + \sigma^2) \,.
-\end{align}
-for \textit{any} $p(L\cond\bmphi)$. 
+  \expect{Q} = \lambda \mu, \qquad 
+  \variance{Q} = \lambda \mu_2 = \lambda (\mu^2 + \sigma^2) \,.
+\end{align}
+for any probability $p(L\cond\bmphi)$.
 %
-\hans{Stopped editing HERE 161202 18h30}
 Furthermore, higher-order quantities such as the skewness and kurtosis
 are suppressed by powers of $\lambda^{-1/2}$.
-% 
-From Theorem 4.3.1 of \cite{bening2002generalized}, we can conclude
+%%%%%%%%%%%
+\hans{The variance is the second so-called \textit{cumulant}. The
+  relation for all orders is that the $k$-th order cumulant
+  $\kappa_q(Q)$ is equal to $\lambda$ times the $k$-th order moment of
+  $L$, $\mu_k$. The skewness is therefore
+  $\kappa_3(Q)/\kappa_2(Q)^{3/2} = \lambda \mu_3/(\lambda \mu_2)^{3/2}
+  = \lambda^{-1/2} \mu_3/\mu_2^{3/2}$. The magnitude of the correction
+  to (rab) due to $\kappa_3$ and higher orders is unclear.}
+%%%%%%%%%%%
+From Theorem 4.3.1 of Ref.~\cite{bening2002generalized}, we conclude
 that the asymptotic distribution of $\Lumtot$ when
 $\lambda \to \infty$, given the first and second moment of $\Lum$, is
+a gaussian or normal distribution,
 \begin{align}
   \lleq{rab}
-  P(\Lumtot \cond \bmphi) 
+  P(\Lumtot \cond \lambda,\bmphi) 
   &= \GaussianDist \left( \Lumtot \cond \expect{Q}, \variance{Q} \right) %
-    = \GaussianDist \left( \Lumtot \cond \lambda \mu, \lambda (\mu^2 + \sigma^2) \right).
-\end{align}
-This result is a drastic simplification as it eliminates both the sum
-over $N$ in \refeq{qab} and the integral over $\bmL$ in \refeq{qaf}.
-That allows us to model $\Lumtot$ in each bin \emph{without} making
-concrete assumptions about the functional form of
-$P(\Lum \cond \bmphi)$; i.e., we may now view $\bmphi$ as the
-collection of first and second moments in each bin in addition to
-$\lambda$. Continuing the notation of \refeq{qadd},
+    = \GaussianDist \left( \Lumtot \cond \lambda \mu_1, \lambda (\mu^2 {+} \sigma^2) \right).
+\end{align}
+This drastic simplification eliminates both the sum over $N$ and the
+integral over $\bmL$ in \refeq{asd} and allows us to model $\Lumtot$
+in each bin without having to make concrete assumptions about the
+functional form of $P(\Lum \cond \bmphi)$. Correspondingly, we may now
+redefine $\bmphi$ as the moment $\mu$ and variance $\sigma^2$ of
+$p(L\cond\bmphi)$,
 \begin{align}
   \lleq{rac}
-  \bmphi =  \left(\lambda_b, \mu_b, \sigma^2_b \right)_{b=1}^B.
+  \bmphi =  (\mu, \sigma^2).
 \end{align}
@@
 \end{align}
-although the actual frequencies may be quite different from a
+although the actual data histograms may be quite different from a
 Gaussian. Our motivation is that this Gaussian represents our state of
-knowledge \emph{before} observing $\bmy$ well. To make the
-calculations easier, we assume a conjugate prior for unknown $\mu$ and
-$\sigma^2$. The standard result for the posterior is
+knowledge \emph{before} data $\bmy$ is observed. To simplify the
+calculations, we assume a conjugate prior for unknown $\mu$ and
+$\sigma^2$. The standard result for the posterior is a Gaussian for
+$\mu$ and an Inverse Gamma distribution for $\sigma^2$,
 \begin{align}
   \lleq{raf}
-  p(\mu, \sigma^2 \cond \bmy) &= \GaussianDist(\mu \cond \mu_0^\prime, \frac{\sigma^2}{n_0^\prime}) \InvGammaDist\left( \sigma^2 \cond \frac{\nu_0^\prime}{2}, \frac{\nu_0^\prime \sigma_0^{\prime 2}}{2} \right)
-\end{align}
-where the hyperparameters are updated from prior to posterior based on $\bmy$ as
-\begin{align}
-  \lleq{rag}
-  \expectest{\lum} &= \frac{1}{n} \sum_{i=1}^n \lum_i,\\
-  \lleq{rah}  n_0^\prime &= n_0 + n,\\
-  \lleq{rai}    \mu_0^\prime &= \frac{n_0 \mu_0 + n \expectest{\lum}}{n^\prime},\\
-  \lleq{raj}    \nu_0^\prime &= \nu_0 + n,\\
-  \lleq{rak}    \nu_0^\prime \sigma_0^{\prime 2} &= \nu_0 \sigma_0^{ 2} + \sum_{i=1}^n \left(\lum_j - \expectest{\lum}\right)^2 + \frac{n_0 n}{n^\prime} \left(\mu_0 - \expectest{\lum}\right)\,.
-  \end{align}
+  p(\mu, \sigma^2 \cond \bmy) 
+  &= \GaussianDist\left(\mu \cond \mu_0^\prime, \sigma^2/n_0^\prime\right) %
+    \InvGammaDist\left( \sigma^2 \,\Bigl|\, \tfrac{1}{2}\nu_0^\prime, 
+    \tfrac{1}{2}\nu_0^\prime \sigma_0^{\prime 2} \right)
+\end{align}
+where the hyperparameters $(n_0',\mu_0',\nu_0')$ are updated from
+prior to posterior based on $\bmy$ with the sample mean
+$\langle\lum\rangle = (1/n) \sum_{j=1}^n \lum_j$,
+\begin{align}
+  %\lleq{rag}
+  \lleq{rah}  n_0^\prime 
+  &= n_0 + n,\\
+  \lleq{rai}    \mu_0^\prime 
+  &= \frac{n_0 \mu_0 + n \langle\lum\rangle}{n_0^\prime},\\
+  \lleq{raj}    \nu_0^\prime 
+  &= \nu_0 + n,\\
+  \lleq{rak}    
+  \nu_0^\prime \sigma_0^{\prime 2} 
+  &= \nu_0 \sigma_0^{ 2} 
+    + \sum_{j=1}^n \left(\lum_j - \langle\lum\rangle\right)^2 
+    + \frac{n_0 n}{n_0^\prime} \left(\mu_0 - \langle\lum\rangle\right)^2\,.
+\end{align}
+\hans{H: The last term in (\ref{rak}) was missing a square. Even so, I
+  have not yet been able to verify that equation (rak) from
+  Wikipedia's Normal Distribution entry agrees with Eq.(200) of
+  Murphy's 2007 paper.} %
 By default, we take a noninformative prior and set
-  $n_0 = \mu_0 = \nu_0 = 0$, so the value $\sigma_0^2>0$ is
-  irrelevant.
+$n_0 = \mu_0 = \nu_0 = 0$, so the value $\sigma_0^2>0$ is irrelevant.
   % The posterior
@@
   % the Jacobian determinant is one.
-
-  Finally all ingredients for the prediction of $\Lumtot$ in one bin are
-  ready, and \refeq{qaa} simplifies to
+The above equations also show that the data $\bmy$ appears only in the
+form of $n$ and the first two sample moments
+$m_1 = \langle\ell\rangle$ and $m_2 = \langle\ell^2\rangle$.
+Inserting all these ingredients into Eq.~(\ref{asc}), we obtain
 \begin{align}
   \lleq{ral}
-  p(\Lumtot \cond \bmy) &= \int \rmdx{\lambda} \rmdx{\mu} \rmdx{\sigma^2} \GaussianDist \left( \Lumtot \cond \lambda \mu, \lambda \sigma^2 \right) p(\lambda \cond \bmy) p( \mu, \sigma^2 \cond \bmy) \,,
-\end{align}
-where $ p(\lambda \cond \bmy)$ is given by \refeq{pms} and
-$ p( \mu, \sigma^2 \cond \bmy)$ by
-\refeq{raf}--\refeq{rak}.
+  p(\Lumtot \cond \bmy) 
+  &= \int d\lambda\; d\mu\; d\sigma^2\;
+    p(\lambda \cond n)\; p( \mu, \sigma^2 \cond \bmy)\;
+    \GaussianDist \left( \Lumtot \cond \lambda \mu, \lambda (\sigma^2{+}\mu^2) \right)
+    \,,
+\end{align}
+where $ p(\lambda \cond n)$ is given by \refeq{pms} and
+$ p( \mu, \sigma^2 \cond \bmy)$ by Eqs.~(\ref{raf})--(\ref{rak}).
 
@@
 
-Unfortunately we could not integrate \refeq{ral} analytically but
-since the integrand is unimodal and strongly peaked the Laplace or
+While the integral in \refeq{ral} cannot be integrated analytically,
+the integrand is unimodal and strongly peaked, so that the Laplace or
 saddle-point approximation~\cite[Ch. 27]{mackay2003information} can be
-applied in two ways. First, if only the value
-$p(\Lumtot=\lumtot \cond \bmy)$ for some particular $q$ is sought,
-then one can approximate it directly. Second, we can compute the
-moments of $\Lumtot$ through the \emph{moment-generating
-  function}\fredmargin{ref}
+applied. If only the value $p(\Lumtot{=}\lumtot \cond n,m_1,m_2)$ at the
+total data luminosity $q$ is sought, then one can approximate it
+directly. In addition, we can compute the moments of $\Lumtot$ through
+the \emph{moment generating function} \cite{stuart1994kendall}
 \begin{align}
@@
   \lleq{rat}
-  \qty(\lambda, \mu, \sigma^2) = 1.0005 (n, \expectest{\lum}, \widehat{\variance{\lum}}).
-\end{align}
-\fred{Did we define expectation value and variance? Should be in notation or intro section}
-\fred{At least for us, I should do a comparison plot to evaluate $p(\Lumtot \cond \bmy)$ with both methods }
-\fred{We need a plot, the one to compare predictions and 250 real replicas or 100 virtual replicas?}
+  \qty(\lambda, \mu, \sigma^2) = 1.0005 (n, \langle\lum\rangle, \langle\lum^2\rangle{-}\langle\lum\rangle^2).
+\end{align}
+\hans{H: If threefold integrals must be carried out for the expectation values, why not do so
+for Eq (ral) directly rather than just the expectation values?} %
+\fred{Did we define expectation value and variance? Should be in notation or intro section. Hans: see footnote 2.} %
+\fred{At least for us, I should do a comparison plot to evaluate $p(\Lumtot \cond \bmy)$ with both methods } %
+\fred{We need a plot, the one to compare predictions and 250 real replicas or 100 virtual replicas?} %
 \fred{Advantage of this approach: no need to store all packets anymore, sum and sum of squares can be accumulated online $\Rightarrow$ reduced computation and storage requirements}

### 2016-12-07 real/draft.tex
[draft] Sec 6.1 modified to uniform sigma^2 prior plus smaller edits. Old text following \end{document} removed.
[draft] `data` is plural

--- 2016-12-03 real/draft.tex
@@
 
-\item Luminosity data generated by a single simulation provides a
+\item Luminosity data generated by a single simulation provide a
   basis for predicting its uncertainty, i.e.\ how much this luminosity
@@
 which will often come in the form of a list of $n$ measurements,
-$\bmy = (y_1,y_2,\ldots,y_n)$. In order to understand or explain this
+$\bmy = (y_1,y_2,\ldots,y_n)$. In order to understand or explain these
 data, the observer constructs one or more \textit{models} or
@@
 where $p(\bmtheta)$, the so-called \textit{prior}, is the probability
-assigned to the values of the parameters before any data was taken
+assigned to the values of the parameters before any data were taken
 into account and the \textit{evidence}
@@
 \end{align}
-Within a given model $\hypo$, the data $\bmy$ is only one of many
+Within a given model $\hypo$, the data $\bmy$ are only one of many
 possible realisations of the corresponding random variables; for
@@
 \end{align}
-Data $n$ is naturally presumed to have arisen from the same Poisson
+The observed $n$ is naturally presumed to have arisen from the same Poisson
 distribution with the same $\lambda$, so $p(n\cond\lambda) =
@@
   % the Jacobian determinant is one.
-The above equations also show that the data $\bmy$ appears only in the
+The above equations also show that the data $\bmy$ appear only in the
 form of $n$ and the first two sample moments

### 2016-12-16 real/code/Asymptotic.jl
[code] Minore playing around

@@ -80,7 +80,7 @@ function targetfactory(target::TARGET, n::Int64, xmean::Float64, xsumsq::Float64
         var = σSq + μ^2
         exponent = λ*μ^2 / (2*var)
         tmp = exp(-exponent) * sqrt(λ * var / (2 * π)) + 1/2 * λ * μ * (1 + erf(sqrt(exponent)))
-        tmp > 0.0 || error(λ, μ, σSq )
+        tmp > 0.0 || error("Negative input ", λ, μ, σSq )
         log(tmp)
     end
 

### 2016-12-16 real/code/tst.jl

@@ -1,66 +1,38 @@
-import optim
-import PyCall
+import Distributions, PyCall, TardisStats
+using FredPlots
 
-function main()
-    # define order of linear models
-    αOrder = 1
-    βOrder = 1
+α0 = 1.5; β0 = 60.; λ0 = 500
+νmin=0.
+νmax=10.
 
-    # read in data and create posterior
-    dim = αOrder + βOrder
-    ∇res, Hres = optim.allocations(dim)
-    frame, (P, ∇P!, HP!) = optim.problem(αOrder=αOrder, βOrder=βOrder, run=99)
+# define order of linear models
+αOrder = 1
+βOrder = 1
+npackets=1000
 
-    # compute the posterior mode to estimate the evidence
-    @time maxP, mode, ret = optim.run_nlopt(frame, P, ∇P!, HP!, αOrder, βOrder, xtol_rel=1e-3)
-    HP!(mode, Hres)
-    evidence = optim.laplace(maxP, Hres)
-    println("got $maxP at $mode (returned $ret) and evidence $evidence")
+# read in data from tardis and create posterior
+dim = αOrder + βOrder
+∇res, Hres = TardisStats.optim.allocations(dim)
+dataframe, (P, ∇P!, HP!) = TardisStats.optim.problem(αOrder=αOrder, βOrder=βOrder, run=99, npackets=npackets)
 
-    # the posterior mean with the normalized posterior
-    pm = optim.PosteriorMean(0.1, 0.2, 6.0)
-    Pmean, ∇Pmean!, HPmean! = optim.targetfactory(frame, αOrder, βOrder, evidence=evidence, pm=pm)
-    println(Pmean(mode))
-    @time maxPmean, mode, ret = optim.run_nlopt(frame, Pmean, ∇Pmean!, HPmean!, αOrder, βOrder,
-                                          init=mode, xtol_rel=1e-8)
-    println("max of normalized posterior = $maxPmean at $mode")
+# now change the energies
+PyCall.@pyimport numpy.random as nprand
+nprand.seed(1358)
+PyCall.@pyimport scipy.stats as stats
+dist0 = stats.gamma(α0, scale=1/β0)
+# crucial: change dataframe in place, else optimization runs with old energies!
+dataframe[:energies][:] = dist0[:rvs](size=length(dataframe[:energies]))
+# Plots.histogram(dataframe[:energies])
 
-    return
+# the observed number of packets in the bin
[…]

### 2016-12-16 real/draft.tex
[draft] Fix error before suff. statistics

@@
 where the data $(n,\bml)$ must naturally follow the same
-Gamma-distribution likelihood as $(N,\bmL)$, i.e.\
+Gamma likelihood as $(N,\bmL)$, i.e.\
 \begin{align}
@@
   & = \prod_{j=1}^n \GammaDist(\ell_j\cond\alpha,\beta)
-    \ =\ \frac{\beta^{n\alpha} p^{\alpha-1}\,e^{-\beta q}}{[\Gamma(\alpha)]^n}
-\end{align}
-with compact representation of the data in terms of
-$q = \sum_{j=1}^n \ell_j$ and $r = \prod_{j=1}^n \ell_j$.
+    \ =\ \frac{\beta^{n\alpha} r^{\alpha-1}\,e^{-\beta q}}{[\Gamma(\alpha)]^n}
+\end{align}
+with the data summarized by the sufficient statistics $n$,
+$q \equiv \sum_{j=1}^n \ell_j$, and $r \equiv \prod_{j=1}^n \ell_j$.
 
@@
 important. There is, however, some explicit prior information which we
-want to build into our analysis. The distribution
-$p(\ell \cond \alpha, \beta)$ should always vanish at the origin,
-requiring that $\alpha$ must be larger than 1.  For $\beta$, we
-likewise choose a uniform prior
+want to build into our analysis.
+The distribution $p(\ell \cond \alpha, \beta)$ should always vanish at
+the origin, requiring that $\alpha$ must be larger than 1.  For
+$\beta$, we likewise choose a uniform prior
+%
+\fred{This paragraph is from the old TARDIS section. In general, we
+  don't impose the constraint on $\alpha$. Only for TARDIS it seemed
+  to match and prevented the fitting from running away if very few
+  samples were present}
+%
 \begin{align}
@@
 \end{document}
+
+% Local Variables:
+% compile-command:"rubber --pdf -W refs --synctex draft"
+% End:

### 2016-12-16 real/draft.tex
[draft] Add comments, fix minor errors

@@
 
+\fred{This is a quick summary of Bayesian inference, we could refer to some text books for more details}
 We first sketch the generic notation and relations for Bayesian
@@
 descriptions of the simulation.
+\fred{I find this paragraph confusing. Which relationship?}
 
@@
 $\Delta_b = [\nu_b {-} \tfrac{1}{2} \Delta\nu \,,\, \nu_b {+}
-\tfrac{1}{2} \Delta\nu], b = 1,2,\ldots, B$. Let $n_b$ be the number
+\tfrac{1}{2} \Delta\nu], b = 1,2,\ldots, B$. \fred{The binning may be dictated by experimental data} Let $n_b$ be the number
 of packets having a frequency falling into a particular bin $b$, and
@@
     \GammaDist(Q\cond N\alpha, \beta)\;
-    \frac{\beta^{n\alpha} p^{\alpha-1}\,e^{-\beta q}}{[\Gamma(\alpha)]^n}
+    \frac{\beta^{n\alpha} r^{\alpha-1}\,e^{-\beta q}}{[\Gamma(\alpha)]^n}
 \end{align}
@@
     p(Q\cond N,\bmL)\;
-    p(\bmL\cond n,\bmphi)
-\end{align}
-with $p(\bmL\cond n,\bmphi) = \prod_{j=1}^n p(L_j\cond\bmphi)$ as
+    p(\bmL\cond N,\bmphi)
+\end{align}
+with $p(\bmL\cond N,\bmphi) = \prod_{j=1}^N p(L_j\cond\bmphi)$ as
 before. Again all quantities are related to a single bin in frequency.
@@
 $\lambda \to \infty$, given the first and second moment of $\Lum$, is
-a gaussian or normal distribution,
+a Gaussian or normal distribution,
 \begin{align}
@@
 is that a Gaussian represents our state of knowledge \emph{before}
-data $\bmy$ is observed. With this assignment, the definition of
-$\bmphi$ reduces to $(\mu,\sigma^2)$ and the posterior becomes, using
-Bayes' Theorem,
+data $\bmy$ is observed. With this assignment,
+$\bmphi =(\mu,\sigma^2)$ and the posterior becomes, using Bayes'
+Theorem,
 \begin{align}
@@
   unnecessary.}  %
-In keeping with prior information that luminositites are always
+In keeping with prior information that luminosities are always
 positive, we assign an improper uniform prior
 $p(\mu\cond\mu_{\rm max}) = 1/\mu_{\rm max}$ independent of $n$; as
-always, the posterior is independent of $\mu_{\rm max}$. For
-$\sigma^2$, we use an $n$-independent Inverse Gamma distribution prior
-with hyperparameters $(a_0,b_0)$ which reads
+always, the posterior is independent of
+$\mu_{\rm max}$.
+\fred{only true if $\mu_{\rm max}$ large enough to contain posterior}
+For $\sigma^2$, we use an $n$-independent Inverse Gamma
+distribution prior with hyperparameters $(a_0,b_0)$ which reads
 \begin{align}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%
-\hans{Some comments on the numerical values of hyperparameters $a$ in Poisson, $a,b$
-  in inverse gamma needed.} %
+\hans{Some comments on the numerical values of hyperparameters $a$ in
+  Poisson, $a,b$ in inverse gamma needed.} %
 \hans{I would omit the text and calculations below. Apart from the
@@
   the values of $\mu$ will be negative. Could this be the reason why
-  the numerics pick up a factor two?}
+  the numerics pick up a factor two?}%
+\fred{No, what I used before was equivalent to a uniform prior that
+  extends into the range $\mu<0$. The normalization cancels out and
+  the negative region doesn't contribute. With the uniform prior, you
+  assume that your upper cut-off on $\mu$ is large enough to contain
+  the posterior but at the same time large values don't actually
+  contribute so it doesn't matter if you include them or not. That
+  means I can't pick up a factor of two there, it would only change
+  the normalization but not the \emph{shape} of $p(Q)$}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2016-12-19 real/code/TardisPaper.jl
[code] Add TardisPaper
Add new self-contained module to implement just what is in the current
version of the paper draft. We can now simplify the code with a lot more
assumptions but I want to keep the old code for now, and a rewrite is
much cleaner.

@@ -1,0 +1,55 @@
+module TardisPaper
+
+module GammaPosterior
+"""reload("TardisPaper"); TardisPaper.GammaPosterior.test()"""
+
+export log_posterior, make_log_posterior
+
+using Base.Test, Distributions, Optim
+
+""" posterior P(α, β | n, q, r) for the parameters of the Gamma
+distribution given the sufficient statistics of the samples."""
+
+log_posterior(α::Real, β::Real, n::Real, q::Real, logr::Real) = n*(α*log(β)-lgamma(α)) + (α-1)*logr - β*q
+function make_log_posterior(n::Real, q::Real, logr::Real)
+    function (θ::Vector)
+        α::Real, β::Real = θ
+        log_posterior(α, β, n, q, logr)
+    end
+end
+
+function test()
+    # draw samples from a Gamma distribution
+    α = 1.5
+    β = 60
+    n = 500
+    srand(1612)
+    dist = Gamma(α, 1/β)
+    samples = rand(dist, n)
+    q = sum(samples)
+    logr = sum(log(samples))
+
+    @test log_posterior(α, β, n, q, logr) ≈ 1407.87731905469
+    @inferred log_posterior(α, β, n, q, logr)
+
+    f = make_log_posterior(n, q, logr)
+    @test f([α, β]) == log_posterior(α, β, n, q, logr)
+
+    # box minimization with more samples
+    samples = rand(dist, n*50)
+        q = sum(samples)
+    logr = sum(log(samples))
+
+    f = make_log_posterior(length(samples), q, logr)
+    negf(θ) = -f(θ)
+    lower = [1.0, 0]
+    upper = [Inf, Inf]
+    initial_θ = 1.05 * [α, β]
+    res = optimize(DifferentiableFunction(negf), initial_θ, lower, upper, Fminbox(), optimizer = LBFGS)
+
+    @test isapprox(Optim.minimizer(res)[1], α; rtol=1e-3)
+    @test isapprox(Optim.minimizer(res)[2], β; rtol=5e-2)
+end
+
+end # GammaPosterior
+end # TardisPaper

### 2016-12-19 real/code/TardisPaper.jl
[code] Add log_gamma function

@@ -1,7 +1,7 @@
 module TardisPaper
 
-module GammaPosterior
-"""reload("TardisPaper"); TardisPaper.GammaPosterior.test()"""
+"""reload("TardisPaper"); TardisPaper.GammaIntegrand.test()"""
+module GammaIntegrand
 
 export log_posterior, make_log_posterior
 
@@ -9,7 +9,6 @@ using Base.Test, Distributions, Optim
 
 """ posterior P(α, β | n, q, r) for the parameters of the Gamma
 distribution given the sufficient statistics of the samples."""
-
 log_posterior(α::Real, β::Real, n::Real, q::Real, logr::Real) = n*(α*log(β)-lgamma(α)) + (α-1)*logr - β*q
 function make_log_posterior(n::Real, q::Real, logr::Real)
     function (θ::Vector)
@@ -17,6 +16,12 @@ function make_log_posterior(n::Real, q::Real, logr::Real)
         log_posterior(α, β, n, q, logr)
     end
 end
+
+""" Gamma(x | α, β) """
+log_gamma(x::Real, α::Real, β::Real) = α*log(β) - lgamma(α) + (α-1)*log(x) - β*x
+
+""" p(Q | α, β, N) """
+log_predict(Q::Real, α::Real, β::Real, N::Real) = log_gamma(Q, N*α, β)
 
 function test()
     # draw samples from a Gamma distribution
@@ -29,6 +34,13 @@ function test()
     q = sum(samples)
     logr = sum(log(samples))
 
+    # compare to Distributions
+    x = 0.04
+    dist = Gamma(α, 1/β)
+    @test log_gamma(x, α, β) ≈ log(pdf(dist, x))
+    @test log_gamma(x, α, β) ≈ logpdf(dist, x)
+    @test log_predict(x, α, β, n) ≈ logpdf(Gamma(n*α, 1/β), x)
+
     @test log_posterior(α, β, n, q, logr) ≈ 1407.87731905469
     @inferred log_posterior(α, β, n, q, logr)
 
@@ -37,7 +49,7 @@ function test()
 
     # box minimization with more samples
     samples = rand(dist, n*50)
-        q = sum(samples)
+    q = sum(samples)
     logr = sum(log(samples))
 
     f = make_log_posterior(length(samples), q, logr)
@@ -47,9 +59,11 @@ function test()
     initial_θ = 1.05 * [α, β]
[…]

### 2016-12-20 real/code/TardisPaper.jl
[code] add Integration module with Cubature

@@ -1,10 +1,77 @@
 module TardisPaper
 
+"""use Cubature or Laplace"""
+module Integrate
+using Cubature
+
+function by_cubature(logf::Function, xmin::Real, xmax::Real; kwargs...)
+    # count function evaluations
+    counter = 0
+    h = t -> (counter += 1; exp(logf(t)))
+
+    # transform variables
+    if xmax === Inf
+        h = t -> (counter += 1; exp(logf(xmin + t/(1-t))) * 1/(1-t)^2)
+        xmin, xmax = 0, 1
+    end
+    estimate, σ = hquadrature(h, xmin, xmax; kwargs...)
+    return estimate, σ, counter
+end
+
+function by_cubature(logf::Function, xmin::Vector, xmax::Vector; kwargs...)
+    # count function evaluations
+    counter = 0
+
+    # keep independent limits in case of transform
+    a = copy(xmin)
+    b = copy(xmax)
+
+    # count parameters to transform and their index in the arg vector
+    ninf = sum(x -> x == Inf, xmax)
+    indices = Vector{Int16}(ninf)
+    offset = 1
+    for (i, xi) in enumerate(xmax)
+        if xi == Inf
+            indices[offset] = i
+            offset += 1
+            a[i] = 0
+            b[i] = 1
+        end
+    end
+
+    if ninf == 0
+        h = function(t) counter += 1; exp(logf(t)) end
+    else
+        # transform variables, mutate x in place
+        x = Vector(xmax)
+        trafo = function(t)
+            x .= t
+            jacobian = 1.0
+            for i in indices
+                x[i] = xmin[i] + t[i]/(1-t[i])
+                jacobian *= 1/(1-t[i])^2
+            end
+            return jacobian
+        end
+        h = function(t)
+            counter += 1
+            jacobian = trafo(t)
+            exp(logf(x)) * jacobian
+        end
+    end
+    estimate, σ = hcubature(h, a, b; kwargs...)
+    return estimate, σ, counter
+end
+
+end # Integrate
+
+
 """reload("TardisPaper"); TardisPaper.GammaIntegrand.test()"""
 module GammaIntegrand
 
 export log_posterior, make_log_posterior
 
+import ..Integrate
 using Base.Test, Distributions, Optim
 
 """ posterior P(α, β | n, q, r) for the parameters of the Gamma
@@ -62,6 +129,24 @@ function test()
[…]

### 2016-12-20 real/code/GammaIntegrand.jl
[code] Move submodules to individual files

@@ -1,0 +1,84 @@
+"""reload("TardisPaper"); TardisPaper.GammaIntegrand.test()"""
+module GammaIntegrand
+
+export log_posterior, make_log_posterior
+
+import ..Integrate
+using Base.Test, Distributions, Optim
+
+""" posterior P(α, β | n, q, r) for the parameters of the Gamma
+distribution given the sufficient statistics of the samples."""
+log_posterior(α::Real, β::Real, n::Real, q::Real, logr::Real) = n*(α*log(β)-lgamma(α)) + (α-1)*logr - β*q
+function make_log_posterior(n::Real, q::Real, logr::Real)
+    function (θ::Vector)
+        α::Real, β::Real = θ
+        log_posterior(α, β, n, q, logr)
+    end
+end
+
+""" Gamma(x | α, β) """
+log_gamma(x::Real, α::Real, β::Real) = α*log(β) - lgamma(α) + (α-1)*log(x) - β*x
+
+""" p(Q | α, β, N) """
+log_predict(Q::Real, α::Real, β::Real, N::Real) = log_gamma(Q, N*α, β)
+
+function test()
+    # draw samples from a Gamma distribution
+    α = 1.5
+    β = 60
+    n = 500
+    srand(1612)
+    dist = Gamma(α, 1/β)
+    samples = rand(dist, n)
+    q = sum(samples)
+    logr = sum(log(samples))
+
+    # compare to Distributions
+    x = 0.04
+    dist = Gamma(α, 1/β)
+    @test log_gamma(x, α, β) ≈ log(pdf(dist, x))
+    @test log_gamma(x, α, β) ≈ logpdf(dist, x)
+    @test log_predict(x, α, β, n) ≈ logpdf(Gamma(n*α, 1/β), x)
+
+    @test log_posterior(α, β, n, q, logr) ≈ 1407.87731905469
+    @inferred log_posterior(α, β, n, q, logr)
+
+    f = make_log_posterior(n, q, logr)
+    @test f([α, β]) == log_posterior(α, β, n, q, logr)
+
+    # box minimization with more samples
+    samples = rand(dist, n*50)
+    q = sum(samples)
+    logr = sum(log(samples))
+
+    f = make_log_posterior(length(samples), q, logr)
+    negf(θ) = -f(θ)
+    lower = [1.0, 0]
+    upper = [Inf, Inf]
+    initial_θ = 1.05 * [α, β]
[…]

### 2016-12-20 real/code/Integrate.jl

@@ -1,0 +1,64 @@
+"""use Cubature or Laplace"""
+module Integrate
+using Cubature
+
+function by_cubature(logf::Function, xmin::Real, xmax::Real; kwargs...)
+    # count function evaluations
+    counter = 0
+    h = t -> (counter += 1; exp(logf(t)))
+
+    # transform variables
+    if xmax === Inf
+        h = t -> (counter += 1; exp(logf(xmin + t/(1-t))) * 1/(1-t)^2)
+        xmin, xmax = 0, 1
+    end
+    estimate, σ = hquadrature(h, xmin, xmax; kwargs...)
+    return estimate, σ, counter
+end
+
+function by_cubature(logf::Function, xmin::Vector, xmax::Vector; kwargs...)
+    # count function evaluations
+    counter = 0
+
+    # keep independent limits in case of transform
+    a = copy(xmin)
+    b = copy(xmax)
+
+    # count parameters to transform and their index in the arg vector
+    ninf = sum(x -> x == Inf, xmax)
+    indices = Vector{Int16}(ninf)
+    offset = 1
+    for (i, xi) in enumerate(xmax)
+        if xi == Inf
+            indices[offset] = i
+            offset += 1
+            a[i] = 0
+            b[i] = 1
+        end
+    end
+
+    if ninf == 0
+        h = function(t) counter += 1; exp(logf(t)) end
+    else
+        # transform variables, mutate x in place
+        x = Vector(xmax)
+        trafo = function(t)
+            x .= t
+            jacobian = 1.0
+            for i in indices
+                x[i] = xmin[i] + t[i]/(1-t[i])
+                jacobian *= 1/(1-t[i])^2
+            end
+            return jacobian
+        end
+        h = function(t)
+            counter += 1
+            jacobian = trafo(t)
+            exp(logf(x)) * jacobian
+        end
+    end
+    estimate, σ = hcubature(h, a, b; kwargs...)
+    return estimate, σ, counter
+end
+
+end # Integrate

### 2016-12-20 real/code/TardisPaper.jl

@@ -1,154 +1,6 @@
 module TardisPaper
 
-"""use Cubature or Laplace"""
-module Integrate
-using Cubature
-
-function by_cubature(logf::Function, xmin::Real, xmax::Real; kwargs...)
-    # count function evaluations
-    counter = 0
-    h = t -> (counter += 1; exp(logf(t)))
-
-    # transform variables
-    if xmax === Inf
-        h = t -> (counter += 1; exp(logf(xmin + t/(1-t))) * 1/(1-t)^2)
-        xmin, xmax = 0, 1
-    end
-    estimate, σ = hquadrature(h, xmin, xmax; kwargs...)
-    return estimate, σ, counter
-end
-
-function by_cubature(logf::Function, xmin::Vector, xmax::Vector; kwargs...)
-    # count function evaluations
-    counter = 0
-
-    # keep independent limits in case of transform
-    a = copy(xmin)
-    b = copy(xmax)
-
-    # count parameters to transform and their index in the arg vector
-    ninf = sum(x -> x == Inf, xmax)
-    indices = Vector{Int16}(ninf)
-    offset = 1
-    for (i, xi) in enumerate(xmax)
-        if xi == Inf
-            indices[offset] = i
-            offset += 1
-            a[i] = 0
-            b[i] = 1
-        end
-    end
-
-    if ninf == 0
-        h = function(t) counter += 1; exp(logf(t)) end
-    else
-        # transform variables, mutate x in place
-        x = Vector(xmax)
-        trafo = function(t)
-            x .= t
-            jacobian = 1.0
-            for i in indices
-                x[i] = xmin[i] + t[i]/(1-t[i])
-                jacobian *= 1/(1-t[i])^2
-            end
-            return jacobian
-        end
-        h = function(t)
-            counter += 1
-            jacobian = trafo(t)
-            exp(logf(x)) * jacobian
-        end
-    end
-    estimate, σ = hcubature(h, a, b; kwargs...)
-    return estimate, σ, counter
-end
-
-end # Integrate
-
-
-"""reload("TardisPaper"); TardisPaper.GammaIntegrand.test()"""
-module GammaIntegrand
-
-export log_posterior, make_log_posterior
-
-import ..Integrate
-using Base.Test, Distributions, Optim
-
[…]

### 2016-12-20 real/code/GammaIntegrand.jl
[code] Add Poisson and optimization of integrand over N

@@ -3,10 +3,10 @@ module GammaIntegrand
 
 export log_posterior, make_log_posterior
 
-import ..Integrate
+import ..Integrate, ..Poisson
 using Base.Test, Distributions, Optim
 
-""" posterior P(α, β | n, q, r) for the parameters of the Gamma
+""" posterior p(α, β | n, q, r) for the parameters of the Gamma
 distribution given the sufficient statistics of the samples."""
 log_posterior(α::Real, β::Real, n::Real, q::Real, logr::Real) = n*(α*log(β)-lgamma(α)) + (α-1)*logr - β*q
 function make_log_posterior(n::Real, q::Real, logr::Real)
@@ -22,11 +22,54 @@ log_gamma(x::Real, α::Real, β::Real) = α*log(β) - lgamma(α) + (α-1)*log(x)
 """ p(Q | α, β, N) """
 log_predict(Q::Real, α::Real, β::Real, N::Real) = log_gamma(Q, N*α, β)
 
+
+"""
+Collect data that the target density needs
+"""
+type OptData
+    Q::Real
+    α::Real
+    β::Real
+    a::Real
+    n::Integer
+    count::Integer
+
+    function OptData(Q::Real, α::Real, β::Real, a::Real, n::Integer)
+        if Q < 0 Error("Invalid Q < 0: $Q") end
+        if α < 1 Error("Invalid α < 1: $α") end
+        if β < 0 Error("Invalid β < 0: $β") end
+        if a < 0 Error("Invalid a < 0: $a") end
+        new(Q, α, β, a, n, 0)
+    end
+end
+
+"""
+Create the target with data embedded via a closure. Negate for minimization
+"""
+function factory(Q::Real, α::Real, β::Real, a::Real, n::Integer)
+    popt = OptData(Q, α, β, a, n)
+    f(N::Real) = (popt.count += 1; -Poisson.log_predictive(N, popt.n, popt.a) - log_gamma(popt.Q, N * popt.α, popt.β))
+    return popt, f
+end
+
+"""
+Find the arg max_N NegativeBinomial(N|n-a+1, 1/2)*Gamma(X|Nα, β) using Brent's method.
+
+Use the plug-in values of α and β to avoid costly integration.
+"""
+function solve(Q::Real, α::Real, β::Real, a::Real, n::Integer;
+               ε=1e-2, min=0.0, max=0.0, trace=false)
+    popt, f = factory(Q, α, β, a, n)
[…]

### 2016-12-20 real/code/Poisson.jl

@@ -1,0 +1,17 @@
+module Poisson
+
+using Base.Test
+
+"""(pmh)"""
+function log_predictive(N::Real, n::Integer, a::Real)
+    tmp = N + n -a + 1
+    lgamma(tmp) - lgamma(N+1) - lgamma(n-a+1) - tmp * log(2)
+end
+
+function test()
+    N=7; n=11; a=0;
+
+    @test log_predictive(N, n, a) ≈ log(binomial(N+n-a, N)) - (N+n-a+1)*log(2)
+end
+
+end #Poisson

### 2016-12-20 real/code/TardisPaper.jl

@@ -1,6 +1,7 @@
 module TardisPaper
 
 include("Integrate.jl")
+include("Poisson.jl")
 include("GammaIntegrand.jl")
 
 end # TardisPaper

### 2016-12-20 real/code/GammaIntegrand.jl
(wip) transfer to laptop

@@ -1,9 +1,9 @@
 """reload("TardisPaper"); TardisPaper.GammaIntegrand.test()"""
 module GammaIntegrand
 
-export log_posterior, make_log_posterior
+export log_gamma, log_gamma_predict, log_poisson_predict, log_posterior, make_log_posterior
 
-import ..Integrate, ..Poisson
+import ..Integrate
 using Base.Test, Distributions, Optim
 
 """ posterior p(α, β | n, q, r) for the parameters of the Gamma
@@ -20,8 +20,13 @@ end
 log_gamma(x::Real, α::Real, β::Real) = α*log(β) - lgamma(α) + (α-1)*log(x) - β*x
 
 """ p(Q | α, β, N) """
-log_predict(Q::Real, α::Real, β::Real, N::Real) = log_gamma(Q, N*α, β)
+log_gamma_predict(Q::Real, α::Real, β::Real, N::Real) = log_gamma(Q, N*α, β)
 
+"""(pmh) p(N | n)"""
+function log_poisson_predict(N::Real, n::Integer, a::Real)
+    tmp = N + n -a + 1
+    lgamma(tmp) - lgamma(N+1) - lgamma(n-a+1) - tmp * log(2)
+end
 
 """
 Collect data that the target density needs
@@ -48,7 +53,7 @@ Create the target with data embedded via a closure. Negate for minimization
 """
 function factory(Q::Real, α::Real, β::Real, a::Real, n::Integer)
     popt = OptData(Q, α, β, a, n)
-    f(N::Real) = (popt.count += 1; -Poisson.log_predictive(N, popt.n, popt.a) - log_gamma(popt.Q, N * popt.α, popt.β))
+    f(N::Real) = (popt.count += 1; -log_poisson_predict(N, popt.n, popt.a) - log_gamma_predict(popt.Q, N, popt.α, popt.β))
     return popt, f
 end
 
@@ -65,6 +70,10 @@ function solve(Q::Real, α::Real, β::Real, a::Real, n::Integer;
 end
 
 function test()
+    N=7; n=11; a=0;
+
+    @test log_poisson_predict(N, n, a) ≈ log(binomial(N+n-a, N)) - (N+n-a+1)*log(2)
+
     # draw samples from a Gamma distribution
     α = 1.5
     β = 60
@@ -81,7 +90,7 @@ function test()
     dist = Gamma(α, 1/β)
     @test log_gamma(x, α, β) ≈ log(pdf(dist, x))
[…]

### 2016-12-20 real/code/Predict.jl

@@ -1,0 +1,91 @@
+module Predict
+
+using ..GammaIntegrand
+import Optim
+
+# """
+
+# Compute contribution to `res` for `N`. Return updated `res` and bool
+# if search should continue in this direction.
+
+# """
+# function search_step!(frame::DataFrame, Pmean, ∇Pmean!, HPmean!, Hres::Matrix, pm::PosteriorMean,
+#                       αOrder::Integer, βOrder::Integer, nb::NegBinom, res::Real, θ::Vector, ε::Real;
+#                       xtol_rel=1e-5, optimize=true)
+# function search_step!(N::Real, Q::Real, α::Real, β::Real, a::Real, n::Real, ε::Real)
+#     # N>0 required to search
+#     if N <= 0
+#         return res, false
+#     end
+
+#     # have to leave the log scale now
+#     latest = exp(log_predictive(N, n, a) + log_predict(Q, α, β))
+#     res += latest
+#     # println("$(pm): latest=$latest, res=$res")
+
+#     return res, (latest / res) > ε
+# end
+
+"""
+(gmhb), just sum and consider α, β fixed.
+"""
+function by_sum(Q::Real, α::Real, β::Real, a::Real, n::Real; Ninit::Real=0, ε::Real=1e-3)
+    if Ninit == 0
+        # perform optimization, optim only does minimization but we
+        # actually optimize to find the N that will likely give the
+        # largest contribution. It's only approximate but should be a
+        # good starting value
+        res = GammaIntegrand.solve(Q, α, β, a, n)
+        Ninit = convert(Integer, ceil(Optim.minimizer(res)))
+    end
+
+    @assert(Ninit >= 1)
+    println("initial N=$Ninit")
+
+    # initialization
+    Nup = Ninit
+    Ndown = Ninit
+    res = 0.0
+
+    """Add contribution at `N` to `res`."""
+    function search(N::Real)
+        # N>0 required to search
+        if N <= 0
+            return false
+        end
+
+        # have to leave the log scale now
+        latest = exp(log_poisson_predict(N, n, a) + log_gamma_predict(Q, α, β, N))
+        res += latest
+        # println("$(pm): latest=$latest, res=$res")
+
+        return (latest / res) > ε
+    end
+
+    # start at initial N, then go up and down until contribution negligible
+    _ = search(Ninit)
+    print("$res from Ninit=$Ninit")
+    goup, godown = true, true
[…]

### 2016-12-20 real/code/TardisPaper.jl

@@ -1,7 +1,7 @@
 module TardisPaper
 
 include("Integrate.jl")
-include("Poisson.jl")
+# include("Poisson.jl")
 include("GammaIntegrand.jl")
-
+include("Predict.jl")
 end # TardisPaper

### 2016-12-20 real/code/optim.jl

@@ -108,12 +108,12 @@ function run_nlopt(frame, f,  ∇f!, Hf!, αOrder, βOrder; init=nothing, xtol_r
     return res
 end
 
-# """
+"""
 
-# Compute contribution to `res` for `N`. Return updated `res` and bool
-# if search should continue in this direction.
+Compute contribution to `res` for `N`. Return updated `res` and bool
+if search should continue in this direction.
 
-# """
+"""
 function search_step!(frame::DataFrame, Pmean, ∇Pmean!, HPmean!, Hres::Matrix, pm::PosteriorMean,
                       αOrder::Integer, βOrder::Integer, nb::NegBinom, res::Real, θ::Vector, ε::Real;
                       xtol_rel=1e-5, optimize=true)

### 2017-01-24 real/code/Predict.jl
[Precict] Update comment and limits on cubature integration

@@ -115,10 +115,11 @@ Predict Q by summing over N and integrating over α, β  with cubature rules
 Arguments
 ---------
 
+`ε`: tolerance for truncating the sum over N
 `reltol`: tolerance for cubature integration
 """
 
-function by_cubature(Q, a, n, q, logr; αmin=1.3, αmax=1.8, βmin=40, βmax=100,
+function by_cubature(Q, a, n, q, logr; αmin=1.0, αmax=2, βmin=10, βmax=100,
                      Ninit=0, ε=1e-3, reltol=1e-3)
     # to avoid overflow in Cubature, evaluate target at mode and subtract it. The
     # value within a few order of magnitude of

### 2017-01-24 real/code/GammaIntegrand.jl
[code] Implement asymptotic approx. with Laplace
finally it agrees with summation

@@ -1,8 +1,9 @@
 """reload("TardisPaper"); TardisPaper.GammaIntegrand.test()"""
 module GammaIntegrand
 
-export heuristicN, log_gamma, log_gamma_predict, log_poisson_predict, log_posterior, make_log_posterior
-export optimize_log_posterior, optimize_log_posterior_predict
+export heuristicN, log_gamma, log_gamma_predict, log_inv_gamma, log_normal, log_poisson_predict, log_posterior
+export make_asymptotic, make_log_posterior
+export optimize_log_posterior, optimize_log_posterior_predict, optimize_integrand_λμσ²
 
 import ..Integrate
 
@@ -34,24 +35,37 @@ function log_poisson_predict(N::Real, n::Integer, a::Real)
 end
 
 "Return Optimize.result struct"
-function optimize_integrand(target; αmin=1.0, αmax=Inf, βmin=0.0, βmax=Inf, αinit=1.5, βinit=50.0)
+function optimize_integrand(target, lower, upper, initial)
     min_target = x -> -target(x)
-
-    lower = [αmin, βmin]
-    upper = [αmax, βmax]
-    initial_θ = [αinit, βinit]
     # TODO couldn't get autodiff to work here. Probably because of the box constraints
-    res = optimize(DifferentiableFunction(min_target), initial_θ,
+    res = optimize(DifferentiableFunction(min_target), initial,
                    lower, upper, Fminbox(), optimizer=LBFGS)
                    # optimizer_o=OptimizationOptions(autodiff=true))
     # get gradient and Hessian into one result
-    storage = DiffBase.HessianResult(initial_θ)
+    storage = DiffBase.HessianResult(initial)
     ForwardDiff.hessian!(storage, min_target, Optim.minimizer(res))
     res, storage
 end
 
+function optimize_integrand_αβ(target; αmin=1.0, αmax=Inf, βmin=0.0, βmax=Inf, αinit=1.5, βinit=50.0)
+    lower = [αmin, βmin]
+    upper = [αmax, βmax]
+    initial = [αinit, βinit]
+    optimize_integrand(target, lower, upper, initial)
+end
+
+function optimize_integrand_λμσ²(target;
+                                λmin=0.0, λmax=Inf, λinit=50.0,
[…]

### 2017-01-24 real/code/Predict.jl

@@ -159,4 +159,30 @@ function by_cubature(Q, a, n, q, logr; αmin=1.0, αmax=2, βmin=10, βmax=100,
     iterate(Ninit, f, ε)
 end
 
+"""
+Predict Q by the asymptotic expression (ral) integrating over λ, μ, and σ² with Laplace.
+
+Arguments
+---------
+
+`a₀`: hyperprior for inverse Gamma shape
+`b₀`: hyperprior for inverse Gamma rate
+"""
+function asymptotic_by_laplace(Q, a, n, first, second; a₀=0, b₀=0)
+    # create integrand
+    f = make_asymptotic(Q, n, a, first, second; a₀=a₀, b₀=b₀)
+
+    # optimize integrand
+    res, diffstore = optimize_integrand_λμσ²(f; λinit=1.05*(n-a+1), μinit=1.05*first, σ²init=1.05*n/2*(second-first^2))
+
+    # Hessian at mode
+    H = DiffBase.hessian(diffstore)
+    invH = inv(H)
+    mode = Optim.minimizer(res)
+    info("mode ± std. err: λ=", mode[1], "±", invH[1,1], ", μ=" , mode[2], "±", invH[2,2], ", σ²=", mode[3], "±", invH[3,3])
+
+    # Laplace approximation on log, go back to linear
+    exp(Integrate.by_laplace(-Optim.minimum(res), H))
+end
+
 end #Predict

### 2017-01-24 real/code/TardisPaperTest.jl

@@ -36,7 +36,13 @@ function predict()
     end
 
     # P(Q|...)
-    @test_approx_eq_eps(res_cuba[1], res_laplace[1], 1e-2)
+    @test_approx_eq_eps(res_cuba[1], res_laplace[1], 3e-2)
+
+    first = α/β
+    second = α/β^2 * (1+α)
+    res_asym_laplace = Predict.asymptotic_by_laplace(Q, a, n, first, second)
+    hello((res_asym_laplace, 0, 0), "asympt. Laplace")
+    @test_approx_eq_eps(res_asym_laplace, res_laplace[1], 3e-2)
 end
 
 function predict_alpha_fixed()
@@ -143,12 +149,45 @@ function gamma_integrand()
     @test isapprox(estimate, 1; atol=σ)
 end
 
+function asymptotic()
+    μ = 1.3
+    σ = 1.44
+    x = 2.011
+    normal = Distributions.Normal(μ, σ)
+
+    @test_approx_eq(log_normal(x, μ, σ^2), logpdf(normal, x))
+
+    a = 1.5
+    b = 0.013
+    invgamma = Distributions.InverseGamma(a, b)
+    @test_approx_eq(log_inv_gamma(x, a, b), logpdf(invgamma, x))
+
+    λ = 24.4
+    Q = λ*μ
+    n = 15
+    a = 0.5
+
+    first = 1.02*μ
+    second = 1.01*σ^2
+    f = make_asymptotic(Q, n, a, first, second)
+    normalQ = Distributions.Normal(λ*μ, sqrt(λ*(σ^2 + μ^2)))
+    normalμ = Distributions.Normal(first, sqrt(σ^2/n))
+    # second arg: rate vs. scale. Irrelevant if =1
+    gamma = Distributions.Gamma(n-a+1, 1)
+    invgamma = Distributions.InverseGamma(n/2, n/2 * (second - first^2))
+
+    @test_approx_eq(+f([λ, μ, σ^2]), logpdf(normalQ, Q) + logpdf(gamma, λ) + logpdf(normalμ, μ) + logpdf(invgamma, σ^2))
+
+    Predict.asymptotic_by_laplace(Q, a, n, first, second)
+end
+
 function run()
     @testset "all tests" begin
         integrate_by_cubature()
         predict()
         predict_alpha_fixed()
         gamma_integrand()
+        asymptotic()
     end
 end
 

### 2017-01-24 real/code/Predict.jl
[test] Fix check of asymptotic expansion

@@ -109,8 +109,8 @@ function by_laplace(Q, a, n, q, logr; Ninit=0, ε=1e-3)
 end
 
 """
-Predict Q by summing over N and integrating over α, β  with cubature rules
-(spk)
+Predict Q by summing over N and integrating over α, β with cubature rules
+(spk).
 
 Arguments
 ---------

### 2017-01-24 real/code/TardisPaperTest.jl

@@ -13,11 +13,11 @@ end
 function predict()
     α = 1.5
     β = 60.
-    n = 1000
+    n = 400
     a = 1/2
     Q = n*α/β
-    ε = 1e-3
-    srand(1612)
+    # ε = 1e-3
+    srand(11289)
 
     dist = Gamma(α, 1/β)
     samples = rand(dist, n)
@@ -38,8 +38,8 @@ function predict()
     # P(Q|...)
     @test_approx_eq_eps(res_cuba[1], res_laplace[1], 3e-2)
 
-    first = α/β
-    second = α/β^2 * (1+α)
+    first = q/n
+    second = mapreduce(x->x^2, +, samples)/n
     res_asym_laplace = Predict.asymptotic_by_laplace(Q, a, n, first, second)
     hello((res_asym_laplace, 0, 0), "asympt. Laplace")
     @test_approx_eq_eps(res_asym_laplace, res_laplace[1], 3e-2)

### 2017-01-25 real/code/GammaIntegrand.jl
[code] Allow 0<alpha<1, more debug logging, higher accuracy in predict's test

@@ -47,7 +47,7 @@ function optimize_integrand(target, lower, upper, initial)
     res, storage
 end
 
-function optimize_integrand_αβ(target; αmin=1.0, αmax=Inf, βmin=0.0, βmax=Inf, αinit=1.5, βinit=50.0)
+function optimize_integrand_αβ(target; αmin=0.0, αmax=Inf, βmin=0.0, βmax=Inf, αinit=1.5, βinit=50.0)
     lower = [αmin, βmin]
     upper = [αmax, βmax]
     initial = [αinit, βinit]
@@ -93,8 +93,8 @@ type OptData
 
     function OptData(Q::Real, α::Real, β::Real, a::Real, n::Integer)
         if Q < 0 Error("Invalid Q < 0: $Q") end
-        if α < 1 Error("Invalid α < 1: $α") end
-        if β < 0 Error("Invalid β < 0: $β") end
+        if α <= 0 Error("Invalid α <= 0: $α") end
+        if β <= 0 Error("Invalid β <= 0: $β") end
         if a < 0 Error("Invalid a < 0: $a") end
         new(Q, α, β, a, n, 0)
     end

### 2017-01-25 real/code/Predict.jl

@@ -5,7 +5,6 @@ using ..GammaIntegrand, ..Integrate
 import DiffBase, Distributions, Optim
 
 using Logging
-@Logging.configure(level=INFO)
 
 """ Find initial value of N that will likely give the largest
 contribution to (gmhb). It's only approximate but should be a good
@@ -23,7 +22,7 @@ function initialize_N(Q, α, β, a, n, N=0)
     end
 
     @assert(N >= 1)
-    info("initial N=$N")
+    debug("initial N=$N")
     return N
 end
 
@@ -119,7 +118,7 @@ Arguments
 `reltol`: tolerance for cubature integration
 """
 
-function by_cubature(Q, a, n, q, logr; αmin=1.0, αmax=2, βmin=10, βmax=100,
+function by_cubature(Q, a, n, q, logr; αmin=1e-2, αmax=5, βmin=1e-5, βmax=100,
                      Ninit=0, ε=1e-3, reltol=1e-3)
     # to avoid overflow in Cubature, evaluate target at mode and subtract it. The
     # value within a few order of magnitude of
@@ -127,7 +126,7 @@ function by_cubature(Q, a, n, q, logr; αmin=1.0, αmax=2, βmin=10, βmax=100,
     α, β = Distributions.params(fit_dist)
     # from scale to rate parameter
     β = 1/β
-    info("MLE estimates $α, $β")
+    debug("MLE estimates $α, $β")
     logf_mode = log_posterior(α, β, n, q, logr)
 
     # TODO take limits like 5*std. err. from Gaussian approximation
@@ -149,11 +148,9 @@ function by_cubature(Q, a, n, q, logr; αmin=1.0, αmax=2, βmin=10, βmax=100,
 
     # integrate by Cubature on linear scale
     f = N -> begin
-        # res, diffstore = optimize_log_posterior_predict(n, q, logr, Q, N, Z; kwargs...)
-        # integral = Integrate.by_laplace(-Optim.minimum(res), DiffBase.hessian(diffstore))
         integrand = GammaIntegrand.make_log_posterior_predict(n, q, logr, Q, N, logZ)
         integral, σ, ncalls = Integrate.by_cubature(integrand, lower, upper; reltol=reltol)
-        debug(integral, σ, ncalls)
+        debug("integrand: $integral  $σ after $ncalls calls")
         exp(log_poisson_predict(N, n, a)) * integral
[…]

### 2017-01-25 real/code/TardisPaperTest.jl

@@ -26,7 +26,7 @@ function predict()
 
     hello(res, alg) = println("$alg: P($Q|$n, $a, x) = $(res[1]) for N=$(res[2])...$(res[3])")
 
-    res_cuba = Predict.by_cubature(Q, 0.5, n, q, logr)
+    res_cuba = Predict.by_cubature(Q, 0.5, n, q, logr; reltol=1e-8)
     hello(res_cuba, "cubature")
     res_laplace = Predict.by_laplace(Q, a, n, q, logr)
     hello(res_laplace, "Laplace")

### 2017-01-25 real/draft.tex
[draft] p -> r
[draft] add comment on scale vs. rate par. for inv. gamma
[draft] comparison plot exact vs asymptotic
* remove old hedgehog plot
* remove paragraphs on conjugate model

@@
 \usepackage{graphicx}    % standard graphics
+\graphicspath{
+  {./figures/}
+}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 $p(X \cond n, \bmx)$ can be well approximated by a unimodal
-distribution; see \reffig{hedgehog}.
-
-\begin{figure}[ht]
-  \centering
-  % \includegraphics[width=0.48\textwidth]{hedgehog}
-  \caption{Left: Prediction $p(\Lumtot \cond  \nu,  \npack, \bmnu, \bml)$, right: $p(x \cond  \nu,  \npack, \bmnu, \bmx)$. I sum over $N$ in both cases. }
-  \fred{I still need to pretty this up. I used the old binned approach
-    with 50 samples in a bin from which I computed the mean and
-    variance to plug into \refeq{gmr} but I computed the sum over $N$
-    instead of the integral.}
-\label{fig:hedgehog}
-\end{figure}
+distribution.
 
@@
 distribution prior with hyperparameters $(a_0,b_0)$ which reads
+\fred{Inconsistentcy. Now $b_0$ is a scale parameter but before $\beta$ was a \emph{rate} parameter in definition of $\Gamma$}
 \begin{align}
@@
 approximation~\cite[Ch. 27]{mackay2003information} can be applied.
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=0.48\textwidth]{asymptotic}
+  \caption{Numerical prediction of $\Lumtot$ with the sum over $N$
+    from \refeq{spk} or the asymptotic expression from \refeq{ral} for $n$
+    samples from $\GammaDist(\alpha=1.5, \beta=60)$}
+\label{fig:asymptotic}
+\end{figure}
+
+\fred{In \reffig{asymptotic}, we contrast the asymptotic expression
+  with the ``exact'' expression \refeq{spk} where the integrals over
+  $\bmphi$ are done numerically to a requested relative accuracy of
+  $10^{-5}$ through the Cubature
+  package\cite{cubature,genz1980remarks,berntsen1991adaptive}. We
+  computed $P(\Lumtot \cond n, \bml)$ on a grid of $\Lumtot$ values
+  and checked if the probability integrates to one using Simpson's
+  rule. With the exact expression, we obtain 0.997 ($n=10$) and 0.996
+  ($n=80$), and 0.698 and 0.953 for the asymptotic expression. For
+  $n=10$, the large discrepancy arises mostly because the constraint
+  $P(Q \leq 0 \cond n, \bml) = 0$ is not respected in \refeq{ral}
+  whereas it is satisfied by
+  \refeq{spk}. Following~\cite{tierney_accurate_1986}, we renormalized
+  the results to exactly integrate to one for ease of comparison in
+  \reffig{asymptotic}. Note that the lack of normalization may not be
+  relevant; e.g. if \refeq{ral} is used as the likelihood for a single
+  bin in Bayesian parameter inference.
+
+  We verified that for $n>80$, the two methods agree even better than
+  shown there and that the normalization of the asymptotic result
+  approaches one. }
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
   the normalization but not the \emph{shape} of $p(Q)$}
+\fred{The error is fixed and I remove the next sections as Hans suggested.}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-To simplify the calculations, we assume a conjugate prior for unknown
-$\mu$ and $\sigma^2$. The standard result for the posterior is a
-Gaussian for $\mu$ and an Inverse Gamma distribution for $\sigma^2$,
-\begin{align}
-  \lleq{raf}
-  p(\mu, \sigma^2 \cond \bmy)
-  &= \GaussianDist\left(\mu \cond \mu_0^\prime, \sigma^2/n_0^\prime\right) %
-    \InvGammaDist\left( \sigma^2 \,\Bigl|\, \tfrac{1}{2}\nu_0^\prime,
-    \tfrac{1}{2}\nu_0^\prime \sigma_0^{\prime 2} \right)
-\end{align}
-where the hyperparameters $(n_0',\mu_0',\nu_0')$ are updated from
-prior to posterior based on $\bmy$ with the sample mean
-$\langle\lum\rangle = (1/n) \sum_{j=1}^n \lum_j$,
-\begin{align}
-  %\lleq{rag}
-  \lleq{rah}  n_0^\prime
-  &= n_0 + n,\\
-  \lleq{rai}    \mu_0^\prime
-  &= \frac{n_0 \mu_0 + n \langle\lum\rangle}{n_0^\prime},\\
-  \lleq{raj}    \nu_0^\prime
-  &= \nu_0 + n,\\
-  \lleq{rak}
-  \nu_0^\prime \sigma_0^{\prime 2}
-  &= \nu_0 \sigma_0^{ 2}
-    + \sum_{j=1}^n \left(\lum_j - \langle\lum\rangle\right)^2
-    + \frac{n_0 n}{n_0^\prime} \left(\mu_0 - \langle\lum\rangle\right)^2\,.
-\end{align}
-By default, we take a noninformative prior and set $n_0 = \mu_0 =
-\nu_0 = 0$, so the value $\sigma_0^2>0$ is irrelevant. %
-  % The posterior
-  % $ p(\expect{\Lum}, \expect{\Lum^2} \cond \bmy)$ is given by
-  % \refeq{raf} with $\mu, \sigma^2$ replaced by
-  % $\expect{\Lum}, \expect{\Lum^2}$ as indicated in \refeq{rae} because
-  % the Jacobian determinant is one.
-The above equations also show that the data $\bmy$ appear only in the
-form of $n$ and the first two sample moments
-$m_1 = \langle\ell\rangle$ and $m_2 = \langle\ell^2\rangle$.
-Inserting all these ingredients into Eq.~(\ref{asc}), we obtain
-\begin{align}
-  \lleq{ral}
-  p(\Lumtot \cond \bmy)
-  &= \int \rmdx{\lambda}\; \rmdx{\mu}\; \rmdx{\sigma^2}\;
-    p(\lambda \cond n)\; p( \mu, \sigma^2 \cond \bmy)\;
-    \GaussianDist \left( \Lumtot \cond \lambda \mu, \lambda (\sigma^2{+}\mu^2) \right)
-    \,,
-\end{align}
-where $ p(\lambda \cond n)$ is given by \refeq{pms} and
-$ p( \mu, \sigma^2 \cond \bmy)$ by Eqs.~(\ref{raf})--(\ref{rak}).
 

### 2017-01-25 real/code/TardisPaperPlots.jl

@@ -1,0 +1,87 @@
+module TardisPaperPlots
+using Logging
+@Logging.configure(level=INFO)
+
+import TardisPaper.Predict, TardisPaper.Integrate
+using Distributions, LaTeXStrings, Plots
+
+const font = Plots.font("TeX Gyre Heros")
+Plots.pyplot(guidefont=font, xtickfont=font, ytickfont=font, legendfont=font)
+Plots.PyPlotBackend()
+
+srand(16142)
+
+"save and replot"
+savepdf(fname) = begin Plots.pdf("../figures/$fname"); plot!() end
+
+function compute_asymptotic(;n=400, Qs=false, Qmin=1e-3, Qmax=2, nQ=50, α=1.5, β=60.0, a=1/2, ε=1e-3, reltol=1e-3)
+    dist = Gamma(α, 1/β)
+    samples = rand(dist, n)
+    q = sum(samples)
+    logr = sum(log(samples))
+
+    if Qs === false
+        meanQ = n*α/β
+        Qs = linspace(Qmin*meanQ, Qmax*meanQ, nQ)
+    end
+
+    res_cuba = map(Q->Predict.by_cubature(Q, 0.5, n, q, logr; reltol=reltol, ε=ε)[1], Qs)
+
+
+    first = q/n
+    second = mapreduce(x->x^2, +, samples)/n
+    res_asym_laplace = map(Q->Predict.asymptotic_by_laplace(Q, a, n, first, second), Qs)
+
+    hello(tag, norm, mean, stderr) = info(tag, ": norm=$norm, Q=$(mean)±$(stderr)")
+
+    norm, mean, stderr = Integrate.simpson(Qs, res_cuba)
+    hello("cubature", norm, mean, stderr)
+    res_cuba ./= norm
+
+    norm, mean, stderr = Integrate.simpson(Qs, res_asym_laplace)
+    hello("laplace", norm, mean, stderr)
+    res_asym_laplace ./= norm
+
+    Qs, res_cuba, res_asym_laplace
+end
+
+function plot_asymptotic_single(Qs, res_cuba, res_asym_laplace; kwargs...)
+    plot!(Qs, res_cuba; label=L"\sum_N", ls=:solid, leg=true, color=:red, kwargs...)
+    plot!(Qs, res_asym_laplace; label=L"\int dλ", ls=:dash, color=:blue, kwargs...)
+    xlabel!(L"Q")
+    # ylabel!(latexstring("\$P(Q|n=$n, \\ell)\$"))
[…]

### 2017-01-30 real/code/Asymptotic.jl
[real] Fix several bugs in moments calculation

@@ -79,7 +79,7 @@ function targetfactory(target::TARGET, n::Int64, xmean::Float64, xsumsq::Float64
         λ, μ, σSq = x
         var = σSq + μ^2
         exponent = λ*μ^2 / (2*var)
-        tmp = exp(-exponent) * sqrt(λ * var / (2 * π)) + 1/2 * λ * μ * (1 + erf(sqrt(exponent)))
+        tmp = exp(-exponent) * sqrt(λ * var / (2π)) + 1/2 * λ * μ * (1 + erf(sqrt(exponent)))
         tmp > 0.0 || error("Negative input ", λ, μ, σSq )
         log(tmp)
     end
@@ -89,15 +89,15 @@ function targetfactory(target::TARGET, n::Int64, xmean::Float64, xsumsq::Float64
         λ, μ, σSq = x
         var = σSq + μ^2
         exponent = λ*μ^2 / (2*var)
-        tmp = exp(-exponent) * sqrt(2 / π * λ^3 * var) * μ + 1/2 * λ*(λ*μ^2 + var)*(1 + erf(sqrt(exponent)))
-        log(1/2 * tmp)
+        tmp = exp(-exponent) * sqrt(λ^3 * var / (2π)) * μ + 1/2 * λ*(λ*μ^2 + var)*(1 + erf(sqrt(exponent)))
+        log(tmp)
     end
 
     """log(-d^3/dt^3 f(target=0|x))"""
     function logthird(x::Vector)
         λ, μ, σSq = x
         var = σSq + μ^2
-        exponent = λ*μ^2 / 2var
+        exponent = λ*μ^2 / (2var)
         tmp = exp(-exponent) / sqrt(2π) * λ^(3/2) * sqrt(var) * (λ*μ^2 + 2var)
         tmp += 1/2 * λ^2 * μ * (λ*μ^2 + 3var)*(1 + erf(sqrt(exponent)))
         log(tmp)
@@ -109,7 +109,7 @@ function targetfactory(target::TARGET, n::Int64, xmean::Float64, xsumsq::Float64
         var = σSq + μ^2
         exponent = λ*μ^2 / (2*var)
         tmp = exp(-exponent) / sqrt(2π) * λ^(5/2) * μ * sqrt(var) * (λ*μ^2 + 5var)
[…]

### 2017-01-31 real/code/Predict.jl
[code] asymptotic_by_cubature
add, test and plot its results and compare to the other integration methods

@@ -91,7 +91,7 @@ function by_laplace(Q, a, n, q, logr; Ninit=0, ε=1e-3)
 
     mode = Optim.minimizer(res)
     invH = inv(H)
-    info("mode ± std. err: α=" , mode[1], "±", invH[1,1], ", β=", mode[2], "±", invH[2,2])
+    debug("mode ± std. err: α=" , mode[1], "±", invH[1,1], ", β=", mode[2], "±", invH[2,2])
 
     # find initial N at posterior mode of α, β
     α, β = Optim.minimizer(res)
@@ -176,10 +176,54 @@ function asymptotic_by_laplace(Q, a, n, first, second; a₀=0, b₀=0)
     H = DiffBase.hessian(diffstore)
     invH = inv(H)
     mode = Optim.minimizer(res)
-    debug("mode ± std. err: λ=", mode[1], "±", invH[1,1], ", μ=" , mode[2], "±", invH[2,2], ", σ²=", mode[3], "±", invH[3,3])
+    debug(" asympt. Laplace, mode ± std. err: λ=", mode[1], "±", invH[1,1], ", μ=" , mode[2], "±", invH[2,2], ", σ²=", mode[3], "±", invH[3,3])
 
     # Laplace approximation on log, go back to linear
     exp(Integrate.by_laplace(-Optim.minimum(res), H))
+end
+
+function asymptotic_by_cubature(Q, a, n, first, second;a₀=0, b₀=0, reltol=1e-3,
+                                λmin=0.0, λmax=0.0,
+                                μmin=0.0, μmax=0.0,
+                                σ²min=0.0, σ²max=0.0)
+    # create integrand
+    f = make_asymptotic(Q, n, a, first, second; a₀=a₀, b₀=b₀)
+
+    # to avoid overflow in Cubature, evaluate target at approximate mode and subtract it. The
+    # value within a few order of magnitude of
+    mode = [n, first, second-first^2]
+    logf_mode = f(mode)
+    # logf_mode = 0.0
+    debug("asymptotic_by_cubature logf($mode)=$(logf_mode)")
+    target = x->(res = f(x) - logf_mode; println(res); res)
+    target = x->f(x) - logf_mode
+
+    # get ranges from marginal posterior distributions
+    k = 5
+
[…]

### 2017-01-31 real/code/TardisPaperPlots.jl

@@ -14,7 +14,17 @@ srand(16142)
 "save and replot"
 savepdf(fname) = begin Plots.pdf("../figures/$fname"); plot!() end
 
-function compute_asymptotic(;n=400, Qs=false, Qmin=1e-3, Qmax=2, nQ=50, α=1.5, β=60.0, a=1/2, ε=1e-3, reltol=1e-3)
+function normalize!(Qs, res, tag)
+    # norm only useful from first call
+    norm, _, _ = Integrate.simpson(Qs, res)
+    res ./= norm
+    # but mean and σ are affected by norm
+    _, mean, stderr = Integrate.simpson(Qs, res)
+    info(tag, ": norm=$norm, Q=$(mean)±$(stderr)")
+end
+
+function compute_prediction(;n=400, Qs=false, Qmin=1e-3, Qmax=2, nQ=50, α=1.5, β=60.0, a=1/2, ε=1e-3, reltol=1e-3)
+    info("Computing for n=$n")
     dist = Gamma(α, 1/β)
     samples = rand(dist, n)
     q = sum(samples)
@@ -25,24 +35,28 @@ function compute_asymptotic(;n=400, Qs=false, Qmin=1e-3, Qmax=2, nQ=50, α=1.5, 
         Qs = linspace(Qmin*meanQ, Qmax*meanQ, nQ)
     end
 
-    res_cuba = map(Q->Predict.by_cubature(Q, 0.5, n, q, logr; reltol=reltol, ε=ε)[1], Qs)
+    # res_known = map(Q->Predict.by_sum(Q, α, β, a, n;ε=ε)[1], Qs)
+    # normalize!(Qs, res_known, "known α, β")
 
+    dist_fit = Distributions.fit_mle(Gamma, samples)
+    res_mle = map(Q->Predict.by_sum(Q, shape(dist_fit), 1/scale(dist_fit), a, n;ε=ε)[1], Qs)
+    normalize!(Qs, res_mle, "max. likelihood")
 
     first = q/n
     second = mapreduce(x->x^2, +, samples)/n
     res_asym_laplace = map(Q->Predict.asymptotic_by_laplace(Q, a, n, first, second), Qs)
+    normalize!(Qs, res_asym_laplace, "asympt. Laplace")
 
-    hello(tag, norm, mean, stderr) = info(tag, ": norm=$norm, Q=$(mean)±$(stderr)")
+    res_laplace = map(Q->Predict.by_laplace(Q, a, n, q, logr; ε=ε)[1], Qs)
[…]

### 2017-01-31 real/code/TardisPaperTest.jl

@@ -3,6 +3,9 @@ module TardisPaperTest
 import TardisPaper.Integrate, TardisPaper.Predict
 using TardisPaper.GammaIntegrand
 using Base.Test, DiffBase, Distributions, ForwardDiff, Optim # Compat
+using Base.Test, DiffBase, Distributions, ForwardDiff, Logging, Optim
+
+@Logging.configure(level=INFO)
 
 function integrate_by_cubature()
     res, σ, ncalls = Integrate.by_cubature(x -> log(x^2), 0, 1)
@@ -26,7 +29,7 @@ function predict()
 
     hello(res, alg) = println("$alg: P($Q|$n, $a, x) = $(res[1]) for N=$(res[2])...$(res[3])")
 
-    res_cuba = Predict.by_cubature(Q, 0.5, n, q, logr; reltol=1e-8)
+    res_cuba = Predict.by_cubature(Q, 0.5, n, q, logr; reltol=1e-5)
     hello(res_cuba, "cubature")
     res_laplace = Predict.by_laplace(Q, a, n, q, logr)
     hello(res_laplace, "Laplace")
@@ -43,6 +46,10 @@ function predict()
     res_asym_laplace = Predict.asymptotic_by_laplace(Q, a, n, first, second)
     hello((res_asym_laplace, 0, 0), "asympt. Laplace")
     @test_approx_eq_eps(res_asym_laplace, res_laplace[1], 3e-2)
+
+    res_asym_cubature = Predict.asymptotic_by_cubature(Q, a, n, first, second; reltol=1e-3)
+    hello((res_asym_cubature, 0, 0), "asympt. cubature")
+    @test_approx_eq_eps(res_asym_cubature, res_laplace[1], 3e-2)
 end
 
 function predict_alpha_fixed()

### 2017-01-31 real/draft.tex
[real] Fix several bugs in moments calculation
[draft] Discuss asymptotic plot in more detail

@@
 \newcommand{\refsec}[1]{Sec.~\ref{sec:#1}}
-
+\DeclareMathOperator{\erfc}{erfc}
 \DeclareMathOperator{\Expect}{\mathbb{E}}
@@
   &= \GaussianDist \left( \Lumtot \cond \expect{Q}, \variance{Q} \right) %
-    = \GaussianDist \left( \Lumtot \cond \lambda \mu_1, \lambda (\mu^2 {+} \sigma^2) \right).
+    = \GaussianDist \left( \Lumtot \cond \lambda \mu, \lambda (\mu^2 {+} \sigma^2) \right).
 \end{align}
@@
 
-\fred{In \reffig{asymptotic}, we contrast the asymptotic expression
-  with the ``exact'' expression \refeq{spk} where the integrals over
-  $\bmphi$ are done numerically to a requested relative accuracy of
-  $10^{-5}$ through the Cubature
+In \reffig{asymptotic}, we contrast the asymptotic expression with the
+``exact'' expression \refeq{spk} where the integrals over $\bmphi$ are
+done numerically to a requested relative accuracy of $10^{-5}$ through
+the Cubature
 package\cite{cubature,genz1980remarks,berntsen1991adaptive}. We
-  computed $P(\Lumtot \cond n, \bml)$ on a grid of $\Lumtot$ values
-  and checked if the probability integrates to one using Simpson's
+computed $P(\Lumtot \cond n, \bml)$ on a grid of $\Lumtot$ values and
+checked if the probability integrates to one using Simpson's
 rule. With the exact expression, we obtain 0.997 ($n=10$) and 0.996
 ($n=80$), and 0.698 and 0.953 for the asymptotic expression. For
-  $n=10$, the large discrepancy arises mostly because the constraint
-  $P(Q \leq 0 \cond n, \bml) = 0$ is not respected in \refeq{ral}
-  whereas it is satisfied by
-  \refeq{spk}. Following~\cite{tierney_accurate_1986}, we renormalized
-  the results to exactly integrate to one for ease of comparison in
-  \reffig{asymptotic}. Note that the lack of normalization may not be
-  relevant; e.g. if \refeq{ral} is used as the likelihood for a single
-  bin in Bayesian parameter inference.
-
-  We verified that for $n>80$, the two methods agree even better than
-  shown there and that the normalization of the asymptotic result
-  approaches one. }
+$n=10$, the large discrepancy arises mostly because the positivity
+constraint $P(Q \leq 0 \cond n, \bml) = 0$ is not respected in
+\refeq{ral} since the Gaussian approximation \refeq{rae} allows
+negative $\lambda$ and $\Lumtot$, whereas the results \refeq{spk}
+based on the $\GammaDist$ model imply $\lambda \geq 0$ and $Q \geq 0$.
+
+For that reason, we discourage the asymptotic approximation for small
+$n$ but we note that in general, the lack of numerical normalization
+may not be a concern. For example, if \refeq{ral} is used as the
+likelihood for a single bin in Bayesian parameter inference, the
+normalization or evidence is
+irrelevant. Following~\cite{tierney_accurate_1986}, we renormalized
+the results to integrate to one for ease of comparison in
+\reffig{asymptotic}.  We verified that for $n>80$, the two methods
+agree even better than shown there and that the normalization of the
+asymptotic result approaches one.
+
+The main advantage of the approximation is computational efficiency. A
+single evaluation of $P(Q \cond n, \bml)$ takes only about 100 calls
+to the integrand of \refeq{ral} regardless of the value of $n$. In
+contrast, a the numerical integration of a single term for fixed $N$
+in \refeq{spk} requires 1000s of function evaluations, and the number
+of terms grows like $\sqrt{n}$ for a given accuracy of
+$P(Q \cond n, \bml)$.
 
@@
   means I can't pick up a factor of two there, it would only change
-  the normalization but not the \emph{shape} of $p(Q)$}
-\fred{The error is fixed and I remove the next sections as Hans suggested.}
+  the normalization but not the \emph{shape} of $p(Q)$} \fred{The
+  error is fixed and I remove the next sections as Hans suggested.}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
   \intertext{where }
-  f\left(t \cond  \lambda, \mu, \sigma^2\right) &\equiv \int \rmdx{\Lumtot} \GaussianDist \left( \Lumtot \cond \lambda \mu, \lambda \sigma^2 \right) e^{-t \Lumtot}= \frac{1}{2} e^{t \lambda (t \sigma^2/2 - \mu)} \left[ 1 + \erf\qty( \frac{\sqrt{\lambda}(\mu - t \sigma^2)}{\sqrt{2 \sigma^2}}) \right].
+  f\left(t \cond  \lambda, \mu, \sigma^2\right) &\equiv \int \rmdx{\Lumtot} \GaussianDist \left( \Lumtot \cond \lambda \mu, \lambda (\mu^2 + \sigma^2) \right) e^{-t \Lumtot}\\
+  &= \frac{1}{2} \exp \qty(t/2 \lambda (t (\mu^2 + \sigma^2)- 2\mu)) \erfc \qty( \frac{\sqrt{\lambda}(t \qty(\mu^2 + \sigma^2) - \mu)}{\sqrt{2 \qty(\mu^2 + \sigma^2)}}).
 \end{align}
@@
 \end{align}
-\fred{Did we define expectation value and variance? Should be in notation or intro section. Hans: see footnote 2.} %
 \fred{At least for us, I should do a comparison plot to evaluate $p(\Lumtot \cond \bmy)$ with both methods } %

### 2017-01-31 real/code/Moments.jl
(wip)[code] Add Moments.jl
Copy over moments calculation from Asymptotic.jl

@@ -1,0 +1,145 @@
+module Moments
+
+using Optim
+import ForwardDiff
+# chunk = ForwardDiff.Chunk{3}()
+"""
+Enumeration to specify which target function to create
+* PosteriorParam p(\lambda, \mu, \sigma^2 | data)
+"""
+@enum TARGET PosteriorParam NormalAndPosterior FirstMoment SecondMoment ThirdMoment FourthMoment
+
+"""
+Create a function for Laplace approximation
+"""
+function targetfactory(target::TARGET, n::Int64, xmean::Float64, xsecond::Float64;
+                       a::Float64=1, Q::Float64=-1.0)
+    0 <= a <= 1 || error("need a in [0,1]")
+    (xmean > 0) || error("nonpositive xmean $xmean")
+    (xsecond > 0) || error("nonpositive xsecond $xsecond")
+
+    """ p(\lambda | data) p(\mu, \sigma^2 | data)"""
+    function logposterior_param(x::Vector)
+        # all values must be positive
+        any(v->v<0, x) && return -Inf
+        λ, μ, σSq = x
+
+        # gamma(λ), set β=1
+        α = n-a+1
+        resgamma = -lgamma(α) + (α-1)*log(λ) - λ
+
+        # Eggers: flat prior in μ, InvGamma(σ^2|a0, b0)
+        σSqn = σSq / n
+        an = 0 + n/2
+        bn = 0 + n/2 * (xsecond - xmean^2)
+        resnormal = -1/2 * (log(2π*σSqn) + (μ-xmean)/σSqn)
+        resinvgamma = an * log(bn) - lgamma(an) - (an + 1)*log(σSq) - bn / σSq
+
+        # λ         μ           σ^2 | μ
+        resgamma + resnormal + resinvgamma
+    end
+
+    """N(Q | \lambda \mu, \lambda(\mu^2 + \sigma^2))"""
+    function lognormal(Q::Real, x::Vector)
+        λ, μ, σSq = x
+        # println("Q = $Q ", λ, μ, σSq )
+        var = λ*(μ^2 + σSq)
+        -1/2(log(2π*var) + (Q-λ*μ)^2 / var)
+    end
+
+    """log(-d/dt f(target=0|x))"""
+    function logfirst(x::Vector)
+        λ, μ, σSq = x
+        var = σSq + μ^2
+        exponent = λ*μ^2 / (2*var)
[…]

### 2017-01-31 real/code/TardisPaper.jl

@@ -2,5 +2,6 @@ module TardisPaper
 
 include("Integrate.jl")
 include("GammaIntegrand.jl")
+include("Moments.jl")
 include("Predict.jl")
 end # TardisPaper

### 2017-01-31 real/code/TardisPaperTest.jl

@@ -1,8 +1,7 @@
 module TardisPaperTest
 
-import TardisPaper.Integrate, TardisPaper.Predict
+import TardisPaper.Integrate, TardisPaper.Predict, TardisPaper.Moments
 using TardisPaper.GammaIntegrand
-using Base.Test, DiffBase, Distributions, ForwardDiff, Optim # Compat
 using Base.Test, DiffBase, Distributions, ForwardDiff, Logging, Optim
 
 @Logging.configure(level=INFO)
@@ -188,6 +187,11 @@ function asymptotic()
     Predict.asymptotic_by_laplace(Q, a, n, first, second)
 end
 
+# function moments()
+#     Moments.uncertainty(100, )
+#     @test false
+# end
+
 function run()
     @testset "all tests" begin
         integrate_by_cubature()
@@ -195,6 +199,7 @@ function run()
         predict_alpha_fixed()
         gamma_integrand()
         asymptotic()
+        # moments()
     end
 end
 

### 2017-02-01 real/code/TardisPaperPlots.jl
(wip)[code] Plot spectrum

@@ -3,6 +3,7 @@ using Logging
 @Logging.configure(level=INFO)
 
 import TardisPaper.Predict, TardisPaper.Integrate
+import Tardis
 using Distributions, LaTeXStrings, Plots
 
 const font = Plots.font("TeX Gyre Heros")
@@ -132,6 +133,34 @@ function cuba_vs_laplace(res)
     xlabel!(L"Q")
     ylabel!(L"P(Q)")
     savepdf("cuba_vs_laplace")
+end
+
+function prepare_spectrum()
+    # read in normalized spectrum
+    raw_data = Tardis.readdata()
+    frame = Tardis.filter_positive(raw_data...)
+    Tardis.transform_data!(frame)
+
+    # update energies by drawing from Gamma distribution.
+    # actual samples are only approximately from Gamma
+    dist = Gamma(1.5, 1/60)
+    for i in 1:length(frame[:energies])
+        frame[:energies][i] = rand(dist)
+    end
+    frame
+end
+
+function plot_spectrum(frame)
+    using StatsBase
+    h=fit(Histogram, frame[:nus], weights(frame[:energies]); nbins=200)
+    # off by one
+    # plot(h.edges[1], h.weights)
+
+    # works, but doesn't show the right graph
+    plot(h)
+
+    # manually subtract by one
+    plot(0.01:0.02:4.3, h.weights)
 end
 
 end # module

### 2017-02-12 real/draft.tex
Commit-All: "yes" -- End of Magit header --

@@
 %% CHANGE THIS DATE AS YOU MODIFY THE FILE
-\newcommand{\vdate}{161206}
+\newcommand{\vdate}{170212}
 \pagestyle{myheadings}
@@
 luminosities $Q_b, b = 1,\ldots,B$ if many simulations had been
-carried out. At this introductory level, we assume that the different
-luminosities are independent, so that the prediction of the entire
-spectrum (including all its uncertainty) is the product of predictions
-for each frequency bin,
+carried out. At this introductory level, we assume that the
+luminosities at different frequencies are independent, so that the
+prediction of the entire spectrum (including all its uncertainty) is
+the product of predictions for each frequency bin,
 \begin{align}
@@
 The uninteresting variables $V$ in the present case are the number of
-packets $N$, the set of luminosities $\bmL$ of the $N$ packets
+packets $N$, the set of individual luminosities $\bmL$ of the $N$ packets
 whose likelihood is modelled by a distribution determined by
@@
   \int \rmdx{\bmL}\; p(Q\cond N,\bmL)\; p(\bmL\cond N,\bmphi)
-  &= \int \rmdx{\bmL}\;\delta(Q - \textstyle\sum_{j=1}^N L_j)\; \prod_j p(L_j\cond \bmphi).
+  &= \int \rmdx{\bmL}\;\delta(Q - \textstyle\sum_{j=1}^N L_j)\; \prod_{j=1}^N p(L_j\cond \bmphi).
 \end{align}
@@
 functional form of each probability. In Section \ref{sec:example}, we
-shall therefore set out a specific example
-
-\section{Example A: modelling single-packet luminosity distributions}
+shall therefore set out a specific example.
+
+\section{Example: modelling single-packet luminosity distributions}
 \label{sec:example}
@@
 First, the parameter $\lambda$ is eliminated as follows. The number of
-packets $N$ is Poisson-distributed
+packets $N$ is Poisson-distributed,
 \begin{align}
@@
   course a parameter $\lambda_{\rm max}$. As long as
-  $\lambda_{\rm max}$ is large enough, it too would cancel.}, The
+  $\lambda_{\rm max}$ is large enough, it would also cancel.}, The
 posterior is
@@
 and inserting Eqs.~(\ref{pmb}) and (\ref{pmc}), the prediction for $N$
-given $n$ is
+given $n$ is, for all possible values of $\lambda$,
 \begin{align}
@@
 \end{align}
-so that Eq.~(\ref{sll}) simplifies to
+(a negative binomial distribution), so that Eq.~(\ref{sll}) simplifies to
 \begin{align}
@@
     p(\bmL\cond N,\bmphi)\,
-    p(\bmphi\cond n,\bml),
+    p(\bmphi\cond n,\bml).
 \end{align}
@@
 
-Which specific probability function to choose for $p(L\cond\bmtheta)$
+Which specific probability function to choose for $p(L_j\cond\bmphi)$
 depends, of course, on the specific simulation data. In this section,
@@
   \int \rmdx{\bmL}\; p(Q\cond N,\bmL)\; p(\bmL\cond N,\bmphi)
-  &= \GammaDist(Q\cond N\alpha, \beta).
-\end{align}
-and so (\ref{gmh}) becomes in this example
+  &= \GammaDist(Q\cond N\alpha, \beta),
+\end{align}
+and so Eq.~(\ref{gmh}) becomes in this example
 \begin{align}
@@
 % since $\alpha_{\rm max}$ and $\beta_{\rm max}$ cancel in (\ref{spf}).
-The posterior is
+With these choices, the posterior is found to be
 \begin{align}
@@
   p(\alpha,\beta\cond n,\bml)
-  &= \frac{1}{C}\; \frac{\beta^{n\alpha} r^{\alpha-1}\,e^{-\beta q}}{[\Gamma(\alpha)]^n}\\
+  &= \frac{1}{C}\; \frac{\beta^{n\alpha} r^{\alpha-1}\,e^{-\beta q}}{[\Gamma(\alpha)]^n},\\
   \lleq{spj}
   C &= \int_1^{\alpha_{\rm max}} d\alpha \int_0^{\beta_{\rm max}} d\beta\;
-    \frac{\beta^{n\alpha} r^{\alpha-1}\,e^{-\beta q}}{[\Gamma(\alpha)]^n}
+    \frac{\beta^{n\alpha} r^{\alpha-1}\,e^{-\beta q}}{[\Gamma(\alpha)]^n},
 \end{align}
@@
     \GammaDist(Q\cond N\alpha, \beta)\;
-    \frac{\beta^{n\alpha} r^{\alpha-1}\,e^{-\beta q}}{[\Gamma(\alpha)]^n}
+    \frac{\beta^{n\alpha} r^{\alpha-1}\,e^{-\beta q}}{[\Gamma(\alpha)]^n}.
 \end{align}
@@
 and continue to larger and smaller values of $N$ until the
-contribution to $P(Q \cond n,\bml)$ becomes negligible. In case the
+contribution to $p(Q \cond n,\bml)$ becomes negligible. In case the
 relevant contribution is from large $N$ and $n$, we interpret $N$ as a
@@
 
-We infer $P(X \cond n,\bml)$ and only in the end transform back to a
+We infer $p(X \cond n,\bml)$ and only in the end transform back to a
 probability density in $\Lumtot$ via
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Asymptotic approximation} \label{sec:asymptotic}
+\section{Large packet number approximations} \label{sec:asymptotic}
 
@@
     p(\bmphi\cond n,\bml)\;
-    p(Q\cond \lambda,\bmphi) \\
+    p(Q\cond \lambda,\bmphi), \\
   \lleq{asd}
@@
     p(Q\cond N,\bmL)\;
-    p(\bmL\cond N,\bmphi)
+    p(\bmL\cond N,\bmphi),
 \end{align}
@@
 %
-Higher-order quantities such as the skewness and kurtosis
-are suppressed by powers of $\lambda^{-1/2}$.
+Higher-order quantities such as the skewness and kurtosis are
+suppressed by powers of $\lambda^{-1/2}$.\footnote{ The variance is
+  the second so-called \textit{cumulant}. The relation for all orders
+  is that the $k$-th order cumulant $\kappa_q(Q)$ is equal to
+  $\lambda$ times the $k$-th order moment of $L$, $\mu_k$. The
+  skewness of $p(Q\cond \lambda,\bmphi)$ is therefore
+  $\kappa_3(Q)/\kappa_2(Q)^{3/2} = \lambda \mu_3/(\lambda \mu_2)^{3/2}
+  = \lambda^{-1/2} \mu_3/\mu_2^{3/2}$.}  %
 %%%%%%%%%%%
-\hans{The variance is the second so-called \textit{cumulant}. The
-  relation for all orders is that the $k$-th order cumulant
-  $\kappa_q(Q)$ is equal to $\lambda$ times the $k$-th order moment of
-  $L$, $\mu_k$. The skewness is therefore
-  $\kappa_3(Q)/\kappa_2(Q)^{3/2} = \lambda \mu_3/(\lambda \mu_2)^{3/2}
-  = \lambda^{-1/2} \mu_3/\mu_2^{3/2}$. The magnitude of the correction
-  to (rab) due to $\kappa_3$ and higher orders is unclear.}
+\hans{The magnitude of the correction to Eq.~(\ref{rab}) due to
+  $\kappa_3$ and higher orders is unclear.}
 %%%%%%%%%%%
@@
   \lleq{rab}
-  P(\Lumtot \cond \lambda,\bmphi)
+  p(\Lumtot \cond \lambda,\bmphi)
   &= \GaussianDist \left( \Lumtot \cond \expect{Q}, \variance{Q} \right) %
@@
 distribution prior with hyperparameters $(a_0,b_0)$ which reads
-\fred{Inconsistentcy. Now $b_0$ is a scale parameter but before $\beta$ was a \emph{rate} parameter in definition of $\Gamma$}
+%%%%%%%
+\fred{Inconsistency. Now $b_0$ is a scale parameter but before $\beta$
+  was a \emph{rate} parameter in definition of $\Gamma$} 
+%%%%%%%
+\hans{Not really. This is an inverse gamma distribution, not a gamma
+  distribution.  Transformation from $\sigma$ to gamma variable
+  $v = 1/\sigma_2$ would make $b_0$ a rate variable for the latter.}
 \begin{align}

### 2017-02-12 real/draft.tex
Section on Poisson-Gamma convolution

@@
 %
-Higher-order quantities such as the skewness and kurtosis are
-suppressed by powers of $\lambda^{-1/2}$.\footnote{ The variance is
-  the second so-called \textit{cumulant}. The relation for all orders
-  is that the $k$-th order cumulant $\kappa_q(Q)$ is equal to
-  $\lambda$ times the $k$-th order moment of $L$, $\mu_k$. The
-  skewness of $p(Q\cond \lambda,\bmphi)$ is therefore
+Higher-order quantities such as the skewness and kurtosis\footnote{
+  The variance is the second so-called \textit{cumulant}. The relation
+  for all orders is that the $k$-th order cumulant $\kappa_q(Q)$ is
+  equal to $\lambda$ times the $k$-th order moment of $L$,
+  $\mu_k$. The skewness of $p(Q\cond \lambda,\bmphi)$ is therefore
   $\kappa_3(Q)/\kappa_2(Q)^{3/2} = \lambda \mu_3/(\lambda \mu_2)^{3/2}
   = \lambda^{-1/2} \mu_3/\mu_2^{3/2}$.}  %
+are suppressed by powers of $\lambda^{-1/2}$.
 %%%%%%%%%%%
-\hans{The magnitude of the correction to Eq.~(\ref{rab}) due to
-  $\kappa_3$ and higher orders is unclear.}
+\hans{H comment (not for text): The magnitude of the correction to
+  Eq.~(\ref{rab}) due to $\kappa_3$ and higher orders is unclear.}
 %%%%%%%%%%%
@@
 approximation~\cite[Ch. 27]{mackay2003information} can be applied.
+
+\subsection{Poisson-gamma convolution}\label{sec:poisson-gamma}
+
+\hans{H: This subsection is temporary. It is motivated by the question
+  whether the use of a gaussian for the single-packet luminosity
+  $p(L_j\cond\bmphi)$ in (\ref{rae}) is necessary and whether making
+  it a gaussian is a source of bias since $L_j$ cannot be
+  negative. The original choice of a Gamma distribution as in
+  (\ref{spc}) is more appropriate and could eliminate one source of
+  systematic bias. Below, I have derived a closed form for
+  $p(Q\cond \lambda,\bmphi)$ using the gamma-distributed
+  $p(L_j\cond \alpha,\beta)$ of (\ref{spc}). The final result
+  (\ref{ffin}) below could be used to replace the gaussian result
+  (\ref{rab}) for insertion into (\ref{asd}). Inclusion of the entire
+  derivation below is not envisaged; rather, we would include only
+  salient issues and equation in the text itself. For the moment, the
+  question is what Fred has to say about this version ie whether it
+  can be implemented numerically and how it compares to the above
+  results.
+
+  The limit of this section to Gaussian can be investigated later if necessary.
+}
+
+We start again from (\ref{asd}) and use (\ref{spc})
+\begin{align}
+  \lleq{asdd}
+  p(Q\cond \lambda,\bmphi)
+  &= \sum_N
+    p(N\cond \lambda)\;
+    \int \rmdx{\bmL}\;
+    p(Q\cond N,\bmL)\;
+    p(\bmL\cond N,\bmphi)\\
+  \lleq{spcd}
+  p(L_j\cond \alpha,\beta) 
+  &= \frac{\beta^\alpha L_j^{\alpha-1}e^{-\beta L_j}}{\Gamma(\alpha)}.
+\end{align}
+The moment generating function of $p(L_j\cond \alpha,\beta)$ is
+\begin{align}
+  M[t\cond p(L_j\cond \alpha,\beta) ]
+  &= E[e^{tL_j}] = \int_0^\infty dL_j e^{tL_j}\,p(L_j\cond \alpha,\beta) 
+    = (1 - t/\beta)^{-\alpha}
+\end{align}
+The convolution of $N$ Gamma distributions yields the moment generating
+function
+\begin{align}
+  M[t\cond p(Q\cond \alpha,\beta) ]
+  &= (1-t/\beta)^{-N\alpha}
+\end{align}
+and so we obtain the well-known result that the convolved distribution
+is also a Gamma distribution with shape parameter $N\alpha$,
+\begin{align}
+  p(Q\cond \alpha,\beta) 
+  &= \frac{\beta^{N\alpha} Q^{N\alpha-1}e^{-\beta Q}}{\Gamma(N\alpha)}.
+\end{align}
+Inserting this into the $N$-sum, we have
+\begin{equation}
+  \lleq{gms}
+  \boxed{
+  p(Q\cond \lambda,\alpha,\beta) 
+  = \frac{e^{-\lambda - \beta Q}}{Q}
+    \sum_{N=0}^\infty 
+    \frac{[\lambda \beta^\alpha Q^\alpha]^N}{N!\;\Gamma(N\alpha)}.
+    }
+\end{equation}
+This series itself should converge quickly because of the
+$N!\;\Gamma(N\alpha)$ in the denominator. It is possible to find an
+analytic solution for some special cases of $\alpha$.
+\begin{itemize}
+\item For $\alpha=1$ we have
+  \begin{align}
+    p(Q\cond \lambda,\alpha{=}1,\beta) 
+    &= \frac{e^{-\lambda - \beta Q}}{Q}
+      \sum_{N=0}^\infty 
+      \frac{[\lambda \beta Q]^N}{N!\;(N-1)!}.
+  \end{align}
+  Comparing the above to the series definition of the modified Bessel function,
+  \begin{align}
+    I_1(z) &= \sum_{k=0}^\infty \frac{(z/2)^{2k+1}}{k!\;(k+1)!},
+  \end{align}
+  we can identify $z = 2\sqrt{\lambda\beta Q}$ to get
+  \begin{equation}
+    \boxed{
+      p(Q\cond \lambda,\alpha{=}1,\beta) 
+      = \sqrt{\frac{\lambda\beta}{Q}}  e^{-\lambda - \beta Q} 
+      \;I_1\!\left(2\sqrt{\lambda\beta Q}\right)
+      }
+  \end{equation}
+  as derived by Fisher and Cornish (1960). We often prefer to use the 
+  corresponding hypergeometric function formulation, in terms of which
+  \begin{align}
+    p(Q\cond \lambda,\alpha{=}1,\beta) 
+    &= \lambda\beta\; e^{-\lambda - \beta Q} 
+      \;{}_0F_1(2; \lambda\beta Q).
+  \end{align}
+
+\item For integer $\alpha = 2,3,\ldots$, we rely on a paper by Withers and
+  Nadarajah (2011) from which we get an expression for the (p,q) hypergeometric 
+  function, defined as
+  \begin{align}
+    {}_pF_q\left( a_1,a_2,\ldots, a_p\;;\; b_1,b_2,\ldots,b_q\;;\; z \right)
+    &= \sum_{k=0}^\infty \frac{z^k}{k!} \;
+      \frac{(a_1)_k(a_2)_k\cdots (a_p)_k}{(b_1)_k(b_2)_k\cdots (b_q)_k}
+  \end{align}
+  in terms of the Pochhammer symbol $(a)_k = \Gamma(a+k)/\Gamma(a)$. Withers
+  and Nadarajah found that (and I checked)
+  \begin{align}
+    (N\alpha)! 
+    &= (\alpha^\alpha)^N 
+      \left(\frac{1}{\alpha}\right)_k
+      \left(\frac{2}{\alpha}\right)_k
+      \left(\frac{\alpha{-}1}{\alpha}\right)_k
+      \left(\frac{\alpha}{\alpha}\right)_k
+  \end{align}
+  which, when inserted into the appropriate hypergeometric function
+  ${}_0F_\alpha$, yields
+  \begin{align}
+    \lleq{qpfa}
+    {}_0F_\alpha\left[ 0,0,\ldots,0 \;;\;
+    \frac{1}{\alpha},
+    \frac{2}{\alpha},
+    \ldots,
+    \frac{\alpha{-}1}{\alpha},
+    \frac{\alpha}{\alpha}
+    \;;\; z \right]
+    &= \sum_{k=0}^\infty \frac{z^k}{k!} \;\frac{(\alpha^\alpha)^k}{(k\alpha)!}
+  \end{align}
+  and with the general relation (see DLMF website)
+  \begin{align}
+    z\frac{d}{dz} {}_0F_p(\;;\;b_1,b_2,\ldots,b_q\;;\;z)
+    &= \frac{z}{b_1b_2\cdots b_q}
+      {}_0F_p(\;;\;b_1{+}1,b_2{+}1,\ldots,b_q{+}1\;;\;z)
+  \end{align}
+  we obtain, with the assignment
+  \begin{align}
+    z &= \lambda (\beta Q / \alpha)^\alpha \\
+    \lleq{qpfe}
+    p(Q\cond \lambda,\alpha,\beta)
+      &= \frac{\alpha e^{-\lambda - \beta Q} }{Q} 
+        z\frac{d}{dz} \;
+        {}_0F_\alpha\left[ 0,0,\ldots,0 \;;\;
+        \frac{1}{\alpha},
+        \frac{2}{\alpha},
+        \ldots,
+        \frac{\alpha{-}1}{\alpha},
+        \frac{\alpha}{\alpha}
+        \;;\; z \right]
+        \nonumber\\
+      &= \frac{\alpha e^{-\lambda - \beta Q} }{Q} 
+        \frac{z}{(\alpha-1)!\;\alpha^{\alpha-1}} \;
+        {}_0F_\alpha\left[ 0,0,\ldots,0 \;;\;
+        1{+}\frac{1}{\alpha},
+        1{+}\frac{2}{\alpha},
+        \ldots,
+        1{+}\frac{\alpha{-}1}{\alpha},
+        1{+}\frac{\alpha}{\alpha}
+        \;;\; z \right]
+        \nonumber\\
+  \end{align}
+  which on insertion of $z$ and simplification yields for integer $\alpha$
+  \begin{equation}
+    \lleq{qfp}
+    \boxed{
+      p(Q\cond \lambda,\alpha,\beta)
+      = 
+    \frac{\lambda\beta^\alpha Q^{\alpha-1} e^{-\lambda - \beta Q} }{(\alpha-1)!}\;
+    {}_0F_\alpha\left[ \;;\;
+        \frac{\alpha{+}1}{\alpha},
+        \frac{\alpha{+}2}{\alpha},
+        \ldots,
+        \frac{2\alpha{-}1}{\alpha},
+        2
+        \;;\; \lambda\left(\frac{\beta Q}{\alpha}\right)^\alpha \;\right]
+      }
+  \end{equation}
+  which includes the $I_1$ Bessel function for $\alpha=1$. Mathematica
+  has the function ${}_pF_q$ available. Where it is not available, the
+  above result should reduce, of course, to the series (\ref{gms}).
+\end{itemize}
+
+\subsection{Numerical implementation}\label{sec:asympt-numeric}
 
@@
 
-\subsection{Numerical implementation}\label{sec:asympt-numeric}
+\subsection{Moments of the luminosity prediction}\label{sec:predict-moments}
 
@@
   just the expectation values?} %
+
+\hans{H: I would agree to remove this section; I do not see value added.}
 

### 2017-02-13 real/code/TardisPaperPlots.jl
[Plots] First spectrum plot

@@ -2,26 +2,28 @@ module TardisPaperPlots
 using Logging
 @Logging.configure(level=INFO)
 
-import TardisPaper.Predict, TardisPaper.Integrate
+import TardisPaper.Predict, TardisPaper.Integrate, TardisPaper.Moments, TardisPaper.SmallestInterval
 import Tardis
-using Distributions, LaTeXStrings, Plots
+using DataFrames, Distributions, LaTeXStrings, Plots, StatPlots, StatsBase
 
 const font = Plots.font("TeX Gyre Heros")
 Plots.pyplot(guidefont=font, xtickfont=font, ytickfont=font, legendfont=font)
 Plots.PyPlotBackend()
+# Plots.gr()
 
 srand(16142)
 
 "save and replot"
 savepdf(fname) = begin Plots.pdf("../figures/$fname"); plot!() end
 
-function normalize!(Qs, res, tag)
+function normalize!(Qs, res, tag="")
     # norm only useful from first call
     norm, _, _ = Integrate.simpson(Qs, res)
     res ./= norm
     # but mean and σ are affected by norm
     _, mean, stderr = Integrate.simpson(Qs, res)
-    info(tag, ": norm=$norm, Q=$(mean)±$(stderr)")
+    (tag != "") && info(tag, ": norm=$norm, Q=$(mean)±$(stderr)")
+    norm, mean, stderr
 end
 
 function compute_prediction(;n=400, Qs=false, Qmin=1e-3, Qmax=2, nQ=50, α=1.5, β=60.0, a=1/2, ε=1e-3, reltol=1e-3)
@@ -78,10 +80,10 @@ function compute_all_predictions()
     kwargs[:Qs] = linspace(0.5, 3.5, 100)
     res[n] = compute_prediction(;kwargs...)
 
-    n = 200
-    kwargs[:n] = n
-    kwargs[:Qs] = linspace(2.5, 8, 100)
-    res[n] = compute_prediction(;kwargs...)
+    # n = 200
+    # kwargs[:n] = n
+    # kwargs[:Qs] = linspace(2.5, 8, 100)
+    # res[n] = compute_prediction(;kwargs...)
 
     # n = 500
     # kwargs[:n] = n
@@ -135,8 +137,8 @@ function cuba_vs_laplace(res)
     savepdf("cuba_vs_laplace")
 end
 
-function prepare_spectrum()
-    # read in normalized spectrum
+function prepare_frame()
+    # read in normalized sp
     raw_data = Tardis.readdata()
[…]

### 2017-02-13 real/code/Moments.jl
[moments] macros to avoid boilerplate code

@@ -5,6 +5,31 @@ import ..Integrate
 using Optim
 import ForwardDiff
 # chunk = ForwardDiff.Chunk{3}()
+
+"Warn if `arg` negative and return `log(0)`. Else return `nothing`"
+macro check_positive(arg)
+    quote
+        # evaluate arg only once to avoid unnecessary side effects
+        t0 = $arg
+        if t0 < 0
+            warn($(string(arg)), " = $t0 < 0")
+            return -Inf
+        end
+    end
+end
+
+"Create several variables and check them for positivity"
+macro unroll_args(x)
+    esc(quote
+        λ, μ, σSq = x
+        @check_positive(λ)
+        @check_positive(μ)
+        @check_positive(σSq)
+        var = σSq + μ^2
+        exponent = λ*μ^2 / (2*var)
+    end)
+end
+
 """
 Enumeration to specify which target function to create
 * PosteriorParam p(\lambda, \mu, \sigma^2 | data)
@@ -39,37 +64,26 @@ function targetfactory(target::TARGET, n::Int64, Qfirst::Float64, Qsecond::Float
 
     """N(Q | \lambda \mu, \lambda(\mu^2 + \sigma^2))"""
     function lognormal(Q::Real, x::Vector)
-        λ, μ, σSq = x
-        # println("Q = $Q ", λ, μ, σSq )
-        var = λ*(μ^2 + σSq)
+        @unroll_args(x)
         -1/2(log(2π*var) + (Q-λ*μ)^2 / var)
     end
 
     """log(-d/dt f(target=0|x))"""
     function logfirst(x::Vector)
-        λ, μ, σSq = x
-        λ > 0 || error("Negative λ ", λ)
-        μ > 0 || error("Negative μ ", μ)
-        σSq > 0 || error("Negative σSq ", σSq)
-        var = σSq + μ^2
-        exponent = λ*μ^2 / (2*var)
+        @unroll_args(x)
         log(exp(-exponent) * sqrt(λ * var / (2π)) + 1/2 * λ * μ * (1 + erf(sqrt(exponent))))
     end
 
     """log(d²/dt² f(target=0|x))"""
     function logsecond(x::Vector)
-        λ, μ, σSq = x
-        var = σSq + μ^2
-        exponent = λ*μ^2 / (2*var)
[…]

### 2017-02-13 real/code/TardisPaperPlots.jl
[PaperPlots] compute second moment directly
Avoid negative values if computed from variance and mean

@@ -159,6 +159,7 @@ function analyze_samples(frame; npackets=typemax(Int64), nbins=20)
     edges = linspace(0.0, nus[end], nbins+1)
     n = Array{Int64}(nbins)
     means = zeros(nbins)
+    seconds = zeros(nbins)
     variances = zeros(nbins)
     logr = zeros(nbins)
     for (i, edge) in enumerate(edges[1:end-1])
@@ -167,9 +168,10 @@ function analyze_samples(frame; npackets=typemax(Int64), nbins=20)
         n[i] = nmax - nmin + 1
         # println("$nmin, $nmax")
         if n[i] == 0
-            means[i] = variances[i] = logr[i] = 0.0
+            means[i] = seconds[i] = variances[i] = logr[i] = 0.0
         else
             means[i], variances[i] = mean_and_var(energies[nmin:nmax])
+            seconds[i] = mapreduce(x->x^2, +, energies[nmin:nmax]) / n[i]
             # remove Bessel correction
             variances[i] *= (n[i] - 1) / n[i]
             logr[i] = n[i] * mean(log, energies[nmin:nmax])
@@ -181,13 +183,13 @@ function analyze_samples(frame; npackets=typemax(Int64), nbins=20)
 
     return DataFrame(left_edge=edges[1:end-1],
                      center=(edges[2:end] + edges[1:end-1])/2,
-                     n=n, mean=means, variance=variances, logr=logr)
+                     n=n, mean=means, second=seconds, variance=variances, logr=logr)
 
 end
 
-function Qrange(n, Qmean, Qvar; k=5, nbins=100, minQ=0.0)
+function Qrange(n, Qmean, Qsecond; k=5, nbins=100, minQ=0.0)
     # estimate moments
-    μ, σ = Moments.uncertainty(n, Qmean, Qvar+Qmean^2)
+    μ, σ = Moments.uncertainty(n, Qmean, Qsecond)
 
     # TODO what if n small? n == 1?
     # use MLE result with enlargement or some other
@@ -213,9 +215,9 @@ n < 10: cubature sum
 10 <= n <= 80: Laplace sum
 80 < n: asymptotic Laplace
 """
-function predict_Q_bin(n, Qmean, Qvar, logr; a=0.5, nbins=50)
-    (n == 0) && error("hit empty bin")
[…]

### 2017-02-14 real/code/GammaIntegrand.jl
[code] Fix corner cases for nb=0

@@ -121,7 +121,9 @@ Use the plug-in values of α and β to avoid costly integration.
 function heuristicN(Q::Real, α::Real, β::Real, a::Real, nb::Integer;
                     ε=1e-2, min=0.0, max=0.0, trace=false)
     popt, f = factory(Q, α, β, a, nb)
-    if (max <= 0.0) max = 10nb end
+    if (max <= 0.0)
+        max = (nb > 0) ? 10nb : 50
+    end
     optimize(f, min, max, rel_tol=ε, show_trace=trace, extended_trace=trace)
 end
 

### 2017-02-14 real/code/Moments.jl

@@ -130,13 +130,17 @@ function uncertainty(n::Int64, Qfirst::Float64, Qsecond::Float64, nb::Int64=n;
     # ERROR: Linesearch failed to converge
     init = 1.005 * triple_mode(nb, Qfirst, Qsecond)
 
+    # corner case nb == 0
+    if init[1] == 0.0
+        init[1] = 0.2
+    end
+
     # Hessian
     H = Array{Float64}(3,3)
 
     function integrate(moment::TARGET; Q=-1.0)
         target = targetfactory(moment, n, Qfirst, Qsecond, nb; a=a, Q=Q)
         res = optimize(target, init, Newton(), Optim.Options(autodiff=true))
-        info(Optim.minimizer(res))
         ForwardDiff.hessian!(H, target, Optim.minimizer(res))
         exp(Integrate.by_laplace(-Optim.minimum(res), H))
     end

### 2017-02-14 real/code/TardisPaperPlots.jl
[plots] Comparison binned vs unbinned
Iron out lot of corner cases to handle nb=0 and friends properly

@@ -2,7 +2,7 @@ module TardisPaperPlots
 using Logging
 @Logging.configure(level=INFO)
 
-import TardisPaper.Predict, TardisPaper.Integrate, TardisPaper.Moments, TardisPaper.SmallestInterval
+import TardisPaper.GammaIntegrand, TardisPaper.Predict, TardisPaper.Integrate, TardisPaper.Moments, TardisPaper.SmallestInterval
 import Tardis
 using DataFrames, Distributions, LaTeXStrings, Plots, StatPlots, StatsBase
 
@@ -11,14 +11,19 @@ Plots.pyplot(guidefont=font, xtickfont=font, ytickfont=font, legendfont=font)
 Plots.PyPlotBackend()
 # Plots.gr()
 
-srand(16142)
-
 "save and replot"
 savepdf(fname) = begin Plots.pdf("../figures/$fname"); plot!() end
 
-function normalize!(Qs, res, tag="")
+"""# Arguments
+
+`δcontrib`: For `N=0`, the δ function has a contribution that affects
+the normalization and thus mean and std. err
+
+"""
+function normalize!(Qs, res, tag=""; δcontrib=0.0)
     # norm only useful from first call
     norm, _, _ = Integrate.simpson(Qs, res)
+    norm += δcontrib
     res ./= norm
     # but mean and σ are affected by norm
     _, mean, stderr = Integrate.simpson(Qs, res)
@@ -26,7 +31,9 @@ function normalize!(Qs, res, tag="")
     norm, mean, stderr
 end
 
-function compute_prediction(;n=400, Qs=false, Qmin=1e-3, Qmax=2, nQ=50, α=1.5, β=60.0, a=1/2, ε=1e-3, reltol=1e-3)
+function compute_prediction(;n=400, Qs=false, Qmin=1e-3, Qmax=2, nQ=50, α=1.5, β=60.0, a=1/2, ε=1e-3, reltol=1e-3, seed=16142)
+    srand(seed)
+
     info("Computing for n=$n")
     dist = Gamma(α, 1/β)
     samples = rand(dist, n)
@@ -137,11 +144,14 @@ function cuba_vs_laplace(res)
     savepdf("cuba_vs_laplace")
 end
 
-function prepare_frame()
+"Fixed seed for reproducible spectra"
+function prepare_frame(seed=1612)
     # read in normalized sp
     raw_data = Tardis.readdata()
     frame = Tardis.filter_positive(raw_data...)
     Tardis.transform_data!(frame)
+
[…]

### 2017-02-14 real/code/TardisPaperPlots.jl
[plots] Predict for n >= 2

@@ -231,9 +231,8 @@ n < 10: cubature
 n > 10: Laplace
 """
 function predict_Q_bin(n, Lmean, Lsecond, logr, nb=n; a=0.5, nbins=50)
-    # n > 8 || error("Hit bin with n = $n < 9 packets. Method unstable!")
-    # TODO is 9 the right cut off?
-    if n < 9
+    # n=2 is the minimum to constrain two parameters
+    if n < 2
         return zeros(0), zeros(0)
     end
 
@@ -310,14 +309,14 @@ function analyze_spectrum(binposterior=true; a=0.5, kwargs...)
             continue
         end
 
-        # if row[:n] == 0
-        #     scatter(Qs, res)
-        #     savepdf("pred0")
-        # end
-
         # upscale
         Qs, res = SmallestInterval.upscale(Qs, res, 1000)
         row[:mode] = Qs[indmax(res)]
+
+        if row[:n] == 0
+            scatter(Qs, res)
+            savepdf("pred0")
+        end
 
         one_sigma_region, two_sigma_region = analyze_bin(Qs, res; δcontrib=exp(GammaIntegrand.log_poisson_predict(0, row[:n], a)))
         row[:onelo] = Qs[one_sigma_region[1]]
@@ -349,18 +348,18 @@ function plot_spectrum(sp::DataFrame, name="spectrum"; maxQ=-1.0)
                   fill_between=maxQ, fillargs...)
             continue
         end
-
         plot!([row[:left_edge], r[i]], [row[:twohi], row[:twohi]];
               fill_between=row[:twolo], fillcolor=:blue, fillargs...)
 
         y = row[:mode]
-        scatter!([row[:center]], [y]; markersize=0, markercolor=:black,
-                 yerror=([y-row[:onelo]], [row[:onehi]-y])) # asymmetric yerror
+        scatter!([row[:center]], [y]; markersize=1, markercolor=:black)
+
+        # scatter!([row[:center]], [y]; markersize=0, markercolor=:black,
+        #          yerror=([y-row[:onelo]], [row[:onehi]-y])) # asymmetric yerror
         # scatter!([row[:center]], [y]; markersize=0,
         #          xerror=[row[:center]-row[:left_edge]], # symmetric xerror
         #          yerror=([y-row[:onelo]], [row[:onehi]-y])) # asymmetric yerror
     end
-
     ylims!(-0.02, maxQ)
 
     savepdf(name)

### 2017-02-15 real/draft.tex
[draft] Clarify motivation for large n

@@
 
-\subsection{Single-packet description}
+\subsection{Single-packet description} \label{sec:single-packet}
 
@@
 \section{Large packet number approximations} \label{sec:asymptotic}
+
+In practice, the situation may arise that many packets fall into a
+bin. One can then avoid modeling the luminosity distribution
+explicitly as done in \refsec{single-packet} as it becomes irrelevant
+in the limit $n \to \infty$, essentially due to the central limit
+theorem. As a bonus, the numerical effort to evaluate the basic
+\refeq{sll} drastically simplifies.
 

### 2017-02-15 real/draft.tex
[draft] Revise Poisson subsection

@@
 \end{align}
-The observed $n$ is naturally presumed to have arisen from the same Poisson
-distribution with the same $\lambda$, so $p(n\cond\lambda) =
-e^{-\lambda}\lambda^{n}/n!$. %
-% and since $n$ is just one realisation of $N$, it likewise is
-% poissonian $p(n\cond\lambda) = e^{-\lambda}\lambda^{n}/n!$ with the
-% same $\lambda$.
-The posterior for $\lambda$ is found from the data $n$ via Bayes'
-Theorem,
+The observed $n$ is naturally presumed to have arisen from the same
+Poisson distribution with the same $\lambda$.  The posterior for
+$\lambda$ is found from the data $n$ via Bayes' Theorem,
 \begin{align}
@@
 meaning that $\int_0^\infty \rmdx{\lambda}\,p(\lambda)$ is known only up to
-a constant $c$. A hyperparameter choice $a=0$ yields the uniform
-prior, $a=1/2$ gives the Jeffreys and reference prior, and $a=1$ is
-the transformation-group result advocated by
-Jaynes~\cite[Ch. 12]{jaynes2003probability}. An improper prior is not
+a constant $c$. An improper prior is not
 a problem because $c$ cancels out when calculating the posterior,
@@
 and inserting Eqs.~(\ref{pmb}) and (\ref{pmc}), the prediction for $N$
-given $n$ is, for all possible values of $\lambda$,
+given $n$ is a negative binomial distribution
 \begin{align}
@@
   &= \int_0^\infty \rmdx{\lambda}\, p(N\cond\lambda)\,p(\lambda\cond n)
-  \ =\ \binom{N{+}n{-}a}{N} \left(\frac{1}{2}\right)^{N+n-a+1}
+  \ =\ \binom{N{+}n{-}a}{N} \left(\frac{1}{2}\right)^{N+n-a+1}\, .
     %\ =\ \frac{(N+n-a)!}{N!\,(n-a)!\,2^{N+n+1-a}} \, .
 \end{align}
-(a negative binomial distribution), so that Eq.~(\ref{sll}) simplifies to
+A hyperparameter choice $a=0$ yields the uniform prior, $a=1/2$ gives
+the Jeffreys and reference prior, and $a=1$ is the
+transformation-group result advocated by
+Jaynes~\cite[Ch. 12]{jaynes2003probability}. The uniform prior has no
+solid information-theoretic foundation in this case reducing the
+choice to $a \in \{1/2, 1\}$. We examine the consequence for the
+extreme case that no packet is observed: $P(N=0 \cond n=0) =
+2^{a-1}$. For $a=1$, any $N>0$ is excluded which is undesirable unless
+we accept that repeating the simulation will never yield a packet in
+the current bin. Hence we are led to Jeffreys prior, $a=1/2$.  With
+\refeq{pmh}, \refeq{sll} simplifies to
 \begin{align}

### 2017-02-15 real/draft.tex
[draft] Remove a few comments, (re)move unnecessary sections

@@
 $\mu_{\rm max}$.
-\fred{only true if $\mu_{\rm max}$ large enough to contain posterior}
 For $\sigma^2$, we use an $n$-independent Inverse Gamma
 distribution prior with hyperparameters $(a_0,b_0)$ which reads
-%%%%%%%
-\fred{Inconsistency. Now $b_0$ is a scale parameter but before $\beta$
-  was a \emph{rate} parameter in definition of $\Gamma$}
-%%%%%%%
-\hans{Not really. This is an inverse gamma distribution, not a gamma
-  distribution.  Transformation from $\sigma$ to gamma variable
-  $v = 1/\sigma_2$ would make $b_0$ a rate variable for the latter.}
 \begin{align}
@@
 approximation~\cite[Ch. 27]{mackay2003information} can be applied.
-
-\subsection{Poisson-gamma convolution}\label{sec:poisson-gamma}
-
-\hans{H: This subsection is temporary. It is motivated by the question
-  whether the use of a gaussian for the single-packet luminosity
-  $p(L_j\cond\bmphi)$ in (\ref{rae}) is necessary and whether making
-  it a gaussian is a source of bias since $L_j$ cannot be
-  negative. The original choice of a Gamma distribution as in
-  (\ref{spc}) is more appropriate and could eliminate one source of
-  systematic bias. Below, I have derived a closed form for
-  $p(Q\cond \lambda,\bmphi)$ using the gamma-distributed
-  $p(L_j\cond \alpha,\beta)$ of (\ref{spc}). The final result
-  (\ref{ffin}) below could be used to replace the gaussian result
-  (\ref{rab}) for insertion into (\ref{asd}). Inclusion of the entire
-  derivation below is not envisaged; rather, we would include only
-  salient issues and equation in the text itself. For the moment, the
-  question is what Fred has to say about this version ie whether it
-  can be implemented numerically and how it compares to the above
-  results.
-
-  The limit of this section to Gaussian can be investigated later if necessary.
-}
-
-We start again from (\ref{asd}) and use (\ref{spc})
-\begin{align}
-  \lleq{asdd}
-  p(Q\cond \lambda,\bmphi)
-  &= \sum_N
-    p(N\cond \lambda)\;
-    \int \rmdx{\bmL}\;
-    p(Q\cond N,\bmL)\;
-    p(\bmL\cond N,\bmphi)\\
-  \lleq{spcd}
-  p(L_j\cond \alpha,\beta)
-  &= \frac{\beta^\alpha L_j^{\alpha-1}e^{-\beta L_j}}{\Gamma(\alpha)}.
-\end{align}
-The moment generating function of $p(L_j\cond \alpha,\beta)$ is
-\begin{align}
-  M[t\cond p(L_j\cond \alpha,\beta) ]
-  &= E[e^{tL_j}] = \int_0^\infty dL_j e^{tL_j}\,p(L_j\cond \alpha,\beta)
-    = (1 - t/\beta)^{-\alpha}
-\end{align}
-The convolution of $N$ Gamma distributions yields the moment generating
-function
-\begin{align}
-  M[t\cond p(Q\cond \alpha,\beta) ]
-  &= (1-t/\beta)^{-N\alpha}
-\end{align}
-and so we obtain the well-known result that the convolved distribution
-is also a Gamma distribution with shape parameter $N\alpha$,
-\begin{align}
-  p(Q\cond \alpha,\beta)
-  &= \frac{\beta^{N\alpha} Q^{N\alpha-1}e^{-\beta Q}}{\Gamma(N\alpha)}.
-\end{align}
-Inserting this into the $N$-sum, we have
-\begin{equation}
-  \lleq{gms}
-  \boxed{
-  p(Q\cond \lambda,\alpha,\beta)
-  = \frac{e^{-\lambda - \beta Q}}{Q}
-    \sum_{N=0}^\infty
-    \frac{[\lambda \beta^\alpha Q^\alpha]^N}{N!\;\Gamma(N\alpha)}.
-    }
-\end{equation}
-This series itself should converge quickly because of the
-$N!\;\Gamma(N\alpha)$ in the denominator. It is possible to find an
-analytic solution for some special cases of $\alpha$.
-\begin{itemize}
-\item For $\alpha=1$ we have
-  \begin{align}
-    p(Q\cond \lambda,\alpha{=}1,\beta)
-    &= \frac{e^{-\lambda - \beta Q}}{Q}
-      \sum_{N=0}^\infty
-      \frac{[\lambda \beta Q]^N}{N!\;(N-1)!}.
-  \end{align}
-  Comparing the above to the series definition of the modified Bessel function,
-  \begin{align}
-    I_1(z) &= \sum_{k=0}^\infty \frac{(z/2)^{2k+1}}{k!\;(k+1)!},
-  \end{align}
-  we can identify $z = 2\sqrt{\lambda\beta Q}$ to get
-  \begin{equation}
-    \boxed{
-      p(Q\cond \lambda,\alpha{=}1,\beta)
-      = \sqrt{\frac{\lambda\beta}{Q}}  e^{-\lambda - \beta Q}
-      \;I_1\!\left(2\sqrt{\lambda\beta Q}\right)
-      }
-  \end{equation}
-  as derived by Fisher and Cornish (1960). We often prefer to use the
-  corresponding hypergeometric function formulation, in terms of which
-  \begin{align}
-    p(Q\cond \lambda,\alpha{=}1,\beta)
-    &= \lambda\beta\; e^{-\lambda - \beta Q}
-      \;{}_0F_1(2; \lambda\beta Q).
-  \end{align}
-
-\item For integer $\alpha = 2,3,\ldots$, we rely on a paper by Withers and
-  Nadarajah (2011) from which we get an expression for the (p,q) hypergeometric
-  function, defined as
-  \begin{align}
-    {}_pF_q\left( a_1,a_2,\ldots, a_p\;;\; b_1,b_2,\ldots,b_q\;;\; z \right)
-    &= \sum_{k=0}^\infty \frac{z^k}{k!} \;
-      \frac{(a_1)_k(a_2)_k\cdots (a_p)_k}{(b_1)_k(b_2)_k\cdots (b_q)_k}
-  \end{align}
-  in terms of the Pochhammer symbol $(a)_k = \Gamma(a+k)/\Gamma(a)$. Withers
-  and Nadarajah found that (and I checked)
-  \begin{align}
-    (N\alpha)!
-    &= (\alpha^\alpha)^N
-      \left(\frac{1}{\alpha}\right)_k
-      \left(\frac{2}{\alpha}\right)_k
-      \left(\frac{\alpha{-}1}{\alpha}\right)_k
-      \left(\frac{\alpha}{\alpha}\right)_k
-  \end{align}
-  which, when inserted into the appropriate hypergeometric function
-  ${}_0F_\alpha$, yields
-  \begin{align}
-    \lleq{qpfa}
-    {}_0F_\alpha\left[ 0,0,\ldots,0 \;;\;
-    \frac{1}{\alpha},
-    \frac{2}{\alpha},
-    \ldots,
-    \frac{\alpha{-}1}{\alpha},
-    \frac{\alpha}{\alpha}
-    \;;\; z \right]
-    &= \sum_{k=0}^\infty \frac{z^k}{k!} \;\frac{(\alpha^\alpha)^k}{(k\alpha)!}
-  \end{align}
-  and with the general relation (see DLMF website)
-  \begin{align}
-    z\frac{d}{dz} {}_0F_p(\;;\;b_1,b_2,\ldots,b_q\;;\;z)
-    &= \frac{z}{b_1b_2\cdots b_q}
-      {}_0F_p(\;;\;b_1{+}1,b_2{+}1,\ldots,b_q{+}1\;;\;z)
-  \end{align}
-  we obtain, with the assignment
-  \begin{align}
-    z &= \lambda (\beta Q / \alpha)^\alpha \\
-    \lleq{qpfe}
-    p(Q\cond \lambda,\alpha,\beta)
-      &= \frac{\alpha e^{-\lambda - \beta Q} }{Q}
-        z\frac{d}{dz} \;
-        {}_0F_\alpha\left[ 0,0,\ldots,0 \;;\;
-        \frac{1}{\alpha},
-        \frac{2}{\alpha},
-        \ldots,
-        \frac{\alpha{-}1}{\alpha},
-        \frac{\alpha}{\alpha}
-        \;;\; z \right]
-        \nonumber\\
-      &= \frac{\alpha e^{-\lambda - \beta Q} }{Q}
-        \frac{z}{(\alpha-1)!\;\alpha^{\alpha-1}} \;
-        {}_0F_\alpha\left[ 0,0,\ldots,0 \;;\;
-        1{+}\frac{1}{\alpha},
-        1{+}\frac{2}{\alpha},
-        \ldots,
-        1{+}\frac{\alpha{-}1}{\alpha},
-        1{+}\frac{\alpha}{\alpha}
-        \;;\; z \right]
-        \nonumber\\
-  \end{align}
-  which on insertion of $z$ and simplification yields for integer $\alpha$
-  \begin{equation}
-    \lleq{qfp}
-    \boxed{
-      p(Q\cond \lambda,\alpha,\beta)
-      =
-    \frac{\lambda\beta^\alpha Q^{\alpha-1} e^{-\lambda - \beta Q} }{(\alpha-1)!}\;
-    {}_0F_\alpha\left[ \;;\;
-        \frac{\alpha{+}1}{\alpha},
-        \frac{\alpha{+}2}{\alpha},
-        \ldots,
-        \frac{2\alpha{-}1}{\alpha},
-        2
-        \;;\; \lambda\left(\frac{\beta Q}{\alpha}\right)^\alpha \;\right]
-      }
-  \end{equation}
-  which includes the $I_1$ Bessel function for $\alpha=1$. Mathematica
-  has the function ${}_pF_q$ available. Where it is not available, the
-  above result should reduce, of course, to the series (\ref{gms}).
-\end{itemize}
 
@@
 
+\section{Conclusion} \label{sec:conclusion}
+
+TODO
+
+\section{Appendix} \label{sec:appendix}
+
 \subsection{Moments of the luminosity prediction}\label{sec:predict-moments}
@@
 
-\section{Conclusion} \label{sec:conclusion}
-
-TODO
-
-\section{Appendix} \label{sec:appendix}
-
-\fred{Will be removed later}
-
-We use the shorthand $\alpha_n \equiv \alpha(\nu_n \cond
-\bmphi), \beta_n \equiv \beta(\nu_n \cond \bmphi), \alpha \equiv
-\alpha(\nu \cond \bmphi), \beta \equiv \beta(\nu \cond \bmphi)$.
-
-\subsection{Gradients} \label{sec:gradients}
-
-\begin{align}
-  \label{eq:grad-posterior}
-  \left( \firstDeriv{\alpha_i}, \firstDeriv{\beta_i} \right) \log p(\bmx \cond \alpha, \beta) &=
-  \left(
-    \sum_n \left[ \log \beta_n - \Psi(\alpha_n) + \log x_n \right] \nu_n^i \;,
-    \sum_n \left[ \frac{\alpha_n}{\beta_n} - x_n \right]\nu_n^i
-  \right)
-\end{align}
-
-\begin{align}
-  \label{eq:grad-prediction}
-  \left( \firstDeriv{\alpha_i}, \firstDeriv{\beta_i}, \firstDeriv{N} \right) \log p(X \cond N \alpha, \beta)  &=
-  \left(
-    \left[ \log \beta - \Psi(N \alpha) + \log X \right] N \nu^i,
-    \left[ \frac{N\alpha}{\beta} - X \right]\nu^i,
-    \left[ \log \beta -\Psi(N \alpha) + \log X \right] \alpha
-  \right)
-\end{align}
-
-\begin{align}
-  \label{eq:grad-nb}
-  \firstDeriv{N} \log p(N \cond n) = \Psi(N+n-a+1) - \Psi(N+1) - \log 2
-\end{align}
-
-\subsection{Hessians} \label{sec:hessians}
-
-\begin{align}
-  \label{eq:hess-posterior}
-    - \left(
-    \begin{array}{ccc}
-      \secPartial{\alpha_i}{\alpha_j} & \secPartial{\alpha_i}{\beta_j}\\
-      \dots & \secPartial{\beta_i}{\beta_j}
-    \end{array}
-  \right) \log p(\bmx \cond \alpha, \beta)
-    &= \left(
-    \begin{array}{cc}
-      \sum_n \Psi'(\alpha_n) \nu_n^{i+j} & -\sum_n \frac{\nu_n^{i+j}}{\beta_n}\\
-      \dots & \sum_n \frac{\alpha_n}{\beta_n^2} \nu_n^{i+j}
-    \end{array}
-  \right)
-\end{align}
-
- \begin{align}
-  \label{eq:hess-prediction}
-  & - \left(
-    \begin{array}{ccc}
-      \secPartial{\alpha_i}{\alpha_j} & \secPartial{\alpha_i}{\beta_j} & \secPartial{\alpha_i}{N}\\
-      \dots & \secPartial{\beta_i}{\beta_j} & \secPartial{\beta_i}{N}\\
-      \dots & \dots & \secDeriv{N}
-    \end{array}
-  \right) \log p(X \cond N \alpha, \beta)
-  \\
-  &=
-  \left(
-    \begin{array}{ccc}
-      N^2 \Psi'(N \alpha) \nu^{i+j} & -\frac{N \nu^{i+j}}{\beta} & \left[ -\log \beta +\Psi(N \alpha) + N \alpha \Psi'(N \alpha) - \log X \right] \nu^{i}\\
-      \dots &  \frac{N\alpha}{\beta^2} \nu^{i+j} & -\frac{\alpha}{\beta} \nu^{i} \\
-      \dots & \dots & \alpha^2 \Psi'(N \alpha)
-    \end{array}
-  \right)
-\end{align}
-\begin{align}
-  \label{eq:hess-nb}
-   - \secDeriv{N} \log p(N \cond n) = \Psi'(N+1) - \Psi'(N+n-a+1)
-\end{align}
+\section{Delendum}
+
+ \subsection{Poisson-gamma convolution}\label{sec:poisson-gamma}
+
+\hans{H: This subsection is temporary. It is motivated by the question
+  whether the use of a gaussian for the single-packet luminosity
+  $p(L_j\cond\bmphi)$ in (\ref{rae}) is necessary and whether making
+  it a gaussian is a source of bias since $L_j$ cannot be
+  negative. The original choice of a Gamma distribution as in
+  (\ref{spc}) is more appropriate and could eliminate one source of
+  systematic bias. Below, I have derived a closed form for
+  $p(Q\cond \lambda,\bmphi)$ using the gamma-distributed
+  $p(L_j\cond \alpha,\beta)$ of (\ref{spc}). The final result
+  (\ref{qfp}) below could be used to replace the gaussian result
+  (\ref{rab}) for insertion into (\ref{asd}). Inclusion of the entire
+  derivation below is not envisaged; rather, we would include only
+  salient issues and equation in the text itself. For the moment, the
+  question is what Fred has to say about this version ie whether it
+  can be implemented numerically and how it compares to the above
+  results.
+
+  The limit of this section to Gaussian can be investigated later if necessary.
+}
+
+We start again from (\ref{asd}) and use (\ref{spc})
+\begin{align}
+  \lleq{asdd}
+  p(Q\cond \lambda,\bmphi)
+  &= \sum_N
+    p(N\cond \lambda)\;
+    \int \rmdx{\bmL}\;
+    p(Q\cond N,\bmL)\;
+    p(\bmL\cond N,\bmphi)\\
+  \lleq{spcd}
+  p(L_j\cond \alpha,\beta)
+  &= \frac{\beta^\alpha L_j^{\alpha-1}e^{-\beta L_j}}{\Gamma(\alpha)}.
+\end{align}
+The moment generating function of $p(L_j\cond \alpha,\beta)$ is
+\begin{align}
+  M[t\cond p(L_j\cond \alpha,\beta) ]
+  &= E[e^{tL_j}] = \int_0^\infty dL_j e^{tL_j}\,p(L_j\cond \alpha,\beta)
+    = (1 - t/\beta)^{-\alpha}
+\end{align}
+The convolution of $N$ Gamma distributions yields the moment generating
+function
+\begin{align}
+  M[t\cond p(Q\cond \alpha,\beta) ]
+  &= (1-t/\beta)^{-N\alpha}
+\end{align}
+and so we obtain the well-known result that the convolved distribution
+is also a Gamma distribution with shape parameter $N\alpha$,
+\begin{align}
+  p(Q\cond \alpha,\beta)
+  &= \frac{\beta^{N\alpha} Q^{N\alpha-1}e^{-\beta Q}}{\Gamma(N\alpha)}.
+\end{align}
+Inserting this into the $N$-sum, we have
+\begin{equation}
+  \lleq{gms}
+  \boxed{
+  p(Q\cond \lambda,\alpha,\beta)
+  = \frac{e^{-\lambda - \beta Q}}{Q}
+    \sum_{N=0}^\infty
+    \frac{[\lambda \beta^\alpha Q^\alpha]^N}{N!\;\Gamma(N\alpha)}.
+    }
+\end{equation}
+This series itself should converge quickly because of the
+$N!\;\Gamma(N\alpha)$ in the denominator. It is possible to find an
+analytic solution for some special cases of $\alpha$.
+\begin{itemize}
+\item For $\alpha=1$ we have
+  \begin{align}
+    p(Q\cond \lambda,\alpha{=}1,\beta)
+    &= \frac{e^{-\lambda - \beta Q}}{Q}
+      \sum_{N=0}^\infty
+      \frac{[\lambda \beta Q]^N}{N!\;(N-1)!}.
+  \end{align}
+  Comparing the above to the series definition of the modified Bessel function,
+  \begin{align}
+    I_1(z) &= \sum_{k=0}^\infty \frac{(z/2)^{2k+1}}{k!\;(k+1)!},
+  \end{align}
+  we can identify $z = 2\sqrt{\lambda\beta Q}$ to get
+  \begin{equation}
+    \boxed{
+      p(Q\cond \lambda,\alpha{=}1,\beta)
+      = \sqrt{\frac{\lambda\beta}{Q}}  e^{-\lambda - \beta Q}
+      \;I_1\!\left(2\sqrt{\lambda\beta Q}\right)
+      }
+  \end{equation}
+  as derived by Fisher and Cornish (1960). We often prefer to use the
+  corresponding hypergeometric function formulation, in terms of which
+  \begin{align}
+    p(Q\cond \lambda,\alpha{=}1,\beta)
+    &= \lambda\beta\; e^{-\lambda - \beta Q}
+      \;{}_0F_1(2; \lambda\beta Q).
+  \end{align}
+
+\item For integer $\alpha = 2,3,\ldots$, we rely on a paper by Withers and
+  Nadarajah (2011) from which we get an expression for the (p,q) hypergeometric
+  function, defined as
+  \begin{align}
+    {}_pF_q\left( a_1,a_2,\ldots, a_p\;;\; b_1,b_2,\ldots,b_q\;;\; z \right)
+    &= \sum_{k=0}^\infty \frac{z^k}{k!} \;
+      \frac{(a_1)_k(a_2)_k\cdots (a_p)_k}{(b_1)_k(b_2)_k\cdots (b_q)_k}
+  \end{align}
+  in terms of the Pochhammer symbol $(a)_k = \Gamma(a+k)/\Gamma(a)$. Withers
+  and Nadarajah found that (and I checked)
+  \begin{align}
+    (N\alpha)!
+    &= (\alpha^\alpha)^N
+      \left(\frac{1}{\alpha}\right)_k
+      \left(\frac{2}{\alpha}\right)_k
+      \left(\frac{\alpha{-}1}{\alpha}\right)_k
+      \left(\frac{\alpha}{\alpha}\right)_k
+  \end{align}
+  which, when inserted into the appropriate hypergeometric function
+  ${}_0F_\alpha$, yields
+  \begin{align}
+    \lleq{qpfa}
+    {}_0F_\alpha\left[ 0,0,\ldots,0 \;;\;
+    \frac{1}{\alpha},
+    \frac{2}{\alpha},
+    \ldots,
+    \frac{\alpha{-}1}{\alpha},
+    \frac{\alpha}{\alpha}
+    \;;\; z \right]
+    &= \sum_{k=0}^\infty \frac{z^k}{k!} \;\frac{(\alpha^\alpha)^k}{(k\alpha)!}
+  \end{align}
+  and with the general relation (see DLMF website)
+  \begin{align}
+    z\frac{d}{dz} {}_0F_p(\;;\;b_1,b_2,\ldots,b_q\;;\;z)
+    &= \frac{z}{b_1b_2\cdots b_q}
+      {}_0F_p(\;;\;b_1{+}1,b_2{+}1,\ldots,b_q{+}1\;;\;z)
+  \end{align}
+  we obtain, with the assignment
+  \begin{align}
+    z &= \lambda (\beta Q / \alpha)^\alpha \\
+    \lleq{qpfe}
+    p(Q\cond \lambda,\alpha,\beta)
+      &= \frac{\alpha e^{-\lambda - \beta Q} }{Q}
+        z\frac{d}{dz} \;
+        {}_0F_\alpha\left[ 0,0,\ldots,0 \;;\;
+        \frac{1}{\alpha},
+        \frac{2}{\alpha},
+        \ldots,
+        \frac{\alpha{-}1}{\alpha},
+        \frac{\alpha}{\alpha}
+        \;;\; z \right]
+        \nonumber\\
+      &= \frac{\alpha e^{-\lambda - \beta Q} }{Q}
+        \frac{z}{(\alpha-1)!\;\alpha^{\alpha-1}} \;
+        {}_0F_\alpha\left[ 0,0,\ldots,0 \;;\;
+        1{+}\frac{1}{\alpha},
+        1{+}\frac{2}{\alpha},
+        \ldots,
+        1{+}\frac{\alpha{-}1}{\alpha},
+        1{+}\frac{\alpha}{\alpha}
+        \;;\; z \right]
+        \nonumber\\
+  \end{align}
+  which on insertion of $z$ and simplification yields for integer $\alpha$
+  \begin{equation}
+    \lleq{qfp}
+    \boxed{
+      p(Q\cond \lambda,\alpha,\beta)
+      =
+    \frac{\lambda\beta^\alpha Q^{\alpha-1} e^{-\lambda - \beta Q} }{(\alpha-1)!}\;
+    {}_0F_\alpha\left[ \;;\;
+        \frac{\alpha{+}1}{\alpha},
+        \frac{\alpha{+}2}{\alpha},
+        \ldots,
+        \frac{2\alpha{-}1}{\alpha},
+        2
+        \;;\; \lambda\left(\frac{\beta Q}{\alpha}\right)^\alpha \;\right]
+      }
+  \end{equation}
+  which includes the $I_1$ Bessel function for $\alpha=1$. Mathematica
+  has the function ${}_pF_q$ available. Where it is not available, the
+  above result should reduce, of course, to the series (\ref{gms}).
+\end{itemize}
 

### 2017-02-15 real/draft.tex
[draft] Clarify prior on Gamma parameters
And many other minor modifications

@@
 specific choices for the various probabilities which are judged best
-to represent the data at hand, and thence to carry out the sum and
+to represent the problem at hand, and thence to carry out the sum and
 multiple integrals. In practice, of course, such calculations may be
 time-consuming, and there will often be more than one choice for the
-functional form of each probability. In Section \ref{sec:example}, we
-shall therefore set out a specific example.
+functional form of each probability. And those choices affect the
+computational effort which may be critical if \refeq{slm} has to be
+repeatedly evaluatated for thousands of bins.
 
@@
 single-packet luminosity distributions can be utilised to find exact
-expressions for Eq.~(\ref{sll}).
-
-\subsection{Multiplicity prediction}
+expressions for Eq.~(\ref{sll}). Our choices are motivated by our
+original application \tardis{} but we emphasize that much of the
+analysis below, especially \refsec{multiplicity}, is generally
+applicable.
+
+\subsection{Multiplicity prediction} \label{sec:multiplicity}
 
@@
 
+\fred{Here and above, $n$ does not have to be the same in
+  $P(N \cond n)$ and $p(n,\bml\cond \alpha,\beta)$. We can use packets
+  from other bins if we assume $\alpha, \beta$ are the same in those bins.}
+
 The only missing piece to explicitly evaluate \refeq{spf} is the prior
@@
 The authors of \cite{moala2013bayesian} present an overview of
-possible choices and conclude that the point estimate given by the
-mode of the posterior is nearly insensitive to the prior choice
-starting from around $n = 30$. This requirement is nearly always
-satisfied and so the functional form of $p(\alpha,\beta)$ is not
-important. There is, however, some explicit prior information which we
-want to build into our analysis.
-The distribution $p(\ell \cond \alpha, \beta)$ should always vanish at
-the origin, requiring that $\alpha$ must be larger than 1.  For
-$\beta$, we likewise choose a uniform prior
-%
-\fred{This paragraph is from the old TARDIS section. In general, we
-  don't impose the constraint on $\alpha$. Only for TARDIS it seemed
-  to match and prevented the fitting from running away if very few
-  samples were present}
-%
+possible choices indicating that their \emph{modified MDIP} prior
+performed best in the few examples considered there. We do not follow
+their recommendation because their modification is not well motivated
+and it seems to accidentally perform better because of the particular
+examples considered. The other conclusion they have is that the
+Jeffreys, reference, and other noninformative priors with similarly
+solid motivation have comparable performance. They all exhibit the
+standard form for a scale or rate parameter $\propto 1/\beta$ and only
+differ in the $\alpha$ dependence. We disregard the reference prior as
+it forces us to identify either $\alpha$ or $\beta$ as \emph{the}
+nuisance parameter but we consider \emph{both} as nuisance
+parameters. Hence we select the Jeffreys prior
 \begin{align}
@@
   p(\alpha,\beta)
-  &= \frac{1}{(\alpha_{\rm max}{-}1)\beta_{\rm max}}
-  \qquad 1 < \alpha < \alpha_{\rm max},\; 0 < \beta < \beta_{\rm max}.
+  &=  C \frac{\sqrt{\alpha \Psi'(\alpha)-1}}{\beta} \, .
+  % \frac{1}{(\alpha_{\rm max}{-}1)\beta_{\rm max}}
+  % \qquad 1 < \alpha < \alpha_{\rm max},\; 0 < \beta < \beta_{\rm max}.
 \end{align}
 % since $\alpha_{\rm max}$ and $\beta_{\rm max}$ cancel in (\ref{spf}).
-With these choices, the posterior is found to be
-\begin{align}
-  \lleq{spi}
-  p(\alpha,\beta\cond n,\bml)
-  &= \frac{1}{C}\; \frac{\beta^{n\alpha} r^{\alpha-1}\,e^{-\beta q}}{[\Gamma(\alpha)]^n},\\
+With these choices, the posterior normalization is
+\begin{align}
   \lleq{spj}
-  C &= \int_1^{\alpha_{\rm max}} d\alpha \int_0^{\beta_{\rm max}} d\beta\;
+  p(n, \bml) &= C \int_0^{\alpha_{\rm max}} d\alpha \int_0^{\beta_{\rm max}} d\beta\;
   \frac{\beta^{n\alpha} r^{\alpha-1}\,e^{-\beta q}}{[\Gamma(\alpha)]^n},
 \end{align}
-and hence
+where the upper integration limits
+$(\alpha_{\rm max}, \beta_{\rm max})$ should formally extend to
+$\infty$ but in practice have to be chosen by trial and error to be
+large enough to contain the relevant contribution to the integral.
+The prior's constant of proportionality $C$ cancels in the final
+result and we may set $C\equiv 1$ to obtain
 \begin{align}
@@
   p(Q\cond n,\bml)
-  &= \frac{1}{C}
+  &= \frac{1}{p(n, \bml)}
     \sum_{N=0}^\infty \;
@@
     \GammaDist(Q\cond N\alpha, \beta)\;
-    \frac{\beta^{n\alpha} r^{\alpha-1}\,e^{-\beta q}}{[\Gamma(\alpha)]^n}.
-\end{align}
-
-\subsection{Numerical implementation} \label{sec:numerical}
-
-\fred{Editing will be done by Fred}
-
-In our numerical implementation, the constants
-$(\alpha_{\rm max}, \beta_{\rm max})$ are chosen by trial and error to
-be large enough to contain the relevant region of
-$p(\ell\cond \alpha,\beta)$.
-%
-To perform the integrals over $(\alpha,\beta)$ to find $C$ and the
-Gamma-weighted numerator in (\ref{spe}), we used the fact that the
-posterior $p(\alpha,\beta \cond n,\bml)$ is approximately Gaussian in
-both $\alpha$ and $\beta$ for large $n$, so we can apply Laplace's
-method for both. This requires the Hessian at the mode and the
-gradient helps in locating the mode. The relevant analytic expressions
-are summarized in \refsec{appendix}.
-
-Regarding the sum over $N$, we start at
-$N = \lfloor n-a+1 \rfloor = \arg \max_N p(N|n)$ %
-\fred{Actually we should say $p(N \cond n, a)$} %
-and continue to larger and smaller values of $N$ until the
-contribution to $p(Q \cond n,\bml)$ becomes negligible. In case the
-relevant contribution is from large $N$ and $n$, we interpret $N$ as a
-continuous variable, replace factorials by $\Gamma$ function in
-\refeq{spj}, and apply Laplace's method to integrate over $N$ also.
-
-\fred{Our implementation is available online on \href{https://github.com/tardis-sn/XXX}{github}}
-
-\subsection{Results} \label{sec:gammaresults}
-
-\fred{Graphics output, comments, conclusions by Fred}
-
-\subsection{Specifics of \tardis} \label{sec:tardis}
+    \frac{\beta^{n\alpha} r^{\alpha-1}\,e^{-\beta q}}{[\Gamma(\alpha)]^n} \frac{\sqrt{\alpha \Psi'(\alpha)-1}}{\beta}.
+\end{align}
+
+\fred{Check below if \refeq{spk} used with flat prior}
+
+\subsection{Preprocessing \tardis{} output } \label{sec:tardis}
 
@@
 
-A histogram of the luminosities from a \tardis{} run indicates a
-distribution strongly skewed with a mode close to the maximum
+A histogram plot of the luminosities from a \tardis{} run indicates a
+strongly skewed distribution with its single mode close to the maximum
 luminosity $\lmax$, while there is a long tail for small $\ell$. These
@@
 
+\subsection{Numerical implementation} \label{sec:numerical}
+
+\fred{Editing will be done by Fred}
+
+In our numerical implementation, the constants
+$(\alpha_{\rm max}, \beta_{\rm max})$ are chosen by trial and error to
+be large enough to contain the relevant region of
+$p(\ell\cond \alpha,\beta)$.
+%
+To perform the integrals over $(\alpha,\beta)$ to find $C$ and the
+Gamma-weighted numerator in (\ref{spe}), we used the fact that the
+posterior $p(\alpha,\beta \cond n,\bml)$ is approximately Gaussian in
+both $\alpha$ and $\beta$ for large $n$, so we can apply Laplace's
+method for both. This requires the Hessian at the mode and the
+gradient helps in locating the mode. The relevant analytic expressions
+are summarized in \refsec{appendix}.
+
+Regarding the sum over $N$, we start at
+$N = \lfloor n-a+1 \rfloor = \arg \max_N p(N|n)$ %
+\fred{Actually we should say $p(N \cond n, a)$} %
+and continue to larger and smaller values of $N$ until the
+contribution to $p(Q \cond n,\bml)$ becomes negligible. In case the
+relevant contribution is from large $N$ and $n$, we interpret $N$ as a
+continuous variable, replace factorials by $\Gamma$ function in
+\refeq{spj}, and apply Laplace's method to integrate over $N$ also.
+
+\fred{Our implementation is available online on \href{https://github.com/tardis-sn/XXX}{github}}
+
  \subsection{Poisson-gamma convolution}\label{sec:poisson-gamma}

### 2017-02-16 real/draft.tex
[draft] Polish "numerical implementation"
Remove several comments

@@
 
-\fred{Check below if \refeq{spk} used with flat prior}
-
 \subsection{Preprocessing \tardis{} output } \label{sec:tardis}
@@
 of $L$ by
-%%%%%%%%%%%%%%%%
-\fredmargin{citation} %
-%%%%%%%%%%%%%%%
 \begin{align}
@@
 \end{align}
-\hans{I have inserted a uniform prior for $\mu$ here; see comments
-  below. The original Gaussian prior has been retained below but I
-  would suggest that the compliations introduced by it are
-  unnecessary.}  %
 In keeping with prior information that luminosities are always
@@
 \end{align}
-with data sample moments $\langle\lum^k\rangle$ defined as $(1/n)
-\sum_{j=1}^n \lum_j^k$.
+with data sample moments $\langle\lum^k\rangle$ defined as
+$(1/n) \sum_{j=1}^n \lum_j^k$. Since we assume $n \gg 1$, we may
+choose $a_0 = b_0 = 0$ resulting in a noninformative prior on $\sigma^2$.
 %
@@
 \end{align}
-While the integral cannot be carried out analytically, the integrand
-is unimodal and strongly peaked, so that the Laplace or saddle-point
-approximation~\cite[Ch. 27]{mackay2003information} can be applied.
 
@@
 
-In \reffig{asymptotic}, we contrast the asymptotic expression with the
-``exact'' expression \refeq{spk} where the integrals over $\bmphi$ are
-done numerically to a requested relative accuracy of $10^{-5}$ through
-the Cubature
-package\cite{cubature,genz1980remarks,berntsen1991adaptive}. We
+In \reffig{asymptotic}, we compare the asymptotic expression
+\refeq{ral} with the ``exact'' expression \refeq{spk} where the
+integrals over $\bmphi$ are done numerically to a requested relative
+accuracy of $10^{-5}$ through the Cubature
+package\cite{cubature,genz1980remarks,berntsen1991adaptive}.
+
+For large $n$ where the asymptotic expression \refeq{ral} is valid,
+asymptotic normality justifies the use of the Laplace or saddle-point
+approximation~\cite[Ch. 27]{mackay2003information}. We implemented
+\fredmargin{\href{https://github.com/tardis-sn/XXX}{github}}
+the
+above formulae in the julia language~\cite{julia14} and used Newton's
+method from the \texttt{Optim.jl} package in combination with
+automatic differentiation~\cite{RevelsLubinPapamarkou2016} to optimize
+the respective integrand and to compute the Hessian at the mode as
+required for the Laplace method. As initial values for optimization,
+we slightly offset the sample estimates
+\begin{align}
+  \lleq{rat}
+  \qty(\lambda, \mu, \sigma^2) = 1.0005 (n, \langle\lum\rangle, \langle\lum^2\rangle{-}\langle\lum\rangle^2)
+\end{align}
+to avoid issues if the optimizer cannot improve on the initial guess
+and found convergence within $\order{10}$ steps.
+
+As a test case, we generated
+$\ell_i \sim \GammaDist(\alpha=1.5, \beta=60), i=1 \dots n$, and
 computed $P(\Lumtot \cond n, \bml)$ on a grid of $\Lumtot$ values and
@@
 based on the $\GammaDist$ model imply $\lambda \geq 0$ and $Q \geq 0$.
+%
+\fred{Update values after Jeffreys implementing prior}
 
@@
 agree even better than shown there and that the normalization of the
-asymptotic result approaches one.
+asymptotic result approaches one. Integrating the asymptotic
+expression with cubature instead of Laplace gave very similar results
+but turns out to be less reliable and requires many more iterations.
 
 The main advantage of the approximation is computational efficiency. A
-single evaluation of $P(Q \cond n, \bml)$ takes only about 100 calls
-to the integrand of \refeq{ral} regardless of the value of $n$. In
-contrast, a the numerical integration of a single term for fixed $N$
-in \refeq{spk} requires 1000s of function evaluations, and the number
-of terms grows like $\sqrt{n}$ for a given accuracy of
-$P(Q \cond n, \bml)$.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\hans{Some comments on the numerical values of hyperparameters $a$ in
-  Poisson, $a,b$ in inverse gamma needed.} %
-\hans{I would omit the text and calculations below. Apart from the
-  extra parameters, the main reason is that the Gaussian prior in
-  $\mu$ is not consistent with the fact that all luminosities are
-  positive, while the gaussian permits negative $\mu$. This may not be
-  important numerically, but the uniform prior in $\mu$ does
-  circumvent this problem and is moreover less informative than the
-  Gaussian prior. } %
-\hans{Note also that letting $\mu_0{=}0$ and $ n_0\to 0$ below implies
-  a prior gaussian of infinite width centered on zero, i.e.\ half of
-  the values of $\mu$ will be negative. Could this be the reason why
-  the numerics pick up a factor two?}%
-\fred{No, what I used before was equivalent to a uniform prior that
-  extends into the range $\mu<0$. The normalization cancels out and
-  the negative region doesn't contribute. With the uniform prior, you
-  assume that your upper cut-off on $\mu$ is large enough to contain
-  the posterior but at the same time large values don't actually
-  contribute so it doesn't matter if you include them or not. That
-  means I can't pick up a factor of two there, it would only change
-  the normalization but not the \emph{shape} of $p(Q)$} \fred{The
-  error is fixed and I remove the next sections as Hans suggested.}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%
+single evaluation of $P(Q \cond n, \bml)$ with the Laplace methods
+requires only about 100 calls to the integrand of \refeq{ral}
+regardless of the value of $n$. In contrast, the numerical integration
+of a single term for fixed $N$ in \refeq{spk} requires 1000s of
+function evaluations, and the number of terms grows like $\sqrt{n}$
+for a given accuracy of $P(Q \cond n, \bml)$.
 
@@
 \subsection{Moments of the luminosity prediction}\label{sec:predict-moments}
-
-\hans{H: If threefold integrals must be carried out for the
-  expectation values, why not do so for Eq (ral) directly rather than
-  just the expectation values?} %
-
 \hans{H: I would agree to remove this section; I do not see value added.}
@@
 \refeq{spj}, and apply Laplace's method to integrate over $N$ also.
-
-\fred{Our implementation is available online on \href{https://github.com/tardis-sn/XXX}{github}}
 

### 2017-02-16 real/code/GammaIntegrand.jl
[GammaIntegrand] Add Gamma Jeffreys prior

@@ -10,9 +10,14 @@ import ..Integrate
 
 using DiffBase, ForwardDiff, Optim
 
+Ψ′ = trigamma
+
+"P(α, β), (sph)"
+log_prior(α::Real, β::Real) = 1/2 * log(α*Ψ′(α)-1) - log(β)
+
 """ posterior p(α, β | n, q, r) for the parameters of the Gamma
-distribution given the sufficient statistics of the samples."""
-log_posterior(α::Real, β::Real, n::Real, q::Real, logr::Real, evidence=0.0) = n*(α*log(β)-lgamma(α)) + (α-1)*logr - β*q - evidence
+distribution given the sufficient statistics of the samples. (spg)"""
+log_posterior(α::Real, β::Real, n::Real, q::Real, logr::Real, evidence=0.0) = n*(α*log(β)-lgamma(α)) + (α-1)*logr - β*q - evidence # + log_prior(α, β)
 function make_log_posterior(n::Real, q::Real, logr::Real, evidence=0.0)
     function (θ::Vector)
         α, β = θ

### 2017-02-16 real/code/Moments.jl
[Moments] Fix variable name

@@ -39,7 +39,7 @@ Enumeration to specify which target function to create
 """
 Create a function for Laplace approximation
 """
-function targetfactory(target::TARGET, n::Int64, Qfirst::Float64, Qsecond::Float64, nb::Int64=nb;
+function targetfactory(target::TARGET, n::Int64, Qfirst::Float64, Qsecond::Float64, nb::Int64=n;
                        a::Float64=1/2, Q::Float64=-1.0)
     0 <= a <= 1 || error("need a in [0,1]")
     (Qfirst > 0) || error("nonpositive Qfirst $Qfirst")

### 2017-02-17 real/draft.tex
[draft/tardis] Start updating notation

@@
   paper, in my view, eg the definition of $X = 1 - L/\ell_{\rm max}$.}
-
-The \tardis{} packages simulates supernovae with %
-\fred{Wolfgang's favorite description of his package}. %
+\fred{I removed a lot of cruft and swapped the symbols $x \leftrightarrow \ell$ so it's in line with the previous section}
+
 Here we focus on the real packets as opposed to the virtual packets
-output by tardis.
-
-A histogram plot of the luminosities from a \tardis{} run indicates a
-strongly skewed distribution with its single mode close to the maximum
-luminosity $\lmax$, while there is a long tail for small $\ell$. These
-features can be accomodated by a $\GammaDist$ distribution if we
-transform the luminosities to
+and denote the raw output by \tardis{} for the $i$-{th} packet as $\qty(\nu_i^\prime, \ell_i^\prime)$.  A histogram plot \fredmargin{include plot} of the luminosities
+from a \tardis{} run indicates a strongly skewed distribution with its
+single mode close to the maximum luminosity $\lmax$, while there is a
+long tail for small $\ell$. These features can be accomodated by a
+$\GammaDist$ distribution if we transform the luminosities to
 \begin{align}

### 2017-02-17 real/draft.tex
[draft] Revamp tardis section

--- 2017-02-16 real/draft.tex
@@
 
-\subsection{Preprocessing \tardis{} output } \label{sec:tardis}
-
-\hans{H: This subsection needs much more work. I have just copied some
-  relevant text in here which may be useful. Much of it should be
-  removed. Some connection to TARDIS should preferably be kept in the
-  paper, in my view, eg the definition of $X = 1 - L/\ell_{\rm max}$.}
-
-The \tardis{} packages simulates supernovae with %
-\fred{Wolfgang's favorite description of his package}. %
-Here we focus on the real packets as opposed to the virtual packets
-output by tardis.
-
-A histogram plot of the luminosities from a \tardis{} run indicates a
-strongly skewed distribution with its single mode close to the maximum
-luminosity $\lmax$, while there is a long tail for small $\ell$. These
-features can be accomodated by a $\GammaDist$ distribution if we
-transform the luminosities to
+\subsection{Preprocessing simulation output } \label{sec:tardis}
+
+\fred{I removed a lot of cruft and swapped the symbols
+  $x \to \ell, \ell \to \ell^\prime$ so it's in line with the previous section}
+
+We consider \tardis{} as an example to demonstrate both how
+transforming the output of a simulation helps in predictions, and what
+further complications may arise in practice.
+
+\tardis{} can output two kinds of packets, real and virtual. Here we
+focus on the real packets and denote the raw output of \tardis{} by
+primed symbols, for example for the $i$-{th} packet
+$\qty(\nu_i^\prime, \ell_i^\prime)$.  A histogram plot
+\fredmargin{include plot} of the luminosities from a \tardis{} run
+indicates a strongly skewed distribution with its single mode close to
+the maximum luminosity $\lmax^\prime$, while there is a long tail for
+small $\ell$. These features can be accomodated by a $\GammaDist$
+distribution as in \refsec{single-packet} if we transform the
+luminosities to
 \begin{align}
   \lleq{gmf}
-  x_j \equiv 1- \frac{\ell_j}{\lmax} > 0
-\end{align}
-and similarly $X_j = 1 - (L_j/\lmax)$ and $\Lumtot \to X$. We assume
+  \ell_i \equiv 1- \frac{\ell_i^\prime}{\lmax^\prime} > 0
+\end{align}
+and similarly $L_i \equiv 1 - (L_i^\prime/\lmax^\prime)$. We assume
 the total number of samples across all bins is large, hence the effect
-of considering $\lmax$ a constant independent of the data is
-negligible. We checked that leaving $\lmax$ as a free parameter in a
-fit gave equivalent results.
+of considering $\lmax^\prime$ a constant independent of the data is
+negligible. We checked that leaving $\lmax^\prime$ as a free parameter
+in a fit gave equivalent results.
 %
@@
   \lleq{gme}
-  \lmax \equiv (1+\epsilon) \times \max_i \ell_i,
-\end{align}
-where $ \epsilon=10^{-6}$.
-
-We infer $p(X \cond n,\bml)$ and only in the end transform back to a
-probability density in $\Lumtot$ via
+  \lmax^\prime \equiv (1+\epsilon) \times \max_i \ell_i^\prime,
+\end{align}
+where $ \epsilon=10^{-6}$. We infer $p(\Lumtot \cond n,\bml)$ and only
+in the end transform back to a probability density in the raw
+coordinates via
 \begin{align}
   \lleq{gmg}
-  p(\Lumtot  \cond n, \bml)
-  &= p(X  \cond n, \bmx) \frac{1}{\lmax} \,.
-\end{align}
-To simplify the numerics, we also rescale the frequencies by a
-constant from $\mathcal{O}(10^{15})$ to $\mathcal{O}(1)$.  %
-
-A major numerical simplification arises because of the $\Lumtot \to X$
-transformation. Rewriting the \refeq{sll} formally as
+  p(\Lumtot^\prime  \cond n, \bml^\prime)
+  &= p(\Lumtot  \cond n, \bml) \frac{1}{\lmax^\prime} \,.
+\end{align}
+We also rescale the frequencies by a constant from
+$\mathcal{O}(10^{15})$ to $\mathcal{O}(1)$ but the most important
+numerical simplification arises because of the transformation of the
+luminosities. Rewriting \refeq{sll} formally as
 \begin{align}
@@
 \end{align}
-we see that the right-hand side is a mixture density. In our
-application, the distribution of a single-packet luminosity is narrow
-in the sense that the standard deviation relative to the mean is at
-the percent level. This implies that $p(Q \cond N, n, \bml)$ has one
-mode from each $N$ and all modes are clearly separated. The
-transformation $\Lumtot \to X$ effects a strong overlap between the
-distributions $p(X \cond N, n, \bmx)$ with similar $N$, so
-$p(X \cond n, \bmx)$ can be well approximated by a unimodal
-distribution.
-
-We chose the Gamma distribution because it is \emph{stable}; i.e.,
-the sum of $N$ Gamma variates is again a Gamma variate
-\begin{align}
-  \lleq{gmb}
-    X_j \sim \GammaDist(\alpha, \beta) \Rightarrow X \equiv \sum_{j=1}^N X_j \sim \GammaDist(N \alpha, \beta).
-\end{align}
-Assuming $\alpha = \alpha(\nu \cond \bmphi)$ and $\beta = \beta(\nu
-\cond \bmphi)$ are functions of $\nu$ that depend on parameters
-$\bmphi$, we set
-\begin{align}
-  \lleq{gmib}
-  p(\bmX \cond \bmphi, N, \nu) &= \prod_{j=1}^N \GammaDist(X_j \cond \alpha, \beta ),\\
-  p(\bmx \cond \bmphi, \npack, \bmnu) &= \prod_{i=1}^{\npack} \GammaDist(x_i \cond \alpha_i, \beta_i).
-\end{align}
-where $\alpha_i \equiv \alpha(\nu_i \cond \bmphi), \beta_i \equiv
-\beta(\nu_i \cond \bmphi)$.
-
-\subsection{Linear model}
-
-\fred{Parts may still be relevant, especially if we put a plot of an
-  entire spectrum in the paper.}
-
-For simple functions $\alpha(\nu \cond \bmphi)$, one can find the
-minimum value of $\alpha$ over the entire range of $\nu$
-analytically. If this is too complicated, we propose to require
-$\alpha(\nu_i \cond \bmphi) > 1$ only at each packet frequency
-$\nu_i$. In the evaluation of $p(\bmx \cond \bmphi, \npack, \bmnu)$,
-one has to loop over all packets and compute $\alpha$ and $\beta$
-anyways, so the constraint imposes essentially no computational
-overhead.
-
-In the absence of a physical model to determine
-$\alpha(\nu \cond \bmphi)$ and $\beta(\nu \cond \bmphi)$, we propose
-to start with a simple polynomial ansatz
-$\alpha(\nu \cond \bmphi) = \sum_{k=0}^\Kalpha \alpha_k \nu^k,
-\beta(\nu \cond \bmphi) = \sum_{k=0}^\Kbeta \beta_k \nu^k$ such that
-\begin{align}
-  \lleq{gmp}
-  \bmphi = (\alpha_0, \dots \alpha_\Kalpha , \beta_0, \dots \beta_\Kbeta)
-\end{align}
-The usual trade off between parsimony and the ability to accurately
-model the frequency dependence determines the orders of the
-polynomials, $\Kalpha$ and $\Kbeta$. This is outside the scope of this
-paper. Here we choose $\Kalpha = 3$ and $\Kbeta = 1$.
+we see that the right-hand side is a mixture density. In the \tardis{}
+application, the distribution of a single-packet luminosity,
+$P\qty(\ell_i \cond \bmphi)$ is narrow in the sense that the standard
+deviation relative to the mean is at the percent level. This implies
+that each $p(Q^\prime \cond N, n, \bml^\prime)$ contributes one narrow
+peak and all peaks are clearly separated in the superposition
+$ p(Q^\prime \cond n,\bml^\prime)$. The transformation
+$\ell^\prime \to \ell$ effects a strong overlap between the
+distributions $p(\Lumtot \cond N, n, \bml)$ with similar $N$, so
+$p(\Lumtot \cond n, \bml)$ can be well approximated by a unimodal
+distribution and thus $\Lumtot$ instead of $\Lumtot^\prime$ the
+preferred coordinate basis.
+%
+\fredmargin{show hedgehog plot}
+
+\subsection{Towards an unbinned analysis} \label{sec:unbinned}
+
+The parameters $\qty(\alpha, \beta)$ of the single-packet distribution
+$\GammaDist(\ell_i \cond \alpha, \beta)$ may, of course, depend on the
+frequency. % $\nu_i$
+This effect could be captured, for example, by a (linear) model such
+$\alpha = \alpha(\nu \cond \bmphi)$ and
+$\beta = \beta(\nu \cond \bmphi)$ are functions of $\nu$ that depend
+on parameters $\bmphi$ such that integrals over $\alpha, \beta$ would
+turn into integrals over $\bmphi$. But for the sake of clarity, let us
+consider only the simplest case where $\alpha, \beta$ are the same in
+every bin such that \refeq{spk} becomes
+\begin{align}
+  \lleq{gmi}
+  p(Q\cond \npack,\bmy)
+  &=     \sum_{N=0}^\infty \;
+    p(N\cond n)\;
+    \int \rmdx{\alpha}\,\rmdx{\beta}\;
+    \GammaDist(Q\cond N\alpha, \beta)\;
+    p(\alpha, \beta \cond \npack, \bmy) \, .
+\end{align}
+In words, the posterior $ p(\alpha, \beta \cond \npack, \bmy)$ uses
+all packets to essentially fix $\alpha, \beta$ if $\npack \gg 1$ and
+allows meaningful predictions of $\Lumtot$ even in bins where
+\emph{no} packet is observed; i.e., $n=0$. The only binning left is
+performed to count $n$ the number of packets that fall into a given
+frequency bin. This strategy is applicable, for example, when
+absorption or emission lines in the spectrum abruptly alter the number
+of packets for nearby frequencies but the luminosities are unaffected
+by those line features. Note that if the dependence on $\bmphi$ can be
+modeled with few parameters, the total number of parameters may be
+drastically reduced: instead of having $\alpha, \beta$ separately for
+each of a few thousand bins, there are potentially only a handful of
+elements in $\bmphi$ for the entire spectrum. In that case, the
+integral in \refeq{gmi} does not depend on the bin but only on $N$ so
+its value may be cached to speed up the simultaneous evaluation across
+all bins.
 
@@
 \subsection{Moments of the luminosity prediction}\label{sec:predict-moments}
-\hans{H: I would agree to remove this section; I do not see value added.}
+\fred{Shouldn't we use $n, \bml$ instead of $\bmy$ to be consistent?
+  $\bmy$ would mean we use all samples from all bins}
 
@@
   \expect{\Lumtot} &= \int \rmdx{\lambda} \rmdx{\mu} \rmdx{\sigma^2} p(\lambda \cond n) p( \mu, \sigma^2 \cond \bmy) \qty(-\eval{\dv{f(t \cond  \lambda, \mu, \sigma^2)}{t}}_{t=0})\\
-  \variance{\Lumtot} &= \int \rmdx{\lambda} \rmdx{\mu} \rmdx{\sigma^2} p(\lambda \cond n) p( \mu, \sigma^2 \cond \bmy) \qty(\eval{\dv[2]{f(t \cond  \lambda, \mu, \sigma^2)}{t}}_{t=0}) - \expect{\Lumtot}^2
-\end{align}
-We implemented the above formulae in the julia language~\cite{julia14}
-and used Newton's method from the \texttt{Optim.jl} package in
-combination with automatic
-differentiation~\cite{RevelsLubinPapamarkou2016} to optimize the
-respective integrand and compute the Hessian at the mode. As initial
-values, we slightly offset the sample estimates to avoid issues if the
-optimizer cannot improve on the initial guess and found convergence
-within $\order{10}$ steps:
-\begin{align}
-  \lleq{rat}
-  \qty(\lambda, \mu, \sigma^2) = 1.0005 (n, \langle\lum\rangle, \langle\lum^2\rangle{-}\langle\lum\rangle^2).
-\end{align}
-\fred{At least for us, I should do a comparison plot to evaluate $p(\Lumtot \cond \bmy)$ with both methods } %
+  \variance{\Lumtot} &= \int \rmdx{\lambda} \rmdx{\mu} \rmdx{\sigma^2} p(\lambda \cond n) p( \mu, \sigma^2 \cond \bmy) \qty{\eval{\dv[2]{f(t \cond  \lambda, \mu, \sigma^2)}{t}}_{t=0} - \qty(\eval{\dv{f(t \cond  \lambda, \mu, \sigma^2)}{t}}_{t=0})^2} \, .
+\end{align}
 \fred{We need a plot, the one to compare predictions and 250 real replicas or 100 virtual replicas?} %

### 2017-02-17 real/code/GammaIntegrand.jl
[code] Add log_poisson

@@ -1,7 +1,7 @@
 """reload("TardisPaper"); TardisPaper.GammaIntegrand.test()"""
 module GammaIntegrand
 
-export heuristicN, log_gamma, log_gamma_predict, log_inv_gamma, log_normal, log_poisson_predict, log_posterior
+export heuristicN, log_gamma, log_gamma_predict, log_inv_gamma, log_normal, log_poisson, log_poisson_predict, log_posterior
 export make_asymptotic, make_log_posterior
 export optimize_log_posterior, optimize_log_posterior_predict, optimize_integrand_λμσ²
 export triple_mode, triple_ranges
@@ -33,6 +33,9 @@ log_gamma(x::Real, α::Real, β::Real) = α*log(β) - lgamma(α) + (α-1)*log(x)
 
 """ p(Q | α, β, N) """
 log_gamma_predict(Q::Real, α::Real, β::Real, N::Real) = log_gamma(Q, N*α, β)
+
+"Poisson(N | λ)"
+log_poisson(N, λ) = N*log(λ) - λ - lgamma(N+1)
 
 """(pmh) p(N | n)"""
 function log_poisson_predict(N::Real, n::Integer, a::Real)

### 2017-02-17 real/code/TardisPaperTest.jl

@@ -114,6 +114,9 @@ function gamma_integrand()
     @test log_gamma(x, α, β) ≈ logpdf(dist, x)
     @test log_gamma_predict(x, α, β, n) ≈ logpdf(Gamma(n*α, 1/β), x)
 
+    λ = 3.3
+    @test log_poisson(N, λ) ≈ logpdf(Poisson(λ), N)
+
     @test log_posterior(α, β, n, q, logr) ≈ 1407.87731905469
     # TODO was type stable before, why not anymore?
     # @inferred log_posterior(α, β, n, q, logr)

### 2017-02-17 real/code/TardisPaperTest.jl
[predict] Add by_sum with lambda fixed

@@ -70,13 +70,17 @@ function predict_alpha_fixed()
     a = 1/2
     Q = n*α/β
     ε = 1e-3
+    λ = n
 
     res, Ndown, Nup = Predict.by_sum(Q, α, β, a, n; ε=ε)
-    target = 0.0
-    for N in Ndown:Nup
-        target += exp(log_poisson_predict(N, n, a) + log_gamma_predict(Q, α, β, N))
-    end
+    res2, Ndown2, Nup2 = Predict.by_sum(Q, α, β, λ; ε=ε)
+
+    # variance should be reduced if λ fixed
+    @test (Nup2 - Ndown2) <= (Nup - Ndown)
+
+    target = mapreduce(N -> exp(log_poisson_predict(N, n, a) + log_gamma_predict(Q, α, β, N)), +, 0.0, max(1, Ndown):Nup)
     @test res == target
+    @test res2 == mapreduce(N -> exp(log_poisson(N, λ) + log_gamma_predict(Q, α, β, N)), +, 0.0, Ndown2:Nup2)
 
     # adding more terms shouldn't change much
     for N in Nup+1:Nup+10

### 2017-02-17 real/code/Predict.jl
[Predict] increase default cubature precision
fix corner case n=0 in initialization

@@ -81,7 +81,7 @@ end
 Predict Q by summing over N and considering λ, α, β fixed.
 """
 function by_sum(Q::Real, α::Real, β::Real, λ::Real; Ninit::Real=0, ε::Real=1e-3)
-    Ninit = initialize_N(Q, α, β, 0.5, λ, Ninit)
+    Ninit = initialize_N(Q, α, β, 0.5, convert(Int64, ceil(λ)), Ninit)
     f = N -> exp(log_poisson(N, λ) + log_gamma_predict(Q, α, β, N))
     iterate(Ninit, f, ε)
 end
@@ -128,7 +128,7 @@ Arguments
 """
 
 function by_cubature(Q, a, n, q, logr, nb=n; αmin=1e-2, αmax=5, βmin=1e-5, βmax=100,
-                     Ninit=0, ε=1e-3, reltol=1e-3)
+                     Ninit=0, ε=1e-3, reltol=1e-5)
     # to avoid overflow in Cubature, evaluate target at mode and subtract it. The
     # value within a few order of magnitude of
     fit_dist = Distributions.fit_mle(Distributions.Gamma, Distributions.GammaStats(q, logr, n))
@@ -192,7 +192,7 @@ function asymptotic_by_laplace(Q, a, n, first, second, nb=n; a₀=0, b₀=0)
 end
 
 function asymptotic_by_cubature(Q, a, n, first, second, nb=n;
-                                reltol=1e-3, a₀=0, b₀=0,
+                                reltol=1e-5, a₀=0, b₀=0,
                                 λmin=0.0, λmax=0.0,
                                 μmin=0.0, μmax=0.0,
                                 σ²min=0.0, σ²max=0.0)

### 2017-02-17 real/draft.tex
[draft] Add preliminary spectra w/ uncertainties
[draft, plots] Add prediction plots for nb=0,2

@@
   \item Statistics of replicas: again compare \refeq{spe} to \refeq{ral}. Teach a lesson: from a single data set, prediction will be centered there
-  \item Question: does Wolfgang want to see a plot in which ``$\lambda$ is divided out''? That is , just plot how mean
 \end{enumerate}
@@
 \end{enumerate}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=0.48\textwidth]{spectrum_bin}
+  \includegraphics[width=0.48\textwidth]{spectrum_all}
+  \caption{Spectrum with smallest 2 $\sigma$ credibility interval. Left: only packets from bin contribute. Right: a}
+\label{fig:spectrum}
+\end{figure}
 
@@
 
+Let's consider the extreme case $\npack \gg 1$ so we can neglect the
+uncertainty on $\alpha, \beta$ and
+$p(\alpha, \beta \cond \npack, \bmy) =
+\delta(\alpha-\alpha_0)\delta(\beta - \beta_0)$ such that
+\begin{align}
+  \lleq{gmj}
+  p(Q\cond \npack,\bmy)
+  &=     \sum_{N=0}^\infty p(N\cond n) \GammaDist(Q\cond N\alpha_0, \beta_0)\\
+  &=     \sum_{N=0}^\infty p(N\cond n) p(Q\cond N, \alpha_0, \beta_0) \, .
+\end{align}
+The case $N=0$ deserves special attention. To preserve the normalization of $p(Q\cond \npack,\bmy)$, we need to define
+\begin{align}
+  \lleq{gmk}
+   p(Q\cond N, \alpha, \beta) &\equiv
+   \begin{cases}
+     \delta(Q),     &N=0;\\
+     \GammaDist(Q\cond N\alpha, \beta), & N > 0 \, .
+   \end{cases}
+ \end{align}
+ %
+ \fred{Notation is sloppy before, it has nothing to do with unbinned
+   stuff or $\npack \gg 1$. This is just where it became apparent to
+   me.}
+ %
+ Intuitively, we can only observe $Q=0$ for $N=0$ packets bu
+ mathematically, $\GammaDist(Q\cond \alpha=0, \beta) = 0$ which,
+ without the extra contribution from the Dirac $\delta$ function,
+ would imply
+ $p(Q\cond \npack,\bmy) = \sum_{N>0}^\infty p(N \cond n) < 1$. Higher
+ moments of $p(Q \cond N, \alpha, \beta)$ are unaffected by this
+ subtlety.
+
+ In \reffig{comp-unc}, we illustrate the impact of the uncertainty due
+ to $\lambda$ or $\alpha, \beta$ for $n=2$ packets and continue to use
+ the true values $\alpha_0 = 1.5, \beta_0 = 60$. The standard error
+ increases from 0.042 ($\lambda, \alpha, \beta$ fixed) over 0.059
+ ($\alpha, \beta$ fixed) to 0.11 (nothing fixed). The case where all
+ parameters are known is usually not attainable in practice and only
+ shown for comparison. But the case $\alpha, \beta$ fixed can be
+ reached easily if packets from other bins are used. Note that $n=2$
+ is the minimum number of packets to perform the binned analysis.
+
+ \begin{figure}[ht]
+  \centering
+  \includegraphics[width=0.44\textwidth]{comp_unc_0}
+  \includegraphics[width=0.44\textwidth]{comp_unc_2}
+  \caption{Prediction considering $\lambda, \alpha, \beta$ fixed (green), $\alpha, \beta$ fixed (red), or nothing fixed (blue).}
+\label{fig:comp-unc}
+\end{figure}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
   Eq.~(\ref{rab}) due to $\kappa_3$ and higher orders is unclear.}
+\fred{These guys \href{http://stats.stackexchange.com/questions/83069/what-is-the-maximum-entropy-probability-density-function-for-a-positive-continuo}{claim} that our truncated Gauss is the max-ent distribution for our prior information.
+I'm not sure they are right, they are not convinced themselves.
+We have to check.}
 %%%%%%%%%%%
@@
     p(\sigma^2\cond n)}
-  {p(\bml\cond n)}
+  {p(\bml\cond n)} \, .
 \end{align}
@@
 $p(\mu\cond\mu_{\rm max}) = 1/\mu_{\rm max}$ independent of $n$; as
-always, the posterior is independent of
-$\mu_{\rm max}$.
-For $\sigma^2$, we use an $n$-independent Inverse Gamma
-distribution prior with hyperparameters $(a_0,b_0)$ which reads
+always, the posterior is independent of $\mu_{\rm max}$.  For
+$\sigma^2$, we use an $n$-independent Inverse Gamma prior with
+hyperparameters $(a_0,b_0)$ which reads
 \begin{align}

### 2017-02-17 real/code/TardisPaperPlots.jl

@@ -31,14 +31,32 @@ function normalize!(Qs, res, tag=""; δcontrib=0.0)
     norm, mean, stderr
 end
 
+type GammaStatistics{T<:Real}
+    q::T
+    logr::T
+    first::T
+    second::T
+    n::Int64
+
+    function GammaStatistics(samples::Vector)
+        q = sum(samples)
+        logr = sum(log(samples))
+        n = length(samples)
+        first = q/n
+        second = mapreduce(x->x^2, +, samples)/n
+        new(q, logr, first, second, n)
+    end
+end
+
 function compute_prediction(;n=400, Qs=false, Qmin=1e-3, Qmax=2, nQ=50, α=1.5, β=60.0, a=1/2, ε=1e-3, reltol=1e-3, seed=16142)
     srand(seed)
 
     info("Computing for n=$n")
     dist = Gamma(α, 1/β)
     samples = rand(dist, n)
-    q = sum(samples)
-    logr = sum(log(samples))
+    gstats = GammaStatistics(samples)
+    # q = sum(samples)
+    # logr = sum(log(samples))
 
     if Qs === false
         meanQ = n*α/β
@@ -52,18 +70,18 @@ function compute_prediction(;n=400, Qs=false, Qmin=1e-3, Qmax=2, nQ=50, α=1.5, 
     res_mle = map(Q->Predict.by_sum(Q, shape(dist_fit), 1/scale(dist_fit), a, n;ε=ε)[1], Qs)
     normalize!(Qs, res_mle, "max. likelihood")
 
-    first = q/n
-    second = mapreduce(x->x^2, +, samples)/n
-    res_asym_laplace = map(Q->Predict.asymptotic_by_laplace(Q, a, n, first, second), Qs)
+    # first = q/n
+    # second = mapreduce(x->x^2, +, samples)/n
+    res_asym_laplace = map(Q->Predict.asymptotic_by_laplace(Q, a, n, gstats.first, gstats.second), Qs)
     normalize!(Qs, res_asym_laplace, "asympt. Laplace")
 
-    res_laplace = map(Q->Predict.by_laplace(Q, a, n, q, logr; ε=ε)[1], Qs)
+    res_laplace = map(Q->Predict.by_laplace(Q, a, n, gstats.q, gstats.logr; ε=ε)[1], Qs)
[…]

### 2017-02-21 real/draft.tex
[draft] minor edits, subsection removed

@@
 %% CHANGE THIS DATE AS YOU MODIFY THE FILE
-\newcommand{\vdate}{170212}
+\newcommand{\vdate}{170221}
 \pagestyle{myheadings}
@@
 %
-Prediction of any ``future'' data $Y$ is made not by considering the
+Prediction of any ``future'' data $Y$ given the present data $\bmy$ is made not by considering the
 likelihood $p(Y\cond\bmtheta_{\rm max})$ using single
@@
 \end{align}
-The above relations are true for any type of data and can be applied
-to actual experimental data and the pertinent physical parameters. The
-scope of this paper, however, is limited to the relationship between
-simulated data generated by radiative transfer simulation codes.
-Likewise, all parameters considered below relate strictly to model
-descriptions of the simulation.
-\fred{I find this paragraph confusing. Which relationship?}
+While in this paper, we apply the above relations to simulated data
+only, they are valid also for actual experimental data and the
+pertinent physical parameters.
 
@@
 and luminosities form the raw simulation data
-$\bmy = \{ (\nu_i,\lum_i): i = 1,\ldots,\npack\}$.  We divide up the
-frequency spectrum into bins of width $\Delta\nu$ centered on
-frequencies $\nu_b$ resulting in bin intervals
+$\bmy = \{ (\nu_i,\lum_i): i = 1,\ldots,\npack\}$. Typically, the
+frequency spectrum is partitioned into bins of width $\Delta\nu$
+centered on frequencies $\nu_b$ resulting in bin intervals
 $\Delta_b = [\nu_b {-} \tfrac{1}{2} \Delta\nu \,,\, \nu_b {+}
-\tfrac{1}{2} \Delta\nu], b = 1,2,\ldots, B$. \fred{The binning may be dictated by experimental data} Let $n_b$ be the number
+\tfrac{1}{2} \Delta\nu], b = 1,2,\ldots, B$.  Let $n_b$ be the number
 of packets having a frequency falling into a particular bin $b$, and
@@
 separate Poisson distribution $p(N\cond\lambda)$ for each bin, where
-the magnitude of $\lambda$ will capture most of the variability of the
-spectrum. Given $N$, the single-packet luminosity distribution is
-modelled in the second step by means of a packet likelihood for
-$\Lum_{b,j}$ with packet parameters, a packet prior and, of course,
-packet parameter posterior. The single-packet distribution contains
-the crucial information which allows us to extend previous methods to
-cases where the number of packets in a bin is small.
+the change in magnitude of $\lambda$ from bin to bin will capture most
+of the variability of the spectrum. Given $N$, the single-packet
+luminosity distribution is modelled in the second step by means of a
+packet likelihood for $\Lum_{b,j}$ with packet parameters, a packet
+prior and, of course, a packet parameter posterior. The single-packet
+distribution contains the crucial information which allows us to
+extend previous methods to cases where the number of packets in a bin
+is small.
 %%%%%%%%%%%%%%%%%%%
@@
 whose likelihood is modelled by a distribution determined by
-likelihood parameters $\bmphi$,
+generic likelihood parameters $\bmphi$,
 \begin{align}
@@
 time-consuming, and there will often be more than one choice for the
-functional form of each probability. And those choices affect the
-computational effort which may be critical if \refeq{slm} has to be
-repeatedly evaluatated for thousands of bins.
+functional form of each probability which may critically affect the
+computational effort if \refeq{sll} has to be repeatedly evaluated for
+thousands of bins.
 
@@
 
-First, the parameter $\lambda$ is eliminated as follows. The number of
-packets $N$ is Poisson-distributed,
+First, we eliminate the parameter $\lambda$ of the Poisson
+multiplicity distribution for the number of packets $N$,
 \begin{align}
@@
 \end{align}
-The observed $n$ is naturally presumed to have arisen from the same
-Poisson distribution with the same $\lambda$.  The posterior for
-$\lambda$ is found from the data $n$ via Bayes' Theorem,
+The simulation number of packets $n$ is naturally presumed to have
+arisen from the same Poisson distribution with the same $\lambda$.
+The posterior for $\lambda$ is found from the data $n$ via Bayes'
+Theorem,
 \begin{align}
@@
 \end{align}
-We use an improper prior
+We use a prior
 \begin{align}
   \lleq{pmf}
-  p(\lambda) = c \lambda^{-a},
-\end{align}
-meaning that $\int_0^\infty \rmdx{\lambda}\,p(\lambda)$ is known only up to
-a constant $c$. An improper prior is not
-a problem because $c$ cancels out when calculating the posterior,
-provided that $n \ge 1$; i.e., at least one packet is observed in the
-bin.\footnote{For a uniform prior, the case $a{=}0$, we require of
-  course a parameter $\lambda_{\rm max}$. As long as
-  $\lambda_{\rm max}$ is large enough, it would also cancel.}, The
-posterior is
+  p(\lambda) = c \lambda^{-a}
+\end{align}
+which is ``improper'', meaning that
+$\int_0^\infty \rmdx{\lambda}\,p(\lambda)$ is known only up to a
+constant $c$. An improper prior is not a problem because $c$ cancels
+out when calculating the posterior, provided that $n \ge 1$; i.e., at
+least one packet is observed in the bin.\footnote{For a uniform prior,
+  the case $a{=}0$, we require of course a parameter
+  $\lambda_{\rm max}$. As long as $\lambda_{\rm max}$ is large enough,
+  it would also cancel.}, The posterior is
 \begin{align}
@@
 we accept that repeating the simulation will never yield a packet in
-the current bin. Hence we are led to Jeffreys prior, $a=1/2$.  With
-\refeq{pmh}, \refeq{sll} simplifies to
+the current bin. Hence we are led to Jeffreys prior, $a=1/2$.
+
+With \refeq{pmh}, \refeq{sll} simplifies to
 \begin{align}
@@
 Next, we consider detailed modelling of $p(\bmL\cond N,\bmphi)$ which
-was written in Eq.~(\ref{slk}) as a product single-packet
+was written in Eq.~(\ref{slk}) as a product of single-packet
 distributions $p(L_j\cond \bmphi)$, meaning that single packets are
@@
 independent of $N$.
-
+% 
 Which specific probability function to choose for $p(L_j\cond\bmphi)$
 depends, of course, on the specific simulation data. In this section,
-we choose it to be the same Gamma distribution for all $L_j$,
+we choose it to be the same Gamma distribution for all $L_j$
 with parameters $\bmphi = (\alpha,\beta)$,
@@
   \lleq{spc}
-  p(L_j\cond \bmphi) &\sim \GammaDist(L_j\cond \alpha,\beta)
+  p(L_j\cond \bmphi) &= \GammaDist(L_j\cond \alpha,\beta)
 \end{align}
 which has the nice property that the convolution (\ref{slm}) results
-in yet another Gamma distribution,
+in another Gamma distribution,
 \begin{align}
@@
   from other bins if we assume $\alpha, \beta$ are the same in those bins.}
-
-The only missing piece to explicitly evaluate \refeq{spf} is the prior
-$p(\alpha,\beta)$.
-% To gain some insight, consider the standard $\GammaDist$ model with
-% $\Kalpha = \Kbeta = 0$.
-% There is a conjugate prior but
-% unfortunately the normalization constant cannot be evaluated in closed
-% form, so it is thus of no use for us.
-% How about a noninformative prior?
+\hans{Please explain more fully; not clear.}
+
+It only remains to explicitly evaluate the prior $p(\alpha,\beta)$
+appearing in \refeq{spf}.
+%
 The authors of \cite{moala2013bayesian} present an overview of
@@
 parameters. Hence we select the Jeffreys prior
+\hans{What is $\Psi$?}
 \begin{align}
@@
 % since $\alpha_{\rm max}$ and $\beta_{\rm max}$ cancel in (\ref{spf}).
-With these choices, the posterior normalization is
+With these choices, the posterior normalization is %
+\hans{``Posterior normalization'' = Evidence? With improper priors,
+  should the upper limits not be $\infty$ below?}
 \begin{align}
@@
 
-We consider \tardis{} as an example to demonstrate both how
+\hans{This and the next section interrupt the actual calculation of
+  exact results in Sections
+  \ref{sec:multiplicity}--\ref{sec:single-packet} and the asymptotic
+  ones in Section \ref{sec:asymptotic}, so I would suggest moving them
+  to the end just before the conclusions. }
+
+In this section, we use \tardis{} as an example to demonstrate both how
 transforming the output of a simulation helps in predictions, and what
@@
 the maximum luminosity $\lmax^\prime$, while there is a long tail for
-small $\ell$. These features can be accomodated by a $\GammaDist$
+small $\ell$. These features can be accommodated by a $\GammaDist$
 distribution as in \refsec{single-packet} if we transform the
@@
 and similarly $L_i \equiv 1 - (L_i^\prime/\lmax^\prime)$. We assume
-the total number of samples across all bins is large, hence the effect
+the total number of samples across all bins is large, so that the effect
 of considering $\lmax^\prime$ a constant independent of the data is
@@
 \end{align}
-where $ \epsilon=10^{-6}$. We infer $p(\Lumtot \cond n,\bml)$ and only
-in the end transform back to a probability density in the raw
-coordinates via
+where $ \epsilon=10^{-6}$. We carry out all calculations with
+$p(\Lumtot \cond n,\bml)$ and only in the end transform back to a
+probability density in the raw coordinates via
 \begin{align}
@@
 \end{align}
-We also rescale the frequencies by a constant from
-$\mathcal{O}(10^{15})$ to $\mathcal{O}(1)$ but the most important
+While we also rescale the frequencies by a constant from
+$\mathcal{O}(10^{15})$ to $\mathcal{O}(1)$, the most important
 numerical simplification arises because of the transformation of the
@@
 application, the distribution of a single-packet luminosity,
-$P\qty(\ell_i \cond \bmphi)$ is narrow in the sense that the standard
+$p\qty(\ell_i' \cond \bmphi)$ is narrow in the sense that the standard
 deviation relative to the mean is at the percent level. This implies
@@
 $p(\Lumtot \cond n, \bml)$ can be well approximated by a unimodal
-distribution and thus $\Lumtot$ instead of $\Lumtot^\prime$ the
+distribution and thus $\Lumtot$ rather than $\Lumtot^\prime$ is the
 preferred coordinate basis.
@@
 frequency. % $\nu_i$
-This effect could be captured, for example, by a (linear) model such
-$\alpha = \alpha(\nu \cond \bmphi)$ and
-$\beta = \beta(\nu \cond \bmphi)$ are functions of $\nu$ that depend
-on parameters $\bmphi$ such that integrals over $\alpha, \beta$ would
-turn into integrals over $\bmphi$. But for the sake of clarity, let us
-consider only the simplest case where $\alpha, \beta$ are the same in
-every bin such that \refeq{spk} becomes
+This effect could be modelled, for example, by a functional
+parametrization $\alpha = \alpha(\nu \cond \bmphi)$ and
+$\beta = \beta(\nu \cond \bmphi)$ in dependence on the frequency $\nu$
+and on hyperparameters $\bmphi$ such that integrals over
+$\alpha, \beta$ would turn into integrals over $\bmphi$. For the
+sake of clarity, let rather us consider only the simplest case where
+$\alpha, \beta$ are the same in every bin such that \refeq{spk}
+becomes
 \begin{align}
@@
 modeled with few parameters, the total number of parameters may be
-drastically reduced: instead of having $\alpha, \beta$ separately for
+drastically reduced: instead of determining $\alpha, \beta$ separately for
 each of a few thousand bins, there are potentially only a handful of
@@
 \end{align}
-The case $N=0$ deserves special attention. To preserve the normalization of $p(Q\cond \npack,\bmy)$, we need to define
+The case $N=0$ deserves special attention. To preserve the
+normalization of $p(Q\cond \npack,\bmy)$, we need to define
 \begin{align}
@@
  %
+\hans{I'm ok with this now.}
  \fred{Notation is sloppy before, it has nothing to do with unbinned
@@
  %
- Intuitively, we can only observe $Q=0$ for $N=0$ packets bu
+ Intuitively, we can only observe $Q=0$ for $N=0$ packets but
  mathematically, $\GammaDist(Q\cond \alpha=0, \beta) = 0$ which,
@@
 theorem. As a bonus, the numerical effort to evaluate the basic
-\refeq{sll} drastically simplifies.
+\refeq{sll} simplifies drastically.
 
@@
 
- \subsection{Poisson-gamma convolution}\label{sec:poisson-gamma}
-
-\hans{H: This subsection is temporary. It is motivated by the question
-  whether the use of a gaussian for the single-packet luminosity
-  $p(L_j\cond\bmphi)$ in (\ref{rae}) is necessary and whether making
-  it a gaussian is a source of bias since $L_j$ cannot be
-  negative. The original choice of a Gamma distribution as in
-  (\ref{spc}) is more appropriate and could eliminate one source of
-  systematic bias. Below, I have derived a closed form for
-  $p(Q\cond \lambda,\bmphi)$ using the gamma-distributed
-  $p(L_j\cond \alpha,\beta)$ of (\ref{spc}). The final result
-  (\ref{qfp}) below could be used to replace the gaussian result
-  (\ref{rab}) for insertion into (\ref{asd}). Inclusion of the entire
-  derivation below is not envisaged; rather, we would include only
-  salient issues and equation in the text itself. For the moment, the
-  question is what Fred has to say about this version ie whether it
-  can be implemented numerically and how it compares to the above
-  results.
-
-  The limit of this section to Gaussian can be investigated later if necessary.
-}
-
-We start again from (\ref{asd}) and use (\ref{spc})
-\begin{align}
-  \lleq{asdd}
-  p(Q\cond \lambda,\bmphi)
-  &= \sum_N
-    p(N\cond \lambda)\;
-    \int \rmdx{\bmL}\;
-    p(Q\cond N,\bmL)\;
-    p(\bmL\cond N,\bmphi)\\
-  \lleq{spcd}
-  p(L_j\cond \alpha,\beta)
-  &= \frac{\beta^\alpha L_j^{\alpha-1}e^{-\beta L_j}}{\Gamma(\alpha)}.
-\end{align}
-The moment generating function of $p(L_j\cond \alpha,\beta)$ is
-\begin{align}
-  M[t\cond p(L_j\cond \alpha,\beta) ]
-  &= E[e^{tL_j}] = \int_0^\infty dL_j e^{tL_j}\,p(L_j\cond \alpha,\beta)
-    = (1 - t/\beta)^{-\alpha}
-\end{align}
-The convolution of $N$ Gamma distributions yields the moment generating
-function
-\begin{align}
-  M[t\cond p(Q\cond \alpha,\beta) ]
-  &= (1-t/\beta)^{-N\alpha}
-\end{align}
-and so we obtain the well-known result that the convolved distribution
-is also a Gamma distribution with shape parameter $N\alpha$,
-\begin{align}
-  p(Q\cond \alpha,\beta)
-  &= \frac{\beta^{N\alpha} Q^{N\alpha-1}e^{-\beta Q}}{\Gamma(N\alpha)}.
-\end{align}
-Inserting this into the $N$-sum, we have
-\begin{equation}
-  \lleq{gms}
-  \boxed{
-  p(Q\cond \lambda,\alpha,\beta)
-  = \frac{e^{-\lambda - \beta Q}}{Q}
-    \sum_{N=0}^\infty
-    \frac{[\lambda \beta^\alpha Q^\alpha]^N}{N!\;\Gamma(N\alpha)}.
-    }
-\end{equation}
-This series itself should converge quickly because of the
-$N!\;\Gamma(N\alpha)$ in the denominator. It is possible to find an
-analytic solution for some special cases of $\alpha$.
-\begin{itemize}
-\item For $\alpha=1$ we have
-  \begin{align}
-    p(Q\cond \lambda,\alpha{=}1,\beta)
-    &= \frac{e^{-\lambda - \beta Q}}{Q}
-      \sum_{N=0}^\infty
-      \frac{[\lambda \beta Q]^N}{N!\;(N-1)!}.
-  \end{align}
-  Comparing the above to the series definition of the modified Bessel function,
-  \begin{align}
-    I_1(z) &= \sum_{k=0}^\infty \frac{(z/2)^{2k+1}}{k!\;(k+1)!},
-  \end{align}
-  we can identify $z = 2\sqrt{\lambda\beta Q}$ to get
-  \begin{equation}
-    \boxed{
-      p(Q\cond \lambda,\alpha{=}1,\beta)
-      = \sqrt{\frac{\lambda\beta}{Q}}  e^{-\lambda - \beta Q}
-      \;I_1\!\left(2\sqrt{\lambda\beta Q}\right)
-      }
-  \end{equation}
-  as derived by Fisher and Cornish (1960). We often prefer to use the
-  corresponding hypergeometric function formulation, in terms of which
-  \begin{align}
-    p(Q\cond \lambda,\alpha{=}1,\beta)
-    &= \lambda\beta\; e^{-\lambda - \beta Q}
-      \;{}_0F_1(2; \lambda\beta Q).
-  \end{align}
-
-\item For integer $\alpha = 2,3,\ldots$, we rely on a paper by Withers and
-  Nadarajah (2011) from which we get an expression for the (p,q) hypergeometric
-  function, defined as
-  \begin{align}
-    {}_pF_q\left( a_1,a_2,\ldots, a_p\;;\; b_1,b_2,\ldots,b_q\;;\; z \right)
-    &= \sum_{k=0}^\infty \frac{z^k}{k!} \;
-      \frac{(a_1)_k(a_2)_k\cdots (a_p)_k}{(b_1)_k(b_2)_k\cdots (b_q)_k}
-  \end{align}
-  in terms of the Pochhammer symbol $(a)_k = \Gamma(a+k)/\Gamma(a)$. Withers
-  and Nadarajah found that (and I checked)
-  \begin{align}
-    (N\alpha)!
-    &= (\alpha^\alpha)^N
-      \left(\frac{1}{\alpha}\right)_k
-      \left(\frac{2}{\alpha}\right)_k
-      \left(\frac{\alpha{-}1}{\alpha}\right)_k
-      \left(\frac{\alpha}{\alpha}\right)_k
-  \end{align}
-  which, when inserted into the appropriate hypergeometric function
-  ${}_0F_\alpha$, yields
-  \begin{align}
-    \lleq{qpfa}
-    {}_0F_\alpha\left[ 0,0,\ldots,0 \;;\;
-    \frac{1}{\alpha},
-    \frac{2}{\alpha},
-    \ldots,
-    \frac{\alpha{-}1}{\alpha},
-    \frac{\alpha}{\alpha}
-    \;;\; z \right]
-    &= \sum_{k=0}^\infty \frac{z^k}{k!} \;\frac{(\alpha^\alpha)^k}{(k\alpha)!}
-  \end{align}
-  and with the general relation (see DLMF website)
-  \begin{align}
-    z\frac{d}{dz} {}_0F_p(\;;\;b_1,b_2,\ldots,b_q\;;\;z)
-    &= \frac{z}{b_1b_2\cdots b_q}
-      {}_0F_p(\;;\;b_1{+}1,b_2{+}1,\ldots,b_q{+}1\;;\;z)
-  \end{align}
-  we obtain, with the assignment
-  \begin{align}
-    z &= \lambda (\beta Q / \alpha)^\alpha \\
-    \lleq{qpfe}
-    p(Q\cond \lambda,\alpha,\beta)
-      &= \frac{\alpha e^{-\lambda - \beta Q} }{Q}
-        z\frac{d}{dz} \;
-        {}_0F_\alpha\left[ 0,0,\ldots,0 \;;\;
-        \frac{1}{\alpha},
-        \frac{2}{\alpha},
-        \ldots,
-        \frac{\alpha{-}1}{\alpha},
-        \frac{\alpha}{\alpha}
-        \;;\; z \right]
-        \nonumber\\
-      &= \frac{\alpha e^{-\lambda - \beta Q} }{Q}
-        \frac{z}{(\alpha-1)!\;\alpha^{\alpha-1}} \;
-        {}_0F_\alpha\left[ 0,0,\ldots,0 \;;\;
-        1{+}\frac{1}{\alpha},
-        1{+}\frac{2}{\alpha},
-        \ldots,
-        1{+}\frac{\alpha{-}1}{\alpha},
-        1{+}\frac{\alpha}{\alpha}
-        \;;\; z \right]
-        \nonumber\\
-  \end{align}
-  which on insertion of $z$ and simplification yields for integer $\alpha$
-  \begin{equation}
-    \lleq{qfp}
-    \boxed{
-      p(Q\cond \lambda,\alpha,\beta)
-      =
-    \frac{\lambda\beta^\alpha Q^{\alpha-1} e^{-\lambda - \beta Q} }{(\alpha-1)!}\;
-    {}_0F_\alpha\left[ \;;\;
-        \frac{\alpha{+}1}{\alpha},
-        \frac{\alpha{+}2}{\alpha},
-        \ldots,
-        \frac{2\alpha{-}1}{\alpha},
-        2
-        \;;\; \lambda\left(\frac{\beta Q}{\alpha}\right)^\alpha \;\right]
-      }
-  \end{equation}
-  which includes the $I_1$ Bessel function for $\alpha=1$. Mathematica
-  has the function ${}_pF_q$ available. Where it is not available, the
-  above result should reduce, of course, to the series (\ref{gms}).
-\end{itemize}
-
 \bibliographystyle{plain}

### 2017-03-08 real/code/TardisPaperPlots.jl
[PaperPlots] plot raw tardis samples

@@ -162,11 +162,15 @@ function cuba_vs_laplace(res)
     savepdf("cuba_vs_laplace")
 end
 
-"Fixed seed for reproducible spectra"
-function prepare_frame(seed=1612)
+function tardis_raw_data()
     # read in normalized sp
     raw_data = Tardis.readdata()
     frame = Tardis.filter_positive(raw_data...)
+end
+
+"Fixed seed for reproducible spectra"
+function prepare_frame(seed=1612)
+    frame = tardis_raw_data()
     Tardis.transform_data!(frame)
 
     srand(seed)
@@ -443,6 +447,39 @@ function prepare_compare_uncertainties(kwargs...)
     res, resαβ, resλαβ = compare_uncertainties(Qs; nb=nb, λ=3.465736e-01, kwargs...)
     plot_compare_uncertainties(Qs, res, resαβ, resλαβ; nb=nb)
     savepdf("comp_unc_$nb")
+end
+
+function plot_tardis_samples()
+    # first plot the raw data but sort so we have the same samples in both plots
+    frame = tardis_raw_data()
+    sort!(frame, cols=:nus)
+
+    imin = 50050
+    imax = imin + 1000
+    nbins = 60
+
+    layout = (1,2)
+
+    Plots.histogram(frame[:energies][imin:imax]/1e38, nbins=nbins, yticks=nothing,
+                    xlabel=L"\ell^\prime/10^{38}", normed=true, layout=layout)
+
+    # now the transformed data, sorted again but that doesn't change anything
+    Tardis.transform_data!(frame)
+
+    samples = frame[:energies][imin:imax]
+    dist_fit = Distributions.fit_mle(Gamma, samples)
+    α = shape(dist_fit)
+    β = 1/scale(dist_fit)
+    println("α = $α, β = $β")
+
+    Plots.histogram!(samples, xlabel=L"\ell", normed=true, nbins=nbins,
+                    lab="", layout=layout, subplot=2)
+    Plots.plot!(dist_fit, label=@sprintf("Gamma(%.2f, %.1f)", α, β), leg=true,
+                layout=layout, subplot=2, yticks=nothing, xticks=[0.0, 0.05, 0.1])
+    # no tick labels
+    # yticks!([])
+
+    savepdf("tardis_input_trafo")
 end
 
 end # module

### 2017-03-08 real/draft.tex
[draft] Major clean up
* Remove most comments
* remove todo's at beginning
* little fixes
* add plot of tardis histogram

@@
 %% CHANGE THIS DATE AS YOU MODIFY THE FILE
-\newcommand{\vdate}{170221}
+\newcommand{\vdate}{170308}
 \pagestyle{myheadings}
@@
 \newcommand{\lum}{\ell}
-% \newcommand{\rmdx}[1]{\mbox{d} #1 \,} % differential
 \newcommand{\rmdx}[1]{\dd{#1}} % differential
@@
 
-\section{What's left to do}
-
-\subsection{why asymptotic approx. has larger variance}
-\begin{enumerate}
-  \item For $\alpha=1.5, \beta=60, n=14000$ still get 50 \% larger uncertainty from asymptotic compared to summation over $N$.
-Is it due very asymmetric one-packet distribution?
-Try larger $\alpha$, for example $\alpha = 30$.
-  \item  Another option: $\int \rmdx{\alpha} \rmdx{\beta}$, use true values in simulation. Where does uncertainty come from? From $\lambda$ or $\mu, \sigma^2$?
-  \item Are finite number of terms in $\sum_N$ a problem? Check
-
-\end{enumerate}
-
-\subsection{Plots}
-
-\begin{enumerate}
-  \item generate a toy spectrum, take frequencies from actual tardis,
-  generate luminosities from a $\GammaDist(\alpha, \beta)$, no $\nu$
-  dependence of $\alpha, \beta$ for clarity , plot spectrum with
-  uncertainty bands, show it early in the paper, then describe where
-  it comes from
-  \item compare asymptotic and $\sum_N$ $p(\Lumtot \cond \bmy)$ on a
-  plot, add what happens if only $\sqrt{n}$ uncertainty is used: that
-  would neglect uncertainty on the single-packet distribution, show
-  case when parameters determined from a single bin with few samples,
-  should make large difference. Fix $\lambda$, what are the uncertainties now?
-  \item Statistics of replicas: again compare \refeq{spe} to \refeq{ral}. Teach a lesson: from a single data set, prediction will be centered there
-\end{enumerate}
-
-\subsection{Other issues that need to be resolved}
-
-\begin{enumerate}
-
-\item Clean up integrals using \verb=\rmdx= rather than $d$.
-
-\item Write everything in terms of frequency $\nu$ or wavelength
-  $\lambda$
-
-\item Now that we have removed direct analysis of \tardis\ from the
-  paper, there is no need to define an overall parameterization of the
-  Gamma parameters $\alpha(\nu_b) = \sum_k a_k \nu_b^k$ and
-  $\beta(\nu_b) = \sum_k b_k \nu_b^k$. However, the possibility of
-  linking the analyses of all the bins by this and similar methods
-  should be highlighted. Perhaps in the Conclusions?
-
-\item \textbf{\tardis} move tardis section into separate document in
-  this git repo. Better to keep it out of the paper. Cite tardis when
-  showing the toy spectrum. Mention that in reality $\alpha$ and
-  $\beta$ are $\nu$ dependent
-
-\end{enumerate}
-
 \section{Introduction (Wolfgang)}
@@
   \includegraphics[width=0.48\textwidth]{spectrum_all}
-  \caption{Spectrum with smallest 2 $\sigma$ credibility interval. Left: only packets from bin contribute. Right: a}
+  \caption{Spectrum with most likely value of $Q$ (dot) and smallest
+    2$\sigma$ credibility interval (blue band). Left: only packets
+    from single bin contribute to the single-packet luminosity
+    distribution in that bin. If $n<8$ packets observed, no reliable
+    prediction is available and bins are masked out (red band) Right:
+    Same packets as on the left but the single-packet luminosity
+    distribution is inferred from all packets. }
 \label{fig:spectrum}
@@
 example, starting the simulation with a different seed would result in
-different output data. We denote the general random variables by the
-corresponding upper-case symbols,
+different output data. Following standard notation, we denote the
+general random variables by the corresponding upper-case symbols,
 \begin{align}
@@
   \lleq{sdf}
-  \Lumtot_b &= \sum_{j=1}^{N_b} \Lum_{b,j},
+  \Lumtot_b &= \sum_{j=1}^{N_b} \Lum_{b,j}.
 \end{align}
 Lower-case symbols such as $\bmy, n_b,\bml_b$ etc.\ are strictly
-reserved for the simulated data at hand.
+reserved for the realisations; i.e., the simulated data at hand.
 
@@
 %
-In the first step, the number of packets $N$ is modelled by means of a
-separate Poisson distribution $p(N\cond\lambda)$ for each bin, where
-the change in magnitude of $\lambda$ from bin to bin will capture most
-of the variability of the spectrum. Given $N$, the single-packet
+In the first step, the number of packets $N$ is modelled by a Poisson
+distribution $p(N\cond\lambda)$, separately for each bin, where the
+change in magnitude of $\lambda$ from bin to bin will capture most of
+the variability of the spectrum. Given $N$, the single-packet
 luminosity distribution is modelled in the second step by means of a
@@
 distribution contains the crucial information which allows us to
-extend previous methods to cases where the number of packets in a bin
-is small.
-%%%%%%%%%%%%%%%%%%%
-\fred{Refs: Zech, maybe others, too}
+extend previous
+methods~\cite{bohm_comparison_2012,bohm_statistics_2014} to cases
+where the number of packets in a bin is small and even zero.
 
@@
 $\int_0^\infty \rmdx{\lambda}\,p(\lambda)$ is known only up to a
-constant $c$. An improper prior is not a problem because $c$ cancels
-out when calculating the posterior, provided that $n \ge 1$; i.e., at
-least one packet is observed in the bin.\footnote{For a uniform prior,
-  the case $a{=}0$, we require of course a parameter
-  $\lambda_{\rm max}$. As long as $\lambda_{\rm max}$ is large enough,
-  it would also cancel.}, The posterior is
+constant $c$.\footnote{For a uniform prior, the case $a{=}0$, we
+  additionally require an upper limit $\lambda_{\rm max}$. As long as
+  $\lambda_{\rm max}$ is large enough, it would also cancel.}, The
+posterior is
 \begin{align}
@@
 The evidence is
-$p(n) =  \int_0^\infty \rmdx{\lambda}\, p(n\cond\lambda)\,p(\lambda) = c\,(n-a)!/n!$
+$p(n) =  \int_0^\infty \rmdx{\lambda}\, p(n\cond\lambda)\,p(\lambda) = c\,(n-a)!/n!$.
 % \begin{align}
@@
 % \end{align}
-and inserting Eqs.~(\ref{pmb}) and (\ref{pmc}), the prediction for $N$
-given $n$ is a negative binomial distribution
+Upon inserting Eqs.~(\ref{pmb}) and (\ref{pmc}), $c$ cancels out and
+the integral over $\lambda$ can be performed analytically resulting in
+a negative binomial distribution that is proper for any $n$
 \begin{align}
@@
 transformation-group result advocated by
-Jaynes~\cite[Ch. 12]{jaynes2003probability}. The uniform prior has no
-solid information-theoretic foundation in this case reducing the
-choice to $a \in \{1/2, 1\}$. We examine the consequence for the
-extreme case that no packet is observed: $P(N=0 \cond n=0) =
-2^{a-1}$. For $a=1$, any $N>0$ is excluded which is undesirable unless
-we accept that repeating the simulation will never yield a packet in
-the current bin. Hence we are led to Jeffreys prior, $a=1/2$.
+Jaynes~\cite[Ch. 12]{jaynes2003probability}. To decide on $a$, we note
+that the uniform prior has no solid information-theoretic foundation
+in this case reducing the choice to $a \in \{1/2, 1\}$. We examine the
+consequence for the extreme case that no packet is observed:
+$P(N=0 \cond n=0) = 2^{a-1}$. For $a=1$, any $N>0$ is excluded which
+is undesirable unless we accept that repeating the simulation will
+never yield a packet in the current bin. Hence we settle on Jeffreys
+prior, $a=1/2$, for any numerical evaluation carried out below.
 
@@
 assumed to be mutually independent and to follow the same distribution
-independent of $N$.
+independent of $N$ and the exact frequency within the bin under consideration.
 %
@@
 \end{align}
-which has the nice property that the convolution (\ref{slm}) results
-in another Gamma distribution,
+which has the nice property that the convolution \refeq{slm} for
+$N > 0$ results in another Gamma distribution,
 \begin{align}
@@
   \int \rmdx{\bmL}\; p(Q\cond N,\bmL)\; p(\bmL\cond N,\bmphi)
-  &= \GammaDist(Q\cond N\alpha, \beta),
-\end{align}
-and so Eq.~(\ref{gmh}) becomes in this example
+  &= \GammaDist(Q\cond N\alpha, \beta).
+\end{align}
+For the special case $N=0$, there is no luminosity because there is no
+packet, so we define
+\begin{align}
+  \lleq{spl}
+   p(Q\cond N, \alpha, \beta) &\equiv
+   \begin{cases}
+     \delta(Q),     &N=0,\\
+     \GammaDist(Q\cond N\alpha, \beta), & N > 0 \, ,
+   \end{cases}
+ \end{align}
+ with the Dirac $\delta$ function and so \refeq{gmh} simplifies to
 \begin{align}
@@
     \int \rmdx{\alpha}\,\rmdx{\beta}\;
-    \GammaDist(Q\cond N\alpha, \beta)\;
+    p(Q\cond N, \alpha, \beta)\;
+    % \GammaDist(Q\cond N\alpha, \beta)\;
     p(\alpha,\beta\cond n,\bml).
@@
 
-\fred{Here and above, $n$ does not have to be the same in
-  $P(N \cond n)$ and $p(n,\bml\cond \alpha,\beta)$. We can use packets
-  from other bins if we assume $\alpha, \beta$ are the same in those bins.}
-\hans{Please explain more fully; not clear.}
-
-It only remains to explicitly evaluate the prior $p(\alpha,\beta)$
-appearing in \refeq{spf}.
+It only remains to specify the prior $p(\alpha,\beta)$ appearing in
+\refeq{spf}.
 %
@@
 their recommendation because their modification is not well motivated
-and it seems to accidentally perform better because of the particular
-examples considered. The other conclusion they have is that the
-Jeffreys, reference, and other noninformative priors with similarly
-solid motivation have comparable performance. They all exhibit the
-standard form for a scale or rate parameter $\propto 1/\beta$ and only
-differ in the $\alpha$ dependence. We disregard the reference prior as
-it forces us to identify either $\alpha$ or $\beta$ as \emph{the}
-nuisance parameter but we consider \emph{both} as nuisance
-parameters. Hence we select the Jeffreys prior
-\hans{What is $\Psi$?}
+and it seems to only accidentally perform better because of the
+particular examples they considered. The other conclusion they have is
+that the Jeffreys, reference, and other noninformative priors with
+similarly solid motivation have comparable performance. They all
+exhibit the standard form for a scale or rate parameter
+$\propto 1/\beta$ and only differ in the $\alpha$ dependence. We
+disregard the reference prior as it forces us to identify either
+$\alpha$ or $\beta$ as \emph{the} nuisance parameter but we consider
+\emph{both} as nuisance parameters. Hence we select the Jeffreys prior
 \begin{align}
@@
   p(\alpha,\beta)
-  &=  C \frac{\sqrt{\alpha \Psi'(\alpha)-1}}{\beta} \, .
-  % \frac{1}{(\alpha_{\rm max}{-}1)\beta_{\rm max}}
-  % \qquad 1 < \alpha < \alpha_{\rm max},\; 0 < \beta < \beta_{\rm max}.
-\end{align}
-% since $\alpha_{\rm max}$ and $\beta_{\rm max}$ cancel in (\ref{spf}).
-With these choices, the posterior normalization is %
-\hans{``Posterior normalization'' = Evidence? With improper priors,
-  should the upper limits not be $\infty$ below?}
+  &=  C \frac{\sqrt{\alpha \Psi'(\alpha)-1}}{\beta} \, ,
+\end{align}
+where $\Psi$ is the digamma function. With these choices, the
+posterior is nondegenerate for $n \geq 2$ and the evidence is
 \begin{align}
   \lleq{spj}
-  p(n, \bml) &= C \int_0^{\alpha_{\rm max}} d\alpha \int_0^{\beta_{\rm max}} d\beta\;
-  \frac{\beta^{n\alpha} r^{\alpha-1}\,e^{-\beta q}}{[\Gamma(\alpha)]^n},
-\end{align}
-where the upper integration limits
-$(\alpha_{\rm max}, \beta_{\rm max})$ should formally extend to
-$\infty$ but in practice have to be chosen by trial and error to be
-large enough to contain the relevant contribution to the integral.
+  p(n, \bml) &= C \int_0^{\infty} \rmdx{\alpha} \int_0^{\infty} \rmdx{\beta}\;
+  \frac{\beta^{n\alpha} r^{\alpha-1}\,e^{-\beta q}}{[\Gamma(\alpha)]^n}\;
+  \frac{\sqrt{\alpha \Psi'(\alpha)-1}}{\beta}.
+\end{align}
 The prior's constant of proportionality $C$ cancels in the final
@@
   &= \frac{1}{p(n, \bml)}
-    \sum_{N=0}^\infty \;
+    \sum_{N=1}^\infty \;
     p(N\cond n)\;
@@
     \GammaDist(Q\cond N\alpha, \beta)\;
-    \frac{\beta^{n\alpha} r^{\alpha-1}\,e^{-\beta q}}{[\Gamma(\alpha)]^n} \frac{\sqrt{\alpha \Psi'(\alpha)-1}}{\beta}.
-\end{align}
-
-\subsection{Preprocessing simulation output } \label{sec:tardis}
-
-\fred{I removed a lot of cruft and swapped the symbols
-  $x \to \ell, \ell \to \ell^\prime$ so it's in line with the previous section}
-
-\hans{This and the next section interrupt the actual calculation of
-  exact results in Sections
-  \ref{sec:multiplicity}--\ref{sec:single-packet} and the asymptotic
-  ones in Section \ref{sec:asymptotic}, so I would suggest moving them
-  to the end just before the conclusions. }
-
-In this section, we use \tardis{} as an example to demonstrate both how
-transforming the output of a simulation helps in predictions, and what
-further complications may arise in practice.
-
-\tardis{} can output two kinds of packets, real and virtual. Here we
-focus on the real packets and denote the raw output of \tardis{} by
-primed symbols, for example for the $i$-{th} packet
-$\qty(\nu_i^\prime, \ell_i^\prime)$.  A histogram plot
-\fredmargin{include plot} of the luminosities from a \tardis{} run
-indicates a strongly skewed distribution with its single mode close to
-the maximum luminosity $\lmax^\prime$, while there is a long tail for
-small $\ell$. These features can be accommodated by a $\GammaDist$
-distribution as in \refsec{single-packet} if we transform the
-luminosities to
-\begin{align}
-  \lleq{gmf}
-  \ell_i \equiv 1- \frac{\ell_i^\prime}{\lmax^\prime} > 0
-\end{align}
-and similarly $L_i \equiv 1 - (L_i^\prime/\lmax^\prime)$. We assume
-the total number of samples across all bins is large, so that the effect
-of considering $\lmax^\prime$ a constant independent of the data is
-negligible. We checked that leaving $\lmax^\prime$ as a free parameter
-in a fit gave equivalent results.
-%
-Numerically, we take the largest luminosity of all \tardis{} packets
-across the entire frequency range and increase it slightly to avoid
-numerical difficulties at the end point
-\begin{align}
-  \lleq{gme}
-  \lmax^\prime \equiv (1+\epsilon) \times \max_i \ell_i^\prime,
-\end{align}
-where $ \epsilon=10^{-6}$. We carry out all calculations with
-$p(\Lumtot \cond n,\bml)$ and only in the end transform back to a
-probability density in the raw coordinates via
-\begin{align}
-  \lleq{gmg}
-  p(\Lumtot^\prime  \cond n, \bml^\prime)
-  &= p(\Lumtot  \cond n, \bml) \frac{1}{\lmax^\prime} \,.
-\end{align}
-While we also rescale the frequencies by a constant from
-$\mathcal{O}(10^{15})$ to $\mathcal{O}(1)$, the most important
-numerical simplification arises because of the transformation of the
-luminosities. Rewriting \refeq{sll} formally as
-\begin{align}
-  \lleq{gmhb}
-  p(Q  \cond n,\bml)
-  & = \sum_{N} p(N \cond n)\; p(Q \cond N, n, \bml),
-\end{align}
-we see that the right-hand side is a mixture density. In the \tardis{}
-application, the distribution of a single-packet luminosity,
-$p\qty(\ell_i' \cond \bmphi)$ is narrow in the sense that the standard
-deviation relative to the mean is at the percent level. This implies
-that each $p(Q^\prime \cond N, n, \bml^\prime)$ contributes one narrow
-peak and all peaks are clearly separated in the superposition
-$ p(Q^\prime \cond n,\bml^\prime)$. The transformation
-$\ell^\prime \to \ell$ effects a strong overlap between the
-distributions $p(\Lumtot \cond N, n, \bml)$ with similar $N$, so
-$p(\Lumtot \cond n, \bml)$ can be well approximated by a unimodal
-distribution and thus $\Lumtot$ rather than $\Lumtot^\prime$ is the
-preferred coordinate basis.
-%
-\fredmargin{show hedgehog plot}
-
-\subsection{Towards an unbinned analysis} \label{sec:unbinned}
-
-The parameters $\qty(\alpha, \beta)$ of the single-packet distribution
-$\GammaDist(\ell_i \cond \alpha, \beta)$ may, of course, depend on the
-frequency. % $\nu_i$
-This effect could be modelled, for example, by a functional
-parametrization $\alpha = \alpha(\nu \cond \bmphi)$ and
-$\beta = \beta(\nu \cond \bmphi)$ in dependence on the frequency $\nu$
-and on hyperparameters $\bmphi$ such that integrals over
-$\alpha, \beta$ would turn into integrals over $\bmphi$. For the
-sake of clarity, let rather us consider only the simplest case where
-$\alpha, \beta$ are the same in every bin such that \refeq{spk}
-becomes
-\begin{align}
-  \lleq{gmi}
-  p(Q\cond \npack,\bmy)
-  &=     \sum_{N=0}^\infty \;
-    p(N\cond n)\;
-    \int \rmdx{\alpha}\,\rmdx{\beta}\;
-    \GammaDist(Q\cond N\alpha, \beta)\;
-    p(\alpha, \beta \cond \npack, \bmy) \, .
-\end{align}
-In words, the posterior $ p(\alpha, \beta \cond \npack, \bmy)$ uses
-all packets to essentially fix $\alpha, \beta$ if $\npack \gg 1$ and
-allows meaningful predictions of $\Lumtot$ even in bins where
-\emph{no} packet is observed; i.e., $n=0$. The only binning left is
-performed to count $n$ the number of packets that fall into a given
-frequency bin. This strategy is applicable, for example, when
-absorption or emission lines in the spectrum abruptly alter the number
-of packets for nearby frequencies but the luminosities are unaffected
-by those line features. Note that if the dependence on $\bmphi$ can be
-modeled with few parameters, the total number of parameters may be
-drastically reduced: instead of determining $\alpha, \beta$ separately for
-each of a few thousand bins, there are potentially only a handful of
-elements in $\bmphi$ for the entire spectrum. In that case, the
-integral in \refeq{gmi} does not depend on the bin but only on $N$ so
-its value may be cached to speed up the simultaneous evaluation across
-all bins.
-
-Let's consider the extreme case $\npack \gg 1$ so we can neglect the
-uncertainty on $\alpha, \beta$ and
-$p(\alpha, \beta \cond \npack, \bmy) =
-\delta(\alpha-\alpha_0)\delta(\beta - \beta_0)$ such that
-\begin{align}
-  \lleq{gmj}
-  p(Q\cond \npack,\bmy)
-  &=     \sum_{N=0}^\infty p(N\cond n) \GammaDist(Q\cond N\alpha_0, \beta_0)\\
-  &=     \sum_{N=0}^\infty p(N\cond n) p(Q\cond N, \alpha_0, \beta_0) \, .
-\end{align}
-The case $N=0$ deserves special attention. To preserve the
-normalization of $p(Q\cond \npack,\bmy)$, we need to define
-\begin{align}
-  \lleq{gmk}
-   p(Q\cond N, \alpha, \beta) &\equiv
-   \begin{cases}
-     \delta(Q),     &N=0;\\
-     \GammaDist(Q\cond N\alpha, \beta), & N > 0 \, .
-   \end{cases}
- \end{align}
- %
-\hans{I'm ok with this now.}
- \fred{Notation is sloppy before, it has nothing to do with unbinned
-   stuff or $\npack \gg 1$. This is just where it became apparent to
-   me.}
- %
- Intuitively, we can only observe $Q=0$ for $N=0$ packets but
- mathematically, $\GammaDist(Q\cond \alpha=0, \beta) = 0$ which,
- without the extra contribution from the Dirac $\delta$ function,
- would imply
- $p(Q\cond \npack,\bmy) = \sum_{N>0}^\infty p(N \cond n) < 1$. Higher
- moments of $p(Q \cond N, \alpha, \beta)$ are unaffected by this
- subtlety.
-
- In \reffig{comp-unc}, we illustrate the impact of the uncertainty due
- to $\lambda$ or $\alpha, \beta$ for $n=2$ packets and continue to use
- the true values $\alpha_0 = 1.5, \beta_0 = 60$. The standard error
- increases from 0.042 ($\lambda, \alpha, \beta$ fixed) over 0.059
- ($\alpha, \beta$ fixed) to 0.11 (nothing fixed). The case where all
- parameters are known is usually not attainable in practice and only
- shown for comparison. But the case $\alpha, \beta$ fixed can be
- reached easily if packets from other bins are used. Note that $n=2$
- is the minimum number of packets to perform the binned analysis.
-
- \begin{figure}[ht]
-  \centering
-  \includegraphics[width=0.44\textwidth]{comp_unc_0}
-  \includegraphics[width=0.44\textwidth]{comp_unc_2}
-  \caption{Prediction considering $\lambda, \alpha, \beta$ fixed (green), $\alpha, \beta$ fixed (red), or nothing fixed (blue).}
-\label{fig:comp-unc}
-\end{figure}
+    \frac{\beta^{n\alpha} r^{\alpha-1}\,e^{-\beta q}}{[\Gamma(\alpha)]^n} \frac{\sqrt{\alpha \Psi'(\alpha)-1}}{\beta}
+\end{align}
+for $Q > 0$ and for $Q=0$ there is the additional contribution from
+$N=0$ through the $\delta$ function from \refeq{spl}.
 
@@
 \end{align}
-for any probability $p(L\cond\bmphi)$.
+for any density $p(L\cond\bmphi)$.
 %
@@
   Eq.~(\ref{rab}) due to $\kappa_3$ and higher orders is unclear.}
-\fred{These guys \href{http://stats.stackexchange.com/questions/83069/what-is-the-maximum-entropy-probability-density-function-for-a-positive-continuo}{claim} that our truncated Gauss is the max-ent distribution for our prior information.
-I'm not sure they are right, they are not convinced themselves.
-We have to check.}
 %%%%%%%%%%%
@@
 that the asymptotic distribution of $\Lumtot$ when
-$\lambda \to \infty$, given the first and second moment of $\Lum$, is
+$\lambda \to \infty$, given the finite first and second moment of $\Lum$, is
 a Gaussian or normal distribution,
@@
 \end{align}
-This drastic simplification eliminates both the sum over $N$ and the
-integral over $\bmL$ in \refeq{asd}.
-
-It remains to find the integrals over $\lambda$ and $\bmphi$ in
-\refeq{asc}. As before, the posterior for $\lambda$ is the Gamma
-distribution of \refeq{pms}. %
+It is remarkable that this result eliminates both the sum over $N$ and
+the integral over $\bmL$ in \refeq{asd}.  What remains is to find the
+integrals over $\lambda$ and $\bmphi$ in \refeq{asc}. As before, we
+assume the posterior for $\lambda$ is the Gamma distribution of
+\refeq{pms} since this is independent of $p(L\cond\bmphi)$.
 %
@@
 model for $p(L\cond\bmphi)$ the only relevant information is contained
-in the first moment $\mu$ and variance $\sigma^2$.  As set out in
-Jaynes~\cite[Ch. 7.10]{jaynes2003probability}, knowledge of $\mu$ and
-$\sigma^2$ implies that we can assign a Gaussian distribution to
-$p(L\cond\bmphi)$
+in the first moment $\mu$ and variance $\sigma^2$. The maximum-entropy
+distribution for $\Lum$ given knowledge of only $\mu$ and $\sigma^2$
+is a Gaussian distribution, and its parameters are just $\mu$ and
+$\sigma^2$ if the contribution from $\Lum < 0$ is negligible,
+which we
+may safely assume in the asymptotic regime, so we set
+%
+\fred{I checked this myself and clarified \href{http://stats.stackexchange.com/questions/83069/what-is-the-maximum-entropy-probability-density-function-for-a-positive-continuo/265763}{this unclear answer} on stats.stackexchange }
+%
 \begin{align}
@@
 \end{align}
-even though the actual data histograms may not be Gaussian. The reason
-is that a Gaussian represents our state of knowledge \emph{before}
-data $\bmy$ is observed. With this assignment,
-$\bmphi =(\mu,\sigma^2)$ and the posterior becomes, using Bayes'
-Theorem,
+and $\bmphi =(\mu,\sigma^2)$ and the posterior
+becomes, using Bayes' Theorem,
 \begin{align}
@@
 
+\section{Practical considerations} \label{sec:prac-consid}
+
+In this section, we use \tardis{} as an example to demonstrate both how
+transforming the output of a simulation helps in predictions, and what
+further complications may arise in practice.
+
+\subsection{Preprocessing simulation output } \label{sec:tardis}
+
+\tardis{} can output two kinds of packets, real and virtual. Here we
+focus on the real packets and denote the raw output of \tardis{} by
+primed symbols, for example for the $i$-{th} packet
+$\qty(\nu_i^\prime, \ell_i^\prime)$.  A histogram plot of the
+luminosities from a \tardis{} run indicates a strongly skewed
+distribution with its single mode close to the maximum luminosity
+$\lmax^\prime$, while there is a long tail for small $\ell$; see
+\reffig{tardis}. These features can be accommodated by a $\GammaDist$
+distribution as in \refsec{single-packet} if we transform the
+luminosities to
+\begin{align}
+  \lleq{gmf}
+  \ell_i \equiv 1- \frac{\ell_i^\prime}{\lmax^\prime} > 0
+\end{align}
+and similarly $L_i \equiv 1 - (L_i^\prime/\lmax^\prime)$. We assume
+the total number of samples across all bins is large, so that the effect
+of considering $\lmax^\prime$ a constant independent of the data is
+negligible. We checked that leaving $\lmax^\prime$ as a free parameter
+in a fit gave equivalent results.
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=0.7\textwidth]{tardis_input_trafo}
+  \caption{Histogram of 1000 packet luminosities in one frequency bin
+    from \tardis{} before (left) and after (right) transforming. The
+    maximum-likelihood fit of $\GammaDist(\alpha, \beta)$ is overlaid.}
+\label{fig:tardis}
+\end{figure}
+
+Numerically, we take the largest luminosity of all \tardis{} packets
+across the entire frequency range and increase it slightly to avoid
+numerical difficulties at the end point
+\begin{align}
+  \lleq{gme}
+  \lmax^\prime \equiv (1+\epsilon) \times \max_i \ell_i^\prime,
+\end{align}
+where $ \epsilon=10^{-6}$. We carry out all calculations with
+$p(\Lumtot \cond n,\bml)$ and only in the end transform back to a
+probability density in the raw coordinates via
+\begin{align}
+  \lleq{gmg}
+  p(\Lumtot^\prime  \cond n, \bml^\prime)
+  &= p(\Lumtot  \cond n, \bml) \frac{1}{\lmax^\prime} \,.
+\end{align}
+While we also rescale the frequencies by a constant from
+$\mathcal{O}(10^{15})$ to $\mathcal{O}(1)$, the most important
+numerical simplification arises because of the transformation of the
+luminosities. Rewriting \refeq{sll} formally as
+\begin{align}
+  \lleq{gmhb}
+  p(Q  \cond n,\bml)
+  & = \sum_{N} p(N \cond n)\; p(Q \cond N, n, \bml),
+\end{align}
+we see that the right-hand side is a mixture density. In the \tardis{}
+application, the distribution of a single-packet luminosity,
+$p\qty(\ell_i' \cond \bmphi)$ is narrow in the sense that the standard
+deviation relative to the mean is at the percent level. This implies
+that each $p(Q^\prime \cond N, n, \bml^\prime)$ contributes one narrow
+peak and all peaks are clearly separated in the superposition
+$ p(Q^\prime \cond n,\bml^\prime)$. The transformation
+$\ell^\prime \to \ell$ effects a strong overlap between the
+distributions $p(\Lumtot \cond N, n, \bml)$ with similar $N$, so
+$p(\Lumtot \cond n, \bml)$ can be well approximated by a unimodal
+distribution and thus $\Lumtot$ rather than $\Lumtot^\prime$ is the
+preferred coordinate basis.
+
+\subsection{Towards an unbinned analysis} \label{sec:unbinned}
+
+The parameters $\qty(\alpha, \beta)$ of the single-packet distribution
+$\GammaDist(\ell_i \cond \alpha, \beta)$ may, of course, depend on the
+frequency. % $\nu_i$
+This effect could be modelled, for example, by a functional
+parametrization $\alpha = \alpha(\nu \cond \bmphi)$ and
+$\beta = \beta(\nu \cond \bmphi)$ in dependence on the frequency $\nu$
+and on hyperparameters $\bmphi$ such that integrals over
+$\alpha, \beta$ would turn into integrals over $\bmphi$. For the sake
+of clarity, let us consider only the simplest case where
+$\alpha, \beta$ are the same in every bin such that \refeq{spe}
+becomes
+\begin{align}
+  \lleq{gmi}
+  p(Q\cond \npack,\bml)
+  &=     \sum_{N=0}^\infty \;
+  p(N\cond n)\;
+  \int \rmdx{\alpha}\,\rmdx{\beta}\;
+  p(\Lumtot \cond N, \alpha, \beta)\;
+  p(\alpha, \beta \cond \npack, \bmy) \, .
+\end{align}
+%   If it assumed that $p(L_j\cond \bmphi)$ does not change across
+% \emph{different} bins, we may include packets from \emph{all} bins to
+% further constrain $\alpha, \beta$ and the only required change is to
+% set $n \to \npack$ in \refeq{spg}. Note that we would still keep $n$
+% as before in $p(N \cond n)$ or else we would squash the entire
+% spectrum into one bin.
+
+In words, the posterior $ p(\alpha, \beta \cond \npack, \bmy)$ uses
+all $\npack$ rather than $n$ packets to determine $\alpha, \beta$. For
+$\npack \gg 1$, this allows meaningful predictions of $\Lumtot$ even
+in bins where \emph{no} packet is observed; i.e., $n=0$. The only
+binning left is performed to count $n$ the number of packets that fall
+into a given frequency bin. This strategy is applicable, for example,
+when absorption or emission lines in the spectrum abruptly alter the
+number of packets for nearby frequencies but the luminosities are
+unaffected by those line features. Note that if the dependence on
+$\bmphi$ can be modeled with few parameters, the total number of
+parameters may be drastically reduced: instead of determining
+$\alpha, \beta$ separately for each of a few thousand bins, there are
+potentially only a handful of elements in $\bmphi$ for the entire
+spectrum. In that case, the integral in \refeq{gmi} does not depend on
+the bin but only on $N$ so its value may be cached to speed up the
+simultaneous evaluation across all bins.
+
+Let's consider the extreme case $\npack \to \infty$ so we can neglect
+the uncertainty on $\alpha, \beta$ and
+$p(\alpha, \beta \cond \npack, \bml) =
+\delta(\alpha-\alpha_0)\delta(\beta - \beta_0)$ such that
+\begin{align}
+  \lleq{gmj}
+  p(Q\cond n,\bml)
+  &=     \sum_{N=0}^\infty p(N\cond n) p(Q\cond N, \alpha_0, \beta_0) \, .
+\end{align}
+For illustration only, let us consider the Poisson rate $\lambda$ were
+known. Then learning $n$ would not improve our knowledge of $\lambda$
+and we would have
+\begin{align}
+  \lleq{gmk}
+  p(Q\cond \lambda,\bml)
+  &=     \sum_{N=0}^\infty p(N\cond \lambda) p(Q\cond N, \alpha_0, \beta_0) \, ,
+\end{align}
+where $p(N \cond \lambda)$ is the Poisson distribution \refeq{pmb}.
+In \reffig{comp-unc}, we illustrate the impact of the uncertainty due
+to $\lambda$ or $\alpha, \beta$ for $n=2$ packets and continue to use
+as true values $\alpha_0 = 1.5, \beta_0 = 60$. The standard error
+increases from 0.042 ($\lambda, \alpha, \beta$ fixed) over 0.059
+($\alpha, \beta$ fixed) to 0.11 (nothing fixed).
+%
+\fred{I guess this is for $n=2$? To what is $\lambda$ fixed? Redo the
+  numbers for the actual prior. Add a plot for larger $n$ to show that
+  the Poisson completely dominates}
+%
+The case where all
+parameters are known is not attainable in practice and only shown for
+comparison. But the case where $\alpha, \beta$ are fixed can be reached easily
+if many packets from other bins can be combined. Note that $n=2$ is the
+minimum number of packets to perform the binned analysis.
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=0.44\textwidth]{comp_unc_0}
+  \includegraphics[width=0.44\textwidth]{comp_unc_2}
+  \caption{Prediction considering $\lambda, \alpha, \beta$ fixed (green), $\alpha, \beta$ fixed (red), or nothing fixed (blue).}
+  \label{fig:comp-unc}
+\end{figure}
+
 \section{Conclusion} \label{sec:conclusion}
 
-TODO
+\begin{enumerate}
+  \item Poisson uncertainty dominates
+  \item can predict reliably for small $n$
+\end{enumerate}
 

### 2017-03-08 real/code/PaperPlots.jl
[code] new PaperPlots.jl to create all plots and numbers

@@ -1,0 +1,4 @@
+include("TardisPlotUtils.jl")
+
+# fig:tardis
+TardisPlotUtils.plot_tardis_samples()

### 2017-03-08 real/code/TardisPlotUtils.jl

@@ -1,4 +1,4 @@
-module TardisPaperPlots
+module TardisPlotUtils
 using Logging
 @Logging.configure(level=INFO)
 

### 2017-03-09 real/code/GammaIntegrand.jl
[GammaIntegrand] Add Jeffreys prior for alpha, beta

@@ -17,7 +17,8 @@ log_prior(α::Real, β::Real) = 1/2 * log(α*Ψ′(α)-1) - log(β)
 
 """ posterior p(α, β | n, q, r) for the parameters of the Gamma
 distribution given the sufficient statistics of the samples. (spg)"""
-log_posterior(α::Real, β::Real, n::Real, q::Real, logr::Real, evidence=0.0) = n*(α*log(β)-lgamma(α)) + (α-1)*logr - β*q - evidence # + log_prior(α, β)
+log_posterior(α::Real, β::Real, n::Real, q::Real, logr::Real, evidence=0.0) = n*(α*log(β)-lgamma(α)) + (α-1)*logr - β*q - evidence + log_prior(α, β)
+
 function make_log_posterior(n::Real, q::Real, logr::Real, evidence=0.0)
     function (θ::Vector)
         α, β = θ

### 2017-03-09 real/code/PaperPlots.jl
[code] Tune uncertainty comparison plot

@@ -1,4 +1,7 @@
 include("TardisPlotUtils.jl")
 
+# fig:comp-unc
+TardisPlotUtils.prepare_compare_uncertainties()
+
 # fig:tardis
 TardisPlotUtils.plot_tardis_samples()

### 2017-03-09 real/code/Predict.jl

@@ -129,6 +129,7 @@ Arguments
 
 function by_cubature(Q, a, n, q, logr, nb=n; αmin=1e-2, αmax=5, βmin=1e-5, βmax=100,
                      Ninit=0, ε=1e-3, reltol=1e-5)
+    debug("Predicting for Q=$Q")
     # to avoid overflow in Cubature, evaluate target at mode and subtract it. The
     # value within a few order of magnitude of
     fit_dist = Distributions.fit_mle(Distributions.Gamma, Distributions.GammaStats(q, logr, n))
@@ -146,7 +147,7 @@ function by_cubature(Q, a, n, q, logr, nb=n; αmin=1e-2, αmax=5, βmin=1e-5, β
     upper = [αmax, βmax]
 
     Z, σ, ncalls = Integrate.by_cubature(make_log_posterior(n, q, logr, logf_mode), lower, upper; reltol=reltol)
-    debug("cubature: $(logf_mode), $Z, $σ, $ncalls")
+    debug("cubature evidence: $(logf_mode), $Z, $σ, $ncalls")
 
     # actual evidence larger by logf because we already subtracted it
     # in log_posterior. We need evidence on log scale

### 2017-03-09 real/code/TardisPlotUtils.jl

@@ -1,6 +1,6 @@
 module TardisPlotUtils
 using Logging
-@Logging.configure(level=INFO)
+@Logging.configure(level=DEBUG)
 
 import TardisPaper.GammaIntegrand, TardisPaper.Predict, TardisPaper.Integrate, TardisPaper.Moments, TardisPaper.SmallestInterval
 import Tardis
@@ -420,33 +420,42 @@ function compare_uncertainties(Qs;nb=2, λ=nb, n=nb, α=1.5, β=60.0, a=1/2, ε=
     res, resαβ, resλαβ
 end
 
-function plot_compare_uncertainties(Qs, res, resαβ, resλαβ; nb=2)
+function plot_compare_uncertainties(Qs, res, resαβ, resλαβ; nb=2, kwargs...)
     # plot(Qs, res, xlabel="Q", ylabel="P(Q | n=2)", leg=true, label=L"p(N | n=2) \int d \alpha d \beta p(Q| N, \alpha, \beta) p(\alpha, \beta | n, y)")
     # plot!(Qs, resαβ, label=L"p(N | n=2) p(Q | N, \alpha_0, \beta_0)")
     # plot!(Qs, resλαβ, label=L"p(N | \lambda=2) p(Q | N, \alpha_0, \beta_0)")
 
-    plot(Qs, resλαβ, label=L"\lambda, \alpha, \beta fixed", xlabel="Q", ylabel="P(Q | n=$nb)", leg=true)
-    plot!(Qs, resαβ, label=L"\alpha, \beta fixed")
-    (res[end] > 0) && plot!(Qs, res, label=L"nothing fixed")
-    plot!()
+    plot!(Qs, resλαβ; label=L"\lambda, \alpha, \beta fixed", xlabel="Q",
+          leg=true, kwargs...)
+    plot!(Qs, resαβ; label=L"\alpha, \beta fixed", kwargs...)
+    (res[end] > 0) && plot!(Qs, res; label=L"nothing fixed", kwargs...)
+    # plot!(title="n=$nb")
 end
 
 function prepare_compare_uncertainties(kwargs...)
-    K = 150
+    # clean canvas with two plots next to each other
+    plot_kwargs = Dict(:layout=>(1,2), :subplot=>2)
+    plot(;plot_kwargs...)
 
-    Qs = linspace(0.001, 0.8, K)
+    # K = 150
+    # Qs = linspace(0.001, 0.8, K)
+    K = 10
[…]

### 2017-03-09 real/code/GammaIntegrand.jl
[code] Revamp comparison plot
add n=20, three subfigures

@@ -17,7 +17,7 @@ log_prior(α::Real, β::Real) = 1/2 * log(α*Ψ′(α)-1) - log(β)
 
 """ posterior p(α, β | n, q, r) for the parameters of the Gamma
 distribution given the sufficient statistics of the samples. (spg)"""
-log_posterior(α::Real, β::Real, n::Real, q::Real, logr::Real, evidence=0.0) = n*(α*log(β)-lgamma(α)) + (α-1)*logr - β*q - evidence + log_prior(α, β)
+log_posterior(α, β, n, q, logr, evidence=0.0) = n*(α*log(β)-lgamma(α)) + (α-1)*logr - β*q - evidence + log_prior(α, β)
 
 function make_log_posterior(n::Real, q::Real, logr::Real, evidence=0.0)
     function (θ::Vector)

### 2017-03-09 real/code/TardisPlotUtils.jl

@@ -1,12 +1,14 @@
 module TardisPlotUtils
 using Logging
-@Logging.configure(level=DEBUG)
+# @Logging.configure(level=DEBUG)
+@Logging.configure(level=INFO)
 
 import TardisPaper.GammaIntegrand, TardisPaper.Predict, TardisPaper.Integrate, TardisPaper.Moments, TardisPaper.SmallestInterval
 import Tardis
 using DataFrames, Distributions, LaTeXStrings, Plots, StatPlots, StatsBase
 
-const font = Plots.font("TeX Gyre Heros")
+# const font = Plots.font("TeX Gyre Heros") # Arial sans serif
+const font = Plots.font("cmr10") # computer modern roman (LaTeX default)
 Plots.pyplot(guidefont=font, xtickfont=font, ytickfont=font, legendfont=font)
 Plots.PyPlotBackend()
 # Plots.gr()
@@ -420,40 +422,56 @@ function compare_uncertainties(Qs;nb=2, λ=nb, n=nb, α=1.5, β=60.0, a=1/2, ε=
     res, resαβ, resλαβ
 end
 
-function plot_compare_uncertainties(Qs, res, resαβ, resλαβ; nb=2, kwargs...)
-    # plot(Qs, res, xlabel="Q", ylabel="P(Q | n=2)", leg=true, label=L"p(N | n=2) \int d \alpha d \beta p(Q| N, \alpha, \beta) p(\alpha, \beta | n, y)")
-    # plot!(Qs, resαβ, label=L"p(N | n=2) p(Q | N, \alpha_0, \beta_0)")
-    # plot!(Qs, resλαβ, label=L"p(N | \lambda=2) p(Q | N, \alpha_0, \beta_0)")
+function plot_compare_uncertainties(Qs, res, resαβ, resλαβ; nb=2, upscale=200, kwargs...)
+    if upscale > 0
+        Qsscaled, resλαβ = SmallestInterval.upscale(Qs, resλαβ, upscale)
+        _, resαβ = SmallestInterval.upscale(Qs, resαβ, upscale)
+        if res[end] > 0
+            _, res = SmallestInterval.upscale(Qs, res, upscale)
+        end
+        # avoid ArgumentError: The length of the range in dimension 1 (1000) did not equal the size of the interpolation object in that direction (5)
+        Qs = Qsscaled
+    end
 
[…]

### 2017-03-09 real/draft.tex
[draft] Add and describe comparison plot

@@
   \lleq{gmj}
-  p(Q\cond n,\bml)
+  p(Q\cond n, \alpha_0, \beta_0)
   &=     \sum_{N=0}^\infty p(N\cond n) p(Q\cond N, \alpha_0, \beta_0) \, .
@@
   \lleq{gmk}
-  p(Q\cond \lambda,\bml)
+  p(Q\cond \lambda, \alpha_0, \beta_0)
   &=     \sum_{N=0}^\infty p(N\cond \lambda) p(Q\cond N, \alpha_0, \beta_0) \, ,
@@
 In \reffig{comp-unc}, we illustrate the impact of the uncertainty due
-to $\lambda$ or $\alpha, \beta$ for $n=2$ packets and continue to use
-as true values $\alpha_0 = 1.5, \beta_0 = 60$. The standard error
-increases from 0.042 ($\lambda, \alpha, \beta$ fixed) over 0.059
-($\alpha, \beta$ fixed) to 0.11 (nothing fixed).
-%
-\fred{I guess this is for $n=2$? To what is $\lambda$ fixed? Redo the
-  numbers for the actual prior. Add a plot for larger $n$ to show that
-  the Poisson completely dominates}
-%
-The case where all
-parameters are known is not attainable in practice and only shown for
-comparison. But the case where $\alpha, \beta$ are fixed can be reached easily
-if many packets from other bins can be combined. Note that $n=2$ is the
-minimum number of packets to perform the binned analysis.
+to $\lambda$ or $\alpha, \beta$ for $n=0,2, 20$ packets and continue
+to use as true values $\alpha_0 = 1.5, \beta_0 = 60$. For $n>0$, we
+set $\lambda=n$ and for $n=0$, we set $\lambda = 0.347$ by solving
+\begin{align}
+  \lleq{gml}
+  p(N=0 \cond \lambda) &= p(N=0 \cond n=0)
+\end{align}
+such that the contribution at $N=0$ is identical for the Poisson
+likelihood and Poisson posterior predictive distribution.
+
+It is interesting to note that the Jeffreys prior for $\alpha, \beta$
+leads to an extra mode at $\Lumtot = 0$ visible in the case $n=2$. We
+checked that this is true for every individual term in the
+\refeq{gmhb}, and the superposition of $p(\Lumtot \cond N, n, \bml)$
+for various $N$ further enhances this effect. The reason for the
+boundary mode is that $\GammaDist(\Lumtot \cond \alpha, \beta)$ peaks
+at the boundary for $\alpha<1$, and peaks away from $\Lumtot=0$ for
+$\alpha>1$. As $n$ increases, the likelihood concentrates around
+$\alpha=1.5$, hence the distribution becomes unimodal.
+
+Already for $n=20$, it becomes apparent in the right panel of
+\reffig{comp-unc} that the uncertainty for $\Lumtot$ is dominated by
+the Poisson uncertainty: both $p(Q \cond n, \alpha_0, \beta_0)$ and
+$p(Q \cond n, \bml)$ yield a standard error of $\approx 0.19$.
+
+The case where all parameters are known is not attainable in practice
+and only shown for comparison. But the case where $\alpha, \beta$ are
+fixed can be reached easily if many packets from other bins can be
+combined. Note that $n=2$ is the minimum number of packets to perform
+the binned analysis.
 
@@
   \centering
-  \includegraphics[width=0.44\textwidth]{comp_unc_0}
-  \includegraphics[width=0.44\textwidth]{comp_unc_2}
-  \caption{Prediction considering $\lambda, \alpha, \beta$ fixed (green), $\alpha, \beta$ fixed (red), or nothing fixed (blue).}
+  \includegraphics[width=\textwidth]{comp_unc}
+  \caption{Prediction of $\Lumtot$ for $n=0,2,20$ considering $\lambda, \alpha, \beta$ fixed (green), $\alpha, \beta$ fixed (red), or nothing fixed (blue).}
   \label{fig:comp-unc}

### 2017-03-10 real/draft.tex
[draft] Editing unbinned section

@@
 to use as true values $\alpha_0 = 1.5, \beta_0 = 60$. For $n>0$, we
-set $\lambda=n$ and for $n=0$, we set $\lambda = 0.347$ by solving
+set $\lambda=n$ as the mode of the posterior $p(\lambda \cond n)$, and
+for $n=0$, we set $\lambda = 0.347$ by solving
 \begin{align}
@@
 the Poisson uncertainty: both $p(Q \cond n, \alpha_0, \beta_0)$ and
-$p(Q \cond n, \bml)$ yield a standard error of $\approx 0.19$.
+$p(Q \cond n, \bml)$ yield a standard error of $\approx 0.19$. This is
+necessarily larger that the standard error of
+$p(\Lumtot \cond \lambda, \alpha_0, \beta_0)$ because the variance of
 
 The case where all parameters are known is not attainable in practice
-and only shown for comparison. But the case where $\alpha, \beta$ are
-fixed can be reached easily if many packets from other bins can be
-combined. Note that $n=2$ is the minimum number of packets to perform
+and only shown for reference. But the case where $\alpha, \beta$ are
+fixed can be approximated easily if many packets from other bins can
+be combined.
+
+Note that $n=2$ is the minimum number of packets to perform
 the binned analysis.
+\fred{I was playing around in mathematica and it seems even for $n=1$ the posterior has a well defined mode. I don't if it is proper, though. In any case it has huge variance.}
 
@@
 \end{align}
-\fred{We need a plot, the one to compare predictions and 250 real replicas or 100 virtual replicas?} %
 \fred{Advantage of this approach: no need to store all packets anymore, sum and sum of squares can be accumulated online $\Rightarrow$ reduced computation and storage requirements}
-
-\section{Delendum}
-
-\subsection{Numerical implementation} \label{sec:numerical}
-
-\fred{Editing will be done by Fred}
-
-In our numerical implementation, the constants
-$(\alpha_{\rm max}, \beta_{\rm max})$ are chosen by trial and error to
-be large enough to contain the relevant region of
-$p(\ell\cond \alpha,\beta)$.
-%
-To perform the integrals over $(\alpha,\beta)$ to find $C$ and the
-Gamma-weighted numerator in (\ref{spe}), we used the fact that the
-posterior $p(\alpha,\beta \cond n,\bml)$ is approximately Gaussian in
-both $\alpha$ and $\beta$ for large $n$, so we can apply Laplace's
-method for both. This requires the Hessian at the mode and the
-gradient helps in locating the mode. The relevant analytic expressions
-are summarized in \refsec{appendix}.
-
-Regarding the sum over $N$, we start at
-$N = \lfloor n-a+1 \rfloor = \arg \max_N p(N|n)$ %
-\fred{Actually we should say $p(N \cond n, a)$} %
-and continue to larger and smaller values of $N$ until the
-contribution to $p(Q \cond n,\bml)$ becomes negligible. In case the
-relevant contribution is from large $N$ and $n$, we interpret $N$ as a
-continuous variable, replace factorials by $\Gamma$ function in
-\refeq{spj}, and apply Laplace's method to integrate over $N$ also.
 

### 2017-03-10 real/code/TardisPlotUtils.jl
[code] Update uncertainty comp. plot

@@ -433,20 +433,33 @@ function plot_compare_uncertainties(Qs, res, resαβ, resλαβ; nb=2, upscale=2
         Qs = Qsscaled
     end
 
-    plot!(Qs, resλαβ; label="\$\\lambda, \\alpha, \\beta\$ fixed", xlabel="Q", kwargs...)
-    plot!(Qs, resαβ; label="\$\\alpha, \\beta\$ fixed", kwargs...)
-    (res[end] > 0) && plot!(Qs, res; label="nothing fixed", kwargs...)
+    plot!(Qs, resλαβ; label=L"p(Q | \lambda, \alpha_0, \beta_0)", xlabel=L"Q", kwargs...)
+    plot!(Qs, resαβ; label=L"p(Q | n, \alpha_0, \beta_0)", kwargs...)
+    # requires Latex for text in matplotlib, enable in ~/.config/matplotlib/matplotlibrc
+    (res[end] > 0) && plot!(Qs, res; label=L"p(Q | n, \bm{\ell})", kwargs...)
 end
 
 function prepare_compare_uncertainties(kwargs...)
     # clean canvas with two plots next to each other
     plot_kwargs = Dict(:layout=>(1,3), :size=>(900, 300), :legend=>false,
-                       :yticks=>nothing, :grid=>false)
+                       :yticks=>nothing,
+                       :grid=>false)
     plot(;plot_kwargs...)
 
     seed = 61
-    K = 70
+    K = 40
     Qmin = 0.001
+
+    # position of `n=?` label in relative coordinates
+    place_label(nb, subplot; xmax=0.0) = begin
+        # fix ranges after autoscaling
+        sub = Plots.plot!().subplots[subplot]
+        x = sub.attr[:xaxis][:extrema]
+        y = sub.attr[:yaxis][:extrema]
+        xmax = xmax > 0.0? xmax : x.emax
+
+        annotate!([(x.emin + 0.7*(xmax-x.emin), y.emin + 0.2*(y.emax-y.emin), text("\$n=$nb\$", :bottom))]; plot_kwargs...)
+    end
 
     plot_kwargs[:subplot] = 1
     nb = 0
@@ -455,15 +468,18 @@ function prepare_compare_uncertainties(kwargs...)
[…]

### 2017-03-10 real/draft.tex
[draft] tune last subsection

@@
   \lleq{gmj}
-  p(Q\cond n, \alpha_0, \beta_0)
+  p(\Lumtot \cond n, \alpha_0, \beta_0)
   &=     \sum_{N=0}^\infty p(N\cond n) p(Q\cond N, \alpha_0, \beta_0) \, .
@@
   \lleq{gmk}
-  p(Q\cond \lambda, \alpha_0, \beta_0)
+  p(\Lumtot \cond \lambda, \alpha_0, \beta_0)
   &=     \sum_{N=0}^\infty p(N\cond \lambda) p(Q\cond N, \alpha_0, \beta_0) \, ,
@@
 \refeq{gmhb}, and the superposition of $p(\Lumtot \cond N, n, \bml)$
-for various $N$ further enhances this effect. The reason for the
-boundary mode is that $\GammaDist(\Lumtot \cond \alpha, \beta)$ peaks
-at the boundary for $\alpha<1$, and peaks away from $\Lumtot=0$ for
-$\alpha>1$. As $n$ increases, the likelihood concentrates around
-$\alpha=1.5$, hence the distribution becomes unimodal.
-
-Already for $n=20$, it becomes apparent in the right panel of
-\reffig{comp-unc} that the uncertainty for $\Lumtot$ is dominated by
-the Poisson uncertainty: both $p(Q \cond n, \alpha_0, \beta_0)$ and
-$p(Q \cond n, \bml)$ yield a standard error of $\approx 0.19$. This is
-necessarily larger that the standard error of
-$p(\Lumtot \cond \lambda, \alpha_0, \beta_0)$ because the variance of
-
-The case where all parameters are known is not attainable in practice
-and only shown for reference. But the case where $\alpha, \beta$ are
-fixed can be approximated easily if many packets from other bins can
-be combined.
-
-Note that $n=2$ is the minimum number of packets to perform
+for various $N$ further enhances this effect.  The peak of
+$\GammaDist(\Lumtot \cond \alpha, \beta)$ is at $\Lumtot=0$ for
+$\alpha<1$, and at $\Lumtot > 0$ for $\alpha>1$. The Jeffreys prior
+for $\alpha$ gives mass to both regions but as $n$ increases, the
+likelihood concentrates around $\alpha=1.5$, hence the distribution
+becomes unimodal.
+
+For $n=20$, the right panel of \reffig{comp-unc} shows that
+$p(Q \cond n, \bml)$ and $p(Q \cond n, \alpha_0, \beta_0)$ are very
+similar so our reconstruction works correctly. While there is a small
+shift in location, the variances are very similar:
+$\variance{Q \cond n, \alpha_0, \beta_0}=0.036 \approx \variance{Q
+  \cond n, \bml} = 0.034$.  This implies that the uncertainty due to
+$\alpha, \beta$ is almost negligible compared to the effect of the
+Poisson process. The simple explanation is that there is only one
+observation to infer $\lambda$ but $n$ to infer $\alpha, \beta$ so for
+large $n$ the Poisson uncertainty is dominant.
+
+If $\lambda$ is known, the variance is signficantly smaller because
+for the Poisson, $\variance{N \cond \lambda} = \lambda$, and for the
+Negative Binomial $\variance{N \cond n} = 2(n + a-1)$, so for large
+$n$, we have
+\begin{align}
+  \lleq{gmm}
+  \variance{N \cond n} \approx 2 \variance{N \cond \lambda} \, ,
+\end{align}
+and this roughly explains
+$\variance{Q \cond \lambda, \alpha_0, \beta_0 } = 0.021$.  The case
+where all parameters are known is not attainable in practice and only
+shown for comparison. But the case where $\alpha, \beta$ are fixed can
+be reached easily if many packets from other bins can be
+combined. Note that $n=2$ is the minimum number of packets to perform
 the binned analysis.
@@
   \includegraphics[width=\textwidth]{comp_unc}
-  \caption{Prediction of $\Lumtot$ for $n=0,2,20$ considering $\lambda, \alpha, \beta$ fixed (green), $\alpha, \beta$ fixed (red), or nothing fixed (blue).}
+  \caption{Prediction of $\Lumtot$ for $n=0,2,20$ considering
+    $\lambda, \alpha, \beta$ fixed (green), $\alpha, \beta$ fixed
+    (red), or nothing fixed (blue). All densities are properly
+    normalized to one.}
   \label{fig:comp-unc}
@@
   $\bmy$ would mean we use all samples from all bins}
+\fred{We need a reference to the appendix in the main text.}
 
